### 2021-08-31 scripts/R/plot_results.rmd
move sourcing out of functions script, add plotting script, fix Dockerfile

---
title: "Plot results"
author: nadeaus
date: 31.08.21
output: html_notebook
---

Setup
```{r}
require(dplyr)
require(ggplot2)
source("scripts/R/functions/plot_functions.R")
accuracy_simulation_results <- read.delim("output/2021-08-30_estimator_accuracy_MLE.txt")
```

Simulation accuracy heatmap
```{r}
# Summarize results for plotting
error.cols <- c("RMSE.z.h", "RMSE.h.MWA.h")

data.long <- do.call(
  what = tidyr::gather,
  args = c(error.cols, list(data = accuracy_simulation_results, key = "inference.method", value = "error")))

data.summary <- as.data.frame(
  data.long %>%
    dplyr::group_by(inference.method, alpha, H2, var.z, p) %>%
    dplyr::summarise(mean.error = mean(error), sd.error = sd(error), N.reps = dplyr::n()))

# Create facet variable labels for plot
data.summary$inference.method <- factor(
  data.summary$inference.method,
  levels = c("RMSE.h.MWA.h", "RMSE.z.h"),
  labels = c("'Estimated env. part of spVL: '*bolditalic(hat(e))",
             "'Measured spVL, scaled: '*bolditalic(z) - bold(bar(z))"))

# Plot RMSE
rmse_plot <- plotHeatmap(
  df = data.summary,
  N.sim = data.summary$N.reps[[1]],
  is.MLE = T,
  scale.limits = c(0, 1.5),
  legend.name = "RMSE",
  midpoint = 0.75,
  fill.hi = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[1],
  fill.lo = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[3],
  fill.mid = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[2],
  x.var = "alpha",
  y.var = "H2",
  fill.var = "mean.error",
  facet.x = "inference.method",
  facet.y = NA,
  is.oob.squish = F,
  is.real.scale = F,
  labeller = label_parsed) +
  labs(x = expression("Selection strength (" * alpha * ")"),
       y = expression("spVL heritability (" * italic('H'^2) * ")"),
       title = "A") +
  theme(
    text = element_text(size = 8),
    panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
    strip.background = element_rect(colour = "black", size = 0.5))

show(rmse_plot)

ggsave(plot = rmse_plot,
       filename = "manuscript/figures/simulation_RMSE.png",
       width = 4.25,
       height = 2,  # maintain height as 2 for consistent font size in manuscript
       units = "in")
```

### 2021-08-31 scripts/R/functions/POUMM_utility_functions.R

@@ -1,0 +1,152 @@
+calculateSigmaFromParameterization <- function(tree, alpha, H2, var.z) {
+  var.gv <- H2 * var.z
+  t.avg <- mean(POUMM::nodeTimes(tree = tree, tipsOnly = TRUE))
+  sigma <- sqrt((2 * alpha * var.gv) / (1 - exp(-2 * alpha * t.avg)))
+  return(sigma)
+}
+
+permutePOUMMparameters <- function(N.simulations = 1, alphas, sigmas, thetas, 
+                                   sigmaes = NA, H2s = NA, tree = NA, ...) {
+  # Creates parameter dataframe with one row of parameter values per simulation. 
+  # If H2 is specified, sigmae is determined from sigmae. 
+  # If H2 is not specified and sigmae is, H2 is determined from sigmae.  
+  #   Returns dataframe. 
+  checkValidPOUMMParameters(alphas, sigmas, thetas, sigmaes, H2s, tree)
+  if (isDoubleList(H2s)) {
+    param.grid <- makeParamGridWithH2(alphas = alphas, 
+                                      sigmas = sigmas, 
+                                      thetas = thetas, 
+                                      H2s = H2s, 
+                                      tree = tree)
+  } else if (isDoubleList(sigmaes)) {
+    param.grid <- makeParamGridWithSigmae(alphas = alphas, 
+                                          sigmas = sigmas, 
+                                          thetas = thetas, 
+                                          sigmaes = sigmaes, 
+                                          tree = tree)
+  } else {
+    param.grid <- expand.grid(alphas = alphas, 
+                              sigmas = sigmas, 
+                              thetas = thetas,
+                              sigmaes = sigmaes,
+                              H2s = H2s)
+  }
+  param.grid <- do.call(what = "rbind", 
+                        args = replicate(n = N.simulations, 
+                                         expr = param.grid, 
+                                         simplify = FALSE))
+  return(param.grid)
+}
+
+checkValidPOUMMParameters <- function(alphas, sigmas, thetas, H2s, sigmaes, tree, ...) {
+  if (!isDoubleList(alphas)) {
+    stop("alphas must be a vector of doubles with at least one entry.")
[…]

### 2021-08-31 scripts/R/functions/plot_functions.R

@@ -1,0 +1,373 @@
+chooseErrorFunction <- function(ERROR_METRIC) {
+  # Returns function for error metric calculation.
+  if (ERROR_METRIC == "MAPE__") {
+    return(calculateMAPE)
+  } else if (ERROR_METRIC == "RMSE") {
+    return(calculateRMSE)
+  } else if (ERROR_METRIC == "MBRAE") {
+    return(calculateMBRAE)
+  } else if (ERROR_METRIC == "UMBRAE") {
+    return(calculateUMBRAE)
+  } else {
+    stop("Unknown error metric.")
+  }
+}  # TODO: implement a scale-invariant error metric (sMAPE?)
+
+addNextHighestValueToDataFrame <- function(df, colname) {
+  values <- unique(df[, colname])
+  if (length(values) < 2) {
+    stop("Attempting to add list of values with all the same value.")
+  }
+  nextcol <- paste(colname, "next", sep=".")
+  df[, nextcol] <- unlist(lapply(df[, colname], getNextHighestValueInList, values))
+  return(df)
+}
+
+getNextHighestValueInList <- function(value, list) {
+  list <- sort(list, decreasing = F)
+  next.highest.value <- list[which(list > value)[1]]
+  if (is.na(next.highest.value)) {
+    next.highest.value <- max(list) + max(list) - list[length(list) - 1]
+  }
+  return(next.highest.value)
+}
+
+plotHeatmap <- function(df, plotname = NULL, N.sim, is.MLE,
+                        scale.limits, legend.name, fill.lo, fill.hi,
+                        fill.mid = NA, midpoint = NA, x.var, y.var, fill.var,
+                        facet.x = NA, facet.y = NA, is.oob.squish = T,
+                        is.real.scale = T, labeller = plot_labeller) {
+  if (!(is.na(facet.x)) | !(is.na(facet.y))) {
+    facet.spec <- constructFacettingSpec(facet.x, facet.y, labeller)
+    no_facet <- F
+  } else {
+    no_facet <- T
+  }
+  oob.choice <- getOOBFunction(is.oob.squish)
+  color.scale <- constructColorScale(fill.lo, fill.hi, legend.name, scale.limits, oob.choice, fill.mid, midpoint)
+
+  # get tile bounds
+  if(is.real.scale) {
+    df <- addNextHighestValueToDataFrame(df = df, colname = x.var)
[…]

### 2021-08-31 scripts/R/functions/simulate_epidemic.R

@@ -24,9 +24,6 @@
 # TODO: p-values dataframe has duplicated columns from merge
 # ==============================================================================
 
-source("functions/POUMM_utility_functions.R")
-source("functions/utility_functions.R")
-
 simulateEpidemic <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delta = NA, H2, H2.h, var.z, is.MLE = T) {
   # Simulate data for GWAS
   sigma <- calculateSigmaFromParameterization(

### 2021-08-31 scripts/R/simulate_for_estimator_accuracy.R

@@ -9,12 +9,13 @@ require(yaml)
 source("scripts/R/functions/simulate_epidemic.R")
 source("scripts/R/functions/simulation_control_functions.R")
 source("scripts/R/functions/utility_functions.R")
+source("scripts/R/functions/POUMM_utility_functions.R")
 
 # Set seed for replicability of simulation
 set.seed(1)
 
 # Read parameters for data simulation from config file
-config_values <- yaml::read_yaml(file = "config-simulation.yaml")
+config_values <- yaml::read_yaml(file = "../../config-simulation.yaml")
 N <- config_values$N
 
 # Designate output file

### 2021-08-31 scripts/R/plot_results.rmd
add GWAS TPR simulation code, runnable in docker

@@
 source("scripts/R/functions/plot_functions.R")
-accuracy_simulation_results <- read.delim("output/2021-08-30_estimator_accuracy_MLE.txt")
+accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
+
+View(accuracy_simulation_results %>% filter(alpha == 5))
 ```
@@
   is.MLE = T,
-  scale.limits = c(0, 1.5),
+  scale.limits = c(0, 1),
   legend.name = "RMSE",
-  midpoint = 0.75,
+  midpoint = 0.5,
   fill.hi = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[1],

### 2021-08-31 scripts/R/functions/GWAS_utility_functions.R

@@ -1,0 +1,84 @@
+summarizeAssociationResults <- function(epidemic) {
+  # Calculate p-values for association tests
+  h.MWA.p.values = runAssociationTests(
+    pred.vars = epidemic$simulated.h.values$genotypes,
+    response.var = epidemic$inferred.h.values$h.MWA)
+  h.OU.p.values = runAssociationTests(
+    pred.vars = epidemic$simulated.h.values$genotypes,
+    response.var = epidemic$inferred.h.values$h.OU)
+  h.plus.e.p.values = runAssociationTests(
+    pred.vars = epidemic$simulated.h.values$genotypes,
+    response.var = epidemic$simulated.h.values$h + epidemic$simulated.e.values$e)
+  norm.z.p.values = runAssociationTests(
+    pred.vars = epidemic$simulated.h.values$genotypes,
+    response.var = epidemic$z - mean(epidemic$z))
+
+  # make p-value dataframe
+  is.causal <- getCausalVariants(variant.names = epidemic$simulated.h.values$variant.names)
+  p.values <- data.frame(
+    variant = epidemic$simulated.h.values$variant.names,
+    is.causal.variant = is.causal,
+    h.MWA.p.value = h.MWA.p.values,
+    h.OU.p.value = h.OU.p.values,
+    h.plus.e.p.value = h.plus.e.p.values,
+    norm.z.p.value = norm.z.p.values)
+  return(p.values)
+}
+
+runAssociationTests <- function(pred.vars, response.var) {
+  # Returns a vector of p-values, 1 for each column in the pred.vars dataframe.
+  p.values <- apply(pred.vars, 2, fitLinearModel, response.var = response.var)
+  return(as.vector(p.values))
+}
+
+fitLinearModel <- function(pred.var, response.var) {
+  # Estimates the linear slope coefficient, returns p-value for it.
+  linear.model <- lm(response.var ~ pred.var)
+  linear.model.coefficients <- summary(linear.model)$coefficients
+  if (dim(linear.model.coefficients)[1] == 1) {
+    warning("No linear model p value estimated if response variable constant.")
+    p.value <- NA
+  } else {
+    p.value <- linear.model.coefficients[2, "Pr(>|t|)"]
+  }
+  return(p.value)
+}
+
+calculateGWASpower <- function(p.values, causal.variants = NA, alpha) {
+  if (!(is.na(causal.variants))) {
+    p.values <- p.values[causal.variants]
+  }
[…]

### 2021-08-31 scripts/R/functions/PLINK_utility_functions.R

@@ -1,0 +1,98 @@
+generateMapFile <- function(variant.names, K, out, filename) {
+  # generate dummy map dataframe for plink, write out to file
+  map <- data.frame(
+    CHR = rep(1, length.out = K),
+    variantID = variant.names,
+    POS = rep(0, length.out = K))
+  write.table(
+    map, file = paste(out, paste(filename, "map", sep = "."), sep = "/"),
+    quote = F, sep = " ", row.names = F, col.names = F)
+}
+
+generatePedFile <- function(G, N, out, filename) {
+  # generate ped dataframe with dummy phenotype
+  ped <- list(
+    FID = rep(1, length.out = N),
+    IID = 1:N,
+    IID_father = rep(0, length.out = N),
+    IID_mother = rep(0, length.out = N),
+    sex = rep(0, length.out = N),
+    phenotype = rep(-9, length.out = N))
+  
+  # add causal variant calls to ped dataframe
+  for (j in 1:dim(G)[2]) {
+    allele.1 <- rep("N", N) 
+    allele.1[G[, j] > 0] <- "Y"
+    allele.2 <- rep("N", N)
+    allele.2[G[, j] == 2] <- "Y"
+    ped[[6 + 2 * j - 1]] <- allele.1
+    ped[[6 + 2 * j]] <- allele.2
+  }
+  attributes(ped) <- list(row.names=c(NA_integer_, N),
+                        class="data.frame",
+                        names=make.names(names(ped),
+                                         unique=TRUE))  # for memory efficient coversion to df
+  write.table(
+    ped, file = paste(out, paste(filename, "ped", sep = "."), sep = "/"),
+    quote = F, sep = " ", row.names = F, col.names = F)
+}
+
+generatePhenoFile <- function(epidemic, N, out, filename) {
+  # generate pheno file with all phenotypes (h estimators)
+  pheno <- data.frame(
+    FID = rep(1, length.out = N),
+    IID = 1:N,
+    z = epidemic$z - mean(epidemic$z),
+    h.OU = epidemic$inferred.h.values$h.OU,
+    h.MWA = epidemic$inferred.h.values$h.MWA,
+    h.plus.e = epidemic$simulated.h.values$h + epidemic$simulated.e.values$e)
+  write.table(
+    pheno, file = paste(out, paste(filename, "pheno", sep = "."), sep = "/"),
+    quote = F, sep = " ", row.names = F, col.names = T)
+}
+
[…]

### 2021-08-31 scripts/R/functions/simulation_control_functions.R

@@ -1,40 +1,5 @@
 # ==============================================================================
 # Control functions for testing GWAS methods on simulated data
-
-### FOR TESTING:
-# N <- 50
-# tree <- ape::rtree(N)
-# g0 <- 4
-# alpha <- 30
-# theta <- 4
-# sigma <- 5
-# M <- 10
-# K <- 100
-# d <- 2
-# p <- 0.2
-# delta <- 0.2
-# var.z <- 0.6
-# H2 = 0.75
-# is.MLE = T
-# plink.path = "/Users/nadeaus/Documents/PLINK/plink"
-# out = "~/Downloads"
-# chunk.size <- 10
-# filename <- "test"
-
-# N = 500
-# g0 = 4.5
-# alpha = 0.1
-# theta = 4.5
-# M = 20
-# K = 20
-# d = 2
-# p = 0.05
-# H2 = 0.75
-# H2.h = 0.25
-# var.z = 0.5
-# N.reps = 20
-# is.MLE <- T
-
 # ==============================================================================
 
 addSimulationParamsToResults <- function(df, epidemic, g0, alpha, theta, K, M, H2, H2.h, var.z, is.MLE) {
@@ -140,7 +105,7 @@ simulateVHEZValues <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delta
 }
 
 simulateVEHZValuesError <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delta = NA, H2, H2.h, var.z, is.MLE = T) {
-  print(paste0("Simulating data and fitting POUMM with these parameters: alpha=", alpha, ", H2=", H2, ", H2.h=", H2.h, ", var.z=", var.z))
+  print(paste0("Simulating for accuracy with these parameters: alpha=", alpha, ", H2=", H2, ", H2.h=", H2.h, ", var.z=", var.z))
   # Simulate data
   epidemic <- simulateEpidemic(
     tree = tree,
@@ -209,7 +174,6 @@ simulateGWASPValuesWithPLINK <- function(tree, N, g0, alpha, theta, M, K, d, p =
   } else {
     runID <- "NA"
   }
-  source("functions/PLINK_utility_functions.R")
 
   # Simulate data
   epidemic <- simulateEpidemic(
@@ -292,8 +256,7 @@ simulateGWASPValuesWithPLINK <- function(tree, N, g0, alpha, theta, M, K, d, p =
 }
 
[…]

### 2021-08-31 scripts/R/functions/utility_functions.R

@@ -39,14 +39,9 @@ writeParametersToStdOut <- function(params) {
 
 makeDataFrameFromMapplyResults <- function(mapply.results) {
   dataframe <- as.data.frame(t(as.data.frame(mapply.results)))
+  cols_to_unnest <- colnames(dataframe)
   dataframe$runID <- 1:dim(dataframe)[1]
-  dataframe <- tidyr::unnest(dataframe, cols = c(
-    p, sigma, RMSE.z.v, RMSE.v.MWA.v, RMSE.v.OU.v, MAPE.z.v, MAPE.v.MWA.v,
-    MAPE.v.OU.v, UMBRAE.v.MWA.z, UMBRAE.v.OU.z, RMSE.z.h, RMSE.h.MWA.h,
-    RMSE.h.OU.h, MAPE.z.h, MAPE.h.MWA.h, MAPE.h.OU.h, UMBRAE.h.MWA.z,
-    UMBRAE.h.OU.z, g0.AVG, alpha.MLE, sigma.MLE, theta.MLE, sigmae.MLE,
-    g0, alpha, theta, sigmae, K, M, delta, H2.h, H2, var.z, empirical.H2,
-    empirical.H2.h))
+  dataframe <- tidyr::unnest(dataframe, cols = all_of(cols_to_unnest))
   return(dataframe)
 }
 

### 2021-08-31 scripts/R/simulate_for_gwas_tpr.R

@@ -1,0 +1,107 @@
+# ==============================================================================
+# Test GWAS TPR improvement using MWA estimation method.
+# ==============================================================================
+
+# To run: Rscript simulate_for_gwas_tpr.R"
+
+# Source necessary functions
+source("scripts/R/functions/simulate_epidemic.R")
+source("scripts/R/functions/simulation_control_functions.R")
+source("scripts/R/functions/utility_functions.R")
+source("scripts/R/functions/plot_functions.R")
+source("scripts/R/functions/PLINK_utility_functions.R")
+source("scripts/R/functions/POUMM_utility_functions.R")
+source("scripts/R/functions/GWAS_utility_functions.R")
+
+# Set seed for replicability of simulation
+set.seed(1)
+
+# Read parameters for data simulation from config file
+# config_values <- yaml::read_yaml(file = "../../config-simulation.yaml")
+config_values <- yaml::read_yaml(file = "config-simulation.yaml")
+N <- config_values$N
+
+# Designate output file
+outfile <- generateOutfileName(outdir = "output", description = "GWAS_TPR_MLE")
+
+# Simulate HIV phylogeny
+tree <- simulateHIVTreeExpBL(N)
+
+# Set constant parameters for data simulation
+param.list <- list(
+  N = N,
+  g0 = config_values$g0,
+  alpha = config_values$alpha,
+  theta = config_values$theta,
+  M = config_values$M,
+  K = config_values$K,
+  d = config_values$d,
+  delta = config_values$delta,
+  H2 = config_values$H2,
+  H2.h = config_values$H2.h,
+  var.z = config_values$var.z,
+  N.reps = config_values$N.reps)
+
+param.df <- permuteParameters(
+  N.reps = param.list$N.reps,
+  parameterlist = param.list)
+
+# Simulate data and calculate estimator error
+data.list <- mapply(
+  simulateGWASPValues,
+  N = param.df$N,
+  g0 = param.df$g0,
+  alpha = param.df$alpha,
+  theta = param.df$theta,
+  M = param.df$M,
+  K = param.df$K,
+  d = param.df$d,
+  delta = param.df$delta,
+  H2 = param.df$H2,
+  H2.h = param.df$H2.h,
+  var.z = param.df$var.z,
+  MoreArgs = list(tree = tree,
+                  is.MLE = config_values$is.MLE))
+
+# Clean up and save simulation results
[…]

### 2021-09-28 scripts/R/plot_results.rmd
migrate code to make full figure 1, add figures in manuscript directory

@@
 ```{r}
+knitr::opts_knit$set(root.dir = "/Users/nadeaus/Repos/poumm-gwas")
+```
+
+```{r}
 require(dplyr)
@@
 source("scripts/R/functions/plot_functions.R")
+source("scripts/R/functions/utility_functions.R")
 accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
-
-View(accuracy_simulation_results %>% filter(alpha == 5))
+tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
 ```
@@
 ```
+
+Simulation accuracy heatmap annotated with inferred POUMM values for HIV spVL
+```{r}
+rmse_plot_with_hiv_values <- rmse_plot +
+  annotate(
+    geom = "point",
+    x = 8.046, y = 3.9)  # values are hand-calculated because the breaks scale is defined as seq(0.5, n.breaks - 0.5, 1) in plotting function
+
+show(rmse_plot_with_hiv_values)
+
+ggsave(plot = rmse_plot_with_hiv_values,
+       filename = "manuscript/figures/simulation_RMSE_with_HIV_value_highlighted.png",
+       width = 4.25,
+       height = 2,  # maintain height as 2 for consistent font size in manuscript
+       units = "in")
+```
+
+Simulation TPR heatmap
+```{r}
+data <- tidyr::gather(
+  data = tpr_simulation_results,
+  key = "inference.method",
+  value = "p.value",
+  h.plus.e.p.value, norm.z.p.value, h.OU.p.value, h.MWA.p.value)
+
+# Calculate TPR for each simulation
+sig.level <- correctSignificanceLevelForMultTesting(
+  significance.level = 0.05,
+  significance.correction = "Bonferroni",
+  N.tests = 20)
+data$TP <- ifelse(
+  data$p.value < sig.level & data$is.causal.variant, T, F)
+data$FP <- ifelse(
+  data$p.value < sig.level & !(data$is.causal.variant), T, F)
+data.TPR <- as.data.frame(
+  data %>%
+    dplyr::group_by(runID, inference.method, alpha, sigma, H2, var.z, p, delta) %>%
+    dplyr::summarise(
+      TPR = sum(TP, na.rm = T)/sum(is.causal.variant),
+      FP = sum(FP), K = dplyr::n(),
+      N.noncausal.variants = sum(!(is.causal.variant))))
+
+# Summarize replicate simulations
+data.summary <- as.data.frame(
+  data.TPR %>%
+    dplyr::group_by(inference.method, alpha, sigma, H2, var.z, p, delta) %>%
+    dplyr::summarise(
+      mean.TPR = mean(TPR),
+      sd.TPR = sd(TPR),
+      mean.FP = mean(FP),
+      sd.FP = sd(FP),
+      N.noncausal.variants = mean(N.noncausal.variants),
+      N.sim = dplyr::n()))
+
+# Create facet variable labels for plot
+data.summary$inference.method <- factor(
+  data.summary$inference.method,
+  levels = c("h.plus.e.p.value", "h.MWA.p.value", "h.OU.p.value", "norm.z.p.value"),
+  labels = c("e", "italic(hat(e))", "hat(e)[OU]", "'scaled spVL'"))
+
+# Plot TPR 
+tpr_plot <- makeTPRPlot(
+  df = data.summary[
+    data.summary$inference.method %in% 
+      c("e", "italic(hat(e))", "'scaled spVL'"), ],
+  is.MLE = T,
+  midpoint = 0.5,
+  fill.lo = RColorBrewer::brewer.pal(name = 'RdBu', n = 3)[3],
+  fill.hi = RColorBrewer::brewer.pal(name = 'RdBu', n = 3)[1],
+  fill.mid = RColorBrewer::brewer.pal(name = 'RdBu', n = 3)[2]) +
+  facet_grid(. ~ inference.method,
+             labeller = label_parsed) +
+  labs(x = expression("Selection strength (" * alpha * ")"),
+       y = expression("spVL heritability (" * 'H'^2 * ")"),
+       title = "B") +
+  theme(
+    text = element_text(size = 8),
+    panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
+    strip.background = element_rect(colour = "black", size = 0.5))
+
+show(tpr_plot)
+
+ggsave(plot = tpr_plot,
+       filename = "manuscript/figures/simulation_TPR.png",
+       width = 5.25,
+       height = 2,  # maintain height as 2 for consistent font size in manuscript
+       units = "in")
+```
+TPR heatmap annotated with inferred POUMM values for HIV spVL
+```{r}
+tpr_plot_with_hiv_values <- tpr_plot +
+annotate(
+geom = "point",
+x = 8.046, y = 3.9)  # values are hand-calculated because the breaks scale is defined as seq(0.5, n.breaks - 0.5, 1) in plotting function
+
+show(tpr_plot_with_hiv_values)
+
+ggsave(plot = tpr_plot_with_hiv_values,
+filename = "manuscript/figures/simulation_TPR_with_HIV_value_highlighted.png",
+width = 5.25,
+height = 2,  # maintain height as 2 for consistent font size in manuscript
+units = "in")
+```
+
+Combined figure
+```{r}
+png(
+  filename = "manuscript/figures/simulation_results.png",
+  width = 5, height = 4, unit = "in", res = 300)
+gridExtra::grid.arrange(
+  rmse_plot_with_hiv_values, tpr_plot_with_hiv_values,
+  nrow = 2,
+  layout_matrix = rbind(c(NA, 1, 1, 1),
+                        c(2, 2, 2, 2)))
+dev.off()
+```

### 2021-10-12 scripts/bash/build-tree.sh
add scripts to build container for tree generation

@@ -1,0 +1,29 @@
+#!/bin/bash
+# This script generates an alignment, builds a tree.
+
+# Data file is newfasta2019-10-25.fas which contains 2012 pol sequences corresponding to 2012/2125 genotyped samples (the remaining host genotypes are not associated with a pol sequence)
+# The sequences are not aligned (some are longer than others) even though gap characters have been introduced --> what does this mean??
+
+# Generate alignment from fasta file
+### PUT IN RIGHT ALIGNMENT!
+./muscle3.8.31_i86linux64 \
+-in sequences.fasta \
+-out output/pathogen.fasta \
+-fasta \
+-maxiters 3
+
+# Build a ML tree with the GTR+F+R4 model and 100 bootstrap replicates
+iqtree \
+-s output/pathogen.fasta \
+-m GTR+F+R4 \
+#-b 100 \
+-pre output/pathogen \
+-nt AUTO
+
+# Hodcroft stripped sequences of codons in positions associated with drug resistance mutations (but no sig. differences to when these left in apparently)
+# 38 reference pol sequences subtypes A-K from Los Alamos HIV database used as outgroup
+# Used RAxML to build the tree, RAxML default model is hard to tell from quick look at documentation
+# Used 100 bootstrap replicates
+
+
+

### 2021-10-26 scripts/R/calculate_spvl.R
Add calculate spVL script

@@ -1,0 +1,131 @@
+# This script is to calcuate spVL for the SHCS data in several ways.
+#   * liberal filter = exclude meaurements after treatment
+#   * strict filter = exclude measurements possibly < 6 mo. after infection and after treatment or AIDS
+#   * liberal 1: liberal filter --> take first of remaining measurements 
+#   * liberal 2: liberal filter --> take mean of remaining measurements
+#   * strict 1: strict filter  --> take mean of remaining measurements
+# Note that spVL measured to be 0 is replaced with log10(25)
+
+# Run as: Rscript scripts/R/calculate_spvl.R
+
+require(dplyr)
+require(ggplot2)
+require(lubridate)
+require(tidyr)
+
+OUTDIR <- "output"
+DATADIR <- "/Volumes/stadler/SHCSData/data"
+LAB_FN <- "LAB_rna_measurement.csv"
+TAILOR_1_FN <- "TAILOR_rna_measurement.csv"
+TAILOR_2_FN <- "RNA_measurement_additional_info.csv"
+
+# Load data --------------------------------------------------------------------
+print("loading data")
+lab_data_raw <- read.table(paste(DATADIR, LAB_FN, sep = "/"), sep = ",", header = T)
+tailor_data1 <- read.table(paste(DATADIR, TAILOR_1_FN, sep = "/"), sep = ",", header = T)
+tailor_data2 <- read.table(paste(DATADIR, TAILOR_2_FN, sep = "/"), sep = ",", header = T)
+tailor_data_raw <- merge(x = tailor_data1, y = tailor_data2)
+
+# Parse dates -----------------------------------------------------------------
+print("Parsing dates.")
+
+tailor_data <- tailor_data_raw %>%
+  mutate(seroc_date = as.Date(seroc_date, format = "%d.%m.%y"),
+         haart_start_date = as.Date(haart_start_date, format = "%d.%m.%y"),
+         tri_start_date = as.Date(tri_start_date, format = "%d.%m.%y"),
+         first_c_date = as.Date(first_c_date, format = "%d.%m.%y"),
+         art_start_date = as.Date(art_start_date, format = "%d.%m.%y"),
+         cd4_200_fd = as.Date(cd4_200_fd, format = "%d.%m.%y"),
+         regdate = as.Date(regdate, format = "%d.%m.%y"),
+         hiv_posdate = as.Date(hiv_posdate, format = "%d.%m.%y"),
[…]

### 2021-10-26 scripts/R/calculate_spvl.R
add script to filter alignment, add spVL and outgroup info to sequence headers

@@ -117,7 +117,7 @@ spvl_values <- rbind(
 names_from = spvl_method,
 values_from = spVL)
 
-write.csv(x = spvl_values, file = paste(OUTDIR, "spvl_values.csv", sep = "/"))
+write.csv(x = spvl_values, file = paste(OUTDIR, "spvl_values.csv", sep = "/"), row.names = F)
 
 p <- ggplot(data = spvl_values,
        aes(x = spvl_lenient_filter_first_measurement,

### 2021-10-26 scripts/R/filter_alignment.R

@@ -1,0 +1,94 @@
+# This script is to filter the HIV pol sequence alignment to focal and outgroup subtypes only, annotate outgroup sequences
+# Data file is newfasta2019-10-25.fas which contains 2012 pol sequences corresponding to 2012/2125 genotyped samples (the remaining host genotypes are not associated with a pol sequence)
+# The sequences are not aligned (some are longer than others) even though gap characters have been introduced --> what does this mean??
+# Dates in the fasta file correspond to "sampledate" in the subtyping data.
+# Run as: Rscript scripts/R/filter_alignment.R
+
+library(dplyr)
+library(ape)
+library(lubridate)
+library(tidyr)
+
+OUTDIR <- "../../output"
+DATADIR <- "/Volumes/stadler/SHCSData/data"
+FASTA_FN <- "newfasta2019-10-25.fas"
+SPVL_FN <- "spvl_values.csv"
+OUTGROUP_TYPE <- "A"
+
+paste("Loading subtype data.")
+subtypes_raw <- read.delim(
+  file = paste(DATADIR, "subtype.csv", sep = "/"),
+  header = T,
+  sep = ",",
+  colClasses = rep("character", 6))
+subtypes <- subtypes_raw %>%
+  mutate(sampledate = as.Date(sampledate, format = "%d.%m.%y"),
+         seq_dt = as.Date(seq_dt, format = "%d.%m.%y")) %>%
+  select(id, sampledate, subtype)
+
+paste("Loading spVL data.")
+spvl_values <- read.csv(file = paste(OUTDIR, SPVL_FN, sep = "/"))
+
+paste("Loading sequence file.")
+sequences <- ape::read.FASTA(file = paste(DATADIR, FASTA_FN, sep = "/"))
+
+# Parse sequence names to get patient IDs, collection dates
+temp <- strsplit(x = names(sequences), split = "\\|")
+temp2 <- matrix(unlist(temp), ncol = 3, byrow = T)
+seq_data <- data.frame(orig_label = names(sequences), id = temp2[, 2], sampledate = temp2[, 3])
+seq_data$sampledate <- as.Date(seq_data$sampledate)
+
+# Match sequence data to subtype data
+tip_data <- merge(
+  x = seq_data, y = subtypes,
+  by = c("id", "sampledate"),
+  all.x = T, all.y = F) %>%
+  filter(subtype != "No sequence") %>%  # don't consider failed sequences
[…]

### 2021-10-26 scripts/bash/build-tree.sh
redo code for tree building, add documentation

@@ -1,11 +1,8 @@
 #!/bin/bash
 # This script generates an alignment, builds a tree.
-
-# Data file is newfasta2019-10-25.fas which contains 2012 pol sequences corresponding to 2012/2125 genotyped samples (the remaining host genotypes are not associated with a pol sequence)
-# The sequences are not aligned (some are longer than others) even though gap characters have been introduced --> what does this mean??
+# Run with: bash /scripts/bash/build-tree.sh
 
 # Generate alignment from fasta file
-### PUT IN RIGHT ALIGNMENT!
 ./muscle3.8.31_i86linux64 \
 -in sequences.fasta \
 -out output/pathogen.fasta \

### 2021-10-26 scripts/bash/build-tree.sh
remove bootstrapping for speed, for now

@@ -9,11 +9,11 @@
 -fasta \
 -maxiters 3
 
-# Build a ML tree with the GTR+F+R4 model and 100 bootstrap replicates
+# Build a ML tree with the GTR+F+R4 model and (TODO) 100 bootstrap replicates
 iqtree \
 -s output/pathogen.fasta \
 -m GTR+F+R4 \
-#-b 100 \
+-redo \
 -pre output/pathogen \
 -nt AUTO
 

### 2021-10-26 scripts/R/filter_alignment.R
add filtering for subtype b-carrying individuals of european descent

@@ -1,4 +1,4 @@
-# This script is to filter the HIV pol sequence alignment to focal and outgroup subtypes only, annotate outgroup sequences
+# This script is to filter the HIV pol sequence alignment to focal and outgroup subtypes only, annotate sequences with date, trait value and focal/outgroup status.
 # Data file is newfasta2019-10-25.fas which contains 2012 pol sequences corresponding to 2012/2125 genotyped samples (the remaining host genotypes are not associated with a pol sequence)
 # The sequences are not aligned (some are longer than others) even though gap characters have been introduced --> what does this mean??
 # Dates in the fasta file correspond to "sampledate" in the subtyping data.

### 2021-10-26 scripts/R/filter_gwas_individuals.R

@@ -1,0 +1,79 @@
+# This script is to filter individuals for GWAS to those of European ancestry, carrying subtype B HIV.
+# Run as: Rscript scripts/R/filter_gwas_individuals.R
+
+library(dplyr)
+library(gridExtra)
+
+OUTDIR <- "output"
+DATADIR <- "/Volumes/stadler/SHCSData/data"
+
+# Load principle components data
+pc_scores <- read.delim(
+  file = paste(DATADIR, "spVL.shcs.rs.hm3.merge.SNPqc.mds", sep = "/"),
+  header = T,
+  sep = "") %>%
+  mutate(Population = case_when(
+    Pop == "Test" ~ "Swiss HIV Cohort Study",
+    Pop == "YRI" ~ "Yoruba in Ibadan, Nigeria (YRI)",
+    Pop == "JPT+CHB" ~ "Japanese in Tokyo, Japan (JPT) +\nHan Chinese in Beijing, China (CHB)",
+    Pop == "CEU" ~ "CEPH/Utah Collection (CEU) ",
+    Pop == "MKK" ~ "Maasai in Kinyawa, Kenya (MKK)",
+    Pop == "LWK" ~ "Luhya in Webuye, Kenya (LWK)",
+    Pop == "CHD" ~ "Chinese in Metropolitan Denver, CO, USA (CHD)",
+    Pop == "GIH" ~ "Gujarati Indians in Houston, TX, USA (GIH)",
+    Pop == "TSI" ~ "Toscani in Italia (TSI)",
+    Pop == "MEX" ~ "Mexican Ancestry in LA, CA, USA (MXL)",
+    Pop == "ASW" ~ "African Ancestry in SW USA (ASW)",
+    T ~ Pop),
+  is_european = PC1 > 0.01 & PC1 < 0.025
+    & PC2 > 0.035 & PC2 < 0.05
+    & PC3 > -0.03 & PC3 < 0.03)
+
+# Load subtype data from sequence metadata
+seq_metadata <- read.csv(file = paste(OUTDIR, "seq_metadata.csv", sep = "/"))
+
+gwas_metadata <- merge(
+  x = seq_metadata, y = pc_scores,
+  by.x = "id", by.y = "IID",
+  all = T) %>%
+  filter(subtype == "B" | Pop != "Test")
+
+# Plot combined SHCS and HapMap samples along the first few PCs
+pop_colors <- RColorBrewer::brewer.pal(n = 11, name = "Set3")
[…]

### 2021-10-27 hiv-gwas.tex
Update on Overleaf.

\documentclass[]{article}
\usepackage[margin=1in]{geometry}
\usepackage{biblatex}
\addbibresource{references.bib}
\usepackage{graphicx}
\usepackage{bm} % for bold math characters
\usepackage{booktabs} % for prettier hlines in table
\usepackage{subcaption} % for subfigures
\usepackage[affil-it]{authblk} % for footnote-style author affiliation
\usepackage{setspace} % for 1.5-spaced lines not including footnotes, floats
\usepackage{tikz} % for graphical model drawing
\usepackage{float} % to enforce figure positioning in subsection they're declared in
\usepackage{amsmath}
\usepackage{longtable} % for table that spans multiple pages

% Tikz preamble from Michael Landis
\usepackage{amssymb} %maths
\usepackage{amsmath} %maths
%\usepackage{subfig} % can't be used together with subcaption package
\usepackage{skull}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{fit}
\usetikzlibrary{calc}
\usetikzlibrary{automata}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{decorations.pathreplacing} % for curly bracehttps://www.overleaf.com/project/6127a9ffb8327d4ab8b9fd4d
\usetikzlibrary{shapes.multipart}
\tikzset{>=latex}
\tikzstyle{snode}=[black,draw=black,line width=1.5pt,shape=circle,fill=white,minimum size=8mm]
\tikzstyle{obnode}=[black,draw=black,line width=1.5pt,shape=circle,fill=black!20!white,minimum size=8mm]
\tikzstyle{detnode}=[black,draw=black,line width=1.5pt,densely dotted,shape=circle,fill=white,minimum size=8mm]
\tikzstyle{constnode}=[black,draw=black,line width=1.5pt,shape=rectangle,fill=white,minimum size=8mm]
\tikzstyle{blnode}=[white,draw=black,line width=1pt,shape=circle,fill=black,minimum size=1mm,font=\scriptsize,inner sep=1pt]
\tikzstyle{ylnode}=[black,draw=black,line width=1pt,shape=circle,fill=yellow,minimum size=1mm,font=\scriptsize,inner sep=1pt]
\tikzstyle{taro}=[->,line width=2pt,color=black]
\tikzstyle{baro}=[<->,line width=2pt,color=black]
\tikzstyle{bline}=[line width=2pt,color=black]
\tikzstyle{dtaro}=[->,line width=2pt, densely dotted,color=black]
\tikzstyle{dline}=[line width=2pt, densely dotted,color=black]
\tikzstyle{smod}=[black, draw=black, line width=2pt, fill=white, shape=rectangle, rounded corners, minimum size=10mm, minimum width=20mm]
\tikzstyle{obmod}=[black, draw=black, line width=2pt, fill=black!20!white, shape=rectangle, rounded corners, minimum size=10mm, minimum width=20mm, minimum width=20mm]

% Set up supplemental figure numbering
\newcommand{\beginsupplement}{%
	\setcounter{table}{0}
	\renewcommand{\thetable}{S\arabic{table}}%
	\setcounter{figure}{0}
	\renewcommand{\thefigure}{S\arabic{figure}}%
}

% For proof
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}

% for colored rows in table
\usepackage{color, colortbl}
\definecolor{Gray}{gray}{0.9}

% For multi-line table cells
\newcommand{\specialcell}[2][c]{\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\begin{document}

\begin{center}
	\Large{A new phylogeny-aware GWAS framework to estimate direct human genetic effects on HIV set-point viral load}
\end{center}
\vspace{4pt}

\begin{flushleft}
	Sarah Nadeau\textsuperscript{1,2}, Christian Thorball\textsuperscript{2,3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F. Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Manuel Battegay\textsuperscript{10}, Matthias Cavassini\textsuperscript{11}, Alexandra Calmy\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
\end{flushleft}

\noindent
\textsuperscript{1}\emph{Department of Biosystems Science and Engineering, ETH Zürich, Basel} \\
\textsuperscript{2}\emph{Swiss Institute of Bioinformatics, Lausanne} \\
\textsuperscript{3}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
\textsuperscript{4}\emph{Institute of Medical Virology, University of Zürich, Zürich} \\
\textsuperscript{5}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zürich, University of Zürich, Zürich} \\
\textsuperscript{6}\emph{Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva} \\
\textsuperscript{7}\emph{Division of Immunology and Allergy, University Hospital Lausanne, Lausanne} \\
\textsuperscript{8}\emph{Molecular Virology, Department of Biomedicine - Petersplatz, University of Basel, Basel} \\
\textsuperscript{9}\emph{Department of Infectious Diseases, Berne University Hospital and University of Berne, Berne} \\
\textsuperscript{10}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
\textsuperscript{11}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
\textsuperscript{12}\emph{HIV/AIDS Unit, Infectious Disease Service, Geneva University Hospital, Geneva} \\
\textsuperscript{13}\emph{Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen} \\
\textsuperscript{14}\emph{Division of Infectious Diseases, Regional Hospital Lugano, Lugano} \\
\textsuperscript{*}\emph{Co-last authors}

\begin{doublespace}

\begin{abstract}
	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because human genetic variants can not only directly effect disease outcome, but also have indirect effects via selection on the pathogen. Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL). However, it is unclear to what extent the identified alleles are directly protective against all HIV strains. HIV strain-specific factors account for an estimated 21 - 29\% of spVL variability, but typical GWAS methods do not consider these effects. We propose a new method to consider the full genome of each patient's infecting strain, remove strain-specific effects on spVL, and thus better estimate the direct effect of human alleles on HIV spVL. In simulations, we show our method can increase GWAS power to detect truly associated human variants when viral effects are highly heritable. When we apply our method to real patient data from Switzerland, we recover weaker signals of association in the \emph{CCR5} gene region and among the top-associated variants in the major histocompatability complex (MHC) compared to a standard GWAS. The top-associated SNPs do not change. This method could also be used investigate whether GWAS-detected associations in other chronic infections are robust to a correction for pathogen-mediated effects.
\end{abstract}

\section{Introduction}

% What is GWAS and why is it important
A key goal of genome-wide association studies (GWAS) is to understand the genetic basis of phenotypic variation among individuals. In a typical GWAS, millions of genetic variants from across the human genome are screened for statistical association with a trait of interest. Ideally, this procedure identifies variants that are located in or are in linkage disequilibrium with alleles that directly effect the trait. If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \parencite{Okada2014}. Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify high-risk individuals \parencite{Dudbridge2013}.

% Why is spVL and important trait
For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with severity of disease course. spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \parencite{Mellors1996}. In untreated individuals, spVL is predictive of duration of asymptomatic infection \parencite{Mellors1996} and infectiousness \parencite{Quinn2000}. If viral load can be reduced to undetectable levels, an individual is effectively uninfectious \parencite{HHSARTGuidelines2019}. Notably, spVL varies by orders of magnitude between individuals \parencite{Mellors1996}. Thus, spVL measurements point to a wide range in natural HIV control by the human immune system.

% How do we measure the extent to which genetic factors determine spVL
We are interested in understanding to what extent spVL, as a proxy for natural HIV control, is determined by human and viral genetic factors. Heritability is a key measure of how genetically-determined a trait is. Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e. due to inherited differences. Narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e. the sum of independent effects from all genetic variants.

% How heritable is spVL (human side)
GWAS can be used to measure narrow-sense heritability from human genetic variants. Several GWAS have been done for spVL \parencite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}. The largest to-date by \citet{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%. All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}). Both associations are biologically relevant: the MHC encodes proteins that display viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry. In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed. On the other hand, CCR5 proteins are a signal to the virus that help it find the right cells to infect.

% How heritable is spVL (viral side)
In addition to these human genetic factors, is also well-recognized that viral genetic factors effect spVL. On the viral side, broad-sense heritability is typically measured. Estimates differ depending on the methods employed and the cohort studied (see \citealt{Mitov2018} for a discussion of this uncertainty). Estimates using phylogenetic methods on large UK and Swiss cohorts by \cite{Mitov2018} and \cite{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%. However, variation in the MHC is known to exert strong selective pressure on the virus \parencite{Kloverpris2016}. If the virus can change it's ``puzzle piece'' shape to escape MHC proteins, infected cells can go undetected. This means that MHC variants effect spVL largely via selection on the virus \parencite{Bartha2017}. In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.

% The gap
Most of the GWAS for human genetic determinants of spVL conducted so far \parencite{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL. In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors. This has several potential negative consequences: (i) variability due to viral effects would make it more challenging to identify human variants of small effect, (ii) we can't distinguish between human variants with direct effects versus those with effects mediated by the virus via selection, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.

% Overview of relevant methods and why we don't use them
Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry in a cohort can give rise to spurious genetic correlations with a trait. Corrections for these correlations are well-developed and widely accepted  \parencite{Astle2009, Price2006}. More recently, \cite{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS. Two state-of-the-art methods exist for this \cite{Earle2016, collins_phylogenetic_nodate}. 
% \cite{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework. \cite{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry. 
However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait. In contrast, we want to estimate effects of $human$ genetic variants on a trait, taking into account pathogen effects. \parencite{Naret2018} developed a relevant method for this task. The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing. This should correct for trait correlations due to shared pathogen ancestry. However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).

% Introduction to our method
In this work,  we draw from the field of phylogenetic comparative methods to develop a new GWAS approach that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny. Our approach should help identify human genetic variants with direct effects on disease traits. Furthermore, we think this approach represents the first GWAS framework to identify human genetic variants associated with a disease trait that accounts for non-independence of samples due to the underlying pathogen phylogeny.

In the following we (i) describe a statistical model for the spVL trait, (ii) derive a maximum likelihood estimate for the viral part of spVL under this model, and (iii) describe a new infectious disease GWAS framework using this information. In simulations, we demonstrate that this framework can improve GWAS power to detect human genetic variants effecting a disease trait. Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Studey (SHCS) and show that only certain associations with spVL are robust to a correction for viral effects. We suggest that this framework could be applied to other infectious disease GWAS.

\subsection{A statistical model for spVL}

Our premise is that spVL is the sum of three parts: direct host genetic effects, viral genetic effects, and environmental effects. Uniquely, the viral part is heritable from one transmission partner to another \parencite{Mitov2018, Bertels2018}. To characterize these parts, we use a phylogenetic mixed model (PMM, \cite{Housworth2004}). PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts. The heritable viral part is modeled by a random process occurring in continuous time along the branches of the viral phylogeny. The non-heritable host and environmental parts are modeled as gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).

So far, spVL has been modelled using PMMs with two different types of random processes. The Brownian Motion (BM) process assumes that trait values are unbounded (i.e. spVL can attain any value). In contrast, the Ornstein-Uhlenbeck (OU) process assumes the trait fluctuates around an optimal value (i.e. extreme spVL values are very unlikely) \cite{Butler2004}. We argue the OU assumption is more realistic for spVL. \citet{Fraser2014} hypothesized that spVL is under stabilizing selection to maximize viral transmission. Accordingly, both \citet{Mitov2018} and \citet{Bertels2018} found higher statistical support for the OU process compared to BM for spVL evolution. Therefore, we assume the OU process. This model, called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM), is described in detail by \citet{Mitov2018}, but we review the main points here in the spVL context.

Under the POUMM, the spVL trait $z$ is the sum of direct host genetic effects $g_h$, viral genetic efects $g_v$, and environmental effects $\epsilon$. We can group the phylogeny-independent effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$: 
\begin{equation}
    z = g_{v} + e
\end{equation}
$g_v$ is a viral trait that evolves along the phylogeny according to an OU process. The OU process is defined by a stochastic differential equation with two terms. The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \cite{Butler2004}:
\begin{align}
\begin{split}
	&dg_v(t) = \alpha[\theta - g_v(t)]dt + \sigma dW_t \\
	&g_v(0) = g_0 
	\label{eq:OUprocess}
\end{split}
\end{align}

Here the parameter $\alpha$ represents selection strength towards an evolutionarily optimal value represented by parameter $\theta$. The parameter $\sigma$ measures the intensity of stochastic fluctuations in the evolutionary process. Finally, $dW_t$ is the Wiener process underlying Brownian motion. The OU process is a Gaussian process, meaning that $g_v(t)$ is a Gaussian random variable. Assuming $g_v(t)$ starts at initial value $g_0$ at time $t = 0$ at the root of the phylogeny, we can write the expectation for $g_v(t)$ at time $t$: 
\begin{equation}
   E[g_v(t)] = g_0e^{-\alpha t} + (1 - e^{-\alpha t})\theta \label{eq:OUmean}
\end{equation}
and the variance in $g_v(t)$ if we were to repeat the random evolutionary process many times \parencite{Butler2004}:
\begin{equation}
    Var[g_v(t)] = \frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) \label{eq:OUVar}
\end{equation}

$g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny. The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \parencite{Butler2004}: 
\begin{equation}
	Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
	\label{eq:OUcov}
\end{equation}
Finally, we remember that $e$ is the non-heritable, environmental part of spVL. $e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent. The expectation of $e$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level. The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.
\begin{align}
\begin{split}
	E(e) = 0 \\
	Var(e) = \sigma^2_e
\end{split}
\end{align}

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.7\linewidth]{figures/model_figure}
		\caption{A statistical model of the spVL trait. The virus-determined part of spVL, $g_v$, evolves according to an OU process along the viral phylogeny. The dashed line shows the evolutionarily optimal trait value $\theta$. The environmentally-determined part of spVL, $e$, including direct host genetic effects and other environmental effects, is modeled as a Gaussian random variable whose value is added at the tips of the phylogeny.}
		\label{fig:spVLModel}
	\end{center}
\end{figure}

\subsection{Estimating and removing the virus-determined part of spVL}

Given the assumptions of the POUMM, we can derive a maximum-likelihood estimate for the virus-determined part of spVL for each individual in a cohort at the time they were sampled. Our goal is to estimate and remove this effect from measured spVL prior to GWAS.

Let $\bm{g}_v(\bm{t})$ be a vector of $g_v$ values, one for each individual in the cohort. $\bm{t}$ are the sampling times of each individual relative to the root of the phylogeny. To simplify notation, we omit the $\bm{t}$ from here on. $\bm{g}_v$ is a realization of a Gaussian random vector $\bm{G_v} \sim \mathcal{N}\big(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU}\big)$. The expectation $\bm{\mu}_{OU}$ is defined by equation \ref{eq:OUmean}, the diagonal elements of the covariance matrix $\boldsymbol{\Sigma}_{OU}$ are defined by equation \ref{eq:OUVar}, and the off-diagonal elements of $\boldsymbol{\Sigma}_{OU}$ by equation \ref{eq:OUcov}.

Similarly, let $\bm{e}$ be a vector of the environmental part of spVL for each individual.  $\bm{e}$ is a realization of a Gaussian random vector $\bm{E} \sim \mathcal{N}\big(\bm{0}, \boldsymbol{\Sigma}_E\big)$, where $\boldsymbol{\Sigma}_E$ is a diagonal matrix with diagonal elements equal to $\sigma^2_e$.

Considering that $\bm{G_v}$ and $\bm{E}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{e}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{e}$:
\begin{equation}
	f\big(\bm{g}_v, \bm{e}\big) \propto \mathcal{N}\big(\bm{g}_v; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
	\label{eq:pdfGprop}
\end{equation}

where the expected value of $\bm{g}_v$ and the covariance matrix $\boldsymbol{\Sigma}_G$ are defined as:
\begin{align}
	Exp(\bm{g}_v) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_E^{-1} \bm{z}\big) \label{eq:MuG}\\
	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_E^{-1}\big)^{-1} \label{eq:SigmaG}
\end{align}

\begin{proof}
	\begin{align}\label{eq:pdfG}
	\begin{split}
		f\big(\bm{g}_v,\ \bm{e}\big) &= f\big(\bm{g}_v|\ \bm{e}\big) \times f\big(\bm{e}\big) \\
	&= f\big(\bm{g}_v\big) \times f\big(\bm{e}\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{e};\ \bm{0}, \mathbf{\Sigma}_E\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_E\big) \\
	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_e\big)
	\end{split}
\end{align}

	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
\end{proof}

Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the viral part of spVL, taking into account all available information - the viral phylogeny, the inferred POUMM parameters, and measured spVL values. This estimator is an inverse-variance weighted average of information from the POUMM evolutionary model ($\bm{\mu}_{OU}$) and measured spVL ($\bm{z}$). Intuitively, the estimate for $g_v$ is closer to measured spVL if spVL is not very heritable, and is closer to the expected value under the POUMM if spVL is highly heritable.

Given that we have an estimate for $\bm{g}_v$, we can finally estimate $\bm{e}$, representing the spVL value with viral effects removed: 
\begin{equation}
	\hat{\bm{e}} = \bm{z} - Exp(\bm{g}_v)
	\label{eq:EHat}
\end{equation}

\subsection{A new infectious disease GWAS framework}

We propose that by estimating and removing spVL variability due to the pathogen we can better measure the direct effects of human genetic variants on spVL. Our proposed framework is as follows:

\begin{enumerate}
	\item Sample host genotypes, host trait values, and pathogen genetic sequences from a cohort.
	\item Construct a pathogen phylogeny using the pathogen genetic sequences.
	\item Fit the POUMM to the trait values and the phylogeny. This produces estimates for parameters $g_0, \alpha, \theta, \sigma,\text{ and }\sigma_e$.
	\item Generate maximum-likelihood estimates for the pathogen-determined and environmental parts of the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
	\item Perform GWAS, using the estimated environmental part of the trait as the response variable instead of raw trait values. 
\end{enumerate}

\section{Results}

\subsection{Simulation study}

To test the theoretical best-case performance of our framework, we simulated data under the POUMM. Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression used for each parameter.

\subsubsection{Estimator accuracy}

First, we evaluated estimator accuracy with respect to $\bm{g}_h$, the sum of additive host genetic effects. This would be the ideal GWAS response variable to identify direct host determinants of spVL. We compute the accuracy of our proposed GWAS endpoint $\bm{\hat{e}}$ and the typical GWAS endpoint, measured spVL, using root mean squared error (RMSE). In our simulations we expect (scaled) measured spVL, $\bm{z} - \bar{\bm{z}}$, where $\bar{\bm{z}}$ is the average spVL, to be Gaussian distributed around $\bm{g}_h$ with variance equal to $\sigma^2_e + \sigma^2_{g_v} \approx 0.55$ (see table \ref{tab:sim-params} for simulation parameters). This should yield an RMSE of $\approx$ 0.74 for $\bm{z} - \bar{\bm{z}}$ as an estimate of $\bm{e}$, which matches simulation results (figure \ref{fig:simulationResults}A, scaled spVL). By incorporating information from the viral phylogenetic tree in calculating $\bm{\hat{e}}$, we were better able to estimate $\bm{e}$. The greatest improvement was possible in simulation scenarios in which spVL was highly heritable, meaning the virus contributed a great deal of variability to spVL. Estimates were also better in scenarios with lower selection strength since the correlation in $g_v$ among related tips is higher when $g_v$ can drift further from the long-term optimum and the stochastic fluctuations characterized by $\sigma$ are correspondingly lower to maintain the same viral genetic variance $\sigma^2_{g_v}$ and the same broad-sense heritability $H^2$.

\subsubsection{Theoretical GWAS improvement}

Next, we characterize the evolutionary scenarios in which our framework improves GWAS power. More accurate estimation of the non-viral part of spVL does not guarantee improved GWAS power. For example, if the viral part of spVL were constant across all viruses, association results (significance and effect size) would not be effected by a viral contribution to the trait. To determine under what evolutionary scenarios our method can improve GWAS power, we performed GWAS on the simulated data. Under our simulation set-up a standard GWAS would have $\approx$ 30\% power to detect a host variant explaining 1.25\% of the variability in spVL. This matches simulation results (figure \ref{fig:simulationResults}B, scaled spVL). Using our proposed framework, we were able to improve GWAS power in simulated scenarios with selection strength $<$ 10-20 time$^{-1}$ and spVL heritability $>$ 35-45\% (figure \ref{fig:simulationResults}B, $\bm{\hat{e}}$). If were were able to perfectly estimate $\bm{e}$, GWAS power would increase across all values of selection strength so long as the trait is highly heritable (figure \ref{fig:simulationResults}, $\bm{e}$). In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by removing an estimated pathogen-determined effect.

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
		\caption{(A) Root mean square error (RMSE) with respect to true (simulated) values for the host part of spVL ($\bm{{g_h}}$). The left plot is for the POUMM-based estimate of the non-viral part of spVL ($\bm{\hat{e}}$) and the right plot is for spVL scaled such that it is centered at zero ($\bm{z} - \bm{\bar{z}}$). Each plot tile represents the average RMSE over 20 simulated datasets of 500 samples. (B) True positive rate (TPR) in GWAS performed using true (simulated) non-viral spVL values ($\bm{e}$), the POUMM-based estimate for the non-viral part of spVL ($\bm{\hat{e}}$), and total spVL scaled such that it is centered at zero ($\bm{z} - \bm{\bar{z}}$). Each plot tile represents the average TPR over 20 simulated datasets of 500 samples.}
		\label{fig:simulationResults}
	\end{center}
\end{figure}

\subsection{GWAS on the Swiss HIV Cohort}

Finally, we demonstrate our framework on empirical data from the Swiss HIV Cohort Study (SHCS). In the ``standard GWAS" we followed a typical GWAS methodology,  i.e. regressing measured spVL ($\bm{z}$) on variant dosage for each sampled host genetic variant. In the ``GWAS with phylogenetic correction", we use our proposed framework, i.e. regressing the estimated environmental part of spVL ($\bm{\hat{e}}$) on variant dosage. In agreement with previous GWAS for host genetic determinants of spVL, the top associations were in the MHC and \emph{CCR5} regions for both GWAS (figure \ref{fig:gwas-results}).

\begin{figure}[H]
	\begin{center}
		\includegraphics[width=\linewidth]{figures/200207_gwas_results}
		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load. The pullout plots give regional association results from the GWAS with phylogenetic correction in the \emph{CCR5} region (positions 45.4 - 47Mb on chromosome 3) and the MHC region (positions 29.5 - 33.5Mb on chromosome 6). Base positions are with reference to genome build GRCh37. The color scale in the regional association plot is the difference in -log$_{10}$ p-value between the GWAS with phylogenetic correction and the standard GWAS. Red color means the phylogenetic correction decreased variant significance and blue color means the phylogenetic correction increased variant significance. The dashed lines show genome-wide significance (p-value = 5x10$^{-8}$).}
		\label{fig:gwas-results}
	\end{center}
\end{figure}

The overall distribution of p-values did not change noticeably between the standard GWAS and the GWAS with a phylogenetic correction (figure \ref{fig:qq-plots}). In the \emph{CCR5} and MHC regions, most variants decreased in significance in the GWAS with phylogenetic correction. However, the top-associated variants were constant between analyses. rs112243036 was most significant in the MHC, with a standard GWAS p-value of 9.7x10$^{-16}$ and corrected GWAS p-value of 2.5x10$^{-14}$. In the \emph{CCR5} region, rs7430431 was most significant, with a standard GWAS p-value of 7.4x10$^{-9}$ and a corrected GWAS p-value of 1.7x10$^{-7}$. Some variants in the MHC did increase in significance with a phylogenetic correction. For example, SNP rs9265880 had the greatest increase in significance from a p-value of 2.2x10$^{-5}$ in the standard GWAS to 6x10$^{-7}$ in the GWAS with phylogenetic correction.

The top-associated SNPs from the \citet{McLaren2015} paper were excluded from the main analysis because of too-high missing rates (3\% for rs1015164 and 4\% for rs59440261). In order to directly compare our results with \citet{McLaren2015}, we ran an additional GWAS for these two SNPs anyways. Both SNPs became less significant with phylogenetic correction (table \ref{tab:comp-gwas}).

Again following the procedure of \citet{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region. Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).

\begin{table}[H]
	\begin{center}
		\begin{tabular}{p{1cm}p{2cm}p{2.5cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr} {llclcccc}
			\hline
			& & McLaren et al.  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
			\hline
			Region & Variant & p-value & Beta & p-value & Beta & p-value \\
			\hline
			MHC & rs59440261 & 2.0 x 10$^{-83}$ & -0.414 & 1.6x10$^{-11}$ & -0.229 & 1.2x10$^{-10}$ \\
			\emph{CCR5} & rs1015164 & 1.5 x 10$^{-19}$ & 0.148 & 1.3x10$^{-6}$ & 0.073 & 3.2x10$^{-5}$ \\			
			\hline
		\end{tabular}
		\caption{Top association results from \citet{McLaren2015} compared to association results in the current study.}
		\label{tab:comp-gwas}
	\end{center}
\end{table}

\begin{table}[H]
	\begin{center}
	\begin{tabular} {p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}
		\hline
		 & \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
		 \hline
		Allele & Beta & p-value & Beta & p-value & Beta & p-value \\
		\hline	
		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
		\rowcolor{Gray}
		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
		\rowcolor{Gray}
		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
		\rowcolor{Gray}
		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
		\rowcolor{Gray}
		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
		\rowcolor{Gray}
		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
		\rowcolor{Gray}
		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
		\rowcolor{Gray}
		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\
		\hline
	\end{tabular}
	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citet{McLaren2015} compared to results from the current study. The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al. study. The full table is available in the supplement (table \ref{tab:full-mclaren-comparison}). Grey rows are alleles that increased in significance with a phylogenetic correction.}
	\label{tab:lim-mclaren-comparison}
	\end{center}
\end{table}

\section{Discussion}
\begin{itemize}
	\item From the field of phylogenetic comparative methods there are several methods to account for non-independence among samples related by an underlying phylogeny (e.g. independent contrasts, phylogenetic generalized least squares). Compared to these methods, our method is nice for GWAS because we do not change the association model itself, instead we just adjust the trait values prior to association testing.
\end{itemize}

%\begin{itemize}
%	\item Compare inferred POUMM parameters to previous studies \parencite{Mitov2018, Bertels2018}
%	\item Standard GWAS highlights previously identified regions associated with spVL: $CCR5$ and MHC
%	\item We expect non-viral only GWAS to increase signals of association in $CCR5$ (because noise in spVL from the virus reduced) and decrease signals of association in MHC (because viral mutations mediate MHC effects). It \emph{seems} like the results support these hypotheses, but I am waiting on POUMM parameter estimates from a longer MCMC run.
%	\item The phylogenetic correction to GWAS leads to the biggest p-value changes in the $CCR5$ and MHC regions
%	\item SNPs near the FUT1 gene (on chromosome 6, highlighted as one of the top-20 p-value changes in the first pipeline run) increased in significance, and the Lewis-X antigen produced by this gene is known to play a role in early HIV infection \cite{Naarding2005}.
%	\item Address method caveats: 
%	\begin{itemize}
%		\item We don't model virus-host interaction effects
%		\item We don't account for uncertainty in non-viral component of spVL estimates
%	\end{itemize}
%	\item Implications for future research: what other infectious disease traits are heritable?
%\end{itemize}

\section{Methods}

\subsection{Simulation model}

Whenever possible, we tried to parameterize our simulation model for spVL using empirical data. We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \parencite{Mitov2018}. After allotting 25\% of this variance to the host part of spVL $g_h$ (following \citealt{McLaren2015}), we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities. $g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$. Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see table \ref{tab:sim-params}). We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \parencite{Hodcroft2014} using the R package ape \parencite{Paradis2018}. $g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \parencite{Mitov2017a-POUMM} and sampling values at the tips. For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (table \ref{tab:POUMMparams}). To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively. Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.

\begin{figure}[H]
	\begin{tikzpicture}

	% Variance 
	\begin{scope}[xshift = 0cm, yshift = 0cm]
	\node[anchor = west] (alabel) at (-1, 4) {a) Variance partitioning};
	\node[constnode] (H2h) at (0, 0) {0.25};
	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {$H^2_{\bar{t}}$};
	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{\epsilon}$};
	\draw[taro] (H2h) -- (varh);
	\draw[taro] (varz) -- (varh);
	\draw[taro] (varz) -- (varv);
	\draw[taro] (H2) -- (varv);
	\draw[taro] (varh) -- (vare);
	\draw[taro] (varv) -- (vare);
	\draw[taro] (varz) -- (vare);
	\end{scope}

	% Virus
	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
	\node[anchor = west] (blabel) at (-1, 4.5) {b) Viral part of spVL};
	\node[constnode] (theta) at (0, 1.5) {$\theta$};
	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
	\node[constnode] (alpha) at (3.6, 1.5) {$\alpha$};
	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
	\node[snode] (v) at (3.6, 3.25) {$\vec{g_v}$};
	\draw[taro] (theta) -- (v);
	\draw[taro] (g0) -- (v);
	\draw[taro] (alpha) -- (v);
	\draw[taro] (sigma) -- (v);
	\draw[taro] (tree) -- (v);
	\draw[taro] (varv) -- (sigma);
	\draw[taro] (tbar) -- (sigma);
	\draw[taro] (tbar) -- (tree);
	\draw[taro] (alpha) -- (sigma);
	\node[white] (varpic) at (3.6+0.9, -1.75) {};
	\draw[taro] (varpic) -- (varv);
	\node (vdist) at ($(v) + (1.8, 0.3)$) {Normal};
	\draw (v) -- (vdist);
	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
	\end{scope}

	% Host 
	\begin{scope}[xshift = 0cm, yshift = -9cm]
	\node[anchor = west] (clabel) at (-1, 5.55) {c) Host part of spVL};
	\node[constnode] (d) at (1.8, 1.5) {2};
	\node[detnode] (p) at (3.6, 1.5) {$p$};
	\node[snode] (G) at (1.8, 3) {$G$};
	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{g_h}$};
	\node[constnode] (delta) at (3.6, 3) {$\delta$};
	\node[detnode] (h) at (3.6, 4.5) {$\vec{g_h}$};
	\node[constnode] (M) at (5.4, 3) {$M$};
	\draw[taro] (d) -- (G);
	\draw[taro] (p) -- (G);
	\draw[taro] (varh) -- (p);
	\draw[taro] (delta) -- (p);
	\draw[taro] (M) -- (p);
	\draw[taro] (M) -- (h);
	\draw[taro] (delta) -- (h);
	\draw[taro] (G) -- (h);
	\node[white] (varpic) at (3.6, -1.5) {};
	\draw[taro] (varpic) -- (varh);
	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Normal};
	\draw (h) -- (hdist);
	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
	\draw (G) -- (gdist);
	\end{scope}

	% Environment
	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
	\node[anchor = west] (dlabel) at (-1, 3) {d) Environmental part of spVL};
	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{\epsilon}$};
	\draw[taro] (zero) -- (e);
	\draw[taro] (vare) -- (e);
	\node[white] (varpic) at (0, -1.5) {};
	\draw[taro] (varpic) -- (vare);
	\node (edist) at ($(e) + (1.8, 0.3)$) {Normal};
	\draw (e) -- (edist);
	\end{scope}

	\end{tikzpicture}
	\caption{A graphical representation of the simulation model. Variables in solid squares are constant in the simulation scheme, whereas variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables. Arrows represent dependencies among variables and the dashed square represents repetition \parencite{Hohna2014}.}
	\label{fig:sim-design}
\end{figure}

\begin{table}[H]
	\begin{tabular}{p{1.5cm}ll} \toprule 
		Variable & Expression & Definition \\ \midrule 
		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & total spVL variance \\ 
		$H^2_{\bar{t}}$ & varied & spVL viral heritability at $\bar{t}$ \\
		$\sigma^2_{g_h}$ & $\sigma_{g_h}^2 = 0.25*\sigma_z^2$ & variance in host part of spVL \\
		$\sigma^2_{g_v}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & variance in viral part of spVL at $\bar{t}$ \\ 
		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{g_v} - \sigma^2_{g_h}$ & variance in environmental part of spVL \\ 
		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & mean root-tip time in viral phylogeny \\ 
		$\bm{g}_v$ & $\bm{g}_v \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & viral part of spVL for all individuals \\ 
		$\theta$ & 4.5 log copies/mL & optimal spVL value \\
		$g_{0}$ & 4.5 log copies/mL & $g_v$ at the root of the phylogeny  \\ 
		$\alpha$ & varied & selection strength of OU process \\ 
		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & time-unit standard deviation of OU process \\ 
		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & viral phylogeny \\ 
		$g_{h_i}$ & $g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & host part of spVL for individual $i$ \\ 
		$G_{N \times M}$ & $G_{ij} \sim Binom(2, p)\ \forall i \in {1...N}, \forall j \in {1...M}$ & host genotype matrix \\
		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{\bar{t}}^2\sigma^2_z}{2\delta^2M}}$ &  host variant allele frequency \\ 
		$\delta$ & 0.2 & host variant effect size \\ 
		$M$ & 20 & number of causal host variants \\ 
		$\epsilon_i$ & $ \epsilon_i \sim Norm(0, \sigma^2_{\epsilon})$ & environmental part of spVL for individual $i$ \\ \bottomrule 
	\end{tabular}
	\caption{Simulation model parameters.}
	\label{tab:sim-params}
\end{table}

\subsection{Swiss HIV-1 data}

Genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, \citealt{Schoeni-Affolter2010}).

We retained data from 1,396 individuals with non-recombinant subtype B $pol$ gene sequences of at least 750 characters, paired RNA measurements allowing for calculation of spVL, European ancestry, and not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}). Viral subtype was determined by the SHCS using the REGA HIV subtyping tool version 2.0 \parencite{DeOliveira2005}. We calculated spVL as the arithmetic mean of viral RNA measurements made prior to the start of antiretroviral treatment. Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}). Kinship was evaluated using PLINK version 2.00a2LM \parencite{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375.

For phylogenetic inference, we included the 1,396 subtype B sequences and 5 randomly chosen subtype A sequences from the SHCS as an outgroup. We used MUSCLE version 3.8.31 \parencite{edgar_muscle:_2004} to align the $pol$ sequences with --maxiters 3 and otherwise default settings. We trimmed the alignment to 1505 characters to standardize sequence lengths. We used IQ-TREE version 1.6.9 to construct an approximate maximum likelihood tree with -m GTR+F+R4 for a general time reversible substitution model with empirical base frequencies and four free substitution rate categories. Otherwise, we used the default IQ-TREE settings. After rooting the tree based on the subtype A samples, we removed the outgroup.

Initial host genotyping quality control and imputation were done by the SHCS. Subsequent genotyping quality control was performed using PLINK version 2.3 \parencite{Chang2015}. We used the options --maf 0.01, --geno 0.01, and --hwe 0.00005 to remove variants with minor allele frequency less than 0.01, missing call rate greater than 0.01, or Hardy-Weinberg equilibrium exact test p-value less than 5x10$^-5$. After quality filtering, approximately XXX million genetic variants from XXX individuals were retained for GWAS (Table \ref{tab:variant-filtering}).

\subsection{POUMM parameter inference}

We used the R package POUMM version 2.1.6 \parencite{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the approximate maximum-likelihood phylogeny and calculated spVL values. The Bayesian inference method implemented in this package requires specification of a prior distribution for each parameter. We used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$. We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000. The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution. Inferred parameters, posterior distributions for inferred parameters, and MCMC trace plots are given in table \ref{tab:POUMM-parameters} and figure \ref{fig:poumm-fit}, respectively.

\subsection{Phylogenetic spVL correction}
We corrected calcualted spVL values using the method described in this paper. For each of the 1,396 individuals in our cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}. For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the maximum posterior probability estimates generated as described above.

\subsection{Association testing}
We performed two GWAS using the same genotype data. For the first ``standard GWAS" we used total calculated spVL ($\bm{z}$) as the dependent variable for association testing, replicating prior GWAS for host genetic determinants of spVL. For the second ``GWAS with phylogenetic correction" we replaced total spVL with the estimated non-viral component of spVL ($\hat{\bm{e}}$) as the dependent variable. Association testing was performed using a linear association model in PLINK version 2.3 \parencite{Chang2015} with sex and the top 5 principle components of host genetic variation included as covariates. The sex and principle components covariates were included to reduce residual variance in spVL and control for confounding from host population structure, respectively. Common HLA allele genotypes were imputed using the SNP2HLA pipeline \parencite{Jia2013a} with a reference panel of 124 individuals from the HapMap CEU population. Association tests for imputed HLA alleles were performed in the same way as for other variants.

\section{Acknowledgments}

\begin{itemize}
	\item Thanks to Michael Landis, who shared a LaTeX template for graphical model drawing.
	\item Esposito Mauro and Alexandra Scherrer from the SHCS data center were especially helpful in retrieving data.
\end{itemize}

%Additional ideas for supplemental materials:
%\begin{itemize}
%	\item Table of genotype samples by project (genotyping chip, number of samples, spVL distribution)
%	\item Link to github with all the code used to run simulations and generate figures
%	\item Example spVL plots
%	\item simulated TPR plot annotated with actual inferred POUMM params
%	\item Simulation results with different parameterizations
%	\item Whole pipeline replicated with a stricter spVL definition (and correspondingly fewer samples)
%	\item Comparative QQ-plots from 1500 and standard GWAS
%	\item Proof of robustness of method to inferred tree
%\end{itemize}

\newpage
\section{Supplemental Material}
\beginsupplement

\subsection{Expected results from simulations}

Here we show the root mean square error (RMSE) of $\bm{z} - \bm{\bar{z}}$ as an estimate for $\bm{h}$ should be $\approx$ 0.74 in our simulation setup. First we write the expression for the RMSE:

\begin{equation}
    RMSE = \sqrt{\frac{\sum_i^N(z - \bar{z} - h)^2}{N}}
\end{equation}

Note that $z - \bar{z}$ differs from $h$ due to a viral effect and an environmental effect. So the term inside the square root equals the combined variance of these two effects:

\begin{equation}
    RMSE = \sqrt{\sigma^2_{g_v} + \sigma^2_\epsilon}
\end{equation}

We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{g_h}$, are fixed parameters in our simulation setup.

\begin{equation}
\begin{split}
    \sigma^2_{g_h} + \sigma^2_{g_v} + \sigma^2_\epsilon = \sigma^2_z \\
    0.25 * \sigma^2_{z} + \sigma^2_{g_v} + \sigma^2_\epsilon = \sigma^2_z \\
    \sigma^2_{g_v} + \sigma^2_\epsilon = 0.75 * \sigma^2_z \\
    \sigma^2_{g_v} + \sigma^2_\epsilon = 0.75 * 0.73 \\
    \sigma^2_{g_v} + \sigma^2_\epsilon = 0.55
\end{split}
\end{equation}

Therefore, we can expect the RMSE for $\bm{z} - \bm{\bar{z}}$ as an estimate for $\bm{h}$ to be around $\sqrt{0.55} \approx 0.74$.

\begin{table}[H]
	\begin{center}
		\begin{tabular}{cccc} \hline 
			Parameter & MLE & Maximum posterior probability & 95\% HPD \\ \hline 
			${g_{0}}$ & 4.17 & 4.23 & 2.73, 5.88 \\
			$\theta$ & 4.32 & 4.33 & 4.25, 4.42 \\
			$\sigma$ & 5.15 & 4.98 & 3.45, 6.61 \\
			$\alpha$ & 45.46 & 43.98 & 23.80, 63.36 \\
			$\sigma_e$ & 0.55 & 0.56 & 0.49, 0.63 \\ 
			$H^2_{\bar{t}}$ & 0.49 & 0.47 & 0.33, 0.60 \\ \hline
		\end{tabular}
		\caption{Inferred POUMM parameters.}
		\label{tab:POUMM-parameters}
	\end{center}
\end{table}

\begin{figure}[H]
	\centering
	\begin{subfigure}[b]{0.75\textwidth}
		\includegraphics[width=\linewidth]{figures/200207_POUMM_posteriors_zoomed.png}
		\caption{Marginal posterior distributions.}
	\end{subfigure}
~
	\begin{subfigure}[b]{0.75\textwidth}
		\includegraphics[width=\linewidth]{figures/200207_POUMM_traces.png}
		\caption{MCMC trace plots.}
	\end{subfigure}
	\caption{POUMM parameter inference.}\label{fig:poumm-fit}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/200207_spvl_on_tree}
	\caption{Inferred viral phylogeny with tips colored by estimated viral part of spVL, non-viral part of spVL, and measured total spVL.}
	\label{fig:spvl-on-tree}
\end{figure}

\begin{table}[H]
    \begin{center}
    \begin{tabular}{lc}
    \hline
    Sample filter & Number of samples remaining \\
    \hline
    Subtype B pol sequences & 1516 \\
    With paired spVL measurement & 1516 \\
    $>$ 750 characters in sequence & 1493 \\
    Individual is of European ancestry & 1396 \\
    Kinship coefficient $>$ 0.09375 & TODO \\ 
    \hline
    \end{tabular}
    \caption{Number of samples for GWAS after sequential filtering steps.}
    \label{tab:sample-filtering}
    \end{center}
\end{table}

\begin{table}[H]
	\begin{center}
		\begin{tabular}{lc}
			\hline
			Variant filter & Number of variants remaining \\
			\hline 
			Raw data & 76,979,521 \\
			Missing call rate $>$ 0.01 & -69,526,564 \\
			Hardy-Weinburg p-value $<$ 5x10$^-5$ & -535 \\
			Minor allele frequency $<$ 0.01 & -2,897,654 \\
			Variants for GWAS & 4,554,768 \\ \hline			
		\end{tabular}
		\caption{Number of variants for GWAS after sequential filtering steps.}
		\label{tab:variant-filtering}
	\end{center}
\end{table}

\begin{figure}[H]
	\includegraphics[width=\linewidth]{figures/pca_plot}
	\caption{Thresholds for exclusion of individuals of non-Western European ancestry. The lower plots are filtered to show only European HapMap samples and the rectangles show thresholds for sample exclusion.}
	\label{fig:PCA}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\linewidth]{figures/200207_qq_plots.png}
	\caption{Quartile-quartile plots from association tests.}
	\label{fig:qq-plots}
\end{figure}

\newpage
\begin{longtable}[H]{p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
		\hline
		& \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS} & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
		\hline
	 	Allele & Beta & p-value & Beta &p-value & Beta& p-value \\ 
		\hline
		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\ 
		HLA-C*02:02 & -0.223 & $5.45 \times 10^{-8}$ & -0.089 & $1.31 \times 10^{-1}$ & -0.039 & $2.50 \times 10^{-1}$ \\ 
		HLA-A*25:01 & -0.301 & $6.80 \times 10^{-8}$ & -0.090 & $9.93 \times 10^{-2}$ & -0.033 & $2.96 \times 10^{-1}$ \\ 
		HLA-B*52:01 & -0.435 & $2.60 \times 10^{-7}$ & 0.238 & $7.63 \times 10^{-2}$ & 0.150 & $5.33 \times 10^{-2}$ \\ 
		HLA-C*01:02 & -0.231 & $2.22 \times 10^{-6}$ & -0.198 & $5.07 \times 10^{-4}$ & -0.113 & $5.70 \times 10^{-4}$ \\ 
		HLA-A*23:01 & 0.269 & $1.66 \times 10^{-5}$ & 0.020 & $7.42 \times 10^{-1}$ & 0.015 & $6.59 \times 10^{-1}$ \\ 
		HLA-B*55:01 & 0.302 & $2.32 \times 10^{-5}$ & 0.685 & $2.48 \times 10^{-1}$ & 0.436 & $2.03 \times 10^{-1}$ \\ 
		HLA-B*35:02 & 0.358 & $2.42 \times 10^{-5}$ &  &  &  &  \\ 
		HLA-A*31:01 & -0.219 & $4.36 \times 10^{-5}$ & -0.052 & $4.50 \times 10^{-1}$ & -0.017 & $6.78 \times 10^{-1}$ \\ 
		HLA-B*35:01 & 0.156 & $4.82 \times 10^{-5}$ & 0.167 & $2.13 \times 10^{-4}$ & 0.083 & $1.33 \times 10^{-3}$ \\ 
		HLA-B*49:01 & 0.253 & $1.86 \times 10^{-4}$ & 2.173 & $9.79 \times 10^{-4}$ & 1.253 & $9.82 \times 10^{-4}$ \\ 
		HLA-B*35:03 & 0.215 & $7.13 \times 10^{-4}$ & -0.241 & $8.00 \times 10^{-1}$ & 0.261 & $6.35 \times 10^{-1}$ \\ 
		HLA-B*40:01 & 0.148 & $9.37 \times 10^{-4}$ & 0.519 & $1.06 \times 10^{-2}$ & 0.349 & $2.91 \times 10^{-3}$ \\ 
		HLA-B*18:01 & 0.136 & $9.48 \times 10^{-4}$ & 0.208 & $1.73 \times 10^{-3}$ & 0.133 & $5.09 \times 10^{-4}$ \\ 
		HLA-A*29:02 & 0.167 & $1.40 \times 10^{-3}$ & 0.150 & $6.17 \times 10^{-2}$ & 0.092 & $4.72 \times 10^{-2}$ \\ 
		HLA-C*14:02 & -0.228 & $1.50 \times 10^{-3}$ & 0.266 & $7.59 \times 10^{-1}$ & 0.235 & $6.38 \times 10^{-1}$ \\ 
		HLA-A*11:01 & -0.109 & $2.70 \times 10^{-3}$ & -0.090 & $8.96 \times 10^{-2}$ & -0.052 & $8.92 \times 10^{-2}$ \\ 
		HLA-A*24:02 & 0.094 & $3.60 \times 10^{-3}$ & 0.133 & $3.79 \times 10^{-3}$ & 0.074 & $5.52 \times 10^{-3}$ \\ 
		HLA-B*37:01 & 0.222 & $4.50 \times 10^{-3}$ & 0.600 & $3.22 \times 10^{-3}$ & 0.311 & $8.11 \times 10^{-3}$ \\ 
		HLA-B*44:03 & 0.119 & $5.00 \times 10^{-3}$ & -0.114 & $2.77 \times 10^{-1}$ & -0.080 & $1.88 \times 10^{-1}$ \\ 
		HLA-C*03:03 & 0.110 & $6.50 \times 10^{-3}$ & -0.078 & $1.86 \times 10^{-1}$ & -0.031 & $3.66 \times 10^{-1}$ \\ 
		HLA-B*58:01 & -0.253 & $7.10 \times 10^{-3}$ & 6.895 & $7.93 \times 10^{-2}$ & 4.841 & $3.27 \times 10^{-2}$ \\ 
		HLA-A*32:01 & -0.123 & $8.30 \times 10^{-3}$ & 0.089 & $5.31 \times 10^{-1}$ & 0.071 & $3.86 \times 10^{-1}$ \\ 
		HLA-C*16:01 & 0.137 & $8.60 \times 10^{-3}$ & 0.057 & $4.46 \times 10^{-1}$ & 0.013 & $7.60 \times 10^{-1}$ \\ 
		HLA-C*03:04 & 0.093 & $1.25 \times 10^{-2}$ & 0.126 & $3.52 \times 10^{-2}$ & 0.095 & $5.77 \times 10^{-3}$ \\ 
		HLA-B*38:01 & -0.148 & $1.35 \times 10^{-2}$ & -3.459 & $3.01 \times 10^{-1}$ & -2.224 & $2.49 \times 10^{-1}$ \\ 
		HLA-A*33:01 & -0.233 & $1.45 \times 10^{-2}$ &  &  &  &  \\ 
		HLA-A*30:01 & -0.159 & $2.89 \times 10^{-2}$ & 0.072 & $9.04 \times 10^{-1}$ & 0.008 & $9.82 \times 10^{-1}$ \\ 
		HLA-C*12:03 & -0.077 & $4.59 \times 10^{-2}$ & -0.114 & $2.95 \times 10^{-2}$ & -0.067 & $2.68 \times 10^{-2}$ \\ 
		HLA-A*01:01 & 0.051 & $4.78 \times 10^{-2}$ & -0.039 & $4.88 \times 10^{-1}$ & -0.050 & $1.24 \times 10^{-1}$ \\ 
		HLA-B*14:01 & -0.187 & $5.51 \times 10^{-2}$ & 7.071 & $5.00 \times 10^{-2}$ & 3.974 & $5.62 \times 10^{-2}$ \\ 
		HLA-B*40:02 & -0.130 & $1.01 \times 10^{-1}$ & -0.343 & $3.88 \times 10^{-1}$ & -0.105 & $6.47 \times 10^{-1}$ \\ 
		HLA-C*07:04 & 0.125 & $1.02 \times 10^{-1}$ & 0.111 & $3.00 \times 10^{-1}$ & 0.080 & $1.98 \times 10^{-1}$ \\ 
		HLA-C*05:01 & 0.043 & $1.93 \times 10^{-1}$ & -0.022 & $6.51 \times 10^{-1}$ & -0.019 & $4.82 \times 10^{-1}$ \\ 
		HLA-A*03:01 & 0.025 & $3.29 \times 10^{-1}$ & 0.002 & $9.51 \times 10^{-1}$ & -0.001 & $9.81 \times 10^{-1}$ \\ 
		HLA-B*51:01 & -0.036 & $3.52 \times 10^{-1}$ & -0.046 & $3.13 \times 10^{-1}$ & -0.044 & $1.01 \times 10^{-1}$ \\ 
		HLA-A*68:01 & 0.046 & $3.71 \times 10^{-1}$ & 0.036 & $6.09 \times 10^{-1}$ & 0.009 & $8.27 \times 10^{-1}$ \\ 
		HLA-B*44:02 & 0.030 & $3.73 \times 10^{-1}$ & -0.051 & $3.90 \times 10^{-1}$ & -0.031 & $3.57 \times 10^{-1}$ \\ 
		HLA-B*39:01 & 0.073 & $3.85 \times 10^{-1}$ &  &  &  &  \\ 
		HLA-A*02:01 & 0.017 & $3.94 \times 10^{-1}$ & -0.018 & $5.57 \times 10^{-1}$ & -0.008 & $6.77 \times 10^{-1}$ \\ 
		HLA-C*15:02 & -0.046 & $4.20 \times 10^{-1}$ & 0.064 & $3.26 \times 10^{-1}$ & 0.033 & $3.78 \times 10^{-1}$ \\ 
		HLA-B*50:01 & 0.066 & $4.68 \times 10^{-1}$ & 1.522 & $1.77 \times 10^{-2}$ & 0.970 & $8.73 \times 10^{-3}$ \\ 
		HLA-B*15:01 & -0.021 & $5.71 \times 10^{-1}$ & -0.073 & $1.71 \times 10^{-1}$ & -0.021 & $4.93 \times 10^{-1}$ \\ 
		HLA-A*30:02 & 0.050 & $5.92 \times 10^{-1}$ & 0.047 & $9.28 \times 10^{-1}$ & -0.004 & $9.89 \times 10^{-1}$ \\ 
		HLA-A*26:01 & -0.003 & $9.57 \times 10^{-1}$ & -0.645 & $1.96 \times 10^{-1}$ & -0.529 & $6.59 \times 10^{-2}$ \\
		\hline
		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citet{McLaren2015} compared to results from the current study.}
		\label{tab:full-mclaren-comparison}
\end{longtable}

\begin{table}[H]
	\begin{center}
		\begin{tabular}{cccl}
			\toprule 
			Parameter & Value & Uncertainty & Reference \\
			\midrule 
			$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citealt{Mitov2018} \\
			$\theta$ & 4.45 & (4.41 - 4.49) HPD& \citealt{Mitov2018}  \\
			$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citealt{Bertels2018} \\
			$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citealt{Blanquart2017} \\
			$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citealt{Mitov2018} \\
			$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citealt{Bertels2018} \\
			$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citealt{Blanquart2017}** \\
			$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citealt{Mitov2018} \\
			$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citealt{Blanquart2017} \\
			$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citealt{Mitov2018} \\
			$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citealt{Blanquart2017} \\
			\bottomrule 
		\end{tabular}
		\caption{POUMM parameters previously inferred from HIV data. Notes for each study: \protect\citet{Mitov2018} = 8,483 UK HIV cohort individuals, pol tree; \protect\citet{Bertels2018} = 3,036 SHCS individuals, pol tree; \protect\citet{Blanquart2017} = 1,581 subtype B individuals from BE, CH, FR, NL, UK, whole genome tree, **limited $\alpha$ to $\le$ 10}
		\label{tab:POUMMparams}
	\end{center}
\end{table}

\begin{table}[H]
	\begin{center}
			\begin{tabular}{ccl} \toprule 
			Measurement & Value & Reference \\ \midrule 
			$\bar{z}$ & $\approx$ 4.5 & \citealt{Mitov2018} \\
			$\bar{z}$ &  4.4 & \citealt{Blanquart2017} \\
			$\bar{z}$ & $\approx$ 4.5 & \citealt{Bonhoeffer2015} \\
			$\sigma^2_z$ &  0.73 &  \citealt{Mitov2018} \\
			$\sigma^2_z$ &  0.50 & \citealt{Blanquart2017} \\
			$\sigma^2_z$ & $\approx$ 0.5 & \citealt{Bonhoeffer2015} \\ \bottomrule 
		\end{tabular}
		\caption{Summary statistics for log spVL in previously sampled populations. $\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$). Values from \protect\citet{Blanquart2017, Mitov2018} are empirical while values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
		\label{tab:spVLmeasurements}
	\end{center}
\end{table}
%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}

\begin{figure}[H]
	\includegraphics[width=\linewidth]{images/HIV_tree_branch_lengths}
	\caption{Left plot: branch length distributions for UK HIV tree and our simulated HIV tree prior to pruning down to 500 tips. The mean root-to-tip distance for the UK tree is 0.14 and the mean branch length is 0.02. The mean root-to-tip distance for the full simulated HIV tree is 0.14 and the mean branch length is 0.01. Right plot: branch length distribution for the tree of 500 tips used for simulation. The mean root-to-tip distance for the pruned HIV tree is 0.14 and the mean branch length is 0.03.}
	\label{fig:branchlengths}
\end{figure}

\end{doublespace}

\typeout{}
\printbibliography

\end{document}

%\citet{Mitov2018} and \citet{Bertels2018} both found higher statistical support for the POUMM model of spVL evolution as opposed to the corresponding model assuming Brownian motion only. This makes sense in the context of the hypothesis proposed by \citet{Fraser2014}, in which spVL is under stabilizing selection to maximize transmissibility of the virus. On the other hand, the assumptions that $e$ are normally distributed, mutually indpendent, and have an expected value of 0 are consistent with the idea that genetic effects from several unlinked host loci and environmental effects from many independent factors are each equally likely to increase an individual's spVL as decrease it. \todo{This statement is not supported}
%
%\todo{TBD once we have real data, here are some thoughts}
%The success of our method in identifying new host determinants of spVL depends on how much host-determined variation in spVL actually exists. In the largest GWAS for HIV spVL to date, \citet{McLaren2015} estimated that additive effects from common human genetic variants explain 20\% of the variability in spVL. Of this quantity, 94.5\% was attributable to variants in just two regions in the human genome. \todo{Were there variants other than HLA in the MHC region?} Associations were with alleles in the CCR5 gene, which encodes a co-receptor for HIV entry into CD4+ T-cells, and the human leukocyte antigen (HLA) genes, which encode proteins that present certain host and pathogen-derived peptides at the surface of cells. These associations have been well-replicated in several GWAS \citet{Naranbhai2017}. This results suggest that nearly all host-determined variation in spVL can be explained by previously discovered variants. 
%
%In a joint analysis of both host and viral genotypes, \citet{Bartha2017} showed that nearly all of the variation in spVL explained by human genetic variation overlaps that explained by HIV genetic variation. If this is the case then our method to remove the viral contribution to spVL prior to GWAS should yield only associations between host variants and spVL that are independent of the virus. Additionally, effect size estimates for associations will reflect only the effect of the host variant, not any correlated effect from the virus.

%\subsection{New method to estimate the viral and non-viral parts of spVL}
%
%Here we derive a method to estimate the viral part of spVL for each infected host by assuming that spVL values at the tips of a phylogenetic tree $\mathcal{T}$ have evolved under the POUMM. In particular, for every tip $i$ in the tree, we assume that the host’s spVL value, $z_i$, is formed as a sum of three independent parts: a  heritable part determined by additive genetic effects from the virus, $g_{v_i}$, a non-heritable part determined by additive genetic effects from the host, $g_{h_i}$, and a further non-heritable part, $\epsilon_i$, which encompasses any other factor effecting spVL (Equation \ref{eq:totalspVL}). For the purposes of our derivation, it is convent to combine the non-heritable parts of spVL, $g_{h_i} + \epsilon_i$, into a single variable $e_i$.
%
%\begin{align} 
%z _i &= g_{v_i} + g_{h_i} + \epsilon_i \\
%\xrightarrow{e_i = g_{h_i} + \epsilon_i } z_i &= g_{v_i} + e_i  \label{eq:totalspVL}
%\end{align}
%
%According to the POUMM, the unobserved viral component of spVL for each host, $g_{v_i}$, evolves along the tree according to an Ornstein-Uhlenbeck (OU) process. The OU process was first introduced by \citet{Lande1976} as a model of evolution under both drift and stabilizing selection, so we are assuming $g_{v_i}$ continually evolves towards an evolutionary optimum in each viral lineage but is subject to random evolutionary fluctuations. Mathematically, $g_{v_i}$ is a realization of a Gaussian distributed random variable $G_{v_i} \sim \mathcal{N}(\mu_{OU_i}, \sigma^2_{OU_i})$, where:
%
%\begin{equation} \label{eq:muOU}
%\mu_{OU_i} = z_0e^{-\alpha t_{0i}} + (1 - e^{-\alpha t_{0i}})\theta \\
%\end{equation}
%\begin{equation}\label{eq:sigOU}
%	\sigma^2_{OU_i} = \sigma^2\frac{(1 - e^{-2\alpha t_{0i}})}{2\alpha}
%\end{equation}
%
%$z_0$ is the ancestral spVL value at the root of the tree, $\theta$ is the optimal spVL value that maximizes viral fitness,  $\alpha$ is the strength of selection on spVL, and $\sigma$ is the standard deviation of random fluctuations in $g_{v}$. We refer to these OU parameters ($z_0, \theta, \alpha, \textrm{and } \sigma$) collectively as $\Theta$. Finally, $t_{0i}$ is the phylogenetic distance from the root of the tree to tip $i$. We use the following notation to denote that the distribution of $G_{v}$ depends on the tree, $\mathcal{T}$, and the OU parameters $\Theta$: $G_{v|\Theta, \mathcal{T}}$. Generalizing to $N$ total tips in the tree, $\vec{G}_{v|\Theta, \mathcal{T}} = (G_{v_1}, ..., G_{v_N})^T$ is a random vector for the unobserved viral component of spVL at the tips. We know that $\vec{G}_{v|\Theta, \mathcal{T}} \sim \mathcal{N}(\vec{\mu}_{OU}, \mathbf{\Sigma}_{OU})$, where the elements of the variance-covariance matrix $\mathbf{\Sigma}_{OU}$ are calculated as follows:
%
%\begin{align} 
%\bm{\Sigma}_{OU_{ii}} &= \sigma^2\frac{(1 - e^{-2\alpha t_{0i}})}{2\alpha} \label{eq:SigmaOUii} \\
%\bm{\Sigma}_{OU_{ij}}&= \sigma^2\frac{e^{-\alpha\tau_{ij}}(1 - e^{-2\alpha t_{0(ij)}})}{2\alpha} \label{eq:SigmaOUij}
%\end{align}
%
%$\tau_{ij}$ is the phylogenetic distance between tips $i$ and $j$ and $t_{0(ij)}$ is the phylogenetic distance from the root of the tree to the most recent common ancestor of $i$ and $j$. Equations \ref{eq:muOU}, \ref{eq:SigmaOUii}, and \ref{eq:SigmaOUij} completely describe the generating process for the unobserved viral part of spVL under the assumptions of the POUMM. 
%
%On the other hand, the non-heritable component of spVL for each host, $e_i$ is assumed to be normally distributed in the population with an expected value of zero.  Furthermore, $e_i$ is uncorrelated among hosts and independent of the heritable viral part of spVL. Mathematically, $e_i$, is a realization of another Gaussian distributed random variable $E_i \sim \mathcal{N}(0, \sigma^2_e)$, which is independent of $G_{v_j}$ for any tip $j$. $\vec{E}_{|\sigma_e} = (E_1, ..., E_N)^T$ is a random vector for the unobserved non-viral part of spVL at the tips. The parameter $\sigma_e$ is the standard deviation of the non-viral component of spVL in the population. 
% 
% \begin{theorem}\label{thmGValues}
% 	Given the POUMM assumptions described, the pair of random vectors $<\vec{G}_{|\Theta,\mathcal{T},\sigma_e, \vec{z}},\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} >$ satisfies the following:
% 	\begin{enumerate}
% 		\renewcommand{\theenumi}{\alph{enumi}}
% 		\item Each realisation $\vec{g_v}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ of $\vec{G_v}_{|\Theta,\mathcal{T},\sigma_e, \vec{z}}$ is also a realization of $\vec{G}_{|\Theta, \mathcal{T}}$, i.e. it results from simulating the OU branching process defined by $\Theta$ on $\mathcal{T}$;
% 		\item $\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} = \vec{z} - \vec{G_v}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$, i.e. each realization $\vec{e}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ of $\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$  satisfies $\vec{z} = \vec{g}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} + \vec{e}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ for its corresponding realization $\vec{g}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$  of $\vec{G}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ in the same pair.
% 	\end{enumerate}
% 	Then the pdf of $<\vec{G}_{|\Theta,\mathcal{T},\sigma_e, \vec{z}},\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} >$ is proportional to a $N$-variate Gaussian function, i.e. 
% 	\begin{equation}\label{eqPdfGValues1}
% 	f\big(<\vec{g}_{|\\Theta,\mathcal{T},\sigma_e, \vec{z}}, \vec{e}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}>\big) \propto \mathcal{N}\big(\vec{g}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}\ ;\ \vec{\mu}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z}), \mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z})\big),
% 	\end{equation}
% 	\noindent where the matrix $\mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z})$ and the  vector $\vec{\mu}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z})$ are defined as follows:
% 	\begin{equation}\label{eqMuSigmaGValues}
% 	\begin{array}{rcl}
% 	\mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T},\sigma_e) & = & \bigg(\big[\mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\big]^{-1}+\mathbf{\Sigma}_{e}(\sigma_e)^{-1}\bigg)^{-1}\\
% 	\vec{\mu}_{MWA}(\Theta,\mathcal{T},\sigma_e, \vec{z}) & = & \mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T},\sigma_e) \bigg(\big[\mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\big]^{-1}\vec{\mu}_{OU}(\Theta,\mathcal{T}) +  \mathbf{\Sigma}_{e}(\sigma_e)^{-1} \vec{z}\bigg).
% 	\end{array}
% 	\end{equation}
% \end{theorem}
% \begin{proof}
% 	\begin{equation}\label{eqPdfGValues2}
% 	\begin{array}{rcl}
% 	f\big(<\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}},\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}>\big) & = & f\big(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ |\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big) \times f\big(\vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big)\\
% 	& = &\mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
% 	& &  \mathcal{N}\bigg(\vec{z} - \vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{0}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
% 	& = & \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
% 	& &  \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\  \vec{z}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
% 	\end{array}
% 	\end{equation}
% 	
% 	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
% 	
% \end{proof}
%
%Removing the viral part of spVL from total spVL using the above method results in the following ordinary least squares slope and intercept coefficient estimates for GWAS:
%
%\begin{equation} \label{eq: OLS-soln}
%	\hat{\beta}_i = (\mathbf{X}_i^T\mathbf{X}_i)^{-1}\mathbf{X}_i^T(\vec{z} - \vec{\mu}_{MWA})
%\end{equation}
%
%where $\mathbf{X}_i$ is the $N \times 2$ design matrix of host genotypes for genetic variant $i$.
%\begin{equation*}
%	\mathbf{X}_i = \begin{bmatrix}
%	1 & g_{i1}  \\
%	\vdots & \vdots \\
%	1 & g_{iN} 
%	\end{bmatrix}
%\end{equation*}
%
%%  serves three purposes. 1) We can estimate for each patient the effect the virus has on their spVL; 2) we can  account for non-independence between spVL samples due to the viral phylogeny; and 3) we can generate corrected estimates for each host genetic variant's effect on spVL, excluding any effect of the virus.
% 
%\subsection{PGLS}
%
%We made two notable changes to the typical PGLS framework (see \citealt{Garamszegi2014} for a nice overview of PGLS). First, we assumed a variance-covariance structure for regression residuals according to the POUMM. This matrix equals $\mathbf{\Sigma}_{OU} + \mathbf{\Sigma}_{e}$. Second, we use the variances on the diagonal of this matrix as sample weights to account for non-ultrametricity of the HIV tree. We implemented PGLS including these two modifications using the nmle package in R \parencite{Pinheiro2019}. The resulting general least squares slope and intercept coefficient estimates for GWAS are:
%
%\begin{equation}
%	\hat{\beta}_i = (\mathbf{X}_i^T(\mathbf{\Sigma}_{OU} + \mathbf{\Sigma}_e)^{-1}\mathbf{X}_i)^{-1}\mathbf{X}_i^T(\mathbf{\Sigma}_{OU} + \mathbf{\Sigma}_e)^{-1}\vec{z}
%\end{equation}

% \subsection{Comparison to PGLS}
% 
%We now adopt a second method to correct GWAS for the viral phylogeny by borrowing a method long used by evolutionary biologists. Phylogenetic generalized least squares (PGLS) regression is a method to account for phylogenetic non-independence between samples using generalized least squares regression. The method has typically been used to test evolutionary hypothesis regarding the relationships between different species traits, In method to estimate the variance-covariance structure of trait data from a  However, an alternate approach already exists to An alternate approach to accounting for non-independence between samples in GWAS would be to use phylogenetic least squares regression (PGLS). To the best of our knowledge, PGLS has not yet been applied to GWAS for infectious disease.
% 
%A key assumption of ordinary least squares (OLS) regression is that the residual differences between predicted and actual outcomes should be independent and identically distributed 
% 
%Given a linear model of assocation between host genetic variants and spVL, we compare the least-squares regression coefficient estimates using 1) our new method to estimate just the non-viral part of spVL and 2) PGLS to assume a phylogentically-informed correlation structure among residuals.

%The inferred POUMM parameters $\alpha$, $\sigma$, $\theta$, $g_{0}$, and $\sigma_e$ define an expected multivariate normal distribution for the viral part of spVL at each tip sample:
%
%\begin{equation} \label{eq:muOU}
%	\mu_{OU} = g_0e^{-\alpha t_{0i}} + (1 - e^{-\alpha t_{0i}})\theta \\
%\end{equation}
%
%\begin{align} 
%	\bm{\Sigma}_{ii} &= \sigma^2\frac{(1 - e^{-2\alpha t_{0i}})}{2\alpha} \label{eq:SigmaOUii} \\
%	\bm{\Sigma}_{ij} &= \sigma^2\frac{e^{-\alpha\tau_{ij}}(1 - e^{-2\alpha t_{0(ij)}})}{2\alpha} \label{eq:SigmaOUij}
%\end{align}
%
%with $t_{0i}$ equal to the root-to-tip distance on the phylogenetic tree and $t_{0(ij)}$ equal to the root-to-most recent common ancestor distance for samples $i$ and $j$. 
%Equation \ref{eq:muOU} describes the expected value for $g_v$ for each sample $i$. The variances in the variance-covariance matrix $\bm{\Sigma}_{OU}$, $\bm{\Sigma}_{ii}$ describe the uncertainty in each expected $g_v$ value, while the covariances $\bm{\Sigma}_{ij}$ describe how much shared ancestry each pair of samples has (Equations \ref{eq:SigmaOUii},  \ref{eq:SigmaOUij}).

% We know that $\vec{E}_{|\Theta}\sim \mathcal{N}\bigg(\vec{0}, \mathbf{\Sigma}_e(\Theta)\bigg)$, where $\mathbf{\Sigma}_{e}(\Theta)$ denotes the $N\times N$ diagonal matrix with each diagonal element equal to $\sigma_e^2$;
% We know that  $\vec{Z}_{|\Theta,\mathcal{T}}\sim \mathcal{N}\bigg(\vec{\mu}_{G}(\Theta_{OU},\mathcal{T}), \mathbf{\Sigma}_{G}(\Theta_{OU},\mathcal{T})+\mathbf{\Sigma}_{e}\bigg)$;
% 
%The parameters of the POUMM are $g_{0}$, the ancestral trait value for spVL, $\theta$, the optimal trait value for spVL to maximize viral fitness, $\alpha$, the strength of selection on spVL, $\sigma$, the time-unit standard deviation of random evolutionary fluctuations in spVL, and $sigma_e$, the standard deviation on the non-viral component of spVL in the population.

%The heritable component of spVL, $g_v$, is assumed to evolve continuously within HIV lineages according to an Ornstein-Uhlenbeck (OU) process. This assumption may be justified by imagining that the host part of spVL, $g_h$ is the sum of contributions from many alleles, each of which may have a positive or negative effect on spVL. Even in the standard GWAS framework, non-heritable noise in phenotype is assumed to be normally distributed with an expected value of zero, justifying our assumption that $e$ is distributed in this way.

%We estimate the viral part of spVL as a covariance-weighted average of the expected $g_v$ values under the POUMM ($\vec{\mu}_{OU}$) and observed spVL values, where the covariance structure for observed spVL is assumed to be a diagonal matrix with diagonal entries equal to the POUMM parameter $\sigma_e$ (Equation \ref{eq:MWA}).

%\begin{equation} \label{eq:MWA}
%\vec{\mu}_{MWA} = (\bm{\Sigma}_{OU}^{-1} + \bm{\Sigma}_{e}^{-1})^{-1} \times (\bm{\Sigma}_{OU}^{-1}\vec{\mu}_{OU} + \bm{\Sigma}_{e}^{-1}\vec{z})
%\end{equation}

%used a standard GWAS methodology. First, a large cohort of phenotypically diverse individuals was genotyped and phenotyped. Next, a separate regression model was fit for each of the millions of surveyed genetic markers. Under the most basic model, the phenotype is simply the sum of some additive genetic effect from the particular variant being tested and a residual term encompassing all other genetic and environmental determinants. Since a hypothesis test for significant association is performed for each of millions of genetic variants, GWAS performed in this way incurs a heavy multiple-testing burden. Despite a strong correction for multiple testing, \citet{McLaren2015} were able to confirm two

%So far, it is well established that both human host and viral genetic factors contribute to spVL. \citet{McLaren2015} found that 25\% of the variability in spVL in a mixed European cohort could be explained by additive genetic effects from the human genome. More recently, \citet{Mitov2018} established that at least 20\% of the variability in spVL in a UK cohort could be explained by viral genetic variation.  Furthermore, depending on the methodological assumptions made and the cohort studied, estimates for the broad-sense viral heritability can vary widely \parencite{Mitov2018}. Finally, human and viral genetic variants may actually explain redundant variation in spVL \parencite{Bartha2017}.

%Several GWAS have been performed to try to identify specific human and viral genetic variants that explain variation in spVL. The largest GWAS for human genetic factors so far, conducted by \citet{McLaren2015}, confirmed two regions of the human genome are associated with variability in spVL – . This study used a standard GWAS framework, in which each surveyed genetic variant (millions in total) was seperately tested for a significant linear relationship with spVL based on thousands of genotyped and phenotyped individuals. This method did not take into account heterogeneity in the viral population. On the viral side, \cite{Bartha2013} performed an analogous GWAS for viral genetic factors. After correcting for multiple testing, they did not find any viral genetic variants significantly associated with spVL. Combining genomic information from both humans and viruses, \cite{Bartha2013} also tested for pairwise associations between human genetic variants and viral genetic variants. In this way the authors could pinpoint sites of genomic conflict where the host's particular HLA alleles selected for certain viral escape mutations.

%So far, the GWAS for human genetic variants associated with spVL have not accounted for viral genetic variation in a holistic way. We wondered whether we could improve the signal of association between known regions of the human genome associated with spVL or discover new associations with spVL by explicitly taking viral genetic variation into account in the GWAS model.

%\subsection{Previous GWAS approaches}
%
%Accounting for extraneous factors in GWAS is not a new idea; a broad range of methods have previously been proposed to adjust the simplistic GWAS framework described. These adjustments typically have two purposes, either to reduce residual variance in the phenotype or to account for non-independence between samples. The reduction of residual variance is often accomplished by including covariates in the regression model that describe extraneous factors influencing the measured phenotype. So rather than regressing phenotype on allele dosage alone, a researcher might regress phenotype on allele dosage, participant age, and assay type.
%
%Non-independence between samples in GWAS is often thought of only in terms of genetic relatedness of sampled individuals, in which case it is called \emph{population structure}. This broad term encompasses a continuum of relatedness. At one extreme, individuals in a GWAS may share distant common ancestry and at the other extreme, they may be close family relatives. If the phenotype of interest is differently distributed among ancestry or family groups, genetic variants that are common among members of a group simply through inheritance may appear associated with the phenotype of interest, despite not having any causal role in determining the phenotype. 
%
%Several approaches have been commonly used to account for population structure in GWAS, including adjusting the GWAS significance threshold to account for potential spurious findings \parencite{Devlin1999}, performing GWAS only on more homogenous population sub-groups \parencite{Pritchard2000}, and adjusting the regression model itself to include estimates of individual relatedness \parencite{Price2006, Kang2010}. Of these approaches, we highlight the linear mixed model developed by \citet{Kang2010} because it is an increasingly popular technique that has a direct analog in the field of phylogenetic comparative analysis, from which we derive our method to account for pathogen relatedness.
%
%\subsection{New methods to correct GWAS for the viral phylogeny}
%
%Since spVL is heritable from one HIV infection partner to another, we expect spVL samples to have a correlation structure based on the underlying HIV transmission tree. By estimating the transmission tree via the viral phylogeny (which we can infer from HIV sequencing data), we can develop an estimate for the correlation structure of spVL samples. Based on this idea, we propose two methods to account for the correlation structure among spVL samples due to the viral phylogeny.
%
%First, we try to partition measured spVL for each patient into a part determined by the virus (``viral part") and a part determined by non-viral factors (``non-viral part", e.g. additive genetic effects from the host, age, sex, assay error, etc.) Once we have partitioned spVL values in this way we can perform GWAS using only the non-viral part of spVL, effectively removing non-independence and noise due to the viral phylogeny from spVL samples. To partition spVL into viral and non-viral parts, we rely on the a phylogenetic mixed model (PMM, \citealt{Housworth2004, Lynch1991}). Under the PMM, a continuous trait is assumed to be composed of two independent parts: a heritable part and a non-heritable part. The heritable part is transmitted from parent to offspring and is subject to the long-term evolutionary forces of selection and drift. In contrast, the non-heritable part of a trait is independent among samples and may reflect the influence of local conditions or rapid evolution unconstrained by phylogenetic background. Several mathematical models for evolution of the heritable part can be applied in order to estimate the relative contributions of heritable and non-heritable parts to the total trait value. It has recently been shown that an Ornstein-Uhlenbeck (OU) process is the best-fitting model for evolution of the heritable viral part of spVL \parencite{Mitov2018} \todo{Justify use of OU model better}. The mixed model of evolution including an OU process for the heritable part is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM, \citealt{Mitov2017a}). By fitting such a model to spVL measurements and a viral phylogeny, we can co-infer the OU parameters that describe the evolutionary trajectory of the viral part of spVL as well as the variance of the non-heritable, non-viral part of spVL in the population. Using these inferred parameters, we develop a method to estimate for each sample the viral part of spVL (see ``Methods"). After partitioning each sampled spVL value into a viral and a non-viral part in this way, we can perform a modified GWAS in which only the non-viral part of spVL is regressed on variant dosage. 
%
%Rather than pre-process spVL values to remove the viral part as just explained, a second method to correct GWAS for a correlation structure among spVL samples would be to directly modify the regression formula using phylogenetic generalized least squares (PGLS) regression. As previously mentioned, a typical GWAS for host determinants of spVL fits a linear regression model for each candidate host genetic variant. Under this model, spVL is regressed on variant dosage and ordinary least squares regression is used to generate slope and intercept estimates. However, a key assumption of ordinary least squares regression is that residual error should be independent among samples. Since spVL samples are not independent but are instead related according to the underlying viral phylogeny, the residuals may not be independent either. To account for non-independence among residuals, \citet{Grafen1989} suggested the PGLS method, in which the phylogeny is translated into a variance-covariance structure among samples. It is assumed that residual errors are not independent, but instead follow this phylogenetic variance-covariance structure. General least squares regression can then be used to generate phylogenetically-corrected slope and intercept estimates for the regression of spVL on variant dosage. We compare our first method for a phylogenetically-corrected GWAS to PGLS using a variance-covariance structure derived from the same modelling assumptions. Namely, we perform PGLS using a variance-covariance structure generated by fitting the POUMM to spVL data and the viral phylogeny.

%In infectious disease GWAS, the pathogen's effect on phenotype is often grouped with other, non-genetic effects as residual phenotypic variation. However, several methods have recently been developed to tailor GWAS to the study of microbial genomes, including pathogen genomes.

%Intuitively, this information gives us a measure of how surprised we should be to observe particular spVL values given the underlying HIV phylogeny. When we observe spVL values that are very surprising given the sampled individual's location on the phylogeny, we can infer that non-viral factors are contributing to the deviation from our expectation. For a more rigorous description of our POUMM-based method to estimate the viral and non-viral parts of spVL, see the ``Methods" section.

%\todo{Add justifying citations to this paragraph}At this point we should address several of what we feel are the most important assumptions made by the POUMM. Fundamentally, we assume that the spVL phenotype is the sum of independent heritable (i.e. viral) and non-heritable (i.e. non-viral) parts. This neglects the possibility of of interaction effects between the virus and the host or the host's environment. We also assume there is no discontinuity in the virus between infection partners. This assumption ignores the possibility of quasi-species transmission, in which only a subset of the potentially heterogeneous viral population circulating in the infecting host is transmitted to their infection partner. \todo{To be continued with other assumptions, need to think about what impact they would have}

%\subsection{Accounting for the viral phylogeny in G2G GWAS}
%So far we have discussed confounding in HIV GWAS due to human (host) population factors. However, in certain infectious disease  contexts the viral population can also confound GWA analyses. In the genome-to-genome (G2G) GWAS framework described by \citet{Bartha2013}, host variants are tested for pairwise associations with viral variants to determine sites of genetic conflict between the human host and HIV pathogen. In this case, confounding may arise from population stratification in both the host and viral populations (Figure \ref{fig:gwas-covariates}b). \cite{Naret2018} describe two potential types of stratification that may occur in the G2G context. So-called ``major stratification" occurs when host population is correlated with viral population and vice-versa. ``Minor stratification"  occurs when host population is correlated with with viral variant frequencies or when viral population is correlated with host variant frequencies. Although \cite{Naret2018} envision only discrete viral populations, the same phenomena could easily apply when considering the full viral phylogeny. 
%
%Several methods have been developed to correct for confounding from the viral phylogeny in HIV G2G GWAS. One of two methods originally introduced by \citet{Bhattacharya2007} involves determining whether viral mutations are inferred to occur disproportionately along branches of the phylogeny where a particular host variant is present. The other method is a model-selection approach in which phylogenetic models with and without selection pressure from host variants are compared for goodness-of-fit \parencite{Bhattacharya2007, Carlson2008}. Lastly, the more recently proposed method by \citet{Naret2018} uses phylogenetic principle components to describe the major axes of variation in the viral population (an approach analogous to using principle components to account for host population structure). However, the method proposed by \citet{Naret2018} does not have a secondary benefit of the other two methods, which is to restore statistical independence to samples that are fundamentally related by a phylogeny. 
%
%\subsection{Towards a full correlation model for HIV GWAS}
%
%As \citet{Bhattacharya2007} recognized, HIV samples are related by descent according to the transmission tree. Such relationships are often investigated by inferring a viral phylogeny from sequencing data. Recently, several studies using phylogentic methods determined that host spVL values are correlated among hosts with closely related versions of the virus \parencite{Alizon2010, Blanquart2017, Hollingsworth2010, Mitov2018}. This correlation can be quantified in terms of the heritability of spVL, or, the fraction of variance in spVL among hosts that can be explained by underlying viral genetic variation. \citet{Mitov2018} demonstrated that spVL is at least 20\% heritable, meaning that the virus has a strong effect on host spVL. Based on this correlation between the virus and spVL (Figure \ref{fig:gwas-covariates}c), we argue it is necessary to expand the correlation model typically assumed in standard GWAS. Expanding upon previous work in the G2G GWAS framework, we develop a full correlation model for spVL GWAS (Figure \ref{fig:gwas-covariates}d). From the standard GWAS framework, we take that host variant frequencies are correlated with host population and that host population is correlated with spVL. From the G2G framework, we take that the viral phylogeny is correlated with host population. Finally, we assume that the viral phylogeny effects spVL based on recent spVL heritability studies. If we assume that host population stratification and noise factors are already taken into account in GWAS for host determinants of spVL, we may reduce the full correlation model to Figure \ref{fig:gwas-covariates}e. It is then apparent that accounting for host population stratification is sufficient to control for confounding effects of the viral phylogeny. However, the viral phylogeny remains both a source of noise and a cause of non-independence between samples in GWAS for host determinants of spVL.
%
%\begin{figure}[h!]
%	\begin{tikzpicture}
%	\begin{scope}[xshift = 0cm, yshift = 0cm]
%	\draw (-.75, 2) node {\textbf{a) standard GWAS}}; 
%	\node[smod, minimum size = 3.75cm, minimum width = 3cm] (a) at (-0.75, -0.75/2) {};
%	\node[snode] (z) at (0, 0) {$z$};
%	\node[snode] (ch) at ($(z) + (-1.5, +.75)$) {$P$};
%	\draw[taro] (ch) -- (z);
%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
%	\draw[taro] (gh) -- (z);
%	\node[snode] (n) at ($(z) + (0, -1.5)$) {$n$};
%	\draw[taro] (n) -- (z);
%	\draw[bline] (gh) -- (ch);
%	\end{scope}
%	
%	\begin{scope}[xshift = 5cm, yshift = 0cm]
%	\draw (0.2, 2) node {\textbf{b) Genome-to-genome GWAS}}; 
%	\node[smod, minimum size = 3cm, minimum width = 4.5cm] (a) at (0, 0) {};
%	\node[white] (z) at (0, 0) {$z$};  % White (hidden) node z to anchor coordinate system
%	\node[snode] (ch) at ($(z) + (-1.5, +.75)$) {$P$};
%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
%	\node[snode] (gv) at ($(z) + (+1.5, -.75)$) {$g_v$};
%	\draw[taro] (gh) -- (gv);
%	\draw[bline] (ch) -- (gh);
%	\draw[bline] (cv) -- (gv);
%	\draw[bline] (ch) -- (cv);
%	\draw[dline] (ch) -- (gv);
%	\draw[dline] (cv) -- (gh);
%	\end{scope}
%	
%	\begin{scope}[xshift = 10cm, yshift = 0cm]
%	\draw (0.9, 2) node {\textbf{c) spVL Heritability}}; 
%	\node[smod, minimum size = 2.25cm, minimum width = 3cm] (a) at (0.75, 0.75/2) {};
%	\node[snode] (z) at (0, 0) {$z$};
%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
%	\draw[taro] (cv) -- (z);
%	\end{scope}
%	
%	\begin{scope}[xshift = 0cm, yshift = -5cm]
%	\draw (0, 2) node {\textbf{d) Full correlation model}}; 
%	\node[smod, minimum size = 3.75cm, minimum width = 4.5cm] (a) at (0, -.75/2) {};
%	\node[snode] (z) at (0, 0) {$z$};
%	\node[snode] (ch) at ($(z) + (-1.5, +.75)$) {$P$};
%	\draw[taro] (ch) -- (z);
%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
%	\draw[taro] (gh) -- (z);
%	\node[snode] (n) at ($(z) + (0, -1.5)$) {$n$};
%	\draw[taro] (n) -- (z);
%	\draw[bline] (gh) -- (ch);
%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
%	\draw[taro] (cv) -- (z);
%	\draw[bline] (cv) -- (ch);
%	\end{scope}
%	
%	\begin{scope}[xshift = 6.5cm, yshift = -5cm]
%	\draw (1.5, 2) node {\textbf{e) Without standard GWAS covariates}}; 
%	\node[smod, minimum size = 3cm, minimum width = 4.5cm] (a) at (0, 0) {};
%	\node[snode] (z) at (0, 0) {$z$};
%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
%	\draw[taro] (gh) -- (z);
%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
%	\draw[taro] (cv) -- (z);
%	\end{scope}
%	\end{tikzpicture}
%	\caption{Correlation and dependency structures for HIV GWAS. Solid lines are correlations between variables and arrows show directional dependency. Dotted lines indicate minor stratification (see section "Accounting for the viral phylogeny...").
%		a) The host population ($P$) is a confounding factor for the association between host variants ($g_h$) and spVL ($z$). Noise factors ($n$) contribute residual variance to spVL. 
%		b) The host population and the viral phylogeny ($\Psi$) are both confounding factors for the association between host and viral ($g_v$) variants. 
%		c) spVL depends on the viral phylogeny.
%		d) Combining models a-c yields the full correlation model.
%		e) Model d may be reduced by controlling for host population stratification and noise factors}
%	\label{fig:gwas-covariates}
%\end{figure}
%
%\todo{Naret doesn't have any justification for minor stratification - how could this arise?}
%

%
%
%# LEFT OFF HERE
%
%\subsection{The standard GWAS framework}
%
%GWA studies conduct an unbiased scan of the genome for genetic variants associated with a phenotype of interest.  Although GWAS are frequently used to explore the genetic architecture of complex phenotypes, they suffer from several known sources of error. We discuss two such sources of error now: confounding from latently associated variables (a source of false positive associations) and high residual variance (a source of false negative associations). 
%
%The first source of error, confounding, arises when a latent variable effects the phenotype of interest and is also correlated with genetic variants being tested (Figure \ref{fig:gwas-covariates}a). A classic example of a confounding variable for GWAS is ancestry. If allele frequencies differ between human populations of different ancestry and phenotypes also differ between the same populations, false positive associations may result since ancestry-associated genetic variants will appear to explain variation in phenotype that is truly due to ancestry. This phenomenon is called population stratification and several methods have been developed to address the problem. In the context of GWAS for host determinants of spVL, a principle components method \cite{Price2006} has most often been applied. In this method, the top few principle components of host genetic variation are included in the association model as proxy variables for demographic and geographic factors. This effectively changes the association test hypothesis from ``does the host variant explain a significant amount of variation in spVL?", to ``does the host variant explain a significant amount of variation in spVL \emph{beyond that which can be explained by demographic or geographic factors?}" Previous GWAS for host determinants of spVL conducted by \citet{Price2006, McLaren2015, Bartha2013, Fellay2007, Pereyra2010, Pelak2010, VanManen2011} have employed this method. Several studies have additionally used other methods to account for host population stratification, including structured association \parencite{Pritchard2000} and genomic control \parencite{Devlin1999}. In structured association, the sample population is a priori divided into groups that are assumed to be homogeneous with respect to potentially confounding variables. GWA testing is then conducted independently within each group before results are finally aggregated across groups. \citet{McLaren2015, Pereyra2010} both applied a form of structured association. So far only \citet{Dalmasso2008} have used genomic control to account for population stratification in spVL GWAS. In genomic control, the significance threshold for association testing is uniformly adjusted for all variants by a factor which accounts for the degree of stratification present. Although methods to account for population stratification vary, all GWAS for host determinants of spVL to-date have taken the problem into account. 
%
%The second source of error typically considered in GWAS for host determinants of spVL arises from ``noise" variables that contribute extra variance to the phenotype of interest. These variables effect the phenotype, but are not correlated with the genetic variant being tested (Figure \ref{fig:gwas-covariates}a). As long as the phenotype is quantitative (as in the case of spVL) or the sample is randomly drawn from the population, including these factors as covariates in the association model helps remove residual variance and increase power to detect true associations \todo{Investigate this claim further} \parencite{Mefford2012}. In the context of GWAS for host determinants of spVL, sex, cohort, and age have all been recognized to be noise variables that effect spVL but should not be correlated with host genetic variants \parencite{Bartha2013, Dalmasso2008, Fellay2007, Fellay2009, Pelak2010}.

%Since the viral and non-viral parts of spVL are unobservable, we developed a simulation model under the POUMM assumptions as a simplistic test of our method. Figure \ref{fig:sim-design} provides a graphical overview of the complete model. When possible, we tried to parameterize the model using values from previously reported HIV cohort data. 
%
%First, we fixed the total variance in simulated spVL to be 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \parencite{Mitov2018}. After allotting 25\% of this variance to $g_h$ (following \citealt{McLaren2015}), we partitioned the remaining variance between $g_v$ and $\epsilon$ in different ratios in order to assess our estimator’s performance across a range of possible spVL heritability values (Figure \ref{fig:sim-design}a, Table \ref{tab:sim-params}). Second, we generated the host part of spVL by summing contributions from $M$ causal host variants (Figure \ref{fig:sim-design}b). \todo{Add binomial description to figure} Host genotypes were simulated according to a binomial distribution with a constant allele frequency shared across all variants. Third, we generated $g_v$ by simulating an Ornstein-Uhlenbeck (OU) process of evolution along a viral phylogenetic tree (Figure \ref{fig:sim-design}c). We use a randomly generated tree with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \parencite{Hodcroft2014}. To test under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively. For the other OU parameters, we used previously inferred values from \parencite{Bertels2018, Blanquart2017, Mitov2018}. Finally, $\epsilon$ values were generated from a normal distribution with mean zero.

%\begin{proof}
%	\begin{equation}\label{eqPdfGValues2}
%	\begin{array}{rcl}
%	f\big(<\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}},\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}>\big) & = & f\big(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ |\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big) \times f\big(\vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big)\\
%	& = &\mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
%	& &  \mathcal{N}\bigg(\vec{z} - \vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{0}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
%	& = & \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
%	& &  \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\  \vec{z}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
%	\end{array}
%	\end{equation}
%	
%	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
%\end{proof}

%\begin{table}[H]
%	\begin{center}
%		\begin{tabular}{c|l|c|c}
%			\toprule 
%			\textbf{Parameter} & \textbf{Description} & \textbf{Posterior mean} & \textbf{95\% HPD} \\
%			\midrule 
%			$z_0$ & spVL value at the root of the transmission tree & 4.9 & 2.8, 7.3 \\
%			$\theta$ & evolutionarily optimal spVL value & 4.3 & 4.2, 4.4 \\
%			$\sigma$ & time-unit standard deviation of the OU process & 4.9 & 3.5, 6.4 \\
%			$\alpha$ & selection strength & 44.5 & 28.4, 60.9 \\
%			$\sigma_e$ & standard deviation of the non-heritable part of spVL & 0.56 & 0.50, 0.63 \\
%			\toprule 
%		\end{tabular}
%		\caption{POUMM parameters.}
%		\label{tab:POUMMParamDesc}
%	\end{center}
%\end{table}

%\begin{table}[H] \label{tab:comp-gwas}
%	\begin{center}
%	\begin{tabular}{llclccc} \toprule 
%	& \multicolumn{2}{c}{\specialcell[l]{\citet{McLaren2015} study \\ N = 6,315}} & \multicolumn{3}{c}{\specialcell[l]{Current study \\ N = 1,370}} & \\ \midrule 
%	Region & Variant & \emph{P} value & Variant & Effect size & \emph{P} value & R$^2$ \\ \midrule
%	MHC & rs59440261 & 2.0 x 10$^{-83}$ & rs202036493 & \specialcell[t]{-0.31 \\ (-0.52)} & \specialcell[t]{1.3 x 10$^{-13}$ \\  (9.2 x 10$^{-15}$)} & 1 \\
%	% MHC & rs59440261 & 2.0 x 10$^{-83}$ & rs2923003 & -0.31 & 1.26511 x 10$^{-13}$ & 1 \\
%	% MHC & - & - & rs112243036 & -0.20 & 2.37153 x 10$^{-14}$ & 0.592 \\
%	\emph{CCR5} & rs1015164 & 1.5 x 10$^{-19}$ & rs7637813 & \specialcell[t]{0.09 \\ (0.17)} & \specialcell[t]{2.9 x 10$^{-6}$ \\  (2.0 x 10$^{-7}$)} & 0.867 \\
%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs2187671 & \specialcell[t]{0.08 \\ (0.15)} & \specialcell[t]{4.7 x 10$^{-5}$ \\  (4.9 x 10$^{-6}$)} & 0.867 \\
%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs34602794 & \specialcell[t]{0.08 \\ (0.15)} & \specialcell[t]{4.7 x 10$^{-5}$ \\  (4.9 x 10$^{-6}$)} & 0.867 \\
%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs2157056 & \specialcell[t]{0.08 \\ (0.15)} & \specialcell[t]{4.7 x 10$^{-5}$ \\  (4.9 x 10$^{-6}$)} & 0.867 \\
%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs4683217 & \specialcell[t]{0.08 \\ (0.14)} & \specialcell[t]{1.4 x 10$^{-4}$ \\  (1.3 x 10$^{-5}$)} & 0.867 \\
%	% NOTE: these p-values are from analysis /Volumes/stadler/SHCSData/analysis/200127/1500*
%	 \toprule
%	\end{tabular}
%	\caption{Association results for the top SNPs in the MHC and \emph{CCR5} regions from the GWAS by \citet{McLaren2015} compared to association results for the best proxy SNPs for these SNPs available in the current study. The proxy SNPs were chosen based on high linkage disequilibrium (as measured by R$^2$) with the McLaren et al. SNPs. For the current study, values without parentheses are from the GWAS with phylogenetic correction and values in parentheses are from the standard GWAS.}
%	\end{center}
%\end{table}

%\begin{table}[H]
%	\begin{center}
%		\begin{tabular}{c|c|c|c}
%			\toprule 
%			\textbf{Filter} & \textbf{Current study} & \textbf{\citealt{Bartha2013}} & \textbf{\citealt{McLaren2015}} \\
%			\midrule 
%			SNP missingness & $>$ 0.01 & $>$ 0.01 &  $>$ 0.05 \\
%			Genotype missingness & $>$ 0.02 & Not specified & $>$ 0.02 \\
%			IBD & $>$ 0.125 & Not specified & $>$ 0.125 \\ 
%			SNP minor allele frequency & $<$ 0.01 & $<$ 0.01 & $<$ 0.01 \\
%			Deviation from Hardy-Weinberg equilibrium & p $<$ 5E-5 & p $<$ 5E-5 & p $<$ 1E-7 \\
%			Inbreeding coefficient & NA & NA & $<$ -0.1 or $>$ 0.1 \\
%			\midrule
%			Remaining samples & 2125 & 698 & 6315 \\
%			Remaining variants & $\sim$ 4.4 million & $\sim$ 7 million & $\sim$ 8 million \\
%			\bottomrule 
%		\end{tabular}
%		\caption{Genotype data quality control prior to GWAS.}
%		\label{tab:GWASQC}
%	\end{center}
%\end{table}

%\begin{table}[H]
%	\centering
%	\begin{tabular}{rlrrr}
%		\hline
%		& stat & PostMean & HPD\_lower & HPD\_upper \\ 
%		\hline
%		1 & alpha & 30.15 & 10.15 & 49.83 \\ 
%		2 & theta & 4.38 & 4.25 & 4.55 \\ 
%		3 & sigmae & 0.62 & 0.56 & 0.68 \\ 
%		4 & g0 & 3.98 & 3.14 & 4.84 \\ 
%		5 & sigma & 3.56 & 2.04 & 5.15 \\ 
%		\hline
%	\end{tabular}
%	\caption{Pipeline rerun 2 results.}
%\end{table}

%\begin{table}[ht]
%	\centering
%	\begin{tabular}{rlrrr}
%		\hline
%		& stat & PostMean & HPD\_lower & HPD\_upper \\ 
%		\hline
%		1 & alpha & 30.31 & 10.54 & 51.07 \\ 
%		2 & theta & 4.38 & 4.24 & 4.55 \\ 
%		3 & sigmae & 0.62 & 0.56 & 0.68 \\ 
%		4 & g0 & 3.99 & 3.15 & 4.79 \\ 
%		5 & sigma & 3.58 & 2.07 & 5.17 \\ 
%		\hline
%	\end{tabular}
%	\caption{Pipeline rerun diff POUMM seed results.}
%\end{table}
%
%\begin{table}[ht]
%	\centering
%	\begin{tabular}{rlrrr}
%		\hline
%		& stat & PostMean & HPD\_lower & HPD\_upper \\ 
%		\hline
%		1 & alpha & 33.15 & 14.66 & 53.74 \\ 
%		2 & theta & 4.37 & 4.25 & 4.49 \\ 
%		3 & sigmae & 0.62 & 0.56 & 0.68 \\ 
%		4 & g0 & 4.49 & 3.30 & 5.67 \\ 
%		5 & sigma & 3.69 & 2.17 & 5.18 \\ 
%		\hline
%	\end{tabular}
%	\caption{Pipeline rerun diff tree results.}
%\end{table}

%\begin{table}[H] \label{tab:comp-gwas}
%	\begin{center}
%		\begin{tabular}{p{1cm}p{2cm}p{1cm}p{2.5cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr} {llclcccc}
%			\hline
%			& & & McLaren et al.  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
%			\hline
%			Region & Variant & R$^2$ & p-value & Beta & p-value & Beta & p-value \\
%			\hline
%			MHC & rs59440261 & - & 2.0 x 10$^{-83}$ & -0.41 & 1.8x10$^{-11}$ & -0.25 & 3.3x10$^{-10}$ \\
%						& rs202036493 & 1 & - & -0.52 & 9.2 x 10$^{-15}$ & -0.31 & 1.3 x 10$^{-13}$ \\
%			\emph{CCR5} & rs1015164 & - & 1.5 x 10$^{-19}$ & 0.15 & 1.1x10$^{-6}$ & 0.08 & 3.0x10$^{-5}$ \\
%						& rs7637813 & 0.867&  - & 0.17 & 2.0 x 10$^{-7}$ & 0.09 & 2.9 x 10$^{-6}$  \\
%			\hline
%		\end{tabular}
%		\caption{Top association results from \citet{McLaren2015} compared to association results for the best proxy SNPs available in the current study. Proxy SNPs were chosen based on linkage disequilibrium (as measured by R$^2$) with the McLaren et al. SNPs.}
%	\end{center}
%\end{table}

### 2021-10-27 scripts/R/filter_gwas_individuals.R
improve alignment generation, sequence filtering

@@ -3,6 +3,7 @@
 
 library(dplyr)
 library(gridExtra)
+library(ggplot2)
 
 OUTDIR <- "output"
 DATADIR <- "/Volumes/stadler/SHCSData/data"
@@ -70,10 +71,9 @@ write.csv(x = gwas_individual_filtering_summary,
           row.names = F)
 
 gwas_individuals_to_keep <- gwas_metadata %>% filter(is_european, Pop == "Test")
-print(paste("Keeping", length(unique(gwas_individuals_to_keep$id)), "individuals for GWAS based on ancestry, viral subtype."))
+print(paste("Keeping", length(unique(gwas_individuals_to_keep$id)), "individuals for GWAS."))
 
 write.csv(
   x = gwas_individuals_to_keep %>% select(id),
   file = paste(OUTDIR, "gwas_individuals_to_keep.csv", sep = "/"),
-  row.names = F,
-  col.names = F)
+  row.names = F)

### 2021-10-27 scripts/R/filter_sequences.R

@@ -2,14 +2,14 @@
 # Data file is newfasta2019-10-25.fas which contains 2012 pol sequences corresponding to 2012/2125 genotyped samples (the remaining host genotypes are not associated with a pol sequence)
 # The sequences are not aligned (some are longer than others) even though gap characters have been introduced --> what does this mean??
 # Dates in the fasta file correspond to "sampledate" in the subtyping data.
-# Run as: Rscript scripts/R/filter_alignment.R
+# Run as: Rscript scripts/R/filter_sequences.R
 
 library(dplyr)
 library(ape)
 library(lubridate)
 library(tidyr)
 
-OUTDIR <- "../../output"
+OUTDIR <- "output"
 DATADIR <- "/Volumes/stadler/SHCSData/data"
 FASTA_FN <- "newfasta2019-10-25.fas"
 SPVL_FN <- "spvl_values.csv"
@@ -55,31 +55,51 @@ tip_data_spvl <- merge(
   by = "id",
   all.x = T, all.y = F)
 
+# Get length of sequences (non-gap characters)
+get_seq_length <- function(seq) {
+  non_gap_seq <- seq[!(seq %in% c("n", "N", "-"))]
+  return(length(non_gap_seq))
+}
+
+n_chars_in_seqs <- unlist(lapply(FUN = get_seq_length, X = as.character(sequences)))
+seq_length_data <- data.frame(n_chars_in_seq = n_chars_in_seqs, orig_label = names(n_chars_in_seqs))
+tip_data_spvl_with_length <- merge(x = tip_data_spvl, y = seq_length_data, by = "orig_label", all = T)
+
 # Write out summary of sequence data
-seq_subtype_summary <- tip_data_spvl %>%
+seq_subtype_summary <- tip_data_spvl_with_length %>%
   group_by(subtypes) %>%
-  summarize(n_seqs = n(), n_seqs_with_spvl = sum(!is.na(spvl_lenient_filter_mean_measurement))) %>%
+  summarize(n_seqs = n(),
+            n_seqs_with_spvl = sum(!is.na(spvl_lenient_filter_mean_measurement)),
+            n_long_enough_seqs_with_spvl = sum(!is.na(spvl_lenient_filter_mean_measurement) & n_chars_in_seq > 750)) %>%
   arrange(desc(n_seqs))
 
-seq_date_summary <- tip_data_spvl %>%
+seq_date_summary <- tip_data_spvl_with_length %>%
[…]

### 2021-10-27 scripts/R/root_tree.R

@@ -1,0 +1,48 @@
+# This script is to root the HIV tree, then write out the tree w/o outgroup.
+# Run as: Rscript scripts/R/root_tree.R
+library(dplyr)
+library(ggplot2)
+library(ggtree)
+library(tidyr)
+
+OUTDIR <- "output"
+DATADIR <- "/Volumes/stadler/SHCSData/data"
+TREE_FN <- "pathogen.treefile"
+
+paste("loading tree")
+tree <- ape::read.tree(file = paste(OUTDIR, TREE_FN, sep = "/"))
+
+# Parse outgroup info from tip labels
+tip_data <- data.frame(
+  label = tree$tip.label) %>%
+  tidyr::separate(label, into = c("id", "sampledate", "is_outgroup", "trait"), sep = "_", remove = F)
+
+# Plot tree with outgroup
+p <- ggtree(tr = tree) %<+% tip_data +
+  geom_tippoint(aes(color = is_outgroup))
+
+ggsave(plot = p, paste(OUTDIR, "pathogen_tree_with_outgroup.png", sep = "/"), height = 10, width = 7, units = "in")
+
+# Plot tree with trait
+p2 <- ggtree(tr = tree) %<+% tip_data +
+  geom_tippoint(aes(color = as.numeric(trait))) +
+  scale_color_gradient(low = "green", high = "red", name = "Trait value")
+
+ggsave(plot = p2, paste(OUTDIR, "pathogen_tree_with_trait.png", sep = "/"), height = 10, width = 7, units = "in")
+
+# Root the tree
+paste("Rooting tree according to outgroup and writing to file")
+outgroup_ids <- as.character(tip_data[tip_data$is_outgroup == "outgroup", "label"])
+rooted_tree <- ape::root(
+  phy = tree,
+  outgroup = outgroup_ids)
+
+# Write out the rooted tree with only B seqeunces for POUMM fitting
+rooted_tree_no_outgroup <- ape::drop.tip(
+  phy = rooted_tree,
+  tip = outgroup_ids)
+
+treename <- "pathogen_no_outgroup.newick"
+ape::write.tree(
+  phy = rooted_tree_no_outgroup,
+  file = paste(OUTDIR, treename, sep = "/"))

### 2021-10-27 scripts/bash/build-tree.sh

@@ -9,9 +9,13 @@
 -fasta \
 -maxiters 3
 
+# Trim the alignment, first removing wrapping
+sed -e 's/\(^>.*$\)/#\1#/' output/pathogen.fasta | tr -d "\r" | tr -d "\n" | sed -e 's/$/#/' | tr "#" "\n" | sed -e '/^$/d' > output/pathogen_oneline.fasta
+cut -c 1-1505 output/pathogen_oneline.fasta > output/pathogen_trimmed.fasta
+
 # Build a ML tree with the GTR+F+R4 model and (TODO) 100 bootstrap replicates
 iqtree \
--s output/pathogen.fasta \
+-s output/pathogen_trimmed.fasta \
 -m GTR+F+R4 \
 -redo \
 -pre output/pathogen \

### 2021-10-28 scripts/R/filter_gwas_individuals.R
change output format to txt not csv for PLINK

@@ -73,7 +73,10 @@ write.csv(x = gwas_individual_filtering_summary,
 gwas_individuals_to_keep <- gwas_metadata %>% filter(is_european, Pop == "Test")
 print(paste("Keeping", length(unique(gwas_individuals_to_keep$id)), "individuals for GWAS."))
 
-write.csv(
+write.table(
   x = gwas_individuals_to_keep %>% select(id),
-  file = paste(OUTDIR, "gwas_individuals_to_keep.csv", sep = "/"),
-  row.names = F)
+  file = paste(OUTDIR, "gwas_individuals_to_keep.txt", sep = "/"),
+  row.names = F,
+  col.names = F,
+  sep = "\t",
+  quote = F)

### 2021-10-28 scripts/R/fit_poumm.R
add script to estimate parameters of the POUMM using VM's R package

@@ -1,0 +1,92 @@
+# This script is to fit the POUMM.
+# Run as: Rscript scripts/R/fit_poumm.R
+
+OUTDIR <- "output"
+
+N_SAMPLES_MCMC <- 4E6
+N_ADAPT_MCMC <- 2E5
+THIN_MCMC <- 1000
+ACCEPTANCE_RATE_MCMC <- 0.01
+TREE_FN <- paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/")
+SEED <- 1500
+
+require(parallel)
+require(ape)
+require(tidyr)
+require(POUMM)
+require(bayesmeta)
+require(la)
+
+set.seed(SEED)  # Seed for POUMM parameter inference
+
+# Write out log file
+con <- file(paste(OUTDIR, "fit_poumm.log", sep = "/"))
+open(con, "w")
+writeLines(
+  con = con,
+  text = paste(
+    "n_samples_mcmc", N_SAMPLES_MCMC, "\n",
+    "n_adapt_mcmc", N_ADAPT_MCMC, "\n",
+    "thin_mcmc", THIN_MCMC, "\n",
+    "acceptance_rate_mcmc:", ACCEPTANCE_RATE_MCMC, "\n"))
+close(con)
+
+# Load the tree (Important: tree must be without outgroup)
+print("loading the tree")
+tree <- ape::read.tree(file = TREE_FN)
+
+# Parse trait data from tip labels
+tip_data <- data.frame(
+  label = tree$tip.label) %>%
+  tidyr::separate(label, into = c("id", "sampledate", "is_outgroup", "trait"), sep = "_", remove = F) %>%
+  mutate(trait = as.numeric(trait))
+
+# Fit the POUMM
+print("Inferring the POUMM parameters")
+POUMM_spec <- POUMM::specifyPOUMM_ATH2tMeanSeG0(
+  # par vector is c(alpha, theta, H2tMean, sigmae, g0)
+  z = tip_data$trait,
+  tree = tree,
+  nSamplesMCMC = N_SAMPLES_MCMC,
+  parPriorMCMC = function(par) {
+      dexp(par[1], rate = 0.02, log = T) +
+      dnorm(par[2], mean = 4.5, sd = sqrt(3), log = T) + 
+      dunif(par[3],  min = 0, max = 1, log = T) + 
+      bayesmeta::drayleigh(par[4], scale = 1/sqrt(2*0.02), log = T) + 
[…]

### 2021-10-28 scripts/R/correct_spvl.R
add script to apply phylo correction to trait based on POUMM fit

@@ -1,0 +1,46 @@
+# This script is to apply phylogenetic correction to spVL using POUMM parameters.
+# Run as: Run as: /home/nadeaus/programs/R-3.6.2/bin/Rscript /links/groups/stadler/SHCSData/analysis/200207_redo_whole_pipeline/correct_spvl.R
+
+OUTDIR <- "output"
+DATADIR <- "/Volumes/stadler/SHCSData/"
+
+require(ape)
+require(POUMM)
+
+source("scripts/R/functions/POUMM_utility_functions.R")
+
+# Load the tree
+print("Loading POUMM fit and tree.")
+tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
+
+# Parse trait data from tip labels (this ensures tips and trait values are in same order)
+tip_data <- data.frame(
+  label = tree$tip.label) %>%
+  tidyr::separate(label, into = c("id", "sampledate", "is_outgroup", "trait"), sep = "_", remove = F) %>%
+  mutate(trait = as.numeric(trait))
+
+# Load the estimated POUMM parameters
+POUMM_summary <- read.csv(file = paste(OUTDIR, "poumm_parameter_estimates.csv", sep = "/"), row.names = 1)
+
+# Make data structure for inferred POUMM params
+POUMM_params <- data.frame(
+  alpha  = POUMM_summary['alpha', 'PostMean'],
+  theta  = POUMM_summary['theta', 'PostMean'],
+  sigma  = POUMM_summary['sigma', 'PostMean'],
+  sigmae = POUMM_summary['sigmae', 'PostMean'],
+  g0     = POUMM_summary['g0', 'PostMean'])  # using posterior means
+
+# Apply phylo-correction
+paste("Inferring maximum likelihood viral and non-viral parts of trait.")
+inference_results <- do.call(
+  what = inferVHfromPOUMMfit,
+  args = c(POUMM_params, list(tree = tree, z = tip_data$trait)))
+
+inference_results <- cbind(tip_data, inference_results)
+
+# Write out phylo-corrected trait values
+write.csv(
+  x = inference_results,
+  file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"),
+  row.names = F)
+

### 2021-10-28 scripts/R/correct_trait.R
add script to make phenotype file for GWAS

@@ -1,11 +1,14 @@
 # This script is to apply phylogenetic correction to spVL using POUMM parameters.
-# Run as: Run as: /home/nadeaus/programs/R-3.6.2/bin/Rscript /links/groups/stadler/SHCSData/analysis/200207_redo_whole_pipeline/correct_spvl.R
+# Run as: Run as: Rscript scripts/R/correct_trait.R
 
 OUTDIR <- "output"
 DATADIR <- "/Volumes/stadler/SHCSData/"
 
 require(ape)
 require(POUMM)
+require(dplyr)
+require(ggtree)
+require(ggplot2)
 
 source("scripts/R/functions/POUMM_utility_functions.R")
 
@@ -44,3 +47,15 @@ write.csv(
   file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"),
   row.names = F)
 
+# Plot tree with phylo-corrected trait
+p <- ggtree(tr = tree) %<+% inference_results +
+  geom_tippoint(aes(color = as.numeric(v.MWA))) +
+  scale_color_gradient(low = "green", high = "red", name = "Phylo-estimated\npathogen trait value")
+
+ggsave(plot = p, paste(OUTDIR, "pathogen_tree_with_phylo-estimated_pathogen_trait.png", sep = "/"), height = 10, width = 7, units = "in")
+
+p2 <- ggtree(tr = tree) %<+% inference_results +
+  geom_tippoint(aes(color = as.numeric(h.MWA))) +
+  scale_color_gradient(low = "green", high = "red", name = "Phylo-estimated\nenvironmental trait value")
+
+ggsave(plot = p2, paste(OUTDIR, "pathogen_tree_with_phylo-estimated_env_trait.png", sep = "/"), height = 10, width = 7, units = "in")

### 2021-10-28 scripts/R/make_gwas_phenotypes_file.R

@@ -1,0 +1,32 @@
+# This script is to make a .fam (phenotype & family relationships) file with corrected and total trait values for PLINK.
+# Run as: Run as: Rscript scripts/R/make_gwas_phenotypes_file.R
+
+require(tidyr)
+require(dplyr)
+
+OUTDIR <- "output"
+DATADIR <- "/Volumes/stadler/SHCSData/data"
+FAM_FN <- "spVL.shcs.rs_with_sex_SN.fam"
+
+# Load files
+print("Loading fam and corrected trait data.")
+fam <- read.delim(file =  paste(DATADIR, FAM_FN, sep = "/"), sep = "", header = F)
+colnames(fam) <- c("FID", "IID", "IID_father", "IID_mother", "sex", "phenotype")  # https://www.cog-genomics.org/plink/1.9/formats#fam
+
+corrected_trait <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
+
+print("Merging corrected trait into fam file.")
+phenotypes <- merge(
+  fam, corrected_trait %>% select(id, h.MWA, trait),
+  by.x = "IID", by.y = "id",
+  all.x = T, all.y = F)  # may have used more individuals for POUMM fitting than have genotype data
+phenotypes <- phenotypes %>%
+  replace_na(replace = list("h.MWA" = -9, "trait" = -9)) %>%
+  select(FID, IID, h.MWA, trait)
+
+write.table(
+  x = phenotypes,
+  file = paste(OUTDIR, "gwas_phenotypes.txt", sep = "/"),
+  row.names = F,
+  col.names = T,
+  quote = F)

### 2021-10-28 scripts/R/filter_gwas_individuals.R
add scripts for container to run GWAS filtering, keep file needs 2 cols for FID, IID to work

@@ -74,7 +74,7 @@ gwas_individuals_to_keep <- gwas_metadata %>% filter(is_european, Pop == "Test")
 print(paste("Keeping", length(unique(gwas_individuals_to_keep$id)), "individuals for GWAS."))
 
 write.table(
-  x = gwas_individuals_to_keep %>% select(id),
+  x = gwas_individuals_to_keep %>% select(id) %>% mutate(fid = id),  # two-column format for IID, FID otherwise PLINK throws out all samples
   file = paste(OUTDIR, "gwas_individuals_to_keep.txt", sep = "/"),
   row.names = F,
   col.names = F,

### 2021-10-28 scripts/bash/prep-gwas-files.sh

@@ -1,0 +1,28 @@
+#!/bin/bash
+# This script filters genotype data for GWAS.
+# Run with: bash /scripts/bash/prep-gwas-files.sh
+
+DATA_PREFIX="spVL.shcs.rs"
+FAM_W_SEX="spVL.shcs.rs_with_sex_SN.fam"
+
+# Filter data to clean dataset
+./plink2 \
+--bfile data/$DATA_PREFIX \
+--fam data/$FAM_W_SEX \
+--keep output/gwas_individuals_to_keep.txt \
+--king-cutoff 0.09375 \
+--maf 0.01 \
+--geno 0.01 \
+--hwe 0.00005 \
+--make-bed \
+--memory 10000 \
+--out output/${DATA_PREFIX}.filtered
+
+# Summarize the QC-passed genome dataset
+./plink2 \
+--bfile output/${DATA_PREFIX}.filtered \
+--freq \
+--missing \
+--hardy \
+--memory 10000 \
+--out output/${DATA_PREFIX}.filtered

### 2021-10-28 scripts/R/plot_simulation_results.rmd
rename file
move all figures to top-level 'figures' folder

--- 2021-09-28 scripts/R/plot_results.rmd
@@
 ---
-title: "Plot results"
+title: "Plot simulation results"
 author: nadeaus
@@
 ggsave(plot = rmse_plot,
-       filename = "manuscript/figures/simulation_RMSE.png",
+       filename = "figures/simulation_RMSE.png",
        width = 4.25,
@@
 ggsave(plot = rmse_plot_with_hiv_values,
-       filename = "manuscript/figures/simulation_RMSE_with_HIV_value_highlighted.png",
+       filename = "figures/simulation_RMSE_with_HIV_value_highlighted.png",
        width = 4.25,
@@
 ggsave(plot = tpr_plot,
-       filename = "manuscript/figures/simulation_TPR.png",
+       filename = "figures/simulation_TPR.png",
        width = 5.25,

### 2021-10-28 scripts/R/correct_trait.R

@@ -1,6 +1,7 @@
 # This script is to apply phylogenetic correction to spVL using POUMM parameters.
 # Run as: Run as: Rscript scripts/R/correct_trait.R
 
+FIGDIR <- "figures"
 OUTDIR <- "output"
 DATADIR <- "/Volumes/stadler/SHCSData/"
 
@@ -52,10 +53,10 @@ p <- ggtree(tr = tree) %<+% inference_results +
   geom_tippoint(aes(color = as.numeric(v.MWA))) +
   scale_color_gradient(low = "green", high = "red", name = "Phylo-estimated\npathogen trait value")
 
-ggsave(plot = p, paste(OUTDIR, "pathogen_tree_with_phylo-estimated_pathogen_trait.png", sep = "/"), height = 10, width = 7, units = "in")
+ggsave(plot = p, paste(FIGDIR, "pathogen_tree_with_phylo-estimated_pathogen_trait.png", sep = "/"), height = 10, width = 7, units = "in")
 
 p2 <- ggtree(tr = tree) %<+% inference_results +
   geom_tippoint(aes(color = as.numeric(h.MWA))) +
   scale_color_gradient(low = "green", high = "red", name = "Phylo-estimated\nenvironmental trait value")
 
-ggsave(plot = p2, paste(OUTDIR, "pathogen_tree_with_phylo-estimated_env_trait.png", sep = "/"), height = 10, width = 7, units = "in")
+ggsave(plot = p2, paste(FIGDIR, "pathogen_tree_with_phylo-estimated_env_trait.png", sep = "/"), height = 10, width = 7, units = "in")

### 2021-10-28 scripts/R/filter_gwas_individuals.R

@@ -5,6 +5,7 @@ library(dplyr)
 library(gridExtra)
 library(ggplot2)
 
+FIGDIR <- "figures"
 OUTDIR <- "output"
 DATADIR <- "/Volumes/stadler/SHCSData/data"
 
@@ -57,7 +58,7 @@ p1 <- ggplot(
   guides(color = "none") +
   facet_grid(PC_number ~ .)
 
-ggsave(plot = p1, filename = paste(OUTDIR, "host_genotype_pca.png", sep = "/"))
+ggsave(plot = p1, filename = paste(FIGDIR, "host_genotype_pca.png", sep = "/"))
 
 # Get list of samples filtered based on European ancestry
 gwas_individual_filtering_summary <- gwas_metadata %>%

### 2021-10-28 scripts/R/fit_poumm.R

@@ -1,6 +1,7 @@
 # This script is to fit the POUMM.
 # Run as: Rscript scripts/R/fit_poumm.R
 
+FIGDIR <- "figures"
 OUTDIR <- "output"
 
 N_SAMPLES_MCMC <- 4E6
@@ -79,7 +80,7 @@ p <- plots$densplot +
   labs(x = "Value", y = "Density") +
   scale_y_continuous(expand = c(0, 0)) +
   theme(legend.position = "bottom")
-ggsave(plot = p, filename = paste(OUTDIR, "poumm_parameter_estimates.png", sep = "/"), width = 6, height = 4, units = "in")
+ggsave(plot = p, filename = paste(FIGDIR, "poumm_parameter_estimates.png", sep = "/"), width = 6, height = 4, units = "in")
 
 estimates <- summary(POUMM_fit)[stat %in% c("alpha", "theta", "sigma", "sigmae", "g0", "H2tMean")]
 estimates_pretty <- estimates %>%

### 2021-10-28 scripts/R/root_tree.R

@@ -5,6 +5,7 @@ library(ggplot2)
 library(ggtree)
 library(tidyr)
 
+FIGDIR <- "figures"
 OUTDIR <- "output"
 DATADIR <- "/Volumes/stadler/SHCSData/data"
 TREE_FN <- "pathogen.treefile"
@@ -21,14 +22,14 @@ tip_data <- data.frame(
 p <- ggtree(tr = tree) %<+% tip_data +
   geom_tippoint(aes(color = is_outgroup))
 
-ggsave(plot = p, paste(OUTDIR, "pathogen_tree_with_outgroup.png", sep = "/"), height = 10, width = 7, units = "in")
+ggsave(plot = p, paste(FIGDIR, "pathogen_tree_with_outgroup.png", sep = "/"), height = 10, width = 7, units = "in")
 
 # Plot tree with trait
 p2 <- ggtree(tr = tree) %<+% tip_data +
   geom_tippoint(aes(color = as.numeric(trait))) +
   scale_color_gradient(low = "green", high = "red", name = "Trait value")
 
-ggsave(plot = p2, paste(OUTDIR, "pathogen_tree_with_trait.png", sep = "/"), height = 10, width = 7, units = "in")
+ggsave(plot = p2, paste(FIGDIR, "pathogen_tree_with_trait.png", sep = "/"), height = 10, width = 7, units = "in")
 
 # Root the tree
 paste("Rooting tree according to outgroup and writing to file")

### 2021-10-28 manuscript/hiv-gwas.tex
Update on Overleaf.

--- 2021-10-27 hiv-gwas.tex
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{figures/200207_gwas_results}
+		\includegraphics[width=\linewidth]{figures_archived/200207_gwas_results}
 		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load.
@@
 	\centering
-	\begin{subfigure}[b]{0.75\textwidth}
-		\includegraphics[width=\linewidth]{figures/200207_POUMM_posteriors_zoomed.png}
+		\includegraphics[width=\linewidth]{figures/poumm_parameter_estimates.png}
 		\caption{Marginal posterior distributions.}
-	\end{subfigure}
-~
-	\begin{subfigure}[b]{0.75\textwidth}
-		\includegraphics[width=\linewidth]{figures/200207_POUMM_traces.png}
-		\caption{MCMC trace plots.}
-	\end{subfigure}
-	\caption{POUMM parameter inference.}\label{fig:poumm-fit}
+\label{fig:poumm-parameters}
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{figures/200207_spvl_on_tree}
-	\caption{Inferred viral phylogeny with tips colored by estimated viral part of spVL, non-viral part of spVL, and measured total spVL.}
+\includegraphics[width = 0.3\linewidth]{figures/pathogen_tree_with_trait.png}
+\includegraphics[width = 0.3\linewidth]{figures/pathogen_tree_with_phylo-estimated_env_trait.png}
+\includegraphics[width = 0.3\linewidth]{figures/pathogen_tree_with_phylo-estimated_pathogen_trait.png}
+	\caption{Inferred viral phylogeny with tips colored by measured total spVL, estimated non-viral part of spVL, and estimated viral part of spVL.}
 	\label{fig:spvl-on-tree}
@@
 			\hline 
-			Raw data & 76,979,521 \\
-			Missing call rate $>$ 0.01 & -69,526,564 \\
-			Hardy-Weinburg p-value $<$ 5x10$^-5$ & -535 \\
-			Minor allele frequency $<$ 0.01 & -2,897,654 \\
-			Variants for GWAS & 4,554,768 \\ \hline			
+			Raw data & 76979521 \\
+			Missing genotype rate $>$ 0.01 & 7452957 \\
+			Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 7452425 \\
+			Minor allele frequency $<$ 0.01 & 4556770 \\ \hline			
 		\end{tabular}
@@
 \begin{figure}[H]
-	\includegraphics[width=\linewidth]{figures/pca_plot}
-	\caption{Thresholds for exclusion of individuals of non-Western European ancestry.
-The lower plots are filtered to show only European HapMap samples and the rectangles show thresholds for sample exclusion.}
+	\includegraphics[width=\linewidth]{figures/host_genotype_pca.png}
+	\caption{SHCS individuals and HapMap3 individuals plotted along the top three principle components of genetic variation.
+Points with black borders are within the thresholds used to select individuals of likely European ancestry.}
 	\label{fig:PCA}
@@
 	\centering
-	\includegraphics[width=0.75\linewidth]{figures/200207_qq_plots.png}
+	\includegraphics[width=0.75\linewidth]{figures_archived/200207_qq_plots.png}
 	\caption{Quartile-quartile plots from association tests.}

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
standardize manuscript figures, start with model example figure

---
title: "Make manuscript figures"
output: html_notebook
---

This script is to make cleaner, consistently formatted figures for the manuscript.

Load libraries
```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(POUMM)
library(ape)
library(cowplot)

FIGDIR <- "figures"
```

Define formatting
```{r}
diverging_palette <- "Set2"
gradient_palette <- "PiYG"
shared_theme <- theme_bw()
single_col_width <- 8.7  # cm
double_col_width <- 11.4  # cm

gv_name <- "Pathogen part of trait"
gh_name <- "Host part of trait"
e_name <- "Non-pathogen part of trait"
```

Load data
```{r}

```

Model illustration
```{r}
# Simulate many OU trajectories along tree
generate_trait_trajectory <- function(v0, alpha, theta, sigma, N_steps) {
  POUMM::rTrajectoryOU(
    z0 = v0,
    t = 1,
    alpha = alpha,
    theta = theta,
    sigma = sigma,
    steps = N_steps)
}

# Generate tree
set.seed(10)
tree <- ape::rtree(n = 3)
node_times <- POUMM::nodeTimes(tree)
edge_colors = RColorBrewer::brewer.pal(n = 4, name = diverging_palette)

# Define OU process
scale_factor <- 1000
v0 <- 3.98
theta <- 4.38
alpha <- 30.15 / scale_factor
sigma <- 3.56 / sqrt(scale_factor)

# Initialize data structures
traceplot <- as.data.frame(matrix(ncol = 4, nrow = 0))
colnames(traceplot) <- c("lineage", "time", "v", "color")
v_nodes <- rep(v0, length(node_times))

# Generate OU trajectory for each edge on tree
for (j in 1:nrow(tree$edge)) {
  from = tree$edge[j, 1]
  to = tree$edge[j, 2]
  N_steps = round(tree$edge.length[j]*scale_factor) - 1
  print(paste(N_steps, "steps on branch", j))

  trajectory <- POUMM::rTrajectoryOU(
    z0 = v_nodes[from],
    t = 0.01,
    alpha = alpha,
    theta = theta,
    sigma = sigma,
    steps = N_steps)

  v_nodes[to] <- trajectory[N_steps]
  v <- c(v_nodes[from], trajectory)

  new_data <- data.frame(
    lineage = rep(j, N_steps + 1),
    time = seq(node_times[from], node_times[to], length.out = N_steps + 1),
    v = v,
    color = rep(edge_colors[j], N_steps + 1))
  traceplot <- rbind(traceplot, new_data)
}

mu_OU <- function(t) v0 * exp(-alpha * t) + theta * (1 - exp(-alpha * t))

OU_plot <- ggplot(
  data = traceplot,
  aes(x = time, y = v, group = lineage, color = I(color))) +
  geom_line(size = 1.25) +
  labs(
    x = expression('Substitutions site'^-1 * ' year'^-1),
    y = expression(italic(g)[v])) +
  shared_theme +
  stat_function(fun = mu_OU, color = "black", linetype = "dashed")

#TODO: I have no idea how the edge coloring works, this is a really shitty hardcoded soln
tree_data = data.frame(node = c(1, 2, 3, 4, 5), color = c(edge_colors[2], edge_colors[3], edge_colors[4], "blue", edge_colors[1]))

tree_plot <- ggtree::ggtree(
  tree,
  ladderize = F,
  size = 1.5) %<+% tree_data + aes(color = I(color)) +
  labs(x = expression('Substitutions site'^-1 * ' year'^-1))

e_data <- data.frame(x = c(-1, 1))

e_plot <-  ggplot(data = e_data, aes(x)) +
  stat_function(fun = dnorm,
                n = 101,
                args = list(mean = 0, sd = 0.2),
                size = 1) +
  ylab("Density") +
  shared_theme +
  labs(x = expression(italic(e)))

# Align OU and tree plots
left_side <- plot_grid(OU_plot, tree_plot, align = "v", axis = "lr", nrow = 2, rel_heights = c(2, 1), labels = c("A", "C"))
right_side <- plot_grid(NULL, e_plot, NULL, nrow = 3, rel_heights = c(1, 2, 1), labels = c("", "B", ""))

png(
  filename = paste(FIGDIR, "model_figure.png", sep = "/"),
  width = double_col_width, height = single_col_width, units = "cm", res = 300)
  plot_grid(left_side, NULL, right_side, ncol = 3, rel_widths = c(1.5, 0.2, 1))
dev.off()

```

Simulation results
```{r}

```

Inferred phylogeny
```{r}

```

POUMM parameter estimates
```{r}

```

Estimated trait values
```{r}

```

PCA of cohort genotypes
```{r}

```

Manhattan plot of GWAS results
```{r}

```

Q-Q plot of GWAS results
```{r}

```

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
fiddle with OU parameters for model figure

@@
 tree <- ape::rtree(n = 3)
+tree$edge.length <- tree$edge.length * 6
 node_times <- POUMM::nodeTimes(tree)
@@
 # Define OU process
-scale_factor <- 1000
-v0 <- 3.98
-theta <- 4.38
-alpha <- 30.15 / scale_factor
-sigma <- 3.56 / sqrt(scale_factor)
+scale_factor <- 40  # 1000
+v0 <- 3.5
+theta <- 4.5
+alpha <- 50 / scale_factor
+sigma <- 3 / sqrt(scale_factor)
 

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
update simulation results figure: cowplot, standard labels

@@
 FIGDIR <- "figures"
+
+source("scripts/R/functions/plot_functions.R")
+source("scripts/R/functions/utility_functions.R")
 ```
@@
 e_name <- "Non-pathogen part of trait"
+ehat_name <- "Estimated non-pathogen part of trait"
+z_minus_zbar_name <- "Scaled trait value"
 ```
@@
 ```{r}
-
+accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
+tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
 ```
@@
 ```{r}
+# Get POUMM values for HIV spVL to annotate plot with
+spvl_data <- data.frame(
+        inference.method = "h.MWA.p.value",
+        H2 = 3.9,
+        alpha = 8.046)  # values are hand-calculated because the breaks scale is defined as seq(0.5, n.breaks - 0.5, 1) in plotting function
 
+spvl_data$inference.method <- factor(
+        spvl_data$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+
+# Summarize results for plotting
+error.cols <- c("RMSE.z.h", "RMSE.h.MWA.h")
+
+data.long <- do.call(
+        what = tidyr::gather,
+        args = c(error.cols, list(data = accuracy_simulation_results, key = "inference.method", value = "error")))
+
+data.summary <- as.data.frame(
+        data.long %>%
+                dplyr::group_by(inference.method, alpha, H2, var.z, p) %>%
+                dplyr::summarise(mean.error = mean(error), sd.error = sd(error), N.reps = dplyr::n()))
+
+# Create facet variable labels for plot
+data.summary$inference.method <- factor(
+        data.summary$inference.method,
+        levels = c("RMSE.h.MWA.h", "RMSE.z.h"),
+        labels = c(ehat_name, z_minus_zbar_name))
+
+# Plot RMSE
+rmse_plot <- plotHeatmap(
+        df = data.summary,
+        N.sim = data.summary$N.reps[[1]],
+        is.MLE = T,
+        scale.limits = c(0, 1),
+        legend.name = "RMSE",
+        midpoint = 0.5,
+        fill.hi = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[1],
+        fill.lo = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[3],
+        fill.mid = RColorBrewer::brewer.pal(name = 'PiYG', n = 3)[2],
+        x.var = "alpha",
+        y.var = "H2",
+        fill.var = "mean.error",
+        facet.x = "inference.method",
+        facet.y = NA,
+        is.oob.squish = F,
+        is.real.scale = F,
+        labeller = label_wrap_gen(width = 25)) +
+        labs(x = expression("Selection strength (" * alpha * ")"),
+             y = expression("spVL heritability (" * italic('H'^2) * ")")) +
+        theme(
+                text = element_text(size = 8),
+                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
+                strip.background = element_rect(colour = "black", size = 0.5)) +
+        geom_point(data = spvl_data, fill = "black")
+
+data <- tidyr::gather(
+        data = tpr_simulation_results,
+        key = "inference.method",
+        value = "p.value",
+        h.plus.e.p.value, norm.z.p.value, h.OU.p.value, h.MWA.p.value)
+
+# Calculate TPR for each simulation
+sig.level <- correctSignificanceLevelForMultTesting(
+        significance.level = 0.05,
+        significance.correction = "Bonferroni",
+        N.tests = 20)
+data$TP <- ifelse(data$p.value < sig.level & data$is.causal.variant, T, F)
+data$FP <- ifelse(data$p.value < sig.level & !(data$is.causal.variant), T, F)
+data.TPR <- as.data.frame(
+        data %>%
+                dplyr::group_by(runID, inference.method, alpha, sigma, H2, var.z, p, delta) %>%
+                dplyr::summarise(
+                        TPR = sum(TP, na.rm = T)/sum(is.causal.variant),
+                        FP = sum(FP), K = dplyr::n(),
+                        N.noncausal.variants = sum(!(is.causal.variant))))
+
+# Summarize replicate simulations
+data.summary <- as.data.frame(
+        data.TPR %>%
+                dplyr::group_by(inference.method, alpha, sigma, H2, var.z, p, delta) %>%
+                dplyr::summarise(
+                        mean.TPR = mean(TPR),
+                        sd.TPR = sd(TPR),
+                        mean.FP = mean(FP),
+                        sd.FP = sd(FP),
+                        N.noncausal.variants = mean(N.noncausal.variants),
+                        N.sim = dplyr::n()))
+
+# Create facet variable labels for plot
+data.summary$inference.method <- factor(
+        data.summary$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+
+# Plot TPR
+tpr_plot <- makeTPRPlot(
+        df = data.summary %>% filter(inference.method %in% c(e_name, ehat_name, z_minus_zbar_name)),
+        is.MLE = T,
+        midpoint = 0.5,
+        fill.lo = RColorBrewer::brewer.pal(name = 'RdBu', n = 3)[3],
+        fill.hi = RColorBrewer::brewer.pal(name = 'RdBu', n = 3)[1],
+        fill.mid = RColorBrewer::brewer.pal(name = 'RdBu', n = 3)[2]) +
+        facet_grid(. ~ inference.method, labeller = label_wrap_gen(width = 25)) +
+        labs(x = expression("Selection strength (" * alpha * ")"),
+             y = expression("spVL heritability (" * 'H'^2 * ")")) +
+        theme(
+                text = element_text(size = 8),
+                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
+                strip.background = element_rect(colour = "black", size = 0.5)) +
+        geom_point(data = spvl_data, fill = "black")
+
+top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1, 3.25), labels = c("", "A"))
+png(filename = "figures/simulation_results.png", width = double_col_width, height = single_col_width, unit = "cm", res = 300)
+plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))
+dev.off()
 ```

### 2021-10-28 manuscript/hiv-gwas.tex

@@
 	\begin{center}
-		\includegraphics[width=0.7\linewidth]{figures/model_figure}
+		\includegraphics[width=0.6\linewidth]{figures/model_figure}
 		\caption{A statistical model of the spVL trait.
-The virus-determined part of spVL, $g_v$, evolves according to an OU process along the viral phylogeny.
-The dashed line shows the evolutionarily optimal trait value $\theta$.
-The environmentally-determined part of spVL, $e$, including direct host genetic effects and other environmental effects, is modeled as a Gaussian random variable whose value is added at the tips of the phylogeny.}
+(A) The virus-determined part of spVL, $g_v$, evolves according to an OU process along (C) the viral phylogeny.
+The dashed line in (A) shows the expected value of the trait at each point in time.
+(B) The environmentally-determined part of spVL, $e$, including host genetic effects, interaction effects, and other environmental effects, is modeled as a Gaussian random variable whose value is independently drawn for each host at the tips of the phylogeny.}
 		\label{fig:spVLModel}

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
mess with plot sizes

@@
 shared_theme <- theme_bw()
-single_col_width <- 8.7  # cm
-double_col_width <- 11.4  # cm
+linewidth <- 16.5  # cm
+height <- 11  # cm
 
@@
   filename = paste(FIGDIR, "model_figure.png", sep = "/"),
-  width = double_col_width, height = single_col_width, units = "cm", res = 300)
+  width = linewidth, height = height, units = "cm", res = 300)
   plot_grid(left_side, NULL, right_side, ncol = 3, rel_widths = c(1.5, 0.2, 1))
@@
              y = expression("spVL heritability (" * italic('H'^2) * ")")) +
-        theme(
-                text = element_text(size = 8),
-                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
-                strip.background = element_rect(colour = "black", size = 0.5)) +
+        shared_theme +
+        theme(axis.text.x = element_text(angle = 90)) +
         geom_point(data = spvl_data, fill = "black")
@@
              y = expression("spVL heritability (" * 'H'^2 * ")")) +
-        theme(
-                text = element_text(size = 8),
-                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
-                strip.background = element_rect(colour = "black", size = 0.5)) +
+        shared_theme +
+        theme(axis.text.x = element_text(angle = 90)) +
         geom_point(data = spvl_data, fill = "black")
@@
 top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1, 3.25), labels = c("", "A"))
-png(filename = "figures/simulation_results.png", width = double_col_width, height = single_col_width, unit = "cm", res = 300)
+png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)
 plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))

### 2021-10-28 scripts/R/functions/plot_functions.R

@@ -305,11 +305,6 @@ makeTPRPlot <- function(df, is.MLE, scale.limits = c(0, 1),
     fill.var = "mean.TPR",
     facet.y = facet.y,
     is.real.scale = F)
-  if (is.MLE) {
-    p <- p + ggtitle("MLE")
-  } else {
-    p <- p + ggtitle("no MLE")
-  }
   return(p)
 }
 

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
fix x-axis labels in GWAS simulation plot

@@
         shared_theme +
-        theme(axis.text.x = element_text(angle = 90)) +
+        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
         geom_point(data = spvl_data, fill = "black")

### 2021-10-28 manuscript/hiv-gwas.tex
Update on Overleaf.

--- 2021-10-28 manuscript/hiv-gwas.tex
@@
 	\begin{center}
-		\includegraphics[width=0.7\linewidth]{figures/model_figure}
-		\caption{A statistical model of the spVL trait.
-The virus-determined part of spVL, $g_v$, evolves according to an OU process along the viral phylogeny.
-The dashed line shows the evolutionarily optimal trait value $\theta$.
-The environmentally-determined part of spVL, $e$, including direct host genetic effects and other environmental effects, is modeled as a Gaussian random variable whose value is added at the tips of the phylogeny.}
+		\includegraphics[width=0.75\linewidth]{figures/model_figure}
+		\caption{The POUMM statistical model for the spVL trait.
+(A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny.
+The dashed line in (A) shows the expected value of the trait at each point in time under the model.
+(B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable.
+Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
 		\label{fig:spVLModel}
@@
 		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
-		\caption{(A) Root mean square error (RMSE) with respect to true (simulated) values for the host part of spVL ($\bm{{g_h}}$).
-The left plot is for the POUMM-based estimate of the non-viral part of spVL ($\bm{\hat{e}}$) and the right plot is for spVL scaled such that it is centered at zero ($\bm{z} - \bm{\bar{z}}$).
-Each plot tile represents the average RMSE over 20 simulated datasets of 500 samples.
-(B) True positive rate (TPR) in GWAS performed using true (simulated) non-viral spVL values ($\bm{e}$), the POUMM-based estimate for the non-viral part of spVL ($\bm{\hat{e}}$), and total spVL scaled such that it is centered at zero ($\bm{z} - \bm{\bar{z}}$).
-Each plot tile represents the average TPR over 20 simulated datasets of 500 samples.}
+		\caption{Results from the simulation study.
+We simulated pathogen and non-pathogen (host \& environmental) effects on the spVL trait under the POUMM with different trait heritability values (y-axis) and selection strengths (x-axis).
+For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS using these estimates.
+(A) shows how well we can estimate the host effect using our method (left side) compared to just the scaled trait value (right side).
+(B) shows how GWAS can theoretically be improved by using the true (normally unknown) non-pathogen effects on spVL (left).
+It can still be improved in certain scenarios using our estimate for this value (middle) compared to the current standard - using just the (scaled) trait value (right).
+Tile color represents the average value over 20 simulated datasets of 500 samples for each heritability and selection strength value.
+The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data.
+RMSE = Root mean square error, TPR = True positive rate.}
+		% (RMSE) with respect to simulated values for the host part of spVL ($\bm{{g_h}}$) as a function of spVL trait heritability and selection strength on the trait.
+The left plot is for the POUMM-based estimate of the non-pathogen part of the trait ($\bm{\hat{e}}$) and the right plot is for the raw trait value scaled by the mean trait value, such that it is centered at zero ($\bm{z} - \bm{\bar{z}}$).
+(B) True positive rate (TPR) in GWAS performed using true (simulated) values for the non-pathogen part of the trait ($\bm{e}$), the POUMM-based estimate for the non-pathogen part of the trait ($\bm{\hat{e}}$), and the raw trait value scaled by the mean trait value ($\bm{z} - \bm{\bar{z}}$).
+The color of each plot tile in (A) and (B) represents the average value over 20 simulated datasets of 500 samples.
+The black point shows our estimates for the heritability of and selection strength on the spVL trait based on Swiss data.}
 		\label{fig:simulationResults}

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
add spvl on tree figure

@@
 library(cowplot)
+library(treeio)
 
 FIGDIR <- "figures"
+OUTDIR <- "output"
 
@@
 e_name <- "Non-pathogen part of trait"
+gvhat_name <- "Estimated pathogen part of trait"
 ehat_name <- "Estimated non-pathogen part of trait"
@@
 tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
+tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
+traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
+tree <- treeio::full_join(tree, traits, by = "label")
 ```
@@
 ```{r}
+legend_width <- 15
 
+# Plot tree with tip colors by spVL components
+legend_x_pos <- 0.99
+legend_y_pos <- 0.45
+
+v_tree <- ggtree(tree, ladderize = F) +
+        geom_tippoint(aes(color = v.MWA)) +
+        scale_color_viridis_c(
+                option = "plasma",
+                direction = -1,
+                limits = c(min(traits$v.MWA), max(traits$v.MWA))) +
+        theme(legend.position = "bottom") +
+        guides(color = guide_colorbar(title.position="top", title.hjust = 0.5)) +
+        labs(color = yulab.utils::str_wrap(gvhat_name, width = legend_width))
+
+h_tree <- ggtree(tree, ladderize = F) +
+        geom_tippoint(aes(color = h.MWA)) +
+        scale_color_viridis_c() +
+        theme(legend.position = "bottom") +
+        guides(color = guide_colorbar(title.position="top", title.hjust = 0.5)) +
+        labs(color = yulab.utils::str_wrap(ehat_name, width = legend_width))
+
+spvl_tree_same_scale <- ggtree(tree, ladderize = F) +
+        geom_tippoint(aes(color = trait)) +
+        scale_color_viridis_c(
+                option = "plasma",
+                direction = -1,
+                limits = c(min(traits$v.MWA), max(traits$v.MWA))) +
+        theme(legend.position = "bottom") +
+        guides(color = guide_colorbar(title.position="top", title.hjust = 0.5)) +
+        labs(color = yulab.utils::str_wrap("Calculated spVL", width = legend_width)) +
+        scale_fill_identity(name = '', guide = 'legend', labels = c('oob'))
+
+png(file = paste(FIGDIR, "spvl_on_tree.png", sep = "/"),
+    height = height, width = linewidth,
+    units = "cm",
+    res = 300)
+plot_grid(v_tree + theme(legend.position = "none"),
+          h_tree + theme(legend.position = "none"),
+          spvl_tree_same_scale + theme(legend.position = "none"),
+          get_legend(v_tree), get_legend(h_tree), get_legend(spvl_tree_same_scale),
+          nrow = 2, align = "v", axis = "tb",
+          rel_heights = c(3, 1))
+dev.off()
 ```

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
add subtitle labels to spvl tree figure

@@
     res = 300)
-plot_grid(v_tree + theme(legend.position = "none"),
+plot_grid(spvl_tree_same_scale + theme(legend.position = "none"),
           h_tree + theme(legend.position = "none"),
-          spvl_tree_same_scale + theme(legend.position = "none"),
-          get_legend(v_tree), get_legend(h_tree), get_legend(spvl_tree_same_scale),
+          v_tree + theme(legend.position = "none"),
+          get_legend(spvl_tree_same_scale), get_legend(h_tree), get_legend(v_tree),
           nrow = 2, align = "v", axis = "tb",
-          rel_heights = c(3, 1))
+          rel_heights = c(3, 1),
+          labels = c("A", "B", "C", "", "", ""))
 dev.off()

### 2021-10-28 manuscript/hiv-gwas.tex

@@
 \begin{figure}
-\includegraphics[width = 0.3\linewidth]{figures/pathogen_tree_with_trait.png}
-\includegraphics[width = 0.3\linewidth]{figures/pathogen_tree_with_phylo-estimated_env_trait.png}
-\includegraphics[width = 0.3\linewidth]{figures/pathogen_tree_with_phylo-estimated_pathogen_trait.png}
-	\caption{Inferred viral phylogeny with tips colored by measured total spVL, estimated non-viral part of spVL, and estimated viral part of spVL.}
+\includegraphics[width = 0.3\linewidth]{figures/spvl_on_tree.png}
+	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by calculated spVL and estimated pathogen and non-pathogen parts of spVL.}
 	\label{fig:spvl-on-tree}

### 2021-10-28 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 \begin{figure}
-\includegraphics[width = 0.3\linewidth]{figures/spvl_on_tree.png}
-	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by calculated spVL and estimated pathogen and non-pathogen parts of spVL.}
+\begin{center}
+\includegraphics[width = 0.7\linewidth]{figures/spvl_on_tree.png}
+	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}
 	\label{fig:spvl-on-tree}
+	\end{center}
 \end{figure}

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
add standardized poumm param estimates figure

@@
 tree <- treeio::full_join(tree, traits, by = "label")
+POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
 ```
@@
 ```{r}
-
+# Generate figures, write out estimates
+plots <- plot(POUMM_fit, doZoomIn = TRUE, doPlot = F)
+p <- plots$densplot +
+        scale_fill_manual(
+                values = RColorBrewer::brewer.pal(n = 3, name = diverging_palette),
+                aesthetics = c("fill", "color"),
+                labels = c("Prior", "Chain 1", "Chain 2"),
+                name = element_blank()) +
+        theme_bw() +
+        labs(x = "Value", y = "Density") +
+        scale_y_continuous(expand = c(0, 0)) +
+        theme(legend.position = "bottom")
+ggsave(plot = p, filename = paste(FIGDIR, "poumm_parameter_estimates.png", sep = "/"),
+       width = linewidth, height = height, units = "cm")
 ```

### 2021-10-28 scripts/R/make_manuscript_figures.rmd
add standardized host pca fig

@@
 POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
+pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv"))
 ```
@@
 
-Estimated trait values
-```{r}
-
-```
-
 PCA of cohort genotypes
 ```{r}
+# Plot combined SHCS and HapMap samples along the first few PCs
+pop_colors <- RColorBrewer::brewer.pal(n = 11, name = "Set3")
+names(pop_colors) <- unique(pc_scores$Population)
+swap_col <- pop_colors[9]
+pop_colors[9] <- pop_colors[4]
+pop_colors[4] <- swap_col
 
+p1 <- ggplot(
+        data = pc_scores %>% pivot_longer(cols = c(PC2, PC3), names_to = "PC_number", values_to = "PC_value"),
+        aes(x = PC1, y = PC_value, fill = Population, color = is_european)) +
+        geom_point(pch=21, size=3) +
+        shared_theme +
+        scale_fill_manual(values = pop_colors) +
+        scale_color_manual(values = c("transparent", "black")) +
+        labs(y = element_blank()) +
+        guides(color = "none") +
+        facet_grid(PC_number ~ .)
+
+ggsave(plot = p1, filename = paste(FIGDIR, "host_genotype_pca.png", sep = "/"),
+       width = linewidth, height = height, units = "cm")
 ```

### 2021-10-28 scripts/R/filter_gwas_individuals.R

@@ -4,6 +4,7 @@
 library(dplyr)
 library(gridExtra)
 library(ggplot2)
+library(tidyr)
 
 FIGDIR <- "figures"
 OUTDIR <- "output"
@@ -73,6 +74,11 @@ write.csv(x = gwas_individual_filtering_summary,
 
 gwas_individuals_to_keep <- gwas_metadata %>% filter(is_european, Pop == "Test")
 print(paste("Keeping", length(unique(gwas_individuals_to_keep$id)), "individuals for GWAS."))
+
+write.csv(
+  x = pc_scores,
+  file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv"),
+  row.names = F)
 
 write.table(
   x = gwas_individuals_to_keep %>% select(id) %>% mutate(fid = id),  # two-column format for IID, FID otherwise PLINK throws out all samples

### 2021-10-28 manuscript/hiv-gwas.tex

@@
 		\includegraphics[width=\linewidth]{figures/poumm_parameter_estimates.png}
-		\caption{Marginal posterior distributions.}
+		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on Swiss data.
+We ran two different MCMC chains to ensure the estimates converged.
+The black point on the x-axis shows the posterior mean value, which was used to estimate the pathogen- and non-pathogen effects on spVL.}
 \label{fig:poumm-parameters}
@@
 
-\begin{figure}
+\begin{figure}[H]
 \begin{center}
-\includegraphics[width = 0.7\linewidth]{figures/spvl_on_tree.png}
+\includegraphics[width = \linewidth]{figures/spvl_on_tree.png}
 	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}

### 2021-10-29 scripts/R/make_gwas_phenotypes_file.R
add scripts to generate container to do GWAS, don't have -9 phenotype for any samples

@@ -5,12 +5,11 @@ require(tidyr)
 require(dplyr)
 
 OUTDIR <- "output"
-DATADIR <- "/Volumes/stadler/SHCSData/data"
-FAM_FN <- "spVL.shcs.rs_with_sex_SN.fam"
+FAM_FN <- "spVL.shcs.rs.filtered.fam"
 
 # Load files
 print("Loading fam and corrected trait data.")
-fam <- read.delim(file =  paste(DATADIR, FAM_FN, sep = "/"), sep = "", header = F)
+fam <- read.delim(file =  paste(OUTDIR, FAM_FN, sep = "/"), sep = "", header = F)
 colnames(fam) <- c("FID", "IID", "IID_father", "IID_mother", "sex", "phenotype")  # https://www.cog-genomics.org/plink/1.9/formats#fam
 
 corrected_trait <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
@@ -20,9 +19,10 @@ phenotypes <- merge(
   fam, corrected_trait %>% select(id, h.MWA, trait),
   by.x = "IID", by.y = "id",
   all.x = T, all.y = F)  # may have used more individuals for POUMM fitting than have genotype data
-phenotypes <- phenotypes %>%
-  replace_na(replace = list("h.MWA" = -9, "trait" = -9)) %>%
-  select(FID, IID, h.MWA, trait)
+if(any(is.na(phenotypes$h.MWA) | is.na(phenotypes$trait))) {
+  stop("Some genomes after filtering don't have a phenotype!")
+}
+phenotypes <- phenotypes %>% select(FID, IID, h.MWA, trait)
 
 write.table(
   x = phenotypes,

### 2021-10-29 scripts/bash/run-gwas.sh

@@ -1,0 +1,68 @@
+# This script runs GWAS with both phenotypes, sex and top 5 PCs of genetic variation as co-variates.
+# Run as: bash scripts/bash/run-gwas.sh
+
+DATA_PREFIX="spVL.shcs.rs.filtered"
+MEM=10000
+
+# Get variants in LD to exclude from PCA calculation
+./plink2 \
+--bfile output/$DATA_PREFIX \
+--indep-pairwise 50 5 0.2 \
+--memory $MEM \
+--out output/$DATA_PREFIX
+
+# Get top 5 PCs for covariates based on SNPs in linkage equilibrium
+./plink2 \
+--bfile output/$DATA_PREFIX \
+--exclude output/${DATA_PREFIX}.prune.out \
+--pca \
+--out output/$DATA_PREFIX \
+--memory $MEM
+awk '{print $1,$2,$3,$4,$5,$6,$7}' output/$DATA_PREFIX.eigenvec >  output/${DATA_PREFIX}.pc5.txt
+
+# Run GWAS
+./plink2 \
+--bfile output/$DATA_PREFIX \
+--pheno output/gwas_phenotypes.txt \
+--glm sex \
+--covar output/spVL.shcs.rs.filtered.pc5.txt \
+--memory $MEM \
+--out output/$DATA_PREFIX
+
+# Extract SNP-only results, without covariates
+head -1 output/${DATA_PREFIX}.trait.glm.linear > output/${DATA_PREFIX}.trait.glm.linear.nocovariates
+grep -h 'ADD' output/${DATA_PREFIX}.trait.glm.linear >> output/${DATA_PREFIX}.trait.glm.linear.nocovariates
+
+head -1 output/${DATA_PREFIX}.h.MWA.glm.linear > output/${DATA_PREFIX}.h.MWA.glm.linear.nocovariates
+grep -h 'ADD' output/${DATA_PREFIX}.h.MWA.glm.linear >> output/${DATA_PREFIX}.h.MWA.glm.linear.nocovariates
+
+# Paste results from both trait values together
+paste "output/${DATA_PREFIX}.trait.glm.linear.nocovariates" "output/${DATA_PREFIX}.h.MWA.glm.linear.nocovariates" | awk '{print $1, $2, $3, $9, $12, $22, $25}' > output/gwas_results.txt
+HEADER="CHROM POS ID BETA_standard P_standard BETA_corrected P_corrected"
[…]

### 2021-10-29 scripts/R/filter_gwas_individuals.R
fix output file location

@@ -77,7 +77,7 @@ print(paste("Keeping", length(unique(gwas_individuals_to_keep$id)), "individuals
 
 write.csv(
   x = pc_scores,
-  file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv"),
+  file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"),
   row.names = F)
 
 write.table(

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
add trait value, qq-plot, manhattan plot figures

@@
 library(treeio)
+library(ggExtra)
 
@@
 z_minus_zbar_name <- "Scaled trait value"
+standard_gwas_name <- "GWAS with standard trait value"
+corrected_gwas_name <- "GWAS with estimated non-pathogen part of trait"
+genome_wide_significance <- 5E-8
 ```
@@
 POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
-pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv"))
+pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"))
+gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
 ```
@@
 
+Estimated trait values
+TODO: think about some way of depicting phylogenetic relationships here too?
+```{r}
+p <- ggplot(data = traits, aes(x = trait, y = h.MWA)) +
+      geom_point() +
+      shared_theme +
+      labs(x = "Calculated spVL trait", y = ehat_name)
+p2 <- ggExtra::ggMarginal(p, type="histogram")
+ggsave(plot = p2, filename = paste(FIGDIR, "trait_values.png", sep = "/"),
+      width = linewidth * 2/3, height = height * 2/3, units = "cm")
+```
+
 PCA of cohort genotypes
@@
 
+Q-Q plot of GWAS results
+```{r}
+qq_data_standard <- data.frame(
+        observed_log10p = -log10(sort(gwas_results$P_standard, decreasing = F)),
+        expected_log10p = -log10(ppoints(nrow(gwas_results))),
+        gwas_type = standard_gwas_name)
+qq_data_phylo <- data.frame(
+        observed_log10p = -log10(sort(gwas_results$P_corrected, decreasing = F)),
+        expected_log10p = -log10(ppoints(nrow(gwas_results))),
+        gwas_type = corrected_gwas_name)
+qq_results <- rbind(qq_data_standard, qq_data_phylo)
+
+p <- ggplot(data = qq_results, aes(x = expected_log10p, y = observed_log10p)) +
+        geom_point() +
+        geom_abline(slope = 1, intercept = 0, linetype = "dashed",
+                    color = RColorBrewer::brewer.pal(n = 3, name = diverging_palette)[[1]]) +
+        facet_wrap(. ~ gwas_type) +
+        labs(x = expression("expected log"[10]*"(p)"), y = expression("observed log"[10]*"(p)")) +
+        shared_theme
+ggsave(plot = p, filename = paste(FIGDIR, "qq_plots.png", sep = "/"),
+       width = linewidth, height = height * 2/3, units = "cm")
+```
+
 Manhattan plot of GWAS results
 ```{r}
+# Add genome position to data
+chr_lengths <- gwas_results %>%
+        group_by(CHROM) %>%
+        summarise(chr_len = max(POS))
+chr_starts <- chr_lengths %>%
+        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
+        select(-chr_len)
+gwas_results_annotated <- left_join(gwas_results, chr_starts, by = "CHROM")
+gwas_results_annotated <- gwas_results_annotated %>%
+        arrange(CHROM, POS) %>%
+        mutate(BP_position = CHR_start + POS)
 
+# This is a super hacky way to get labels in between x ticks
+# Hadley Wickam disapproves :/
+max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 22, "BP_position"])
+x_axis_ticks <- chr_starts
+x_axis_ticks <- rbind(
+        x_axis_ticks,
+        data.frame(
+                CHROM = c(23, 24),
+                CHR_start = c(max_BP_position, max_BP_position + 1000)))
+x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
+x_axis_ticks[x_axis_ticks$CHROM == 22, "midpoint"] <-
+        x_axis_ticks[x_axis_ticks$CHROM == 22, "CHR_start"] +
+                (max(gwas_results_annotated[gwas_results_annotated$CHROM == 22, "BP_position"]) -
+                        x_axis_ticks[x_axis_ticks$CHROM == 22, "CHR_start"]) / 2
+x_axis_ticks_tmp <- tidyr::pivot_longer(
+        data = x_axis_ticks,
+        cols = c("midpoint", "CHR_start"),
+        names_to = "marker",
+        values_to = "tick_position") %>%
+        mutate(CHROM = as.character(as.integer(CHROM)))
+x_axis_ticks_tmp[seq(from = 2, to = 46, by = 2), "CHROM"] <- ""
+x_axis_ticks_tmp[x_axis_ticks_tmp$CHROM == "23", "CHROM"] <- ""
+x_axis_ticks_tmp[x_axis_ticks_tmp$CHROM == "24", "CHROM"] <- ""
+
+# Format data for plotting
+results_long <- tidyr::pivot_longer(
+        data = gwas_results_annotated,
+        cols = c("P_standard", "P_corrected"),
+        names_to = "gwas_type",
+        names_prefix = "P_",
+        values_to = "P")
+
+# Fix facet labels for plotting
+results_long$gwas_type <- factor(
+        results_long$gwas_type,
+        levels = c("standard", "corrected"),
+        labels = c(standard_gwas_name, corrected_gwas_name))
+
+# Make comparative GWAS overview plot
+overview_comparison <- ggplot(
+        data = results_long,
+        aes(x = BP_position, y = -log10(P))) +
+        geom_point(aes(color = CHROM %% 2 == 0)) +
+        scale_color_manual(values = c("grey", "black")) +
+        scale_x_continuous(
+                breaks = x_axis_ticks$tick_position,
+                labels = x_axis_ticks$CHROM,
+                limits = c(0, NA),
+                expand = c(0, 0)) +
+        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
+        labs(x = "Chromosome", y = expression('-Log'[10]*'(p)')) +
+        geom_hline(
+                yintercept = -log10(genome_wide_significance),
+                linetype = "dotted") +
+        theme_bw() +
+        theme(
+                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
+                panel.grid.major.x = element_blank(),
+                axis.ticks.x = element_line(color = rep(c(NA, "black"), 23)),
+                panel.grid.minor.x = element_blank(),
+                legend.position = "none",
+                plot.margin = unit(c(5.5, 5.5, 25, 5.5), "pt"),
+                axis.text.x = element_text(size = 6)) +
+        facet_grid(. ~ gwas_type)
+
+# Make plot showing p-value changes in the CCR5 and MHC regions
+regional_results<- gwas_results_annotated %>% filter(
+        (CHROM == 3 & POS >= 45.5E6 & POS <= 47E6) | (CHROM == 6 & POS >= 29.5E6 & POS <= 33.5E6))
+
+# Fix facet labels for plotting
+regional_results$CHROM <- factor(
+        regional_results$CHROM,
+        levels = c("3", "6"),
+        labels = c("italic(CCR5)*' region'", "'MHC region'"))
+
+p_vals_regional <- ggplot(
+        data = regional_results,
+        aes(x = POS/1E6, y = -log10(P_corrected))) +
+        geom_point(aes(color = -neg_log10_P_standard_minus_corrected)) +
+        scale_color_gradient2(
+                name = expression(Delta*'[-Log'[10]*'(p)]'),
+                low = "#B2182B",
+                mid = "#FFFFBF",
+                high = "#2166AC",
+                midpoint = 0) +
+        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
+        labs(
+                x = "Base position (Mb)",
+                y = expression('-Log'[10]*'(p)')) +
+        geom_hline(
+                yintercept = -log10(genome_wide_significance),
+                linetype = "dotted") +
+        theme_bw() +
+        theme(
+                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
+                panel.grid.major.x = element_blank(),
+                axis.ticks.x = element_line(color = rep(c(NA, "black"), 23)),
+                panel.grid.minor.x = element_blank(),
+                plot.margin = unit(c(25, 5.5, 5.5, 5.5), "pt")) +
+        facet_grid(. ~ CHROM, labeller = label_parsed, scales = "free")
+
+# Assemble the plots as subplots
+png(
+        filename = paste(FIGDIR, "gwas_results.png", sep = "/"),
+        width = linewidth, height = height, unit = "cm", res = 300)
+gridExtra::grid.arrange(
+        overview_comparison,
+        p_vals_regional,
+        nrow = 2, heights = c(1.5, 2))
+dev.off()
 ```
 
-Q-Q plot of GWAS results
+Debugging
 ```{r}
+# Are trait values very different from what was inferred last time?
+old_traits <- read.delim(file = "/Users/nadeaus/Polybox/200207_phenotypes.txt", sep = "", header = T)
+trait_comparison <- merge(
+        x = old_traits, y = traits,
+        by.x = "IID", by.y = "id",
+        all = T)
 
+ggplot(data = trait_comparison, aes(x = total_spvl, y = trait)) + geom_point()  # calculated spVL indistinguishable
+ggplot(data = trait_comparison, aes(x = corrected_spvl, y = h.MWA)) + geom_point()  # corrected spVL very similar, well correlated
+
+# What about GWAS phenotypes file?
+gwas_phenotypes <- read.delim(file = "output/gwas_phenotypes.txt", sep = "", header = T)
+gwas_pheno_comparison <- merge(
+        x = old_traits, y = gwas_phenotypes,
+        by = "IID",
+        all = T)
+ggplot(data = gwas_pheno_comparison, aes(x = corrected_spvl, y = h.MWA)) + geom_point()  # also well correlated but previous GWAS had more samples?
 ```

### 2021-10-29 manuscript/hiv-gwas.tex

@@
 \usepackage{longtable} % for table that spans multiple pages
+\usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
 
@@
 
-\begin{figure}[H]
-	\begin{tikzpicture}
-
-	% Variance 
-	\begin{scope}[xshift = 0cm, yshift = 0cm]
-	\node[anchor = west] (alabel) at (-1, 4) {a) Variance partitioning};
-	\node[constnode] (H2h) at (0, 0) {0.25};
-	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
-	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {$H^2_{\bar{t}}$};
-	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
-	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
-	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{\epsilon}$};
-	\draw[taro] (H2h) -- (varh);
-	\draw[taro] (varz) -- (varh);
-	\draw[taro] (varz) -- (varv);
-	\draw[taro] (H2) -- (varv);
-	\draw[taro] (varh) -- (vare);
-	\draw[taro] (varv) -- (vare);
-	\draw[taro] (varz) -- (vare);
-	\end{scope}
-
-	% Virus
-	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
-	\node[anchor = west] (blabel) at (-1, 4.5) {b) Viral part of spVL};
-	\node[constnode] (theta) at (0, 1.5) {$\theta$};
-	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
-	\node[constnode] (alpha) at (3.6, 1.5) {$\alpha$};
-	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
-	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
-	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
-	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
-	\node[snode] (v) at (3.6, 3.25) {$\vec{g_v}$};
-	\draw[taro] (theta) -- (v);
-	\draw[taro] (g0) -- (v);
-	\draw[taro] (alpha) -- (v);
-	\draw[taro] (sigma) -- (v);
-	\draw[taro] (tree) -- (v);
-	\draw[taro] (varv) -- (sigma);
-	\draw[taro] (tbar) -- (sigma);
-	\draw[taro] (tbar) -- (tree);
-	\draw[taro] (alpha) -- (sigma);
-	\node[white] (varpic) at (3.6+0.9, -1.75) {};
-	\draw[taro] (varpic) -- (varv);
-	\node (vdist) at ($(v) + (1.8, 0.3)$) {Normal};
-	\draw (v) -- (vdist);
-	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
-	\end{scope}
-
-	% Host 
-	\begin{scope}[xshift = 0cm, yshift = -9cm]
-	\node[anchor = west] (clabel) at (-1, 5.55) {c) Host part of spVL};
-	\node[constnode] (d) at (1.8, 1.5) {2};
-	\node[detnode] (p) at (3.6, 1.5) {$p$};
-	\node[snode] (G) at (1.8, 3) {$G$};
-	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
-	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
-	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{g_h}$};
-	\node[constnode] (delta) at (3.6, 3) {$\delta$};
-	\node[detnode] (h) at (3.6, 4.5) {$\vec{g_h}$};
-	\node[constnode] (M) at (5.4, 3) {$M$};
-	\draw[taro] (d) -- (G);
-	\draw[taro] (p) -- (G);
-	\draw[taro] (varh) -- (p);
-	\draw[taro] (delta) -- (p);
-	\draw[taro] (M) -- (p);
-	\draw[taro] (M) -- (h);
-	\draw[taro] (delta) -- (h);
-	\draw[taro] (G) -- (h);
-	\node[white] (varpic) at (3.6, -1.5) {};
-	\draw[taro] (varpic) -- (varh);
-	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Normal};
-	\draw (h) -- (hdist);
-	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
-	\draw (G) -- (gdist);
-	\end{scope}
-
-	% Environment
-	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
-	\node[anchor = west] (dlabel) at (-1, 3) {d) Environmental part of spVL};
-	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
-	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
-	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{\epsilon}$};
-	\draw[taro] (zero) -- (e);
-	\draw[taro] (vare) -- (e);
-	\node[white] (varpic) at (0, -1.5) {};
-	\draw[taro] (varpic) -- (vare);
-	\node (edist) at ($(e) + (1.8, 0.3)$) {Normal};
-	\draw (e) -- (edist);
-	\end{scope}
-
-	\end{tikzpicture}
-	\caption{A graphical representation of the simulation model.
-Variables in solid squares are constant in the simulation scheme, whereas variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
-Arrows represent dependencies among variables and the dashed square represents repetition \parencite{Hohna2014}.}
-	\label{fig:sim-design}
-\end{figure}
-
 \begin{table}[H]
@@
 
-\begin{table}[H]
-	\begin{center}
-		\begin{tabular}{cccc} \hline 
-			Parameter & MLE & Maximum posterior probability & 95\% HPD \\ \hline 
-			${g_{0}}$ & 4.17 & 4.23 & 2.73, 5.88 \\
-			$\theta$ & 4.32 & 4.33 & 4.25, 4.42 \\
-			$\sigma$ & 5.15 & 4.98 & 3.45, 6.61 \\
-			$\alpha$ & 45.46 & 43.98 & 23.80, 63.36 \\
-			$\sigma_e$ & 0.55 & 0.56 & 0.49, 0.63 \\ 
-			$H^2_{\bar{t}}$ & 0.49 & 0.47 & 0.33, 0.60 \\ \hline
-		\end{tabular}
-		\caption{Inferred POUMM parameters.}
-		\label{tab:POUMM-parameters}
-	\end{center}
-\end{table}
+\newpage
+
+\begin{figure}[H]
+	\begin{tikzpicture}
+
+	% Variance 
+	\begin{scope}[xshift = 0cm, yshift = 0cm]
+	\node[anchor = west] (alabel) at (-1, 4) {A. Variance partitioning};
+	\node[constnode] (H2h) at (0, 0) {0.25};
+	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
+	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {$H^2_{\bar{t}}$};
+	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
+	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
+	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{\epsilon}$};
+	\draw[taro] (H2h) -- (varh);
+	\draw[taro] (varz) -- (varh);
+	\draw[taro] (varz) -- (varv);
+	\draw[taro] (H2) -- (varv);
+	\draw[taro] (varh) -- (vare);
+	\draw[taro] (varv) -- (vare);
+	\draw[taro] (varz) -- (vare);
+	\end{scope}
+
+	% Virus
+	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
+	\node[anchor = west] (blabel) at (-1, 4.5) {B. Viral part of spVL};
+	\node[constnode] (theta) at (0, 1.5) {$\theta$};
+	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
+	\node[constnode] (alpha) at (3.6, 1.5) {$\alpha$};
+	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
+	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
+	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
+	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
+	\node[snode] (v) at (3.6, 3.25) {$\vec{g_v}$};
+	\draw[taro] (theta) -- (v);
+	\draw[taro] (g0) -- (v);
+	\draw[taro] (alpha) -- (v);
+	\draw[taro] (sigma) -- (v);
+	\draw[taro] (tree) -- (v);
+	\draw[taro] (varv) -- (sigma);
+	\draw[taro] (tbar) -- (sigma);
+	\draw[taro] (tbar) -- (tree);
+	\draw[taro] (alpha) -- (sigma);
+	\node[white] (varpic) at (3.6+0.9, -1.75) {};
+	\draw[taro] (varpic) -- (varv);
+	\node (vdist) at ($(v) + (1.8, 0.3)$) {Normal};
+	\draw (v) -- (vdist);
+	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
+	\end{scope}
+
+	% Host 
+	\begin{scope}[xshift = 0cm, yshift = -9cm]
+	\node[anchor = west] (clabel) at (-1, 5.55) {C. Host part of spVL};
+	\node[constnode] (d) at (1.8, 1.5) {2};
+	\node[detnode] (p) at (3.6, 1.5) {$p$};
+	\node[snode] (G) at (1.8, 3) {$G$};
+	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
+	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
+	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{g_h}$};
+	\node[constnode] (delta) at (3.6, 3) {$\delta$};
+	\node[detnode] (h) at (3.6, 4.5) {$\vec{g_h}$};
+	\node[constnode] (M) at (5.4, 3) {$M$};
+	\draw[taro] (d) -- (G);
+	\draw[taro] (p) -- (G);
+	\draw[taro] (varh) -- (p);
+	\draw[taro] (delta) -- (p);
+	\draw[taro] (M) -- (p);
+	\draw[taro] (M) -- (h);
+	\draw[taro] (delta) -- (h);
+	\draw[taro] (G) -- (h);
+	\node[white] (varpic) at (3.6, -1.5) {};
+	\draw[taro] (varpic) -- (varh);
+	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Normal};
+	\draw (h) -- (hdist);
+	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
+	\draw (G) -- (gdist);
+	\end{scope}
+
+	% Environment
+	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
+	\node[anchor = west] (dlabel) at (-1, 3) {D. Environmental part of spVL};
+	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
+	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
+	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{\epsilon}$};
+	\draw[taro] (zero) -- (e);
+	\draw[taro] (vare) -- (e);
+	\node[white] (varpic) at (0, -1.5) {};
+	\draw[taro] (varpic) -- (vare);
+	\node (edist) at ($(e) + (1.8, 0.3)$) {Normal};
+	\draw (e) -- (edist);
+	\end{scope}
+
+	\end{tikzpicture}
+	\caption{A graphical representation of the simulation model.
+Variables in solid squares are constant in the simulation scheme, whereas variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
+Arrows represent dependencies among variables and the dashed square represents repetition \parencite{Hohna2014}.}
+	\label{fig:sim-design}
+\end{figure}
 
@@
 
-\begin{table}[H]
-    \begin{center}
-    \begin{tabular}{lc}
-    \hline
-    Sample filter & Number of samples remaining \\
-    \hline
-    Subtype B pol sequences & 1516 \\
-    With paired spVL measurement & 1516 \\
-    $>$ 750 characters in sequence & 1493 \\
-    Individual is of European ancestry & 1396 \\
-    Kinship coefficient $>$ 0.09375 & TODO \\ 
-    \hline
-    \end{tabular}
-    \caption{Number of samples for GWAS after sequential filtering steps.}
-    \label{tab:sample-filtering}
-    \end{center}
-\end{table}
-
-\begin{table}[H]
-	\begin{center}
-		\begin{tabular}{lc}
-			\hline
-			Variant filter & Number of variants remaining \\
-			\hline 
-			Raw data & 76979521 \\
-			Missing genotype rate $>$ 0.01 & 7452957 \\
-			Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 7452425 \\
-			Minor allele frequency $<$ 0.01 & 4556770 \\ \hline			
-		\end{tabular}
-		\caption{Number of variants for GWAS after sequential filtering steps.}
-		\label{tab:variant-filtering}
-	\end{center}
-\end{table}
-
 \begin{figure}[H]
@@
 	\centering
-	\includegraphics[width=0.75\linewidth]{figures_archived/200207_qq_plots.png}
+	\includegraphics[width=\linewidth]{figures_archived/200207_qq_plots.png}
 	\caption{Quartile-quartile plots from association tests.}
@@
 
+\end{doublespace}  % don't doublspace tables
+
 \newpage
-\begin{longtable}[H]{p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
+
+\begin{table}[H]
+\caption{POUMM parameter estimates for spVL based on Swiss data. HPD = Highest posterior density.}
+	\begin{tabular}{lccc} \hline 
+	Parameter & Posterior mean & 95\% HPD \\ \hline 
+	${g_{0}}$ & 4.23 & (1.72, 6.71) \\
+	$\theta$ & 4.47 & (4.37, 4.58) \\
+	$\sigma$ & 5.25 & (2.37, 7.9) \\
+	$\alpha$ & 57.65 & (19.49, 95.2) \\
+	$\sigma_e$ & 0.54 & (0.43, 0.65) \\ 
+	$H^2_{\bar{t}}$ & 0.45 & (0.24, 0.67) \\ \hline
+	\end{tabular}
+	\label{tab:POUMM-parameters}
+\end{table}
+
+\begin{table}[H]
+\caption{POUMM parameter estimates for spVL from previous studies.}
+	\begin{tabular}{lcccp{0.35\linewidth}}
+	\hline 
+	Parameter & Value & Uncertainty & Reference & Notes \\
+	\hline 
+	$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citealt{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
+	$\theta$ & 4.45 & (4.41 - 4.49) HPD & \citealt{Mitov2018} & \\
+	$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citealt{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
+	$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citealt{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
+	$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citealt{Mitov2018} & \\
+	$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citealt{Bertels2018} & \\
+	$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citealt{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
+	$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citealt{Mitov2018} & \\
+	$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citealt{Blanquart2017} & \\
+	$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citealt{Mitov2018} & \\
+	$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citealt{Blanquart2017} & \\
+	\hline 
+	\end{tabular}
+	\label{tab:POUMMparams}
+\end{table}
+
+\begin{table}[H]
+\caption{Summary statistics for log spVL in previously sampled populations.
+$\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$).
+Values from \protect\citet{Blanquart2017, Mitov2018} are empirical; values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
+	\begin{tabular}{lcc} \hline 
+	Measurement & Value & Reference \\ \hline 
+	$\bar{z}$ & $\approx$ 4.5 & \citealt{Mitov2018} \\
+	$\bar{z}$ &  4.4 & \citealt{Blanquart2017} \\
+	$\bar{z}$ & $\approx$ 4.5 & \citealt{Bonhoeffer2015} \\
+	$\sigma^2_z$ &  0.73 &  \citealt{Mitov2018} \\
+	$\sigma^2_z$ &  0.50 & \citealt{Blanquart2017} \\
+	$\sigma^2_z$ & $\approx$ 0.5 & \citealt{Bonhoeffer2015} \\ \hline 
+	\end{tabular}
+	\label{tab:spVLmeasurements}
+\end{table}
+%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}
+
+\begin{table}[H]
+\caption{Number of samples for GWAS after sequential filtering steps.}
+    \begin{tabular}{lc}
+    \hline
+    Sample filter & Number of samples remaining \\
+    \hline
+    Subtype B pol sequences & 1516 \\
+    With paired spVL measurement & 1516 \\
+    $>$ 750 characters in sequence & 1493 \\
+    Individual is of European ancestry & 1396 \\
+    Kinship coefficient $>$ 0.09375 & 1392 \\ 
+    \hline
+    \end{tabular}
+    \label{tab:sample-filtering}
+\end{table}
+
+\begin{table}[H]
+\caption{Number of variants for GWAS after sequential filtering steps.}
+	\begin{tabular}{lc}
+	\hline
+	Variant filter & Number of variants remaining \\
+	\hline 
+	Raw data & 76979521 \\
+	Missing genotype rate $>$ 0.01 & 7452957 \\
+	Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 7452425 \\
+	Minor allele frequency $<$ 0.01 & 4556770 \\ \hline			
+	\end{tabular}
+	\label{tab:variant-filtering}
+\end{table}
+
+\begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
+		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citet{McLaren2015} compared to results from the current study.}
 		\hline
@@
 		\hline
-		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citet{McLaren2015} compared to results from the current study.}
 		\label{tab:full-mclaren-comparison}
@@
 
-\begin{table}[H]
-	\begin{center}
-		\begin{tabular}{cccl}
-			\toprule 
-			Parameter & Value & Uncertainty & Reference \\
-			\midrule 
-			$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citealt{Mitov2018} \\
-			$\theta$ & 4.45 & (4.41 - 4.49) HPD& \citealt{Mitov2018}  \\
-			$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citealt{Bertels2018} \\
-			$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citealt{Blanquart2017} \\
-			$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citealt{Mitov2018} \\
-			$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citealt{Bertels2018} \\
-			$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citealt{Blanquart2017}** \\
-			$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citealt{Mitov2018} \\
-			$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citealt{Blanquart2017} \\
-			$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citealt{Mitov2018} \\
-			$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citealt{Blanquart2017} \\
-			\bottomrule 
-		\end{tabular}
-		\caption{POUMM parameters previously inferred from HIV data.
-Notes for each study: \protect\citet{Mitov2018} = 8,483 UK HIV cohort individuals, pol tree; \protect\citet{Bertels2018} = 3,036 SHCS individuals, pol tree; \protect\citet{Blanquart2017} = 1,581 subtype B individuals from BE, CH, FR, NL, UK, whole genome tree, **limited $\alpha$ to $\le$ 10}
-		\label{tab:POUMMparams}
-	\end{center}
-\end{table}
-
-\begin{table}[H]
-	\begin{center}
-			\begin{tabular}{ccl} \toprule 
-			Measurement & Value & Reference \\ \midrule 
-			$\bar{z}$ & $\approx$ 4.5 & \citealt{Mitov2018} \\
-			$\bar{z}$ &  4.4 & \citealt{Blanquart2017} \\
-			$\bar{z}$ & $\approx$ 4.5 & \citealt{Bonhoeffer2015} \\
-			$\sigma^2_z$ &  0.73 &  \citealt{Mitov2018} \\
-			$\sigma^2_z$ &  0.50 & \citealt{Blanquart2017} \\
-			$\sigma^2_z$ & $\approx$ 0.5 & \citealt{Bonhoeffer2015} \\ \bottomrule 
-		\end{tabular}
-		\caption{Summary statistics for log spVL in previously sampled populations.
-$\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$).
-Values from \protect\citet{Blanquart2017, Mitov2018} are empirical while values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
-		\label{tab:spVLmeasurements}
-	\end{center}
-\end{table}
-%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}
-
-\begin{figure}[H]
-	\includegraphics[width=\linewidth]{images/HIV_tree_branch_lengths}
-	\caption{Left plot: branch length distributions for UK HIV tree and our simulated HIV tree prior to pruning down to 500 tips.
+% \begin{figure}[H]
+% 	\includegraphics[width=\linewidth]{images/HIV_tree_branch_lengths}
+% 	\caption{Left plot: branch length distributions for UK HIV tree and our simulated HIV tree prior to pruning down to 500 tips.
 The mean root-to-tip distance for the UK tree is 0.14 and the mean branch length is 0.02.
@@
 The mean root-to-tip distance for the pruned HIV tree is 0.14 and the mean branch length is 0.03.}
-	\label{fig:branchlengths}
-\end{figure}
-
-\end{doublespace}
+% 	\label{fig:branchlengths}
+% \end{figure}
 

### 2021-10-29 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 		\caption{Results from the simulation study.
-We simulated pathogen and non-pathogen (host \& environmental) effects on the spVL trait under the POUMM with different trait heritability values (y-axis) and selection strengths (x-axis).
-For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS using these estimates.
-(A) shows how well we can estimate the host effect using our method (left side) compared to just the scaled trait value (right side).
-(B) shows how GWAS can theoretically be improved by using the true (normally unknown) non-pathogen effects on spVL (left).
-It can still be improved in certain scenarios using our estimate for this value (middle) compared to the current standard - using just the (scaled) trait value (right).
-Tile color represents the average value over 20 simulated datasets of 500 samples for each heritability and selection strength value.
+We simulated pathogen genetic effects, host genetic effects, and environmental effects on the spVL trait under the POUMM with different trait heritability values (y-axis) and selection strengths (x-axis).
+For each simulated dataset, we applied our method to estimate the non-pathogen effects (host genetic \& environmental effects) and performed GWAS using these estimates.
+(A) shows how well we can estimate the host genetic effect using our method (left) compared to the whole trait value (scaled such that it has mean 0) (right).
+(B) shows the theoretical improvement to GWAS using the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) in different simulated scenarios.
+For comparison, we show GWAS results using the whole trait value (right).
+Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples for each heritability and selection strength value.
 The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data.
 RMSE = Root mean square error, TPR = True positive rate.}
-		% (RMSE) with respect to simulated values for the host part of spVL ($\bm{{g_h}}$) as a function of spVL trait heritability and selection strength on the trait.
-The left plot is for the POUMM-based estimate of the non-pathogen part of the trait ($\bm{\hat{e}}$) and the right plot is for the raw trait value scaled by the mean trait value, such that it is centered at zero ($\bm{z} - \bm{\bar{z}}$).
-(B) True positive rate (TPR) in GWAS performed using true (simulated) values for the non-pathogen part of the trait ($\bm{e}$), the POUMM-based estimate for the non-pathogen part of the trait ($\bm{\hat{e}}$), and the raw trait value scaled by the mean trait value ($\bm{z} - \bm{\bar{z}}$).
-The color of each plot tile in (A) and (B) represents the average value over 20 simulated datasets of 500 samples.
-The black point shows our estimates for the heritability of and selection strength on the spVL trait based on Swiss data.}
 		\label{fig:simulationResults}

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
figured out that PLINK output columns differed, fixed run-gwas script and made new plots

@@
 ```
-
-Debugging
-```{r}
-# Are trait values very different from what was inferred last time?
-old_traits <- read.delim(file = "/Users/nadeaus/Polybox/200207_phenotypes.txt", sep = "", header = T)
-trait_comparison <- merge(
-        x = old_traits, y = traits,
-        by.x = "IID", by.y = "id",
-        all = T)
-
-ggplot(data = trait_comparison, aes(x = total_spvl, y = trait)) + geom_point()  # calculated spVL indistinguishable
-ggplot(data = trait_comparison, aes(x = corrected_spvl, y = h.MWA)) + geom_point()  # corrected spVL very similar, well correlated
-
-# What about GWAS phenotypes file?
-gwas_phenotypes <- read.delim(file = "output/gwas_phenotypes.txt", sep = "", header = T)
-gwas_pheno_comparison <- merge(
-        x = old_traits, y = gwas_phenotypes,
-        by = "IID",
-        all = T)
-ggplot(data = gwas_pheno_comparison, aes(x = corrected_spvl, y = h.MWA)) + geom_point()  # also well correlated but previous GWAS had more samples?
-```

### 2021-10-29 scripts/bash/run-gwas.sh

@@ -37,7 +37,7 @@ head -1 output/${DATA_PREFIX}.h.MWA.glm.linear > output/${DATA_PREFIX}.h.MWA.glm
 grep -h 'ADD' output/${DATA_PREFIX}.h.MWA.glm.linear >> output/${DATA_PREFIX}.h.MWA.glm.linear.nocovariates
 
 # Paste results from both trait values together
-paste "output/${DATA_PREFIX}.trait.glm.linear.nocovariates" "output/${DATA_PREFIX}.h.MWA.glm.linear.nocovariates" | awk '{print $1, $2, $3, $9, $12, $22, $25}' > output/gwas_results.txt
+paste "output/${DATA_PREFIX}.trait.glm.linear.nocovariates" "output/${DATA_PREFIX}.h.MWA.glm.linear.nocovariates" | awk '{print $1, $2, $3, $9, $12, $21, $24}' > output/gwas_results.txt
 HEADER="CHROM POS ID BETA_standard P_standard BETA_corrected P_corrected"
 sed -i.bak "1 s/^.*$/$HEADER/" output/gwas_results.txt
 

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
fix black point in fig 2 to match actual POUMM estimate, add gwas results figure

@@
 POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
+poumm_parameter_estimates <- read.csv(file = paste(OUTDIR, "poumm_parameter_estimates.csv", sep = "/"))
 pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"))
@@
 # Get POUMM values for HIV spVL to annotate plot with
+H2_estimate <- poumm_parameter_estimates[poumm_parameter_estimates$stat == "H2tMean", "PostMean"]
+alpha_estimate <- poumm_parameter_estimates[poumm_parameter_estimates$stat == "alpha", "PostMean"]
+
+# Breaks scale is defined as seq(0.5, n.breaks - 0.5, 1) in plotting function, need to translate real values to breaks scale
+get_xval_for_sim_plots <- function(real_vals, estimate) {
+      alpha_vals <-
+      base_val <- rev(real_vals[real_vals < estimate])[1]
+      next_val <- real_vals[real_vals > estimate][1]
+      base_xval <- which(real_vals == base_val)
+      next_xval <- which(real_vals == next_val)
+      xval <- base_xval - 0.5 + (next_xval - base_xval) * (estimate - base_val) / (next_val - base_val)
+      return(xval)
+}
+
 spvl_data <- data.frame(
         inference.method = "h.MWA.p.value",
-        H2 = 3.9,
-        alpha = 8.046)  # values are hand-calculated because the breaks scale is defined as seq(0.5, n.breaks - 0.5, 1) in plotting function
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75), estimate = H2_estimate),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = alpha_estimate))
 

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
squick model figure a bit

@@
       labs(
-        x = expression('Substitutions site'^-1 * ' year'^-1),
+        x = element_blank(),
         y = expression(italic(g)[v])) +
@@
 # Align OU and tree plots
-left_side <- plot_grid(OU_plot, tree_plot, align = "v", axis = "lr", nrow = 2, rel_heights = c(2, 1), labels = c("A", "C"))
+left_side <- plot_grid(OU_plot, tree_plot, align = "v", axis = "lr", nrow = 2, rel_heights = c(3, 1), labels = c("A", "C"))
 right_side <- plot_grid(NULL, e_plot, NULL, nrow = 3, rel_heights = c(1, 2, 1), labels = c("", "B", ""))
@@
       filename = paste(FIGDIR, "model_figure.png", sep = "/"),
-      width = linewidth, height = height, units = "cm", res = 300)
-      plot_grid(left_side, NULL, right_side, ncol = 3, rel_widths = c(1.5, 0.2, 1))
+      width = linewidth * 2/3, height = height * 2/3, units = "cm", res = 300)
+      plot_grid(left_side, NULL, right_side, ncol = 3, rel_widths = c(1.5, 0.1, 1))
 dev.off()

### 2021-10-29 manuscript/hiv-gwas.tex

@@
 
-Our premise is that spVL is the sum of three parts: direct host genetic effects, viral genetic effects, and environmental effects.
-Uniquely, the viral part is heritable from one transmission partner to another \parencite{Mitov2018, Bertels2018}.
-To characterize these parts, we use a phylogenetic mixed model (PMM, \cite{Housworth2004}).
+Our premise is that spVL is the sum of several parts: direct host genetic effects, pathogen genetic effects, interaction effects between host and pathogen genetic factors, and environmental effects.
+Of these, only pathogen genetic effects should be heritable from one transmission partner to another \parencite{Mitov2018, Bertels2018}.
+To characterize these effects, we use a phylogenetic mixed model (PMM, \cite{Housworth2004}).
 PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts.
-The heritable viral part is modeled by a random process occurring in continuous time along the branches of the viral phylogeny.
-The non-heritable host and environmental parts are modeled as gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).
+In our case, the heritable part comes from pathogen genetic effects and the non-heritable part comes from the other effects.
+Then, viral genetic effects are modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny.
+The other, non-heritable effects are modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).
+
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=0.75\linewidth]{figures/model_figure}
+		\caption{The POUMM statistical model for the spVL trait.
+(A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny.
+The dashed line in (A) shows the expected value of the trait at each point in time under the model.
+(B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable.
+Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
+		\label{fig:spVLModel}
+	\end{center}
+\end{figure}
 
@@
 \end{align}
-
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=0.75\linewidth]{figures/model_figure}
-		\caption{The POUMM statistical model for the spVL trait.
-(A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny.
-The dashed line in (A) shows the expected value of the trait at each point in time under the model.
-(B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable.
-Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
-		\label{fig:spVLModel}
-	\end{center}
-\end{figure}
 
@@
 	\centering
-	\includegraphics[width=\linewidth]{figures_archived/200207_qq_plots.png}
-	\caption{Quartile-quartile plots from association tests.}
+	\includegraphics[width=\linewidth]{figures/qq_plots.png}
+	\caption{Quartile-quartile plots from association tests. The dashed green line shows the y = x line.}
 	\label{fig:qq-plots}

### 2021-10-29 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{biblatex}
-\addbibresource{references.bib}
 \usepackage{graphicx}
@@
 \usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
+
+\usepackage[backend=bibtex,style=nature]{biblatex} % Borrowing citation formatting from Nature
+\bibliography{references}
 
@@
 Several GWAS have been done for spVL \parencite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
-The largest to-date by \citet{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
+The largest to-date by \cite{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
 All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}).
@@
 
-Our premise is that spVL is the sum of several parts: direct host genetic effects, pathogen genetic effects, interaction effects between host and pathogen genetic factors, and environmental effects.
-Of these, only pathogen genetic effects should be heritable from one transmission partner to another \parencite{Mitov2018, Bertels2018}.
-To characterize these effects, we use a phylogenetic mixed model (PMM, \cite{Housworth2004}).
+Variation in spVL comes from several sources: direct host genetic effects, pathogen genetic effects, interaction effects between host and pathogen genetic factors, and other environmental effects.
+Of these, only pathogen genetic effects are heritable from one transmission partner to another \parencite{Leventhal2016}.
+To characterize these effects, we use a phylogenetic mixed model (PMM) \cite{Housworth2004}.
 PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts.
-In our case, the heritable part comes from pathogen genetic effects and the non-heritable part comes from the other effects.
-Then, viral genetic effects are modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny.
-The other, non-heritable effects are modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).
+In our case, pathogen genetic effects comprise the heritable part and all other effects comprise the non-heritable part.
+Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny.
+The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).
 
@@
 	\begin{center}
-		\includegraphics[width=0.75\linewidth]{figures/model_figure}
+		\includegraphics[width=0.462\linewidth]{figures/model_figure}
 		\caption{The POUMM statistical model for the spVL trait.
@@
 
-So far, spVL has been modelled using PMMs with two different types of random processes.
-The Brownian Motion (BM) process assumes that trait values are unbounded (i.e.
-spVL can attain any value).
-In contrast, the Ornstein-Uhlenbeck (OU) process assumes the trait fluctuates around an optimal value (i.e.
-extreme spVL values are very unlikely) \cite{Butler2004}.
-We argue the OU assumption is more realistic for spVL.
-\citet{Fraser2014} hypothesized that spVL is under stabilizing selection to maximize viral transmission.
-Accordingly, both \citet{Mitov2018} and \citet{Bertels2018} found higher statistical support for the OU process compared to BM for spVL evolution.
+So far, PMMs with two types of random processes have been used to model spVL evolution.
+The Brownian Motion (BM) process assumes unbounded trait values, i.e.
+spVL can attain any value.
+The Ornstein-Uhlenbeck (OU) process assumes trait values fluctuate around an optimal value, i.e.
+extreme spVL values are unlikely.
+\cite{Mitov2018} and \cite{Bertels2018} previously showed the OU process has higher statistical support for spVL.
+This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \cite{Fraser2014}.
 Therefore, we assume the OU process.
-This model, called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM), is described in detail by \citet{Mitov2018}, but we review the main points here in the spVL context.
-
-Under the POUMM, the spVL trait $z$ is the sum of direct host genetic effects $g_h$, viral genetic efects $g_v$, and environmental effects $\epsilon$.
-We can group the phylogeny-independent effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$:
+The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \cite{Mitov2018}.
+Here, we review the main relevant points in the spVL context.
+
+Under the POUMM, the spVL trait $z$ is the sum of viral genetic efects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$.
+We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$:
 \begin{equation}
@@
 \end{equation}
-Finally, we remember that $e$ is the non-heritable, environmental part of spVL.
+Next, we remember that $e$ is the non-heritable, environmental part of spVL.
 $e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
@@
 
-\subsection{Estimating and removing the virus-determined part of spVL}
-
-Given the assumptions of the POUMM, we can derive a maximum-likelihood estimate for the virus-determined part of spVL for each individual in a cohort at the time they were sampled.
-Our goal is to estimate and remove this effect from measured spVL prior to GWAS.
+Finally, broad-sense trait heritability can be calculated as the fraction of total trait variance that is heritable:
+\begin{equation}
+	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(e)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_e}
+\end{equation}
+
+\subsection{Teasing apart pathogen and non-pathogen effects on spVL}
+
+Given the assumptions of the POUMM, we can estimate a heritable pathogen effect on spVL and a non-heritable, host and environmental effect on spVL.
+Here, we derive a maximum-likelihood estimate for these values for individuals in a cohort, given measured spVL values and a pathogen phylogeny linking the infecting strains.
 
@@
 
-	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
+	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
 \end{proof}
 
-Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the viral part of spVL, taking into account all available information - the viral phylogeny, the inferred POUMM parameters, and measured spVL values.
-This estimator is an inverse-variance weighted average of information from the POUMM evolutionary model ($\bm{\mu}_{OU}$) and measured spVL ($\bm{z}$).
-Intuitively, the estimate for $g_v$ is closer to measured spVL if spVL is not very heritable, and is closer to the expected value under the POUMM if spVL is highly heritable.
-
-Given that we have an estimate for $\bm{g}_v$, we can finally estimate $\bm{e}$, representing the spVL value with viral effects removed: 
+Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the viral effect on spVL, taking into account all available information - measured spVL, the pathogen phylogeny, and inferred POUMM parameters.
+This estimator is an inverse-variance weighted average of measured spVL ($\bm{z}$) and information from the POUMM evolutionary model ($\bm{\mu}_{OU}$).
+In other words, $g_v$ will be closer to measured spVL if spVL is not very heritable.
+If spVl is highly heritable, $g_v$ will be closer to the expected value under the POUMM, i.e.
+take more information from the phylogenetic relationships between infecting strains.
+
+Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{e}$, the spVL value $without$ pathogen effects: 
 \begin{equation}
@@
 
-\subsection{A new infectious disease GWAS framework}
-
-We propose that by estimating and removing spVL variability due to the pathogen we can better measure the direct effects of human genetic variants on spVL. Our proposed framework is as follows:
+We will use this value to try to improve upon standard GWAS methods in infectious disease.
+
+\subsection{A POUMM-based GWAS framework for infectious disease}
+
+We propose to improve standard GWAS for infectious diseases by estimating and removing trait variability due to pathogen genetic effects. Our new framework is as follows:
 
 \begin{enumerate}
-	\item Sample host genotypes, host trait values, and pathogen genetic sequences from a cohort.
-	\item Construct a pathogen phylogeny using the pathogen genetic sequences.
-	\item Fit the POUMM to the trait values and the phylogeny. This produces estimates for parameters $g_0, \alpha, \theta, \sigma,\text{ and }\sigma_e$.
-	\item Generate maximum-likelihood estimates for the pathogen-determined and environmental parts of the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
-	\item Perform GWAS, using the estimated environmental part of the trait as the response variable instead of raw trait values. 
+	\item Sample host genotypes, trait values, and pathogen genome sequence data from a cohort.
+	\item Construct a pathogen phylogeny using the pathogen genome sequences.
+	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with e.g. the R package POUMM \cite{Mitov2017a-POUMM}.
+	\item Generate maximum-likelihood estimates for the pathogen and corresponding non-pathogen effects on the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
+	\item Perform GWAS with only the non-pathogen effects on the trait as the response variable.
 \end{enumerate}
@@
 
-To test the theoretical best-case performance of our framework, we simulated data under the POUMM.
-Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression used for each parameter.
+To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data.
+Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression for each parameter.
+In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength.
+To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly.
+We fixed other variables to plausible values based on the spVL literature.
 
@@
 
-First, we evaluated estimator accuracy with respect to $\bm{g}_h$, the sum of additive host genetic effects.
-This would be the ideal GWAS response variable to identify direct host determinants of spVL.
-We compute the accuracy of our proposed GWAS endpoint $\bm{\hat{e}}$ and the typical GWAS endpoint, measured spVL, using root mean squared error (RMSE).
-In our simulations we expect (scaled) measured spVL, $\bm{z} - \bar{\bm{z}}$, where $\bar{\bm{z}}$ is the average spVL, to be Gaussian distributed around $\bm{g}_h$ with variance equal to $\sigma^2_e + \sigma^2_{g_v} \approx 0.55$ (see table \ref{tab:sim-params} for simulation parameters).
-This should yield an RMSE of $\approx$ 0.74 for $\bm{z} - \bar{\bm{z}}$ as an estimate of $\bm{e}$, which matches simulation results (figure \ref{fig:simulationResults}A, scaled spVL).
-By incorporating information from the viral phylogenetic tree in calculating $\bm{\hat{e}}$, we were better able to estimate $\bm{e}$.
-The greatest improvement was possible in simulation scenarios in which spVL was highly heritable, meaning the virus contributed a great deal of variability to spVL.
-Estimates were also better in scenarios with lower selection strength since the correlation in $g_v$ among related tips is higher when $g_v$ can drift further from the long-term optimum and the stochastic fluctuations characterized by $\sigma$ are correspondingly lower to maintain the same viral genetic variance $\sigma^2_{g_v}$ and the same broad-sense heritability $H^2$.
+First, we evaluated how well our method estimated the additive host genetic effects from the simulated data.
+Additive host genetic effects represent an ideal baseline for GWAS that seek to identify host genetic determinants of a trait.
+Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can generate more accurate estimates than the trait value alone.
+To be fair, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
+In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation set-up.
+By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable and under low selection pressure.
+This is because high heritability means the pathogen greatly effects the trait value.
+Then, when the trait is under weak selection, pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability.
+Thus, our method performs well when an infectious disease trait is highly heritable and trait values are highly correlated amongst closely related members of a transmission chain, but vary widely within the entire cohort.
 
@@
 
-Next, we characterize the evolutionary scenarios in which our framework improves GWAS power.
-More accurate estimation of the non-viral part of spVL does not guarantee improved GWAS power.
+Next, we characterized the evolutionary scenarios under which our framework can actually improve GWAS power.
+We used the true positive rate (TPR) to evaluate the fraction of simulated host genetic variants with direct effects we could recover as being significantly associated with the trait.
+We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the unmodified trait value.
+Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%.
+If were were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable.
+We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects.
+In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.
+
+% More accurate estimation of the non-viral part of spVL does not guarantee improved GWAS power.
 For example, if the viral part of spVL were constant across all viruses, association results (significance and effect size) would not be effected by a viral contribution to the trait.
-To determine under what evolutionary scenarios our method can improve GWAS power, we performed GWAS on the simulated data.
-Under our simulation set-up a standard GWAS would have $\approx$ 30\% power to detect a host variant explaining 1.25\% of the variability in spVL.
+% Under our simulation set-up a standard GWAS would have $\approx$ 30\% power to detect a host variant explaining 1.25\% of the variability in spVL.
 This matches simulation results (figure \ref{fig:simulationResults}B, scaled spVL).
-Using our proposed framework, we were able to improve GWAS power in simulated scenarios with selection strength $<$ 10-20 time$^{-1}$ and spVL heritability $>$ 35-45\% (figure \ref{fig:simulationResults}B, $\bm{\hat{e}}$).
-If were were able to perfectly estimate $\bm{e}$, GWAS power would increase across all values of selection strength so long as the trait is highly heritable (figure \ref{fig:simulationResults}, $\bm{e}$).
-In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by removing an estimated pathogen-determined effect.
 
@@
 		\caption{Results from the simulation study.
-We simulated pathogen genetic effects, host genetic effects, and environmental effects on the spVL trait under the POUMM with different trait heritability values (y-axis) and selection strengths (x-axis).
-For each simulated dataset, we applied our method to estimate the non-pathogen effects (host genetic \& environmental effects) and performed GWAS using these estimates.
-(A) shows how well we can estimate the host genetic effect using our method (left) compared to the whole trait value (scaled such that it has mean 0) (right).
-(B) shows the theoretical improvement to GWAS using the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) in different simulated scenarios.
-For comparison, we show GWAS results using the whole trait value (right).
-Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples for each heritability and selection strength value.
+We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters.
+For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
+(A) shows that our method (left) can generate more accurate estimates of host genetic effects than the trait value, scaled by its mean (right).
+(B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the unmodified trait value (right).
+Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
 The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data.
@@
 
-Finally, we demonstrate our framework on empirical data from the Swiss HIV Cohort Study (SHCS).
+Finally, we applied our framework to empirical data from the Swiss HIV Cohort Study (SHCS).
 In the ``standard GWAS" we followed a typical GWAS methodology,  i.e.
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{figures_archived/200207_gwas_results}
+		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
 		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load.
@@
 
-The top-associated SNPs from the \citet{McLaren2015} paper were excluded from the main analysis because of too-high missing rates (3\% for rs1015164 and 4\% for rs59440261).
-In order to directly compare our results with \citet{McLaren2015}, we ran an additional GWAS for these two SNPs anyways.
+The top-associated SNPs from the \cite{McLaren2015} paper were excluded from the main analysis because of too-high missing rates (3\% for rs1015164 and 4\% for rs59440261).
+In order to directly compare our results with \cite{McLaren2015}, we ran an additional GWAS for these two SNPs anyways.
 Both SNPs became less significant with phylogenetic correction (table \ref{tab:comp-gwas}).
 
-Again following the procedure of \citet{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
+Again following the procedure of \cite{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
 Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).
@@
 		\end{tabular}
-		\caption{Top association results from \citet{McLaren2015} compared to association results in the current study.}
+		\caption{Top association results from \cite{McLaren2015} compared to association results in the current study.}
 		\label{tab:comp-gwas}
@@
 	\end{tabular}
-	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citet{McLaren2015} compared to results from the current study.
+	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.
 The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al.
@@
 The Bayesian inference method implemented in this package requires specification of a prior distribution for each parameter.
-We used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$.
+We used the same, broad prior distributions as in \cite{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$.
 We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000.
@@
 \end{itemize}
+
+\printbibliography
 
@@
 $\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$).
-Values from \protect\citet{Blanquart2017, Mitov2018} are empirical; values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
+Values from \protect\cite{Blanquart2017, Mitov2018} are empirical; values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
 	\begin{tabular}{lcc} \hline 
@@
 \begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
-		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citet{McLaren2015} compared to results from the current study.}
+		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.}
 		\hline
@@
 
-%\citet{Mitov2018} and \citet{Bertels2018} both found higher statistical support for the POUMM model of spVL evolution as opposed to the corresponding model assuming Brownian motion only.
-This makes sense in the context of the hypothesis proposed by \citet{Fraser2014}, in which spVL is under stabilizing selection to maximize transmissibility of the virus.
+%\cite{Mitov2018} and \cite{Bertels2018} both found higher statistical support for the POUMM model of spVL evolution as opposed to the corresponding model assuming Brownian motion only.
+This makes sense in the context of the hypothesis proposed by \cite{Fraser2014}, in which spVL is under stabilizing selection to maximize transmissibility of the virus.
 On the other hand, the assumptions that $e$ are normally distributed, mutually indpendent, and have an expected value of 0 are consistent with the idea that genetic effects from several unlinked host loci and environmental effects from many independent factors are each equally likely to increase an individual's spVL as decrease it.
@@
 %The success of our method in identifying new host determinants of spVL depends on how much host-determined variation in spVL actually exists.
-In the largest GWAS for HIV spVL to date, \citet{McLaren2015} estimated that additive effects from common human genetic variants explain 20\% of the variability in spVL.
+In the largest GWAS for HIV spVL to date, \cite{McLaren2015} estimated that additive effects from common human genetic variants explain 20\% of the variability in spVL.
 Of this quantity, 94.5\% was attributable to variants in just two regions in the human genome.
 \todo{Were there variants other than HLA in the MHC region?} Associations were with alleles in the CCR5 gene, which encodes a co-receptor for HIV entry into CD4+ T-cells, and the human leukocyte antigen (HLA) genes, which encode proteins that present certain host and pathogen-derived peptides at the surface of cells.
-These associations have been well-replicated in several GWAS \citet{Naranbhai2017}.
+These associations have been well-replicated in several GWAS \cite{Naranbhai2017}.
 This results suggest that nearly all host-determined variation in spVL can be explained by previously discovered variants.
 %
-%In a joint analysis of both host and viral genotypes, \citet{Bartha2017} showed that nearly all of the variation in spVL explained by human genetic variation overlaps that explained by HIV genetic variation.
+%In a joint analysis of both host and viral genotypes, \cite{Bartha2017} showed that nearly all of the variation in spVL explained by human genetic variation overlaps that explained by HIV genetic variation.
 If this is the case then our method to remove the viral contribution to spVL prior to GWAS should yield only associations between host variants and spVL that are independent of the virus.
@@
 %According to the POUMM, the unobserved viral component of spVL for each host, $g_{v_i}$, evolves along the tree according to an Ornstein-Uhlenbeck (OU) process.
-The OU process was first introduced by \citet{Lande1976} as a model of evolution under both drift and stabilizing selection, so we are assuming $g_{v_i}$ continually evolves towards an evolutionary optimum in each viral lineage but is subject to random evolutionary fluctuations.
+The OU process was first introduced by \cite{Lande1976} as a model of evolution under both drift and stabilizing selection, so we are assuming $g_{v_i}$ continually evolves towards an evolutionary optimum in each viral lineage but is subject to random evolutionary fluctuations.
 Mathematically, $g_{v_i}$ is a realization of a Gaussian distributed random variable $G_{v_i} \sim \mathcal{N}(\mu_{OU_i}, \sigma^2_{OU_i})$, where:
@@
 % 	
-% 	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
+% 	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
 % 	
@@
 Since a hypothesis test for significant association is performed for each of millions of genetic variants, GWAS performed in this way incurs a heavy multiple-testing burden.
-Despite a strong correction for multiple testing, \citet{McLaren2015} were able to confirm two
+Despite a strong correction for multiple testing, \cite{McLaren2015} were able to confirm two
 
 %So far, it is well established that both human host and viral genetic factors contribute to spVL.
-\citet{McLaren2015} found that 25\% of the variability in spVL in a mixed European cohort could be explained by additive genetic effects from the human genome.
-More recently, \citet{Mitov2018} established that at least 20\% of the variability in spVL in a UK cohort could be explained by viral genetic variation.
+\cite{McLaren2015} found that 25\% of the variability in spVL in a mixed European cohort could be explained by additive genetic effects from the human genome.
+More recently, \cite{Mitov2018} established that at least 20\% of the variability in spVL in a UK cohort could be explained by viral genetic variation.
 Furthermore, depending on the methodological assumptions made and the cohort studied, estimates for the broad-sense viral heritability can vary widely \parencite{Mitov2018}.
@@
 %Several GWAS have been performed to try to identify specific human and viral genetic variants that explain variation in spVL.
-The largest GWAS for human genetic factors so far, conducted by \citet{McLaren2015}, confirmed two regions of the human genome are associated with variability in spVL – .
+The largest GWAS for human genetic factors so far, conducted by \cite{McLaren2015}, confirmed two regions of the human genome are associated with variability in spVL – .
 This study used a standard GWAS framework, in which each surveyed genetic variant (millions in total) was seperately tested for a significant linear relationship with spVL based on thousands of genotyped and phenotyped individuals.
@@
 %Several approaches have been commonly used to account for population structure in GWAS, including adjusting the GWAS significance threshold to account for potential spurious findings \parencite{Devlin1999}, performing GWAS only on more homogenous population sub-groups \parencite{Pritchard2000}, and adjusting the regression model itself to include estimates of individual relatedness \parencite{Price2006, Kang2010}.
-Of these approaches, we highlight the linear mixed model developed by \citet{Kang2010} because it is an increasingly popular technique that has a direct analog in the field of phylogenetic comparative analysis, from which we derive our method to account for pathogen relatedness.
+Of these approaches, we highlight the linear mixed model developed by \cite{Kang2010} because it is an increasingly popular technique that has a direct analog in the field of phylogenetic comparative analysis, from which we derive our method to account for pathogen relatedness.
 %
@@
 Since spVL samples are not independent but are instead related according to the underlying viral phylogeny, the residuals may not be independent either.
-To account for non-independence among residuals, \citet{Grafen1989} suggested the PGLS method, in which the phylogeny is translated into a variance-covariance structure among samples.
+To account for non-independence among residuals, \cite{Grafen1989} suggested the PGLS method, in which the phylogeny is translated into a variance-covariance structure among samples.
 It is assumed that residual errors are not independent, but instead follow this phylogenetic variance-covariance structure.
@@
 However, in certain infectious disease  contexts the viral population can also confound GWA analyses.
-In the genome-to-genome (G2G) GWAS framework described by \citet{Bartha2013}, host variants are tested for pairwise associations with viral variants to determine sites of genetic conflict between the human host and HIV pathogen.
+In the genome-to-genome (G2G) GWAS framework described by \cite{Bartha2013}, host variants are tested for pairwise associations with viral variants to determine sites of genetic conflict between the human host and HIV pathogen.
 In this case, confounding may arise from population stratification in both the host and viral populations (Figure \ref{fig:gwas-covariates}b).
@@
 %Several methods have been developed to correct for confounding from the viral phylogeny in HIV G2G GWAS.
-One of two methods originally introduced by \citet{Bhattacharya2007} involves determining whether viral mutations are inferred to occur disproportionately along branches of the phylogeny where a particular host variant is present.
+One of two methods originally introduced by \cite{Bhattacharya2007} involves determining whether viral mutations are inferred to occur disproportionately along branches of the phylogeny where a particular host variant is present.
 The other method is a model-selection approach in which phylogenetic models with and without selection pressure from host variants are compared for goodness-of-fit \parencite{Bhattacharya2007, Carlson2008}.
-Lastly, the more recently proposed method by \citet{Naret2018} uses phylogenetic principle components to describe the major axes of variation in the viral population (an approach analogous to using principle components to account for host population structure).
-However, the method proposed by \citet{Naret2018} does not have a secondary benefit of the other two methods, which is to restore statistical independence to samples that are fundamentally related by a phylogeny.
+Lastly, the more recently proposed method by \cite{Naret2018} uses phylogenetic principle components to describe the major axes of variation in the viral population (an approach analogous to using principle components to account for host population structure).
+However, the method proposed by \cite{Naret2018} does not have a secondary benefit of the other two methods, which is to restore statistical independence to samples that are fundamentally related by a phylogeny.
 %
@@
 %
-%As \citet{Bhattacharya2007} recognized, HIV samples are related by descent according to the transmission tree.
+%As \cite{Bhattacharya2007} recognized, HIV samples are related by descent according to the transmission tree.
 Such relationships are often investigated by inferring a viral phylogeny from sequencing data.
@@
 This correlation can be quantified in terms of the heritability of spVL, or, the fraction of variance in spVL among hosts that can be explained by underlying viral genetic variation.
-\citet{Mitov2018} demonstrated that spVL is at least 20\% heritable, meaning that the virus has a strong effect on host spVL.
+\cite{Mitov2018} demonstrated that spVL is at least 20\% heritable, meaning that the virus has a strong effect on host spVL.
 Based on this correlation between the virus and spVL (Figure \ref{fig:gwas-covariates}c), we argue it is necessary to expand the correlation model typically assumed in standard GWAS.
@@
 In this method, the top few principle components of host genetic variation are included in the association model as proxy variables for demographic and geographic factors.
-This effectively changes the association test hypothesis from ``does the host variant explain a significant amount of variation in spVL?", to ``does the host variant explain a significant amount of variation in spVL \emph{beyond that which can be explained by demographic or geographic factors?}" Previous GWAS for host determinants of spVL conducted by \citet{Price2006, McLaren2015, Bartha2013, Fellay2007, Pereyra2010, Pelak2010, VanManen2011} have employed this method.
+This effectively changes the association test hypothesis from ``does the host variant explain a significant amount of variation in spVL?", to ``does the host variant explain a significant amount of variation in spVL \emph{beyond that which can be explained by demographic or geographic factors?}" Previous GWAS for host determinants of spVL conducted by \cite{Price2006, McLaren2015, Bartha2013, Fellay2007, Pereyra2010, Pelak2010, VanManen2011} have employed this method.
 Several studies have additionally used other methods to account for host population stratification, including structured association \parencite{Pritchard2000} and genomic control \parencite{Devlin1999}.
@@
 GWA testing is then conducted independently within each group before results are finally aggregated across groups.
-\citet{McLaren2015, Pereyra2010} both applied a form of structured association.
-So far only \citet{Dalmasso2008} have used genomic control to account for population stratification in spVL GWAS.
+\cite{McLaren2015, Pereyra2010} both applied a form of structured association.
+So far only \cite{Dalmasso2008} have used genomic control to account for population stratification in spVL GWAS.
 In genomic control, the significance threshold for association testing is uniformly adjusted for all variants by a factor which accounts for the degree of stratification present.
@@
 %	
-%	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
+%	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
 %\end{proof}
@@
 %	\begin{tabular}{llclccc} \toprule 
-%	& \multicolumn{2}{c}{\specialcell[l]{\citet{McLaren2015} study \\ N = 6,315}} & \multicolumn{3}{c}{\specialcell[l]{Current study \\ N = 1,370}} & \\ \midrule 
+%	& \multicolumn{2}{c}{\specialcell[l]{\cite{McLaren2015} study \\ N = 6,315}} & \multicolumn{3}{c}{\specialcell[l]{Current study \\ N = 1,370}} & \\ \midrule 
 %	Region & Variant & \emph{P} value & Variant & Effect size & \emph{P} value & R$^2$ \\ \midrule
@@
 %	\end{tabular}
-%	\caption{Association results for the top SNPs in the MHC and \emph{CCR5} regions from the GWAS by \citet{McLaren2015} compared to association results for the best proxy SNPs for these SNPs available in the current study.
+%	\caption{Association results for the top SNPs in the MHC and \emph{CCR5} regions from the GWAS by \cite{McLaren2015} compared to association results for the best proxy SNPs for these SNPs available in the current study.
 The proxy SNPs were chosen based on high linkage disequilibrium (as measured by R$^2$) with the McLaren et al.
@@
 %		\end{tabular}
-%		\caption{Top association results from \citet{McLaren2015} compared to association results for the best proxy SNPs available in the current study.
+%		\caption{Top association results from \cite{McLaren2015} compared to association results for the best proxy SNPs available in the current study.
 Proxy SNPs were chosen based on linkage disequilibrium (as measured by R$^2$) with the McLaren et al.

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
standardize gwas results plot, fix italic in simulation results plot

@@
         labs(x = expression("Selection strength (" * alpha * ")"),
-             y = expression("spVL heritability (" * 'H'^2 * ")")) +
+             y = expression("spVL heritability (" * italic('H'^2) * ")")) +
         shared_theme +
@@
 ```{r}
+pt_size <- 1
+
 # Add genome position to data
@@
 
-# This is a super hacky way to get labels in between x ticks
-# Hadley Wickam disapproves :/
+# This is maybe not the best way, but makes sure tick marks and labels are at the center of each chromosome
 max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 22, "BP_position"])
@@
 x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
-x_axis_ticks[x_axis_ticks$CHROM == 22, "midpoint"] <-
-        x_axis_ticks[x_axis_ticks$CHROM == 22, "CHR_start"] +
-                (max(gwas_results_annotated[gwas_results_annotated$CHROM == 22, "BP_position"]) -
-                        x_axis_ticks[x_axis_ticks$CHROM == 22, "CHR_start"]) / 2
-x_axis_ticks_tmp <- tidyr::pivot_longer(
-        data = x_axis_ticks,
-        cols = c("midpoint", "CHR_start"),
-        names_to = "marker",
-        values_to = "tick_position") %>%
-        mutate(CHROM = as.character(as.integer(CHROM)))
-x_axis_ticks_tmp[seq(from = 2, to = 46, by = 2), "CHROM"] <- ""
-x_axis_ticks_tmp[x_axis_ticks_tmp$CHROM == "23", "CHROM"] <- ""
-x_axis_ticks_tmp[x_axis_ticks_tmp$CHROM == "24", "CHROM"] <- ""
 
@@
 overview_comparison <- ggplot(
-        data = results_long,
+        data = results_long %>% sample_n(size = 10000),  ###
         aes(x = BP_position, y = -log10(P))) +
-        geom_point(aes(color = CHROM %% 2 == 0)) +
-        scale_color_manual(values = c("grey", "black")) +
+        geom_point(aes(color = CHROM %% 2 == 0), size = pt_size) +
+        scale_color_manual(values = c("grey", "dark grey")) +
         scale_x_continuous(
-                breaks = x_axis_ticks$tick_position,
+                breaks = x_axis_ticks$midpoint,
                 labels = x_axis_ticks$CHROM,
-                limits = c(0, NA),
-                expand = c(0, 0)) +
+                expand = c(0.01, 0.01),  # neccessary to fit chrom 23 label on
+                guide = guide_axis(n.dodge = 2)) +  # makes the x-axis labels alternate hi/lo
         scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
@@
                 yintercept = -log10(genome_wide_significance),
-                linetype = "dotted") +
-        theme_bw() +
+                linetype = "dashed") +
+        shared_theme +
         theme(
-                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
                 panel.grid.major.x = element_blank(),
-                axis.ticks.x = element_line(color = rep(c(NA, "black"), 23)),
                 panel.grid.minor.x = element_blank(),
                 legend.position = "none",
-                plot.margin = unit(c(5.5, 5.5, 25, 5.5), "pt"),
-                axis.text.x = element_text(size = 6)) +
+                axis.text.x = element_text(size = 5)
+        ) +
         facet_grid(. ~ gwas_type)
@@
 p_vals_regional <- ggplot(
-        data = regional_results,
+        data = regional_results  %>% sample_n(size = 100),  ###
         aes(x = POS/1E6, y = -log10(P_corrected))) +
-        geom_point(aes(color = -neg_log10_P_standard_minus_corrected)) +
+        geom_point(aes(color = -neg_log10_P_standard_minus_corrected), size = pt_size) +
         scale_color_gradient2(
@@
                 yintercept = -log10(genome_wide_significance),
-                linetype = "dotted") +
-        theme_bw() +
+                linetype = "dashed") +
+        shared_theme +
         theme(
-                panel.border = element_rect(colour = "black", size = 0.5, fill = NA),
                 panel.grid.major.x = element_blank(),
-                axis.ticks.x = element_line(color = rep(c(NA, "black"), 23)),
-                panel.grid.minor.x = element_blank(),
-                plot.margin = unit(c(25, 5.5, 5.5, 5.5), "pt")) +
+                panel.grid.minor.x = element_blank()) +
         facet_grid(. ~ CHROM, labeller = label_parsed, scales = "free")
@@
 # Assemble the plots as subplots
+aligned_plots <- align_plots(overview_comparison, p_vals_regional, align = "h", axis = "tb")
+
 png(
@@
         width = linewidth, height = height, unit = "cm", res = 300)
-gridExtra::grid.arrange(
-        overview_comparison,
-        p_vals_regional,
-        nrow = 2, heights = c(1.5, 2))
+plot_grid(aligned_plots[[1]], aligned_plots[[2]], nrow = 2, labels = c("A", "B"))
 dev.off()

### 2021-10-29 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 Finally, we applied our framework to empirical data from the Swiss HIV Cohort Study (SHCS).
-In the ``standard GWAS" we followed a typical GWAS methodology,  i.e.
-regressing measured spVL ($\bm{z}$) on variant dosage for each sampled host genetic variant.
-In the ``GWAS with phylogenetic correction", we use our proposed framework, i.e.
-regressing the estimated environmental part of spVL ($\bm{\hat{e}}$) on variant dosage.
-In agreement with previous GWAS for host genetic determinants of spVL, the top associations were in the MHC and \emph{CCR5} regions for both GWAS (figure \ref{fig:gwas-results}).
+We used data from 1,392 individuals in Switzerland infected with subtype B HIV-1 between 1994 and 2018.
+The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals.
+We followed the framework outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort individuals from these data.
+Figure \ref{fig:spvl-on-tree} shows the calculated
+
+We ran two sets of GWAS association tests on the same host genotype data to compare our framework with a more standard approach.
+In both GWAS, we regressed a trait value on dosage for each candidate host genetic variant to identify host variants correlated with trait values across the cohort.
+The two GWAS differ in the trait value used.
+In the ``GWAS with standard trait value'' we used the unmodified trait value, measured spVL.
+In the ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL.
+Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS.
+Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions.
+Regardless, these regions were consistent in having variants with the strongest signals of association from across the genome.
 
@@
 		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
-		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load.
+		\caption{Results from comparative GWAS.
+Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load.
 The pullout plots give regional association results from the GWAS with phylogenetic correction in the \emph{CCR5} region (positions 45.4 - 47Mb on chromosome 3) and the MHC region (positions 29.5 - 33.5Mb on chromosome 6).
@@
 \label{fig:poumm-parameters}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/spvl_calculation_comparison.png}
+	\caption{A comparison of different ways to calculate spVL based on viral load measurements provided by the SHCS.}
+	\label{fig:spvl-on-tree}
+	\end{center}
 \end{figure}

### 2021-10-29 scripts/R/make_manuscript_figures.rmd
standardize format for spvl comparison

@@
 gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
+spvl_values <- read.csv(file = paste(OUTDIR, "spvl_values.csv", sep = "/"))
 ```
@@
 ```{r}
-p <- ggplot(data = traits, aes(x = trait, y = h.MWA)) +
+p <- ggplot(data = spvl_values,
+            aes(x = spvl_lenient_filter_first_measurement,
+                y = spvl_lenient_filter_mean_measurement,
+                color = spvl_strict_filter_mean_measurement)) +
+        scale_color_gradient(low = "green", high = "red", name = "Mean measurement\nwith stricter filtering") +
+        coord_equal() +
+        geom_point() +
+        labs(x = "First measurement", y = "Mean measurement") +
+        shared_theme
+
+ggsave(plot = p, filename = paste(FIGDIR, "spvl_calculation_comparison.png", sep = "/"),
+       width = linewidth * 2/3, height = height * 2/3, units = "cm")
+
+p1 <- ggplot(data = traits, aes(x = trait, y = h.MWA)) +
       geom_point() +
@@
       labs(x = "Calculated spVL trait", y = ehat_name)
-p2 <- ggExtra::ggMarginal(p, type="histogram")
+p2 <- ggExtra::ggMarginal(p1, type="histogram")
 ggsave(plot = p2, filename = paste(FIGDIR, "trait_values.png", sep = "/"),

### 2021-10-30 manuscript/hiv-gwas.tex
Update on Overleaf.

--- 2021-10-29 manuscript/hiv-gwas.tex
@@
 \begin{center}
-	\Large{A new phylogeny-aware GWAS framework to estimate direct human genetic effects on HIV set-point viral load}
+	\Large{A phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits}
 \end{center}
@@
 \begin{abstract}
-	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because human genetic variants can not only directly effect disease outcome, but also have indirect effects via selection on the pathogen.
-Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL).
-However, it is unclear to what extent the identified alleles are directly protective against all HIV strains.
-HIV strain-specific factors account for an estimated 21 - 29\% of spVL variability, but typical GWAS methods do not consider these effects.
-We propose a new method to consider the full genome of each patient's infecting strain, remove strain-specific effects on spVL, and thus better estimate the direct effect of human alleles on HIV spVL.
-In simulations, we show our method can increase GWAS power to detect truly associated human variants when viral effects are highly heritable.
-When we apply our method to real patient data from Switzerland, we recover weaker signals of association in the \emph{CCR5} gene region and among the top-associated variants in the major histocompatability complex (MHC) compared to a standard GWAS.
-The top-associated SNPs do not change.
-This method could also be used investigate whether GWAS-detected associations in other chronic infections are robust to a correction for pathogen-mediated effects.
+	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits.
+Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits.
+However, GWAS for host genetic determinants of infectious disease traits do not account for potentially confounding or extraneous pathogen effects that are heritable from one infection partner to another in transmission chains.
+We propose a novel method to consider the full genome of each patient's infecting strain, remove strain-specific effects on a trait, and thus better estimate the effect of human genetic variants on infectious disease traits.
+In simulations, we show our method can increase GWAS power to detect truly associated human variants when pathogen effects are highly heritable.
+When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association in the \emph{CCR5} gene region and among the top-associated variants in the major histocompatability complex (MHC) compared to a standard GWAS.
+Although estimated heritability and selection parameters for HIV-1 mean standard GWAS are quite robust to pathogen effects, our simulation study shows that our framework may improve GWAS for other diseases where pathogen effects are stronger and more correlated among close members of transmission chains.
 \end{abstract}
@@
 % How do we measure the extent to which genetic factors determine spVL
-We are interested in understanding to what extent spVL, as a proxy for natural HIV control, is determined by human and viral genetic factors.
+Previous studies have underscored the importance of understanding to what extent spVL, as a proxy for natural HIV control, is determined by human genetic factors \cite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
 Heritability is a key measure of how genetically-determined a trait is.
@@
 GWAS can be used to measure narrow-sense heritability from human genetic variants.
-Several GWAS have been done for spVL \parencite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
+Several GWAS have been done for spVL \cite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
 The largest to-date by \cite{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
@@
 % Introduction to our method
-In this work,  we draw from the field of phylogenetic comparative methods to develop a new GWAS approach that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny.
-Our approach should help identify human genetic variants with direct effects on disease traits.
+In this work,  we draw from the field of phylogenetic comparative methods to develop a new GWAS framework that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny.
+Our approach should help identify human genetic variants that affect disease traits and more accurately estimate their effects.
 Furthermore, we think this approach represents the first GWAS framework to identify human genetic variants associated with a disease trait that accounts for non-independence of samples due to the underlying pathogen phylogeny.
@@
 Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Studey (SHCS) and show that only certain associations with spVL are robust to a correction for viral effects.
-We suggest that this framework could be applied to other infectious disease GWAS.
+Although we developed our framework in the context of HIV-1 spVL, the same approach could readily be applied to other heritable infectious disease traits.
 
@@
 Finally, we applied our framework to empirical data from the Swiss HIV Cohort Study (SHCS).
-In the ``standard GWAS" we followed a typical GWAS methodology,  i.e.
-regressing measured spVL ($\bm{z}$) on variant dosage for each sampled host genetic variant.
-In the ``GWAS with phylogenetic correction", we use our proposed framework, i.e.
-regressing the estimated environmental part of spVL ($\bm{\hat{e}}$) on variant dosage.
-In agreement with previous GWAS for host genetic determinants of spVL, the top associations were in the MHC and \emph{CCR5} regions for both GWAS (figure \ref{fig:gwas-results}).
-
+We used data collected from 1,392 individuals in Switzerland infected with subtype B HIV-1 between 1994 and 2018.
+The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals.
+We followed the framework outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
+Figure \ref{fig:spvl-calc-comparison} shows the calculated spVL values, figure \ref{fig:spvl-on-tree} shows the inferred pathogen phylogeny, and figures \ref{fig:spvl-on-tree}, \ref{fig:trait-comparison} show the estimated non-pathogen effects on spVL calculated for these individuals.
+We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) \ref{tab:POUMM-parameters-our-estimates}.
+To put these values into the context of our simulation study, they are shown as black points on figure \ref{fig:simulationResults}.
+
+We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes.
+In the ``GWAS with standard trait value'' we used the unmodified trait value, our calculated spVL values, as the GWAS response variable.
+In the ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL.
+Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS.
+Q-Q plots show the distribution of p-values are very similar as well (figure \ref{fig:qq-plots}).
+Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions.
+Taking into account phylogenetic information slightly decreased association strength for most variants in \emph{CCR5} region.
+Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance from a p-value of 2.2x10$^{-5}$ to 6x10$^{-7}$.
+% TODO
+However, the two top-associated SNPs are consistent between our study and \cite{McLaren2015} (table \ref{tab:comp-gwas}).
+In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
+
+% TODO
 \begin{figure}[H]
@@
 		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
-		\caption{Association results for standard GWAS and GWAS with a phylogenetic correction applied to the same data for HIV-1 set-point viral load.
-The pullout plots give regional association results from the GWAS with phylogenetic correction in the \emph{CCR5} region (positions 45.4 - 47Mb on chromosome 3) and the MHC region (positions 29.5 - 33.5Mb on chromosome 6).
+		\caption{Results from comparative GWAS.
+(A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables.
+On the left, we used unmodified spVL values.
+On the right, we used our estimates for the non-pathogen effects on spVL.
+The alternating colors correspond to different chromosomes.
+(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
 Base positions are with reference to genome build GRCh37.
-The color scale in the regional association plot is the difference in -log$_{10}$ p-value between the GWAS with phylogenetic correction and the standard GWAS.
-Red color means the phylogenetic correction decreased variant significance and blue color means the phylogenetic correction increased variant significance.
-The dashed lines show genome-wide significance (p-value = 5x10$^{-8}$).}
+The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
+Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
+The dashed lines show genome-wide significance at p = 5x10$^{-8}$).}
 		\label{fig:gwas-results}
@@
 \end{figure}
-
-The overall distribution of p-values did not change noticeably between the standard GWAS and the GWAS with a phylogenetic correction (figure \ref{fig:qq-plots}).
-In the \emph{CCR5} and MHC regions, most variants decreased in significance in the GWAS with phylogenetic correction.
-However, the top-associated variants were constant between analyses.
-rs112243036 was most significant in the MHC, with a standard GWAS p-value of 9.7x10$^{-16}$ and corrected GWAS p-value of 2.5x10$^{-14}$.
-In the \emph{CCR5} region, rs7430431 was most significant, with a standard GWAS p-value of 7.4x10$^{-9}$ and a corrected GWAS p-value of 1.7x10$^{-7}$.
-Some variants in the MHC did increase in significance with a phylogenetic correction.
-For example, SNP rs9265880 had the greatest increase in significance from a p-value of 2.2x10$^{-5}$ in the standard GWAS to 6x10$^{-7}$ in the GWAS with phylogenetic correction.
-
-The top-associated SNPs from the \cite{McLaren2015} paper were excluded from the main analysis because of too-high missing rates (3\% for rs1015164 and 4\% for rs59440261).
-In order to directly compare our results with \cite{McLaren2015}, we ran an additional GWAS for these two SNPs anyways.
-Both SNPs became less significant with phylogenetic correction (table \ref{tab:comp-gwas}).
-
-Again following the procedure of \cite{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
-Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).
 
@@
 
-\begin{table}[H]
-	\begin{center}
-	\begin{tabular} {p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}
-		\hline
-		 & \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
-		 \hline
-		Allele & Beta & p-value & Beta & p-value & Beta & p-value \\
-		\hline	
-		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
-		\rowcolor{Gray}
-		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
-		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
-		\rowcolor{Gray}
-		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
-		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
-		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
-		\rowcolor{Gray}
-		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
-		\rowcolor{Gray}
-		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
-		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
-		\rowcolor{Gray}
-		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
-		\rowcolor{Gray}
-		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
-		\rowcolor{Gray}
-		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\
-		\hline
-	\end{tabular}
-	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.
-The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al.
-study.
-The full table is available in the supplement (table \ref{tab:full-mclaren-comparison}).
-Grey rows are alleles that increased in significance with a phylogenetic correction.}
-	\label{tab:lim-mclaren-comparison}
-	\end{center}
-\end{table}
-
 \section{Discussion}
+% State the importance of your findings
+We present a novel, phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits.
+
+% Contextualize the study
+We estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the SHCS.
+Our estimate is at the high end of previous estimates based on the POUMM in European cohorts, but the uncertainty intervals largely overlap: \cite{Bertels2018} estimated 29\% (N = 2014, CI 12 - 46\%) also from the SHCS and \cite{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort.
+Our sample size (N = 1493 individuals) is quite a bit smaller than these other studies, possibly because we restrict samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases.
+Here our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty.
+
+% State some limitations
+Given our estimates for the heritability of spVL and the selection strength on this trait, our simulation results show that we cannot expect a significant improvement in GWAS power to identify host genetic determinants of this trait (figure \ref{fig:simulationResults}).
+So although there is a growing consensus that spVL is highly heritable (see \cite{Mitov2018} and citations therein), this does not necessarily mean individuals in a GWAS cohort have phylogenetically correlated spVL values.
+
+% Summarize the paper
+% Make recommendations or suggestions for future research
 \begin{itemize}
@@
 %	\item Example spVL plots
-%	\item simulated TPR plot annotated with actual inferred POUMM params
-%	\item Simulation results with different parameterizations
-%	\item Whole pipeline replicated with a stricter spVL definition (and correspondingly fewer samples)
-%	\item Comparative QQ-plots from 1500 and standard GWAS
-%	\item Proof of robustness of method to inferred tree
+%	\item Proof of robustness of method to inferred tree -- NA because I just re-ran the _whole_ thing and it's not that different
 %\end{itemize}
@@
 	\centering
-		\includegraphics[width=\linewidth]{figures/poumm_parameter_estimates.png}
+		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
 		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on Swiss data.
@@
 \label{fig:poumm-parameters}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = 0.462\linewidth]{figures/spvl_calculation_comparison.png}
+	\caption{A comparison of different ways to calculate spVL based on viral load measurements provided by the SHCS.
+The stricter filtering excludes all measurements possibly $<$ 6 months after infection and after treatment or AIDS, whereas the more lenient filtering excludes only measurements after treatment.
+We used the lenient filter, mean measurement values because these correlate well with the values from the stricter filter but allow us to retain many more individuals from the cohort for our study.}
+	\label{fig:spvl-calc-comparison}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = 0.462\linewidth]{figures/trait_values.png}
+	\caption{A comparison of measured (calculated) spVL values versus our estimated non-pathogen effect on spVL for each SHCS cohort member used in the study.
+The histograms show the marginal distribution of each value across the individuals.}
+	\label{fig:trait-comparison}
+	\end{center}
 \end{figure}
@@
 	\end{tabular}
-	\label{tab:POUMM-parameters}
+	\label{tab:POUMM-parameters-our-estimates}
 \end{table}
@@
 	\end{tabular}
-	\label{tab:POUMMparams}
+	\label{tab:POUMMparams-prev-estimated}
 \end{table}
@@
 %\end{table}
+
+% Again following the procedure of \cite{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
+Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).
+
+% \begin{table}[H]
+% 	\begin{center}
+% 	\begin{tabular} {p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}
+% 		\hline
+% 		 & \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
+% 		 \hline
+% 		Allele & Beta & p-value & Beta & p-value & Beta & p-value \\
+% 		\hline	
+% 		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
+% 		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
+% 		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
+% 		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
+% 		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
+% 		\rowcolor{Gray}
+% 		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\
+% 		\hline
+% 	\end{tabular}
+% 	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.
+The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al.
+study.
+The full table is available in the supplement (table \ref{tab:full-mclaren-comparison}).
+Grey rows are alleles that increased in significance with a phylogenetic correction.}
+% 	\label{tab:lim-mclaren-comparison}
+% 	\end{center}
+% \end{table}

### 2021-10-30 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 \usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
+\usepackage{array}
+\newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % for centered, width-specified columns
 
@@
 Taking into account phylogenetic information slightly decreased association strength for most variants in \emph{CCR5} region.
-Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance from a p-value of 2.2x10$^{-5}$ to 6x10$^{-7}$.
-% TODO
-However, the two top-associated SNPs are consistent between our study and \cite{McLaren2015} (table \ref{tab:comp-gwas}).
+Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
+However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used \ref{tab:gwas-top-associations}.
+For comparison, table \ref{tab:comp-gwas-mclaren-snps} shows our GWAS results for the two top-associated SNPs identified by \cite{McLaren2015}.
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
@@
 \begin{table}[H]
-	\begin{center}
-		\begin{tabular}{p{1cm}p{2cm}p{2.5cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr} {llclcccc}
+\centering
+\caption{Top association results from \cite{McLaren2015} compared to results from this study from GWAS with unmodified (standard) spVL values and with our estimates for the non-pathogen effects on spVL.}
+\begin{tabular}{p{1cm}p{1.8cm}p{2.4cm}p{1.7cm}p{1.8cm}p{1.7cm}p{1.8cm}} % {lrrrrrr} {llclcccc}
 \hline
-			& & McLaren et al.  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
+& & McLaren et al.  & \multicolumn{2}{l}{Standard trait value} & \multicolumn{2}{l}{Estimated non-pathogen part of trait} \\ 
   \hline
-			Region & Variant & p-value & Beta & p-value & Beta & p-value \\
+ Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
   \hline
-			MHC & rs59440261 & 2.0 x 10$^{-83}$ & -0.414 & 1.6x10$^{-11}$ & -0.229 & 1.2x10$^{-10}$ \\
-			\emph{CCR5} & rs1015164 & 1.5 x 10$^{-19}$ & 0.148 & 1.3x10$^{-6}$ & 0.073 & 3.2x10$^{-5}$ \\			
+    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
+    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
    \hline
 \end{tabular}
-		\caption{Top association results from \cite{McLaren2015} compared to association results in the current study.}
-		\label{tab:comp-gwas}
-	\end{center}
+\label{tab:comp-gwas-mclaren-snps}
 \end{table}
@@
 
-\begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
-		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.}
+% latex table generated in R 4.1.1 by xtable 1.8-4 package
+% Sat Oct 30 10:31:16 2021
+\begin{table}[ht]
+\caption{Effect size and p-values from top-associated variants in the $\emph{CCR5}$ and MCR regions from the two GWAS performed in our study.
+``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with estimated non-pathogen effects on spVL.}
+\begin{tabular}{lllP{2.2cm}P{2.2cm}P{1.6cm}P{2.2cm}}
   \hline
-		& \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS} & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
+    Region & Position & Variant & Standard effect size & Standard p-value & Corrected effect size & Corrected p-value \\ 
   \hline
-	 	Allele & Beta & p-value & Beta &p-value & Beta& p-value \\ 
+    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
+    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46540932 & rs10865942 & -0.16 & $8.4 \times 10^{-9}$ & -0.081 & $4 \times 10^{-7}$ \\ 
+    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
+    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
+    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
+    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
+    MHC & 31391401 & rs4418214 & -0.34 & $2.4 \times 10^{-14}$ & -0.18 & $2.5 \times 10^{-12}$ \\ 
+    MHC & 31400137 & rs138130755 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    MHC & 31400705 & rs138117378 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    MHC & 31402358 & rs148792134 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    MHC & 31409677 & rs140991764 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
+    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46556835 & rs6808142 & 0.15 & $8.3 \times 10^{-8}$ & 0.082 & $3.3 \times 10^{-7}$ \\ 
+    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
+    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
+    MHC & 31367874 & rs111281598 & -0.37 & $1.5 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
+    MHC & 31376266 & rs73400361 & -0.37 & $1.4 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
+    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
    \hline
-		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
-		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
-		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
-		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
-		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
-		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
-		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
-		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
-		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
-		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
-		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
-		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\ 
-		HLA-C*02:02 & -0.223 & $5.45 \times 10^{-8}$ & -0.089 & $1.31 \times 10^{-1}$ & -0.039 & $2.50 \times 10^{-1}$ \\ 
-		HLA-A*25:01 & -0.301 & $6.80 \times 10^{-8}$ & -0.090 & $9.93 \times 10^{-2}$ & -0.033 & $2.96 \times 10^{-1}$ \\ 
-		HLA-B*52:01 & -0.435 & $2.60 \times 10^{-7}$ & 0.238 & $7.63 \times 10^{-2}$ & 0.150 & $5.33 \times 10^{-2}$ \\ 
-		HLA-C*01:02 & -0.231 & $2.22 \times 10^{-6}$ & -0.198 & $5.07 \times 10^{-4}$ & -0.113 & $5.70 \times 10^{-4}$ \\ 
-		HLA-A*23:01 & 0.269 & $1.66 \times 10^{-5}$ & 0.020 & $7.42 \times 10^{-1}$ & 0.015 & $6.59 \times 10^{-1}$ \\ 
-		HLA-B*55:01 & 0.302 & $2.32 \times 10^{-5}$ & 0.685 & $2.48 \times 10^{-1}$ & 0.436 & $2.03 \times 10^{-1}$ \\ 
-		HLA-B*35:02 & 0.358 & $2.42 \times 10^{-5}$ &  &  &  &  \\ 
-		HLA-A*31:01 & -0.219 & $4.36 \times 10^{-5}$ & -0.052 & $4.50 \times 10^{-1}$ & -0.017 & $6.78 \times 10^{-1}$ \\ 
-		HLA-B*35:01 & 0.156 & $4.82 \times 10^{-5}$ & 0.167 & $2.13 \times 10^{-4}$ & 0.083 & $1.33 \times 10^{-3}$ \\ 
-		HLA-B*49:01 & 0.253 & $1.86 \times 10^{-4}$ & 2.173 & $9.79 \times 10^{-4}$ & 1.253 & $9.82 \times 10^{-4}$ \\ 
-		HLA-B*35:03 & 0.215 & $7.13 \times 10^{-4}$ & -0.241 & $8.00 \times 10^{-1}$ & 0.261 & $6.35 \times 10^{-1}$ \\ 
-		HLA-B*40:01 & 0.148 & $9.37 \times 10^{-4}$ & 0.519 & $1.06 \times 10^{-2}$ & 0.349 & $2.91 \times 10^{-3}$ \\ 
-		HLA-B*18:01 & 0.136 & $9.48 \times 10^{-4}$ & 0.208 & $1.73 \times 10^{-3}$ & 0.133 & $5.09 \times 10^{-4}$ \\ 
-		HLA-A*29:02 & 0.167 & $1.40 \times 10^{-3}$ & 0.150 & $6.17 \times 10^{-2}$ & 0.092 & $4.72 \times 10^{-2}$ \\ 
-		HLA-C*14:02 & -0.228 & $1.50 \times 10^{-3}$ & 0.266 & $7.59 \times 10^{-1}$ & 0.235 & $6.38 \times 10^{-1}$ \\ 
-		HLA-A*11:01 & -0.109 & $2.70 \times 10^{-3}$ & -0.090 & $8.96 \times 10^{-2}$ & -0.052 & $8.92 \times 10^{-2}$ \\ 
-		HLA-A*24:02 & 0.094 & $3.60 \times 10^{-3}$ & 0.133 & $3.79 \times 10^{-3}$ & 0.074 & $5.52 \times 10^{-3}$ \\ 
-		HLA-B*37:01 & 0.222 & $4.50 \times 10^{-3}$ & 0.600 & $3.22 \times 10^{-3}$ & 0.311 & $8.11 \times 10^{-3}$ \\ 
-		HLA-B*44:03 & 0.119 & $5.00 \times 10^{-3}$ & -0.114 & $2.77 \times 10^{-1}$ & -0.080 & $1.88 \times 10^{-1}$ \\ 
-		HLA-C*03:03 & 0.110 & $6.50 \times 10^{-3}$ & -0.078 & $1.86 \times 10^{-1}$ & -0.031 & $3.66 \times 10^{-1}$ \\ 
-		HLA-B*58:01 & -0.253 & $7.10 \times 10^{-3}$ & 6.895 & $7.93 \times 10^{-2}$ & 4.841 & $3.27 \times 10^{-2}$ \\ 
-		HLA-A*32:01 & -0.123 & $8.30 \times 10^{-3}$ & 0.089 & $5.31 \times 10^{-1}$ & 0.071 & $3.86 \times 10^{-1}$ \\ 
-		HLA-C*16:01 & 0.137 & $8.60 \times 10^{-3}$ & 0.057 & $4.46 \times 10^{-1}$ & 0.013 & $7.60 \times 10^{-1}$ \\ 
-		HLA-C*03:04 & 0.093 & $1.25 \times 10^{-2}$ & 0.126 & $3.52 \times 10^{-2}$ & 0.095 & $5.77 \times 10^{-3}$ \\ 
-		HLA-B*38:01 & -0.148 & $1.35 \times 10^{-2}$ & -3.459 & $3.01 \times 10^{-1}$ & -2.224 & $2.49 \times 10^{-1}$ \\ 
-		HLA-A*33:01 & -0.233 & $1.45 \times 10^{-2}$ &  &  &  &  \\ 
-		HLA-A*30:01 & -0.159 & $2.89 \times 10^{-2}$ & 0.072 & $9.04 \times 10^{-1}$ & 0.008 & $9.82 \times 10^{-1}$ \\ 
-		HLA-C*12:03 & -0.077 & $4.59 \times 10^{-2}$ & -0.114 & $2.95 \times 10^{-2}$ & -0.067 & $2.68 \times 10^{-2}$ \\ 
-		HLA-A*01:01 & 0.051 & $4.78 \times 10^{-2}$ & -0.039 & $4.88 \times 10^{-1}$ & -0.050 & $1.24 \times 10^{-1}$ \\ 
-		HLA-B*14:01 & -0.187 & $5.51 \times 10^{-2}$ & 7.071 & $5.00 \times 10^{-2}$ & 3.974 & $5.62 \times 10^{-2}$ \\ 
-		HLA-B*40:02 & -0.130 & $1.01 \times 10^{-1}$ & -0.343 & $3.88 \times 10^{-1}$ & -0.105 & $6.47 \times 10^{-1}$ \\ 
-		HLA-C*07:04 & 0.125 & $1.02 \times 10^{-1}$ & 0.111 & $3.00 \times 10^{-1}$ & 0.080 & $1.98 \times 10^{-1}$ \\ 
-		HLA-C*05:01 & 0.043 & $1.93 \times 10^{-1}$ & -0.022 & $6.51 \times 10^{-1}$ & -0.019 & $4.82 \times 10^{-1}$ \\ 
-		HLA-A*03:01 & 0.025 & $3.29 \times 10^{-1}$ & 0.002 & $9.51 \times 10^{-1}$ & -0.001 & $9.81 \times 10^{-1}$ \\ 
-		HLA-B*51:01 & -0.036 & $3.52 \times 10^{-1}$ & -0.046 & $3.13 \times 10^{-1}$ & -0.044 & $1.01 \times 10^{-1}$ \\ 
-		HLA-A*68:01 & 0.046 & $3.71 \times 10^{-1}$ & 0.036 & $6.09 \times 10^{-1}$ & 0.009 & $8.27 \times 10^{-1}$ \\ 
-		HLA-B*44:02 & 0.030 & $3.73 \times 10^{-1}$ & -0.051 & $3.90 \times 10^{-1}$ & -0.031 & $3.57 \times 10^{-1}$ \\ 
-		HLA-B*39:01 & 0.073 & $3.85 \times 10^{-1}$ &  &  &  &  \\ 
-		HLA-A*02:01 & 0.017 & $3.94 \times 10^{-1}$ & -0.018 & $5.57 \times 10^{-1}$ & -0.008 & $6.77 \times 10^{-1}$ \\ 
-		HLA-C*15:02 & -0.046 & $4.20 \times 10^{-1}$ & 0.064 & $3.26 \times 10^{-1}$ & 0.033 & $3.78 \times 10^{-1}$ \\ 
-		HLA-B*50:01 & 0.066 & $4.68 \times 10^{-1}$ & 1.522 & $1.77 \times 10^{-2}$ & 0.970 & $8.73 \times 10^{-3}$ \\ 
-		HLA-B*15:01 & -0.021 & $5.71 \times 10^{-1}$ & -0.073 & $1.71 \times 10^{-1}$ & -0.021 & $4.93 \times 10^{-1}$ \\ 
-		HLA-A*30:02 & 0.050 & $5.92 \times 10^{-1}$ & 0.047 & $9.28 \times 10^{-1}$ & -0.004 & $9.89 \times 10^{-1}$ \\ 
-		HLA-A*26:01 & -0.003 & $9.57 \times 10^{-1}$ & -0.645 & $1.96 \times 10^{-1}$ & -0.529 & $6.59 \times 10^{-2}$ \\
-		\hline
-		\label{tab:full-mclaren-comparison}
-\end{longtable}
+\end{tabular}
+\label{gwas-top-associations}
+\end{table}
+
+% \begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
+% 		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.}
+% 		\hline
+% 		& \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS} & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
+% 		\hline
+% 	 	Allele & Beta & p-value & Beta &p-value & Beta& p-value \\ 
+% 		\hline
+% 		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
+% 		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
+% 		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
+% 		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
+% 		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
+% 		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
+% 		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
+% 		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
+% 		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
+% 		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
+% 		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
+% 		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\ 
+% 		HLA-C*02:02 & -0.223 & $5.45 \times 10^{-8}$ & -0.089 & $1.31 \times 10^{-1}$ & -0.039 & $2.50 \times 10^{-1}$ \\ 
+% 		HLA-A*25:01 & -0.301 & $6.80 \times 10^{-8}$ & -0.090 & $9.93 \times 10^{-2}$ & -0.033 & $2.96 \times 10^{-1}$ \\ 
+% 		HLA-B*52:01 & -0.435 & $2.60 \times 10^{-7}$ & 0.238 & $7.63 \times 10^{-2}$ & 0.150 & $5.33 \times 10^{-2}$ \\ 
+% 		HLA-C*01:02 & -0.231 & $2.22 \times 10^{-6}$ & -0.198 & $5.07 \times 10^{-4}$ & -0.113 & $5.70 \times 10^{-4}$ \\ 
+% 		HLA-A*23:01 & 0.269 & $1.66 \times 10^{-5}$ & 0.020 & $7.42 \times 10^{-1}$ & 0.015 & $6.59 \times 10^{-1}$ \\ 
+% 		HLA-B*55:01 & 0.302 & $2.32 \times 10^{-5}$ & 0.685 & $2.48 \times 10^{-1}$ & 0.436 & $2.03 \times 10^{-1}$ \\ 
+% 		HLA-B*35:02 & 0.358 & $2.42 \times 10^{-5}$ &  &  &  &  \\ 
+% 		HLA-A*31:01 & -0.219 & $4.36 \times 10^{-5}$ & -0.052 & $4.50 \times 10^{-1}$ & -0.017 & $6.78 \times 10^{-1}$ \\ 
+% 		HLA-B*35:01 & 0.156 & $4.82 \times 10^{-5}$ & 0.167 & $2.13 \times 10^{-4}$ & 0.083 & $1.33 \times 10^{-3}$ \\ 
+% 		HLA-B*49:01 & 0.253 & $1.86 \times 10^{-4}$ & 2.173 & $9.79 \times 10^{-4}$ & 1.253 & $9.82 \times 10^{-4}$ \\ 
+% 		HLA-B*35:03 & 0.215 & $7.13 \times 10^{-4}$ & -0.241 & $8.00 \times 10^{-1}$ & 0.261 & $6.35 \times 10^{-1}$ \\ 
+% 		HLA-B*40:01 & 0.148 & $9.37 \times 10^{-4}$ & 0.519 & $1.06 \times 10^{-2}$ & 0.349 & $2.91 \times 10^{-3}$ \\ 
+% 		HLA-B*18:01 & 0.136 & $9.48 \times 10^{-4}$ & 0.208 & $1.73 \times 10^{-3}$ & 0.133 & $5.09 \times 10^{-4}$ \\ 
+% 		HLA-A*29:02 & 0.167 & $1.40 \times 10^{-3}$ & 0.150 & $6.17 \times 10^{-2}$ & 0.092 & $4.72 \times 10^{-2}$ \\ 
+% 		HLA-C*14:02 & -0.228 & $1.50 \times 10^{-3}$ & 0.266 & $7.59 \times 10^{-1}$ & 0.235 & $6.38 \times 10^{-1}$ \\ 
+% 		HLA-A*11:01 & -0.109 & $2.70 \times 10^{-3}$ & -0.090 & $8.96 \times 10^{-2}$ & -0.052 & $8.92 \times 10^{-2}$ \\ 
+% 		HLA-A*24:02 & 0.094 & $3.60 \times 10^{-3}$ & 0.133 & $3.79 \times 10^{-3}$ & 0.074 & $5.52 \times 10^{-3}$ \\ 
+% 		HLA-B*37:01 & 0.222 & $4.50 \times 10^{-3}$ & 0.600 & $3.22 \times 10^{-3}$ & 0.311 & $8.11 \times 10^{-3}$ \\ 
+% 		HLA-B*44:03 & 0.119 & $5.00 \times 10^{-3}$ & -0.114 & $2.77 \times 10^{-1}$ & -0.080 & $1.88 \times 10^{-1}$ \\ 
+% 		HLA-C*03:03 & 0.110 & $6.50 \times 10^{-3}$ & -0.078 & $1.86 \times 10^{-1}$ & -0.031 & $3.66 \times 10^{-1}$ \\ 
+% 		HLA-B*58:01 & -0.253 & $7.10 \times 10^{-3}$ & 6.895 & $7.93 \times 10^{-2}$ & 4.841 & $3.27 \times 10^{-2}$ \\ 
+% 		HLA-A*32:01 & -0.123 & $8.30 \times 10^{-3}$ & 0.089 & $5.31 \times 10^{-1}$ & 0.071 & $3.86 \times 10^{-1}$ \\ 
+% 		HLA-C*16:01 & 0.137 & $8.60 \times 10^{-3}$ & 0.057 & $4.46 \times 10^{-1}$ & 0.013 & $7.60 \times 10^{-1}$ \\ 
+% 		HLA-C*03:04 & 0.093 & $1.25 \times 10^{-2}$ & 0.126 & $3.52 \times 10^{-2}$ & 0.095 & $5.77 \times 10^{-3}$ \\ 
+% 		HLA-B*38:01 & -0.148 & $1.35 \times 10^{-2}$ & -3.459 & $3.01 \times 10^{-1}$ & -2.224 & $2.49 \times 10^{-1}$ \\ 
+% 		HLA-A*33:01 & -0.233 & $1.45 \times 10^{-2}$ &  &  &  &  \\ 
+% 		HLA-A*30:01 & -0.159 & $2.89 \times 10^{-2}$ & 0.072 & $9.04 \times 10^{-1}$ & 0.008 & $9.82 \times 10^{-1}$ \\ 
+% 		HLA-C*12:03 & -0.077 & $4.59 \times 10^{-2}$ & -0.114 & $2.95 \times 10^{-2}$ & -0.067 & $2.68 \times 10^{-2}$ \\ 
+% 		HLA-A*01:01 & 0.051 & $4.78 \times 10^{-2}$ & -0.039 & $4.88 \times 10^{-1}$ & -0.050 & $1.24 \times 10^{-1}$ \\ 
+% 		HLA-B*14:01 & -0.187 & $5.51 \times 10^{-2}$ & 7.071 & $5.00 \times 10^{-2}$ & 3.974 & $5.62 \times 10^{-2}$ \\ 
+% 		HLA-B*40:02 & -0.130 & $1.01 \times 10^{-1}$ & -0.343 & $3.88 \times 10^{-1}$ & -0.105 & $6.47 \times 10^{-1}$ \\ 
+% 		HLA-C*07:04 & 0.125 & $1.02 \times 10^{-1}$ & 0.111 & $3.00 \times 10^{-1}$ & 0.080 & $1.98 \times 10^{-1}$ \\ 
+% 		HLA-C*05:01 & 0.043 & $1.93 \times 10^{-1}$ & -0.022 & $6.51 \times 10^{-1}$ & -0.019 & $4.82 \times 10^{-1}$ \\ 
+% 		HLA-A*03:01 & 0.025 & $3.29 \times 10^{-1}$ & 0.002 & $9.51 \times 10^{-1}$ & -0.001 & $9.81 \times 10^{-1}$ \\ 
+% 		HLA-B*51:01 & -0.036 & $3.52 \times 10^{-1}$ & -0.046 & $3.13 \times 10^{-1}$ & -0.044 & $1.01 \times 10^{-1}$ \\ 
+% 		HLA-A*68:01 & 0.046 & $3.71 \times 10^{-1}$ & 0.036 & $6.09 \times 10^{-1}$ & 0.009 & $8.27 \times 10^{-1}$ \\ 
+% 		HLA-B*44:02 & 0.030 & $3.73 \times 10^{-1}$ & -0.051 & $3.90 \times 10^{-1}$ & -0.031 & $3.57 \times 10^{-1}$ \\ 
+% 		HLA-B*39:01 & 0.073 & $3.85 \times 10^{-1}$ &  &  &  &  \\ 
+% 		HLA-A*02:01 & 0.017 & $3.94 \times 10^{-1}$ & -0.018 & $5.57 \times 10^{-1}$ & -0.008 & $6.77 \times 10^{-1}$ \\ 
+% 		HLA-C*15:02 & -0.046 & $4.20 \times 10^{-1}$ & 0.064 & $3.26 \times 10^{-1}$ & 0.033 & $3.78 \times 10^{-1}$ \\ 
+% 		HLA-B*50:01 & 0.066 & $4.68 \times 10^{-1}$ & 1.522 & $1.77 \times 10^{-2}$ & 0.970 & $8.73 \times 10^{-3}$ \\ 
+% 		HLA-B*15:01 & -0.021 & $5.71 \times 10^{-1}$ & -0.073 & $1.71 \times 10^{-1}$ & -0.021 & $4.93 \times 10^{-1}$ \\ 
+% 		HLA-A*30:02 & 0.050 & $5.92 \times 10^{-1}$ & 0.047 & $9.28 \times 10^{-1}$ & -0.004 & $9.89 \times 10^{-1}$ \\ 
+% 		HLA-A*26:01 & -0.003 & $9.57 \times 10^{-1}$ & -0.645 & $1.96 \times 10^{-1}$ & -0.529 & $6.59 \times 10^{-2}$ \\
+% 		\hline
+% 		\label{tab:full-mclaren-comparison}
+% \end{longtable}
 

### 2021-10-30 scripts/R/make_manuscript_figures.rmd
remake all figures given more lenient geno filter, write out association results for text, tables

@@
 library(ggExtra)
+library(tidyr)
+library(xtable)
 
@@
 traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
-tree <- treeio::full_join(tree, traits, by = "label")
+tree_w_trait_data <- treeio::full_join(tree, traits, by = "label")
 POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
@@
 
-v_tree <- ggtree(tree, ladderize = F) +
+v_tree <- ggtree(tree_w_trait_data, ladderize = F) +
         geom_tippoint(aes(color = v.MWA)) +
@@
 
-h_tree <- ggtree(tree, ladderize = F) +
+h_tree <- ggtree(tree_w_trait_data, ladderize = F) +
         geom_tippoint(aes(color = h.MWA)) +
@@
 
-spvl_tree_same_scale <- ggtree(tree, ladderize = F) +
+spvl_tree_same_scale <- ggtree(tree_w_trait_data, ladderize = F) +
         geom_tippoint(aes(color = trait)) +
@@
 overview_comparison <- ggplot(
-        data = results_long %>% sample_n(size = 10000),  ###
+        data = results_long,
         aes(x = BP_position, y = -log10(P))) +
@@
 # Make plot showing p-value changes in the CCR5 and MHC regions
-regional_results<- gwas_results_annotated %>% filter(
+regional_results <- gwas_results_annotated %>% filter(
         (CHROM == 3 & POS >= 45.5E6 & POS <= 47E6) | (CHROM == 6 & POS >= 29.5E6 & POS <= 33.5E6))
@@
 p_vals_regional <- ggplot(
-        data = regional_results  %>% sample_n(size = 100),  ###
+        data = regional_results,
         aes(x = POS/1E6, y = -log10(P_corrected))) +
@@
 ```
+
+Get the most highly-associated and most changed SNPs from the GWAS results
+```{r}
+# Most highly associated
+most_highly_associated_mhc_ccr5_standard <- regional_results %>% group_by(CHROM) %>% top_n(n = 5, wt = -P_standard)
+most_highly_associated_mhc_ccr5_corrected <- regional_results %>% group_by(CHROM) %>% top_n(n = 5, wt = -P_corrected)
+
+most_highly_associated_mhc_ccr5 <- rbind(
+        most_highly_associated_mhc_ccr5_standard %>% mutate(type = "strongest association standard GWAS"),
+        most_highly_associated_mhc_ccr5_corrected %>% mutate(type = "strongest association corrected GWAS"))
+
+tex_table_1 <- most_highly_associated_mhc_ccr5 %>%
+        ungroup() %>%
+        mutate(region = case_when(
+                CHROM == "italic(CCR5)*' region'" ~ "\\emph{CCR5}",
+                T ~ "MHC")) %>%
+        select(region, POS, ID, BETA_standard, P_standard, BETA_corrected, P_corrected)
+
+print(xtable(tex_table_1, type = "latex", display=c("s", "s", "s", "s", "g","g","g","g")),
+      file = paste(FIGDIR, "strongest_associations.tex", sep = "/"), math.style.exponents = TRUE)
+
+write.csv(x = most_highly_associated_mhc_ccr5,
+          file = paste(OUTDIR, "strongest_associations.csv", sep = "/"), row.names = F)
+
+# Most changed: higher significance
+most_changed_positive_mhc_ccr5 <- regional_results %>%
+        group_by(CHROM) %>%
+        top_n(n = 1, wt = -neg_log10_P_standard_minus_corrected)
+
+write.csv(x = most_changed_positive_mhc_ccr5,
+          file = paste(OUTDIR, "most_increased_association_mhc_ccr5.csv", sep = "/"), row.names = F)
+
+# Comparison to top SNPs from McLaren et al.
+mclaren_top_snps <- gwas_results %>% filter(ID %in% c('rs59440261', 'rs1015164'))
+tex_table <- mclaren_top_snps %>% select(ID, BETA_standard, P_standard, BETA_corrected, P_corrected)
+print(xtable(tex_table, type = "latex", display=c("s", "s", "g","g","g","g")),
+      file = paste(FIGDIR, "mclaren_top_snp_comparison.tex", sep = "/"), math.style.exponents = TRUE)
+```

### 2021-10-30 scripts/bash/prep-gwas-files.sh

@@ -12,7 +12,7 @@ FAM_W_SEX="spVL.shcs.rs_with_sex_SN.fam"
 --keep output/gwas_individuals_to_keep.txt \
 --king-cutoff 0.09375 \
 --maf 0.01 \
---geno 0.01 \
+--geno 0.05 \
 --hwe 0.00005 \
 --make-bed \
 --memory 10000 \

### 2021-10-30 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 
-\usepackage[backend=bibtex,style=nature]{biblatex} % Borrowing citation formatting from Nature
-\bibliography{references}
+\usepackage{url} % otherwise bibtex hates my references file
+\usepackage[round]{natbib}
+\bibliographystyle{plainnat}
 
@@
 We propose a novel method to consider the full genome of each patient's infecting strain, remove strain-specific effects on a trait, and thus better estimate the effect of human genetic variants on infectious disease traits.
-In simulations, we show our method can increase GWAS power to detect truly associated human variants when pathogen effects are highly heritable.
-When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association in the \emph{CCR5} gene region and among the top-associated variants in the major histocompatability complex (MHC) compared to a standard GWAS.
-Although estimated heritability and selection parameters for HIV-1 mean standard GWAS are quite robust to pathogen effects, our simulation study shows that our framework may improve GWAS for other diseases where pathogen effects are stronger and more correlated among close members of transmission chains.
+In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable.
+When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association with spVL in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to a standard GWAS.
+Our simulation study confirms that estimated heritability and selection parameters for HIV-1 mean standard GWAS are quite robust to pathogen effects.
+Our framework may improve GWAS for other diseases where pathogen effects are stronger and more correlated amongst individuals in a cohort.
 \end{abstract}
@@
 Ideally, this procedure identifies variants that are located in or are in linkage disequilibrium with alleles that directly effect the trait.
-If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \parencite{Okada2014}.
-Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify high-risk individuals \parencite{Dudbridge2013}.
+If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \citep{Okada2014}.
+Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify high-risk individuals \citep{Dudbridge2013}.
 
@@
 For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with severity of disease course.
-spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \parencite{Mellors1996}.
-In untreated individuals, spVL is predictive of duration of asymptomatic infection \parencite{Mellors1996} and infectiousness \parencite{Quinn2000}.
-If viral load can be reduced to undetectable levels, an individual is effectively uninfectious \parencite{HHSARTGuidelines2019}.
-Notably, spVL varies by orders of magnitude between individuals \parencite{Mellors1996}.
-Thus, spVL measurements point to a wide range in natural HIV control by the human immune system.
+spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \citep{Mellors1996}.
+In untreated individuals, spVL is predictive of duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}.
+If viral load can be reduced to undetectable levels, an individual is effectively uninfectious \citep{HHSARTGuidelines2019}.
+Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}.
+Thus, spVL measurements point to a wide range in natural HIV control amongst individuals.
 
 % How do we measure the extent to which genetic factors determine spVL
-Previous studies have underscored the importance of understanding to what extent spVL, as a proxy for natural HIV control, is determined by human genetic factors \cite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
+To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
 Heritability is a key measure of how genetically-determined a trait is.
+Here we distinguish between two different heritability measures that are used in different contexts in infectious disease study.
 Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e.
 due to inherited differences.
-Narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e.
+In the infectious disease case, broad-sense heritability from pathogen factors, which are inherited by recipients from their infection partners, is typically measured.
+On the other hand, narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e.
 the sum of independent effects from all genetic variants.
+GWAS for infectious disease typically measure narrow-sense heritability of a trait based on human genetic variants.
 
 % How heritable is spVL (human side)
-GWAS can be used to measure narrow-sense heritability from human genetic variants.
-Several GWAS have been done for spVL \cite{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
-The largest to-date by \cite{McLaren2015} measured the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
+Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
+The largest to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
 All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}).
@@
 In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed.
-On the other hand, CCR5 proteins are a signal to the virus that help it find the right cells to infect.
+CCR5 proteins are a signal to the virus that help it find the right cells to infect.
 
@@
 In addition to these human genetic factors, is also well-recognized that viral genetic factors effect spVL.
-On the viral side, broad-sense heritability is typically measured.
-Estimates differ depending on the methods employed and the cohort studied (see \citealt{Mitov2018} for a discussion of this uncertainty).
-Estimates using phylogenetic methods on large UK and Swiss cohorts by \cite{Mitov2018} and \cite{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
-However, variation in the MHC is known to exert strong selective pressure on the virus \parencite{Kloverpris2016}.
+As mentioned, heritability from the viral side is typically measured using broad-sense heritability.
+Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty).
+Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
+However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016}.
 If the virus can change it's ``puzzle piece'' shape to escape MHC proteins, infected cells can go undetected.
-This means that MHC variants effect spVL largely via selection on the virus \parencite{Bartha2017}.
+This means that MHC variants effect spVL largely via selection on the virus \citep{Bartha2017}.
 In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.
@@
 % The gap
-Most of the GWAS for human genetic determinants of spVL conducted so far \parencite{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL.
+Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL.
 In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors.
-This has several potential negative consequences: (i) variability due to viral effects would make it more challenging to identify human variants of small effect, (ii) we can't distinguish between human variants with direct effects versus those with effects mediated by the virus via selection, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
+This has several potential negative consequences: (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
 
@@
 Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry in a cohort can give rise to spurious genetic correlations with a trait.
-Corrections for these correlations are well-developed and widely accepted  \parencite{Astle2009, Price2006}.
-More recently, \cite{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS.
-Two state-of-the-art methods exist for this \cite{Earle2016, collins_phylogenetic_nodate}.
-% \cite{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
-\cite{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry.
+Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}.
+More recently, \citep{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS.
+Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}.
+% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
+\citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry.
 However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait.
 In contrast, we want to estimate effects of $human$ genetic variants on a trait, taking into account pathogen effects.
-\parencite{Naret2018} developed a relevant method for this task.
+\citep{Naret2018} developed a relevant method for this task.
 The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing.
@@
 In the following we (i) describe a statistical model for the spVL trait, (ii) derive a maximum likelihood estimate for the viral part of spVL under this model, and (iii) describe a new infectious disease GWAS framework using this information.
-In simulations, we demonstrate that this framework can improve GWAS power to detect human genetic variants effecting a disease trait.
-Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Studey (SHCS) and show that only certain associations with spVL are robust to a correction for viral effects.
+In simulations, we demonstrate that this framework can improve GWAS power to detect host genetic variants effecting a disease trait.
+Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects.
 Although we developed our framework in the context of HIV-1 spVL, the same approach could readily be applied to other heritable infectious disease traits.
@@
 
-Variation in spVL comes from several sources: direct host genetic effects, pathogen genetic effects, interaction effects between host and pathogen genetic factors, and other environmental effects.
-Of these, only pathogen genetic effects are heritable from one transmission partner to another \parencite{Leventhal2016}.
-To characterize these effects, we use a phylogenetic mixed model (PMM) \cite{Housworth2004}.
+Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects.
+Of these, only pathogen effects are heritable from one transmission partner to another \citep{Leventhal2016}.
+To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}.
 PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts.
-In our case, pathogen genetic effects comprise the heritable part and all other effects comprise the non-heritable part.
+In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part.
 Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny.
@@
 extreme spVL values are unlikely.
-\cite{Mitov2018} and \cite{Bertels2018} previously showed the OU process has higher statistical support for spVL.
-This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \cite{Fraser2014}.
+\citep{Mitov2018} and \citep{Bertels2018} previously showed the OU process has higher statistical support for spVL.
+This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \citep{Fraser2014}.
 Therefore, we assume the OU process.
-The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \cite{Mitov2018}.
+The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \citet{Mitov2018}.
 Here, we review the main relevant points in the spVL context.
 
-Under the POUMM, the spVL trait $z$ is the sum of viral genetic efects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$.
+Under the POUMM, the spVL trait $z$ is the sum of viral effects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$.
 We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$:
@@
 The OU process is defined by a stochastic differential equation with two terms.
-The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \cite{Butler2004}:
+The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \citep{Butler2004}:
 \begin{align}
@@
 \end{equation}
-and the variance in $g_v(t)$ if we were to repeat the random evolutionary process many times \parencite{Butler2004}:
+and the variance in $g_v(t)$ if we were to repeat the random evolutionary process many times \citep{Butler2004}:
 \begin{equation}
@@
 $g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny.
-The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \parencite{Butler2004}:
+The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \citep{Butler2004}:
 \begin{equation}
@@
 
-	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
+	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
 \end{proof}
@@
 In other words, $g_v$ will be closer to measured spVL if spVL is not very heritable.
-If spVl is highly heritable, $g_v$ will be closer to the expected value under the POUMM, i.e.
+If spVL is highly heritable, $g_v$ will be closer to the expected value under the POUMM, i.e.
 take more information from the phylogenetic relationships between infecting strains.
@@
 
-We propose to improve standard GWAS for infectious diseases by estimating and removing trait variability due to pathogen genetic effects. Our new framework is as follows:
+We propose to improve standard GWAS for infectious diseases by estimating and removing trait variability due to pathogen effects. Our new framework is as follows:
 
@@
 	\item Construct a pathogen phylogeny using the pathogen genome sequences.
-	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with e.g. the R package POUMM \cite{Mitov2017a-POUMM}.
+	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with e.g. the R package POUMM \citep{Mitov2017a-POUMM}.
 	\item Generate maximum-likelihood estimates for the pathogen and corresponding non-pathogen effects on the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
@@
 First, we evaluated how well our method estimated the additive host genetic effects from the simulated data.
-Additive host genetic effects represent an ideal baseline for GWAS that seek to identify host genetic determinants of a trait.
-Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can generate more accurate estimates than the trait value alone.
-To be fair, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
+Additive host genetic effects represent an ideal (albeit unattainable) baseline for infectious disease GWAS.
+Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate this value compared to the trait value.
+To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
 In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation set-up.
@@
 We used the true positive rate (TPR) to evaluate the fraction of simulated host genetic variants with direct effects we could recover as being significantly associated with the trait.
-We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the unmodified trait value.
+We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the scaled trait value.
 Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%.
@@
 (A) shows that our method (left) can generate more accurate estimates of host genetic effects than the trait value, scaled by its mean (right).
-(B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the unmodified trait value (right).
+(B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
 Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
@@
 Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
-However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used \ref{tab:gwas-top-associations}.
-For comparison, table \ref{tab:comp-gwas-mclaren-snps} shows our GWAS results for the two top-associated SNPs identified by \cite{McLaren2015}.
+However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (figure \ref{tab:gwas-top-associations}).
+Finally table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}.
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
@@
 On the right, we used our estimates for the non-pathogen effects on spVL.
-The alternating colors correspond to different chromosomes.
+The alternating shades correspond to different chromosomes.
 (B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
@@
 Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
-The dashed lines show genome-wide significance at p = 5x10$^{-8}$).}
+The dashed lines show genome-wide significance at p = 5x10$^{-8}$.}
 		\label{fig:gwas-results}
@@
 \centering
-\caption{Top association results from \cite{McLaren2015} compared to results from this study from GWAS with unmodified (standard) spVL values and with our estimates for the non-pathogen effects on spVL.}
+\caption{Top association results from \citet{McLaren2015} compared to results from this study.
+The ``Standard trait value'' and ``Estimated non-pathogen part of trait'' results are from using two different GWAS response variables with the same human genotype data.}
 \begin{tabular}{p{1cm}p{1.8cm}p{2.4cm}p{1.7cm}p{1.8cm}p{1.7cm}p{1.8cm}} % {lrrrrrr} {llclcccc}
@@
 % State the importance of your findings
-We present a novel, phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits.
-
-% Contextualize the study
-We estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the SHCS.
-Our estimate is at the high end of previous estimates based on the POUMM in European cohorts, but the uncertainty intervals largely overlap: \cite{Bertels2018} estimated 29\% (N = 2014, CI 12 - 46\%) also from the SHCS and \cite{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort.
-Our sample size (N = 1493 individuals) is quite a bit smaller than these other studies, possibly because we restrict samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases.
-Here our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty.
-
-% State some limitations
-Given our estimates for the heritability of spVL and the selection strength on this trait, our simulation results show that we cannot expect a significant improvement in GWAS power to identify host genetic determinants of this trait (figure \ref{fig:simulationResults}).
-So although there is a growing consensus that spVL is highly heritable (see \cite{Mitov2018} and citations therein), this does not necessarily mean individuals in a GWAS cohort have phylogenetically correlated spVL values.
+In this paper, we presented a novel, phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits.
+By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait.
+This should help us better understand which host factors are protective against a disease versus which increase susceptibility or disease severity.
+
+% Contextualize the study & address limitations
+Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection.
+Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study.
+Compared to previous studies, this estimate is at the higher end \citep{Mitov2018}.
+Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort.
+We note that our sample size (N = 1493 individuals) is quite a bit smaller than these other studies.
+This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases.
+Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods.
+So although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.
+
+Instead, the main novelty of our approach was to correct the spVL trait prior to GWAS by estimating and removing pathogen effects.
+In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
+Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.
+Unfortunately, given our estimates for the heritability of spVL and the selection strength on this trait, our simulation results show that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (figure \ref{fig:simulationResults}).
+Our method is anticipated to be more useful for disease traits that are highly heritable from infection partner to infection partner and  maintain a high correlation between sampled individuals.
+In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
+So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to be meaningful.
+
+Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests.
+Therefore, it is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}).
+The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project github at TODO.
 
 % Summarize the paper
-% Make recommendations or suggestions for future research
-\begin{itemize}
-	\item From the field of phylogenetic comparative methods there are several methods to account for non-independence among samples related by an underlying phylogeny (e.g.
-independent contrasts, phylogenetic generalized least squares).
-Compared to these methods, our method is nice for GWAS because we do not change the association model itself, instead we just adjust the trait values prior to association testing.
-\end{itemize}
+In summary, we argue that infectious disease GWAS should take the pathogen phylogeny into account when searching for host determinants of a disease trait.
+We give a practical threshold for identifying when GWAS suffer from pathogen effects (heritability of the trait amongst infection partners $>$ 25\%) and provide a method that can help in scenarios where trait values are highly heritable and phylogenetically-structured amongst members of a cohort.
 
 %\begin{itemize}
-%	\item Compare inferred POUMM parameters to previous studies \parencite{Mitov2018, Bertels2018}
+%	\item Compare inferred POUMM parameters to previous studies \citep{Mitov2018, Bertels2018}
 %	\item Standard GWAS highlights previously identified regions associated with spVL: $CCR5$ and MHC
@@
 %	\item The phylogenetic correction to GWAS leads to the biggest p-value changes in the $CCR5$ and MHC regions
-%	\item SNPs near the FUT1 gene (on chromosome 6, highlighted as one of the top-20 p-value changes in the first pipeline run) increased in significance, and the Lewis-X antigen produced by this gene is known to play a role in early HIV infection \cite{Naarding2005}.
+%	\item SNPs near the FUT1 gene (on chromosome 6, highlighted as one of the top-20 p-value changes in the first pipeline run) increased in significance, and the Lewis-X antigen produced by this gene is known to play a role in early HIV infection \citep{Naarding2005}.
 %	\item Address method caveats: 
@@
 Whenever possible, we tried to parameterize our simulation model for spVL using empirical data.
-We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \parencite{Mitov2018}.
-After allotting 25\% of this variance to the host part of spVL $g_h$ (following \citealt{McLaren2015}), we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
+We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}.
+After allotting 25\% of this variance to the host part of spVL $g_h$ (following \citet{McLaren2015}), we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
 $g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
 Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see table \ref{tab:sim-params}).
-We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \parencite{Hodcroft2014} using the R package ape \parencite{Paradis2018}.
-$g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \parencite{Mitov2017a-POUMM} and sampling values at the tips.
-For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (table \ref{tab:POUMMparams}).
+We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}.
+$g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
+For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (table \ref{tab:POUMM-parameters-our-estimates}).
 To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively.
@@
 
-Genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, \citealt{Schoeni-Affolter2010}).
-
-We retained data from 1,396 individuals with non-recombinant subtype B $pol$ gene sequences of at least 750 characters, paired RNA measurements allowing for calculation of spVL, European ancestry, and not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}).
-Viral subtype was determined by the SHCS using the REGA HIV subtyping tool version 2.0 \parencite{DeOliveira2005}.
-We calculated spVL as the arithmetic mean of viral RNA measurements made prior to the start of antiretroviral treatment.
-Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}).
-Kinship was evaluated using PLINK version 2.00a2LM \parencite{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375.
-
-For phylogenetic inference, we included the 1,396 subtype B sequences and 5 randomly chosen subtype A sequences from the SHCS as an outgroup.
-We used MUSCLE version 3.8.31 \parencite{edgar_muscle:_2004} to align the $pol$ sequences with --maxiters 3 and otherwise default settings.
+Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, www.shcs.ch).
+All participants were HIV-1–infected individuals at least 16 years old and written informed consent was obtained from all cohort participants by \citet{Schoeni-Affolter2010}.
+The data was received in anonymized form after the project was approved by the SHCS.
+
+For phylogenetic inference, we retained sequences from 1,493 individuals with non-recombinant subtype B $pol$ gene sequences of at least 750 characters and paired RNA measurements allowing for calculation of spVL, as well as 5 randomly chosen subtype A sequences as an outgroup.
+We used MUSCLE version 3.8.31 \citep{edgar_muscle:_2004} to align the $pol$ sequences with --maxiters 3 and otherwise default settings.
 We trimmed the alignment to 1505 characters to standardize sequence lengths.
@@
 After rooting the tree based on the subtype A samples, we removed the outgroup.
-
+Viral subtype was determined by the SHCS using the REGA HIV subtyping tool version 2.0 \citep{DeOliveira2005}.
+We calculated spVL as the arithmetic mean of viral RNA measurements made prior to the start of antiretroviral treatment.
+
+For GWAS, we retained data from 1,392 of the 1,493 individuals with European ancestry and not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}).
+These were 227 females and 1165 males.
+Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}).
+Kinship was evaluated using PLINK version 2.00a2LM \citep{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375.
 Initial host genotyping quality control and imputation were done by the SHCS.
-Subsequent genotyping quality control was performed using PLINK version 2.3 \parencite{Chang2015}.
-We used the options --maf 0.01, --geno 0.01, and --hwe 0.00005 to remove variants with minor allele frequency less than 0.01, missing call rate greater than 0.01, or Hardy-Weinberg equilibrium exact test p-value less than 5x10$^-5$.
-After quality filtering, approximately XXX million genetic variants from XXX individuals were retained for GWAS (Table \ref{tab:variant-filtering}).
+Subsequent genotyping quality control was performed using PLINK version 2.3 \citep{Chang2015}.
+We used the options --maf 0.01, --geno 0.01, and --hwe 0.00005 to remove variants with minor allele frequency less than 0.01, missing call rate greater than 0.05, or Hardy-Weinberg equilibrium exact test p-value less than 5x10$^-5$.
+After quality filtering, approximately 6.2 million genetic variants from the 1,392 individuals were retained for GWAS (Table \ref{tab:variant-filtering}).
 
@@
 
-We used the R package POUMM version 2.1.6 \parencite{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the approximate maximum-likelihood phylogeny and calculated spVL values.
+We used the R package POUMM version 2.1.6 \citep{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the approximate maximum-likelihood phylogeny and calculated spVL values.
 The Bayesian inference method implemented in this package requires specification of a prior distribution for each parameter.
-We used the same, broad prior distributions as in \cite{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$.
+We used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$.
 We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000.
 The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution.
-Inferred parameters, posterior distributions for inferred parameters, and MCMC trace plots are given in table \ref{tab:POUMM-parameters} and figure \ref{fig:poumm-fit}, respectively.
+Figure \ref{fig:poumm-parameters} shows the posterior distributions for inferred parameters.
+Table \ref{tab:POUMM-parameters-our-estimates} gives the posterior mean values used for subsequent calculations.
 
@@
 We corrected calcualted spVL values using the method described in this paper.
-For each of the 1,396 individuals in our cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}.
-For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the maximum posterior probability estimates generated as described above.
+For each of the 1,392 individuals in the GWAS cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}.
+For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.
 
 \subsection{Association testing}
-We performed two GWAS using the same genotype data.
-For the first ``standard GWAS" we used total calculated spVL ($\bm{z}$) as the dependent variable for association testing, replicating prior GWAS for host genetic determinants of spVL.
-For the second ``GWAS with phylogenetic correction" we replaced total spVL with the estimated non-viral component of spVL ($\hat{\bm{e}}$) as the dependent variable.
-Association testing was performed using a linear association model in PLINK version 2.3 \parencite{Chang2015} with sex and the top 5 principle components of host genetic variation included as covariates.
+We performed two GWAS using the same human genotype data from the SHCS.
+For the first ``GWAS with standard trait value'' we used total calculated spVL ($\bm{z}$) as the response variable for association testing, replicating prior GWAS for host genetic determinants of spVL.
+For the second ``GWAS with estimated non-pathogen part of trait'' we replaced total spVL with the estimated non-viral component of spVL ($\hat{\bm{e}}$) as the response variable.
+Association testing was performed using a linear association model in PLINK version 2.3 \citep{Chang2015} with sex and the top 5 principle components of host genetic variation included as covariates.
 The sex and principle components covariates were included to reduce residual variance in spVL and control for confounding from host population structure, respectively.
-Common HLA allele genotypes were imputed using the SNP2HLA pipeline \parencite{Jia2013a} with a reference panel of 124 individuals from the HapMap CEU population.
-Association tests for imputed HLA alleles were performed in the same way as for other variants.
 
@@
 
-\begin{itemize}
-	\item Thanks to Michael Landis, who shared a LaTeX template for graphical model drawing.
-	\item Esposito Mauro and Alexandra Scherrer from the SHCS data center were especially helpful in retrieving data.
-\end{itemize}
-
-\printbibliography
+This study was funded by ETH Zurich.
+We would like to thank Esposito Mauro and Alexandra Scherrer from the SHCS data center, who were especially helpful in retrieving data, and Michael Landis, who shared a LaTeX template for graphical model drawing.
+Finally and most importantly, we thank the participants of the Swiss HIV Cohort Study for contributing their samples and making this research possible.
+
+\bibliography{manuscript/references_synced}
 
@@
 Variables in solid squares are constant in the simulation scheme, whereas variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
-Arrows represent dependencies among variables and the dashed square represents repetition \parencite{Hohna2014}.}
+Arrows represent dependencies among variables and the dashed square represents repetition \citep{Hohna2014}.}
 	\label{fig:sim-design}
@@
 
-\end{doublespace}  % don't doublspace tables
+\end{doublespace}  % don't doublespace tables
 
@@
 	\hline 
-	$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citealt{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
-	$\theta$ & 4.45 & (4.41 - 4.49) HPD & \citealt{Mitov2018} & \\
-	$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citealt{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
-	$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citealt{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
-	$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citealt{Mitov2018} & \\
-	$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citealt{Bertels2018} & \\
-	$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citealt{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
-	$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citealt{Mitov2018} & \\
-	$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citealt{Blanquart2017} & \\
-	$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citealt{Mitov2018} & \\
-	$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citealt{Blanquart2017} & \\
+	$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
+	$\theta$ & 4.45 & (4.41 - 4.49) HPD & \citet{Mitov2018} & \\
+	$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
+	$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
+	$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citet{Mitov2018} & \\
+	$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citet{Bertels2018} & \\
+	$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
+	$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citet{Mitov2018} & \\
+	$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citet{Blanquart2017} & \\
+	$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citet{Mitov2018} & \\
+	$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citet{Blanquart2017} & \\
 	\hline 
@@
 $\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$).
-Values from \protect\cite{Blanquart2017, Mitov2018} are empirical; values from \protect\cite{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
+Values from \protect\citep{Blanquart2017, Mitov2018} are empirical; values from \protect\citep{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
 	\begin{tabular}{lcc} \hline 
 	Measurement & Value & Reference \\ \hline 
-	$\bar{z}$ & $\approx$ 4.5 & \citealt{Mitov2018} \\
-	$\bar{z}$ &  4.4 & \citealt{Blanquart2017} \\
-	$\bar{z}$ & $\approx$ 4.5 & \citealt{Bonhoeffer2015} \\
-	$\sigma^2_z$ &  0.73 &  \citealt{Mitov2018} \\
-	$\sigma^2_z$ &  0.50 & \citealt{Blanquart2017} \\
-	$\sigma^2_z$ & $\approx$ 0.5 & \citealt{Bonhoeffer2015} \\ \hline 
+	$\bar{z}$ & $\approx$ 4.5 & \citet{Mitov2018} \\
+	$\bar{z}$ &  4.4 & \citet{Blanquart2017} \\
+	$\bar{z}$ & $\approx$ 4.5 & \citet{Bonhoeffer2015} \\
+	$\sigma^2_z$ &  0.73 &  \citet{Mitov2018} \\
+	$\sigma^2_z$ &  0.50 & \citet{Blanquart2017} \\
+	$\sigma^2_z$ & $\approx$ 0.5 & \citet{Bonhoeffer2015} \\ \hline 
 	\end{tabular}
@@
 	Raw data & 76979521 \\
-	Missing genotype rate $>$ 0.01 & 7452957 \\
-	Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 7452425 \\
-	Minor allele frequency $<$ 0.01 & 4556770 \\ \hline			
+	Missing genotype rate $>$ 0.05 & 11590002 \\
+	Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 11589246 \\
+	Minor allele frequency $<$ 0.01 & 6228626 \\ \hline			
 	\end{tabular}
@@
 \end{tabular}
-\label{gwas-top-associations}
+\label{tab:gwas-top-associations}
 \end{table}
@@
 % \begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
-% 		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.}
+% 		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citep{McLaren2015} compared to results from the current study.}
 % 		\hline
@@
 \typeout{}
-\printbibliography
 
@@
 
-%\cite{Mitov2018} and \cite{Bertels2018} both found higher statistical support for the POUMM model of spVL evolution as opposed to the corresponding model assuming Brownian motion only.
-This makes sense in the context of the hypothesis proposed by \cite{Fraser2014}, in which spVL is under stabilizing selection to maximize transmissibility of the virus.
+%\citep{Mitov2018} and \citep{Bertels2018} both found higher statistical support for the POUMM model of spVL evolution as opposed to the corresponding model assuming Brownian motion only.
+This makes sense in the context of the hypothesis proposed by \citep{Fraser2014}, in which spVL is under stabilizing selection to maximize transmissibility of the virus.
 On the other hand, the assumptions that $e$ are normally distributed, mutually indpendent, and have an expected value of 0 are consistent with the idea that genetic effects from several unlinked host loci and environmental effects from many independent factors are each equally likely to increase an individual's spVL as decrease it.
@@
 %The success of our method in identifying new host determinants of spVL depends on how much host-determined variation in spVL actually exists.
-In the largest GWAS for HIV spVL to date, \cite{McLaren2015} estimated that additive effects from common human genetic variants explain 20\% of the variability in spVL.
+In the largest GWAS for HIV spVL to date, \citep{McLaren2015} estimated that additive effects from common human genetic variants explain 20\% of the variability in spVL.
 Of this quantity, 94.5\% was attributable to variants in just two regions in the human genome.
 \todo{Were there variants other than HLA in the MHC region?} Associations were with alleles in the CCR5 gene, which encodes a co-receptor for HIV entry into CD4+ T-cells, and the human leukocyte antigen (HLA) genes, which encode proteins that present certain host and pathogen-derived peptides at the surface of cells.
-These associations have been well-replicated in several GWAS \cite{Naranbhai2017}.
+These associations have been well-replicated in several GWAS \citep{Naranbhai2017}.
 This results suggest that nearly all host-determined variation in spVL can be explained by previously discovered variants.
 %
-%In a joint analysis of both host and viral genotypes, \cite{Bartha2017} showed that nearly all of the variation in spVL explained by human genetic variation overlaps that explained by HIV genetic variation.
+%In a joint analysis of both host and viral genotypes, \citep{Bartha2017} showed that nearly all of the variation in spVL explained by human genetic variation overlaps that explained by HIV genetic variation.
 If this is the case then our method to remove the viral contribution to spVL prior to GWAS should yield only associations between host variants and spVL that are independent of the virus.
@@
 %According to the POUMM, the unobserved viral component of spVL for each host, $g_{v_i}$, evolves along the tree according to an Ornstein-Uhlenbeck (OU) process.
-The OU process was first introduced by \cite{Lande1976} as a model of evolution under both drift and stabilizing selection, so we are assuming $g_{v_i}$ continually evolves towards an evolutionary optimum in each viral lineage but is subject to random evolutionary fluctuations.
+The OU process was first introduced by \citep{Lande1976} as a model of evolution under both drift and stabilizing selection, so we are assuming $g_{v_i}$ continually evolves towards an evolutionary optimum in each viral lineage but is subject to random evolutionary fluctuations.
 Mathematically, $g_{v_i}$ is a realization of a Gaussian distributed random variable $G_{v_i} \sim \mathcal{N}(\mu_{OU_i}, \sigma^2_{OU_i})$, where:
@@
 % 	
-% 	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
+% 	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
 % 	
@@
 %
-%We made two notable changes to the typical PGLS framework (see \citealt{Garamszegi2014} for a nice overview of PGLS).
+%We made two notable changes to the typical PGLS framework (see \citet{Garamszegi2014} for a nice overview of PGLS).
 First, we assumed a variance-covariance structure for regression residuals according to the POUMM.
@@
 Second, we use the variances on the diagonal of this matrix as sample weights to account for non-ultrametricity of the HIV tree.
-We implemented PGLS including these two modifications using the nmle package in R \parencite{Pinheiro2019}.
+We implemented PGLS including these two modifications using the nmle package in R \citep{Pinheiro2019}.
 The resulting general least squares slope and intercept coefficient estimates for GWAS are:
@@
 Since a hypothesis test for significant association is performed for each of millions of genetic variants, GWAS performed in this way incurs a heavy multiple-testing burden.
-Despite a strong correction for multiple testing, \cite{McLaren2015} were able to confirm two
+Despite a strong correction for multiple testing, \citep{McLaren2015} were able to confirm two
 
 %So far, it is well established that both human host and viral genetic factors contribute to spVL.
-\cite{McLaren2015} found that 25\% of the variability in spVL in a mixed European cohort could be explained by additive genetic effects from the human genome.
-More recently, \cite{Mitov2018} established that at least 20\% of the variability in spVL in a UK cohort could be explained by viral genetic variation.
-Furthermore, depending on the methodological assumptions made and the cohort studied, estimates for the broad-sense viral heritability can vary widely \parencite{Mitov2018}.
-Finally, human and viral genetic variants may actually explain redundant variation in spVL \parencite{Bartha2017}.
+\citep{McLaren2015} found that 25\% of the variability in spVL in a mixed European cohort could be explained by additive genetic effects from the human genome.
+More recently, \citep{Mitov2018} established that at least 20\% of the variability in spVL in a UK cohort could be explained by viral genetic variation.
+Furthermore, depending on the methodological assumptions made and the cohort studied, estimates for the broad-sense viral heritability can vary widely \citep{Mitov2018}.
+Finally, human and viral genetic variants may actually explain redundant variation in spVL \citep{Bartha2017}.
 
 %Several GWAS have been performed to try to identify specific human and viral genetic variants that explain variation in spVL.
-The largest GWAS for human genetic factors so far, conducted by \cite{McLaren2015}, confirmed two regions of the human genome are associated with variability in spVL – .
+The largest GWAS for human genetic factors so far, conducted by \citep{McLaren2015}, confirmed two regions of the human genome are associated with variability in spVL – .
 This study used a standard GWAS framework, in which each surveyed genetic variant (millions in total) was seperately tested for a significant linear relationship with spVL based on thousands of genotyped and phenotyped individuals.
 This method did not take into account heterogeneity in the viral population.
-On the viral side, \cite{Bartha2013} performed an analogous GWAS for viral genetic factors.
+On the viral side, \citep{Bartha2013} performed an analogous GWAS for viral genetic factors.
 After correcting for multiple testing, they did not find any viral genetic variants significantly associated with spVL.
-Combining genomic information from both humans and viruses, \cite{Bartha2013} also tested for pairwise associations between human genetic variants and viral genetic variants.
+Combining genomic information from both humans and viruses, \citep{Bartha2013} also tested for pairwise associations between human genetic variants and viral genetic variants.
 In this way the authors could pinpoint sites of genomic conflict where the host's particular HLA alleles selected for certain viral escape mutations.
@@
 %
-%Several approaches have been commonly used to account for population structure in GWAS, including adjusting the GWAS significance threshold to account for potential spurious findings \parencite{Devlin1999}, performing GWAS only on more homogenous population sub-groups \parencite{Pritchard2000}, and adjusting the regression model itself to include estimates of individual relatedness \parencite{Price2006, Kang2010}.
-Of these approaches, we highlight the linear mixed model developed by \cite{Kang2010} because it is an increasingly popular technique that has a direct analog in the field of phylogenetic comparative analysis, from which we derive our method to account for pathogen relatedness.
+%Several approaches have been commonly used to account for population structure in GWAS, including adjusting the GWAS significance threshold to account for potential spurious findings \citep{Devlin1999}, performing GWAS only on more homogenous population sub-groups \citep{Pritchard2000}, and adjusting the regression model itself to include estimates of individual relatedness \citep{Price2006, Kang2010}.
+Of these approaches, we highlight the linear mixed model developed by \citep{Kang2010} because it is an increasingly popular technique that has a direct analog in the field of phylogenetic comparative analysis, from which we derive our method to account for pathogen relatedness.
 %
@@
 additive genetic effects from the host, age, sex, assay error, etc.) Once we have partitioned spVL values in this way we can perform GWAS using only the non-viral part of spVL, effectively removing non-independence and noise due to the viral phylogeny from spVL samples.
-To partition spVL into viral and non-viral parts, we rely on the a phylogenetic mixed model (PMM, \citealt{Housworth2004, Lynch1991}).
+To partition spVL into viral and non-viral parts, we rely on the a phylogenetic mixed model (PMM, \citet{Housworth2004, Lynch1991}).
 Under the PMM, a continuous trait is assumed to be composed of two independent parts: a heritable part and a non-heritable part.
@@
 Several mathematical models for evolution of the heritable part can be applied in order to estimate the relative contributions of heritable and non-heritable parts to the total trait value.
-It has recently been shown that an Ornstein-Uhlenbeck (OU) process is the best-fitting model for evolution of the heritable viral part of spVL \parencite{Mitov2018} \todo{Justify use of OU model better}.
-The mixed model of evolution including an OU process for the heritable part is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM, \citealt{Mitov2017a}).
+It has recently been shown that an Ornstein-Uhlenbeck (OU) process is the best-fitting model for evolution of the heritable viral part of spVL \citep{Mitov2018} \todo{Justify use of OU model better}.
+The mixed model of evolution including an OU process for the heritable part is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM, \citet{Mitov2017a}).
 By fitting such a model to spVL measurements and a viral phylogeny, we can co-infer the OU parameters that describe the evolutionary trajectory of the viral part of spVL as well as the variance of the non-heritable, non-viral part of spVL in the population.
@@
 Since spVL samples are not independent but are instead related according to the underlying viral phylogeny, the residuals may not be independent either.
-To account for non-independence among residuals, \cite{Grafen1989} suggested the PGLS method, in which the phylogeny is translated into a variance-covariance structure among samples.
+To account for non-independence among residuals, \citep{Grafen1989} suggested the PGLS method, in which the phylogeny is translated into a variance-covariance structure among samples.
 It is assumed that residual errors are not independent, but instead follow this phylogenetic variance-covariance structure.
@@
 However, in certain infectious disease  contexts the viral population can also confound GWA analyses.
-In the genome-to-genome (G2G) GWAS framework described by \cite{Bartha2013}, host variants are tested for pairwise associations with viral variants to determine sites of genetic conflict between the human host and HIV pathogen.
+In the genome-to-genome (G2G) GWAS framework described by \citep{Bartha2013}, host variants are tested for pairwise associations with viral variants to determine sites of genetic conflict between the human host and HIV pathogen.
 In this case, confounding may arise from population stratification in both the host and viral populations (Figure \ref{fig:gwas-covariates}b).
-\cite{Naret2018} describe two potential types of stratification that may occur in the G2G context.
+\citep{Naret2018} describe two potential types of stratification that may occur in the G2G context.
 So-called ``major stratification" occurs when host population is correlated with viral population and vice-versa.
 ``Minor stratification"  occurs when host population is correlated with with viral variant frequencies or when viral population is correlated with host variant frequencies.
-Although \cite{Naret2018} envision only discrete viral populations, the same phenomena could easily apply when considering the full viral phylogeny.
+Although \citep{Naret2018} envision only discrete viral populations, the same phenomena could easily apply when considering the full viral phylogeny.
 %
 %Several methods have been developed to correct for confounding from the viral phylogeny in HIV G2G GWAS.
-One of two methods originally introduced by \cite{Bhattacharya2007} involves determining whether viral mutations are inferred to occur disproportionately along branches of the phylogeny where a particular host variant is present.
-The other method is a model-selection approach in which phylogenetic models with and without selection pressure from host variants are compared for goodness-of-fit \parencite{Bhattacharya2007, Carlson2008}.
-Lastly, the more recently proposed method by \cite{Naret2018} uses phylogenetic principle components to describe the major axes of variation in the viral population (an approach analogous to using principle components to account for host population structure).
-However, the method proposed by \cite{Naret2018} does not have a secondary benefit of the other two methods, which is to restore statistical independence to samples that are fundamentally related by a phylogeny.
+One of two methods originally introduced by \citep{Bhattacharya2007} involves determining whether viral mutations are inferred to occur disproportionately along branches of the phylogeny where a particular host variant is present.
+The other method is a model-selection approach in which phylogenetic models with and without selection pressure from host variants are compared for goodness-of-fit \citep{Bhattacharya2007, Carlson2008}.
+Lastly, the more recently proposed method by \citep{Naret2018} uses phylogenetic principle components to describe the major axes of variation in the viral population (an approach analogous to using principle components to account for host population structure).
+However, the method proposed by \citep{Naret2018} does not have a secondary benefit of the other two methods, which is to restore statistical independence to samples that are fundamentally related by a phylogeny.
 %
@@
 %
-%As \cite{Bhattacharya2007} recognized, HIV samples are related by descent according to the transmission tree.
+%As \citep{Bhattacharya2007} recognized, HIV samples are related by descent according to the transmission tree.
 Such relationships are often investigated by inferring a viral phylogeny from sequencing data.
-Recently, several studies using phylogentic methods determined that host spVL values are correlated among hosts with closely related versions of the virus \parencite{Alizon2010, Blanquart2017, Hollingsworth2010, Mitov2018}.
+Recently, several studies using phylogentic methods determined that host spVL values are correlated among hosts with closely related versions of the virus \citep{Alizon2010, Blanquart2017, Hollingsworth2010, Mitov2018}.
 This correlation can be quantified in terms of the heritability of spVL, or, the fraction of variance in spVL among hosts that can be explained by underlying viral genetic variation.
-\cite{Mitov2018} demonstrated that spVL is at least 20\% heritable, meaning that the virus has a strong effect on host spVL.
+\citep{Mitov2018} demonstrated that spVL is at least 20\% heritable, meaning that the virus has a strong effect on host spVL.
 Based on this correlation between the virus and spVL (Figure \ref{fig:gwas-covariates}c), we argue it is necessary to expand the correlation model typically assumed in standard GWAS.
@@
 This phenomenon is called population stratification and several methods have been developed to address the problem.
-In the context of GWAS for host determinants of spVL, a principle components method \cite{Price2006} has most often been applied.
+In the context of GWAS for host determinants of spVL, a principle components method \citep{Price2006} has most often been applied.
 In this method, the top few principle components of host genetic variation are included in the association model as proxy variables for demographic and geographic factors.
-This effectively changes the association test hypothesis from ``does the host variant explain a significant amount of variation in spVL?", to ``does the host variant explain a significant amount of variation in spVL \emph{beyond that which can be explained by demographic or geographic factors?}" Previous GWAS for host determinants of spVL conducted by \cite{Price2006, McLaren2015, Bartha2013, Fellay2007, Pereyra2010, Pelak2010, VanManen2011} have employed this method.
-Several studies have additionally used other methods to account for host population stratification, including structured association \parencite{Pritchard2000} and genomic control \parencite{Devlin1999}.
+This effectively changes the association test hypothesis from ``does the host variant explain a significant amount of variation in spVL?", to ``does the host variant explain a significant amount of variation in spVL \emph{beyond that which can be explained by demographic or geographic factors?}" Previous GWAS for host determinants of spVL conducted by \citep{Price2006, McLaren2015, Bartha2013, Fellay2007, Pereyra2010, Pelak2010, VanManen2011} have employed this method.
+Several studies have additionally used other methods to account for host population stratification, including structured association \citep{Pritchard2000} and genomic control \citep{Devlin1999}.
 In structured association, the sample population is a priori divided into groups that are assumed to be homogeneous with respect to potentially confounding variables.
 GWA testing is then conducted independently within each group before results are finally aggregated across groups.
-\cite{McLaren2015, Pereyra2010} both applied a form of structured association.
-So far only \cite{Dalmasso2008} have used genomic control to account for population stratification in spVL GWAS.
+\citep{McLaren2015, Pereyra2010} both applied a form of structured association.
+So far only \citep{Dalmasso2008} have used genomic control to account for population stratification in spVL GWAS.
 In genomic control, the significance threshold for association testing is uniformly adjusted for all variants by a factor which accounts for the degree of stratification present.
@@
 These variables effect the phenotype, but are not correlated with the genetic variant being tested (Figure \ref{fig:gwas-covariates}a).
-As long as the phenotype is quantitative (as in the case of spVL) or the sample is randomly drawn from the population, including these factors as covariates in the association model helps remove residual variance and increase power to detect true associations \todo{Investigate this claim further} \parencite{Mefford2012}.
-In the context of GWAS for host determinants of spVL, sex, cohort, and age have all been recognized to be noise variables that effect spVL but should not be correlated with host genetic variants \parencite{Bartha2013, Dalmasso2008, Fellay2007, Fellay2009, Pelak2010}.
+As long as the phenotype is quantitative (as in the case of spVL) or the sample is randomly drawn from the population, including these factors as covariates in the association model helps remove residual variance and increase power to detect true associations \todo{Investigate this claim further} \citep{Mefford2012}.
+In the context of GWAS for host determinants of spVL, sex, cohort, and age have all been recognized to be noise variables that effect spVL but should not be correlated with host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Fellay2009, Pelak2010}.
 
@@
 %
-%First, we fixed the total variance in simulated spVL to be 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \parencite{Mitov2018}.
-After allotting 25\% of this variance to $g_h$ (following \citealt{McLaren2015}), we partitioned the remaining variance between $g_v$ and $\epsilon$ in different ratios in order to assess our estimator’s performance across a range of possible spVL heritability values (Figure \ref{fig:sim-design}a, Table \ref{tab:sim-params}).
+%First, we fixed the total variance in simulated spVL to be 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}.
+After allotting 25\% of this variance to $g_h$ (following \citet{McLaren2015}), we partitioned the remaining variance between $g_v$ and $\epsilon$ in different ratios in order to assess our estimator’s performance across a range of possible spVL heritability values (Figure \ref{fig:sim-design}a, Table \ref{tab:sim-params}).
 Second, we generated the host part of spVL by summing contributions from $M$ causal host variants (Figure \ref{fig:sim-design}b).
@@
 Third, we generated $g_v$ by simulating an Ornstein-Uhlenbeck (OU) process of evolution along a viral phylogenetic tree (Figure \ref{fig:sim-design}c).
-We use a randomly generated tree with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \parencite{Hodcroft2014}.
+We use a randomly generated tree with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014}.
 To test under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively.
-For the other OU parameters, we used previously inferred values from \parencite{Bertels2018, Blanquart2017, Mitov2018}.
+For the other OU parameters, we used previously inferred values from \citep{Bertels2018, Blanquart2017, Mitov2018}.
 Finally, $\epsilon$ values were generated from a normal distribution with mean zero.
@@
 %	
-%	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \cite{Petersen2012}.
+%	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
 %\end{proof}
@@
 %	\begin{tabular}{llclccc} \toprule 
-%	& \multicolumn{2}{c}{\specialcell[l]{\cite{McLaren2015} study \\ N = 6,315}} & \multicolumn{3}{c}{\specialcell[l]{Current study \\ N = 1,370}} & \\ \midrule 
+%	& \multicolumn{2}{c}{\specialcell[l]{\citep{McLaren2015} study \\ N = 6,315}} & \multicolumn{3}{c}{\specialcell[l]{Current study \\ N = 1,370}} & \\ \midrule 
 %	Region & Variant & \emph{P} value & Variant & Effect size & \emph{P} value & R$^2$ \\ \midrule
@@
 %	\end{tabular}
-%	\caption{Association results for the top SNPs in the MHC and \emph{CCR5} regions from the GWAS by \cite{McLaren2015} compared to association results for the best proxy SNPs for these SNPs available in the current study.
+%	\caption{Association results for the top SNPs in the MHC and \emph{CCR5} regions from the GWAS by \citep{McLaren2015} compared to association results for the best proxy SNPs for these SNPs available in the current study.
 The proxy SNPs were chosen based on high linkage disequilibrium (as measured by R$^2$) with the McLaren et al.
@@
 %			\toprule 
-%			\textbf{Filter} & \textbf{Current study} & \textbf{\citealt{Bartha2013}} & \textbf{\citealt{McLaren2015}} \\
+%			\textbf{Filter} & \textbf{Current study} & \textbf{\citet{Bartha2013}} & \textbf{\citet{McLaren2015}} \\
 %			\midrule 
@@
 %		\end{tabular}
-%		\caption{Top association results from \cite{McLaren2015} compared to association results for the best proxy SNPs available in the current study.
+%		\caption{Top association results from \citep{McLaren2015} compared to association results for the best proxy SNPs available in the current study.
 Proxy SNPs were chosen based on linkage disequilibrium (as measured by R$^2$) with the McLaren et al.
@@
 
-% Again following the procedure of \cite{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
+% Again following the procedure of \citep{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
 Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).
@@
 % 	\end{tabular}
-% 	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \cite{McLaren2015} compared to results from the current study.
+% 	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citep{McLaren2015} compared to results from the current study.
 The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al.

### 2021-11-16 scripts/R/make_manuscript_figures.rmd
make correlations clearer in spvl comparison figure, ignore some manuscript-related files

@@
 ```{r}
-p <- ggplot(data = spvl_values,
+comp_1 <- ggplot(data = spvl_values,
              aes(x = spvl_lenient_filter_first_measurement,
-                y = spvl_lenient_filter_mean_measurement,
-                color = spvl_strict_filter_mean_measurement)) +
-        scale_color_gradient(low = "green", high = "red", name = "Mean measurement\nwith stricter filtering") +
+                 y = spvl_lenient_filter_mean_measurement)) +
         coord_equal() +
+        lims(x = c(0, 7.25), y = c(0, 7.25)) +
         geom_point() +
-        labs(x = "First measurement", y = "Mean measurement") +
+        labs(x = "First measurement", y = "Mean measurement\n") +
+        geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
         shared_theme
 
-ggsave(plot = p, filename = paste(FIGDIR, "spvl_calculation_comparison.png", sep = "/"),
-       width = linewidth * 2/3, height = height * 2/3, units = "cm")
+comp_2 <- ggplot(data = spvl_values,
+               aes(x = spvl_lenient_filter_first_measurement,
+                   y = spvl_strict_filter_mean_measurement)) +
+        coord_equal() +
+        lims(x = c(0, 7.25), y = c(0, 7.25)) +
+        geom_point() +
+        labs(x = "First measurement", y = "Mean measurement\nwith stricter filtering") +
+        geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+        shared_theme
+
+comp_3 <- ggplot(data = spvl_values,
+                 aes(x = spvl_lenient_filter_mean_measurement,
+                     y = spvl_strict_filter_mean_measurement)) +
+        coord_equal() +
+        lims(x = c(0, 7.25), y = c(0, 7.25)) +
+        geom_point() +
+        labs(x = "Mean measurement", y = "Mean measurement\nwith stricter filtering") +
+        geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
+        shared_theme
+
+comp <- cowplot::plot_grid(comp_1, comp_2, comp_3, nrow = 1)
+
+ggsave(plot = comp, filename = paste(FIGDIR, "spvl_calculation_comparison.png", sep = "/"),
+       width = linewidth, height = height * 2/3, units = "cm")
 

### 2021-11-16 manuscript/hiv-gwas.tex

@@
 \begin{flushleft}
-	Sarah Nadeau\textsuperscript{1,2}, Christian Thorball\textsuperscript{2,3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F.
-Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Manuel Battegay\textsuperscript{10}, Matthias Cavassini\textsuperscript{11}, Alexandra Calmy\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
+	Sarah Nadeau\textsuperscript{1,2}, Christian W.
+Thorball\textsuperscript{3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F.
+Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Hans H.
+Hirsch\textsuperscript{8,10,11}, Matthias Cavassini\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3,15}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
 \end{flushleft}
@@
 \textsuperscript{2}\emph{Swiss Institute of Bioinformatics, Lausanne} \\
-\textsuperscript{3}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
+\textsuperscript{3}\emph{Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland} \\
 \textsuperscript{4}\emph{Institute of Medical Virology, University of Zürich, Zürich} \\
@@
 \textsuperscript{7}\emph{Division of Immunology and Allergy, University Hospital Lausanne, Lausanne} \\
-\textsuperscript{8}\emph{Molecular Virology, Department of Biomedicine - Petersplatz, University of Basel, Basel} \\
+\textsuperscript{8}\emph{Department of Biomedicine, University of Basel, Basel} \\
 \textsuperscript{9}\emph{Department of Infectious Diseases, Berne University Hospital and University of Berne, Berne} \\
-\textsuperscript{10}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
-\textsuperscript{11}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
-\textsuperscript{12}\emph{HIV/AIDS Unit, Infectious Disease Service, Geneva University Hospital, Geneva} \\
+\textsuperscript{10}\emph{Division of Clinical Virology, University Hospital Basel, Basel} \\
+\textsuperscript{11}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
+\textsuperscript{12}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
 \textsuperscript{13}\emph{Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen} \\
 \textsuperscript{14}\emph{Division of Infectious Diseases, Regional Hospital Lugano, Lugano} \\
+\textsuperscript{15}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
 \textsuperscript{*}\emph{Co-last authors}
@@
 Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits.
-However, GWAS for host genetic determinants of infectious disease traits do not account for potentially confounding or extraneous pathogen effects that are heritable from one infection partner to another in transmission chains.
-We propose a novel method to consider the full genome of each patient's infecting strain, remove strain-specific effects on a trait, and thus better estimate the effect of human genetic variants on infectious disease traits.
-In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable.
-When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association with spVL in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to a standard GWAS.
-Our simulation study confirms that estimated heritability and selection parameters for HIV-1 mean standard GWAS are quite robust to pathogen effects.
-Our framework may improve GWAS for other diseases where pathogen effects are stronger and more correlated amongst individuals in a cohort.
+However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.
+We propose a novel method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits.
+In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
+When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to standard GWAS.
+Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects.
+Our framework may improve GWAS for other diseases where pathogen effects are stronger and more phylogenetically correlated amongst individuals in a cohort.
 \end{abstract}
@@
 In a typical GWAS, millions of genetic variants from across the human genome are screened for statistical association with a trait of interest.
-Ideally, this procedure identifies variants that are located in or are in linkage disequilibrium with alleles that directly effect the trait.
+Ideally, this procedure identifies variants that are located in, or are in linkage disequilibrium with, alleles that directly affect the trait.
 If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \citep{Okada2014}.
@@
 Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
-The largest to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
+The largest study to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
 All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}).
-Both associations are biologically relevant: the MHC encodes proteins that display viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry.
+Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry.
 In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed.
-CCR5 proteins are a signal to the virus that help it find the right cells to infect.
+CCR5 proteins help the virus infect target cells.
 
 % How heritable is spVL (viral side)
-In addition to these human genetic factors, is also well-recognized that viral genetic factors effect spVL.
+In addition to these human genetic factors, it is well-recognized that viral genetic factors effect spVL.
 As mentioned, heritability from the viral side is typically measured using broad-sense heritability.
@@
 Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
-However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016}.
-If the virus can change it's ``puzzle piece'' shape to escape MHC proteins, infected cells can go undetected.
-This means that MHC variants effect spVL largely via selection on the virus \citep{Bartha2017}.
+However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021}.
+If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected.
+This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}.
 In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.
@@
 In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors.
-This has several potential negative consequences: (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
+This has several potential negative consequences, namely that (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
 
 % Overview of relevant methods and why we don't use them
-Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry in a cohort can give rise to spurious genetic correlations with a trait.
+Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry, especially between close relatives, can give rise to spurious genetic correlations with a trait.
 Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}.
@@
 	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_E\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_e\big)
+	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_E\big)
 	\end{split}
@@
 Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%.
-If were were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable.
+If we were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable.
 We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects.
@@
 Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
-The black point represents our estimates for the heritability and selection strength of spVL based on Swiss data.
+The black point represents our estimates for the heritability and selection strength of spVL based on Swiss HIV data.
 RMSE = Root mean square error, TPR = True positive rate.}
@@
 Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions.
-Taking into account phylogenetic information slightly decreased association strength for most variants in \emph{CCR5} region.
+Taking into account phylogenetic information slightly decreased association strength for most variants in the \emph{CCR5} region.
 Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
@@
 Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods.
-So although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.
-
-Instead, the main novelty of our approach was to correct the spVL trait prior to GWAS by estimating and removing pathogen effects.
+Although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.
+
+Instead, the main novelty of our approach was to correct the spVL trait prior to performing the GWAS, thereby estimating and removing pathogen effects.
 In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
 Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.
-Unfortunately, given our estimates for the heritability of spVL and the selection strength on this trait, our simulation results show that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (figure \ref{fig:simulationResults}).
-Our method is anticipated to be more useful for disease traits that are highly heritable from infection partner to infection partner and  maintain a high correlation between sampled individuals.
+
+Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals.
 In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
-So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to be meaningful.
+So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
+
+Given our estimates for the heritability of spVL and the selection strength on this trait using Swiss HIV cohort data, our simulation results reveal that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (figure \ref{fig:simulationResults}).
+Indeed the results of our  GWAS method presented here and a standard GWAS are very similar.
+Future applications might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g.
+\cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.
 
@@
 Therefore, it is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}).
-The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project github at TODO.
+The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.
 
@@
 Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}).
-Kinship was evaluated using PLINK version 2.00a2LM \citep{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375.
-Initial host genotyping quality control and imputation were done by the SHCS.
+Kinship was evaluated using PLINK version 2.3 \citep{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375.
+Initial host genotyping quality control and imputation were done as in \cite{Thorball2021GeneticLymphoma}.
 Subsequent genotyping quality control was performed using PLINK version 2.3 \citep{Chang2015}.
@@
 This study was funded by ETH Zurich.
-We would like to thank Esposito Mauro and Alexandra Scherrer from the SHCS data center, who were especially helpful in retrieving data, and Michael Landis, who shared a LaTeX template for graphical model drawing.
+We would like to thank Esposito Mauro and Alexandra Scherrer from the SHCS data center, who were especially helpful in retrieving data, Alexandra Calmy and her team from the Geneva University Hospital for data contributions, and Michael Landis, who shared a LaTeX template for graphical model drawing.
 Finally and most importantly, we thank the participants of the Swiss HIV Cohort Study for contributing their samples and making this research possible.

### 2021-11-17 scripts/R/make_manuscript_figures.rmd
add more descriptive figure for simulation model

@@
 ```
+
+OU process for simulation setup figure
+```{r}
+# Simulate many OU trajectories along tree
+generate_trait_trajectory <- function(v0, alpha, theta, sigma, N_steps) {
+      POUMM::rTrajectoryOU(
+            z0 = v0,
+            t = 1,
+            alpha = alpha,
+            theta = theta,
+            sigma = sigma,
+            steps = N_steps)
+}
+
+# Generate tree
+set.seed(10)
+tree <- ape::rtree(n = 4)
+tree$edge.length <- tree$edge.length * 6
+node_times <- POUMM::nodeTimes(tree)
+edge_colors = RColorBrewer::brewer.pal(n = 4, name = diverging_palette)
+
+# Define OU process
+scale_factor <- 10  # 1000
+v0 <- 4.5
+theta <- 4.5
+alpha <- 60 / scale_factor
+sigma <- 2 / sqrt(scale_factor)
+
+# Initialize data structures
+traceplot <- as.data.frame(matrix(ncol = 4, nrow = 0))
+colnames(traceplot) <- c("lineage", "time", "v", "color")
+v_nodes <- rep(v0, length(node_times))
+
+set.seed(30)
+# Generate OU trajectory for each edge on tree
+for (j in 1:nrow(tree$edge)) {
+      from = tree$edge[j, 1]
+      to = tree$edge[j, 2]
+      N_steps = round(tree$edge.length[j]*scale_factor) - 1
+      print(paste(N_steps, "steps on branch", j))
+
+      trajectory <- POUMM::rTrajectoryOU(
+            z0 = v_nodes[from],
+            t = 0.0001,
+            alpha = alpha,
+            theta = theta,
+            sigma = sigma,
+            steps = N_steps)
+
+      v_nodes[to] <- trajectory[N_steps]
+      v <- c(v_nodes[from], trajectory)
+
+      new_data <- data.frame(
+            lineage = rep(j, N_steps + 1),
+            time = seq(node_times[from], node_times[to], length.out = N_steps + 1),
+            v = v,
+            color = rep(edge_colors[j], N_steps + 1))
+      traceplot <- rbind(traceplot, new_data)
+}
+
+mu_OU <- function(t) v0 * exp(-alpha * t) + theta * (1 - exp(-alpha * t))
+
+OU_plot <- ggplot(
+      data = traceplot,
+      aes(x = time, y = v, group = lineage)) +
+      geom_line(size = 1.25) +
+      theme_void()
+
+#TODO: I have no idea how the edge coloring works, this is a really shitty hardcoded soln
+# tree_data = data.frame(node = c(1, 2, 3, 4, 5), color = c(edge_colors[2], edge_colors[3], edge_colors[4], "blue", edge_colors[1]))
+
+tree_plot <- ggtree::ggtree(
+      tree,
+      ladderize = F,
+      size = 1.5)
+
+# Align OU and tree plots
+aligned_plots <- align_plots(OU_plot, tree_plot, align = "hv", axis = "tblr", rel_heights = c(3, 1))
+
+svglite::svglite(filename = paste(FIGDIR, "ou_process_for_simulation_figure.svg", sep = "/"))
+      ggdraw(aligned_plots[[1]])
+dev.off()
+
+svglite::svglite(filename = paste(FIGDIR, "tree_for_simulation_figure.svg", sep = "/"))
+ggdraw(aligned_plots[[2]])
+dev.off()
+```

### 2021-11-17 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 	\begin{center}
-		\includegraphics[width=0.462\linewidth]{figures/model_figure}
-		\caption{The POUMM statistical model for the spVL trait.
+	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+% 		\includegraphics[width=0.462\linewidth]{figures/model_figure}
+% 		\caption{The POUMM statistical model for the spVL trait.
 (A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny.
@@
 \begin{center}
-\includegraphics[width = 0.462\linewidth]{figures/spvl_calculation_comparison.png}
+\includegraphics[width = 0.7\linewidth]{figures/spvl_calculation_comparison.png}
 	\caption{A comparison of different ways to calculate spVL based on viral load measurements provided by the SHCS.

### 2021-11-18 manuscript/hiv-gwas.tex

@@
 \begin{flushleft}
-	Sarah Nadeau\textsuperscript{1,2}, Christian W.
-Thorball\textsuperscript{3}, Roger Kouyos\textsuperscript{4,5}, Huldrych F.
-Günthard\textsuperscript{4,5}, Jürg Böni\textsuperscript{4}, Sabine Yerly\textsuperscript{6}, Matthieu Perreau\textsuperscript{7}, Thomas Klimkait\textsuperscript{8}, Andri Rauch\textsuperscript{9}, Hans H.
-Hirsch\textsuperscript{8,10,11}, Matthias Cavassini\textsuperscript{12}, Pietro Vernazza\textsuperscript{13}, Enos Bernasconi\textsuperscript{14}, Jacques Fellay\textsuperscript{2,3,15}, Venelin Mitov\textsuperscript{1,2,*}, Tanja Stadler\textsuperscript{1,2,*}, and the Swiss HIV Cohort Study (SHCS)
+	Sarah Nadeau,$^{1,2}$ Christian W.
+Thorball,$^{3}$ Roger Kouyos,$^{4,5}$ Huldrych F.
+Günthard,$^{4,5}$ Jürg Böni,$^{4}$ Sabine Yerly,$^6$ Matthieu Perreau,$^{7}$ Thomas Klimkait,$^{8}$ Andri Rauch,$^{9}$ Hans H.
+Hirsch,$^{8,10,11}$ Matthias Cavassini,$^{12}$ Pietro Vernazza,$^{13}$ Enos Bernasconi,$^{14}$ Jacques Fellay,$^{2,3,15}$ Venelin Mitov,$^{\dagger,1,2}$ Tanja Stadler,$^{\dagger,\ast,1,2}$ and the Swiss HIV Cohort Study (SHCS)
 \end{flushleft}
@@
 \noindent
-\textsuperscript{1}\emph{Department of Biosystems Science and Engineering, ETH Zürich, Basel} \\
-\textsuperscript{2}\emph{Swiss Institute of Bioinformatics, Lausanne} \\
-\textsuperscript{3}\emph{Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland} \\
-\textsuperscript{4}\emph{Institute of Medical Virology, University of Zürich, Zürich} \\
-\textsuperscript{5}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zürich, University of Zürich, Zürich} \\
-\textsuperscript{6}\emph{Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva} \\
-\textsuperscript{7}\emph{Division of Immunology and Allergy, University Hospital Lausanne, Lausanne} \\
-\textsuperscript{8}\emph{Department of Biomedicine, University of Basel, Basel} \\
-\textsuperscript{9}\emph{Department of Infectious Diseases, Berne University Hospital and University of Berne, Berne} \\
-\textsuperscript{10}\emph{Division of Clinical Virology, University Hospital Basel, Basel} \\
-\textsuperscript{11}\emph{Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel} \\
-\textsuperscript{12}\emph{Division of Infectious Diseases, University Hospital Lausanne, Lausanne} \\
-\textsuperscript{13}\emph{Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen} \\
-\textsuperscript{14}\emph{Division of Infectious Diseases, Regional Hospital Lugano, Lugano} \\
-\textsuperscript{15}\emph{Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne} \\
-\textsuperscript{*}\emph{Co-last authors}
+$^{1}$Department of Biosystems Science and Engineering, ETH Zürich, Basel, Switzerland\\
+$^{2}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\ 
+$^{3}$Precision Medicine Unit, Lausanne University Hospital and University of Lausanne, Lausanne, Switzerland\\ 
+$^{4}$Institute of Medical Virology, University of Zurich, Zurich, Switzerland\\ 
+$^{5}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Zurich, University of Zurich, Zurich, Switzerland\\
+$^{6}$Division of Infectious Diseases, Laboratory of Virology, Geneva University Hospital, Geneva, Switzerland\\
+$^{7}$Division of Immunology and Allergy, University Hospital Lausanne, Lausanne, Switzerland\\
+$^{8}$Department of Biomedicine, University of Basel, Basel, Switzerland\\
+$^{9}$Department of Infectious Diseases, Bern University Hospital and University of Bern, Bern, Switzerland\\
+$^{10}$Division of Clinical Virology, University Hospital Basel, Basel, Switzerland\\
+$^{11}$Division of Infectious Diseases and Hospital Epidemiology, University Hospital Basel, Basel, Switzerland\\
+$^{12}$Division of Infectious Diseases, University Hospital Lausanne, Lausanne, Switzerland\\
+$^{13}$Division of Infectious Diseases, Cantonal Hospital St. Gallen, St. Gallen, Switzerland\\
+$^{14}$Division of Infectious Diseases, Regional Hospital Lugano, Lugano, Switzerland\\
+$^{15}$Global Health Institute, School of Life Sciences, École Polytechnique Fédérale de Lausanne, Lausanne, Switzerland\\
+$\dagger$Co-last authors\\
+$*$Corresponding author: tanja.stadler@bsse.ethz.ch
 
@@
 In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
-When we apply our method to HIV-1 cohort data from Switzerland, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatability complex (MHC) gene regions compared to standard GWAS.
+When we apply our method to HIV-1 cohort data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
 Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects.
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 In untreated individuals, spVL is predictive of duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}.
-If viral load can be reduced to undetectable levels, an individual is effectively uninfectious \citep{HHSARTGuidelines2019}.
+If viral load can be reduced to undetectable levels, an individual is effectively uninfectious and the risk of disease progression is massively reduced \citep{HHSARTGuidelines2019}.
 Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}.
@@
 Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry.
-In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that infected cells should be destroyed.
+In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that a cell is infected.
 CCR5 proteins help the virus infect target cells.
@@
 
-In the following we (i) describe a statistical model for the spVL trait, (ii) derive a maximum likelihood estimate for the viral part of spVL under this model, and (iii) describe a new infectious disease GWAS framework using this information.
+In the following we describe a statistical model for the spVL trait, derive a maximum likelihood estimate for the viral part of spVL under this model, and describe a new infectious disease GWAS framework using this information.
 In simulations, we demonstrate that this framework can improve GWAS power to detect host genetic variants effecting a disease trait.
@@
 
-\subsection{A statistical model for spVL}
+\section*{New Approaches}
+\subsection*{A statistical model for spVL}
 
@@
 In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part.
-Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny.
-The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny (Figure \ref{fig:spVLModel}).
+Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A.
+The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.
 
@@
 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+	\caption{A high-level schematic of our POUMM-based simulation framework.
+(A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an OU process.
+(B) shows host genetic effects, which are the sum of independent effects from several causal variants.
+Each variant can be present in 0, 1, or 2 copies and has some effect $\delta$ on the trait value.
+We assume each variant has the same magnitude of effect and that half the causal variants have a positive effect and the other half have a negative effect.
+(C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero with variance $\sigma^2_e$.}
 % 		\includegraphics[width=0.462\linewidth]{figures/model_figure}
@@
 
-\subsection{Teasing apart pathogen and non-pathogen effects on spVL}
+\subsection*{Teasing apart pathogen and non-pathogen effects on spVL}
 
@@
 
-\subsection{A POUMM-based GWAS framework for infectious disease}
+\subsection*{A POUMM-based GWAS framework for infectious disease}
 
@@
 
-\section{Results}
-
-\subsection{Simulation study}
+\section*{Results}
+
+\subsection*{Simulation study}
 
 To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data.
-Figure \ref{fig:sim-design} shows the simulation scheme and table \ref{tab:sim-params} gives the value or expression for each parameter.
+Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and table \ref{tab:sim-params} gives the value or expression for each parameter.
 In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength.
@@
 
-\subsubsection{Estimator accuracy}
+\subsubsection*{Estimator accuracy}
 
@@
 To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
-In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation set-up.
+In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation scheme.
 By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable and under low selection pressure.
@@
 
-\subsubsection{Theoretical GWAS improvement}
+\subsubsection*{Theoretical GWAS improvement}
 
@@
 
-\subsection{GWAS on the Swiss HIV Cohort}
+\subsection*{GWAS on the Swiss HIV Cohort}
 
@@
 
-% TODO
 \begin{figure}[H]
@@
 
-\section{Discussion}
+\section*{Discussion}
 % State the importance of your findings
@@
 Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort.
-We note that our sample size (N = 1493 individuals) is quite a bit smaller than these other studies.
+We note that our sample size (N = 1493 individuals) is smaller than in these other studies.
 This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases.
@@
 
-%\begin{itemize}
-%	\item Compare inferred POUMM parameters to previous studies \citep{Mitov2018, Bertels2018}
-%	\item Standard GWAS highlights previously identified regions associated with spVL: $CCR5$ and MHC
-%	\item We expect non-viral only GWAS to increase signals of association in $CCR5$ (because noise in spVL from the virus reduced) and decrease signals of association in MHC (because viral mutations mediate MHC effects).
-It \emph{seems} like the results support these hypotheses, but I am waiting on POUMM parameter estimates from a longer MCMC run.
-%	\item The phylogenetic correction to GWAS leads to the biggest p-value changes in the $CCR5$ and MHC regions
-%	\item SNPs near the FUT1 gene (on chromosome 6, highlighted as one of the top-20 p-value changes in the first pipeline run) increased in significance, and the Lewis-X antigen produced by this gene is known to play a role in early HIV infection \citep{Naarding2005}.
-%	\item Address method caveats: 
-%	\begin{itemize}
-%		\item We don't model virus-host interaction effects
-%		\item We don't account for uncertainty in non-viral component of spVL estimates
-%	\end{itemize}
-%	\item Implications for future research: what other infectious disease traits are heritable?
-%\end{itemize}
-
-\section{Methods}
-
-\subsection{Simulation model}
+\section*{Materials and Methods}
+
+\subsection*{Simulation model}
 
@@
 Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.
+For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.
 
@@
 		Variable & Expression & Definition \\ \midrule 
-		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & total spVL variance \\ 
-		$H^2_{\bar{t}}$ & varied & spVL viral heritability at $\bar{t}$ \\
-		$\sigma^2_{g_h}$ & $\sigma_{g_h}^2 = 0.25*\sigma_z^2$ & variance in host part of spVL \\
-		$\sigma^2_{g_v}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & variance in viral part of spVL at $\bar{t}$ \\ 
-		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{g_v} - \sigma^2_{g_h}$ & variance in environmental part of spVL \\ 
-		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & mean root-tip time in viral phylogeny \\ 
-		$\bm{g}_v$ & $\bm{g}_v \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & viral part of spVL for all individuals \\ 
-		$\theta$ & 4.5 log copies/mL & optimal spVL value \\
+		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & Total spVL variance \\ 
+		$H^2_h$ & 0.25 & Host heritability of spVL \\
+		$H^2_{\bar{t}}$ & varied & Viral heritability of spVL at $\bar{t}$ \\
+		$\sigma^2_{g_h}$ & $\sigma_{g_h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
+		$\sigma^2_{g_v}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in viral part of spVL at $\bar{t}$ \\ 
+		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{g_v} - \sigma^2_{g_h}$ & Variance in environmental part of spVL \\ 
+		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in viral phylogeny \\ 
+		$\bm{g}_v$ & $\bm{g}_v \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & Viral part of spVL for all individuals \\ 
+		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
 		$g_{0}$ & 4.5 log copies/mL & $g_v$ at the root of the phylogeny  \\ 
-		$\alpha$ & varied & selection strength of OU process \\ 
-		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & time-unit standard deviation of OU process \\ 
-		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & viral phylogeny \\ 
-		$g_{h_i}$ & $g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & host part of spVL for individual $i$ \\ 
-		$G_{N \times M}$ & $G_{ij} \sim Binom(2, p)\ \forall i \in {1...N}, \forall j \in {1...M}$ & host genotype matrix \\
-		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{\bar{t}}^2\sigma^2_z}{2\delta^2M}}$ &  host variant allele frequency \\ 
-		$\delta$ & 0.2 & host variant effect size \\ 
-		$M$ & 20 & number of causal host variants \\ 
-		$\epsilon_i$ & $ \epsilon_i \sim Norm(0, \sigma^2_{\epsilon})$ & environmental part of spVL for individual $i$ \\ \bottomrule 
+		$\alpha$ & varied & Selection strength of OU process \\ 
+		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
+		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Viral phylogeny \\ 
+		$g_{h_i}$ & $g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & Host part of spVL for individual $i$ \\ 
+		$G_{N \times M}$ & $G_{ij} \sim Binom(2, p)\ \forall i \in {1...N}, \forall j \in {1...M}$ & Host genotype matrix \\
+		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
+		$\delta$ & 0.2 & Host variant effect size \\ 
+		$M$ & 20 & Number of causal host variants \\ 
+		$\epsilon_i$ & $ \epsilon_i \sim Norm(0, \sigma^2_{\epsilon})$ & Environmental part of spVL for individual $i$ \\ 
+		$N$ & 500 & Number of simulated samples \\ \bottomrule 
 	\end{tabular}
-	\caption{Simulation model parameters.}
+	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
 	\label{tab:sim-params}
@@
 
-\subsection{Swiss HIV-1 data}
+\subsection*{Swiss HIV-1 data}
 
@@
 
-\subsection{POUMM parameter inference}
+\subsection*{POUMM parameter inference}
 
@@
 
-\subsection{Phylogenetic spVL correction}
+\subsection*{Phylogenetic spVL correction}
 We corrected calcualted spVL values using the method described in this paper.
@@
 
-\subsection{Association testing}
+\subsection*{Association testing}
 We performed two GWAS using the same human genotype data from the SHCS.
@@
 
-\section{Acknowledgments}
+\section*{Acknowledgments}
 
@@
 \newpage
-\section{Supplemental Material}
+\section*{Supplemental Material}
 \beginsupplement
 
-\subsection{Expected results from simulations}
-
-Here we show the root mean square error (RMSE) of $\bm{z} - \bm{\bar{z}}$ as an estimate for $\bm{h}$ should be $\approx$ 0.74 in our simulation setup. First we write the expression for the RMSE:
+\subsection*{Expected results from simulations}
+
+Here we show the root mean square error (RMSE) of $\bm{z} - \bm{\bar{z}}$ as an estimate for $\bm{h}$ should be $\approx$ 0.74 in our simulation scheme. First we write the expression for the RMSE:
 
@@
 
-We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{g_h}$, are fixed parameters in our simulation setup.
+We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{g_h}$, are fixed parameters in our simulation scheme.
 
@@
 	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
-	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {$H^2_{\bar{t}}$};
+	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {\textcolor{red}{$H^2_{\bar{t}}$}};
 	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
@@
 	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
-	\node[anchor = west] (blabel) at (-1, 4.5) {B. Viral part of spVL};
+	\node[anchor = west] (blabel) at (-1, 4.5) {B. Generating viral effects};
 	\node[constnode] (theta) at (0, 1.5) {$\theta$};
 	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
-	\node[constnode] (alpha) at (3.6, 1.5) {$\alpha$};
+	\node[constnode] (alpha) at (3.6, 1.5) {\textcolor{red}{$\alpha$}};
 	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
@@
 	\draw[taro] (varpic) -- (varv);
-	\node (vdist) at ($(v) + (1.8, 0.3)$) {Normal};
+	\node (vdist) at ($(v) + (1.8, 0.3)$) {Gaussian};
 	\draw (v) -- (vdist);
@@
 	\begin{scope}[xshift = 0cm, yshift = -9cm]
-	\node[anchor = west] (clabel) at (-1, 5.55) {C. Host part of spVL};
+	\node[anchor = west] (clabel) at (-1, 5.55) {C. Generating host effects};
 	\node[constnode] (d) at (1.8, 1.5) {2};
@@
 	\draw[taro] (varpic) -- (varh);
-	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Normal};
+	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Gaussian};
 	\draw (h) -- (hdist);
@@
 	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
-	\node[anchor = west] (dlabel) at (-1, 3) {D. Environmental part of spVL};
+	\node[anchor = west] (dlabel) at (-1, 3) {D. Generating environmental effects};
 	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
@@
 	\draw[taro] (varpic) -- (vare);
-	\node (edist) at ($(e) + (1.8, 0.3)$) {Normal};
+	\node (edist) at ($(e) + (1.8, 0.3)$) {Gaussian};
 	\draw (e) -- (edist);
@@
 	\end{tikzpicture}
-	\caption{A graphical representation of the simulation model.
-Variables in solid squares are constant in the simulation scheme, whereas variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
-Arrows represent dependencies among variables and the dashed square represents repetition \citep{Hohna2014}.}
+	\caption{A graphical model representation of our simulation scheme, following the recommendations in \cite{Hohna2014}.
+Variables in solid squares are constants, with the two master control variables that we vary from simulation to simulation highlighted in red.
+Variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
+Arrows represent dependencies among variables and the dashed square represents repetition.
+All parameters are defined in table \ref{tab:sim-params}, as well as the values or expressions used for them.
+(A) shows how the variance in the simulated environmental effect $\sigma^2_{\epsilon}$ is smaller if the master pathogen heritability value $H^2_{\bar{t}}$ is higher and vice-versa.
+(B) shows the OU parameters and the pathogen phylogeny, which generate the Gaussian-distributed pathogen effects.
+The OU parameters $\theta$ and $g_0$ are fixed, whereas $\sigma$ is a deterministic function of the variance in the pathogen effect and the value of $\alpha$.
+In other words, we use $\sigma$ to maintain the desired pathogen heritability while varying $\alpha$.
+We generate a new random phylogeny for each simulation.
+(C) shows how host genotypes are drawn to generate host effects.
+The host genotype matrix $G$ contains the number of copies (0, 1, or 2) for each of $M$ causal variants with effect size $\delta$.
+We assume half the variants have a positive effect and half have a negative effect.
+The allele frequency $p$ for the causal variants set so that we achieve the desired variance in the host effects.
+(D) shows that the environmental effect is drawn from a Gaussian distribution with mean zero and variance as determined in part (A).}
 	\label{fig:sim-design}

### 2021-11-18 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 	\begin{center}
-	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+	\includegraphics[width = \linewidth]{figures/simulation_setup.pdf}
 	\caption{A high-level schematic of our POUMM-based simulation framework.
@@
 (B) shows host genetic effects, which are the sum of independent effects from several causal variants.
-Each variant can be present in 0, 1, or 2 copies and has some effect $\delta$ on the trait value.
-We assume each variant has the same magnitude of effect and that half the causal variants have a positive effect and the other half have a negative effect.
-(C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero with variance $\sigma^2_e$.}
+Each variant can be present in 0, 1, or 2 copies and we assume half have effect size $\delta$ on the trait value and half have effect size $-\delta$.
+(C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero.}
 % 		\includegraphics[width=0.462\linewidth]{figures/model_figure}

### 2021-11-18 manuscript/hiv-gwas.tex

@@
-\documentclass[]{article}
+\documentclass[12pt]{article}
 \usepackage[margin=1in]{geometry}
@@
 \usepackage{amsmath}
-\usepackage{longtable} % for table that spans multiple pages
 \usepackage[singlelinecheck=false]{caption} % to left-justify single-line table captions
@@
 \newcolumntype{P}[1]{>{\centering\arraybackslash}p{#1}} % for centered, width-specified columns 
+\usepackage{makecell} % for multi-row table entries
+\usepackage{tabularx} % for tables that don't spill past margin
 
 \usepackage{url} % otherwise bibtex hates my references file
-\usepackage[round]{natbib}
-\bibliographystyle{plainnat}
+\usepackage[round,sort]{natbib}
 
@@
 
-% Set up supplemental figure numbering
-\newcommand{\beginsupplement}{%
-	\setcounter{table}{0}
-	\renewcommand{\thetable}{S\arabic{table}}%
-	\setcounter{figure}{0}
-	\renewcommand{\thefigure}{S\arabic{figure}}%
-}
-
 % For proof
@@
 
-% for colored rows in table
-\usepackage{color, colortbl}
-\definecolor{Gray}{gray}{0.9}
-
-% For multi-line table cells
-\newcommand{\specialcell}[2][c]{\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
+% Set up supplemental figure numbering
+\newcommand{\beginsupplement}{\setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}}
+
+\startsetups
+    \setcounter{figure}{0}
+    \renewcommand{\thefigure}{S\arabic{figure}}
+    \setcounter{table}{0}
+    \renewcommand{\thetable}{S\arabic{table}}
+\stopsetups
 
@@
 
+\begin{flushright}
+Submission type: Article\\
+Section: Methods
+\end{flushright}
+
 \begin{center}
-	\Large{A phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits}
+	\Large{\textbf{A phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits}}
 \end{center}
@@
 \begin{flushleft}
-	Sarah Nadeau,$^{1,2}$ Christian W.
-Thorball,$^{3}$ Roger Kouyos,$^{4,5}$ Huldrych F.
-Günthard,$^{4,5}$ Jürg Böni,$^{4}$ Sabine Yerly,$^6$ Matthieu Perreau,$^{7}$ Thomas Klimkait,$^{8}$ Andri Rauch,$^{9}$ Hans H.
-Hirsch,$^{8,10,11}$ Matthias Cavassini,$^{12}$ Pietro Vernazza,$^{13}$ Enos Bernasconi,$^{14}$ Jacques Fellay,$^{2,3,15}$ Venelin Mitov,$^{\dagger,1,2}$ Tanja Stadler,$^{\dagger,\ast,1,2}$ and the Swiss HIV Cohort Study (SHCS)
+	Sarah Nadeau$^{1,2}$, Christian W.
+Thorball$^{3}$, Roger Kouyos$^{4,5}$, Huldrych F.
+Günthard$^{4,5}$, Jürg Böni$^{4}$, Sabine Yerly$^6$, Matthieu Perreau$^{7}$, Thomas Klimkait$^{8}$, Andri Rauch$^{9}$, Hans H.
+Hirsch$^{8,10,11}$, Matthias Cavassini$^{12}$, Pietro Vernazza$^{13}$, Enos Bernasconi$^{14}$, Jacques Fellay$^{2,3,15}$, Venelin Mitov$^{\dagger,1,2}$, Tanja Stadler$^{\dagger,\ast,1,2}$, and the Swiss HIV Cohort Study (SHCS)
 \end{flushleft}
@@
 
+\newpage
 \begin{doublespace}
@@
 Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
-However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021}.
+However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}.
 If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected.
@@
 (C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero.}
-% 		\includegraphics[width=0.462\linewidth]{figures/model_figure}
-% 		\caption{The POUMM statistical model for the spVL trait.
-(A) shows how the pathogen effect on spVL, $g_v$, evolves according to an OU process along (C) the pathogen phylogeny.
-The dashed line in (A) shows the expected value of the trait at each point in time under the model.
-(B) shows that the non-pathogen effect on spVL, $e$, is modeled as a Gaussian random variable.
-Each tip on the phylogeny corresponds to a sampled individual with an independent non-pathogen-effect drawn from the distribution shown in (B).}
 		\label{fig:spVLModel}
@@
 To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data.
-Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and table \ref{tab:sim-params} gives the value or expression for each parameter.
+Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
 In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength.
@@
 	\begin{center}
-		\includegraphics[width=0.75\linewidth]{figures/simulation_results}
+		\includegraphics[width=\linewidth]{figures/simulation_results}
 		\caption{Results from the simulation study.
@@
 We followed the framework outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
-Figure \ref{fig:spvl-calc-comparison} shows the calculated spVL values, figure \ref{fig:spvl-on-tree} shows the inferred pathogen phylogeny, and figures \ref{fig:spvl-on-tree}, \ref{fig:trait-comparison} show the estimated non-pathogen effects on spVL calculated for these individuals.
+Figure \ref{fig:spvl-calc-comparison} shows the calculated spVL values, Figure \ref{fig:spvl-on-tree} shows the inferred pathogen phylogeny, and figures \ref{fig:spvl-on-tree}, \ref{fig:trait-comparison} show the estimated non-pathogen effects on spVL calculated for these individuals.
 We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) \ref{tab:POUMM-parameters-our-estimates}.
-To put these values into the context of our simulation study, they are shown as black points on figure \ref{fig:simulationResults}.
+To put these values into the context of our simulation study, they are shown as black points on Figure \ref{fig:simulationResults}.
 
@@
 Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS.
-Q-Q plots show the distribution of p-values are very similar as well (figure \ref{fig:qq-plots}).
+Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}).
 Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions.
@@
 Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
-However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (figure \ref{tab:gwas-top-associations}).
-Finally table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}.
+However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Figure \ref{tab:gwas-top-associations}).
+Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}.
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
@@
 	\begin{center}
-		\includegraphics[width=0.7\linewidth]{figures/gwas_results}
+		\includegraphics[width=\linewidth]{figures/gwas_results}
 		\caption{Results from comparative GWAS.
@@
 The ``Standard trait value'' and ``Estimated non-pathogen part of trait'' results are from using two different GWAS response variables with the same human genotype data.}
-\begin{tabular}{p{1cm}p{1.8cm}p{2.4cm}p{1.7cm}p{1.8cm}p{1.7cm}p{1.8cm}} % {lrrrrrr} {llclcccc}
-\hline
-& & McLaren et al.  & \multicolumn{2}{l}{Standard trait value} & \multicolumn{2}{l}{Estimated non-pathogen part of trait} \\ 
+\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
+\hline
+& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
   \hline
@@
    \hline
-\end{tabular}
+\end{tabularx}
 \label{tab:comp-gwas-mclaren-snps}
@@
 
-Given our estimates for the heritability of spVL and the selection strength on this trait using Swiss HIV cohort data, our simulation results reveal that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (figure \ref{fig:simulationResults}).
+Given our estimates for the heritability of spVL and the selection strength on this trait using Swiss HIV cohort data, our simulation results reveal that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (Figure \ref{fig:simulationResults}).
 Indeed the results of our  GWAS method presented here and a standard GWAS are very similar.
@@
 $g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
-Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see table \ref{tab:sim-params}).
+Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see Table \ref{tab:sim-params}).
 We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}.
 $g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
-For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (table \ref{tab:POUMM-parameters-our-estimates}).
+For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (Table \ref{tab:POUMM-parameters-our-estimates}).
 To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively.
@@
 \begin{table}[H]
-	\begin{tabular}{p{1.5cm}ll} \toprule 
+	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
+	\begin{tabularx}{\linewidth}{p{1.5cm}ll} \toprule 
 		Variable & Expression & Definition \\ \midrule 
@@
 		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Viral phylogeny \\ 
-		$g_{h_i}$ & $g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & Host part of spVL for individual $i$ \\ 
-		$G_{N \times M}$ & $G_{ij} \sim Binom(2, p)\ \forall i \in {1...N}, \forall j \in {1...M}$ & Host genotype matrix \\
+		$g_{h_i}$ & \makecell{$g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \\ \delta \sum_{j = M/2}^{j = M}{G_{ij}}$} & Host part of spVL for individual $i$ \\ 
+		$G_{N \times M}$ & \makecell{$G_{ij} \sim Binom(2, p)\ \\ \forall i \in {1...N}, \forall j \in {1...M}$} & Host genotype matrix \\
 		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
@@
 		$N$ & 500 & Number of simulated samples \\ \bottomrule 
-	\end{tabular}
-	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
+	\end{tabularx}
 	\label{tab:sim-params}
@@
 
-Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, www.shcs.ch).
-All participants were HIV-1–infected individuals at least 16 years old and written informed consent was obtained from all cohort participants by \citet{Schoeni-Affolter2010}.
-The data was received in anonymized form after the project was approved by the SHCS.
+Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, www.shcs.ch, \citet{Schoeni-Affolter2010, Scherrer2021CohortSHCS}).
+All participants were HIV-1–infected individuals at least 16 years old and written informed consent was obtained from all cohort participants.
+The anonymized data was made available for this study after a study proposal was approved by the SHCS.
 
@@
 
-This study was funded by ETH Zurich.
-We would like to thank Esposito Mauro and Alexandra Scherrer from the SHCS data center, who were especially helpful in retrieving data, Alexandra Calmy and her team from the Geneva University Hospital for data contributions, and Michael Landis, who shared a LaTeX template for graphical model drawing.
-Finally and most importantly, we thank the participants of the Swiss HIV Cohort Study for contributing their samples and making this research possible.
-
+This work was supported by ETH Zurich.
+We thank the patients who participate in the SHCS; the physicians and study nurses for excellent patient care; A.
+Scherrer,  E.
+Mauro, and K.
+Kusejko from the SHCS Data Centre for data management; and D.
+Perraudin and M.
+Amstad for administrative assistance.
+We also thank Michael Landis, who shared a LaTeX template for graphical model drawing.
+
+The members of the SHCS are:
+Abela I, Aebi-Popp K, Anagnostopoulos A, Battegay M, Bernasconi E, Braun DL, Bucher HC, Calmy A, Cavassini M, Ciuffi A, Dollenmaier G, Egger M, Elzi L, Fehr J, Fellay J, Furrer H, Fux CA, Günthard HF (President of the SHCS), Hachfeld A, Haerry D (deputy of "Positive Council"), Hasse B, Hirsch HH, Hoffmann M, Hösli I, Huber M, Kahlert CR (Chairman of the Mother & Child Substudy), Kaiser L, Keiser O, Klimkait T, Kouyos RD, Kovari H, Kusejko K (Head of Data Centre), Martinetti G, Martinez de Tejada B, Marzolini C, Metzner KJ, Müller N, Nemeth J, Nicca D, Paioni P, Pantaleo G, Perreau M, Rauch A (Chairman of the Scientific Board), Schmid P, Speck R, Stöckle M (Chairman of the Clinical and Laboratory Committee), Tarr P, Trkola A, Wandeler G, Yerly S.
+
+The Swiss HIV Cohort Study is supported by the Swiss National Science Foundation (grant #201369), by SHCS project #858 and by the SHCS research foundation .
+Furthermore, the SHCS drug resistance database is supported by the Yvonne Jacob Foundation (to HFG).
+The data are gathered by the Five Swiss University Hospitals, two Cantonal Hospitals, 15 affiliated hospitals and 36 private physicians (listed in http://www.shcs.ch/180-health-care-providers).
+
+\bibliographystyle{manuscript/my_natbib}
 \bibliography{manuscript/references_synced}
@@
 \newpage
-\section*{Supplemental Material}
+\startsection
+\section*{Supplemental Material}\label{supp}
 \beginsupplement
+\setcounter{page}{1}
 
@@
 	\begin{scope}[xshift = 0cm, yshift = 0cm]
-	\node[anchor = west] (alabel) at (-1, 4) {A. Variance partitioning};
+	\node[anchor = west] (alabel) at (-1, 4) {\textbf{A} Variance partitioning};
 	\node[constnode] (H2h) at (0, 0) {0.25};
@@
 	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
-	\node[anchor = west] (blabel) at (-1, 4.5) {B. Generating viral effects};
+	\node[anchor = west] (blabel) at (-1, 4.5) {\textbf{B} Generating viral effects};
 	\node[constnode] (theta) at (0, 1.5) {$\theta$};
@@
 	% Host 
-	\begin{scope}[xshift = 0cm, yshift = -9cm]
-	\node[anchor = west] (clabel) at (-1, 5.55) {C. Generating host effects};
+	\begin{scope}[xshift = 0cm, yshift = -7.5cm]
+	\node[anchor = west] (clabel) at (-1, 5.55) {\textbf{C} Generating host effects};
 	\node[constnode] (d) at (1.8, 1.5) {2};
@@
 	% Environment
-	\begin{scope}[xshift = 10cm, yshift = -6.5cm]
-	\node[anchor = west] (dlabel) at (-1, 3) {D. Generating environmental effects};
+	\begin{scope}[xshift = 9cm, yshift = -6.5cm]
+	\node[anchor = west] (dlabel) at (-1, 3) {\textbf{D} Generating environmental effects};
 	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
@@
 Arrows represent dependencies among variables and the dashed square represents repetition.
-All parameters are defined in table \ref{tab:sim-params}, as well as the values or expressions used for them.
+All parameters are defined in Table \ref{tab:sim-params}, as well as the values or expressions used for them.
 (A) shows how the variance in the simulated environmental effect $\sigma^2_{\epsilon}$ is smaller if the master pathogen heritability value $H^2_{\bar{t}}$ is higher and vice-versa.
@@
 \caption{POUMM parameter estimates for spVL based on Swiss data. HPD = Highest posterior density.}
-	\begin{tabular}{lccc} \hline 
+	\begin{tabularx}{\linewidth}{XXX} \hline 
 	Parameter & Posterior mean & 95\% HPD \\ \hline 
@@
 	$H^2_{\bar{t}}$ & 0.45 & (0.24, 0.67) \\ \hline
-	\end{tabular}
+	\end{tabularx}
 	\label{tab:POUMM-parameters-our-estimates}
@@
 \caption{POUMM parameter estimates for spVL from previous studies.}
-	\begin{tabular}{lcccp{0.35\linewidth}}
-	\hline 
-	Parameter & Value & Uncertainty & Reference & Notes \\
-	\hline 
-	$v_0$ & 5.54 & (4.04 - 7.25) HPD & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
-	$\theta$ & 4.45 & (4.41 - 4.49) HPD & \citet{Mitov2018} & \\
-	$\theta$ &  4.0 & (1.6 - 4.3) 95\% CI & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
-	$\theta$ & 4.1 & (3.5 - 4.9) 95\% CI & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
-	$\alpha$ & 28.78 & (16.64 - 46.93) HPD & \citet{Mitov2018} & \\
-	$\alpha$ & 32.7 & (0.03 - 57.6) 95\% CI & \citet{Bertels2018} & \\
-	$\alpha$ & 7.6 & (1.2 - 10) 95\% CI & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
-	$\sigma$ & 2.97 & (1.95 - 4.37) HPD & \citet{Mitov2018} & \\
-	$\sigma$ & 1.3 & (0.66 -1.87) 95\% CI &  \citet{Blanquart2017} & \\
-	$\sigma_e$ & 0.77 & (0.73, 0.8) HPD & \citet{Mitov2018} & \\
-	$\sigma_e$ & 0.61 & (0.54, 0.65) 95\% CI &  \citet{Blanquart2017} & \\
-	\hline 
-	\end{tabular}
+	\begin{tabularx}{\linewidth}{llXXX}
+	\hline 
+	Parameter & Value (Uncertainty) & Reference & Notes \\
+	\hline 
+	$v_0$ & 5.54 (4.04 - 7.25) & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
+	$\theta$ & 4.45 (4.41 - 4.49) & \citet{Mitov2018} & \\
+	$\theta$ &  4.0 (1.6 - 4.)) & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
+	$\theta$ & 4.1 (3.5 - 4.9) & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
+	$\alpha$ & 28.78 (16.64 - 46.93) & \citet{Mitov2018} & \\
+	$\alpha$ & 32.7 (0.03 - 57.6) & \citet{Bertels2018} & \\
+	$\alpha$ & 7.6 (1.2 - 10) & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
+	$\sigma$ & 2.97 (1.95 - 4.37) & \citet{Mitov2018} & \\
+	$\sigma$ & 1.3 (0.66 -1.87) &  \citet{Blanquart2017} & \\
+	$\sigma_e$ & 0.77 (0.73, 0.8) & \citet{Mitov2018} & \\
+	$\sigma_e$ & 0.61 (0.54, 0.65) &  \citet{Blanquart2017} & \\
+	\hline 
+	\end{tabularx}
 	\label{tab:POUMMparams-prev-estimated}
@@
 Values from \protect\citep{Blanquart2017, Mitov2018} are empirical; values from \protect\citep{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
-	\begin{tabular}{lcc} \hline 
+	\begin{tabularx}{\linewidth}{XXX} \hline 
 	Measurement & Value & Reference \\ \hline 
@@
 	$\sigma^2_z$ & $\approx$ 0.5 & \citet{Bonhoeffer2015} \\ \hline 
-	\end{tabular}
+	\end{tabularx}
 	\label{tab:spVLmeasurements}
@@
 \caption{Number of samples for GWAS after sequential filtering steps.}
-    \begin{tabular}{lc}
+    \begin{tabularx}{\linewidth}{XX}
     \hline
@@
     \hline
-    \end{tabular}
+    \end{tabularx}
     \label{tab:sample-filtering}
@@
 \caption{Number of variants for GWAS after sequential filtering steps.}
-	\begin{tabular}{lc}
+	\begin{tabularx}{\linewidth}{XX}
 	\hline
@@
 	Missing genotype rate $>$ 0.05 & 11590002 \\
-	Hardy-Weinburg exact test p-value $<$ 5x10$^-5$ & 11589246 \\
+	Hardy-Weinburg exact test p-value $< 5 \times 10^{-5}$ & 11589246 \\
 	Minor allele frequency $<$ 0.01 & 6228626 \\ \hline			
-	\end{tabular}
+	\end{tabularx}
 	\label{tab:variant-filtering}
@@
 ``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with estimated non-pathogen effects on spVL.}
-\begin{tabular}{lllP{2.2cm}P{2.2cm}P{1.6cm}P{2.2cm}}
+\begin{tabularx}{\linewidth}{XXXXXXX}
   \hline
@@
    \hline
-\end{tabular}
+\end{tabularx}
 \label{tab:gwas-top-associations}
@@
 
-% \begin{longtable}[H]{lp{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}{lrrrrrr}
-% 		\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citep{McLaren2015} compared to results from the current study.}
-% 		\hline
-% 		& \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS} & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
-% 		\hline
-% 	 	Allele & Beta & p-value & Beta &p-value & Beta& p-value \\ 
-% 		\hline
-% 		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
-% 		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
-% 		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
-% 		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
-% 		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
-% 		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
-% 		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
-% 		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
-% 		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
-% 		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
-% 		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
-% 		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\ 
-% 		HLA-C*02:02 & -0.223 & $5.45 \times 10^{-8}$ & -0.089 & $1.31 \times 10^{-1}$ & -0.039 & $2.50 \times 10^{-1}$ \\ 
-% 		HLA-A*25:01 & -0.301 & $6.80 \times 10^{-8}$ & -0.090 & $9.93 \times 10^{-2}$ & -0.033 & $2.96 \times 10^{-1}$ \\ 
-% 		HLA-B*52:01 & -0.435 & $2.60 \times 10^{-7}$ & 0.238 & $7.63 \times 10^{-2}$ & 0.150 & $5.33 \times 10^{-2}$ \\ 
-% 		HLA-C*01:02 & -0.231 & $2.22 \times 10^{-6}$ & -0.198 & $5.07 \times 10^{-4}$ & -0.113 & $5.70 \times 10^{-4}$ \\ 
-% 		HLA-A*23:01 & 0.269 & $1.66 \times 10^{-5}$ & 0.020 & $7.42 \times 10^{-1}$ & 0.015 & $6.59 \times 10^{-1}$ \\ 
-% 		HLA-B*55:01 & 0.302 & $2.32 \times 10^{-5}$ & 0.685 & $2.48 \times 10^{-1}$ & 0.436 & $2.03 \times 10^{-1}$ \\ 
-% 		HLA-B*35:02 & 0.358 & $2.42 \times 10^{-5}$ &  &  &  &  \\ 
-% 		HLA-A*31:01 & -0.219 & $4.36 \times 10^{-5}$ & -0.052 & $4.50 \times 10^{-1}$ & -0.017 & $6.78 \times 10^{-1}$ \\ 
-% 		HLA-B*35:01 & 0.156 & $4.82 \times 10^{-5}$ & 0.167 & $2.13 \times 10^{-4}$ & 0.083 & $1.33 \times 10^{-3}$ \\ 
-% 		HLA-B*49:01 & 0.253 & $1.86 \times 10^{-4}$ & 2.173 & $9.79 \times 10^{-4}$ & 1.253 & $9.82 \times 10^{-4}$ \\ 
-% 		HLA-B*35:03 & 0.215 & $7.13 \times 10^{-4}$ & -0.241 & $8.00 \times 10^{-1}$ & 0.261 & $6.35 \times 10^{-1}$ \\ 
-% 		HLA-B*40:01 & 0.148 & $9.37 \times 10^{-4}$ & 0.519 & $1.06 \times 10^{-2}$ & 0.349 & $2.91 \times 10^{-3}$ \\ 
-% 		HLA-B*18:01 & 0.136 & $9.48 \times 10^{-4}$ & 0.208 & $1.73 \times 10^{-3}$ & 0.133 & $5.09 \times 10^{-4}$ \\ 
-% 		HLA-A*29:02 & 0.167 & $1.40 \times 10^{-3}$ & 0.150 & $6.17 \times 10^{-2}$ & 0.092 & $4.72 \times 10^{-2}$ \\ 
-% 		HLA-C*14:02 & -0.228 & $1.50 \times 10^{-3}$ & 0.266 & $7.59 \times 10^{-1}$ & 0.235 & $6.38 \times 10^{-1}$ \\ 
-% 		HLA-A*11:01 & -0.109 & $2.70 \times 10^{-3}$ & -0.090 & $8.96 \times 10^{-2}$ & -0.052 & $8.92 \times 10^{-2}$ \\ 
-% 		HLA-A*24:02 & 0.094 & $3.60 \times 10^{-3}$ & 0.133 & $3.79 \times 10^{-3}$ & 0.074 & $5.52 \times 10^{-3}$ \\ 
-% 		HLA-B*37:01 & 0.222 & $4.50 \times 10^{-3}$ & 0.600 & $3.22 \times 10^{-3}$ & 0.311 & $8.11 \times 10^{-3}$ \\ 
-% 		HLA-B*44:03 & 0.119 & $5.00 \times 10^{-3}$ & -0.114 & $2.77 \times 10^{-1}$ & -0.080 & $1.88 \times 10^{-1}$ \\ 
-% 		HLA-C*03:03 & 0.110 & $6.50 \times 10^{-3}$ & -0.078 & $1.86 \times 10^{-1}$ & -0.031 & $3.66 \times 10^{-1}$ \\ 
-% 		HLA-B*58:01 & -0.253 & $7.10 \times 10^{-3}$ & 6.895 & $7.93 \times 10^{-2}$ & 4.841 & $3.27 \times 10^{-2}$ \\ 
-% 		HLA-A*32:01 & -0.123 & $8.30 \times 10^{-3}$ & 0.089 & $5.31 \times 10^{-1}$ & 0.071 & $3.86 \times 10^{-1}$ \\ 
-% 		HLA-C*16:01 & 0.137 & $8.60 \times 10^{-3}$ & 0.057 & $4.46 \times 10^{-1}$ & 0.013 & $7.60 \times 10^{-1}$ \\ 
-% 		HLA-C*03:04 & 0.093 & $1.25 \times 10^{-2}$ & 0.126 & $3.52 \times 10^{-2}$ & 0.095 & $5.77 \times 10^{-3}$ \\ 
-% 		HLA-B*38:01 & -0.148 & $1.35 \times 10^{-2}$ & -3.459 & $3.01 \times 10^{-1}$ & -2.224 & $2.49 \times 10^{-1}$ \\ 
-% 		HLA-A*33:01 & -0.233 & $1.45 \times 10^{-2}$ &  &  &  &  \\ 
-% 		HLA-A*30:01 & -0.159 & $2.89 \times 10^{-2}$ & 0.072 & $9.04 \times 10^{-1}$ & 0.008 & $9.82 \times 10^{-1}$ \\ 
-% 		HLA-C*12:03 & -0.077 & $4.59 \times 10^{-2}$ & -0.114 & $2.95 \times 10^{-2}$ & -0.067 & $2.68 \times 10^{-2}$ \\ 
-% 		HLA-A*01:01 & 0.051 & $4.78 \times 10^{-2}$ & -0.039 & $4.88 \times 10^{-1}$ & -0.050 & $1.24 \times 10^{-1}$ \\ 
-% 		HLA-B*14:01 & -0.187 & $5.51 \times 10^{-2}$ & 7.071 & $5.00 \times 10^{-2}$ & 3.974 & $5.62 \times 10^{-2}$ \\ 
-% 		HLA-B*40:02 & -0.130 & $1.01 \times 10^{-1}$ & -0.343 & $3.88 \times 10^{-1}$ & -0.105 & $6.47 \times 10^{-1}$ \\ 
-% 		HLA-C*07:04 & 0.125 & $1.02 \times 10^{-1}$ & 0.111 & $3.00 \times 10^{-1}$ & 0.080 & $1.98 \times 10^{-1}$ \\ 
-% 		HLA-C*05:01 & 0.043 & $1.93 \times 10^{-1}$ & -0.022 & $6.51 \times 10^{-1}$ & -0.019 & $4.82 \times 10^{-1}$ \\ 
-% 		HLA-A*03:01 & 0.025 & $3.29 \times 10^{-1}$ & 0.002 & $9.51 \times 10^{-1}$ & -0.001 & $9.81 \times 10^{-1}$ \\ 
-% 		HLA-B*51:01 & -0.036 & $3.52 \times 10^{-1}$ & -0.046 & $3.13 \times 10^{-1}$ & -0.044 & $1.01 \times 10^{-1}$ \\ 
-% 		HLA-A*68:01 & 0.046 & $3.71 \times 10^{-1}$ & 0.036 & $6.09 \times 10^{-1}$ & 0.009 & $8.27 \times 10^{-1}$ \\ 
-% 		HLA-B*44:02 & 0.030 & $3.73 \times 10^{-1}$ & -0.051 & $3.90 \times 10^{-1}$ & -0.031 & $3.57 \times 10^{-1}$ \\ 
-% 		HLA-B*39:01 & 0.073 & $3.85 \times 10^{-1}$ &  &  &  &  \\ 
-% 		HLA-A*02:01 & 0.017 & $3.94 \times 10^{-1}$ & -0.018 & $5.57 \times 10^{-1}$ & -0.008 & $6.77 \times 10^{-1}$ \\ 
-% 		HLA-C*15:02 & -0.046 & $4.20 \times 10^{-1}$ & 0.064 & $3.26 \times 10^{-1}$ & 0.033 & $3.78 \times 10^{-1}$ \\ 
-% 		HLA-B*50:01 & 0.066 & $4.68 \times 10^{-1}$ & 1.522 & $1.77 \times 10^{-2}$ & 0.970 & $8.73 \times 10^{-3}$ \\ 
-% 		HLA-B*15:01 & -0.021 & $5.71 \times 10^{-1}$ & -0.073 & $1.71 \times 10^{-1}$ & -0.021 & $4.93 \times 10^{-1}$ \\ 
-% 		HLA-A*30:02 & 0.050 & $5.92 \times 10^{-1}$ & 0.047 & $9.28 \times 10^{-1}$ & -0.004 & $9.89 \times 10^{-1}$ \\ 
-% 		HLA-A*26:01 & -0.003 & $9.57 \times 10^{-1}$ & -0.645 & $1.96 \times 10^{-1}$ & -0.529 & $6.59 \times 10^{-2}$ \\
-% 		\hline
-% 		\label{tab:full-mclaren-comparison}
-% \end{longtable}
-
-% \begin{figure}[H]
-% 	\includegraphics[width=\linewidth]{images/HIV_tree_branch_lengths}
-% 	\caption{Left plot: branch length distributions for UK HIV tree and our simulated HIV tree prior to pruning down to 500 tips.
-The mean root-to-tip distance for the UK tree is 0.14 and the mean branch length is 0.02.
-The mean root-to-tip distance for the full simulated HIV tree is 0.14 and the mean branch length is 0.01.
-Right plot: branch length distribution for the tree of 500 tips used for simulation.
-The mean root-to-tip distance for the pruned HIV tree is 0.14 and the mean branch length is 0.03.}
-% 	\label{fig:branchlengths}
-% \end{figure}
-
-\typeout{}
-
+\stopsection
 \end{document}
-
-%\citep{Mitov2018} and \citep{Bertels2018} both found higher statistical support for the POUMM model of spVL evolution as opposed to the corresponding model assuming Brownian motion only.
-This makes sense in the context of the hypothesis proposed by \citep{Fraser2014}, in which spVL is under stabilizing selection to maximize transmissibility of the virus.
-On the other hand, the assumptions that $e$ are normally distributed, mutually indpendent, and have an expected value of 0 are consistent with the idea that genetic effects from several unlinked host loci and environmental effects from many independent factors are each equally likely to increase an individual's spVL as decrease it.
-\todo{This statement is not supported}
-%
-%\todo{TBD once we have real data, here are some thoughts}
-%The success of our method in identifying new host determinants of spVL depends on how much host-determined variation in spVL actually exists.
-In the largest GWAS for HIV spVL to date, \citep{McLaren2015} estimated that additive effects from common human genetic variants explain 20\% of the variability in spVL.
-Of this quantity, 94.5\% was attributable to variants in just two regions in the human genome.
-\todo{Were there variants other than HLA in the MHC region?} Associations were with alleles in the CCR5 gene, which encodes a co-receptor for HIV entry into CD4+ T-cells, and the human leukocyte antigen (HLA) genes, which encode proteins that present certain host and pathogen-derived peptides at the surface of cells.
-These associations have been well-replicated in several GWAS \citep{Naranbhai2017}.
-This results suggest that nearly all host-determined variation in spVL can be explained by previously discovered variants.
-%
-%In a joint analysis of both host and viral genotypes, \citep{Bartha2017} showed that nearly all of the variation in spVL explained by human genetic variation overlaps that explained by HIV genetic variation.
-If this is the case then our method to remove the viral contribution to spVL prior to GWAS should yield only associations between host variants and spVL that are independent of the virus.
-Additionally, effect size estimates for associations will reflect only the effect of the host variant, not any correlated effect from the virus.
-
-%\subsection{New method to estimate the viral and non-viral parts of spVL}
-%
-%Here we derive a method to estimate the viral part of spVL for each infected host by assuming that spVL values at the tips of a phylogenetic tree $\mathcal{T}$ have evolved under the POUMM.
-In particular, for every tip $i$ in the tree, we assume that the host’s spVL value, $z_i$, is formed as a sum of three independent parts: a  heritable part determined by additive genetic effects from the virus, $g_{v_i}$, a non-heritable part determined by additive genetic effects from the host, $g_{h_i}$, and a further non-heritable part, $\epsilon_i$, which encompasses any other factor effecting spVL (Equation \ref{eq:totalspVL}).
-For the purposes of our derivation, it is convent to combine the non-heritable parts of spVL, $g_{h_i} + \epsilon_i$, into a single variable $e_i$.
-%
-%\begin{align} 
-%z _i &= g_{v_i} + g_{h_i} + \epsilon_i \\
-%\xrightarrow{e_i = g_{h_i} + \epsilon_i } z_i &= g_{v_i} + e_i  \label{eq:totalspVL}
-%\end{align}
-%
-%According to the POUMM, the unobserved viral component of spVL for each host, $g_{v_i}$, evolves along the tree according to an Ornstein-Uhlenbeck (OU) process.
-The OU process was first introduced by \citep{Lande1976} as a model of evolution under both drift and stabilizing selection, so we are assuming $g_{v_i}$ continually evolves towards an evolutionary optimum in each viral lineage but is subject to random evolutionary fluctuations.
-Mathematically, $g_{v_i}$ is a realization of a Gaussian distributed random variable $G_{v_i} \sim \mathcal{N}(\mu_{OU_i}, \sigma^2_{OU_i})$, where:
-%
-%\begin{equation} \label{eq:muOU}
-%\mu_{OU_i} = z_0e^{-\alpha t_{0i}} + (1 - e^{-\alpha t_{0i}})\theta \\
-%\end{equation}
-%\begin{equation}\label{eq:sigOU}
-%	\sigma^2_{OU_i} = \sigma^2\frac{(1 - e^{-2\alpha t_{0i}})}{2\alpha}
-%\end{equation}
-%
-%$z_0$ is the ancestral spVL value at the root of the tree, $\theta$ is the optimal spVL value that maximizes viral fitness,  $\alpha$ is the strength of selection on spVL, and $\sigma$ is the standard deviation of random fluctuations in $g_{v}$.
-We refer to these OU parameters ($z_0, \theta, \alpha, \textrm{and } \sigma$) collectively as $\Theta$.
-Finally, $t_{0i}$ is the phylogenetic distance from the root of the tree to tip $i$.
-We use the following notation to denote that the distribution of $G_{v}$ depends on the tree, $\mathcal{T}$, and the OU parameters $\Theta$: $G_{v|\Theta, \mathcal{T}}$.
-Generalizing to $N$ total tips in the tree, $\vec{G}_{v|\Theta, \mathcal{T}} = (G_{v_1}, ..., G_{v_N})^T$ is a random vector for the unobserved viral component of spVL at the tips.
-We know that $\vec{G}_{v|\Theta, \mathcal{T}} \sim \mathcal{N}(\vec{\mu}_{OU}, \mathbf{\Sigma}_{OU})$, where the elements of the variance-covariance matrix $\mathbf{\Sigma}_{OU}$ are calculated as follows:
-%
-%\begin{align} 
-%\bm{\Sigma}_{OU_{ii}} &= \sigma^2\frac{(1 - e^{-2\alpha t_{0i}})}{2\alpha} \label{eq:SigmaOUii} \\
-%\bm{\Sigma}_{OU_{ij}}&= \sigma^2\frac{e^{-\alpha\tau_{ij}}(1 - e^{-2\alpha t_{0(ij)}})}{2\alpha} \label{eq:SigmaOUij}
-%\end{align}
-%
-%$\tau_{ij}$ is the phylogenetic distance between tips $i$ and $j$ and $t_{0(ij)}$ is the phylogenetic distance from the root of the tree to the most recent common ancestor of $i$ and $j$.
-Equations \ref{eq:muOU}, \ref{eq:SigmaOUii}, and \ref{eq:SigmaOUij} completely describe the generating process for the unobserved viral part of spVL under the assumptions of the POUMM.
-%
-%On the other hand, the non-heritable component of spVL for each host, $e_i$ is assumed to be normally distributed in the population with an expected value of zero.
-Furthermore, $e_i$ is uncorrelated among hosts and independent of the heritable viral part of spVL.
-Mathematically, $e_i$, is a realization of another Gaussian distributed random variable $E_i \sim \mathcal{N}(0, \sigma^2_e)$, which is independent of $G_{v_j}$ for any tip $j$.
-$\vec{E}_{|\sigma_e} = (E_1, ..., E_N)^T$ is a random vector for the unobserved non-viral part of spVL at the tips.
-The parameter $\sigma_e$ is the standard deviation of the non-viral component of spVL in the population.
-% 
-% \begin{theorem}\label{thmGValues}
-% 	Given the POUMM assumptions described, the pair of random vectors $<\vec{G}_{|\Theta,\mathcal{T},\sigma_e, \vec{z}},\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} >$ satisfies the following:
-% 	\begin{enumerate}
-% 		\renewcommand{\theenumi}{\alph{enumi}}
-% 		\item Each realisation $\vec{g_v}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ of $\vec{G_v}_{|\Theta,\mathcal{T},\sigma_e, \vec{z}}$ is also a realization of $\vec{G}_{|\Theta, \mathcal{T}}$, i.e.
-it results from simulating the OU branching process defined by $\Theta$ on $\mathcal{T}$;
-% 		\item $\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} = \vec{z} - \vec{G_v}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$, i.e.
-each realization $\vec{e}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ of $\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$  satisfies $\vec{z} = \vec{g}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} + \vec{e}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ for its corresponding realization $\vec{g}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$  of $\vec{G}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}$ in the same pair.
-% 	\end{enumerate}
-% 	Then the pdf of $<\vec{G}_{|\Theta,\mathcal{T},\sigma_e, \vec{z}},\vec{E}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}} >$ is proportional to a $N$-variate Gaussian function, i.e. 
-% 	\begin{equation}\label{eqPdfGValues1}
-% 	f\big(<\vec{g}_{|\\Theta,\mathcal{T},\sigma_e, \vec{z}}, \vec{e}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}>\big) \propto \mathcal{N}\big(\vec{g}_{|\Theta,\mathcal{T}, \sigma_e, \vec{z}}\ ;\ \vec{\mu}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z}), \mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z})\big),
-% 	\end{equation}
-% 	\noindent where the matrix $\mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z})$ and the  vector $\vec{\mu}_{MWA}(\Theta,\mathcal{T}, \sigma_e, \vec{z})$ are defined as follows:
-% 	\begin{equation}\label{eqMuSigmaGValues}
-% 	\begin{array}{rcl}
-% 	\mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T},\sigma_e) & = & \bigg(\big[\mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\big]^{-1}+\mathbf{\Sigma}_{e}(\sigma_e)^{-1}\bigg)^{-1}\\
-% 	\vec{\mu}_{MWA}(\Theta,\mathcal{T},\sigma_e, \vec{z}) & = & \mathbf{\Sigma}_{MWA}(\Theta,\mathcal{T},\sigma_e) \bigg(\big[\mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\big]^{-1}\vec{\mu}_{OU}(\Theta,\mathcal{T}) +  \mathbf{\Sigma}_{e}(\sigma_e)^{-1} \vec{z}\bigg).
-% 	\end{array}
-% 	\end{equation}
-% \end{theorem}
-% \begin{proof}
-% 	\begin{equation}\label{eqPdfGValues2}
-% 	\begin{array}{rcl}
-% 	f\big(<\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}},\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}>\big) & = & f\big(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ |\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big) \times f\big(\vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big)\\
-% 	& = &\mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
-% 	& &  \mathcal{N}\bigg(\vec{z} - \vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{0}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
-% 	& = & \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
-% 	& &  \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\  \vec{z}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
-% 	\end{array}
-% 	\end{equation}
-% 	
-% 	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
-% 	
-% \end{proof}
-%
-%Removing the viral part of spVL from total spVL using the above method results in the following ordinary least squares slope and intercept coefficient estimates for GWAS:
-%
-%\begin{equation} \label{eq: OLS-soln}
-%	\hat{\beta}_i = (\mathbf{X}_i^T\mathbf{X}_i)^{-1}\mathbf{X}_i^T(\vec{z} - \vec{\mu}_{MWA})
-%\end{equation}
-%
-%where $\mathbf{X}_i$ is the $N \times 2$ design matrix of host genotypes for genetic variant $i$.
-%\begin{equation*}
-%	\mathbf{X}_i = \begin{bmatrix}
-%	1 & g_{i1}  \\
-%	\vdots & \vdots \\
-%	1 & g_{iN} 
-%	\end{bmatrix}
-%\end{equation*}
-%
-%%  serves three purposes.
-1) We can estimate for each patient the effect the virus has on their spVL; 2) we can  account for non-independence between spVL samples due to the viral phylogeny; and 3) we can generate corrected estimates for each host genetic variant's effect on spVL, excluding any effect of the virus.
-% 
-%\subsection{PGLS}
-%
-%We made two notable changes to the typical PGLS framework (see \citet{Garamszegi2014} for a nice overview of PGLS).
-First, we assumed a variance-covariance structure for regression residuals according to the POUMM.
-This matrix equals $\mathbf{\Sigma}_{OU} + \mathbf{\Sigma}_{e}$.
-Second, we use the variances on the diagonal of this matrix as sample weights to account for non-ultrametricity of the HIV tree.
-We implemented PGLS including these two modifications using the nmle package in R \citep{Pinheiro2019}.
-The resulting general least squares slope and intercept coefficient estimates for GWAS are:
-%
-%\begin{equation}
-%	\hat{\beta}_i = (\mathbf{X}_i^T(\mathbf{\Sigma}_{OU} + \mathbf{\Sigma}_e)^{-1}\mathbf{X}_i)^{-1}\mathbf{X}_i^T(\mathbf{\Sigma}_{OU} + \mathbf{\Sigma}_e)^{-1}\vec{z}
-%\end{equation}
-
-% \subsection{Comparison to PGLS}
-% 
-%We now adopt a second method to correct GWAS for the viral phylogeny by borrowing a method long used by evolutionary biologists.
-Phylogenetic generalized least squares (PGLS) regression is a method to account for phylogenetic non-independence between samples using generalized least squares regression.
-The method has typically been used to test evolutionary hypothesis regarding the relationships between different species traits, In method to estimate the variance-covariance structure of trait data from a  However, an alternate approach already exists to An alternate approach to accounting for non-independence between samples in GWAS would be to use phylogenetic least squares regression (PGLS).
-To the best of our knowledge, PGLS has not yet been applied to GWAS for infectious disease.
-% 
-%A key assumption of ordinary least squares (OLS) regression is that the residual differences between predicted and actual outcomes should be independent and identically distributed 
-% 
-%Given a linear model of assocation between host genetic variants and spVL, we compare the least-squares regression coefficient estimates using 1) our new method to estimate just the non-viral part of spVL and 2) PGLS to assume a phylogentically-informed correlation structure among residuals.
-
-%The inferred POUMM parameters $\alpha$, $\sigma$, $\theta$, $g_{0}$, and $\sigma_e$ define an expected multivariate normal distribution for the viral part of spVL at each tip sample:
-%
-%\begin{equation} \label{eq:muOU}
-%	\mu_{OU} = g_0e^{-\alpha t_{0i}} + (1 - e^{-\alpha t_{0i}})\theta \\
-%\end{equation}
-%
-%\begin{align} 
-%	\bm{\Sigma}_{ii} &= \sigma^2\frac{(1 - e^{-2\alpha t_{0i}})}{2\alpha} \label{eq:SigmaOUii} \\
-%	\bm{\Sigma}_{ij} &= \sigma^2\frac{e^{-\alpha\tau_{ij}}(1 - e^{-2\alpha t_{0(ij)}})}{2\alpha} \label{eq:SigmaOUij}
-%\end{align}
-%
-%with $t_{0i}$ equal to the root-to-tip distance on the phylogenetic tree and $t_{0(ij)}$ equal to the root-to-most recent common ancestor distance for samples $i$ and $j$. 
-%Equation \ref{eq:muOU} describes the expected value for $g_v$ for each sample $i$.
-The variances in the variance-covariance matrix $\bm{\Sigma}_{OU}$, $\bm{\Sigma}_{ii}$ describe the uncertainty in each expected $g_v$ value, while the covariances $\bm{\Sigma}_{ij}$ describe how much shared ancestry each pair of samples has (Equations \ref{eq:SigmaOUii},  \ref{eq:SigmaOUij}).
-
-% We know that $\vec{E}_{|\Theta}\sim \mathcal{N}\bigg(\vec{0}, \mathbf{\Sigma}_e(\Theta)\bigg)$, where $\mathbf{\Sigma}_{e}(\Theta)$ denotes the $N\times N$ diagonal matrix with each diagonal element equal to $\sigma_e^2$;
-% We know that  $\vec{Z}_{|\Theta,\mathcal{T}}\sim \mathcal{N}\bigg(\vec{\mu}_{G}(\Theta_{OU},\mathcal{T}), \mathbf{\Sigma}_{G}(\Theta_{OU},\mathcal{T})+\mathbf{\Sigma}_{e}\bigg)$;
-% 
-%The parameters of the POUMM are $g_{0}$, the ancestral trait value for spVL, $\theta$, the optimal trait value for spVL to maximize viral fitness, $\alpha$, the strength of selection on spVL, $\sigma$, the time-unit standard deviation of random evolutionary fluctuations in spVL, and $sigma_e$, the standard deviation on the non-viral component of spVL in the population.
-
-%The heritable component of spVL, $g_v$, is assumed to evolve continuously within HIV lineages according to an Ornstein-Uhlenbeck (OU) process.
-This assumption may be justified by imagining that the host part of spVL, $g_h$ is the sum of contributions from many alleles, each of which may have a positive or negative effect on spVL.
-Even in the standard GWAS framework, non-heritable noise in phenotype is assumed to be normally distributed with an expected value of zero, justifying our assumption that $e$ is distributed in this way.
-
-%We estimate the viral part of spVL as a covariance-weighted average of the expected $g_v$ values under the POUMM ($\vec{\mu}_{OU}$) and observed spVL values, where the covariance structure for observed spVL is assumed to be a diagonal matrix with diagonal entries equal to the POUMM parameter $\sigma_e$ (Equation \ref{eq:MWA}).
-
-%\begin{equation} \label{eq:MWA}
-%\vec{\mu}_{MWA} = (\bm{\Sigma}_{OU}^{-1} + \bm{\Sigma}_{e}^{-1})^{-1} \times (\bm{\Sigma}_{OU}^{-1}\vec{\mu}_{OU} + \bm{\Sigma}_{e}^{-1}\vec{z})
-%\end{equation}
-
-%used a standard GWAS methodology.
-First, a large cohort of phenotypically diverse individuals was genotyped and phenotyped.
-Next, a separate regression model was fit for each of the millions of surveyed genetic markers.
-Under the most basic model, the phenotype is simply the sum of some additive genetic effect from the particular variant being tested and a residual term encompassing all other genetic and environmental determinants.
-Since a hypothesis test for significant association is performed for each of millions of genetic variants, GWAS performed in this way incurs a heavy multiple-testing burden.
-Despite a strong correction for multiple testing, \citep{McLaren2015} were able to confirm two
-
-%So far, it is well established that both human host and viral genetic factors contribute to spVL.
-\citep{McLaren2015} found that 25\% of the variability in spVL in a mixed European cohort could be explained by additive genetic effects from the human genome.
-More recently, \citep{Mitov2018} established that at least 20\% of the variability in spVL in a UK cohort could be explained by viral genetic variation.
-Furthermore, depending on the methodological assumptions made and the cohort studied, estimates for the broad-sense viral heritability can vary widely \citep{Mitov2018}.
-Finally, human and viral genetic variants may actually explain redundant variation in spVL \citep{Bartha2017}.
-
-%Several GWAS have been performed to try to identify specific human and viral genetic variants that explain variation in spVL.
-The largest GWAS for human genetic factors so far, conducted by \citep{McLaren2015}, confirmed two regions of the human genome are associated with variability in spVL – .
-This study used a standard GWAS framework, in which each surveyed genetic variant (millions in total) was seperately tested for a significant linear relationship with spVL based on thousands of genotyped and phenotyped individuals.
-This method did not take into account heterogeneity in the viral population.
-On the viral side, \citep{Bartha2013} performed an analogous GWAS for viral genetic factors.
-After correcting for multiple testing, they did not find any viral genetic variants significantly associated with spVL.
-Combining genomic information from both humans and viruses, \citep{Bartha2013} also tested for pairwise associations between human genetic variants and viral genetic variants.
-In this way the authors could pinpoint sites of genomic conflict where the host's particular HLA alleles selected for certain viral escape mutations.
-
-%So far, the GWAS for human genetic variants associated with spVL have not accounted for viral genetic variation in a holistic way.
-We wondered whether we could improve the signal of association between known regions of the human genome associated with spVL or discover new associations with spVL by explicitly taking viral genetic variation into account in the GWAS model.
-
-%\subsection{Previous GWAS approaches}
-%
-%Accounting for extraneous factors in GWAS is not a new idea; a broad range of methods have previously been proposed to adjust the simplistic GWAS framework described.
-These adjustments typically have two purposes, either to reduce residual variance in the phenotype or to account for non-independence between samples.
-The reduction of residual variance is often accomplished by including covariates in the regression model that describe extraneous factors influencing the measured phenotype.
-So rather than regressing phenotype on allele dosage alone, a researcher might regress phenotype on allele dosage, participant age, and assay type.
-%
-%Non-independence between samples in GWAS is often thought of only in terms of genetic relatedness of sampled individuals, in which case it is called \emph{population structure}.
-This broad term encompasses a continuum of relatedness.
-At one extreme, individuals in a GWAS may share distant common ancestry and at the other extreme, they may be close family relatives.
-If the phenotype of interest is differently distributed among ancestry or family groups, genetic variants that are common among members of a group simply through inheritance may appear associated with the phenotype of interest, despite not having any causal role in determining the phenotype.
-%
-%Several approaches have been commonly used to account for population structure in GWAS, including adjusting the GWAS significance threshold to account for potential spurious findings \citep{Devlin1999}, performing GWAS only on more homogenous population sub-groups \citep{Pritchard2000}, and adjusting the regression model itself to include estimates of individual relatedness \citep{Price2006, Kang2010}.
-Of these approaches, we highlight the linear mixed model developed by \citep{Kang2010} because it is an increasingly popular technique that has a direct analog in the field of phylogenetic comparative analysis, from which we derive our method to account for pathogen relatedness.
-%
-%\subsection{New methods to correct GWAS for the viral phylogeny}
-%
-%Since spVL is heritable from one HIV infection partner to another, we expect spVL samples to have a correlation structure based on the underlying HIV transmission tree.
-By estimating the transmission tree via the viral phylogeny (which we can infer from HIV sequencing data), we can develop an estimate for the correlation structure of spVL samples.
-Based on this idea, we propose two methods to account for the correlation structure among spVL samples due to the viral phylogeny.
-%
-%First, we try to partition measured spVL for each patient into a part determined by the virus (``viral part") and a part determined by non-viral factors (``non-viral part", e.g.
-additive genetic effects from the host, age, sex, assay error, etc.) Once we have partitioned spVL values in this way we can perform GWAS using only the non-viral part of spVL, effectively removing non-independence and noise due to the viral phylogeny from spVL samples.
-To partition spVL into viral and non-viral parts, we rely on the a phylogenetic mixed model (PMM, \citet{Housworth2004, Lynch1991}).
-Under the PMM, a continuous trait is assumed to be composed of two independent parts: a heritable part and a non-heritable part.
-The heritable part is transmitted from parent to offspring and is subject to the long-term evolutionary forces of selection and drift.
-In contrast, the non-heritable part of a trait is independent among samples and may reflect the influence of local conditions or rapid evolution unconstrained by phylogenetic background.
-Several mathematical models for evolution of the heritable part can be applied in order to estimate the relative contributions of heritable and non-heritable parts to the total trait value.
-It has recently been shown that an Ornstein-Uhlenbeck (OU) process is the best-fitting model for evolution of the heritable viral part of spVL \citep{Mitov2018} \todo{Justify use of OU model better}.
-The mixed model of evolution including an OU process for the heritable part is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM, \citet{Mitov2017a}).
-By fitting such a model to spVL measurements and a viral phylogeny, we can co-infer the OU parameters that describe the evolutionary trajectory of the viral part of spVL as well as the variance of the non-heritable, non-viral part of spVL in the population.
-Using these inferred parameters, we develop a method to estimate for each sample the viral part of spVL (see ``Methods").
-After partitioning each sampled spVL value into a viral and a non-viral part in this way, we can perform a modified GWAS in which only the non-viral part of spVL is regressed on variant dosage.
-%
-%Rather than pre-process spVL values to remove the viral part as just explained, a second method to correct GWAS for a correlation structure among spVL samples would be to directly modify the regression formula using phylogenetic generalized least squares (PGLS) regression.
-As previously mentioned, a typical GWAS for host determinants of spVL fits a linear regression model for each candidate host genetic variant.
-Under this model, spVL is regressed on variant dosage and ordinary least squares regression is used to generate slope and intercept estimates.
-However, a key assumption of ordinary least squares regression is that residual error should be independent among samples.
-Since spVL samples are not independent but are instead related according to the underlying viral phylogeny, the residuals may not be independent either.
-To account for non-independence among residuals, \citep{Grafen1989} suggested the PGLS method, in which the phylogeny is translated into a variance-covariance structure among samples.
-It is assumed that residual errors are not independent, but instead follow this phylogenetic variance-covariance structure.
-General least squares regression can then be used to generate phylogenetically-corrected slope and intercept estimates for the regression of spVL on variant dosage.
-We compare our first method for a phylogenetically-corrected GWAS to PGLS using a variance-covariance structure derived from the same modelling assumptions.
-Namely, we perform PGLS using a variance-covariance structure generated by fitting the POUMM to spVL data and the viral phylogeny.
-
-%In infectious disease GWAS, the pathogen's effect on phenotype is often grouped with other, non-genetic effects as residual phenotypic variation.
-However, several methods have recently been developed to tailor GWAS to the study of microbial genomes, including pathogen genomes.
-
-%Intuitively, this information gives us a measure of how surprised we should be to observe particular spVL values given the underlying HIV phylogeny.
-When we observe spVL values that are very surprising given the sampled individual's location on the phylogeny, we can infer that non-viral factors are contributing to the deviation from our expectation.
-For a more rigorous description of our POUMM-based method to estimate the viral and non-viral parts of spVL, see the ``Methods" section.
-
-%\todo{Add justifying citations to this paragraph}At this point we should address several of what we feel are the most important assumptions made by the POUMM.
-Fundamentally, we assume that the spVL phenotype is the sum of independent heritable (i.e.
-viral) and non-heritable (i.e.
-non-viral) parts.
-This neglects the possibility of of interaction effects between the virus and the host or the host's environment.
-We also assume there is no discontinuity in the virus between infection partners.
-This assumption ignores the possibility of quasi-species transmission, in which only a subset of the potentially heterogeneous viral population circulating in the infecting host is transmitted to their infection partner.
-\todo{To be continued with other assumptions, need to think about what impact they would have}
-
-%\subsection{Accounting for the viral phylogeny in G2G GWAS}
-%So far we have discussed confounding in HIV GWAS due to human (host) population factors.
-However, in certain infectious disease  contexts the viral population can also confound GWA analyses.
-In the genome-to-genome (G2G) GWAS framework described by \citep{Bartha2013}, host variants are tested for pairwise associations with viral variants to determine sites of genetic conflict between the human host and HIV pathogen.
-In this case, confounding may arise from population stratification in both the host and viral populations (Figure \ref{fig:gwas-covariates}b).
-\citep{Naret2018} describe two potential types of stratification that may occur in the G2G context.
-So-called ``major stratification" occurs when host population is correlated with viral population and vice-versa.
-``Minor stratification"  occurs when host population is correlated with with viral variant frequencies or when viral population is correlated with host variant frequencies.
-Although \citep{Naret2018} envision only discrete viral populations, the same phenomena could easily apply when considering the full viral phylogeny.
-%
-%Several methods have been developed to correct for confounding from the viral phylogeny in HIV G2G GWAS.
-One of two methods originally introduced by \citep{Bhattacharya2007} involves determining whether viral mutations are inferred to occur disproportionately along branches of the phylogeny where a particular host variant is present.
-The other method is a model-selection approach in which phylogenetic models with and without selection pressure from host variants are compared for goodness-of-fit \citep{Bhattacharya2007, Carlson2008}.
-Lastly, the more recently proposed method by \citep{Naret2018} uses phylogenetic principle components to describe the major axes of variation in the viral population (an approach analogous to using principle components to account for host population structure).
-However, the method proposed by \citep{Naret2018} does not have a secondary benefit of the other two methods, which is to restore statistical independence to samples that are fundamentally related by a phylogeny.
-%
-%\subsection{Towards a full correlation model for HIV GWAS}
-%
-%As \citep{Bhattacharya2007} recognized, HIV samples are related by descent according to the transmission tree.
-Such relationships are often investigated by inferring a viral phylogeny from sequencing data.
-Recently, several studies using phylogentic methods determined that host spVL values are correlated among hosts with closely related versions of the virus \citep{Alizon2010, Blanquart2017, Hollingsworth2010, Mitov2018}.
-This correlation can be quantified in terms of the heritability of spVL, or, the fraction of variance in spVL among hosts that can be explained by underlying viral genetic variation.
-\citep{Mitov2018} demonstrated that spVL is at least 20\% heritable, meaning that the virus has a strong effect on host spVL.
-Based on this correlation between the virus and spVL (Figure \ref{fig:gwas-covariates}c), we argue it is necessary to expand the correlation model typically assumed in standard GWAS.
-Expanding upon previous work in the G2G GWAS framework, we develop a full correlation model for spVL GWAS (Figure \ref{fig:gwas-covariates}d).
-From the standard GWAS framework, we take that host variant frequencies are correlated with host population and that host population is correlated with spVL.
-From the G2G framework, we take that the viral phylogeny is correlated with host population.
-Finally, we assume that the viral phylogeny effects spVL based on recent spVL heritability studies.
-If we assume that host population stratification and noise factors are already taken into account in GWAS for host determinants of spVL, we may reduce the full correlation model to Figure \ref{fig:gwas-covariates}e.
-It is then apparent that accounting for host population stratification is sufficient to control for confounding effects of the viral phylogeny.
-However, the viral phylogeny remains both a source of noise and a cause of non-independence between samples in GWAS for host determinants of spVL.
-%
-%\begin{figure}[h!]
-%	\begin{tikzpicture}
-%	\begin{scope}[xshift = 0cm, yshift = 0cm]
-%	\draw (-.75, 2) node {\textbf{a) standard GWAS}}; 
-%	\node[smod, minimum size = 3.75cm, minimum width = 3cm] (a) at (-0.75, -0.75/2) {};
-%	\node[snode] (z) at (0, 0) {$z$};
-%	\node[snode] (ch) at ($(z) + (-1.5, +.75)$) {$P$};
-%	\draw[taro] (ch) -- (z);
-%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
-%	\draw[taro] (gh) -- (z);
-%	\node[snode] (n) at ($(z) + (0, -1.5)$) {$n$};
-%	\draw[taro] (n) -- (z);
-%	\draw[bline] (gh) -- (ch);
-%	\end{scope}
-%	
-%	\begin{scope}[xshift = 5cm, yshift = 0cm]
-%	\draw (0.2, 2) node {\textbf{b) Genome-to-genome GWAS}}; 
-%	\node[smod, minimum size = 3cm, minimum width = 4.5cm] (a) at (0, 0) {};
-%	\node[white] (z) at (0, 0) {$z$};  % White (hidden) node z to anchor coordinate system
-%	\node[snode] (ch) at ($(z) + (-1.5, +.75)$) {$P$};
-%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
-%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
-%	\node[snode] (gv) at ($(z) + (+1.5, -.75)$) {$g_v$};
-%	\draw[taro] (gh) -- (gv);
-%	\draw[bline] (ch) -- (gh);
-%	\draw[bline] (cv) -- (gv);
-%	\draw[bline] (ch) -- (cv);
-%	\draw[dline] (ch) -- (gv);
-%	\draw[dline] (cv) -- (gh);
-%	\end{scope}
-%	
-%	\begin{scope}[xshift = 10cm, yshift = 0cm]
-%	\draw (0.9, 2) node {\textbf{c) spVL Heritability}}; 
-%	\node[smod, minimum size = 2.25cm, minimum width = 3cm] (a) at (0.75, 0.75/2) {};
-%	\node[snode] (z) at (0, 0) {$z$};
-%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
-%	\draw[taro] (cv) -- (z);
-%	\end{scope}
-%	
-%	\begin{scope}[xshift = 0cm, yshift = -5cm]
-%	\draw (0, 2) node {\textbf{d) Full correlation model}}; 
-%	\node[smod, minimum size = 3.75cm, minimum width = 4.5cm] (a) at (0, -.75/2) {};
-%	\node[snode] (z) at (0, 0) {$z$};
-%	\node[snode] (ch) at ($(z) + (-1.5, +.75)$) {$P$};
-%	\draw[taro] (ch) -- (z);
-%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
-%	\draw[taro] (gh) -- (z);
-%	\node[snode] (n) at ($(z) + (0, -1.5)$) {$n$};
-%	\draw[taro] (n) -- (z);
-%	\draw[bline] (gh) -- (ch);
-%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
-%	\draw[taro] (cv) -- (z);
-%	\draw[bline] (cv) -- (ch);
-%	\end{scope}
-%	
-%	\begin{scope}[xshift = 6.5cm, yshift = -5cm]
-%	\draw (1.5, 2) node {\textbf{e) Without standard GWAS covariates}}; 
-%	\node[smod, minimum size = 3cm, minimum width = 4.5cm] (a) at (0, 0) {};
-%	\node[snode] (z) at (0, 0) {$z$};
-%	\node[snode] (gh) at ($(z) + (-1.5, -.75)$) {$g_h$};
-%	\draw[taro] (gh) -- (z);
-%	\node[snode] (cv) at ($(z) + (+1.5, +.75)$) {$\Psi$};
-%	\draw[taro] (cv) -- (z);
-%	\end{scope}
-%	\end{tikzpicture}
-%	\caption{Correlation and dependency structures for HIV GWAS.
-Solid lines are correlations between variables and arrows show directional dependency.
-Dotted lines indicate minor stratification (see section "Accounting for the viral phylogeny...").
-%		a) The host population ($P$) is a confounding factor for the association between host variants ($g_h$) and spVL ($z$). Noise factors ($n$) contribute residual variance to spVL. 
-%		b) The host population and the viral phylogeny ($\Psi$) are both confounding factors for the association between host and viral ($g_v$) variants. 
-%		c) spVL depends on the viral phylogeny.
-%		d) Combining models a-c yields the full correlation model.
-%		e) Model d may be reduced by controlling for host population stratification and noise factors}
-%	\label{fig:gwas-covariates}
-%\end{figure}
-%
-%\todo{Naret doesn't have any justification for minor stratification - how could this arise?}
-%
-
-%
-%
-%# LEFT OFF HERE
-%
-%\subsection{The standard GWAS framework}
-%
-%GWA studies conduct an unbiased scan of the genome for genetic variants associated with a phenotype of interest.
-Although GWAS are frequently used to explore the genetic architecture of complex phenotypes, they suffer from several known sources of error.
-We discuss two such sources of error now: confounding from latently associated variables (a source of false positive associations) and high residual variance (a source of false negative associations).
-%
-%The first source of error, confounding, arises when a latent variable effects the phenotype of interest and is also correlated with genetic variants being tested (Figure \ref{fig:gwas-covariates}a).
-A classic example of a confounding variable for GWAS is ancestry.
-If allele frequencies differ between human populations of different ancestry and phenotypes also differ between the same populations, false positive associations may result since ancestry-associated genetic variants will appear to explain variation in phenotype that is truly due to ancestry.
-This phenomenon is called population stratification and several methods have been developed to address the problem.
-In the context of GWAS for host determinants of spVL, a principle components method \citep{Price2006} has most often been applied.
-In this method, the top few principle components of host genetic variation are included in the association model as proxy variables for demographic and geographic factors.
-This effectively changes the association test hypothesis from ``does the host variant explain a significant amount of variation in spVL?", to ``does the host variant explain a significant amount of variation in spVL \emph{beyond that which can be explained by demographic or geographic factors?}" Previous GWAS for host determinants of spVL conducted by \citep{Price2006, McLaren2015, Bartha2013, Fellay2007, Pereyra2010, Pelak2010, VanManen2011} have employed this method.
-Several studies have additionally used other methods to account for host population stratification, including structured association \citep{Pritchard2000} and genomic control \citep{Devlin1999}.
-In structured association, the sample population is a priori divided into groups that are assumed to be homogeneous with respect to potentially confounding variables.
-GWA testing is then conducted independently within each group before results are finally aggregated across groups.
-\citep{McLaren2015, Pereyra2010} both applied a form of structured association.
-So far only \citep{Dalmasso2008} have used genomic control to account for population stratification in spVL GWAS.
-In genomic control, the significance threshold for association testing is uniformly adjusted for all variants by a factor which accounts for the degree of stratification present.
-Although methods to account for population stratification vary, all GWAS for host determinants of spVL to-date have taken the problem into account.
-%
-%The second source of error typically considered in GWAS for host determinants of spVL arises from ``noise" variables that contribute extra variance to the phenotype of interest.
-These variables effect the phenotype, but are not correlated with the genetic variant being tested (Figure \ref{fig:gwas-covariates}a).
-As long as the phenotype is quantitative (as in the case of spVL) or the sample is randomly drawn from the population, including these factors as covariates in the association model helps remove residual variance and increase power to detect true associations \todo{Investigate this claim further} \citep{Mefford2012}.
-In the context of GWAS for host determinants of spVL, sex, cohort, and age have all been recognized to be noise variables that effect spVL but should not be correlated with host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Fellay2009, Pelak2010}.
-
-%Since the viral and non-viral parts of spVL are unobservable, we developed a simulation model under the POUMM assumptions as a simplistic test of our method.
-Figure \ref{fig:sim-design} provides a graphical overview of the complete model.
-When possible, we tried to parameterize the model using values from previously reported HIV cohort data.
-%
-%First, we fixed the total variance in simulated spVL to be 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}.
-After allotting 25\% of this variance to $g_h$ (following \citet{McLaren2015}), we partitioned the remaining variance between $g_v$ and $\epsilon$ in different ratios in order to assess our estimator’s performance across a range of possible spVL heritability values (Figure \ref{fig:sim-design}a, Table \ref{tab:sim-params}).
-Second, we generated the host part of spVL by summing contributions from $M$ causal host variants (Figure \ref{fig:sim-design}b).
-\todo{Add binomial description to figure} Host genotypes were simulated according to a binomial distribution with a constant allele frequency shared across all variants.
-Third, we generated $g_v$ by simulating an Ornstein-Uhlenbeck (OU) process of evolution along a viral phylogenetic tree (Figure \ref{fig:sim-design}c).
-We use a randomly generated tree with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014}.
-To test under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively.
-For the other OU parameters, we used previously inferred values from \citep{Bertels2018, Blanquart2017, Mitov2018}.
-Finally, $\epsilon$ values were generated from a normal distribution with mean zero.
-
-%\begin{proof}
-%	\begin{equation}\label{eqPdfGValues2}
-%	\begin{array}{rcl}
-%	f\big(<\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}},\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}>\big) & = & f\big(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ |\ \vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big) \times f\big(\vec{e}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\big)\\
-%	& = &\mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
-%	& &  \mathcal{N}\bigg(\vec{z} - \vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{0}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
-%	& = & \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\ \vec{\mu}_{OU}(\Theta,\mathcal{T}), \mathbf{\Sigma}_{OU}(\Theta,\mathcal{T})\bigg) \times \\
-%	& &  \mathcal{N}\bigg(\vec{g}_{|\Theta,\mathcal{T},\sigma_e,\vec{z}}\ ;\  \vec{z}, \mathbf{\Sigma}_e(\sigma_e)\bigg) \\
-%	\end{array}
-%	\end{equation}
-%	
-%	Equations \ref{eqPdfGValues1} and \ref{eqMuSigmaGValues} follow from eq. \ref{eqPdfGValues2} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
-%\end{proof}
-
-%\begin{table}[H]
-%	\begin{center}
-%		\begin{tabular}{c|l|c|c}
-%			\toprule 
-%			\textbf{Parameter} & \textbf{Description} & \textbf{Posterior mean} & \textbf{95\% HPD} \\
-%			\midrule 
-%			$z_0$ & spVL value at the root of the transmission tree & 4.9 & 2.8, 7.3 \\
-%			$\theta$ & evolutionarily optimal spVL value & 4.3 & 4.2, 4.4 \\
-%			$\sigma$ & time-unit standard deviation of the OU process & 4.9 & 3.5, 6.4 \\
-%			$\alpha$ & selection strength & 44.5 & 28.4, 60.9 \\
-%			$\sigma_e$ & standard deviation of the non-heritable part of spVL & 0.56 & 0.50, 0.63 \\
-%			\toprule 
-%		\end{tabular}
-%		\caption{POUMM parameters.}
-%		\label{tab:POUMMParamDesc}
-%	\end{center}
-%\end{table}
-
-%\begin{table}[H] \label{tab:comp-gwas}
-%	\begin{center}
-%	\begin{tabular}{llclccc} \toprule 
-%	& \multicolumn{2}{c}{\specialcell[l]{\citep{McLaren2015} study \\ N = 6,315}} & \multicolumn{3}{c}{\specialcell[l]{Current study \\ N = 1,370}} & \\ \midrule 
-%	Region & Variant & \emph{P} value & Variant & Effect size & \emph{P} value & R$^2$ \\ \midrule
-%	MHC & rs59440261 & 2.0 x 10$^{-83}$ & rs202036493 & \specialcell[t]{-0.31 \\ (-0.52)} & \specialcell[t]{1.3 x 10$^{-13}$ \\  (9.2 x 10$^{-15}$)} & 1 \\
-%	% MHC & rs59440261 & 2.0 x 10$^{-83}$ & rs2923003 & -0.31 & 1.26511 x 10$^{-13}$ & 1 \\
-%	% MHC & - & - & rs112243036 & -0.20 & 2.37153 x 10$^{-14}$ & 0.592 \\
-%	\emph{CCR5} & rs1015164 & 1.5 x 10$^{-19}$ & rs7637813 & \specialcell[t]{0.09 \\ (0.17)} & \specialcell[t]{2.9 x 10$^{-6}$ \\  (2.0 x 10$^{-7}$)} & 0.867 \\
-%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs2187671 & \specialcell[t]{0.08 \\ (0.15)} & \specialcell[t]{4.7 x 10$^{-5}$ \\  (4.9 x 10$^{-6}$)} & 0.867 \\
-%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs34602794 & \specialcell[t]{0.08 \\ (0.15)} & \specialcell[t]{4.7 x 10$^{-5}$ \\  (4.9 x 10$^{-6}$)} & 0.867 \\
-%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs2157056 & \specialcell[t]{0.08 \\ (0.15)} & \specialcell[t]{4.7 x 10$^{-5}$ \\  (4.9 x 10$^{-6}$)} & 0.867 \\
-%	% \emph{CCR5} & rs1015164  & 1.5 x 10$^{-19}$ & rs4683217 & \specialcell[t]{0.08 \\ (0.14)} & \specialcell[t]{1.4 x 10$^{-4}$ \\  (1.3 x 10$^{-5}$)} & 0.867 \\
-%	% NOTE: these p-values are from analysis /Volumes/stadler/SHCSData/analysis/200127/1500*
-%	 \toprule
-%	\end{tabular}
-%	\caption{Association results for the top SNPs in the MHC and \emph{CCR5} regions from the GWAS by \citep{McLaren2015} compared to association results for the best proxy SNPs for these SNPs available in the current study.
-The proxy SNPs were chosen based on high linkage disequilibrium (as measured by R$^2$) with the McLaren et al.
-SNPs.
-For the current study, values without parentheses are from the GWAS with phylogenetic correction and values in parentheses are from the standard GWAS.}
-%	\end{center}
-%\end{table}
-
-%\begin{table}[H]
-%	\begin{center}
-%		\begin{tabular}{c|c|c|c}
-%			\toprule 
-%			\textbf{Filter} & \textbf{Current study} & \textbf{\citet{Bartha2013}} & \textbf{\citet{McLaren2015}} \\
-%			\midrule 
-%			SNP missingness & $>$ 0.01 & $>$ 0.01 &  $>$ 0.05 \\
-%			Genotype missingness & $>$ 0.02 & Not specified & $>$ 0.02 \\
-%			IBD & $>$ 0.125 & Not specified & $>$ 0.125 \\ 
-%			SNP minor allele frequency & $<$ 0.01 & $<$ 0.01 & $<$ 0.01 \\
-%			Deviation from Hardy-Weinberg equilibrium & p $<$ 5E-5 & p $<$ 5E-5 & p $<$ 1E-7 \\
-%			Inbreeding coefficient & NA & NA & $<$ -0.1 or $>$ 0.1 \\
-%			\midrule
-%			Remaining samples & 2125 & 698 & 6315 \\
-%			Remaining variants & $\sim$ 4.4 million & $\sim$ 7 million & $\sim$ 8 million \\
-%			\bottomrule 
-%		\end{tabular}
-%		\caption{Genotype data quality control prior to GWAS.}
-%		\label{tab:GWASQC}
-%	\end{center}
-%\end{table}
-
-%\begin{table}[H]
-%	\centering
-%	\begin{tabular}{rlrrr}
-%		\hline
-%		& stat & PostMean & HPD\_lower & HPD\_upper \\ 
-%		\hline
-%		1 & alpha & 30.15 & 10.15 & 49.83 \\ 
-%		2 & theta & 4.38 & 4.25 & 4.55 \\ 
-%		3 & sigmae & 0.62 & 0.56 & 0.68 \\ 
-%		4 & g0 & 3.98 & 3.14 & 4.84 \\ 
-%		5 & sigma & 3.56 & 2.04 & 5.15 \\ 
-%		\hline
-%	\end{tabular}
-%	\caption{Pipeline rerun 2 results.}
-%\end{table}
-
-%\begin{table}[ht]
-%	\centering
-%	\begin{tabular}{rlrrr}
-%		\hline
-%		& stat & PostMean & HPD\_lower & HPD\_upper \\ 
-%		\hline
-%		1 & alpha & 30.31 & 10.54 & 51.07 \\ 
-%		2 & theta & 4.38 & 4.24 & 4.55 \\ 
-%		3 & sigmae & 0.62 & 0.56 & 0.68 \\ 
-%		4 & g0 & 3.99 & 3.15 & 4.79 \\ 
-%		5 & sigma & 3.58 & 2.07 & 5.17 \\ 
-%		\hline
-%	\end{tabular}
-%	\caption{Pipeline rerun diff POUMM seed results.}
-%\end{table}
-%
-%\begin{table}[ht]
-%	\centering
-%	\begin{tabular}{rlrrr}
-%		\hline
-%		& stat & PostMean & HPD\_lower & HPD\_upper \\ 
-%		\hline
-%		1 & alpha & 33.15 & 14.66 & 53.74 \\ 
-%		2 & theta & 4.37 & 4.25 & 4.49 \\ 
-%		3 & sigmae & 0.62 & 0.56 & 0.68 \\ 
-%		4 & g0 & 4.49 & 3.30 & 5.67 \\ 
-%		5 & sigma & 3.69 & 2.17 & 5.18 \\ 
-%		\hline
-%	\end{tabular}
-%	\caption{Pipeline rerun diff tree results.}
-%\end{table}
-
-%\begin{table}[H] \label{tab:comp-gwas}
-%	\begin{center}
-%		\begin{tabular}{p{1cm}p{2cm}p{1cm}p{2.5cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr} {llclcccc}
-%			\hline
-%			& & & McLaren et al.  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
-%			\hline
-%			Region & Variant & R$^2$ & p-value & Beta & p-value & Beta & p-value \\
-%			\hline
-%			MHC & rs59440261 & - & 2.0 x 10$^{-83}$ & -0.41 & 1.8x10$^{-11}$ & -0.25 & 3.3x10$^{-10}$ \\
-%						& rs202036493 & 1 & - & -0.52 & 9.2 x 10$^{-15}$ & -0.31 & 1.3 x 10$^{-13}$ \\
-%			\emph{CCR5} & rs1015164 & - & 1.5 x 10$^{-19}$ & 0.15 & 1.1x10$^{-6}$ & 0.08 & 3.0x10$^{-5}$ \\
-%						& rs7637813 & 0.867&  - & 0.17 & 2.0 x 10$^{-7}$ & 0.09 & 2.9 x 10$^{-6}$  \\
-%			\hline
-%		\end{tabular}
-%		\caption{Top association results from \citep{McLaren2015} compared to association results for the best proxy SNPs available in the current study.
-Proxy SNPs were chosen based on linkage disequilibrium (as measured by R$^2$) with the McLaren et al.
-SNPs.}
-%	\end{center}
-%\end{table}
-
-% Again following the procedure of \citep{McLaren2015}, we imputed common HLA genotypes to more finely map association signals in the MHC region.
-Some HLA alleles also became more significant with a phylogenetic correction (table \ref{tab:lim-mclaren-comparison}).
-
-% \begin{table}[H]
-% 	\begin{center}
-% 	\begin{tabular} {p{3cm}p{1cm}p{2cm}p{1cm}p{2cm}p{1cm}p{2cm}} % {lrrrrrr}
-% 		\hline
-% 		 & \multicolumn{2}{l}{McLaren et al.}  & \multicolumn{2}{l}{Standard GWAS } & \multicolumn{2}{l}{GWAS with phylogenetic correction} \\ 
-% 		 \hline
-% 		Allele & Beta & p-value & Beta & p-value & Beta & p-value \\
-% 		\hline	
-% 		HLA-B*57:01 & -0.840 & $2.69 \times 10^{-86}$ & -0.479 & $1.11 \times 10^{-12}$ & -0.268 & $5.43 \times 10^{-12}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-C*06:02 & -0.436 & $7.32 \times 10^{-50}$ & -0.173 & $1.10 \times 10^{-4}$ & -0.107 & $3.15 \times 10^{-5}$ \\ 
-% 		HLA-B*27:05 & -0.417 & $4.32 \times 10^{-20}$ & -0.244 & $9.59 \times 10^{-4}$ & -0.122 & $4.40 \times 10^{-3}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-C*07:01 & 0.195 & $1.86 \times 10^{-14}$ & 0.129 & $7.55 \times 10^{-3}$ & 0.086 & $1.99 \times 10^{-3}$ \\ 
-% 		HLA-C*07:02 & 0.194 & $3.75 \times 10^{-13}$ & 0.095 & $3.02 \times 10^{-2}$ & 0.052 & $3.98 \times 10^{-2}$ \\ 
-% 		HLA-B*07:02 & 0.200 & $5.17 \times 10^{-13}$ & 0.129 & $5.29 \times 10^{-3}$ & 0.072 & $7.06 \times 10^{-3}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-B*08:01 & 0.206 & $2.94 \times 10^{-11}$ & -0.009 & $9.20 \times 10^{-1}$ & 0.015 & $7.75 \times 10^{-1}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-C*08:02 & -0.307 & $8.76 \times 10^{-11}$ & 0.324 & $8.55 \times 10^{-1}$ & 0.400 & $6.95 \times 10^{-1}$ \\ 
-% 		HLA-C*04:01 & 0.183 & $9.07 \times 10^{-11}$ & 0.124 & $3.04 \times 10^{-3}$ & 0.058 & $1.67 \times 10^{-2}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-B*13:02 & -0.346 & $3.70 \times 10^{-10}$ & -0.188 & $1.03 \times 10^{-2}$ & -0.112 & $7.90 \times 10^{-3}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-B*14:02 & -0.318 & $2.10 \times 10^{-9}$ & 0.067 & $8.12 \times 10^{-1}$ & 0.118 & $4.71 \times 10^{-1}$ \\ 
-% 		\rowcolor{Gray}
-% 		HLA-C*12:02 & -0.478 & $4.02 \times 10^{-8}$ & 0.111 & $7.63 \times 10^{-1}$ & 0.094 & $6.58 \times 10^{-1}$ \\
-% 		\hline
-% 	\end{tabular}
-% 	\caption{Association results for common class I HLA alleles ranked by p-value from the GWAS by \citep{McLaren2015} compared to results from the current study.
-The table is truncated to show only alleles that reached genome-wide significance in the McLaren et al.
-study.
-The full table is available in the supplement (table \ref{tab:full-mclaren-comparison}).
-Grey rows are alleles that increased in significance with a phylogenetic correction.}
-% 	\label{tab:lim-mclaren-comparison}
-% 	\end{center}
-% \end{table}

### 2021-11-18 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 	\caption{A high-level schematic of our POUMM-based simulation framework.
-(A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an OU process.
+(A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an Ornstein Uhlenbeck process.
 (B) shows host genetic effects, which are the sum of independent effects from several causal variants.
 Each variant can be present in 0, 1, or 2 copies and we assume half have effect size $\delta$ on the trait value and half have effect size $-\delta$.
-(C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero.}
+(C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero.
+Measured trait values are the sum of these three effects for each simulated individual.}
 		\label{fig:spVLModel}
@@
 Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
-The black point represents our estimates for the heritability and selection strength of spVL based on Swiss HIV data.
+The black point represents our estimates for the heritability and selection strength of spVL based on Swiss HIV Cohort Study data.
 RMSE = Root mean square error, TPR = True positive rate.}
@@
 \caption{Effect size and p-values from top-associated variants in the $\emph{CCR5}$ and MCR regions from the two GWAS performed in our study.
-``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with estimated non-pathogen effects on spVL.}
+``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with the estimated non-pathogen part of the trait.}
 \begin{tabularx}{\linewidth}{XXXXXXX}

### 2021-11-22 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
 \usepackage[margin=1in]{geometry}
+\usepackage{lineno} % for line numbers
 \usepackage{graphicx}
@@
 \usepackage[round,sort]{natbib}
+\linespread{1.3}
 
@@
 
-\startsetups
-    \setcounter{figure}{0}
-    \renewcommand{\thefigure}{S\arabic{figure}}
-    \setcounter{table}{0}
-    \renewcommand{\thetable}{S\arabic{table}}
-\stopsetups
-
 \begin{document}
 
-\begin{flushright}
-Submission type: Article\\
-Section: Methods
-\end{flushright}
+% \begin{flushright} % for MBE submission
+% Submission type: Article\\
+% Section: Methods
+% \end{flushright}
 
@@
 \noindent
-$^{1}$Department of Biosystems Science and Engineering, ETH Zürich, Basel, Switzerland\\
+$^{1}$Department of Biosystems Science and Engineering, ETH Zurich, Basel, Switzerland\\
 $^{2}$Swiss Institute of Bioinformatics, Lausanne, Switzerland\\ 
@@
 \newpage
-\begin{doublespace}
+\begin{linenumbers}
+% \begin{doublespace}
 
 \begin{abstract}
-	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all effect disease traits.
+	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all affect disease traits.
 Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits.
 However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.
-We propose a novel method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits.
+We propose a new method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits.
 In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
-When we apply our method to HIV-1 cohort data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
-Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects.
-Our framework may improve GWAS for other diseases where pathogen effects are stronger and more phylogenetically correlated amongst individuals in a cohort.
+When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
+Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects.
+Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
 \end{abstract}
@@
 % Why is spVL and important trait
-For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with severity of disease course.
-spVL is generally defined to be the average concentration of viral particles in host plasma during the asymptomatic phase of infection in the absence of treatment \citep{Mellors1996}.
-In untreated individuals, spVL is predictive of duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}.
+For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with the severity of disease course.
+spVL is generally defined to be the average concentration of viral RNA copies in host plasma during the asymptomatic phase of infection in the absence of treatment (see e.g.
+\cite{Alizon2010}).
+In untreated individuals, spVL is predictive of the duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}.
 If viral load can be reduced to undetectable levels, an individual is effectively uninfectious and the risk of disease progression is massively reduced \citep{HHSARTGuidelines2019}.
@@
 Heritability is a key measure of how genetically-determined a trait is.
-Here we distinguish between two different heritability measures that are used in different contexts in infectious disease study.
+Here we distinguish between two different heritability measures that are used in different contexts in the study of infectious diseases.
 Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e.
@@
 the sum of independent effects from all genetic variants.
-GWAS for infectious disease typically measure narrow-sense heritability of a trait based on human genetic variants.
+GWAS for infectious disease traits typically measure the narrow-sense heritability of a trait based on human genetic variants.
 
@@
 % How heritable is spVL (viral side)
-In addition to these human genetic factors, it is well-recognized that viral genetic factors effect spVL.
+In addition to these human genetic factors, it is well-recognized that viral genetic factors affect spVL.
 As mentioned, heritability from the viral side is typically measured using broad-sense heritability.
 Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty).
-Estimates using phylogenetic methods on large UK and Swiss cohorts by \citep{Mitov2018} and \citep{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
+Estimates using phylogenetic methods on large UK and Swiss cohorts by \citet{Mitov2018} and \citet{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
 However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}.
@@
 This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}.
-In summary, human genetic factors play a role in determining spVL, but these effects may be mediated by viral genetic variation.
+In summary, human genetic factors play a role in determining spVL, but these effects may be due to interaction with specific viral genetic variants.
 
@@
 In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors.
-This has several potential negative consequences, namely that (i) viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, (ii) variability due to viral effects would make it more challenging to identify human variants of small effect, and (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
+This has several potential negative consequences.
+(i) Viral effects may be confounding or extraneous variables that bias estimates of host genetic effects.
+(ii) Variability due to viral effects would make it more challenging to identify human variants of small effect.
+Finally, (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
 
 % Overview of relevant methods and why we don't use them
-Issue (iii) is closely related to a well-known problem in standard GWAS, namely that shared (human) ancestry, especially between close relatives, can give rise to spurious genetic correlations with a trait.
+Issue (iii) is closely related to a well-known problem in standard GWAS.
+Shared (human) ancestry, especially between close relatives, can also give rise to spurious genetic correlations with a trait.
 Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}.
-More recently, \citep{Power2016} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS.
+More recently, \citet{Power2017} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS.
 Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}.
@@
 However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait.
-In contrast, we want to estimate effects of $human$ genetic variants on a trait, taking into account pathogen effects.
-\citep{Naret2018} developed a relevant method for this task.
+In contrast, we want to estimate effects of $human$ genetic variants on a trait, while accounting for pathogen effects.
+\citet{Naret2018} developed a relevant method for this in the context of a genome-to-genome GWAS framework.
 The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing.
@@
 Our approach should help identify human genetic variants that affect disease traits and more accurately estimate their effects.
-Furthermore, we think this approach represents the first GWAS framework to identify human genetic variants associated with a disease trait that accounts for non-independence of samples due to the underlying pathogen phylogeny.
 
 In the following we describe a statistical model for the spVL trait, derive a maximum likelihood estimate for the viral part of spVL under this model, and describe a new infectious disease GWAS framework using this information.
-In simulations, we demonstrate that this framework can improve GWAS power to detect host genetic variants effecting a disease trait.
+In simulations, we show that this framework can improve GWAS power to detect host genetic variants that affect disease traits.
 Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects.
-Although we developed our framework in the context of HIV-1 spVL, the same approach could readily be applied to other heritable infectious disease traits.
+Although we developed our framework in the context of HIV-1 spVL, this approach can readily be applied to other heritable infectious disease traits.
 
@@
 In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part.
-Then, the heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A.
+The heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A.
 The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.
@@
 	\begin{center}
-	\includegraphics[width = \linewidth]{figures/simulation_setup.pdf}
+	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
 	\caption{A high-level schematic of our POUMM-based simulation framework.
-(A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an Ornstein Uhlenbeck process.
-(B) shows host genetic effects, which are the sum of independent effects from several causal variants.
-Each variant can be present in 0, 1, or 2 copies and we assume half have effect size $\delta$ on the trait value and half have effect size $-\delta$.
-(C) shows that environmental effects are assumed to be independent between sampled individuals and are drawn from a distribution centered at zero.
-Measured trait values are the sum of these three effects for each simulated individual.}
+(A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an Ornstein-Uhlenbeck process.
+(B) shows how host genetic effects are the sum of independent effects from several causal variants.
+Each variant can be present in 0, 1, or 2 copies.
+Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
+(C) shows how environmental effects are independently drawn from a Gaussian distribution centered at 0.
+These three effects sum to the trait value for each simulated individual.}
 		\label{fig:spVLModel}
@@
 extreme spVL values are unlikely.
-\citep{Mitov2018} and \citep{Bertels2018} previously showed the OU process has higher statistical support for spVL.
+\citet{Mitov2018} and \citet{Bertels2018} previously showed the OU process has higher statistical support for spVL.
 This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \citep{Fraser2014}.
@@
 The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \citet{Mitov2018}.
-Here, we review the main relevant points in the spVL context.
+Here, we review the main points in the spVL context.
 
@@
 We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$:
+
 \begin{equation}
@@
 The first term represents a deterministic pull towards an optimal trait value and the second term represents stochastic fluctuations modelled by Brownian motion \citep{Butler2004}:
+
 \begin{align}
@@
 Assuming $g_v(t)$ starts at initial value $g_0$ at time $t = 0$ at the root of the phylogeny, we can write the expectation for $g_v(t)$ at time $t$:
+
 \begin{equation}
@@
 The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \citep{Butler2004}:
+
 \begin{equation}
@@
 The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.
+
 \begin{align}
@@
 Finally, broad-sense trait heritability can be calculated as the fraction of total trait variance that is heritable:
+
 \begin{equation}
@@
 Considering that $\bm{G_v}$ and $\bm{E}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{e}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{e}$:
+
 \begin{equation}
@@
 where the expected value of $\bm{g}_v$ and the covariance matrix $\boldsymbol{\Sigma}_G$ are defined as:
+
 \begin{align}
@@
 
-	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citep{Petersen2012}.
+	Equations \ref{eq:MuG} and \ref{eq:SigmaG} follow from eq. \ref{eq:pdfG} and eq. 371, p. 42, section 8.1.8 ``Product of Gaussian densities'' in \citet{Petersen2012}.
 \end{proof}
@@
 Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{e}$, the spVL value $without$ pathogen effects:
+
 \begin{equation}
@@
 Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
+For a description of the full simulation scheme, see Figure \ref{fig:sim-design}.
 In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength.
@@
 Additive host genetic effects represent an ideal (albeit unattainable) baseline for infectious disease GWAS.
-Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate this value compared to the trait value.
+Figure \ref{fig:simulationResults}A shows that our method incorporating phylogenetic information can more accurately estimate these value compared to the trait value.
 To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
 In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation scheme.
-By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable and under low selection pressure.
-This is because high heritability means the pathogen greatly effects the trait value.
-Then, when the trait is under weak selection, pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability.
-Thus, our method performs well when an infectious disease trait is highly heritable and trait values are highly correlated amongst closely related members of a transmission chain, but vary widely within the entire cohort.
+By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable, under low selection pressure, and with relatively moderate stochastic fluctuations compared to outbreak duration.
+This is because high heritability means the trait value is highly pathogen-dependent.
+Then, when the trait is under weak selection, the pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability.
+Thus, our method performs well when an infectious disease trait is highly heritable and trait values are strongly correlated amongst close transmission partners.
 
@@
 Next, we characterized the evolutionary scenarios under which our framework can actually improve GWAS power.
-We used the true positive rate (TPR) to evaluate the fraction of simulated host genetic variants with direct effects we could recover as being significantly associated with the trait.
+We used the true positive rate (TPR) to evaluate the fraction of simulated causal host genetic variants we could recover as being significantly associated with the trait.
 We performed three different GWAS for each simulated dataset: the first represents an ideal in which we can exactly know and remove pathogen effects from trait values, the second is using our method to estimate this value and remove it, and the third represents a standard GWAS using the scaled trait value.
 Figure \ref{fig:simulationResults}B shows that our framework can improve the TPR in simulated scenarios where selection strength $<$ 10 time$^{-1}$ and heritability $>$ 45\%.
-If we were able to perfectly estimate and remove pathogen effects on a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable.
+If we were able to perfectly estimate and remove pathogen effects from a trait, the TPR would increase across all values of selection strength so long as the trait is more than marginally heritable.
 We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects.
@@
 For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
-(A) shows that our method (left) can generate more accurate estimates of host genetic effects than the trait value, scaled by its mean (right).
+(A) shows that our method (left) can generate more accurate estimates of additive host genetic effects than the trait value, scaled by its mean (right).
 (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
@@
 Finally, we applied our framework to empirical data from the Swiss HIV Cohort Study (SHCS).
-We used data collected from 1,392 individuals in Switzerland infected with subtype B HIV-1 between 1994 and 2018.
+We used data collected from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018.
 The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals.
 We followed the framework outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
-Figure \ref{fig:spvl-calc-comparison} shows the calculated spVL values, Figure \ref{fig:spvl-on-tree} shows the inferred pathogen phylogeny, and figures \ref{fig:spvl-on-tree}, \ref{fig:trait-comparison} show the estimated non-pathogen effects on spVL calculated for these individuals.
-We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) \ref{tab:POUMM-parameters-our-estimates}.
+Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
+Figure \ref{fig:spvl-on-tree} shows that this trait is not strongly phylogenetically structured in the cohort, despite high heritability.
+Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
+We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}).
 To put these values into the context of our simulation study, they are shown as black points on Figure \ref{fig:simulationResults}.
@@
 We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes.
-In the ``GWAS with standard trait value'' we used the unmodified trait value, our calculated spVL values, as the GWAS response variable.
+In the ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable.
 In the ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL.
@@
 Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
-However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Figure \ref{tab:gwas-top-associations}).
+However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}).
 Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}.
@@
 (A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables.
-On the left, we used unmodified spVL values.
+On the left, we used unmodified (total) spVL values.
 On the right, we used our estimates for the non-pathogen effects on spVL.
@@
 Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
-The dashed lines show genome-wide significance at p = 5x10$^{-8}$.}
+The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
 		\label{fig:gwas-results}
@@
 \caption{Top association results from \citet{McLaren2015} compared to results from this study.
-The ``Standard trait value'' and ``Estimated non-pathogen part of trait'' results are from using two different GWAS response variables with the same human genotype data.}
+Results from this study are for host variants from the SHCS in GWAS with two different response variables.
+``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
 \begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
@@
 % State the importance of your findings
-In this paper, we presented a novel, phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits.
+In this paper, we presented a new phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits.
 By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait.
@@
 Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study.
-Compared to previous studies, this estimate is at the higher end \citep{Mitov2018}.
-Also using the POUMM, \citep{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citep{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort.
+Compared to previous studies, this estimate is at the higher end (see \citet{Mitov2018} and references therein).
+Also using the POUMM, \citet{Bertels2018} estimated a spVL heritability of 29\% (N = 2014, CI 12 - 46\%) from the same cohort and \citet{Blanquart2017} estimated 31\% (N = 2028, CI 15 - 43\%) from a pan-European cohort.
 We note that our sample size (N = 1493 individuals) is smaller than in these other studies.
 This might be because we restricted samples based on having $pol$ gene sequences with at least 750 non-ambiguous bases.
-Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream GWAS methods.
+Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream trait-correction method.
 Although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.
@@
 Given our estimates for the heritability of spVL and the selection strength on this trait using Swiss HIV cohort data, our simulation results reveal that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (Figure \ref{fig:simulationResults}).
-Indeed the results of our  GWAS method presented here and a standard GWAS are very similar.
-Future applications might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g.
-\cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.
+Our method slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B).
+Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM.
+In real life, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values.
+However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A).
+Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL is robust to our correction for pathogen effects.
 
 Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests.
-Therefore, it is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}).
+It is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}).
 The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.
+Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g.
+\cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.
 
@@
 We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}.
-After allotting 25\% of this variance to the host part of spVL $g_h$ (following \citet{McLaren2015}), we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
+Other studies have estimated slightly lower values though (Table \ref{tab:spVLmeasurements}).
+After allotting 25\% of this variance to the host part of spVL $g_h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
 $g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
@@
 $g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
-For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$, similar to values previously inferred for HIV (Table \ref{tab:POUMM-parameters-our-estimates}).
+For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on fitting the same model to SCHS data (Table \ref{tab:POUMM-parameters-our-estimates}).
+This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}).
 To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively.
@@
 
-Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS, www.shcs.ch, \citet{Schoeni-Affolter2010, Scherrer2021CohortSHCS}).
-All participants were HIV-1–infected individuals at least 16 years old and written informed consent was obtained from all cohort participants.
-The anonymized data was made available for this study after a study proposal was approved by the SHCS.
+Human genotypes, viral load measurements, and HIV-1 $pol$ gene sequences from HIV-1 positive individuals were all collected in the context of other studies by the Swiss HIV Cohort Study (SHCS) (www.shcs.ch, \citet{Schoeni-Affolter2010, Scherrer2021CohortSHCS}).
+All participants were HIV-1–infected individuals 16 years or older and written informed consent was obtained from all cohort participants.
+The anonymized data were made available for this study after the study proposal was approved by the SHCS.
 
@@
 We trimmed the alignment to 1505 characters to standardize sequence lengths.
-We used IQ-TREE version 1.6.9 to construct an approximate maximum likelihood tree with -m GTR+F+R4 for a general time reversible substitution model with empirical base frequencies and four free substitution rate categories.
+We used IQ-TREE version 1.6.9 \citep{Nguyen2014} to construct an approximate maximum likelihood tree with -m GTR+F+R4 for a general time reversible substitution model with empirical base frequencies and four free substitution rate categories.
 Otherwise, we used the default IQ-TREE settings.
@@
 We calculated spVL as the arithmetic mean of viral RNA measurements made prior to the start of antiretroviral treatment.
+For a comparison of several different filtering methods, see Figure \ref{fig:spvl-calc-comparison}.
 
-For GWAS, we retained data from 1,392 of the 1,493 individuals with European ancestry and not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}).
+For GWAS, we retained data from 1,392 of the 1,493 SHCS individuals with European ancestry who were not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}).
 These were 227 females and 1165 males.
@@
 
+\subsection*{Data availability}
+The simulated data underlying this article can be re-generated using the code available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.
+The HIV pathogen genome sequences, clinical data, and human genotypes cannot be shared publicly due to the privacy of individuals who participated in the cohort study.
+The data may be shared on reasonable request to the Swiss HIV Cohort Study at http://www.shcs.ch.
+
 \section*{Acknowledgments}
@@
 
-The Swiss HIV Cohort Study is supported by the Swiss National Science Foundation (grant #201369), by SHCS project #858 and by the SHCS research foundation .
+The Swiss HIV Cohort Study is supported by the Swiss National Science Foundation (grant #201369), by SHCS project #858 and by the SHCS research foundation.
 Furthermore, the SHCS drug resistance database is supported by the Yvonne Jacob Foundation (to HFG).
 The data are gathered by the Five Swiss University Hospitals, two Cantonal Hospitals, 15 affiliated hospitals and 36 private physicians (listed in http://www.shcs.ch/180-health-care-providers).
+% \end{doublespace}
+\end{linenumbers}
 
@@
 \newpage
-\startsection
+\startsetups
+    \setcounter{figure}{0}
+    \renewcommand{\thefigure}{S\arabic{figure}}
+    \setcounter{table}{0}
+    \renewcommand{\thetable}{S\arabic{table}}
+\stopsetups
 \section*{Supplemental Material}\label{supp}
@@
 \setcounter{page}{1}
+
+\begin{doublespace}
 
@@
 \begin{figure}[H]
-	\centering
-		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on Swiss data.
-We ran two different MCMC chains to ensure the estimates converged.
-The black point on the x-axis shows the posterior mean value, which was used to estimate the pathogen- and non-pathogen effects on spVL.}
-\label{fig:poumm-parameters}
-\end{figure}
-
-\begin{figure}[H]
 \begin{center}
@@
 	\label{fig:spvl-calc-comparison}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/spvl_on_tree.png}
+	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}
+	\label{fig:spvl-on-tree}
 	\end{center}
@@
 \begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/spvl_on_tree.png}
-	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}
-	\label{fig:spvl-on-tree}
-	\end{center}
+	\centering
+		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
+		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on SHCS data.
+We ran two different MCMC chains to ensure the estimates converged.
+The black point on the x-axis shows the posterior mean value, which was used to estimate the pathogen- and non-pathogen effects on spVL.}
+\label{fig:poumm-parameters}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{figures/qq_plots.png}
+	\caption{Quartile-quartile plots from association tests. The dashed green line shows the y = x line.}
+	\label{fig:qq-plots}
 \end{figure}
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{figures/qq_plots.png}
-	\caption{Quartile-quartile plots from association tests. The dashed green line shows the y = x line.}
-	\label{fig:qq-plots}
-\end{figure}
-
 \end{doublespace}  % don't doublespace tables
@@
 \begin{table}[H]
-\caption{POUMM parameter estimates for spVL based on Swiss data. HPD = Highest posterior density.}
+\caption{POUMM parameter estimates for spVL based on SHCS data. HPD = Highest posterior density.}
 	\begin{tabularx}{\linewidth}{XXX} \hline 
@@
 
-\begin{table}[H]
-\caption{POUMM parameter estimates for spVL from previous studies.}
-	\begin{tabularx}{\linewidth}{llXXX}
+% latex table generated in R 4.1.1 by xtable 1.8-4 package
+% Sat Oct 30 10:31:16 2021
+\begin{table}[ht]
+\caption{Effect size and p-values from the top most strongly associated variants in the $\emph{CCR5}$ and MCR regions from each of the two GWAS performed in our study.
+``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with the estimated non-pathogen part of the trait.
+Entries above the dividing line are the top-associated variants from the ``Standard'' GWAS and entries below the dividing line are the top-associated variants from the ``Corrected'' GWAS.
+Many entries overlap between the two.}
+\begin{tabularx}{\linewidth}{XXXXXXX}
   \hline
-	Parameter & Value (Uncertainty) & Reference & Notes \\
+    Region & Position & Variant & Standard effect size & Standard p-value & Corrected effect size & Corrected p-value \\ 
   \hline
-	$v_0$ & 5.54 (4.04 - 7.25) & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
-	$\theta$ & 4.45 (4.41 - 4.49) & \citet{Mitov2018} & \\
-	$\theta$ &  4.0 (1.6 - 4.)) & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
-	$\theta$ & 4.1 (3.5 - 4.9) & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
-	$\alpha$ & 28.78 (16.64 - 46.93) & \citet{Mitov2018} & \\
-	$\alpha$ & 32.7 (0.03 - 57.6) & \citet{Bertels2018} & \\
-	$\alpha$ & 7.6 (1.2 - 10) & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
-	$\sigma$ & 2.97 (1.95 - 4.37) & \citet{Mitov2018} & \\
-	$\sigma$ & 1.3 (0.66 -1.87) &  \citet{Blanquart2017} & \\
-	$\sigma_e$ & 0.77 (0.73, 0.8) & \citet{Mitov2018} & \\
-	$\sigma_e$ & 0.61 (0.54, 0.65) &  \citet{Blanquart2017} & \\
+    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
+    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46540932 & rs10865942 & -0.16 & $8.4 \times 10^{-9}$ & -0.081 & $4 \times 10^{-7}$ \\ 
+    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
+    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
+    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
+    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
+    MHC & 31391401 & rs4418214 & -0.34 & $2.4 \times 10^{-14}$ & -0.18 & $2.5 \times 10^{-12}$ \\ 
+    MHC & 31400137 & rs138130755 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    MHC & 31400705 & rs138117378 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    MHC & 31402358 & rs148792134 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    MHC & 31409677 & rs140991764 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
+    \hline
+    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
+    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
+    $\emph{CCR5}$ & 46556835 & rs6808142 & 0.15 & $8.3 \times 10^{-8}$ & 0.082 & $3.3 \times 10^{-7}$ \\ 
+    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
+    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
+    MHC & 31367874 & rs111281598 & -0.37 & $1.5 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
+    MHC & 31376266 & rs73400361 & -0.37 & $1.4 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
+    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
    \hline
 \end{tabularx}
-	\label{tab:POUMMparams-prev-estimated}
+\label{tab:gwas-top-associations}
 \end{table}
@@
 %\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}
+
+\begin{table}[H]
+\caption{POUMM parameter estimates for spVL from previous studies.}
+	\begin{tabularx}{\linewidth}{llXXX}
+	\hline 
+	Parameter & Value (Uncertainty) & Reference & Notes \\
+	\hline 
+	$g_0$ & 5.54 (4.04 - 7.25) & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
+	$\theta$ & 4.45 (4.41 - 4.49) & \citet{Mitov2018} & \\
+	$\theta$ &  4.0 (1.6 - 4.)) & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
+	$\theta$ & 4.1 (3.5 - 4.9) & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
+	$\alpha$ & 28.78 (16.64 - 46.93) & \citet{Mitov2018} & \\
+	$\alpha$ & 32.7 (0.03 - 57.6) & \citet{Bertels2018} & \\
+	$\alpha$ & 7.6 (1.2 - 10) & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
+	$\sigma$ & 2.97 (1.95 - 4.37) & \citet{Mitov2018} & \\
+	$\sigma$ & 1.3 (0.66 -1.87) &  \citet{Blanquart2017} & \\
+	$\sigma_e$ & 0.77 (0.73, 0.8) & \citet{Mitov2018} & \\
+	$\sigma_e$ & 0.61 (0.54, 0.65) &  \citet{Blanquart2017} & \\
+	\hline 
+	\end{tabularx}
+	\label{tab:POUMMparams-prev-estimated}
+\end{table}
 
@@
 
-% latex table generated in R 4.1.1 by xtable 1.8-4 package
-% Sat Oct 30 10:31:16 2021
-\begin{table}[ht]
-\caption{Effect size and p-values from top-associated variants in the $\emph{CCR5}$ and MCR regions from the two GWAS performed in our study.
-``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with the estimated non-pathogen part of the trait.}
-\begin{tabularx}{\linewidth}{XXXXXXX}
-  \hline
-    Region & Position & Variant & Standard effect size & Standard p-value & Corrected effect size & Corrected p-value \\ 
-  \hline
-    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46540932 & rs10865942 & -0.16 & $8.4 \times 10^{-9}$ & -0.081 & $4 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
-    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
-    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
-    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
-    MHC & 31391401 & rs4418214 & -0.34 & $2.4 \times 10^{-14}$ & -0.18 & $2.5 \times 10^{-12}$ \\ 
-    MHC & 31400137 & rs138130755 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31400705 & rs138117378 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31402358 & rs148792134 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31409677 & rs140991764 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46556835 & rs6808142 & 0.15 & $8.3 \times 10^{-8}$ & 0.082 & $3.3 \times 10^{-7}$ \\ 
-    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
-    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
-    MHC & 31367874 & rs111281598 & -0.37 & $1.5 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
-    MHC & 31376266 & rs73400361 & -0.37 & $1.4 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
-    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:gwas-top-associations}
-\end{table}
-
 \stopsection

### 2022-05-30 scripts/R/simulate_for_examples.R
get example trait values from simulation

@@ -1,0 +1,71 @@
+# This script is to repeat a simulation where our method improves GWAS TPR
+# It generates data for an additional figure comparing trait values
+
+# Source necessary functions
+source("scripts/R/functions/simulate_epidemic.R")
+source("scripts/R/functions/simulation_control_functions.R")
+source("scripts/R/functions/utility_functions.R")
+source("scripts/R/functions/plot_functions.R")
+source("scripts/R/functions/PLINK_utility_functions.R")
+source("scripts/R/functions/POUMM_utility_functions.R")
+source("scripts/R/functions/GWAS_utility_functions.R")
+
+# Set seed for replicability of simulation
+set.seed(1)
+
+# Read parameters for data simulation from config file
+# config_values <- yaml::read_yaml(file = "../../config-simulation.yaml")
+config_values <- yaml::read_yaml(file = "config-simulation.yaml")
+N <- config_values$N
+
+# Designate output file
+outfile <- generateOutfileName(outdir = "output_revisions", description = "simulateFullDataset")
+
+# Simulate HIV phylogeny
+tree <- simulateHIVTreeExpBL(N)
+
+# Set constant parameters for data simulation
+param.list <- list(
+  N = N,
+  g0 = config_values$g0,
+  alpha = config_values$alpha,
+  theta = config_values$theta,
+  M = config_values$M,
+  K = config_values$K,
+  d = config_values$d,
+  delta = config_values$delta,
+  H2 = config_values$H2,
+  H2.h = config_values$H2.h,
+  var.z = config_values$var.z,
+  N.reps = config_values$N.reps)
+
+param.df <- permuteParameters(
+  N.reps = param.list$N.reps,
+  parameterlist = param.list)
+
+# Simulate data and calculate estimator error
+data.list <- mapply(
+  simulateFullDataset,
+  N = param.df$N,
+  g0 = param.df$g0,
+  alpha = param.df$alpha,
+  theta = param.df$theta,
+  M = param.df$M,
+  K = param.df$K,
+  d = param.df$d,
+  delta = param.df$delta,
+  H2 = param.df$H2,
+  H2.h = param.df$H2.h,
+  var.z = param.df$var.z,
+  MoreArgs = list(tree = tree,
+                  is.MLE = config_values$is.MLE))
+
+# Clean up and save simulation results
+data <- makeDataFrameFromMapplyResults(mapply.results = data.list)
+
+write.table(
+  data,
+  file = outfile,
+  quote = F,
+  sep = "\t",
+  row.names = F)

### 2022-05-30 scripts/bash/build-tree.sh
output all locally optimal trees to account for phylo uncertainty

@@ -13,10 +13,12 @@
 sed -e 's/\(^>.*$\)/#\1#/' output/pathogen.fasta | tr -d "\r" | tr -d "\n" | sed -e 's/$/#/' | tr "#" "\n" | sed -e '/^$/d' > output/pathogen_oneline.fasta
 cut -c 1-1505 output/pathogen_oneline.fasta > output/pathogen_trimmed.fasta
 
-# Build a ML tree with the GTR+F+R4 model and (TODO) 100 bootstrap replicates
+# Build a ML tree with the GTR+F+R4 model
+# -wt writes out all locally optimal trees
 iqtree \
 -s output/pathogen_trimmed.fasta \
 -m GTR+F+R4 \
+-wt \
 -redo \
 -pre output/pathogen \
 -nt AUTO
@@ -25,6 +27,3 @@ iqtree \
 # 38 reference pol sequences subtypes A-K from Los Alamos HIV database used as outgroup
 # Used RAxML to build the tree, RAxML default model is hard to tell from quick look at documentation
 # Used 100 bootstrap replicates
-
-
-

### 2022-05-30 scripts/bash/get_atomm_host_genotypes.sh
document ATOMM data formatting

@@ -1,0 +1,54 @@
+# ==============================================================================
+# Coverted PLINK BED file to haploid genotype matrix for ATOMM.
+# ==============================================================================
+
+# Run interactivelly with: 
+# docker run -it --rm --mount type=bind,src=$PWD/scripts,dst=/scripts --mount type=bind,src=$PWD/output,dst=/output plink
+# copy-paste these commands
+
+DATA_PREFIX="spVL.shcs.rs.filtered"
+MEM=10000
+
+# Make sure REF allele is the major allele
+./plink2 \
+--bfile output/$DATA_PREFIX \
+--maj-ref 'force' \
+--make-bed \
+--out output/$DATA_PREFIX.majref
+
+# Get variant dosages from .bed file (binary)
+./plink2 \
+--bfile output/$DATA_PREFIX.majref \
+--export A-transpose \
+--memory $MEM \
+--out output/$DATA_PREFIX.Av.majref
+
+# Make genotypes haploid
+head -1 output/$DATA_PREFIX.Av.majref.traw > output/atomm_sequence_host_header.txt
+
+### TO REMOVE ###
+# Down-sample host SNPs for first-pass analysis try
+head -1000 output/$DATA_PREFIX.Av.majref.traw > output/$DATA_PREFIX.Av.majref.traw.subsample
+
+# Reference allele is counted, so make missing data = 2 reference alleles
+sed 's/\tNA/\t2/g' output/$DATA_PREFIX.Av.majref.traw.subsample| tail -n +2 > output/$DATA_PREFIX.Av.haptmp1.traw
+# heterozygotes -> homozygous alternate allele (0)
+sed 's/\t1/\t0/g' output/$DATA_PREFIX.Av.haptmp1.traw | tail -n +2 > output/$DATA_PREFIX.Av.haptmp2.traw
+# homozygous reference allele -> 1
+sed 's/\t2/\t1/g' output/$DATA_PREFIX.Av.haptmp2.traw | tail -n +2 > output/$DATA_PREFIX.Av.hap.traw
+rm output/$DATA_PREFIX.Av.haptmp1.traw output/$DATA_PREFIX.Av.haptmp2.traw
+
+# File looks like this now:
+# CHR     SNP     (C)M    POS     COUNTED ALT     <host1>     <host2>     <host3> 
+# 1       1:768116:A:AGTTTT       0       768116  AGTTTT  A       0       1       1 
+
+# Format for ATOMM
[…]

### 2022-05-30 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 \begin{abstract}
-	Infectious diseases are a unique challenge for genome-wide association studies (GWAS) because pathogen, host, and environmental factors can all affect disease traits.
+	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) may significantly influence a trait of interest.
 Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits.
@@
 More recently, \citet{Power2017} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS.
-Two state-of-the-art methods exist for this \citep{Earle2016, collins_phylogenetic_nodate}.
+Several state-of-the-art methods exist for this, e.g.
+\citep{Earle2016, collins_phylogenetic_nodate}.
 % \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
@@
 Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects.
-Of these, only pathogen effects are heritable from one transmission partner to another \citep{Leventhal2016}.
+n GWAS, hosts are typically selected to not share recent ancestry.
+Further, shared environmental variables are included as covariates to avoid spurious associations.
+We argue similar considerations must be taken
+for pathogen effects that heritable from one transmission partner to another \citep{Leventhal2016}.
 To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}.
@@
 Under the POUMM, the spVL trait $z$ is the sum of viral effects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$.
-We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $e$:
+We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $l$:
 
 \begin{equation}
-    z = g_{v} + e
+    z = g_{v} + l
 \end{equation}
@@
 \end{equation}
-Next, we remember that $e$ is the non-heritable, environmental part of spVL.
-$e$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
-The expectation of $e$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level.
-The parameter $\sigma_e^2$ measures the between-host variance of the environmental effect.
+Next, we remember that $l$ is the non-heritable, environmental part of spVL.
+$l$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
+The expectation of $l$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level.
+The parameter $\sigma_l^2$ measures the between-host variance of the environmental effect.
 
@@
 \begin{split}
-	E(e) = 0 \\
-	Var(e) = \sigma^2_e
+	E(l) = 0 \\
+	Var(l) = \sigma^2_l
 \end{split}
@@
 \begin{equation}
-	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(e)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_e}
+	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(l)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_l}
 \end{equation}
@@
 
-Similarly, let $\bm{e}$ be a vector of the environmental part of spVL for each individual.
-$\bm{e}$ is a realization of a Gaussian random vector $\bm{E} \sim \mathcal{N}\big(\bm{0}, \boldsymbol{\Sigma}_E\big)$, where $\boldsymbol{\Sigma}_E$ is a diagonal matrix with diagonal elements equal to $\sigma^2_e$.
+Similarly, let $\bm{l}$ be a vector of the environmental part of spVL for each individual.
+$\bm{l}$ is a realization of a Gaussian random vector $\bm{L} \sim \mathcal{N}\big(\bm{0}, \boldsymbol{\Sigma}_L\big)$, where $\boldsymbol{\Sigma}_L$ is a diagonal matrix with diagonal elements equal to $\sigma^2_l$.
 
-Considering that $\bm{G_v}$ and $\bm{E}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{e}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{e}$:
+Considering that $\bm{G_v}$ and $\bm{L}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{l}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{l}$:
 
 \begin{equation}
-	f\big(\bm{g}_v, \bm{e}\big) \propto \mathcal{N}\big(\bm{g}_v; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
+	f\big(\bm{g}_v, \bm{l}\big) \propto \mathcal{N}\big(\bm{g}_v; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
 	\label{eq:pdfGprop}
@@
 \begin{align}
-	Exp(\bm{g}_v) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_E^{-1} \bm{z}\big) \label{eq:MuG}\\
-	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_E^{-1}\big)^{-1} \label{eq:SigmaG}
+	Exp(\bm{g}_v) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_L^{-1} \bm{z}\big) \label{eq:MuG}\\
+	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_L^{-1}\big)^{-1} \label{eq:SigmaG}
 \end{align}
@@
 	\begin{split}
-		f\big(\bm{g}_v,\ \bm{e}\big) &= f\big(\bm{g}_v|\ \bm{e}\big) \times f\big(\bm{e}\big) \\
-	&= f\big(\bm{g}_v\big) \times f\big(\bm{e}\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{e};\ \bm{0}, \mathbf{\Sigma}_E\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_E\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_E\big)
+		f\big(\bm{g}_v,\ \bm{l}\big) &= f\big(\bm{g}_v|\ \bm{l}\big) \times f\big(\bm{l}\big) \\
+	&= f\big(\bm{g}_v\big) \times f\big(\bm{l}\big) \\
+	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{l};\ \bm{0}, \mathbf{\Sigma}_L\big) \\
+	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_L\big) \\
+	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_L\big)
 	\end{split}
@@
 
-Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{e}$, the spVL value $without$ pathogen effects:
+Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{l}$, the spVL value $without$ pathogen effects:
 
 \begin{equation}
-	\hat{\bm{e}} = \bm{z} - Exp(\bm{g}_v)
+	\hat{\bm{l}} = \bm{z} - Exp(\bm{g}_v)
 	\label{eq:EHat}
@@
 To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
-In the supplemental material, we show why the scaled trait value is expected to have an RMSE of approximately 0.74 under our simulation scheme.
+As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the spVL trait due to viral genetic effects and environmental effects is fixed.
+Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.
 By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable, under low selection pressure, and with relatively moderate stochastic fluctuations compared to outbreak duration.
@@
 For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.
+All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.
 

### 2022-05-30 scripts/R/make_manuscript_figures.rmd
update figure labels

@@
 tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
+example_simulation_results <- read.delim("output_revisions/2022-05-30_simulateFullDataset_2.txt")
 tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
@@
         labs(x = expression("Selection strength (" * alpha * ")"),
-             y = expression("spVL heritability (" * italic('H'^2) * ")")) +
+             y = expression("Heritability (" * italic('H'^2) * ")")) +
         shared_theme +
@@
 ggsave(plot = p2, filename = paste(FIGDIR, "trait_values.png", sep = "/"),
+      width = linewidth * 2/3, height = height * 2/3, units = "cm")
+```
+
+Example trait values from successful simulation
+```{r}
+p1 <- ggplot(data = example_simulation_results %>% rename("Heritability" = H2), aes(x = h, y = h.MWA)) +
+      geom_point() +
+      facet_grid(. ~ Heritability, labeller = "label_both") + 
+      shared_theme +
+      labs(x = "Simulated host part of trait", y = ehat_name)
+ggsave(plot = p1, filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
       width = linewidth * 2/3, height = height * 2/3, units = "cm")

### 2022-05-30 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 		\label{fig:simulationResults}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=0.7\linewidth]{figures/example_simulation_trait_values}
+		\caption{Example simulated data from two evolutionary scenarios where (left) a phylogenetic correction does not improve GWAS power and (right) where it does.
+Both data sets were simulated with selection strength = 0.1 time $^{-1}$ and all other parameters fixed as in the full simulation study.}
+		\label{fig:simulationTraitExample}
 	\end{center}

### 2022-05-30 scripts/R/make_manuscript_figures.rmd
add example simulation points to fig

@@
 get_xval_for_sim_plots <- function(real_vals, estimate) {
-      alpha_vals <-
-      base_val <- rev(real_vals[real_vals < estimate])[1]
+      base_val <- rev(real_vals[real_vals <= estimate])[1]
       next_val <- real_vals[real_vals > estimate][1]
@@
         alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = alpha_estimate))
-
 spvl_data$inference.method <- factor(
         spvl_data$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+
+# Get example simulation POUMM values to annotate plot with
+example_data1 <- data.frame(
+        inference.method = "h.MWA.p.value",
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75), estimate = 0.15),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))
+example_data1$inference.method <- factor(
+        example_data1$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+example_data2 <- data.frame(
+        inference.method = "h.MWA.p.value",
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.75),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))
+example_data2$inference.method <- factor(
+        example_data2$inference.method,
         levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
@@
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
-        geom_point(data = spvl_data, fill = "black")
+        geom_point(data = spvl_data, fill = "black") + 
+        geom_point(data = example_data1, fill = "black", shape = 1) + 
+        geom_point(data = example_data2, fill = "black", shape = 1)
 

### 2022-05-30 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 The black point represents our estimates for the heritability and selection strength of spVL based on Swiss HIV Cohort Study data.
+The empty points show simulated scenarios that are presented in more detail in Figure \ref{fig:simulationTraitExample}.
 RMSE = Root mean square error, TPR = True positive rate.}
@@
 		\caption{Example simulated data from two evolutionary scenarios where (left) a phylogenetic correction does not improve GWAS power and (right) where it does.
-Both data sets were simulated with selection strength = 0.1 time $^{-1}$ and all other parameters fixed as in the full simulation study.}
+The only difference in the two simulations is heritability; selection strength was fixed to 0.1 time $^{-1}$ and all other parameters were fixed as in the full simulation study.
+These examples correspond to the empty points in Figure \ref{fig:simulationResults}.}
 		\label{fig:simulationTraitExample}

### 2022-05-30 scripts/R/make_manuscript_figures.rmd
adjust sim results fig alignment

@@
 
-top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1, 3.25), labels = c("", "A"))
+top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1.03, 3), labels = c("", "A"))
 png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)

### 2022-06-08 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) may significantly influence a trait of interest.
-Previous GWAS have successfully identified several human genetic variants associated with HIV-1 set point viral load (spVL), among other important infectious disease traits.
+Previous GWAS have successfully identified host genetic variants associated with important infectious disease traits, such as HIV-1 set point viral load (spVL).
 However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.
@@
 % How do we measure the extent to which genetic factors determine spVL
-To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
+To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors as opposed to pathogen genetic factors.
 Heritability is a key measure of how genetically-determined a trait is.
-Here we distinguish between two different heritability measures that are used in different contexts in the study of infectious diseases.
-Broad-sense heritability $H^2$ measures the fraction of total trait variance that is heritable, i.e.
-due to inherited differences.
-In the infectious disease case, broad-sense heritability from pathogen factors, which are inherited by recipients from their infection partners, is typically measured.
-On the other hand, narrow-sense heritability $h^2$ measures the fraction of total trait variance due specifically to additive genetic effects, i.e.
-the sum of independent effects from all genetic variants.
-GWAS for infectious disease traits typically measure the narrow-sense heritability of a trait based on human genetic variants.
-
+On the host side, narrow-sense heritability, the fraction of total trait variance explained by additive genetic effects, is a typical output of GWAS.
 % How heritable is spVL (human side)
-Several GWAS have been done to measure the narrow-sense heritability of spVL and identify associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
-The largest study to-date by \citet{McLaren2015} estimated the narrow-sense heritability of spVL from human genetic variants to be approximately 25\%.
+Several GWAS have measured the narrow-sense heritability of spVL and identified associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
+The largest study to-date by \citet{McLaren2015} estimated that additive human genetic effects account for approximately 25\% of the variation in spVL amongst infected individuals.
 All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}).
@@
 % How heritable is spVL (viral side)
-In addition to these human genetic factors, it is well-recognized that viral genetic factors affect spVL.
-As mentioned, heritability from the viral side is typically measured using broad-sense heritability.
-Estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty).
-Estimates using phylogenetic methods on large UK and Swiss cohorts by \citet{Mitov2018} and \citet{Bertels2018} measured the broad-sense heritability of spVL from the virus to be 21\% - 29\%.
+On the pathogen side, broad-sense heritability $H^2$ is typically measured.
+This is the fraction of total trait variance due to any factors ``inherited'' from infection partner to infection partner (the most obvious being pathogen genetic factors).
+Broad-sense heritability is commonly measured by quantifying how correlated a trait is between infection partners or how phylogenetically structured a trait is based on a reconstructed pathogen phylogeny.
+For HIV spVL, estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty).
+Using phylogenetic methods on large UK and Swiss cohorts, \citet{Mitov2018} and \citet{Bertels2018} estimated that pathogen factors account for 21\% - 29\% of the variation in spVL amongst infected individuals.
 However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}.
@@
 This has several potential negative consequences.
-(i) Viral effects may be confounding or extraneous variables that bias estimates of host genetic effects.
-(ii) Variability due to viral effects would make it more challenging to identify human variants of small effect.
+Viral effects may be (i) confounding or (ii) extraneous variables that produce distorted or spurious associations between host genetic variants and a disease trait in host GWAS.
+%This may occur if the virus affects the disease trait (the dependent variable in host GWAS) or if it affects both the disease trait and the host genotype (the independent variable in host GWAS).
+For example, stratification in both the host and pathogen populations may yield spurious correlations in host GWAS that are really due to independent pathogen factors.
+Then, co-evolution between a pathogen and its host may yield distorted estimates for host genetic effects that are not broadly protective against all pathogen strains.
 Finally, (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
@@
 It is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}).
-The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM} and all the code we wrote is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.
+The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}.
+All code for the simulations and HIV spVL analysis presented in this study is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.
 Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g.
@@
 
+\subsection*{\emph{A. thaliana}-\emph{X. arboricola} data}
+\emph{A.
+thaliana} genotypes were downloaded from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}.
+Quantitative diseases resistance (QDR) score measurements and \emph{X.
+arboricola} genotypes were provided by the authors of \citep{Wang2018Two-wayGenomes}.
+The \emph{X.
+arboricola} QDR trait values were generated in a fully-crossed experimental design, where each host type (accession) was infected with each pathogen strain.
+QDR was measured on 4 leaves for each host-pathogen pairing.
+
+For phylogenetic inference, the \emph{X.
+arboricola} genotype matrix consisted of allele dosage information for 33,610 SNPs in 22 strains.
+We used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from these data.
+We calculated mean QDR across all leaves infected on all hosts for each \emph{X.
+arboricola} strain.
+
+For GWAS, we retained data from 133 \emph{A. thaliana} accessions with associated QDR measurements.
+
 \subsection*{POUMM parameter inference}
 
-We used the R package POUMM version 2.1.6 \citep{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the approximate maximum-likelihood phylogeny and calculated spVL values.
+We used the R package POUMM version 2.1.6 \citep{Mitov2017a-POUMM} to infer the POUMM parameters $g_0, \alpha, \theta, \sigma, \text{ and }\sigma_e$ from the HIV-1 and \emph{X.
+arboricola} phylogenies and associated spVL and QDR trait values.
 The Bayesian inference method implemented in this package requires specification of a prior distribution for each parameter.
-We used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$.
-We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000.
+For HIV-1 spVL, we used the same, broad prior distributions as in \citet{Mitov2018}, namely: $g_{0} \sim \mathcal{N}(4.5,\ 3)$, $\alpha \sim Exp(0.02)$, $\theta \sim \mathcal{N}(4.5,\ 3)$, $H^2_{\bar{t}} \sim \mathcal{U}(0,\ 1)$, and $\sigma^2_e \sim Exp(0.02)$.
+For \emph{X.
+arboricola} QDR, we modified the $g_0$ and $\theta$ priors to match the empirical mean and standard deviation of QDR trait values in the dataset: $g_{0} \sim \mathcal{N}(0.4,\ 0.2)$ and $\theta \sim \mathcal{N}(0.4,\ 0.2)$.
+We ran two MCMC chains for 4x10$^6$ samples each with a target sample acceptance rate of 0.01 and a thinning interval of 1000 for both analyses.
 The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution.
@@
 
-\subsection*{Phylogenetic spVL correction}
-We corrected calcualted spVL values using the method described in this paper.
-For each of the 1,392 individuals in the GWAS cohort, we estimated the viral part of spVL using equation \ref{eq:MuG} and the corresponding non-viral part using equation \ref{eq:EHat}.
+\subsection*{Phylogenetic trait correction}
+We corrected HIV-1 spVL and \emph{X.
+arboricola} QDR trait values using the method described in this paper.
+For each host individual, we estimated the pathogen part of their trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}.
 For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.
+For the \emph{A.
+thaliana}-\emph{X.
+arboricola} trait correction and downstream GWAS, we randomly selected one pathogen pairing for each host to analyze.
+This makes the dataset analogous to the observational HIV-1 dataset.
 All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.
@@
 
+\begin{table}[H]
+\caption{Number of variants for \emph{X. arboricola} QDR GWAS after sequential filtering steps.}
+	\begin{tabularx}{\linewidth}{XX}
+	\hline
+	Variant filter & Number of variants remaining \\
+	\hline 
+	Raw data & 12883854 \\
+	Bi-allelic variants & 11769920 \\ \hline			
+	\end{tabularx}
+	\label{tab:variant-filtering-plants}
+\end{table}
+
 \stopsection

### 2022-06-08 scripts/R/correct_trait_xanthamonas.R
backup work applying method to Xanthomonas data

@@ -1,0 +1,121 @@
+# This script is to apply phylogenetic correction to QDR using POUMM parameters.
+# Run as: Run as: Rscript scripts/R/correct_trait.R
+
+FIGDIR <- "output_arabidopsis_xanthomonas"
+OUTDIR <- "output_arabidopsis_xanthomonas"
+
+require(ape)
+require(POUMM)
+require(dplyr)
+require(ggtree)
+require(ggplot2)
+
+source("scripts/R/functions/POUMM_utility_functions.R")
+
+# Load the tree
+print("Loading POUMM fit and tree.")
+tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_nj.newick", sep = "/"))
+
+# Parse trait data from tip labels (this ensures tips and trait values are in same order)
+tip_data <- data.frame(
+  label = tree$tip.label) %>%
+  tidyr::separate(label, into = c("id", "sampledate", "is_outgroup", "trait"), sep = "\\|", remove = F) %>%
+  mutate(trait = as.numeric(trait))
+
+# Load the estimated POUMM parameters
+POUMM_summary <- read.csv(file = paste(OUTDIR, "poumm_parameter_estimates.csv", sep = "/"), row.names = 1)
+
+# Make data structure for inferred POUMM params
+POUMM_params <- data.frame(
+  alpha  = POUMM_summary['alpha', 'PostMean'],
+  theta  = POUMM_summary['theta', 'PostMean'],
+  sigma  = POUMM_summary['sigma', 'PostMean'],
+  sigmae = POUMM_summary['sigmae', 'PostMean'],
+  g0     = POUMM_summary['g0', 'PostMean'])  # using posterior means
+
+# Apply phylo-correction
+paste("Inferring maximum likelihood pathogen and non-pathogen parts of trait.")
+inference_results <- do.call(
+  what = inferVHfromPOUMMfit,
+  args = c(POUMM_params, list(tree = tree, z = tip_data$trait)))
+
+inference_results <- cbind(tip_data, inference_results)
+
+# Write out phylo-corrected trait values
+write.csv(
+  x = inference_results,
+  file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"),
+  row.names = F)
+
+# Plot tree with phylo-corrected trait
+p <- ggtree(tr = tree) %<+% inference_results +
[…]

### 2022-06-08 scripts/R/fit_poumm_xanthamonas.R

@@ -1,0 +1,113 @@
+# This script is to fit the POUMM.
+# Run as: Rscript scripts/R/fit_poumm.R
+
+FIGDIR <- "output_arabidopsis_xanthomonas"
+OUTDIR <- "output_arabidopsis_xanthomonas"
+
+N_SAMPLES_MCMC <- 4E6
+N_ADAPT_MCMC <- 2E5
+THIN_MCMC <- 1000
+ACCEPTANCE_RATE_MCMC <- 0.01
+TREE_FN <- paste(OUTDIR, "pathogen_nj.newick", sep = "/")
+SEED <- 1500
+
+require(parallel)
+require(ape)
+require(tidyr)
+require(POUMM)
+require(bayesmeta)
+require(dplyr)
+require(ggplot2)
+# require(la)
+
+set.seed(SEED)  # Seed for POUMM parameter inference
+
+# Write out log file
+con <- file(paste(OUTDIR, "fit_poumm.log", sep = "/"))
+open(con, "w")
+writeLines(
+  con = con,
+  text = paste(
+    "n_samples_mcmc", N_SAMPLES_MCMC, "\n",
+    "n_adapt_mcmc", N_ADAPT_MCMC, "\n",
+    "thin_mcmc", THIN_MCMC, "\n",
+    "acceptance_rate_mcmc:", ACCEPTANCE_RATE_MCMC, "\n"))
+close(con)
+
+# Load the tree (Important: tree must be without outgroup)
+print("loading the tree")
+tree <- ape::read.tree(file = TREE_FN)
+
+# Parse trait data from tip labels
+tip_data <- data.frame(
+  label = tree$tip.label) %>%
+  tidyr::separate(label, into = c("id", "sampledate", "is_outgroup", "trait"), sep = "\\|", remove = F) %>%
+  mutate(trait = as.numeric(trait))
+
+# Fit the POUMM
+print("Inferring the POUMM parameters")
+POUMM_spec <- POUMM::specifyPOUMM_ATH2tMeanSeG0(
+  # par vector is c(alpha, theta, H2tMean, sigmae, g0)
+  z = tip_data$trait,
+  tree = tree,
+  nSamplesMCMC = N_SAMPLES_MCMC,
+  parPriorMCMC = function(par) {
+      # alpha
+      dexp(par[1], rate = 0.02, log = T) +
+      # g0 or theta
+      dnorm(par[2], mean = 0.4, sd = 0.2, log = T) + 
[…]

### 2022-06-08 scripts/R/make_nj_tree_xanthamonas.R

@@ -1,0 +1,78 @@
+# This script is to build a Neighbor-Joining tree from allele data by way of a similarity matrix
+
+require(dplyr)
+
+# Load data
+infile <- "data/arabidopsis_xanthomonas/pathogen_alleles.txt"
+metadatafile <- "data/arabidopsis_xanthomonas/full_phenotype_data.txt"
+
+alleles <- read.table(infile, header = T)
+metadata <- read.table(metadatafile, header = T, sep = "\t")
+
+# Generate genetic relatedness matrix
+alleles_clean <- alleles %>% 
+    select(-c(count, freq1, freq2, stat_pro, df, p)) %>%
+    rename_with(~ gsub("_allele$", "", .x))
+G <- t(as.matrix(alleles_clean))
+dist <- ape::dist.gene(G, method = "pairwise")
+
+# Build neighbor-joining tree
+tree_nj <- ape::nj(dist)
+
+# Plot tree
+png("output_arabidopsis_xanthomonas/tree.png")
+plot(tree_nj)
+dev.off()
+
+# Calculate phenotype values
+metadata_clean <- metadata %>%
+    mutate(
+        Strain = recode(
+            Strain,
+            "ME_DV_A37" = "MEDV_A37",
+            "ME_DV_P25" = "MEDV_P25",
+            "ME_DV_A40" = "MEDV_A40",
+            "ME_DV_P39" = "MEDV_P39",
+        )
+    )
+
+metadata_long <- metadata_clean %>% pivot_longer(
+    cols = c("Leaf_1", "Leaf_2", "Leaf_3", "Leaf_4"),
+    names_to = "Leaf",
+    names_prefix = "Leaf_",
+    values_to = "QDR_score"
+)
+
+average_phenotype <- metadata_long %>%
+    mutate(QDR_score = as.integer(QDR_score)) %>%
+    group_by(Strain) %>%
+    summarize(
+        n_leaves = sum(!is.na(QDR_score)),
+        mean_QDR_score = mean(QDR_score, na.rm = T),
+        sd_QDR_score = sd(QDR_score, na.rm = T)
+    )
+
+# Plot phenotype value distribution
+trait_dist <- ggplot(data = average_phenotype, aes(x = mean_QDR_score)) + 
+    geom_histogram(bins = 10) + 
+    labs(x = "Mean QDR score")
+ggsave(
+    plot = trait_dist,
+    file = "output_arabidopsis_xanthomonas/trait_dist.png",
[…]

### 2022-06-08 scripts/R/functions/utility_functions.R
add alternate tree sim function to make trees larger than HIV tree

@@ -201,6 +201,12 @@ simulateHIVTreeGeometricBL <- function(N) {
   return(subsampled_HIV_tree)
 }
 
+simulateRandomTreeExpBL <- function(N) {
+  require(ape)
+  tree <- ape::rtree(n = N, br = rexp, rate = 1)
+  return(tree)
+}
+
 printExpectedH2FromParameterization <- function(tree = NULL, sigma = NULL, alpha = NULL, d = 2, delta, M, p, var.z) {
   # Use to determine M, p, delta values such that H2.h is reasonable value
   if (!(is.null(tree))) {

### 2022-06-08 scripts/R/simulate_for_estimator_accuracy.R

@@ -21,9 +21,14 @@ N <- config_values$N
 # Designate output file
 outfile <- generateOutfileName(outdir = "output", description = "estimator_accuracy_MLE")
 
-# Simulate HIV phylogeny
-print("Simulating HIV phylogeny.")
-tree <- simulateHIVTreeExpBL(N)
+if (config_values$is.HIVTree) {
+  # Simulate HIV phylogeny
+  print("Simulating HIV phylogeny.")
+  tree <- simulateHIVTreeExpBL(N)
+} else {
+  print("Simulating random phylogeny -- be careful with selection strength units!")
+  tree <- simulateRandomTreeExpBL(N)
+}
 
 param.list <- list(
   N = N,

### 2022-06-08 scripts/R/get_atomm_pathogen_genotypes.R
remove atomm file formatting scripts from remote

@@ -1,120 +1,0 @@
-# ==============================================================================
-# Turn pathogen genome alignment into ATOMM-formatted genotype matrix
-# Get phenotype file for ATOMM
-# ==============================================================================
-
-require(ape)
-require(tidyr)
-require(dplyr)
-require(readr)
-
-aln_file <- "output/pathogen_trimmed.fasta"  # prepared alignment
-
-# Read in alignment, convert to matrix
-aln <- ape::read.dna(file = aln_file, format = "fasta", as.matrix = T, as.character = T)
-
-# Get pathogen info, including pathogen-host-phenotype correspondance
-path_info <- data.frame(
-    seq_name = rownames(aln)
-) %>% tidyr::separate(
-    col = seq_name, 
-    into = c('host_id', 'date', 'is_focal', 'spvl'), 
-    sep = "_",
-    remove = F
-)
-
-# Get host order in host genotype file
-hosts_str <- gsub("\n", "", read_file("output/atomm_sequence_host_header.txt"))
-hosts <- strsplit(hosts_str, split = "\t")[[1]]
-hosts <- hosts[7:length(hosts)]
-host_info <- data.frame(
-    atomm_sequence_host_col_index = seq_len(length(hosts)),
-    host_id = hosts
-) %>% tidyr::separate(
-    col = host_id,
-    into = c("host_id", "tmp"),
-    sep = "_"
-) %>% select(-tmp)
-
-# Remove sequences not included in GWAS (hosts filtered out, outgroup)
-path_host_info <- left_join(host_info, path_info, by = "host_id")
-
-rows_to_keep <- which(rownames(aln_raw) %in% path_host_info$seq_name)
-gwas_aln <- aln_raw[rows_to_keep, ]
-
-# Get consensus base at each position (excluding missing '-')
-get_mode <- function(c) {
-   uniqc <- unique(c[c != '-'])
-   mode <- uniqc[which.max(tabulate(match(c, uniqc)))]
-   return(mode)
-}
-consensus <- apply(X = gwas_aln, MARGIN = 2, FUN = get_mode)
-
-# Code SNPs as consensus (0) or non-consensus (1) ('-' assumed to be reference)
-get_snp_code <- function(consensus, r) {
-    non_ref <- r != consensus
-    non_del <- r != '-'
-    is_snp <- non_ref & non_del
[…]

### 2022-06-08 scripts/bash/get_atomm_host_genotypes.sh

@@ -1,54 +1,0 @@
-# ==============================================================================
-# Coverted PLINK BED file to haploid genotype matrix for ATOMM.
-# ==============================================================================
-
-# Run interactivelly with: 
-# docker run -it --rm --mount type=bind,src=$PWD/scripts,dst=/scripts --mount type=bind,src=$PWD/output,dst=/output plink
-# copy-paste these commands
-
-DATA_PREFIX="spVL.shcs.rs.filtered"
-MEM=10000
-
-# Make sure REF allele is the major allele
-./plink2 \
---bfile output/$DATA_PREFIX \
---maj-ref 'force' \
---make-bed \
---out output/$DATA_PREFIX.majref
-
-# Get variant dosages from .bed file (binary)
-./plink2 \
---bfile output/$DATA_PREFIX.majref \
---export A-transpose \
---memory $MEM \
---out output/$DATA_PREFIX.Av.majref
-
-# Make genotypes haploid
-head -1 output/$DATA_PREFIX.Av.majref.traw > output/atomm_sequence_host_header.txt
-
-### TO REMOVE ###
-# Down-sample host SNPs for first-pass analysis try
-head -1000 output/$DATA_PREFIX.Av.majref.traw > output/$DATA_PREFIX.Av.majref.traw.subsample
-
-# Reference allele is counted, so make missing data = 2 reference alleles
-sed 's/\tNA/\t2/g' output/$DATA_PREFIX.Av.majref.traw.subsample| tail -n +2 > output/$DATA_PREFIX.Av.haptmp1.traw
-# heterozygotes -> homozygous alternate allele (0)
-sed 's/\t1/\t0/g' output/$DATA_PREFIX.Av.haptmp1.traw | tail -n +2 > output/$DATA_PREFIX.Av.haptmp2.traw
-# homozygous reference allele -> 1
-sed 's/\t2/\t1/g' output/$DATA_PREFIX.Av.haptmp2.traw | tail -n +2 > output/$DATA_PREFIX.Av.hap.traw
-rm output/$DATA_PREFIX.Av.haptmp1.traw output/$DATA_PREFIX.Av.haptmp2.traw
-
-# File looks like this now:
-# CHR     SNP     (C)M    POS     COUNTED ALT     <host1>     <host2>     <host3> 
-# 1       1:768116:A:AGTTTT       0       768116  AGTTTT  A       0       1       1 
-
-# Format for ATOMM
[…]

### 2022-06-11 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 \begin{abstract}
-	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) may significantly influence a trait of interest.
-Previous GWAS have successfully identified host genetic variants associated with important infectious disease traits, such as HIV-1 set point viral load (spVL).
-However, these GWAS do not account for potentially confounding or extraneous pathogen effects that are heritable from donor to recipient in transmission chains.
-We propose a new method to consider the full genome of each patient's infecting pathogen strain, remove strain-specific effects on a trait based on the pathogen phylogeny, and thus better estimate the effect of human genetic variants on infectious disease traits.
-In simulations, we show our method can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
+	Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait.
+Traditional GWAS assume individual samples are independent observations.
+However, pathogen effects on an infectious disease trait can be heritable from donor to recipient in transmission chains.
+Thus, residuals in GWAS association tests for host genetic effects may not be independent due to shared pathogen ancestry.
+We propose a new method to estimate and remove heritable pathogen effects on a trait based on the pathogen phylogeny prior to host GWAS, thus restoring independence of samples.
+In simulations, we show this pre-processing step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
 When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
@@
 A key goal of genome-wide association studies (GWAS) is to understand the genetic basis of phenotypic variation among individuals.
-In a typical GWAS, millions of genetic variants from across the human genome are screened for statistical association with a trait of interest.
+In a typical GWAS, millions of genetic variants from across an organism's genome are screened for statistical association with a trait of interest.
 Ideally, this procedure identifies variants that are located in, or are in linkage disequilibrium with, alleles that directly affect the trait.
 If GWAS finds a variant strongly associated with a disease trait, the gene product may be a good drug target \citep{Okada2014}.
-Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify high-risk individuals \citep{Dudbridge2013}.
-
-% Why is spVL and important trait
-For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with the severity of disease course.
+Even if no single variant has a strong association, many small associations can be aggregated into a polygenic risk score to identify susceptible individuals \citep{Dudbridge2013}.
+
+% Overview of relevant methods and why we don't use them
+It is well-known that GWAS can be sensitive to confounding variables.
+Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait.
+Corrections for these types of population structure are well-developed and widely accepted  \citep{Astle2009, Price2006}.
+More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}.
+Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \citep{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \citep{collins_phylogenetic_nodate}.
+% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
+\citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry.
+These approaches are designed to quantify genetic effects from one organism on a trait.
+
+In the infectious disease context, genetic effects from two organisms - the host and the pathogen - may affect an infectious disease trait.
+GWAS using paired host-pathogen genotype data have previously been done to elucidate the marginal and interaction effects of host and pathogen genetic variants.
+Methods to account for microbial population structure when testing for marginal host associations or host-pathogen interaction effects include adding the microbial kinship matrix as a random effect in a linear mixed model as in \citep{Wang2018Two-wayGenomes} and using principle components derived from either this matrix or the pathogen phylogeny as covariates in a linear model as in \citep{Naret2018}.
+These methods focus on capturing and accounting for correlations due to the pathogen phylogeny, without further investigating the nature of these correlations.
+%However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).
+
+In this work, we draw from the field of phylogenetic comparative methods to propose a new two-step framework to correct for pathogen population structure and satisfy the GWAS assumption of independent samples.
+It is a pre-processing step for GWAS that first fits an evolutionary model to trait data and the pathogen phylogeny to estimate the correlation structure of the trait due to pathogen effects.
+On the basis of this inference, it estimates and removes pathogen effects on the trait so that downstream GWAS association testing with host genetic variants can be performed as normal under the assumption of independent samples.
+The main advantage of this approach compared to the previously outlined methods to correct for pathogen population structure are that it generates additional information on the evolutionary dynamics of the trait in the pathogen population.
+It is also restricted to a pre-processing step, which allows use of existing highly optimized tools to perform GWAS association tests under a variety of models.
+This framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.
+% We hypothesis this should increase GWAS power for several reasons.
+For one, pathogen effects may be confounding or extraneous variables that produce distorted or spurious associations between host genetic variants and a disease trait in host GWAS.
+We hypothesized that our approach should improve GWAS power to identify host genetic variants broadly associated with disease traits.
+%This may occur if the virus affects the disease trait (the dependent variable in host GWAS) or if it affects both the disease trait and the host genotype (the independent variable in host GWAS).
+%For example, stratification in both the host and pathogen populations may yield spurious correlations in host GWAS that are really due to independent pathogen factors. 
+% Then, co-evolution between a pathogen and its host may yield distorted estimates for host genetic effects that are not broadly protective against all pathogen strains.
+Finally, trait values from a population are not truly independent samples, given that hosts closer in the transmission chain have more similar pathogen strains and therefore more similar trait values.
+
+% % Why is spVL and important trait
+% For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with the severity of disease course.
 spVL is generally defined to be the average concentration of viral RNA copies in host plasma during the asymptomatic phase of infection in the absence of treatment (see e.g.
@@
 
-% How do we measure the extent to which genetic factors determine spVL
-To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors as opposed to pathogen genetic factors.
+% % How do we measure the extent to which genetic factors determine spVL
+% To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors as opposed to pathogen genetic factors.
 Heritability is a key measure of how genetically-determined a trait is.
 On the host side, narrow-sense heritability, the fraction of total trait variance explained by additive genetic effects, is a typical output of GWAS.
-% How heritable is spVL (human side)
-Several GWAS have measured the narrow-sense heritability of spVL and identified associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
+% % How heritable is spVL (human side)
+% Several GWAS have measured the narrow-sense heritability of spVL and identified associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
 The largest study to-date by \citet{McLaren2015} estimated that additive human genetic effects account for approximately 25\% of the variation in spVL amongst infected individuals.
@@
 % How heritable is spVL (viral side)
-On the pathogen side, broad-sense heritability $H^2$ is typically measured.
+% On the pathogen side, broad-sense heritability $H^2$ is typically measured.
 This is the fraction of total trait variance due to any factors ``inherited'' from infection partner to infection partner (the most obvious being pathogen genetic factors).
@@
 
-% The gap
-Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL.
+% % The gap
+% Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL.
 In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors.
 This has several potential negative consequences.
-Viral effects may be (i) confounding or (ii) extraneous variables that produce distorted or spurious associations between host genetic variants and a disease trait in host GWAS.
-%This may occur if the virus affects the disease trait (the dependent variable in host GWAS) or if it affects both the disease trait and the host genotype (the independent variable in host GWAS).
-For example, stratification in both the host and pathogen populations may yield spurious correlations in host GWAS that are really due to independent pathogen factors.
-Then, co-evolution between a pathogen and its host may yield distorted estimates for host genetic effects that are not broadly protective against all pathogen strains.
-Finally, (iii) spVL values from a cohort are not truly independent samples, given that patients closer in the transmission chain have more similar strains and therefore more similar spVL values.
-
-% Overview of relevant methods and why we don't use them
-Issue (iii) is closely related to a well-known problem in standard GWAS.
-Shared (human) ancestry, especially between close relatives, can also give rise to spurious genetic correlations with a trait.
-Corrections for these correlations are well-developed and widely accepted  \citep{Astle2009, Price2006}.
-More recently, \citet{Power2017} emphasized the need to do similar corrections for shared pathogen ancestry in microbial GWAS.
-Several state-of-the-art methods exist for this, e.g.
-\citep{Earle2016, collins_phylogenetic_nodate}.
-% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
-\citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry.
-However, these approaches are only suitable to quantify effects from $pathogen$ genetic variants on a trait.
-In contrast, we want to estimate effects of $human$ genetic variants on a trait, while accounting for pathogen effects.
-\citet{Naret2018} developed a relevant method for this in the context of a genome-to-genome GWAS framework.
-The authors suggest adding principle components derived from the pathogen phylogeny as covariates to the linear regression models for association testing.
-This should correct for trait correlations due to shared pathogen ancestry.
-However, the top principle components capture only some of the information from the full pathogen phylogeny.
-Furthermore, we would like to simultaneously address issues (i) and (ii).
-
-% Introduction to our method
-In this work,  we draw from the field of phylogenetic comparative methods to develop a new GWAS framework that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny.
-Our approach should help identify human genetic variants that affect disease traits and more accurately estimate their effects.
-
-In the following we describe a statistical model for the spVL trait, derive a maximum likelihood estimate for the viral part of spVL under this model, and describe a new infectious disease GWAS framework using this information.
-In simulations, we show that this framework can improve GWAS power to detect host genetic variants that affect disease traits.
-Finally, we apply our framework to human and viral genome data from the Swiss HIV Cohort Study (SHCS) and show that associations with spVL are robust to a correction for viral effects.
-Although we developed our framework in the context of HIV-1 spVL, this approach can readily be applied to other heritable infectious disease traits.
-
-\section*{New Approaches}
-\subsection*{A statistical model for spVL}
-
-Variation in spVL comes from several sources: direct host genetic effects, pathogen effects, interaction effects between the host and the pathogen, and other environmental effects.
-n GWAS, hosts are typically selected to not share recent ancestry.
+% n GWAS, hosts are typically selected to not share recent ancestry.
 Further, shared environmental variables are included as covariates to avoid spurious associations.
 We argue similar considerations must be taken
-for pathogen effects that heritable from one transmission partner to another \citep{Leventhal2016}.
-To characterize these effects, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}.
-PMMs assume continuous traits like spVL are the sum of independent heritable and non-heritable parts.
-In our case, pathogen effects comprise the heritable part and all other effects comprise the non-heritable part.
-The heritable part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A.
+% for pathogen effects that heritable from one transmission partner to another \citep{Leventhal2016}.
+
+% Introduction to our method
+In the following, we describe the evolutionary model for heritable, continuous-valued infectious disease traits upon which our method is based.
+We then derive a maximum likelihood estimate for the pathogen part of a trait under this model, and describe a new infectious disease GWAS framework using this information.
+In simulations, we show that this framework can improve GWAS power to detect host genetic variants that affect disease traits.
+Finally, we apply our framework to paired host-pathogen genotyping data from the Swiss HIV Cohort Study (SHCS) and a previously studied \emph{Arabidosis thaliana}-\emph{Xanthomonas arboricola} pathosystem.
+We show that associations with set-point viral load (spVL) and quantitative disease resistance (QDR) traits, respectively, are robust to a correction for pathogen effects.
+
+\section*{New Approaches}
+\subsection*{A statistical model for heritable, continuous-valued infectious disease traits}
+
+Variation in infectious disease traits like viral load or infection severity can come from several sources.
+These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature.
+GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
+This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
+Broad-sense heritability $H^2$ quantifies the fraction of total trait variance that is ``inherited'' from infection partner to infection partner.
+To characterize $H^2$ and the heritable and non-heritable factors that determine infectious disease traits, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}.
+PMMs assume continuous traits are the sum of independent heritable and non-heritable parts.
+In the infectious disease GWAS case, we assume the heritable part comprises pathogen genetic factors and all other factors are non-heritable.
+The heritable pathogen part is modeled by a random process occurring in continuous time along the branches of the pathogen phylogeny, as in Figure \ref{fig:spVLModel}A.
 The non-heritable part is modeled as Gaussian noise added to sampled individuals at the tips of the phylogeny.
@@
 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
-	\caption{A high-level schematic of our POUMM-based simulation framework.
-(A) shows how pathogen genetic effects on spVL evolve along the pathogen phylogeny according to an Ornstein-Uhlenbeck process.
-(B) shows how host genetic effects are the sum of independent effects from several causal variants.
+	\caption{A high-level schematic of our POUMM-based simulation framework in the context of HIV-1 spVL.
+(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
+(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
 Each variant can be present in 0, 1, or 2 copies.
 Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
-(C) shows how environmental effects are independently drawn from a Gaussian distribution centered at 0.
+(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
 These three effects sum to the trait value for each simulated individual.}
@@
 
-So far, PMMs with two types of random processes have been used to model spVL evolution.
+PMMs have previously been applied to the study of infectious disease traits using two different types of random processes to model trait evolution.
 The Brownian Motion (BM) process assumes unbounded trait values, i.e.
-spVL can attain any value.
+the trait can attain any value.
 The Ornstein-Uhlenbeck (OU) process assumes trait values fluctuate around an optimal value, i.e.
-extreme spVL values are unlikely.
-\citet{Mitov2018} and \citet{Bertels2018} previously showed the OU process has higher statistical support for spVL.
+extreme trait values are unlikely.
+Here, we assume the more flexible OU process as it encompasses a wider variety of evolutionary scenarios.
+For example, \citet{Mitov2018} and \citet{Bertels2018} previously showed the OU process has higher statistical support for HIV-1 spVL.
 This makes sense given that spVL is likely under stabilizing selection to maximize viral transmission potential \citep{Fraser2014}.
-Therefore, we assume the OU process.
 The full model is called the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) and is described in detail by \citet{Mitov2018}.
-Here, we review the main points in the spVL context.
-
-Under the POUMM, the spVL trait $z$ is the sum of viral effects $g_v$, host genetic effects $g_h$, and other environmental or interaction effects $\epsilon$.
-We can group the non-heritable effects $g_h$ and $\epsilon$ into a broader category of ``environmental'' effects $l$:
+Here, we review the main points relevant to our method.
+
+Under the POUMM, the trait $z$ is the sum of heritable genetic effects $g$, i.e.
+due to the pathogen, and non-heritable ``environmental'' effects $\epsilon$, i.e.
+host genetic effects and other environmental or interaction effects:
 
 \begin{equation}
-    z = g_{v} + l
+    z = g + \epsilon
 \end{equation}
-$g_v$ is a viral trait that evolves along the phylogeny according to an OU process.
+$g$ is a pathogen trait that evolves along the phylogeny according to an OU process.
 The OU process is defined by a stochastic differential equation with two terms.
@@
 \begin{split}
-	&dg_v(t) = \alpha[\theta - g_v(t)]dt + \sigma dW_t \\
-	&g_v(0) = g_0 
+	&dg(t) = \alpha[\theta - g(t)]dt + \sigma dW_t \\
+	&g(0) = g_0 
 	\label{eq:OUprocess}
@@
 Finally, $dW_t$ is the Wiener process underlying Brownian motion.
-The OU process is a Gaussian process, meaning that $g_v(t)$ is a Gaussian random variable.
-Assuming $g_v(t)$ starts at initial value $g_0$ at time $t = 0$ at the root of the phylogeny, we can write the expectation for $g_v(t)$ at time $t$:
+The OU process is a Gaussian process, meaning that $g(t)$ is a Gaussian random variable.
+Assuming $g(t)$ starts at initial value $g_0$ at time $t = 0$ at the root of the phylogeny, we can write the expectation for $g(t)$ at time $t$:
 
 \begin{equation}
-   E[g_v(t)] = g_0e^{-\alpha t} + (1 - e^{-\alpha t})\theta \label{eq:OUmean}
+   E[g(t)] = g_0e^{-\alpha t} + (1 - e^{-\alpha t})\theta \label{eq:OUmean}
 \end{equation}
-and the variance in $g_v(t)$ if we were to repeat the random evolutionary process many times \citep{Butler2004}:
+and the variance in $g(t)$ if we were to repeat the random evolutionary process many times \citep{Butler2004}:
 \begin{equation}
-    Var[g_v(t)] = \frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) \label{eq:OUVar}
+    Var[g(t)] = \frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) \label{eq:OUVar}
 \end{equation}
 
-$g_v$ evolves independently in descendent lineages after a divergence event in the phylogeny.
-The covariance between $g_v(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \citep{Butler2004}:
+$g$ evolves independently in descendent lineages after a divergence event in the phylogeny.
+The covariance between $g(t)$ in a lineage $i$ at time $t_i$ and another lineage $j$ at time $t_j$, $Cov\big(g_{i}(t_i), g_{j}(t_j)\big)$, increases with the amount of time between $t_0$ and the divergence of the two lineages, $t_{0(ij)}$, and decreases with the total amount of time the lineages evolve independently, $d_{ij}$ \citep{Butler2004}:
 
 \begin{equation}
-	Cov\big(g_{v_i}(t_i), g_{v_j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
+	Cov\big(g_{i}(t_i), g_{j}(t_j)\big) = \frac{\sigma^2}{2\alpha}[e^{-\alpha d_{ij}}(1 - e^{-2\alpha t_{0(ij)}})]
 	\label{eq:OUcov}
 \end{equation}
-Next, we remember that $l$ is the non-heritable, environmental part of spVL.
-$l$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
-The expectation of $l$ is 0, meaning environmental effects are equally likely to raise or lower spVL from the virus-determined level.
-The parameter $\sigma_l^2$ measures the between-host variance of the environmental effect.
+Next, we remember that $\epsilon$ is the non-heritable part of the trait.
+$\epsilon$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
+The expectation of $\epsilon$ is 0, meaning environmental effects are equally likely to raise or lower the trait from the pathogen-determined level.
+The parameter $\sigma_\epsilon^2$ measures the between-host variance of the environmental effect.
 
@@
 \begin{split}
-	E(l) = 0 \\
-	Var(l) = \sigma^2_l
+	E(\epsilon) = 0 \\
+	Var(\epsilon) = \sigma^2_\epsilon
 \end{split}
@@
 \begin{equation}
-	H^2_{\bar{t}} = \frac{Var[g_v(t)]}{Var[g_v(t)] + Var(l)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_l}
+	H^2_{\bar{t}} = \frac{Var[g(t)]}{Var[g(t)] + Var(\epsilon)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_\epsilon}
+	\label{eq:POUMM-H2}
 \end{equation}
 
-\subsection*{Teasing apart pathogen and non-pathogen effects on spVL}
-
-Given the assumptions of the POUMM, we can estimate a heritable pathogen effect on spVL and a non-heritable, host and environmental effect on spVL.
-Here, we derive a maximum-likelihood estimate for these values for individuals in a cohort, given measured spVL values and a pathogen phylogeny linking the infecting strains.
-
-Let $\bm{g}_v(\bm{t})$ be a vector of $g_v$ values, one for each individual in the cohort.
+\subsection*{Teasing apart pathogen and non-pathogen effects on a trait}
+
+Given the assumptions of the POUMM, we can estimate a heritable pathogen effect on a trait and a corresponding non-heritable, host and environmental effect.
+Here, we derive a maximum-likelihood estimate for these values for individuals in a GWAS cohort, given measured trait values and a pathogen phylogeny linking the infecting strains.
+
+Let $\bm{g}(\bm{t})$ be a vector of $g$ values, one for each individual in the cohort.
 $\bm{t}$ are the sampling times of each individual relative to the root of the phylogeny.
 To simplify notation, we omit the $\bm{t}$ from here on.
-$\bm{g}_v$ is a realization of a Gaussian random vector $\bm{G_v} \sim \mathcal{N}\big(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU}\big)$.
+$\bm{g}$ is a realization of a Gaussian random vector $\bm{G} \sim \mathcal{N}\big(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU}\big)$.
 The expectation $\bm{\mu}_{OU}$ is defined by equation \ref{eq:OUmean}, the diagonal elements of the covariance matrix $\boldsymbol{\Sigma}_{OU}$ are defined by equation \ref{eq:OUVar}, and the off-diagonal elements of $\boldsymbol{\Sigma}_{OU}$ by equation \ref{eq:OUcov}.
-
-Similarly, let $\bm{l}$ be a vector of the environmental part of spVL for each individual.
-$\bm{l}$ is a realization of a Gaussian random vector $\bm{L} \sim \mathcal{N}\big(\bm{0}, \boldsymbol{\Sigma}_L\big)$, where $\boldsymbol{\Sigma}_L$ is a diagonal matrix with diagonal elements equal to $\sigma^2_l$.
-
-Considering that $\bm{G_v}$ and $\bm{L}$ are independent random vectors and that their realizations $\bm{g}_v$ and $\bm{l}$ must sum together to equal the observed spVL values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}_v$ and $\bm{l}$:
+Similarly, let $\bm{\epsilon}$ be a vector of the environmental part of the trait for each individual.
+$\bm{\epsilon}$ is a realization of a Gaussian random vector $\bm{\mathcal{E}} \sim \mathcal{N}\big(\bm{0}, \boldsymbol{\Sigma}_\mathcal{E}\big)$, where $\boldsymbol{\Sigma}_\mathcal{E}$ is a diagonal matrix with diagonal elements equal to $\sigma^2_\epsilon$.
+
+Considering that $\bm{G}$ and $\bm{\mathcal{E}}$ are independent random vectors and that their realizations $\bm{g}$ and $\bm{\epsilon}$ must sum together to equal the observed trait values $\bm{z}$, we can write the following proportionality for the joint probability density of $\bm{g}$ and $\bm{\epsilon}$:
 
 \begin{equation}
-	f\big(\bm{g}_v, \bm{l}\big) \propto \mathcal{N}\big(\bm{g}_v; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
+	f\big(\bm{g}, \bm{\epsilon}\big) \propto \mathcal{N}\big(\bm{g}; \bm{\mu}_{G}, \boldsymbol{\Sigma}_G\big)
 	\label{eq:pdfGprop}
@@
 
-where the expected value of $\bm{g}_v$ and the covariance matrix $\boldsymbol{\Sigma}_G$ are defined as:
+where the expected value of $\bm{g}$ and the covariance matrix $\boldsymbol{\Sigma}_G$ are defined as:
 
 \begin{align}
-	Exp(\bm{g}_v) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_L^{-1} \bm{z}\big) \label{eq:MuG}\\
-	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_L^{-1}\big)^{-1} \label{eq:SigmaG}
+	Exp(\bm{g}) = \bm{\mu}_{G} &=  \boldsymbol{\Sigma}_G\big(\boldsymbol{\Sigma}_{OU}^{-1}\bm{\mu}_{OU} + \boldsymbol{\Sigma}_\mathcal{E}^{-1} \bm{z}\big) \label{eq:MuG}\\
+	\boldsymbol{\Sigma}_G &= \big(\boldsymbol{\Sigma}_{OU}^{-1} + \boldsymbol{\Sigma}_\mathcal{E}^{-1}\big)^{-1} \label{eq:SigmaG}
 \end{align}
@@
 	\begin{split}
-		f\big(\bm{g}_v,\ \bm{l}\big) &= f\big(\bm{g}_v|\ \bm{l}\big) \times f\big(\bm{l}\big) \\
-	&= f\big(\bm{g}_v\big) \times f\big(\bm{l}\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{l};\ \bm{0}, \mathbf{\Sigma}_L\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g}_v;\ \bm{0}, \mathbf{\Sigma}_L\big) \\
-	&= \mathcal{N}\big(\bm{g}_v;\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g}_v;\ \bm{z}, \mathbf{\Sigma}_L\big)
+		f\big(\bm{g},\ \bm{\epsilon}\big) &= f\big(\bm{g}|\ \bm{\epsilon}\big) \times f\big(\bm{\epsilon}\big) \\
+	&= f\big(\bm{g}\big) \times f\big(\bm{\epsilon}\big) \\
+	&= \mathcal{N}\big(\bm{g};\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{\epsilon};\ \bm{0}, \mathbf{\Sigma}_\mathcal{E}\big) \\
+	&= \mathcal{N}\big(\bm{g};\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{z} - \bm{g};\ \bm{0}, \mathbf{\Sigma}_\mathcal{E}\big) \\
+	&= \mathcal{N}\big(\bm{g};\ \bm{\mu}_{OU}, \mathbf{\Sigma}_{OU}\big) \times \mathcal{N}\big(\bm{g};\ \bm{z}, \mathbf{\Sigma}_\mathcal{E}\big)
 	\end{split}
@@
 
-Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the viral effect on spVL, taking into account all available information - measured spVL, the pathogen phylogeny, and inferred POUMM parameters.
-This estimator is an inverse-variance weighted average of measured spVL ($\bm{z}$) and information from the POUMM evolutionary model ($\bm{\mu}_{OU}$).
-In other words, $g_v$ will be closer to measured spVL if spVL is not very heritable.
-If spVL is highly heritable, $g_v$ will be closer to the expected value under the POUMM, i.e.
+Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the pathogen effect on the trait, taking into account all available information - measured trait values, the pathogen phylogeny, and inferred POUMM parameters.
+This estimator is an inverse-variance weighted average of measured trait ($\bm{z}$) and information from the POUMM evolutionary model ($\bm{\mu}_{OU}$).
+In other words, $g$ will be closer to the measured trait value if the trait is not very heritable.
+If the trait is highly heritable, $g$ will be closer to the expected value under the POUMM, i.e.
 take more information from the phylogenetic relationships between infecting strains.
 
-Given the estimator we just derived for $\bm{g}_v$, we can now estimate $\bm{l}$, the spVL value $without$ pathogen effects:
+Given the estimator we just derived for $\bm{g}$, we can now estimate $\bm{\epsilon}$, the trait value $without$ pathogen effects:
 
 \begin{equation}
-	\hat{\bm{l}} = \bm{z} - Exp(\bm{g}_v)
+	\hat{\bm{\epsilon}} = \bm{z} - Exp(\bm{g})
 	\label{eq:EHat}
@@
 \begin{enumerate}
-	\item Sample host genotypes, trait values, and pathogen genome sequence data from a cohort.
+	\item Sample paired host genotypes, pathogen genome sequences, and trait values from a cohort.
 	\item Construct a pathogen phylogeny using the pathogen genome sequences.
@@
 To test the theoretical best-case performance of our method, we simulated data under the POUMM and applied our framework to the simulated data.
-Figure \ref{fig:spVLModel} shows a high-level schematic of our simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
-For a description of the full simulation scheme, see Figure \ref{fig:sim-design}.
-In a nutshell, we simulated independent, additive host genetic effects, independent environmental effects, and heritable pathogen genetic effects under different scenarios of trait heritability and selection strength.
-To maintain the same heritability while varying selection strength, we counter-balanced by varying the intensity of stochastic evolutionary fluctuations accordingly.
-We fixed other variables to plausible values based on the spVL literature.
+We parameterized our simulation scheme with the time-scale and other parameters of an HIV-1 outbreak in mind, with spVL as the trait of interest.
+
+We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in \cite{Hodcroft2014}.
+Then, we simulated pathogen trait values $g$ along this phylogeny using the POUMM package in R \citep{Mitov2017a-POUMM}.
+For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
+The intensity of stochastic fluctuations parameter $\sigma$ was determined based $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:POUMM-H2} given the partition of trait variance described below).
+These values ranged from XX to XX, depending on alpha.
+We recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
+
+We paired each tip's simulated pathogen trait value with a simulated host trait value.
+Simulated hosts had 100 genome positions.
+We sampled alleles (0, 1, or 2) for each position from a binomial distribution with probability 0.13.
+10 random positions had an effect size of 0.2 on the trait and 10 had an effect size of -0.2.
+The other 80 positions had no effect on the trait.
+This parameterization produced roughly normally distributed host trait values centered at 0 with variance equal to 25\% of the total trait variance, which we constrained to 0.73 based on the variance in log spVL values measured by \cite{Mitov2018}.
+We used 25\% host heritability for spVL based on \cite{McLaren2015}.
+
+Finally, we sampled an additional random environmental effect for each tip from a normal distribution centered at 0.
+The variance of this distribution was scaled based on the pathogen heritability of the trait, from 0 (no affect) in the scenario with 75\% pathogen heritability and 25\% host heritability to 0.44 in the scenario with 15\% pathogen heritability and 25\% host heritability.
+Figure \ref{fig:sim-design} shows a high-level schematic of this simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
 
@@
 To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
-As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the spVL trait due to viral genetic effects and environmental effects is fixed.
+As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to viral genetic effects and environmental effects is fixed.
 Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.
@@
 		\includegraphics[width=0.7\linewidth]{figures/example_simulation_trait_values}
-		\caption{Example simulated data from two evolutionary scenarios where (left) a phylogenetic correction does not improve GWAS power and (right) where it does.
-The only difference in the two simulations is heritability; selection strength was fixed to 0.1 time $^{-1}$ and all other parameters were fixed as in the full simulation study.
-These examples correspond to the empty points in Figure \ref{fig:simulationResults}.}
+		\caption{Example simulated data from two evolutionary scenarios where (left) a phylogenetic correction to trait values does not improve GWAS power and (right) where it does.
+These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
+Apparent is that the phylogenetically-corrected trait values (y-axis) correspond better with the true host part of the trait when trait heritability due to the pathogen is higher (right side).
+Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.
+}
 		\label{fig:simulationTraitExample}
@@
 
+\subsection*{GWAS on the \emph{A. thaliana}-\emph{X. arboricola} pathosystem}
+
 \section*{Discussion}
@@
 So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
+Co-infection, or any other non-genetic factor that increases variability in the trait of interest, would reduce the effectiveness of this method.
+The more environmental ``noise'' in the trait value, the less signal there is to tease apart pathogen effects.
+There is also less signal for host genetic effects, so GWAS in general suffer when there are confounding factors affecting a trait of interest.
+In this case, we would recommend adding co-infection status as a covariate in the GWAS association test.
 
@@
 Other studies have estimated slightly lower values though (Table \ref{tab:spVLmeasurements}).
-After allotting 25\% of this variance to the host part of spVL $g_h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between the viral part $g_v$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
-$g_h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
-Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $g_h$ had the appropriate variance (see Table \ref{tab:sim-params}).
+After allotting 25\% of this variance to the host part of spVL $h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between the viral part $g$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
+$h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
+Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $h$ had the appropriate variance (see Table \ref{tab:sim-params}).
 We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}.
-$g_v$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
+$g$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
 For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on fitting the same model to SCHS data (Table \ref{tab:POUMM-parameters-our-estimates}).
 This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}).
-To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g_v$ were attributable to selection and drift, respectively.
+To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g$ were attributable to selection and drift, respectively.
 Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.
@@
 		$H^2_{\bar{t}}$ & varied & Viral heritability of spVL at $\bar{t}$ \\
-		$\sigma^2_{g_h}$ & $\sigma_{g_h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
-		$\sigma^2_{g_v}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in viral part of spVL at $\bar{t}$ \\ 
-		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{g_v} - \sigma^2_{g_h}$ & Variance in environmental part of spVL \\ 
+		$\sigma^2_{h}$ & $\sigma_{h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
+		$\sigma^2_{p}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in viral part of spVL at $\bar{t}$ \\ 
+		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{p} - \sigma^2_{h}$ & Variance in environmental part of spVL \\ 
 		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in viral phylogeny \\ 
@@
 		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
-		$g_{0}$ & 4.5 log copies/mL & $g_v$ at the root of the phylogeny  \\ 
+		$g_{0}$ & 4.5 log copies/mL & $g$ at the root of the phylogeny  \\ 
 		$\alpha$ & varied & Selection strength of OU process \\ 
-		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
+		$\sigma$ & $\sigma = \sqrt{\frac{-2\alpha H^2 \sigma^2_z (0.75 - H^2)}{(H^2 - 1)(1 - exp(-2\alpha\bar{t}))}}$ & Time-unit standard deviation of OU process \\ 
 		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Viral phylogeny \\ 
@@
 \emph{A.
-thaliana} genotypes were downloaded from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}.
-Quantitative diseases resistance (QDR) score measurements and \emph{X.
-arboricola} genotypes were provided by the authors of \citep{Wang2018Two-wayGenomes}.
-The \emph{X.
-arboricola} QDR trait values were generated in a fully-crossed experimental design, where each host type (accession) was infected with each pathogen strain.
-QDR was measured on 4 leaves for each host-pathogen pairing.
-
-For phylogenetic inference, the \emph{X.
-arboricola} genotype matrix consisted of allele dosage information for 33,610 SNPs in 22 strains.
-We used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from these data.
-We calculated mean QDR across all leaves infected on all hosts for each \emph{X.
+thaliana} and \emph{X.
+arboricola} genotyping and quantitative disease resistance (QDR) measurements are described in detail in \citep{Wang2018Two-wayGenomes}.
+Briefly, \citep{Wang2018Two-wayGenomes} infected different \emph{A.
+thaliana} host accessions with different \emph{X.
+arboricola} pathogen strains in a fully-crossed experimental design.
+They infected up to 4 leaves on each of three biological replicates for each host-pathogen pair.
+Then, they scored QDR for each leaf on a scale of 0 (resistant) to 4 (susceptible).
+A genotype matrix with allele dosage of 33,610 SNPs for the 22 \emph{X.
+arboricola} pathogen strains used was downloaded from the supplemental material of \citep{Wang2018Two-wayGenomes}.
+A VCF file with allele dosage of 12,883,854 SNPs for different \emph{A.
+thaliana} accessions was downloaded from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}.
+QDR measurements were provided directly by the authors of \citep{Wang2018Two-wayGenomes}.
+
+For phylogenetic inference, we used the ``dist.gene'' and ``nj'' functions from the ape package in R to construct a pairwise genetic distance matrix and then a neighbor-joining tree from the \emph{X.
+arboricola} pathogen genotype matrix.
+For the trait value to fit the POUMM, we calculated mean QDR across all leaves infected on all hosts for each \emph{X.
 arboricola} strain.
 
-For GWAS, we retained data from 133 \emph{A. thaliana} accessions with associated QDR measurements.
+For GWAS, we retained genotype data from 132 \emph{A.
+thaliana} accessions with associated QDR measurements.
+We used the options --maf 0.1 and --max-alleles 2 to remove variants with minor allele frequency less than 0.1 as in \citep{Wang2018Two-wayGenomes} and multi-allelic variants.
+After filtering, approximately 1.7 million genetic variants from the 132 accessions were retained for GWAS (Table \ref{tab:variant-filtering-plants}).
+For the GWAS trait value, we randomly selected one pathogen pairing per host accession (weighted by number of leaves scored across the three replicates), so that the data is paired rather than fully-crossed and thus analogous to the HIV spVL GWAS analysis.
 
@@
 For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.
-For the \emph{A.
-thaliana}-\emph{X.
-arboricola} trait correction and downstream GWAS, we randomly selected one pathogen pairing for each host to analyze.
-This makes the dataset analogous to the observational HIV-1 dataset.
+As mentioned above, we randomly selected one \emph{X.
+arboricola} pathogen strain pairing per \emph{A.
+thaliana} host accession to be analogous to the human infectious disease case.
+Thus, trait values for GWAS were calculated as average QDR across three replicates for a particular pathogen-host pairing, while the strain-specific trait correction was calculated based on average QDR across all host pairings.
 All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.
@@
 \subsection*{Association testing}
-We performed two GWAS using the same human genotype data from the SHCS.
-For the first ``GWAS with standard trait value'' we used total calculated spVL ($\bm{z}$) as the response variable for association testing, replicating prior GWAS for host genetic determinants of spVL.
-For the second ``GWAS with estimated non-pathogen part of trait'' we replaced total spVL with the estimated non-viral component of spVL ($\hat{\bm{e}}$) as the response variable.
-Association testing was performed using a linear association model in PLINK version 2.3 \citep{Chang2015} with sex and the top 5 principle components of host genetic variation included as covariates.
-The sex and principle components covariates were included to reduce residual variance in spVL and control for confounding from host population structure, respectively.
+We performed two comparative GWAS for each infectious disease trait (HIV-1 spVL and \emph{X.
+arboricola} QDR).
+For the first ``GWAS with standard trait value'' we used the total (uncorrected) trait values ($\bm{z}$) as the response variable for association testing, replicating a standard GWAS set-up.
+For the second ``GWAS with estimated non-pathogen part of trait'' we replaced total trait values with the estimated non-pathogen component of the trait ($\hat{\bm{e}}$) as the response variable.
+Association testing was performed using a linear association model in PLINK version 2.3 \citep{Chang2015} with the top 5 principle components of host genetic variation included as covariates.
+In the HIV-1 spVL case, sex was also used as a covariate.
+These covariates were included to reduce residual variance in the trait values (sex) and control for confounding from host population structure (principle components).
 
@@
 The data may be shared on reasonable request to the Swiss HIV Cohort Study at http://www.shcs.ch.
+The \emph{X.
+arboricola} pathogen genotypes are available in the supplemental material of \citep{Wang2018Two-wayGenomes}, the \emph{A.
+thaliana} host genotypes are available at https://1001genomes.org/, and the \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR measurements are available on request to the authors of \citep{Wang2018Two-wayGenomes}.
 
@@
 Amstad for administrative assistance.
+We thank Joy Bergelson for sharing the \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR measurements.
 We also thank Michael Landis, who shared a LaTeX template for graphical model drawing.
@@
 \begin{equation}
-    RMSE = \sqrt{\sigma^2_{g_v} + \sigma^2_\epsilon}
+    RMSE = \sqrt{\sigma^2_{p} + \sigma^2_\epsilon}
 \end{equation}
 
-We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{g_h}$, are fixed parameters in our simulation scheme.
+We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{h}$, are fixed parameters in our simulation scheme.
 
@@
 \begin{split}
-    \sigma^2_{g_h} + \sigma^2_{g_v} + \sigma^2_\epsilon = \sigma^2_z \\
-    0.25 * \sigma^2_{z} + \sigma^2_{g_v} + \sigma^2_\epsilon = \sigma^2_z \\
-    \sigma^2_{g_v} + \sigma^2_\epsilon = 0.75 * \sigma^2_z \\
-    \sigma^2_{g_v} + \sigma^2_\epsilon = 0.75 * 0.73 \\
-    \sigma^2_{g_v} + \sigma^2_\epsilon = 0.55
+    \sigma^2_{h} + \sigma^2_{p} + \sigma^2_\epsilon = \sigma^2_z \\
+    0.25 * \sigma^2_{z} + \sigma^2_{p} + \sigma^2_\epsilon = \sigma^2_z \\
+    \sigma^2_{p} + \sigma^2_\epsilon = 0.75 * \sigma^2_z \\
+    \sigma^2_{p} + \sigma^2_\epsilon = 0.75 * 0.73 \\
+    \sigma^2_{p} + \sigma^2_\epsilon = 0.55
 \end{split}
@@
 	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {\textcolor{red}{$H^2_{\bar{t}}$}};
-	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{g_h}$};
+	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{h}$};
 	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
@@
 	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
-	\node[snode] (v) at (3.6, 3.25) {$\vec{g_v}$};
+	\node[snode] (v) at (3.6, 3.25) {$\vec{p}$};
 	\draw[taro] (theta) -- (v);
@@
 	\node[constnode] (d) at (1.8, 1.5) {2};
-	\node[detnode] (p) at (3.6, 1.5) {$p$};
+	\node[detnode] (p) at (3.6, 1.5) {$g$};
 	\node[snode] (G) at (1.8, 3) {$G$};
@@
 	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
-	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{g_h}$};
+	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{h}$};
 	\node[constnode] (delta) at (3.6, 3) {$\delta$};
-	\node[detnode] (h) at (3.6, 4.5) {$\vec{g_h}$};
+	\node[detnode] (h) at (3.6, 4.5) {$\vec{h}$};
 	\node[constnode] (M) at (5.4, 3) {$M$};
@@
 We assume half the variants have a positive effect and half have a negative effect.
-The allele frequency $p$ for the causal variants set so that we achieve the desired variance in the host effects.
+The allele frequency $g$ for the causal variants set so that we achieve the desired variance in the host effects.
 (D) shows that the environmental effect is drawn from a Gaussian distribution with mean zero and variance as determined in part (A).}
@@
 	Raw data & 12883854 \\
-	Bi-allelic variants & 11769920 \\ \hline			
+	Bi-allelic variants & 11769920 \\
+	Minor allele frequency $<$ 0.1 & 1743952 \\ \hline
 	\end{tabularx}

### 2022-06-11 scripts/R/functions/POUMM_utility_functions.R
fix error in sigma calculation for simulation parameterization

@@ -1,7 +1,15 @@
+# This function is incorrect! It was used for the initial manuscript submission
+# calculateSigmaFromParameterization_old <- function(tree, alpha, H2, var.z) {
+#   var.gv <- H2 * var.z
+#   t.avg <- mean(POUMM::nodeTimes(tree = tree, tipsOnly = TRUE))
+#   sigma <- sqrt((2 * alpha * var.gv) / (1 - exp(-2 * alpha * t.avg)))
+#   return(sigma)
+# }
+
+# This function is correct! It was used for the manuscript revision
 calculateSigmaFromParameterization <- function(tree, alpha, H2, var.z) {
-  var.gv <- H2 * var.z
   t.avg <- mean(POUMM::nodeTimes(tree = tree, tipsOnly = TRUE))
-  sigma <- sqrt((2 * alpha * var.gv) / (1 - exp(-2 * alpha * t.avg)))
+  sigma <- sqrt((-2 * alpha * H2 * (var.z * (0.75 - H2))) / ((H2 - 1) * (1 - exp(-2 * alpha * t.avg))))
   return(sigma)
 }
 

### 2022-06-11 scripts/R/make_manuscript_figures.rmd
add trees to example simulation figure

@@
 OUTDIR <- "output"
+OUTDIR_XANTHAMONAS <- "output_arabidopsis_xanthomonas"
 
@@
 ```{r}
-accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
-tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
-example_simulation_results <- read.delim("output_revisions/2022-05-30_simulateFullDataset_2.txt")
+accuracy_simulation_results <- read.delim("output_revisions/2022-06-11_estimator_accuracy_MLE.txt")
+tpr_simulation_results <- read.delim("output_revisions/2022-06-11_GWAS_TPR_MLE.txt")
+example_simulation_results <- read.delim("output_revisions/2022-06-11_simulateFullDataset.txt")
 tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
@@
 gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
+gwas_results_xanthomonas <- read.delim(file = paste(OUTDIR_XANTHAMONAS, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
 spvl_values <- read.csv(file = paste(OUTDIR, "spvl_values.csv", sep = "/"))
@@
 top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1.03, 3), labels = c("", "A"))
-png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)
+png(filename = "figures/simulation_results_v2.png", width = linewidth, height = height, unit = "cm", res = 300)
 plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))
@@
 ```{r}
-p1 <- ggplot(data = example_simulation_results %>% rename("Heritability" = H2), aes(x = h, y = h.MWA)) +
+# Re-generate simulation tree, attach trait data to it
+set.seed(1)
+simulation_tree <- simulateHIVTreeExpBL(500)
+for (idx in 501:550) {
+        single_clade <- ape::extract.clade(phy = simulation_tree, node = idx)
+        if (length(single_clade$tip.label) > 8 && length(single_clade$tip.label) < 30) {
+                print(idx)
+        }
+}
+single_clade <- ape::extract.clade(phy = simulation_tree, node = 516)
+
+example_data <- example_simulation_results %>% rename("label" = "sample", "Heritability" = "H2")
+example_data_lo_h2 <- example_data %>% filter(Heritability == 0.15)
+example_data_hi_h2 <- example_data %>% filter(Heritability == 0.65)
+
+# Plot trait values on a clade in the tree
+mar <- 20
+lo_h2_tree <- ggtree(single_clade) %<+% example_data_lo_h2 + 
+        geom_tippoint(aes(color = v), size = 3) + 
+        scale_color_viridis_b(name = "Pathogen\npart of trait") + 
+        theme_tree(plot.margin=margin(mar, mar, mar, mar))
+
+hi_h2_tree <- ggtree(single_clade) %<+% example_data_hi_h2 + 
+        geom_tippoint(aes(color = v), size = 3) + 
+        scale_color_viridis_b(name = "Pathogen\npart of trait") + 
+        theme_tree(plot.margin=margin(mar, mar, mar, mar))
+
+# Plot trait value accuracy
+p1 <- ggplot(data = example_data, aes(x = h, y = h.MWA)) +
       geom_point() +
@@
       labs(x = "Simulated host part of trait", y = ehat_name)
-ggsave(plot = p1, filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
-      width = linewidth * 2/3, height = height * 2/3, units = "cm")
+
+# Arrange plots together
+top_row <- plot_grid(
+        lo_h2_tree + theme(legend.position = "none"),
+        hi_h2_tree + theme(legend.position = "none"),
+        labels = c("A", "B"),
+        nrow = 1
+)
+plot_grid(top_row, get_legend(lo_h2_tree),
+          p1,
+          nrow = 2, align = "v", axis = "tb",
+          rel_heights = c(1, 1.5),
+          rel_widths = c(1, 0.2),
+          labels = c("", "", "C"))
+ggsave(filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
+      width = linewidth, height = height * 1.3, unit = "cm")
 ```
@@
 ```
+
+Q-Q plot of Arabidopsis-Xanthomonas GWAS results
+```{r}
+gwas_results_xanthomonas <- gwas_results_xanthomonas %>% filter(!is.na(P_standard))
+qq_data_standard <- data.frame(
+        observed_log10p = -log10(sort(gwas_results_xanthomonas$P_standard, decreasing = F)),
+        expected_log10p = -log10(ppoints(nrow(gwas_results_xanthomonas))),
+        gwas_type = standard_gwas_name)
+qq_data_phylo <- data.frame(
+        observed_log10p = -log10(sort(gwas_results_xanthomonas$P_corrected, decreasing = F)),
+        expected_log10p = -log10(ppoints(nrow(gwas_results_xanthomonas))),
+        gwas_type = corrected_gwas_name)
+qq_results <- rbind(qq_data_standard, qq_data_phylo)
+
+p <- ggplot(data = qq_results, aes(x = expected_log10p, y = observed_log10p)) +
+        geom_point() +
+        geom_abline(slope = 1, intercept = 0, linetype = "dashed",
+                    color = RColorBrewer::brewer.pal(n = 3, name = diverging_palette)[[1]]) +
+        facet_wrap(. ~ gwas_type) +
+        labs(x = expression("expected log"[10]*"(p)"), y = expression("observed log"[10]*"(p)")) +
+        shared_theme
+ggsave(plot = p, filename = paste(FIGDIR, "qq_plots_xanthomonas.png", sep = "/"),
+       width = linewidth, height = height * 2/3, units = "cm")
+```
+
+Manhattan plot of Arabidopsis-Xanthomonas GWAS results
+```{r}
+pt_size <- 1
+
+# Add genome position to data
+chr_lengths <- gwas_results_xanthomonas %>%
+        group_by(CHROM) %>%
+        summarise(chr_len = max(POS))
+chr_starts <- chr_lengths %>%
+        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
+        select(-chr_len)
+gwas_results_annotated <- left_join(gwas_results_xanthomonas, chr_starts, by = "CHROM")
+gwas_results_annotated <- gwas_results_annotated %>%
+        arrange(CHROM, POS) %>%
+        mutate(BP_position = CHR_start + POS)
+
+# This is maybe not the best way, but makes sure tick marks and labels are at the center of each chromosome
+x_axis_ticks <- chr_starts
+max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 5, "BP_position"])
+x_axis_ticks <- rbind(
+        x_axis_ticks,
+        data.frame(
+                CHROM = c(5, 6),
+                CHR_start = c(max_BP_position, max_BP_position + 1000)))
+x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
+
+# Format data for plotting
+results_long <- tidyr::pivot_longer(
+        data = gwas_results_annotated,
+        cols = c("P_standard", "P_corrected"),
+        names_to = "gwas_type",
+        names_prefix = "P_",
+        values_to = "P")
+
+# Fix facet labels for plotting
+results_long$gwas_type <- factor(
+        results_long$gwas_type,
+        levels = c("standard", "corrected"),
+        labels = c(standard_gwas_name, corrected_gwas_name))
+
+# Make comparative GWAS overview plot
+bonferroni_signifiance_threshold <- 0.05 / nrow(gwas_results_annotated)
+overview_comparison <- ggplot(
+        data = results_long %>% sample_frac(0.1),  # TODO: plot all of results, not sub-sample of them
+        aes(x = BP_position, y = -log10(P))) +
+        geom_point(aes(color = CHROM %% 2 == 0), size = pt_size) +
+        scale_color_manual(values = c("grey", "dark grey")) +
+        scale_x_continuous(
+                breaks = x_axis_ticks$midpoint,
+                labels = x_axis_ticks$CHROM,
+                expand = c(0.01, 0.01)) +
+        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
+        labs(x = "Chromosome", y = expression('-Log'[10]*'(p)')) +
+        geom_hline(
+                yintercept = -log10(bonferroni_signifiance_threshold),
+                linetype = "dashed") +
+        shared_theme +
+        theme(
+                panel.grid.major.x = element_blank(),
+                panel.grid.minor.x = element_blank(),
+                legend.position = "none"
+        ) +
+        facet_grid(. ~ gwas_type)
+ggsave(
+        plot = overview_comparison,
+        filename = paste(FIGDIR, "gwas_results_xanthomonas.png", sep = "/"),
+        width = linewidth, height = height / 2, unit = "cm"
+)
+```
+
+Get the most highly-associated and most changed SNPs from the Xanthomonas GWAS results
+```{r}
+# Most highly associated
+most_highly_associated_standard <- gwas_results_annotated %>% top_n(n = 5, wt = -P_standard)
+most_highly_associated_corrected <- gwas_results_annotated %>% top_n(n = 5, wt = -P_corrected)
+
+most_highly_associated <- rbind(
+        most_highly_associated_standard %>% mutate(type = "strongest association standard GWAS"),
+        most_highly_associated_corrected %>% mutate(type = "strongest association corrected GWAS"))
+
+tex_table_1 <- most_highly_associated %>%
+        ungroup() %>%
+        select(CHROM, POS, ID, BETA_standard, P_standard, BETA_corrected, P_corrected)
+
+print(xtable(tex_table_1, type = "latex", display=c("s", "s", "s", "s", "g","g","g","g")),
+      file = paste(FIGDIR, "strongest_associations_xanthamonas.tex", sep = "/"), math.style.exponents = TRUE)
+
+write.csv(x = most_highly_associated_mhc_ccr5,
+          file = paste(OUTDIR, "strongest_associations_xanthamonas.csv", sep = "/"), row.names = F)
+
+# Most changed: higher significance
+most_changed_positive <- gwas_results_annotated %>%
+        group_by(CHROM) %>%
+        top_n(n = 1, wt = -neg_log10_P_standard_minus_corrected)
+
+write.csv(x = most_changed_positive,
+          file = paste(OUTDIR, "most_increased_association_xanthomonas.csv", sep = "/"), row.names = F)
+```

### 2022-06-11 scripts/R/simulate_for_gwas_tpr.R
update simulate tree to optionally take non-HIV tree

@@ -17,15 +17,20 @@ source("scripts/R/functions/GWAS_utility_functions.R")
 set.seed(1)
 
 # Read parameters for data simulation from config file
-# config_values <- yaml::read_yaml(file = "../../config-simulation.yaml")
-config_values <- yaml::read_yaml(file = "config-simulation.yaml")
+config_values <- yaml::read_yaml(file = "../../config-simulation.yaml")
 N <- config_values$N
 
 # Designate output file
 outfile <- generateOutfileName(outdir = "output", description = "GWAS_TPR_MLE")
 
-# Simulate HIV phylogeny
-tree <- simulateHIVTreeExpBL(N)
+if (config_values$is.HIVTree) {
+  # Simulate HIV phylogeny
+  print("Simulating HIV phylogeny.")
+  tree <- simulateHIVTreeExpBL(N)
+} else {
+  print("Simulating random phylogeny -- be careful with selection strength units!")
+  tree <- simulateRandomTreeExpBL(N)
+}
 
 # Set constant parameters for data simulation
 param.list <- list(

### 2022-06-11 scripts/R/correct_trait_xanthamonas.R
update to calculate trait across three replicates

@@ -74,6 +74,12 @@ phenotypes <- read.table("data/arabidopsis_xanthomonas/full_phenotype_data.txt",
         Leaf_2 = as.integer(Leaf_2),
         Leaf_3 = as.integer(Leaf_3),
         Leaf_4 = as.integer(Leaf_4)
+    ) %>%
+    tidyr::pivot_longer(
+      cols = c("Leaf_1", "Leaf_2", "Leaf_3", "Leaf_4"), 
+      values_to = "QDR_score", 
+      names_to = "Leaf", 
+      names_prefix = "Leaf_"
     )
 
 # Load host genotype accession IDs from 1001 genomes project
@@ -81,28 +87,41 @@ host_genotype_accession_ids <- read.csv("data/arabidopsis_xanthomonas/host_acces
     select(c(V1, V3)) %>%
     rename("id" = "V1", "Name_acc" = "V3")
 
-# Filter to accesssions & strains with genotype data
+# Filter to accesssions & strains with genotype data, scored leaves
 phenotypes_filtered <- phenotypes %>%
+  filter(!is.na(QDR_score)) %>%
   filter(Strain %in% inference_results$id) %>%
   filter(Name_acc %in% host_genotype_accession_ids$Name_acc)
 
-# Pick a random host/pathogen pair for each host plant (from pairings with 4 leaves scored)
+# Calculate mean QDR score across 3 replicates for each plant/pathogen pairing
+mean_phenotypes <- phenotypes_filtered %>%
+  group_by(Strain, Name_acc) %>%
+  summarize(n_leaves_scored = n(), mean_QDR_score = mean(QDR_score), .groups = "drop")
+
+# Pick a random host/pathogen pair for each host plant (prioritizing more leaves scored)
 set.seed(54321)
-phenotypes_for_gwas <- phenotypes_filtered %>%
-    filter(!is.na(Leaf_1), !is.na(Leaf_2), !is.na(Leaf_3), !is.na(Leaf_4)) %>%
+phenotypes_for_gwas <- mean_phenotypes %>%
     group_by(Name_acc) %>%
-    slice_sample(n = 1) %>%
-    mutate(mean_QDR_score = mean(c(Leaf_1, Leaf_2, Leaf_3, Leaf_4)))
+    slice_sample(n = 1, weight_by = n_leaves_scored)
+phenotypes_for_gwas_summary <- phenotypes_for_gwas %>%
+  group_by(Strain) %>%
[…]

### 2022-06-11 scripts/R/make_manuscript_figures.rmd
fix dimensions of simulation example fig

@@
         scale_color_viridis_b(name = "Pathogen\npart of trait") + 
-        theme_tree(plot.margin=margin(mar, mar, mar, mar))
+        theme_tree(plot.margin=margin(0, mar, 0, mar))
 
@@
       shared_theme +
-      labs(x = "Simulated host part of trait", y = ehat_name)
+      labs(x = "Simulated host part of trait", y = "Estimated non-pathogen\npart of trait")
 
@@
 ggsave(filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
-      width = linewidth, height = height * 1.3, unit = "cm")
+      width = linewidth, height = height, unit = "cm")
 ```

### 2022-06-11 scripts/R/functions/POUMM_utility_functions.R
fix sigma calculation back, there was no mistake

@@ -1,15 +1,7 @@
-# This function is incorrect! It was used for the initial manuscript submission
-# calculateSigmaFromParameterization_old <- function(tree, alpha, H2, var.z) {
-#   var.gv <- H2 * var.z
-#   t.avg <- mean(POUMM::nodeTimes(tree = tree, tipsOnly = TRUE))
-#   sigma <- sqrt((2 * alpha * var.gv) / (1 - exp(-2 * alpha * t.avg)))
-#   return(sigma)
-# }
-
-# This function is correct! It was used for the manuscript revision
 calculateSigmaFromParameterization <- function(tree, alpha, H2, var.z) {
+  var.gv <- H2 * var.z
   t.avg <- mean(POUMM::nodeTimes(tree = tree, tipsOnly = TRUE))
-  sigma <- sqrt((-2 * alpha * H2 * (var.z * (0.75 - H2))) / ((H2 - 1) * (1 - exp(-2 * alpha * t.avg))))
+  sigma <- sqrt((2 * alpha * var.gv) / (1 - exp(-2 * alpha * t.avg)))
   return(sigma)
 }
 

### 2022-06-11 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait.
-Corrections for these types of population structure are well-developed and widely accepted  \citep{Astle2009, Price2006}.
+Corrections for these types of population structure in human GWAS cohorts are well-developed and widely accepted  \citep{Astle2009, Price2006}.
 More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}.
@@
 \begin{equation}
-	H^2_{\bar{t}} = \frac{Var[g(t)]}{Var[g(t)] + Var(\epsilon)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_\epsilon}
+	H^2_{t} = \frac{Var[g(t)]}{Var[g(t)] + Var(\epsilon)} = \frac{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t})}{\frac{\sigma^2}{2\alpha}(1 - e^{-2\alpha t}) + \sigma^2_\epsilon}
 	\label{eq:POUMM-H2}
@@
 For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
-The intensity of stochastic fluctuations parameter $\sigma$ was determined based $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:POUMM-H2} given the partition of trait variance described below).
-These values ranged from XX to XX, depending on alpha.
+The intensity of stochastic fluctuations parameter $\sigma$ was determined based $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}).
+The relationship between $\alpha$ and $\sigma$ at different values of $H^2$ is shown in Figure TODO.
 We recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
@@
 	\begin{center}
-		\includegraphics[width=0.7\linewidth]{figures/example_simulation_trait_values}
-		\caption{Example simulated data from two evolutionary scenarios where (left) a phylogenetic correction to trait values does not improve GWAS power and (right) where it does.
+		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
+		\caption{Example simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
 These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
-Apparent is that the phylogenetically-corrected trait values (y-axis) correspond better with the true host part of the trait when trait heritability due to the pathogen is higher (right side).
-Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.
-}
+(A) and (B) show simulated pathogen trait values for a sub-clade of the 500-tip phylogeny with either (A) pathogen heritability of 15\% or (B) 65\%.
+Pathogen trait values at clustered tips are (A) less or (B) more correlated, depending on the heritability.
+(C) shows that our method produces corrected trait values (y-axis) that better match the true simulated host part of the trait (x-axis) when pathogen heritability is high (right side).
+Selection strength was fixed to 5 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
 		\label{fig:simulationTraitExample}
@@
 		$\alpha$ & varied & Selection strength of OU process \\ 
-		$\sigma$ & $\sigma = \sqrt{\frac{-2\alpha H^2 \sigma^2_z (0.75 - H^2)}{(H^2 - 1)(1 - exp(-2\alpha\bar{t}))}}$ & Time-unit standard deviation of OU process \\ 
+		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
 		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Viral phylogeny \\ 

### 2022-06-11 scripts/R/make_manuscript_figures.rmd
fix up & add explanatory methods figures based on simulation

@@
         H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75), estimate = 0.15),
-        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 5))
 example_data1$inference.method <- factor(
@@
         inference.method = "h.MWA.p.value",
-        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.75),
-        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.65),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 5))
 example_data2$inference.method <- factor(
@@
 top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1.03, 3), labels = c("", "A"))
-png(filename = "figures/simulation_results_v2.png", width = linewidth, height = height, unit = "cm", res = 300)
+png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)
 plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))
@@
 
+Plot how sigma changes with alpha at different heritabilities
+```{r}
+set.seed(1)
+tree <- simulateHIVTreeExpBL(500)
+to_plot <- data.frame(alpha = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60))
+to_plot$sigma_H15 <- calculateSigmaFromParameterization(tree, to_plot$alpha, H2 = 0.15, 0.73)
+to_plot$sigma_H75 <- calculateSigmaFromParameterization(tree, to_plot$alpha, H2 = 0.75, 0.73)
+
+ggplot(data = to_plot, aes(x = alpha)) + 
+        geom_point(aes(y = sigma_H15, color = "15%")) + 
+        geom_point(aes(y = sigma_H75, color = "75%")) + 
+        scale_color_discrete(name = expression("Heritability (" * italic('H'^2) * ")")) + 
+        shared_theme + 
+        labs(x = expression("Selection strength (" * alpha * ")"),
+             y = expression("Intensity of stochastic fluctuations (" * sigma * ")"))
+ggsave(filename = "figures/sigma_vs_alpha.png", width = linewidth, height = height, units = "cm")
+```
+
 Example trait values from successful simulation
 ```{r}
-# Re-generate simulation tree, attach trait data to it
+# Re-generate simulation tree
 set.seed(1)
 simulation_tree <- simulateHIVTreeExpBL(500)
-for (idx in 501:550) {
-        single_clade <- ape::extract.clade(phy = simulation_tree, node = idx)
-        if (length(single_clade$tip.label) > 8 && length(single_clade$tip.label) < 30) {
-                print(idx)
-        }
-}
-single_clade <- ape::extract.clade(phy = simulation_tree, node = 516)
 
-example_data <- example_simulation_results %>% rename("label" = "sample", "Heritability" = "H2")
-example_data_lo_h2 <- example_data %>% filter(Heritability == 0.15)
-example_data_hi_h2 <- example_data %>% filter(Heritability == 0.65)
+# Get tree distances
+tree_distances <- as.data.frame(ape::cophenetic.phylo(simulation_tree))
+xy <- t(combn(colnames(tree_distances), 2))
+tree_distances_long <- data.frame(xy, dist=tree_distances[xy]) %>% 
+        rename("label_1" = "X1", "label_2" = "X2") %>%
+        mutate(dist_decile_rank = ntile(dist, 10))
+
+# Plot correlation between tips as function of tree distance for each simulation scenario
+slim_sim_results <- example_simulation_results %>% select(H2, alpha, runID, z, sample)
+tmp <- tree_distances_long %>% merge(slim_sim_results, by.x = "label_1", by.y = "sample") %>% rename("z_1" = "z")
+tmp2 <- tmp %>% merge(slim_sim_results %>% rename("label_2" = "sample")) %>% rename("z_2" = "z")
+
+tip_pair_data_corelations <- tmp2 %>%
+        group_by(H2, alpha, runID, dist_decile_rank) %>%
+        summarize(correlation = cor(z_1, z_2))
+
+ggplot(data = tip_pair_data_corelations, aes(x = factor(dist_decile_rank), y = correlation)) +
+        geom_boxplot() + 
+        facet_grid(alpha ~ H2, labeller = plot_labeller) + 
+        labs(x = "Tree distance decile", y = "Pearson correlation in trait value") + 
+        shared_theme
+ggsave(filename = "figures/tip_trait_correlation.png", width = linewidth, height = height, unit = "cm")
+
+pruned_tree <- ape::keep.tip(phy = simulation_tree, tip = sample(simulation_tree$tip.label, size = 12))  # keep some random tips
+# pruned_tree <- ape::extract.clade(phy = simulation_tree, 516)  # keep some small clade
+
+example_data_lo_h2 <- example_simulation_results %>% filter(H2 == 0.15, alpha == 0.1)
+min_runID <- min(example_data_lo_h2$runID)
+example_data_lo_h2 <- example_data_lo_h2 %>% filter(runID == min_runID)
+example_data_hi_h2 <- example_simulation_results %>% filter(H2 == 0.75, alpha == 0.1)
+min_runID <- min(example_data_hi_h2$runID)
+example_data_hi_h2 <- example_data_hi_h2 %>% filter(runID == min_runID)
+example_data <- rbind(example_data_lo_h2, example_data_hi_h2) %>% rename("label" = "sample", "Heritability" = "H2")
 
@@
 mar <- 20
-lo_h2_tree <- ggtree(single_clade) %<+% example_data_lo_h2 + 
-        geom_tippoint(aes(color = v), size = 3) + 
-        scale_color_viridis_b(name = "Pathogen\npart of trait") + 
+lo_h2_tree <- ggtree(pruned_tree) %<+% example_data_lo_h2 + 
+        geom_tippoint(aes(color = z), size = 3) + 
+        scale_color_viridis_b(name = "Trait value") + 
         theme_tree(plot.margin=margin(0, mar, 0, mar))
 
-hi_h2_tree <- ggtree(single_clade) %<+% example_data_hi_h2 + 
-        geom_tippoint(aes(color = v), size = 3) + 
-        scale_color_viridis_b(name = "Pathogen\npart of trait") + 
+hi_h2_tree <- ggtree(pruned_tree) %<+% example_data_hi_h2 + 
+        geom_tippoint(aes(color = z), size = 3) + 
+        scale_color_viridis_b(name = "Trait value") + 
         theme_tree(plot.margin=margin(0, mar, 0, mar))

### 2022-06-11 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
-(A) and (B) show simulated pathogen trait values for a sub-clade of the 500-tip phylogeny with either (A) pathogen heritability of 15\% or (B) 65\%.
+(A) and (B) show simulated pathogen trait values for 15 randomly selected tips from the simulated phylogeny where pathogen heritability was (A)  15\% or (B) 75\%.
 Pathogen trait values at clustered tips are (A) less or (B) more correlated, depending on the heritability.

### 2022-06-11 scripts/R/make_manuscript_figures.rmd
add y=x line to simulation example figure

@@
 example_data_lo_h2 <- example_simulation_results %>% filter(H2 == 0.15, alpha == 0.1)
-min_runID <- min(example_data_lo_h2$runID)
-example_data_lo_h2 <- example_data_lo_h2 %>% filter(runID == min_runID)
+selected_runID <- max(example_data_lo_h2$runID)
+example_data_lo_h2 <- example_data_lo_h2 %>% filter(runID == selected_runID)
 example_data_hi_h2 <- example_simulation_results %>% filter(H2 == 0.75, alpha == 0.1)
-min_runID <- min(example_data_hi_h2$runID)
-example_data_hi_h2 <- example_data_hi_h2 %>% filter(runID == min_runID)
+selected_runID <- max(example_data_hi_h2$runID)
+example_data_hi_h2 <- example_data_hi_h2 %>% filter(runID == selected_runID)
 example_data <- rbind(example_data_lo_h2, example_data_hi_h2) %>% rename("label" = "sample", "Heritability" = "H2")
@@
       geom_point() +
-      facet_grid(. ~ Heritability, labeller = "label_both") + 
+      facet_wrap(. ~ Heritability, labeller = "label_both") + 
+      geom_abline(intercept = 0, slope = 1) + 
       shared_theme +

### 2022-06-13 manuscript/hiv-gwas.tex
Update on Overleaf.

--- 2022-06-11 manuscript/hiv-gwas.tex
@@
 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
-		\caption{Example simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
+		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
 These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
-(A) and (B) show simulated pathogen trait values for a sub-clade of the 500-tip phylogeny with either (A) pathogen heritability of 15\% or (B) 65\%.
-Pathogen trait values at clustered tips are (A) less or (B) more correlated, depending on the heritability.
-(C) shows that our method produces corrected trait values (y-axis) that better match the true simulated host part of the trait (x-axis) when pathogen heritability is high (right side).
-Selection strength was fixed to 5 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
+(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively.
+Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
+(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%.
+The solid line is the y=x line.
+Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
 		\label{fig:simulationTraitExample}
@@
 \subsection*{GWAS on the \emph{A. thaliana}-\emph{X. arboricola} pathosystem}
+
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/gwas_results}
+		\caption{Results from comparative GWAS.
+(A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables.
+On the left, we used unmodified (total) spVL values.
+On the right, we used our estimates for the non-pathogen effects on spVL.
+The alternating shades correspond to different chromosomes.
+(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
+Base positions are with reference to genome build GRCh37.
+The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
+Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
+The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
+		\label{fig:gwas-results}
+	\end{center}
+\end{figure}
 

### 2022-06-13 scripts/R/make_manuscript_figures.rmd
update figures to include arboricola-thaliana results

@@
 library(xtable)
+library(ggedit)
 
@@
 ```{r}
-accuracy_simulation_results <- read.delim("output_revisions/2022-06-11_estimator_accuracy_MLE.txt")
-tpr_simulation_results <- read.delim("output_revisions/2022-06-11_GWAS_TPR_MLE.txt")
+accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
+tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
 example_simulation_results <- read.delim("output_revisions/2022-06-11_simulateFullDataset.txt")
 tree <- ape::read.tree(file = paste(OUTDIR, "pathogen_no_outgroup.newick", sep = "/"))
+tree_arboricola <- ape::read.tree(file = paste(OUTDIR_XANTHAMONAS, "pathogen_nj.newick", sep = "/"))
 traits <- read.csv(file = paste(OUTDIR, "poumm_corrected_traits.csv", sep = "/"))
+traits_arboricola <- read.csv(file = paste(OUTDIR_XANTHAMONAS, "poumm_corrected_traits.csv", sep = "/"))
 tree_w_trait_data <- treeio::full_join(tree, traits, by = "label")
+tree_w_trait_data_arboricola <- treeio::full_join(tree_arboricola, traits_arboricola, by = "label")
 POUMM_fit <- get(load(file = paste(OUTDIR, "POUMM_fit.RData", sep = "/")))
+POUMM_fit_arboricola <- get(load(file = paste(OUTDIR_XANTHAMONAS, "POUMM_fit.RData", sep = "/")))
 poumm_parameter_estimates <- read.csv(file = paste(OUTDIR, "poumm_parameter_estimates.csv", sep = "/"))
+poumm_parameter_estimates_arboricola <- read.csv(file = paste(OUTDIR_XANTHAMONAS, "poumm_parameter_estimates.csv", sep = "/"))
 pc_scores <- read.csv(file = paste(OUTDIR, "shcs_vs_hapmap_pc_scores.csv", sep = "/"))
@@
 ```{r}
-# Get POUMM values for HIV spVL to annotate plot with
+# Get POUMM values for HIV spVL and X. arboricola QDR to annotate plot with
 H2_estimate <- poumm_parameter_estimates[poumm_parameter_estimates$stat == "H2tMean", "PostMean"]
@@
 
+H2_estimate_arboricola <- poumm_parameter_estimates_arboricola [poumm_parameter_estimates_arboricola $stat == "H2tMean", "PostMean"]
+alpha_estimate_arboricola  <- poumm_parameter_estimates_arboricola [poumm_parameter_estimates_arboricola $stat == "alpha", "PostMean"]
+
 # Breaks scale is defined as seq(0.5, n.breaks - 0.5, 1) in plotting function, need to translate real values to breaks scale
 get_xval_for_sim_plots <- function(real_vals, estimate) {
+      if (estimate < min(real_vals)) { return(0.1) }  # an approximation, plot actually only starts at alpha = 0.1
       base_val <- rev(real_vals[real_vals <= estimate])[1]
@@
 
+qdr_data <- data.frame(
+        inference.method = "h.MWA.p.value",
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75), estimate = H2_estimate_arboricola),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = alpha_estimate_arboricola))
+qdr_data$inference.method <- factor(
+        qdr_data$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+
 # Get example simulation POUMM values to annotate plot with
@@
         H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75), estimate = 0.15),
-        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 5))
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))
 example_data1$inference.method <- factor(
@@
         H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.65),
-        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 5))
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))
 example_data2$inference.method <- factor(
@@
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
-        geom_point(data = spvl_data, fill = "black") + 
-        geom_point(data = example_data1, fill = "black", shape = 1) + 
-        geom_point(data = example_data2, fill = "black", shape = 1)
+        geom_point(data = spvl_data, fill = "black", aes(shape = "spVL")) + 
+        geom_point(data = qdr_data, fill = "black", aes(shape = "QDR")) + 
+        geom_point(data = example_data1, fill = "black", aes(shape = "Example")) + 
+        geom_point(data = example_data2, fill = "black", aes(shape = "Example")) +
+        scale_shape_manual(values = c("QDR" = 16, "spVL" = 4, "Example" = 1), name = element_blank())
 
@@
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
-        geom_point(data = spvl_data, fill = "black") + 
-        geom_point(data = example_data1, fill = "black", shape = 1) + 
-        geom_point(data = example_data2, fill = "black", shape = 1)
+        geom_point(data = spvl_data, fill = "black", aes(shape = "spVL")) + 
+        geom_point(data = qdr_data, fill = "black", aes(shape = "QDR")) + 
+        geom_point(data = example_data1, fill = "black", aes(shape = "Example")) + 
+        geom_point(data = example_data2, fill = "black", aes(shape = "Example")) +
+        scale_shape_manual(values = c("QDR" = 16, "spVL" = 4, "Example" = 1), name = element_blank())
 
-top_row <- plot_grid(NULL, rmse_plot, rel_widths = c(1.03, 3), labels = c("", "A"))
+shape_legend <- get_legend(rmse_plot + guides(fill = "none") + theme(legend.position = "bottom"))
+top_row <- plot_grid(NULL, rmse_plot + guides(shape = "none"), rel_widths = c(1.03, 3), labels = c("", "A"))
 png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)
-plot_grid(top_row, tpr_plot, align = "v", axis = "tb", nrow = 2, labels = c("", "B"))
+plot_grid(
+        top_row, 
+        tpr_plot + guides(shape = "none"), 
+        shape_legend, 
+        rel_heights = c(1, 1, 0.1),
+        align = "v", axis = "tb", nrow = 3, 
+        labels = c("", "B", "")
+)
 dev.off()
@@
         theme(legend.position = "bottom")
+# Remove maximum log-likelihood point
+p <- remove_geom(p, 'point')
 ggsave(plot = p, filename = paste(FIGDIR, "poumm_parameter_estimates.png", sep = "/"),
@@
 ```{r}
-pt_size <- 1
-
-# Add genome position to data
-chr_lengths <- gwas_results_xanthomonas %>%
-        group_by(CHROM) %>%
-        summarise(chr_len = max(POS))
-chr_starts <- chr_lengths %>%
-        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
-        select(-chr_len)
-gwas_results_annotated <- left_join(gwas_results_xanthomonas, chr_starts, by = "CHROM")
-gwas_results_annotated <- gwas_results_annotated %>%
-        arrange(CHROM, POS) %>%
-        mutate(BP_position = CHR_start + POS)
-
-# This is maybe not the best way, but makes sure tick marks and labels are at the center of each chromosome
-x_axis_ticks <- chr_starts
-max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 5, "BP_position"])
-x_axis_ticks <- rbind(
-        x_axis_ticks,
-        data.frame(
-                CHROM = c(5, 6),
-                CHR_start = c(max_BP_position, max_BP_position + 1000)))
-x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
-
-# Format data for plotting
-results_long <- tidyr::pivot_longer(
-        data = gwas_results_annotated,
-        cols = c("P_standard", "P_corrected"),
-        names_to = "gwas_type",
-        names_prefix = "P_",
-        values_to = "P")
-
-# Fix facet labels for plotting
-results_long$gwas_type <- factor(
-        results_long$gwas_type,
-        levels = c("standard", "corrected"),
-        labels = c(standard_gwas_name, corrected_gwas_name))
-
-# Make comparative GWAS overview plot
-bonferroni_signifiance_threshold <- 0.05 / nrow(gwas_results_annotated)
-overview_comparison <- ggplot(
-        data = results_long %>% sample_frac(0.1),  # TODO: plot all of results, not sub-sample of them
-        aes(x = BP_position, y = -log10(P))) +
-        geom_point(aes(color = CHROM %% 2 == 0), size = pt_size) +
-        scale_color_manual(values = c("grey", "dark grey")) +
-        scale_x_continuous(
-                breaks = x_axis_ticks$midpoint,
-                labels = x_axis_ticks$CHROM,
-                expand = c(0.01, 0.01)) +
-        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
-        labs(x = "Chromosome", y = expression('-Log'[10]*'(p)')) +
-        geom_hline(
-                yintercept = -log10(bonferroni_signifiance_threshold),
-                linetype = "dashed") +
-        shared_theme +
-        theme(
-                panel.grid.major.x = element_blank(),
-                panel.grid.minor.x = element_blank(),
-                legend.position = "none"
-        ) +
-        facet_grid(. ~ gwas_type)
-ggsave(
-        plot = overview_comparison,
-        filename = paste(FIGDIR, "gwas_results_xanthomonas.png", sep = "/"),
-        width = linewidth, height = height / 2, unit = "cm"
-)
+# See script: scripts/R/plot_manhattan_euler.R (plotting ran out of memory on my laptop)
 ```
@@
 ```
+Inferred phylogeny Xanthomonas
+```{r}
+legend_width <- 15
+
+# Plot tree with tip colors by spVL components
+legend_x_pos <- 0.99
+legend_y_pos <- 0.45
+
+x_max <- 20000
+v_tree <- ggtree(tree_w_trait_data_arboricola, ladderize = F) +
+        geom_tippoint(aes(color = v.MWA)) +
+        geom_tiplab(aes(label = id), size = 3, hjust = -0.1) + 
+        xlim(c(0, x_max)) +
+        scale_color_viridis_c(
+                option = "plasma",
+                direction = -1,
+                limits = c(0, max(traits_arboricola$trait))) +
+        theme(legend.position = "bottom") +
+        guides(color = guide_colorbar(title.position="top", title.hjust = 0.5)) +
+        labs(color = yulab.utils::str_wrap(gvhat_name, width = legend_width))
+
+h_tree <- ggtree(tree_w_trait_data_arboricola, ladderize = F) +
+        geom_tippoint(aes(color = h.MWA)) +
+        geom_tiplab(aes(label = id), size = 3, hjust = -0.1) + 
+        xlim(c(0, x_max)) +
+        scale_color_viridis_c() +
+        theme(legend.position = "bottom") +
+        guides(color = guide_colorbar(title.position="top", title.hjust = 0.5)) +
+        labs(color = yulab.utils::str_wrap(ehat_name, width = legend_width))
+
+qdr_tree_same_scale <- ggtree(tree_w_trait_data_arboricola, ladderize = F) +
+        geom_tippoint(aes(color = trait)) +
+        geom_tiplab(aes(label = id), size = 3, hjust = -0.1) + 
+        xlim(c(0, x_max)) +
+        scale_color_viridis_c(
+                option = "plasma",
+                direction = -1,
+                limits =  c(0, max(traits_arboricola$trait))) +
+        theme(legend.position = "bottom") +
+        guides(color = guide_colorbar(title.position="top", title.hjust = 0.5)) +
+        labs(color = yulab.utils::str_wrap("Calculated QDR", width = legend_width)) +
+        scale_fill_identity(name = '', guide = 'legend', labels = c('oob'))
+
+png(file = paste(FIGDIR, "qdr_on_tree_arboricola.png", sep = "/"),
+    height = height, width = linewidth,
+    units = "cm",
+    res = 300)
+plot_grid(qdr_tree_same_scale + theme(legend.position = "none"),
+          h_tree + theme(legend.position = "none"),
+          v_tree + theme(legend.position = "none"),
+          get_legend(qdr_tree_same_scale), get_legend(h_tree), get_legend(v_tree),
+          nrow = 2, align = "v", axis = "tb",
+          rel_heights = c(3, 1),
+          labels = c("A", "B", "C", "", "", ""))
+dev.off()
+```
+
+POUMM parameter estimates Xanthomonas
+```{r}
+# Generate figures, write out estimates
+plots <- plot(POUMM_fit_arboricola, doZoomIn = TRUE, doPlot = F)
+p <- plots$densplot +
+        scale_fill_manual(
+                values = RColorBrewer::brewer.pal(n = 3, name = diverging_palette),
+                aesthetics = c("fill", "color"),
+                labels = c("Prior", "Chain 1", "Chain 2"),
+                name = element_blank()) +
+        theme_bw() +
+        labs(x = "Value", y = "Density") +
+        scale_y_continuous(expand = c(0, 0)) +
+        theme(legend.position = "bottom")
+p <- remove_geom(p, 'point')
+ggsave(plot = p, filename = paste(FIGDIR, "poumm_parameter_estimates_arboricola.png", sep = "/"),
+       width = linewidth, height = height, units = "cm")
+```
+
+Estimated trait values Xanthomonas
+```{r}
+p1 <- ggplot(data = traits_arboricola, aes(x = trait, y = h.MWA)) +
+      geom_point() +
+      shared_theme +
+      labs(x = "Calculated QDR trait", y = ehat_name)
+p2 <- ggExtra::ggMarginal(p1, type="histogram")
+ggsave(plot = p2, filename = paste(FIGDIR, "trait_values_arboricola.png", sep = "/"),
+      width = linewidth * 2/3, height = height * 2/3, units = "cm")
+```

### 2022-06-13 scripts/R/plot_manhattan_euler.R

@@ -1,0 +1,98 @@
+# env2lmod
+# module load r/4.1.3
+# bsub -I -R "rusage[mem=40960]" "Rscript plot_manhattan_euler.R"
+
+print("Loading libraries")
+library(ggplot2)
+library(dplyr)
+library(tidyr)
+
+OUTDIR <- "output_arabidopsis_xanthomonas"
+
+print("Sourcing functions")
+source("scripts/R/functions/plot_functions.R")
+source("scripts/R/functions/utility_functions.R")
+
+standard_gwas_name <- "GWAS with standard trait value"
+corrected_gwas_name <- "GWAS with estimated non-pathogen part of trait"
+shared_theme <- theme_bw()
+linewidth <- 16.5  # cm
+height <- 11  # cm
+
+# Load data
+print("Loading data")
+gwas_results_xanthomonas <- read.delim(
+    file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), 
+    sep = "", 
+    header = T
+)
+
+pt_size <- 1
+
+# Add genome position to data
+chr_lengths <- gwas_results_xanthomonas %>%
+        group_by(CHROM) %>%
+        summarise(chr_len = max(POS))
+chr_starts <- chr_lengths %>%
+        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
+        select(-chr_len)
+gwas_results_annotated <- left_join(gwas_results_xanthomonas, chr_starts, by = "CHROM")
+gwas_results_annotated <- gwas_results_annotated %>%
+        arrange(CHROM, POS) %>%
+        mutate(BP_position = CHR_start + POS)
+
+# This is maybe not the best way, but makes sure tick marks and labels are at the center of each chromosome
+x_axis_ticks <- chr_starts
+max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 5, "BP_position"])
+x_axis_ticks <- rbind(
+        x_axis_ticks,
+        data.frame(
+                CHROM = c(5, 6),
+                CHR_start = c(max_BP_position, max_BP_position + 1000)))
+x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
+
+# Format data for plotting
+results_long <- tidyr::pivot_longer(
+        data = gwas_results_annotated,
+        cols = c("P_standard", "P_corrected"),
[…]

### 2022-06-13 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/gwas_results}
-		\caption{Results from comparative GWAS.
-(A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables.
-On the left, we used unmodified (total) spVL values.
-On the right, we used our estimates for the non-pathogen effects on spVL.
-The alternating shades correspond to different chromosomes.
-(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
-Base positions are with reference to genome build GRCh37.
-The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
-Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
-The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
-		\label{fig:gwas-results}
-	\end{center}
-\end{figure}
-
 \section*{Discussion}
@@
 		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on SHCS data.
-We ran two different MCMC chains to ensure the estimates converged.
-The black point on the x-axis shows the posterior mean value, which was used to estimate the pathogen- and non-pathogen effects on spVL.}
+		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on SHCS data. We ran two different MCMC chains to ensure the estimates converged.}
 \label{fig:poumm-parameters}
@@
 	\label{fig:PCA}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/qdr_on_tree_arboricola.png}
+	\caption{Inferred \emph{X. arboricola} phylogeny with tips colored by (A) calculated QDR, (B) estimated non-pathogen effects on QDR and (C) estimated pathogen effects on QDR.}
+	\label{fig:qdr-on-tree}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = 0.462\linewidth]{figures/trait_values_arboricola.png}
+	\caption{A comparison of measured (calculated) QDR values versus our estimated non-pathogen effect on QDR for each selected \emph{A.
+thaliana} host - \emph{X.
+arboricola} pathogen pairing selected for the GWAS.
+The histograms show the marginal distribution of each value across the individual pairings.}
+	\label{fig:trait-comparison-qdr}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates_arboricola.png}
+		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on \emph{A.
+thaliana}-\emph{X.
+arboricola} data.
+We ran two different MCMC chains to ensure the estimates converged.}
+\label{fig:poumm-parameters-qdr}
+\end{figure}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{figures/gwas_results_xanthomonas.png}
+	\caption{Results from comparative GWAS on \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR.
+Points are association p-values for the same \emph{A.
+thaliana} host variants in GWAS with two different response variables.
+On the left, we used unmodified (total) QDR values.
+On the right, we used our estimates for the non-pathogen effects on QDR.
+The alternating shades correspond to different chromosomes.
+The dashed lines show significance at $\alpha$ = 0.05 with a Bonferroni correction for multiple-testing.}
+	\label{fig:gwas-results-qdr}
 \end{figure}
@@
 	\label{tab:POUMM-parameters-our-estimates}
+\end{table}
+
+\begin{table}[H]
+\caption{POUMM parameter estimates for QDR based on \emph{A. thaliana}-\emph{X. arboricola} data data. HPD = Highest posterior density.}
+	\begin{tabularx}{\linewidth}{XXX} \hline 
+	Parameter & Posterior mean & 95\% HPD \\ \hline 
+	${g_{0}}$ & 0.40 & (0.01, 0.78) \\
+	$\theta$ & 0.39 & (0.30, 0.49) \\
+	$\sigma$ & 0.03 & (0.0, 0.06) \\
+	$\alpha$ & 0.03 & (0.0, 0.05) \\
+	$\sigma_e$ & 0.18 & (0.08, 0.27) \\ 
+	$H^2_{\bar{t}}$ & 0.33 & (0.0, 0.77) \\ \hline
+	\end{tabularx}
+	\label{tab:POUMM-parameters-our-estimates-qdr}
 \end{table}

### 2022-06-13 scripts/R/make_manuscript_figures.rmd
correct simulation fig to have H2 = 0.75 not 0.65

@@
         inference.method = "h.MWA.p.value",
-        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.65),
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.75),
         alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60), estimate = 0.1))

### 2022-06-13 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
-The black point represents our estimates for the heritability and selection strength of spVL based on Swiss HIV Cohort Study data.
-The empty points show simulated scenarios that are presented in more detail in Figure \ref{fig:simulationTraitExample}.
-RMSE = Root mean square error, TPR = True positive rate.}
+The points highlight specific heritability and selection strength values from the \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR analysis, HIV-1 spVL analysis, and two simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.
+RMSE = Root mean square error, TPR = True positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
 		\label{fig:simulationResults}
@@
 \begin{center}
-		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
-		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
+\includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
+	\caption{Correlations between trait values in pairs of tips in four of the simulated scenarios from Figure \ref{fig:simulationResults}.
+Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each evolutionary scenario.
+Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength (upper right facet).}
+	\label{fig:tip-trait-correlation}
+	\end{center}
+\end{figure}
+
+% \begin{figure}[H]
+% 	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
+% 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
 These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
@@
 Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
-		\label{fig:simulationTraitExample}
-	\end{center}
-\end{figure}
-
-\subsection*{GWAS on the Swiss HIV Cohort}
-
-Finally, we applied our framework to empirical data from the Swiss HIV Cohort Study (SHCS).
-We used data collected from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018.
+% 		\label{fig:simulationTraitExample}
+% 	\end{center}
+% \end{figure}
+
+\subsection*{Application to HIV-1 set-point viral load}
+
+We applied our framework to empirical data from two different host-pathogen systems.
+First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018.
 The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals.
@@
 We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}).
-To put these values into the context of our simulation study, they are shown as black points on Figure \ref{fig:simulationResults}.
+To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}.
 
 We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes.
-In the ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable.
-In the ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL.
+In the (i) ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable.
+In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL.
 Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS.
@@
 
-\subsection*{GWAS on the \emph{A. thaliana}-\emph{X. arboricola} pathosystem}
+\subsection*{Application to the \emph{A. thaliana}-\emph{X. arboricola} pathosystem}
+
+Next, we applied our framework to data collected from the \emph{A.
+thaliana}-\emph{X.
+arboricola} pathosystem by \cite{Wang2018Two-wayGenomes}.
+\cite{Wang2018Two-wayGenomes} performed a fully-crossed experiment in which they infected genetically diverse \emph{A.
+thaliana} accessions with genetically diverse strains of the phytopathogenic bacteria \emph{X.
+arboricola}.
+They scored quantitative disease resistance (QDR) on a scale of 0 (resistant) to 4 (susceptible) for up to four infected leaves for three replicates of each \emph{A.
+thaliana}-\emph{X.
+arboricola} pairing.
+Figure \ref{fig:qdr-on-tree} shows the inferred \emph{X.
+arboricola} pathogen phylogeny.
+Mean QDR calculated across all \emph{A.
+thaliana} pairings and all replicates for each \emph{X.
+arboricola} strain was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21.
+Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table \ref{tab:POUMM-parameters-our-estimates-qdr}).
+These values deviated significantly from the respective priors (Figure \ref{fig:poumm-parameters-qdr}).
+Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table \ref{tab:POUMM-parameters-our-estimates-qdr}).
+These selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.
+
+Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application).
+Thus, the phylogeny is not very informative for a trait value correction.
+Indeed, mean QDR values for each \emph{X.
+arboricola} strain calculated across all \emph{A.
+thaliana} pairings and all replicates is simply scaled by our method (Figure \ref{fig:trait-comparison-qdr}).
 
@@
 Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection.
-Using this model, we estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in the Swiss HIV Cohort Study.
+We applied this model to two different host-pathogen systems where paired host and pathogen genetic data was generated alongside a measure of pathogen virulence.
+First, we fit the POUMM to set-point viral load data from individuals living with HIV in Switzerland.
+We estimated HIV-1 spVL heritability to be 45\% (95\% HPD 24 - 67\%) in this cohort.
 Compared to previous studies, this estimate is at the higher end (see \citet{Mitov2018} and references therein).
@@
 
-Instead, the main novelty of our approach was to correct the spVL trait prior to performing the GWAS, thereby estimating and removing pathogen effects.
+For comparison, we also fit the POUMM to quantitative disease resistance measurements from \emph{A.
+thaliana} infected with the phytopathogenic bacteria \emph{X.
+arboricola}.
+We estimated \emph{X.
+arboricola} virulence heritability to be 33\% (95\% HPD 0 - 77\%).
+\citep{Wang2018Two-wayGenomes} originally estimated a QDR heritability of 44\% in this dataset, falling within the wide range of our estimate.
+We note that \cite{Wang2018Two-wayGenomes} used a linear mixed model in which the experimental unit is QDR scored on individual leaves, wherease our estimate is based on much coarser binning of QDR scores into a mean score across all leaves on all host accessions and all replicates (N = 22).
+Furthermore, the QDR score trait values were not truly continuous (scores were measured on an integer scale from 0 to 4).
+Thus, these data partially violate the assumptions of the POUMM.
+Compared to HIV-1, we estimate very low selection strength for virulence in \emph{X.
+arboricola}.
+As \cite{Wang2018Two-wayGenomes} explain, \emph{X.
+arboricola} strains with differing virulence can co-inhabit populations of \emph{A.
+thaliana}.
+This might also point to low selection on \emph{X.
+arboricola} virulence.
+Furthermore, expected correlation in virulence between related strains of \emph{X.
+arboricola} was smaller than for HIV-1.
+
+The main novelty of our approach is to use these POUMM parameter estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects.
 In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
 Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.
-
 Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals.
@@
 So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
-Co-infection, or any other non-genetic factor that increases variability in the trait of interest, would reduce the effectiveness of this method.
-The more environmental ``noise'' in the trait value, the less signal there is to tease apart pathogen effects.
-There is also less signal for host genetic effects, so GWAS in general suffer when there are confounding factors affecting a trait of interest.
-In this case, we would recommend adding co-infection status as a covariate in the GWAS association test.
-
-Given our estimates for the heritability of spVL and the selection strength on this trait using Swiss HIV cohort data, our simulation results reveal that we cannot expect a significant improvement in GWAS power for human genetic determinants of spVL (Figure \ref{fig:simulationResults}).
-Our method slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B).
+
+Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A.
+thaliana} quantitative disease resistance to \emph{X.
+arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power (Figure \ref{fig:simulationResults}).
+Certain pairs of samples in the HIV-1 cohort were indeed expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45).
+However, the overall effect on GWAS is small.
+For HIV-1 spVL, our phylogenetic correction slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B).
 Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM.
@@
 Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL is robust to our correction for pathogen effects.
-
-Our method is convenient for GWAS because it is simply a pre-processing step that produces an alternate response variable for GWAS association tests.
-It is still possible to use previously developed, well-documented, and fast tools for the actual association testing (we used PLINK \citep{Chang2015}).
-The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}.
+For \emph{A.
+thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+\cite{Wang2018Two-wayGenomes} found no significant \emph{A.
+thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects.
+As with HIV-1 spVL, our results do not challenge this previous finding.
+
+Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS.
+First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population.
+Then, it is a convenient pre-processing step for GWAS because it simply produces an alternate response variable for GWAS association tests.
+As in the \emph{A.
+thaliana}-\emph{X.
+arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using the corrected trait values in a GWAS.
+In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.
+In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
+
+Our method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}.
 All code for the simulations and HIV spVL analysis presented in this study is available on the project GitHub at https://github.com/cevo-public/POUMM-GWAS.
@@
 
-For GWAS, we retained genotype data from 132 \emph{A.
-thaliana} accessions with associated QDR measurements.
-We used the options --maf 0.1 and --max-alleles 2 to remove variants with minor allele frequency less than 0.1 as in \citep{Wang2018Two-wayGenomes} and multi-allelic variants.
-After filtering, approximately 1.7 million genetic variants from the 132 accessions were retained for GWAS (Table \ref{tab:variant-filtering-plants}).
-For the GWAS trait value, we randomly selected one pathogen pairing per host accession (weighted by number of leaves scored across the three replicates), so that the data is paired rather than fully-crossed and thus analogous to the HIV spVL GWAS analysis.
-
 \subsection*{POUMM parameter inference}
@@
 The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution.
-Figure \ref{fig:poumm-parameters} shows the posterior distributions for inferred parameters.
+Figures \ref{fig:poumm-parameters} and \ref{fig:poumm-parameters-qdr} show the posterior distributions for inferred parameters for HIV-1 spVL and \emph{X.
+arboricola} QDR, respectively.
 Table \ref{tab:POUMM-parameters-our-estimates} gives the posterior mean values used for subsequent calculations.
@@
 We corrected HIV-1 spVL and \emph{X.
-arboricola} QDR trait values using the method described in this paper.
-For each host individual, we estimated the pathogen part of their trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}.
+arboricola} mean QDR trait values using the method described in this paper.
+For each  individual, we estimated the pathogen part of the trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}.
+This is implemented in the function ``POUMM:::gPOUMM'' in the R package POUMM.
+In the HIV-1 case, each sample corresponds to one HIV-1 strain with one spVL value.
+In the \emph{X.
+arboricola} case, each sample corresponds to one \emph{X.
+arboricola} strain and the mean QDR score for that strain.
+To calculate the expected correlation in trait values between tips in the pathogen phylogeny, we used the function ``covVTipsGivenTreePOUMM'' in the same package.
 For the POUMM parameters $\alpha$, $\sigma$, $\theta$, and $\sigma_e$, we used the posterior mean estimates generated as described above.
-As mentioned above, we randomly selected one \emph{X.
-arboricola} pathogen strain pairing per \emph{A.
-thaliana} host accession to be analogous to the human infectious disease case.
-Thus, trait values for GWAS were calculated as average QDR across three replicates for a particular pathogen-host pairing, while the strain-specific trait correction was calculated based on average QDR across all host pairings.
 All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.
@@
 \subsection*{Association testing}
-We performed two comparative GWAS for each infectious disease trait (HIV-1 spVL and \emph{X.
-arboricola} QDR).
+We performed two GWAS using the same human genotype data from the SHCS.
 For the first ``GWAS with standard trait value'' we used the total (uncorrected) trait values ($\bm{z}$) as the response variable for association testing, replicating a standard GWAS set-up.
 For the second ``GWAS with estimated non-pathogen part of trait'' we replaced total trait values with the estimated non-pathogen component of the trait ($\hat{\bm{e}}$) as the response variable.
-Association testing was performed using a linear association model in PLINK version 2.3 \citep{Chang2015} with the top 5 principle components of host genetic variation included as covariates.
-In the HIV-1 spVL case, sex was also used as a covariate.
-These covariates were included to reduce residual variance in the trait values (sex) and control for confounding from host population structure (principle components).
+Association testing was performed using a linear association model in PLINK version 2.3 \citep{Chang2015} with sex and the top 5 principle components of host genetic variation included as covariates.
+The sex and principle components covariates were included to reduce residual variance in spVL and control for confounding from host population structure, respectively.
 
@@
 \begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.462\linewidth]{figures/trait_values_arboricola.png}
-	\caption{A comparison of measured (calculated) QDR values versus our estimated non-pathogen effect on QDR for each selected \emph{A.
-thaliana} host - \emph{X.
-arboricola} pathogen pairing selected for the GWAS.
-The histograms show the marginal distribution of each value across the individual pairings.}
-	\label{fig:trait-comparison-qdr}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
 	\centering
@@
 \begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{figures/gwas_results_xanthomonas.png}
-	\caption{Results from comparative GWAS on \emph{A.
-thaliana}-\emph{X.
-arboricola} QDR.
-Points are association p-values for the same \emph{A.
-thaliana} host variants in GWAS with two different response variables.
-On the left, we used unmodified (total) QDR values.
-On the right, we used our estimates for the non-pathogen effects on QDR.
-The alternating shades correspond to different chromosomes.
-The dashed lines show significance at $\alpha$ = 0.05 with a Bonferroni correction for multiple-testing.}
-	\label{fig:gwas-results-qdr}
+\begin{center}
+\includegraphics[width = 0.462\linewidth]{figures/trait_values_arboricola.png}
+	\caption{A comparison of measured (calculated) mean QDR values across all \emph{A.
+thaliana} accession pairings and replicates  versus our estimated non-pathogen effect on mean QDR for each \emph{X.
+arboricola} pathogen strain.
+The histograms show the marginal distribution of each value across the strains.
+The Pearson correlation coefficient is 1.}
+	\label{fig:trait-comparison-qdr}
+	\end{center}
 \end{figure}
@@
 
-\begin{table}[H]
-\caption{Number of variants for \emph{X. arboricola} QDR GWAS after sequential filtering steps.}
-	\begin{tabularx}{\linewidth}{XX}
-	\hline
-	Variant filter & Number of variants remaining \\
-	\hline 
-	Raw data & 12883854 \\
-	Bi-allelic variants & 11769920 \\
-	Minor allele frequency $<$ 0.1 & 1743952 \\ \hline
-	\end{tabularx}
-	\label{tab:variant-filtering-plants}
-\end{table}
-
 \stopsection

### 2022-06-13 scripts/R/make_manuscript_figures.rmd
update simulation fig to have other 2 example points

@@
         example_data2$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+example_data3 <- data.frame(
+        inference.method = "h.MWA.p.value",
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.15),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60, 70), estimate = 60))
+example_data3$inference.method <- factor(
+        example_data3$inference.method,
+        levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
+        labels = c(e_name, ehat_name, z_minus_zbar_name))
+example_data4 <- data.frame(
+        inference.method = "h.MWA.p.value",
+        H2 = get_xval_for_sim_plots(real_vals = c(0.15, 0.25, 0.35, 0.45, 0.55, 0.65, 0.75, 0.85), estimate = 0.75),
+        alpha = get_xval_for_sim_plots(real_vals = c(0.1, 0.5, 1, 5, 10, 20, 30, 40, 50, 60, 70), estimate = 60))
+example_data4$inference.method <- factor(
+        example_data4$inference.method,
         levels = c("h.plus.e.p.value", "h.MWA.p.value", "norm.z.p.value"),
@@
         geom_point(data = example_data2, fill = "black", aes(shape = "Example")) +
+        geom_point(data = example_data3, fill = "black", aes(shape = "Example")) + 
+        geom_point(data = example_data4, fill = "black", aes(shape = "Example")) +
         scale_shape_manual(values = c("QDR" = 16, "spVL" = 4, "Example" = 1), name = element_blank())
@@
 ```
+Compare variance-covariance matrix under POUMM
+```{r}
+# Get variance-covariance matrix from POUMM
+corr_HIV <- POUMM::covVTipsGivenTreePOUMM(
+        tree = tree, 
+        alpha = poumm_parameter_estimates[[which(poumm_parameter_estimates$stat == 'alpha'), 'PostMean']], 
+        sigma = poumm_parameter_estimates[[which(poumm_parameter_estimates$stat == 'sigma'), 'PostMean']], 
+        sigmae = poumm_parameter_estimates[[which(poumm_parameter_estimates$stat == 'sigmae'), 'PostMean']],
+        corr = T
+)
+corr_arboricola <- POUMM::covVTipsGivenTreePOUMM(
+        tree = tree_arboricola, 
+        alpha = poumm_parameter_estimates_arboricola[[which(poumm_parameter_estimates_arboricola$stat == 'alpha'), 'PostMean']], 
+        sigma = poumm_parameter_estimates_arboricola[[which(poumm_parameter_estimates_arboricola$stat == 'sigma'), 'PostMean']], 
+        sigmae = poumm_parameter_estimates_arboricola[[which(poumm_parameter_estimates_arboricola$stat == 'sigmae'), 'PostMean']],
+        corr = T
+)
+max(corr_HIV[upper.tri(corr_HIV)])
+min(corr_HIV[upper.tri(corr_HIV)])
+max(corr_arboricola[upper.tri(corr_arboricola)])
+min(corr_arboricola[upper.tri(corr_arboricola)])
+```

### 2022-06-13 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 In simulations, we show this pre-processing step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
-When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
+When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between set-point viral load (spVL) and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
 Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects.
+Then, we show that phylogenetic correlations for quantitative disease resistance (QDR) in the \emph{A.
+thaliana}-\emph{X.
+arboricola} pathosystem are not strong enough to justify a phylogenetic correction for GWAS.
 Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
@@
 For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
-The intensity of stochastic fluctuations parameter $\sigma$ was determined based $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}).
-The relationship between $\alpha$ and $\sigma$ at different values of $H^2$ is shown in Figure TODO.
-We recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
+The intensity of stochastic fluctuations parameter $\sigma$ was determined based $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}).
+As shown in Figure \ref{fig:sigma-vs-alpha}, higher $\alpha$ correspond to higher $\sigma$ to maintain constant $H^2$ under this parameterization.
+For each $H^2$ and $\alpha$ considered, we simulated recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
 
@@
 By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable, under low selection pressure, and with relatively moderate stochastic fluctuations compared to outbreak duration.
+Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength.
+These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny.
+This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait.
 This is because high heritability means the trait value is highly pathogen-dependent.
@@
 thaliana}-\emph{X.
-arboricola} QDR analysis, HIV-1 spVL analysis, and two simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.
+arboricola} QDR analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.
 RMSE = Root mean square error, TPR = True positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
 		\label{fig:simulationResults}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
+		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
+These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
+(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively.
+Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
+(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%.
+The solid line is the y=x line.
+Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
+		\label{fig:simulationTraitExample}
 	\end{center}
@@
 
-% \begin{figure}[H]
-% 	\begin{center}
-% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
-% 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
-These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
-(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively.
-Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
-(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%.
-The solid line is the y=x line.
-Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
-% 		\label{fig:simulationTraitExample}
-% 	\end{center}
-% \end{figure}
-
 \subsection*{Application to HIV-1 set-point viral load}
@@
 Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
-Figure \ref{fig:spvl-on-tree} shows that this trait is not strongly phylogenetically structured in the cohort, despite high heritability.
-Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
 We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}).
 To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}.
+The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
+However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.
+Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
 
@@
 \citep{Wang2018Two-wayGenomes} originally estimated a QDR heritability of 44\% in this dataset, falling within the wide range of our estimate.
-We note that \cite{Wang2018Two-wayGenomes} used a linear mixed model in which the experimental unit is QDR scored on individual leaves, wherease our estimate is based on much coarser binning of QDR scores into a mean score across all leaves on all host accessions and all replicates (N = 22).
+We note that \cite{Wang2018Two-wayGenomes} used a linear mixed model in which the experimental unit is QDR scored on individual leaves, whereas our estimate is based on much coarser binning of QDR scores into a mean score across all leaves on all host accessions and all replicates (N = 22).
 Furthermore, the QDR score trait values were not truly continuous (scores were measured on an integer scale from 0 to 4).
@@
 In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
+Finally, additional model complexity can be added to the GWAS association tests.
+For instance, our method does not account for co-infection, although one could add co-infection status as a covariate in the GWAS association test to account for this variable.
 
@@
 This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}).
-To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the OU parameters for selection strength, $\alpha$, and intensity of random fluctuations, $\sigma$, so that different proportions of the variability in $g$ were attributable to selection and drift, respectively.
+To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the heritability $H^2$ and selection strength $\alpha$ parameters.
+The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table \ref{tab:sim-params}, Figure \ref{fig:sigma-vs-alpha}).
 Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.
@@
 \begin{center}
+\includegraphics[width = 0.7\linewidth]{figures/sigma_vs_alpha.png}
+	\caption{Relationship between the selection strength parameter $\alpha$ and the intensity of stochastic evolutionary fluctuations parameter $\sigma$ at two different heritability ($H^2$) values in the  simulation scheme.
+$\sigma$ was determined as a function of $\alpha$ and $H^2$ under the POUMM (function given in Table \ref{tab:sim-params}).}
+	\label{fig:sigma-vs-alpha}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
 \includegraphics[width = 0.7\linewidth]{figures/spvl_calculation_comparison.png}

### 2022-06-13 scripts/R/make_manuscript_figures.rmd
fix labels on tip trait cor figure

@@
 
-ggplot(data = tip_pair_data_corelations, aes(x = factor(dist_decile_rank), y = correlation)) +
+tip_pair_data_to_plot <- tip_pair_data_corelations %>%
+        rename("Selection strength" = alpha) %>%
+        mutate(Heritability = factor(H2, levels = c(0.75, 0.15)))
+
+ggplot(data = tip_pair_data_to_plot, aes(x = factor(dist_decile_rank), y = correlation)) +
         geom_boxplot() + 
-        facet_grid(alpha ~ H2, labeller = plot_labeller) + 
+        facet_grid(Heritability ~ `Selection strength`, labeller = label_both) + 
         labs(x = "Tree distance decile", y = "Pearson correlation in trait value") + 
         shared_theme
-ggsave(filename = "figures/tip_trait_correlation.png", width = linewidth, height = height, unit = "cm")
+ggsave(filename = "figures/tip_trait_correlation.png", width = linewidth, height = height * 2/3, unit = "cm")
 

### 2022-06-14 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 \usepackage{tabularx} % for tables that don't spill past margin
+\usepackage[markup=underlined]{changes}
 
@@
 Traditional GWAS assume individual samples are independent observations.
-However, pathogen effects on an infectious disease trait can be heritable from donor to recipient in transmission chains.
+However, pathogen effects on a trait can be heritable from donor to recipient in transmission chains.
 Thus, residuals in GWAS association tests for host genetic effects may not be independent due to shared pathogen ancestry.
@@
 In simulations, we show this pre-processing step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
-When we apply our method to HIV-1 subtype B data from the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between set-point viral load (spVL) and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions compared to standard GWAS.
-Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 subtype B spVL, standard GWAS are robust to pathogen effects.
-Then, we show that phylogenetic correlations for quantitative disease resistance (QDR) in the \emph{A.
-thaliana}-\emph{X.
-arboricola} pathosystem are not strong enough to justify a phylogenetic correction for GWAS.
-Our framework may improve GWAS for other diseases if pathogen effects are even more phylogenetically correlated amongst individuals in a cohort.
+We applied our framework to data from two different host-pathogen systems.
+For HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions.
+Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects.
+For \emph{A.
+thaliana} quantitative disease resistance (QDR) to \emph{X.
+arboricola}, phylogenetic correlations are not strong enough to justify our phylogenetic correction for GWAS.
+In summary, our framework provides additional information on the evolutionary dynamics of traits in pathogen populations and may improve GWAS if pathogen effects are highly phylogenetically correlated amongst individuals in a cohort.
 \end{abstract}
@@
 This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
-Broad-sense heritability $H^2$ quantifies the fraction of total trait variance that is ``inherited'' from infection partner to infection partner.
+Broad-sense pathogen heritability $H^2$ quantifies the fraction of total variance in a trait that is ``inherited'' from infection partner to infection partner, i.e., due to pathogen factors.
 To characterize $H^2$ and the heritable and non-heritable factors that determine infectious disease traits, we use a phylogenetic mixed model (PMM) \citep{Housworth2004}.
@@
 By incorporating phylogenetic information, we can improve upon this error in scenarios where the trait is highly heritable, under low selection pressure, and with relatively moderate stochastic fluctuations compared to outbreak duration.
-Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength.
+Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations.
 These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny.
 This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait.
-This is because high heritability means the trait value is highly pathogen-dependent.
-Then, when the trait is under weak selection, the pathogen effects can drift far from the long-term optimum and stochastic fluctuations are low to maintain the same heritability.
-Thus, our method performs well when an infectious disease trait is highly heritable and trait values are strongly correlated amongst close transmission partners.
+Figure \ref{fig:tip-trait-correlation} shows in more detail how trait values are correlated at closely clustered tips when an infectious disease trait is highly heritable and selection strength is low/ stochastic fluctuations are low.
 
@@
 For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
-(A) shows that our method (left) can generate more accurate estimates of additive host genetic effects than the trait value, scaled by its mean (right).
+(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
+The RMSE is with reference to the true (simulated) host part of the trait values.
+Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
 (B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
@@
 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
-These examples correspond to the empty points in Figure \ref{fig:simulationResults}.
+These examples correspond to two of the empty points in Figure \ref{fig:simulationResults}.
 (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively.
@@
 \includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
-	\caption{Correlations between trait values in pairs of tips in four of the simulated scenarios from Figure \ref{fig:simulationResults}.
-Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each evolutionary scenario.
-Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength (upper right facet).}
+	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
+These examples correspond to the four empty points in Figure \ref{fig:simulationResults}.
+Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
+Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength/ low stochastic fluctuations (upper left facet).}
 	\label{fig:tip-trait-correlation}
@@
 The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals.
-We followed the framework outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
+We followed the method outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
 Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
@@
 However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}).
-Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}.
+Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date.
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
@@
 
-Next, we applied our framework to data collected from the \emph{A.
+Next, we applied our method to data collected from the \emph{A.
 thaliana}-\emph{X.
@@
 Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table \ref{tab:POUMM-parameters-our-estimates-qdr}).
-These selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.
+The posterior mean selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.
 
@@
 By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait.
-This should help us better understand which host factors are protective against a disease versus which increase susceptibility or disease severity.
+This should help us better understand which host factors are broadly protective against a disease versus which increase susceptibility or disease severity.
 
@@
 Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals.
-In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
-So, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
+In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
+It is possible that traits under strong selection pressure would also be highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
+However, such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect.
+In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
 
@@
 For \emph{A.
-thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+thaliana} QDR, inferred phylogenetic correlations are too weak to justify our phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
 \cite{Wang2018Two-wayGenomes} found no significant \emph{A.
@@
 thaliana}-\emph{X.
-arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using the corrected trait values in a GWAS.
+arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS.
 In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.
@@
 Finally, additional model complexity can be added to the GWAS association tests.
-For instance, our method does not account for co-infection, although one could add co-infection status as a covariate in the GWAS association test to account for this variable.
+For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power.
+In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.
 
@@
 \cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.
+Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model \citep{Rohlfs2014ModelingVariation}.
+Thus, one could also estimate pathogen effects on host gene expression.
+In particular, it will be of interest to understand the evolutionary dynamics of these traits in the respective pathogen populations.
 
@@
 The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table \ref{tab:sim-params}, Figure \ref{fig:sigma-vs-alpha}).
-Finally, the environmental component of spVL $\epsilon$ was generated from a normal distribution with mean 0.
+Finally, the environmental part of spVL $\epsilon$ was generated from a normal distribution with mean 0.
 For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.
+
+We performed GWAS on the simulated data using a linear association model as implemented in the ``lm'' function in R.
+For each simulated dataset, we performed three association tests: (i) using the true (simulated) non-pathogen part of the trait (host + environmental parts), (ii) using the estimated non-pathogen part of the trait according to the method presented in this paper, and (iii) using the total trait value, scaled by its mean.
+We assessed the significance of each associations at a significance level of 0.05 with a Bonferroni correction for multiple testing.
+For our main results (Figure \ref{fig:simulationResults}) we simulated 20 truly associated variants, as described above.
+To also check the false positive rate (FPR), we re-ran the simulations with an additional 80 non-associated variants.
+Across all the association tests in this second simulation setup (7 $H^2$ levels $\times$ 10 $\alpha$ levels $\times$ 100 variants $\times$ 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005  using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value.
+These rates are comparable to the expected FPR of 0.0005 at significance level 0.05 corrected for 100 tests.
+Given the stricter correction for multiple testing in this second simulation setup, the TPR decreased significantly across all three GWAS response variables used.
 
@@
 thaliana} and \emph{X.
-arboricola} genotyping and quantitative disease resistance (QDR) measurements are described in detail in \citep{Wang2018Two-wayGenomes}.
-Briefly, \citep{Wang2018Two-wayGenomes} infected different \emph{A.
+arboricola} genotyping and quantitative disease resistance (QDR) measurements are described in detail in \cite{Wang2018Two-wayGenomes}.
+Briefly, \cite{Wang2018Two-wayGenomes} infected different \emph{A.
 thaliana} host accessions with different \emph{X.
 arboricola} pathogen strains in a fully-crossed experimental design.
-They infected up to 4 leaves on each of three biological replicates for each host-pathogen pair.
+They infected up to 4 leaves on each of three biological replicates for each host-pathogen pairing.
 Then, they scored QDR for each leaf on a scale of 0 (resistant) to 4 (susceptible).
-A genotype matrix with allele dosage of 33,610 SNPs for the 22 \emph{X.
-arboricola} pathogen strains used was downloaded from the supplemental material of \citep{Wang2018Two-wayGenomes}.
-A VCF file with allele dosage of 12,883,854 SNPs for different \emph{A.
-thaliana} accessions was downloaded from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}.
-QDR measurements were provided directly by the authors of \citep{Wang2018Two-wayGenomes}.
+We downloaded the genotype matrix with allele dosage of 33,610 SNPs for the 22 \emph{X.
+arboricola} pathogen strains generated by \cite{Wang2018Two-wayGenomes} from their supplemental material.
+We additionally downloaded a VCF file with allele dosage of 12,883,854 SNPs for the different \emph{A.
+thaliana} accessions from the 1001 Genomes project \citep{Alonso-Blanco20161135Thaliana}.
+QDR measurements were provided directly by the  \cite{Wang2018Two-wayGenomes} authors.
 
@@
 arboricola} pathogen genotype matrix.
+The inferred tree topology (Figure \ref{fig:qdr-on-tree}) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes} generated using the unweighted pair group method with arithmetic mean (UPGMA).
+Compared to UPGMA, the neighbor-joining method we used relaxes the assumptions of a strict molecular clock and sampling all at the same time-point.
 For the trait value to fit the POUMM, we calculated mean QDR across all leaves infected on all hosts for each \emph{X.
@@
 \stopsection
+\newpage
+\section*{Editors’ comments}
+AE: I have now received two reviews of your paper.
+Both reviewers were excited about the methodological concept, but both where also concerned about 1) the perceived narrow application of the methodology and 2) the example based on data which are not accessible.
+The reviewers have suggested ways to perhaps broaden the potential applications of the method and have even suggested additional (alternative) data sets that ARE accessible.
+I strongly encourage you to utilize data that others can access to explore your methods.
+If you want to make insights on HIV evolution using restricted data, then please send such work to a medical journal.
+But for MBE, the expectation (especially for a methodological paper) is that the demonstrations of methods should be using data that are easily and openly accessible.
+Hopefully, you will find these reviews helpful in revising your work.
+
+\textcolor{blue}{We thank the editor for the chance to improve our manuscript and the two reviewers for their constructive feedback.
+We feel the quality of the manuscript improved thanks to their suggestions.
+Below, we highlight the main improvements in the revised manuscript.
+We also discuss two suggestions we did not implement.
+Finally, in the point-by-point responses, we highlight the specific changes to the manuscript based on each of the reviewers comments.}
+
+\textcolor{blue}{The main improvement in the revised manuscript is that we added an analysis of an open dataset and re-phrased our abstract and introduction to focus on the broader applicability of our method, including what differentiates it from previous methods.
+We highlight that our two-step method provides additional insights on the evolutionary dynamics of a trait in a pathogen population and that it is compatible with a variety of downstream GWAS association testing methods.
+We removed the HIV-specific information from the introduction and moved the relevant comparisons to the discussion section.}
+
+\textcolor{blue}{The only suggestions we did not implement were to re-run the HIV GWAS using a linear mixed model and to perform the HIV analysis again using elite controllers.
+In the first case, we rephrased the discussion to make it clearer that our method is envisioned as a pre-processing step that can be used prior to GWAS association testing under any model.
+Given the revised focus of the manuscript away from HIV-1 specific insights, we felt improving the HIV-1 GWAS association testing methodology was beyond the scope of the current manuscript.
+In the second case, our dataset does not contain enough elite controllers to perform the additional analysis.}
+
+\section*{Reviewer 1}
+
+The authors present a new GWAS method that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny.
+I do not have critiques about the framework, as the method the authors have developed seems novel and in principal, groundbreaking.
+However, I am finding difficult to grasp the usability of it outside of pathogens that integrates in host genome.
+What is the broad applicability of the method?
+Is this method only applicable to stVL in the context of HIV?
+What else could be investigated?
+Is it possible to pair transcriptomic data?
+Can author expand on this?
+
+\vspace{5mm}
+\textcolor{blue}{We edited the introduction to make it clearer when this method is applicable: ``This framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.''}
+
+\textcolor{blue}{We also expanded the discussion to include the possibility of transcriptomic applications and highlight that even when this method is not anticipated to increase GWAS power, it is still interesting to fit the POUMM and estimate pathogen effects on different traits: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g.
+\cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.
+Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model \citep{Rohlfs2014ModelingVariation}.
+Thus, one could also estimate pathogen effects on host gene expression.
+In particular, it will be of interest to understand the evolutionary dynamics of these traits in the respective pathogen populations.''}
+
+\begin{enumerate}
+    \item It would have been interesting to see how their methods performs with another type of pathogen, a virus that does not integrate, such SARS-CoV-2, or a bacterial one, such TB. \\
+    \textcolor{blue}{We added an analysis of \emph{A.
+thaliana} quantitative disease resistance to the phytopathogenic bacteria \emph{X.
+arboricola}.
+The findings are highlighted in the discussion: ``For \emph{A.
+thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+\cite{Wang2018Two-wayGenomes} found no significant \emph{A.
+thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects.
+As with HIV-1 spVL, our results do not challenge this previous finding.''}
+    \item In the case of co-infection, as sometimes is the case of HIV and TB, how would the method perform? \\
+    \textcolor{blue}{We addressed this point in the discussion: ``Finally, additional model complexity can be added to the GWAS association tests.
+For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power.
+In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.''}
+    \item In the case of HIV, is it possible to compare the data with elite controllers? \\
+    \textcolor{blue}{Unfortunately our dataset does not include a sufficient number of elite controllers to perform this comparison.
+The SHCS cohort includes 83 individuals classified as elite controllers, as defined by at least three measurements of plasma virus load (VL) $<$ 2000 RNA copies/ml over at least a 12-month period in the absence of antiviral therapy (Pereya et al, Science, 2010).
+spVL values range from 1.3 to 3.1 in this group (all very low).
+After filtering steps, only 8 of these individuals were retained for the GWAS, so the sample size is unfortunately not sufficient to do a comparative GWAS.
+However, we note that McLaren et al.
+2015 found similar signal of association on chromosome 3 and 6 when they filtered to elite controllers in a standard GWAS setup.}
+    \item In the case of determining whether viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, I am not sure I understand the premise of this hypothesis.
+Can authors elaborate on it more in detail?
+\\
+    \textcolor{blue}{We edited the introduction to make it clearer why one should consider pathogen effects in a host GWAS: ``Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait.
+Corrections for these types of population structure in human GWAS cohorts are well-developed and widely accepted  \citep{Astle2009, Price2006}.
+More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}...
+[our method] estimates and removes pathogen effects on the trait so that downstream GWAS association testing with host genetic variants can be performed as normal under the assumption of independent samples''}
+    \item What type of clinical information/trait is included under the environmental effect? How the diversity in environmental effects across the population affects the method? \\
+    \textcolor{blue}{We edited the text to give examples of factors that would be grouped into the environmental effect under the POUMM model we use: ``Variation in infectious disease traits like viral load or infection severity can come from several sources.
+These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature...
+In the infectious disease GWAS case, we assume the heritable part comprises pathogen genetic factors and all other factors are non-heritable.''}
+
+    \textcolor{blue}{In addition to the change described in response 1, we also added text to the new approaches section to make it clearer that our method is complementary to but does not replace other methods to account for diversity in environmental effects (we only account for heritable pathogen genetic effects): ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
+This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.''}
+    \item Lines 194-196:  please expand on the parameters of the simulation in greater detail. \\
+    \textcolor{blue}{We edited this paragraph to provide a full description of the simulation scheme.}  
+    \item Lines 205-207: it would be best, if there is not a limitation in the number of words, to report here “why” instead of saying that it has been shown in SI. \\
+    \textcolor{blue}{We added the suggested explanation: ``As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to viral genetic effects and environmental effects is fixed.
+Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.''}
+    \item Liens 212: “our method performs well when an infectious disease trait is highly heritable”, I found the concept of heritability in the infectious disease filed confusing.
+Are the authors talking about traits in the viral population?
+Is it possible to expand the sentence with more details for clarity?
+\\
+    \textcolor{blue}{We edited this statement to be clear that we mean ``pathogen heritability''.
+We also tried to explain the concept of heritability due to the pathogen better in the introduction: ``Variation in infectious disease traits like viral load or infection severity can come from several sources.
+These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature.
+GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
+This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
+Broad-sense pathogen heritability $H^2$ quantifies the fraction of total variance in a trait that is ``inherited'' from infection partner to infection partner, i.e., due to pathogen factors.''}
+\end{enumerate}
+
+\section*{Reviewer 2}
+
+In this paper, Nadeau and colleagues look to correct for pathogen heritability in studies of host association with infectious diseases.
+They do this by extending an existing method which models trait evolution as a walk along a phylogeny, and provide a new method which which decomposes a pathogen trait into pathogen heritability vs 'other effects' using maximum-likelihood.
+By using only the component corresponding to 'other effects' as the phenotype in a standard host GWAS, they are able to correct for dependence between pathogen samples due to shared ancestry.
+As the authors note, this method is rather convenient as it results in an adjusted phenotype which can be used with standard GWAS tools.
+They apply their method to simulated data, showing a range (high heritability and selection) where a power advantage should be gained.
+They also apply the method to HIV set point viral load (spVL).
+
+The paper is based on a sound idea, and is a technically competent solution, with no major scientific issues I could identify.
+Code is provided; data available via managed access; derivations in methods were appropriately explained.
+The introduction and conclusion are both well-written and do not overstate the results.
+
+The main comment I have is based perhaps more on the impact of the study.
+The study is very strongly focused on analysis of HIV-1 spVL, which is not really the ideal trait for this method (as seen from the simulations in fig 2; and fig S4 showing the trait changing not all that much).
+It has long been established that there are two clear host associations, which indeed are exactly what the authors find before and after their adjustment.
+[If the aim was to try and find more associations for this trait, it would likely be much better to collect lots more samples than change the association methods, as has been found in host GWAS studies of other traits].
+To increase the relevance of this study, the authors may consider whether they can apply their approach to other pathogens/traits.
+At the same time, I also appreciate that host/pathogen data is hard to come by and may have access issues that are too difficult to overcome.
+However, I do think if any of these studies in other pathogens could also be analysed here, it would really add to the paper.
+The authors cite a study of HCV as well as some other viral studies as examples, and I would also point to two more in bacteria (https://www.pnas.org/content/115/24/E5440; https://www.nature.com/articles/s41467-019-09976-3), and one in malaria (https://www.biorxiv.org/content/10.1101/2021.03.30.437659v1).
+Perhaps similar data is now available for SARS-CoV-2 also?
+
+\textcolor{blue}{We were able to access additional data from the \emph{A.
+thaliana}-\emph{X.
+arboricola} pathosystem to demonstrate the broader applicability of our method.
+Unfortunately neither dataset we analyzed had sufficiently highly correlated trait values amongst related pathogen strains to generate novel biological insights.
+We edited the text to emphasize that an advantage of our approach is that one gets information on the evolutionary dynamics of the trait in  the pathogen population, even if GWAS results are not changed.}
+
+\vspace{5mm}
+I also have some more minor comments:
+
+\begin{enumerate}
+    \item How does the method compare to the ATOMM approach in https://www.pnas.org/content/115/24/E5440? Can this be applied to any of the simulated or real data here for comparison. \\
+    \textcolor{blue}{We cannot apply ATOMM directly to our simulated data because we do not simulate pathogen genotypes, rather we simulate only pathogen trait values.
+Further, ATOMM is only implemented for haploid alleles, making it not ideal for the HIV data analyzed.
+Instead, we were able to access the same dataset ATOMM was tested on in their original publication.
+This allows us to compare GWAS results from our method versus the ATOMM method, which accounts for the pathogen and host genome jointly in the same linear mixed model and allows tests for marginal effects of host and pathogen alleles as well as pairwise host allele $\times$ pathogen allele interaction effects.}
+
+    \textcolor{blue}{We present the findings of this comparison in the discussion: ``For \emph{A.
+thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+\cite{Wang2018Two-wayGenomes} found no significant \emph{A.
+thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects.
+As with HIV-1 spVL, our results do not challenge this previous finding.''}
+
+    \textcolor{blue}{We also highlight the linear mixed model approach presented in ATOMM as an alternate means to take pathogen effects into accout: ``As in the \emph{A.
+thaliana}-\emph{X.
+arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using the corrected trait values in a GWAS.
+In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.''}
+    \item As I understand it the POUMM method assumes a 'correct' phylogeny.
+Do the authors have any thoughts about including uncertainty in the phylogeny (for example by using a posterior sample from e.g.
+mrbayes) and whether this may affect their approach?
+Relatedly, does a tree from just pol sequences (L342) give enough resolution?
+\\
+    \textcolor{blue}{TODO: improve phylogeny by masking sites under selection.
+We randomly chose an alternate tree from the collection of optimal trees produced with the IQ-TREE ``wt'' option and re-fit the POUMM.
+Posterior mean parameter estimates were comparable.
+TODO: re-correct trait values and re-run the GWAS using a few different trees.}
+
+    \item It would be useful to include more histograms and scatterplots of traits before/after the adjustment, as in figure S4.
+In particular, doing this plot again for one of the simulations where it does improve power would be informative.
+In my opinion, these could be a main figure.
+\\
+    \textcolor{blue}{We implemented this suggestion and added two new main text figures.
+They are discussed in the section ``Estimator accuracy'': ``Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations.
+These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny.
+This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait.
+Figure \ref{fig:tip-trait-correlation} shows in more detail how trait values are correlated at closely clustered tips when an infectious disease trait is highly heritable and selection strength is low/ stochastic fluctuations are low.''}
+    \item Do the authors know how the computational requirements of their ML estimation method scales with sample size? Would it work for a large SARS-CoV-2 phylogeny for instance? \\
+    \textcolor{blue}{The POUMM R package for estimating the POUMM model parameters runs in less than an hour on a tree of 10,000 tips and has been shown to scale to trees of up to 100,000 tips.
+We additionally verified that the matrix inversion steps required for the last ML estimation step based on the fitted POUMM parameters ran successfully on a simulated tree of 10,000 tips using 4 cores with maximum memory use of 6500MB per core.
+We now highlight the method's scalability in the discussion: ``The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}.''}
+    \item Fig S3 would be useful, but isn't really readable at present. Perhaps consider an interactive tree viewer to make this more useful. \\
+    \textcolor{blue}{We added the same figure for the \emph{A.
+thaliana}-\emph{X.
+arboricola} dataset, which is much more legible due to having fewer samples (Figure \ref{fig:qdr-on-tree}).
+We also added sub-sampled phylogenies from the simulation study annotated with trait values in Figure \ref{fig:simulationTraitExample} to demonstrate what phylogenetically-correlated trait values look like.
+Finally, we now reference the highest correlation values expected between any two tips in the HIV-1 and \emph{X.
+arboricola} phylogenies to add a quantative element to the discussion of phylogenetic correlations: ``The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
+However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability...
+Given the posterior mean estimates for the POUMM parameters [using \emph{X.
+arboricola} data], expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application).
+Thus, the phylogeny is not very informative for a trait value correction.''}
+    \item The authors I think only look at power/TPR. What about the effect on false positive rate? \\
+    \textcolor{blue}{We re-ran the simulations to assess the false positive rate and include the results in the Materials and methods section: ``For our main results (Figure \ref{fig:simulationResults}) we simulated 20 truly associated variants, as described above.
+To also check the false positive rate (FPR), we re-ran the simulations with an additional 80 non-associated variants.
+Across all the association tests in this second simulation setup (7 $H^2$ levels $\times$ 10 $\alpha$ levels $\times$ 100 variants $\times$ 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005  using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value.
+These rates are comparable to the expected FPR of 0.0005 at significance level 0.05 corrected for 100 tests.
+Given the stricter correction for multiple testing in this second simulation setup, the TPR decreased significantly across all three GWAS response variables used.''}
+    \item I don't think panel A of figure 2 is explained well by the caption or text, I could not understand what the RMSE is showing here. What does 'more accurate' look like in this plot? \\
+    \textcolor{blue}{We edited the figure legend to make it clearer: ``(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
+The RMSE is with reference to the true (simulated) host part of the trait values.
+Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.''}
+    \item Is there any explanation why the simulated (actual) and estimated non-pathogen trait in fig 2B start to look quite different at stronger selection strengths. \\
+    \textcolor{blue}{We added a figure (Figure \ref{fig:tip-trait-correlation}) to illustrate how the correlation between clustered tips on the phylogenetic tree decreases with increasing selection strength/ increasing stochasticity.
+Without phylogenetic signal, our method has no information to estimate the pathogen effect on the trait.}
+    \item Using plink and removing samples with divergent ancestry is probably a little outdated for host GWAS.
+A linear mixed model would likely provide more power still.
+In particular, I'd suggest warped-lmm (http://dx.doi.org/10.1038/ncomms5890), which also transforms the phenotype to increase power.
+Furthermore, I would be interested to see histograms/scatter plots of a few phenotypes: a) no adjustment; b) warped-lmm only; c) the author's method; d) both adjustments.
+Perhaps the methods are similar, or using both does increases power beyond either method alone.
+\\
+    \textcolor{blue}{We highlight that our method is a pre-processing step before GWAS and can be combined with many different GWAS models implemented in softwares like PLINK, GEMMA, and warped-lmm.
+Our revised manuscript focuses more on the method rather than generating biological insights for HIV or \emph{X.
+arboricola}, so we feel that updating our example GWAS is beyond the scope of the current study.
+We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS...
+In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
+\end{enumerate}
+
+\vspace{5mm}
+Suggested text changes:
+
+\begin{enumerate}
+    \item The first sentence of the abstract is probably also true for many common diseases analysed in human GWAS (apart from perhaps the pathogen, but gut microbiome may be a factor too). \\
+    \textcolor{blue}{We changed the wording to make it clearer what sets infectious disease GWAS apart: ``Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait.''}
+    \item L84-86.
+There are more than two methods for this, see also https://doi.org/10.1371/journal.pgen.1007758; https://doi.org/10.1093/bioinformatics/bty539; https://doi.org/10.1099/mgen.0.000469.
+Maybe remove 'two', as this will change anyway.\\
+    \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}.
+Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \citep{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \citep{collins_phylogenetic_nodate}.''}
+    \item L110 'Of these, only pathogen effects are heritable from one transmission partner to another'.
+This might be nit-picking but I could see this phrasing being misunderstood by some readers.
+Environment will be partially shared along a transmission chain, probably.
+Interaction effects could also be shared if e.g.
+hosts are likely to have the same ancestry.\\
+    \textcolor{blue}{We improved the phrasing to be more accurate: ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
+This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.''}
+    \item Equation 1 onwards.
+Using 'e' for the environmental effects is easy to get confused with more standard use of the number e = 2.718, especially when raised to a power.
+Consider changing this symbol.\\
+    \textcolor{blue}{We changed the variable to $\epsilon$ and also changed the variable for the pathogen part of the trait to $g$ to match previous work on the POUMM (e.g. \citep{Mitov2017a-POUMM})}.
+    \item Please add the link to the github code in the framework section at the end of the methods.\\
+    \textcolor{blue}{We added the link in the section ``Phylogenetic spVL correction'': ``All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.''}
+    \item L283 'In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.' Could the authors give a biological example of this?
+Would antimicrobial resistance variation fit the bill?\\
+    \textcolor{blue}{We anticipate antimicrobial resistance would be under strong selective pressure, but agree this could produce phylogenetically correlated trait values.
+We added a statement in the discussion: ``It is possible that traits under strong selection pressure would also be phylogenetically structured, for instance antimicrobial resistance, though such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect.''}
+\end{enumerate}
 \end{document}

### 2022-06-14 scripts/R/make_manuscript_figures.rmd
add code to summarize FPR

@@
 accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
-tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
+tpr_simulation_results_old <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
+tpr_simulation_results <- read.delim("output_revisions/2022-06-13_GWAS_TPR_MLE.txt")
 example_simulation_results <- read.delim("output_revisions/2022-06-11_simulateFullDataset.txt")
@@
 gwas_results <- read.delim(file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
-gwas_results_xanthomonas <- read.delim(file = paste(OUTDIR_XANTHAMONAS, "gwas_results.pvals.effectsize.txt", sep = "/"), sep = "", header = T)
 spvl_values <- read.csv(file = paste(OUTDIR, "spvl_values.csv", sep = "/"))
@@
         significance.correction = "Bonferroni",
-        N.tests = 20)
+        N.tests = 100)
 data$TP <- ifelse(data$p.value < sig.level & data$is.causal.variant, T, F)
@@
         scale_shape_manual(values = c("QDR" = 16, "spVL" = 4, "Example" = 1), name = element_blank())
+
+fp_summary <- data %>%
+        group_by(inference.method) %>%
+        summarise(FPR = sum(FP, na.rm = T) / sum(!is.na(FP)), n_non_na = sum(!is.na(FP)))
 
@@
 ```
+Look for false positives in the simulated data
+```{r}
+sig.level <- correctSignificanceLevelForMultTesting(
+        significance.level = 0.05,
+        significance.correction = "Bonferroni",
+        N.tests = 100)
+
+example_simulation_results <- example_simulation_results %>%
+        mutate(
+                FP <- ifelse(p.value < sig.level & !(is.causal.variant), T, F)
+        )
+```
+Compare other ML phylogenies for HIV data
+```{r}
+trees <- ape::read.tree("output_revisions/pathogen.treels.head500")
+
+# Calculate tree distances
+tree_dists <- dist.topo(trees, method = "PH85")
+```

### 2022-06-14 scripts/R/functions/simulation_control_functions.R
remove some hardcoded simulation params (this was done post-manuscript analyses)

@@ -255,7 +255,7 @@ simulateGWASPValuesWithPLINK <- function(tree, N, g0, alpha, theta, M, K, d, p =
   return(p.values)
 }
 
-simulateGWASPValues <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delta = NA, H2, H2.h, var.z, is.MLE = T) {
+simulateGWASPValues <- function(tree, N, g0, alpha, theta, M, K, d = 2, p = NA, delta = NA, H2, H2.h =0.25, var.z, is.MLE = T) {
   print(paste0("Simulating for GWAS + POUMM TPR with these parameters: alpha=", alpha, ", H2=", H2, ", H2.h=", H2.h, ", var.z=", var.z))
   # Simulate data
   epidemic <- simulateEpidemic(
@@ -266,11 +266,11 @@ simulateGWASPValues <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delt
     theta = theta,
     M = M,
     K = K, 
-    d = 2, 
+    d = d, 
     p = p,
     delta = delta,
     H2 = H2,
-    H2.h = 0.25,
+    H2.h = H2.h,
     var.z = var.z,
     is.MLE = T)
 
@@ -286,7 +286,7 @@ simulateGWASPValues <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delt
   return(p.values)
 }
 
-simulatePGLSPValues <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delta = NA, H2, H2.h, var.z, is.MLE = T, is.estimatePOUMMparams = T) {
+simulatePGLSPValues <- function(tree, N, g0, alpha, theta, M, K, d = 2, p = NA, delta = NA, H2, H2.h = 0.25, var.z, is.MLE = T, is.estimatePOUMMparams = T) {
   print(paste0("Simulating for GWAS + PGLS TPR with these parameters: alpha=", alpha, ", H2=", H2, ", H2.h=", H2.h, ", var.z=", var.z))
   # Simulate data
   epidemic <- simulateEpidemic(
@@ -297,11 +297,11 @@ simulatePGLSPValues <- function(tree, N, g0, alpha, theta, M, K, d, p = NA, delt
     theta = theta,
     M = M,
     K = K, 
-    d = 2, 
+    d = d, 
     p = p,
     delta = delta,
     H2 = H2,
-    H2.h = 0.25,
+    H2.h = H2.h,
     var.z = var.z,
     is.MLE = T)
   

### 2022-06-17 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 We propose a new method to estimate and remove heritable pathogen effects on a trait based on the pathogen phylogeny prior to host GWAS, thus restoring independence of samples.
-In simulations, we show this pre-processing step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
-We applied our framework to data from two different host-pathogen systems.
-For HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, we recover slightly weaker but qualitatively similar signals of association between spVL and human genetic variants in the \emph{CCR5} and major histocompatibility complex (MHC) gene regions.
-Our simulation study confirms that based on the estimated heritability and selection parameters for HIV-1 spVL, standard GWAS are robust to pathogen effects.
-For \emph{A.
-thaliana} quantitative disease resistance (QDR) to \emph{X.
-arboricola}, phylogenetic correlations are not strong enough to justify our phylogenetic correction for GWAS.
+In simulations, we show this additional step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
+We applied our framework to data from two different host-pathogen systems, namely HIV in humans and \emph{X.
+arboricola} in \emph{A.
+thaliana}.
+In both systems the heritability and thus phylogenetic correlations turn out to be low enough such that qualitative results of GWAS do not change when accounting for the pathogen shared ancestry through a correction step.
+This means that previous GWAS results applied to these two systems should not be biased due to shared pathogen ancestry.
 In summary, our framework provides additional information on the evolutionary dynamics of traits in pathogen populations and may improve GWAS if pathogen effects are highly phylogenetically correlated amongst individuals in a cohort.
@@
 
-In this work, we draw from the field of phylogenetic comparative methods to propose a new two-step framework to correct for pathogen population structure and satisfy the GWAS assumption of independent samples.
-It is a pre-processing step for GWAS that first fits an evolutionary model to trait data and the pathogen phylogeny to estimate the correlation structure of the trait due to pathogen effects.
-On the basis of this inference, it estimates and removes pathogen effects on the trait so that downstream GWAS association testing with host genetic variants can be performed as normal under the assumption of independent samples.
-The main advantage of this approach compared to the previously outlined methods to correct for pathogen population structure are that it generates additional information on the evolutionary dynamics of the trait in the pathogen population.
-It is also restricted to a pre-processing step, which allows use of existing highly optimized tools to perform GWAS association tests under a variety of models.
-This framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.
+In this work, we draw from the field of phylogenetic comparative methods to propose a new two-step framework  which corrects for pathogen population structure and thus satisfies the GWAS assumption of independent samples.
+The introduced framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.
+We hypothesized that our approach should improve GWAS power to identify host genetic variants broadly associated with disease traits.
+
+In a first step, we fit an evolutionary model to trait data and the pathogen phylogeny.
+This first step provides an estimate of the correlation structure of the trait due to pathogen effects.
+The estimate is used to remove pathogen effects on the trait.
+In the second step, the resulting corrected trait data is used in a GWAS with host genetic variants.
+The GWAS can be performed as normal under the assumption of independent samples.
+The main advantage of this two-step approach compared to the previously outlined methods to correct for pathogen population structure is that it generates additional information on the evolutionary dynamics of the trait in the pathogen population.
+The advances presented here are on the first step, while in the second step existing highly optimized tools to perform GWAS association tests under a variety of models can be employed.
+
+.
 % We hypothesis this should increase GWAS power for several reasons.
 For one, pathogen effects may be confounding or extraneous variables that produce distorted or spurious associations between host genetic variants and a disease trait in host GWAS.
-We hypothesized that our approach should improve GWAS power to identify host genetic variants broadly associated with disease traits.
+
 %This may occur if the virus affects the disease trait (the dependent variable in host GWAS) or if it affects both the disease trait and the host genotype (the independent variable in host GWAS).
@@
 In the following, we describe the evolutionary model for heritable, continuous-valued infectious disease traits upon which our method is based.
-We then derive a maximum likelihood estimate for the pathogen part of a trait under this model, and describe a new infectious disease GWAS framework using this information.
+We derive a maximum likelihood estimate for the pathogen part of a trait under this model.
+We then describe a new infectious disease GWAS framework assessing associations of the trait with host genetic variants using the maximum likelihood estimates.
 In simulations, we show that this framework can improve GWAS power to detect host genetic variants that affect disease traits.
@@
 This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
+We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes.
+
 Broad-sense pathogen heritability $H^2$ quantifies the fraction of total variance in a trait that is ``inherited'' from infection partner to infection partner, i.e., due to pathogen factors.
@@
 \end{equation}
-Next, we remember that $\epsilon$ is the non-heritable part of the trait.
+Next, we recall that $\epsilon$ is the non-heritable part of the trait.
 $\epsilon$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
@@
 For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
-The intensity of stochastic fluctuations parameter $\sigma$ was determined based $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}).
-As shown in Figure \ref{fig:sigma-vs-alpha}, higher $\alpha$ correspond to higher $\sigma$ to maintain constant $H^2$ under this parameterization.
-For each $H^2$ and $\alpha$ considered, we simulated recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
+The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}).
+As shown in Figure \ref{fig:sigma-vs-alpha}, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization.
+For each $H^2$ and $\alpha$ value considered in the simulation, we recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
 
 We paired each tip's simulated pathogen trait value with a simulated host trait value.
-Simulated hosts had 100 genome positions.
+Simulated hosts had 20 genome positions.
 We sampled alleles (0, 1, or 2) for each position from a binomial distribution with probability 0.13.
 10 random positions had an effect size of 0.2 on the trait and 10 had an effect size of -0.2.
-The other 80 positions had no effect on the trait.
 This parameterization produced roughly normally distributed host trait values centered at 0 with variance equal to 25\% of the total trait variance, which we constrained to 0.73 based on the variance in log spVL values measured by \cite{Mitov2018}.
@@
 We estimate approximately 25\% to be the heritability threshold above which GWAS power is negatively impacted by pathogen effects.
-In summary, we show it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.
+In summary, we show that it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.
 
@@
 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
-These examples correspond to two of the empty points in Figure \ref{fig:simulationResults}.
+These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
 (A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively.
@@
 	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
-These examples correspond to the four empty points in Figure \ref{fig:simulationResults}.
+These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
 Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
@@
 In this paper, we presented a new phylogeny-aware GWAS framework to correct for heritable pathogen effects on infectious disease traits.
-By using information from the pathogen phylogeny, we show it is possible to improve GWAS power to detect host genetic variants associated with a disease trait.
-This should help us better understand which host factors are broadly protective against a disease versus which increase susceptibility or disease severity.
+By using information from the pathogen phylogeny, we show that it is possible to improve GWAS power to detect host genetic variants associated with a disease trait.
+This improved power is envisioned to obtain a better understanding of which host factors are broadly protective against a disease versus which increase susceptibility or disease severity.
 
 % Contextualize the study & address limitations
-Our method relies on the POUMM, a model of continuous trait evolution that accounts for heritable and non-heritable effects on a trait, as well as selection.
+The main novelty of our approach is to use these POUMM parameter estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects.
+In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
+Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.
+Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals.
+In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
+It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
+However, such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect.
+In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
+
 We applied this model to two different host-pathogen systems where paired host and pathogen genetic data was generated alongside a measure of pathogen virulence.
@@
 Our aim was to reconstruct a high-quality phylogeny, since the POUMM does not account for phylogenetic uncertainty and the POUMM parameter estimates are key to our downstream trait-correction method.
-Although our heritability estimate is rather high, the confidence interval largely overlaps that of other studies and we note that estimating heritability per se was not our primary focus.
+Although our heritability estimate is rather high, the confidence interval largely overlaps with the intervals  of other studies and we note that estimating heritability per se was not our primary focus.
 
@@
 Thus, these data partially violate the assumptions of the POUMM.
-Compared to HIV-1, we estimate very low selection strength for virulence in \emph{X.
+We estimate very low selection strength for virulence in \emph{X.
 arboricola}.
@@
 
-The main novelty of our approach is to use these POUMM parameter estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects.
-In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
-Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.
-Based on our simulation results, our method is anticipated to be very useful for disease traits that are highly heritable from donor to recipient and  maintain a high correlation between sampled individuals.
-In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
-It is possible that traits under strong selection pressure would also be highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
-However, such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect.
-In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
-
 Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A.
 thaliana} quantitative disease resistance to \emph{X.
-arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power (Figure \ref{fig:simulationResults}).
-Certain pairs of samples in the HIV-1 cohort were indeed expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45).
-However, the overall effect on GWAS is small.
+arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure \ref{fig:simulationResults}).
+Indeed, while certain pairs of samples in the HIV-1 cohort were expected to have phylogenetically correlated spvL values (maximum expected correlation between any two samples was 0.45), the overall effect on GWAS is small.
 For HIV-1 spVL, our phylogenetic correction slightly decreases p-values for variants in $CCR5$ and slightly decreases some and increases other p-values for variants in the MHC (Figure \ref{fig:gwas-results}B).
 Simulations show we shouldn't expect a net p-value decrease, but our simulations represent an ideal scenario since we simulate under the POUMM.
-In real life, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values.
+For the empirical data, un-modeled evolutionary pressures like drug treatment and host-specific HLA alleles might cause the reduced p-values.
 However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A).
@@
 First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population.
-Then, it is a convenient pre-processing step for GWAS because it simply produces an alternate response variable for GWAS association tests.
+Then, it is a convenient pre-processing step for GWAS because it simply produces a corrected response variable for GWAS association tests.
 As in the \emph{A.
@@
 In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.
-In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
-Finally, additional model complexity can be added to the GWAS association tests.
+In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
+Further, additional model complexity can be added to the GWAS association tests.
 For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power.
@@
 Thus, one could also estimate pathogen effects on host gene expression.
-In particular, it will be of interest to understand the evolutionary dynamics of these traits in the respective pathogen populations.
 
 % Summarize the paper
-In summary, we argue that infectious disease GWAS should take the pathogen phylogeny into account when searching for host determinants of a disease trait.
-We give a practical threshold for identifying when GWAS suffer from pathogen effects (heritability of the trait amongst infection partners $>$ 25\%) and provide a method that can help in scenarios where trait values are highly heritable and phylogenetically-structured amongst members of a cohort.
+In summary, we present a coherent infectious disease GWAS framework which takes the pathogen phylogeny into account when searching for host determinants of a disease trait.
+We further show that the pathogen phylogeny only has an impact on the GWAS outputs if heritability of the trait amongst infection partners $>$ 25\%.
+This means that for the systems studied here, spVL in HIV and QDR for X.
+arboricola infections in A.
+thaliana, the phylogeny correction does not change GWAS results and thus previously published results are not biased by the pathogen phylogeny.
 

### 2022-06-20 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 
-We applied our framework to empirical data from two different host-pathogen systems.
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
+	\caption{A high-level schematic of the experimental setup for the two application datasets.
+For application (A) to HIV-1 spVL in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
+We fit the POUMM to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
+For application (B) to \emph{A.
+thaliana}-\emph{X.
+Arboricola} QDR from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
+We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
+	\label{fig:application-exp-setup}
+	\end{center}
+\end{figure}
+
+We applied our framework to empirical data from two different host-pathogen systems with different experimental setups (Figure \ref{fig:application-exp-setup}).
 First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018.
@@
 		\includegraphics[width=\linewidth]{figures/gwas_results}
-		\caption{Results from comparative GWAS.
+		\caption{Results from comparative GWAS on HIV-1 spVL data.
 (A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables.
@@
 arboricola} pairing.
-Figure \ref{fig:qdr-on-tree} shows the inferred \emph{X.
-arboricola} pathogen phylogeny.
-Mean QDR calculated across all \emph{A.
-thaliana} pairings and all replicates for each \emph{X.
-arboricola} strain was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21.
+Our method requires a single trait value per pathogen strain, so we used mean QDR calculated for each pathogen strain across all the host \emph{A.
+thaliana} types (Figure \ref{fig:application-exp-setup}B).
+Figure \ref{fig:qdr-on-tree}A shows the inferred \emph{X.
+arboricola} pathogen phylogeny annotated with the  mean QDR trait value used for each strain.
+Mean QDR  was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21.
 Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table \ref{tab:POUMM-parameters-our-estimates-qdr}).
@@
 Thus, the phylogeny is not very informative for a trait value correction.
-Indeed, mean QDR values for each \emph{X.
-arboricola} strain calculated across all \emph{A.
-thaliana} pairings and all replicates is simply scaled by our method (Figure \ref{fig:trait-comparison-qdr}).
+Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure \ref{fig:trait-comparison-qdr}).
+We anyways selected 22 random host-pathogen strain pairings to perform a comparative GWAS analogous to that for HIV-1 spVL, where each host is infected with a single pathogen strain.
+In the first GWAS, we used the specific QDR measurement for each selected host-pathogen pairing.
+I.e., with reference to Figure \ref{fig:application-exp-setup}, we selected $z_{11}$ for the first sample, $z_{23}$ for the second sample, and so on.
+In the second GWAS, we used our (very uncertain) estimates for the non-pathogen effects on QDR for each pairing.
+Since our method did not utilize phylogenetic information in this case, the estimated non-pathogen part of the trait is simply the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
+thaliana} types.
+I.e., with reference to Figure \ref{fig:application-exp-setup}, we used a scaled version of $z_{11} - \bar{z}_1$ for the first sample, $z_{23} - \bar{z}_2$ for the first sample, and so on.
+Figure \ref{fig:gwas-results-qdr} shows that XXX.
 
@@
 
+\subsection{Phylogenetic uncertainty}
+Our method assumes the phylogeny accurately reflects the evolutionary relationships between pathogen strains.
+Previously, \cite{Hodcroft2014} observed HIV spVL heritability estimates based on $pol$ gene sequences were robust to including or not including resistance-associated codons.
+Our analysis includes these codons.
+For the HIV application, we additionally tested the sensitivity of the inference to phylogenetic uncertainty.
+We inferred the phylogeny again, this time using the IQ-TREE option ``wt'' to output all locally optimal trees.
+We fit the POUMM to two randomly selected trees from this set.
+Posterior mean parameter estimates were within the 95\% HPD intervals inferred from the maximum likelihood tree.
+GWAS results were also comparable.
+
 \subsection*{Data availability}
@@
 
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas.png}
+	\caption{Results from comparative GWAS on \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR data.
+The two facets show association p-values for the same host \emph{A.
+thaliana} variants in GWAS with two different response variables.
+On the left, we used unmodified (total) QDR values.
+On the right, we used our estimates for the non-pathogen effects on QDR.
+In this case, these values are total QDR minus a pathogen strain-specific scaled mean QDR.
+The alternating shades correspond to different chromosomes.
+The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
+	\label{fig:gwas-results-qdr}
+	\end{center}
+\end{figure}
+
 \end{doublespace}  % don't doublespace tables
@@
 \begin{table}[H]
-\caption{Number of samples for GWAS after sequential filtering steps.}
+\caption{Number of samples for HIV-1 spVL GWAS after sequential filtering steps.}
     \begin{tabularx}{\linewidth}{XX}
@@
 \begin{table}[H]
-\caption{Number of variants for GWAS after sequential filtering steps.}
+\caption{Number of variants for HIV-1 spVL GWAS after sequential filtering steps.}
 	\begin{tabularx}{\linewidth}{XX}
@@
 	\label{tab:variant-filtering}
+\end{table}
+
+\begin{table}[H]
+\caption{Number of variants for \emph{A. thaliana} QDR GWAS after sequential filtering steps.}
+	\begin{tabularx}{\linewidth}{XX}
+	\hline
+	Variant filter & Number of variants remaining \\
+	\hline 
+	Raw data & 11769920 \\
+	Minor allele frequency $<$ 0.1 & 1743952 \\
+	NA p-value due to too-high correlation with a covariate & 1070541 \hline			
+	\end{tabularx}
+	\label{tab:variant-filtering-qdr}
 \end{table}
@@
 \\
-    \textcolor{blue}{TODO: improve phylogeny by masking sites under selection.
-We randomly chose an alternate tree from the collection of optimal trees produced with the IQ-TREE ``wt'' option and re-fit the POUMM.
-Posterior mean parameter estimates were comparable.
-TODO: re-correct trait values and re-run the GWAS using a few different trees.}
+    \textcolor{blue}{This is correct, our method assumes the phylogeny accurately reflects the evolutionary history of the pathogen strains.
+We tested the sensitivity of the method to phylogenetic uncertainty in the HIV application.
+This is described in the section ``'': ``''.}
 

### 2022-06-20 scripts/R/plot_manhattan_euler.R
account for NA p-values from too-high correlations with covariates in qq plot

@@ -22,7 +22,7 @@ height <- 11  # cm
 # Load data
 print("Loading data")
 gwas_results_xanthomonas <- read.delim(
-    file = paste(OUTDIR, "gwas_results.pvals.effectsize.txt", sep = "/"), 
+    file = paste("gwas_results.maxmaf.pvals.txt", sep = "/"), 
     sep = "", 
     header = T
 )
@@ -47,8 +47,8 @@ max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 5,
 x_axis_ticks <- rbind(
         x_axis_ticks,
         data.frame(
-                CHROM = c(5, 6),
-                CHR_start = c(max_BP_position, max_BP_position + 1000)))
+                CHROM = c(6),
+                CHR_start = c(max_BP_position)))
 x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
 
 # Format data for plotting
@@ -69,7 +69,7 @@ results_long$gwas_type <- factor(
 print("Making plot")
 bonferroni_signifiance_threshold <- 0.05 / nrow(gwas_results_annotated)
 overview_comparison <- ggplot(
-        data = results_long %>% sample_frac(0.1),  # TODO: plot all of results, not sub-sample of them
+        data = results_long,
         aes(x = BP_position, y = -log10(P))) +
         geom_point(aes(color = CHROM %% 2 == 0), size = pt_size) +
         scale_color_manual(values = c("grey", "dark grey")) +
@@ -93,6 +93,30 @@ overview_comparison <- ggplot(
 print("Saving plot")
 ggsave(
         plot = overview_comparison,
-        filename = paste(OUTDIR, "gwas_results_xanthomonas.png", sep = "/"),
+        filename = paste("gwas_results_xanthomonas.png", sep = "/"),
         width = linewidth, height = height / 2, unit = "cm"
 )
+
+# Make qq-plot
+# Fiter to variants with p-value calculated
+gwas_results_xanthomonas_filtered <- gwas_results_xanthomonas %>% filter(!is.na(P_standard))
+
+qq_data_standard <- data.frame(
+        observed_log10p = -log10(sort(gwas_results_xanthomonas_filtered$P_standard, decreasing = F)),
+        expected_log10p = -log10(ppoints(nrow(gwas_results_xanthomonas_filtered))),
+        gwas_type = standard_gwas_name)
[…]

### 2022-06-22 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 In simulations, we show this additional step can increase GWAS power to detect truly associated host variants when pathogen effects are highly heritable, with strong phylogenetic correlations.
-We applied our framework to data from two different host-pathogen systems, namely HIV in humans and \emph{X.
+We applied our framework to data from two different host-pathogen systems, HIV in humans and \emph{X.
 arboricola} in \emph{A.
 thaliana}.
-In both systems the heritability and thus phylogenetic correlations turn out to be low enough such that qualitative results of GWAS do not change when accounting for the pathogen shared ancestry through a correction step.
+In both systems, the heritability and thus phylogenetic correlations turn out to be low enough such that qualitative results of GWAS do not change when accounting for the pathogen shared ancestry through a correction step.
 This means that previous GWAS results applied to these two systems should not be biased due to shared pathogen ancestry.
@@
 
-In this work, we draw from the field of phylogenetic comparative methods to propose a new two-step framework  which corrects for pathogen population structure and thus satisfies the GWAS assumption of independent samples.
+In this work, we draw from the field of phylogenetic comparative methods to propose a new two-step framework that corrects for pathogen population structure and thus satisfies the GWAS assumption of independent samples.
 The introduced framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.
@@
 In a first step, we fit an evolutionary model to trait data and the pathogen phylogeny.
-This first step provides an estimate of the correlation structure of the trait due to pathogen effects.
+This first step provides an estimate of the correlation structure of the trait due to heritable pathogen effects.
 The estimate is used to remove pathogen effects on the trait.
@@
 The main advantage of this two-step approach compared to the previously outlined methods to correct for pathogen population structure is that it generates additional information on the evolutionary dynamics of the trait in the pathogen population.
-The advances presented here are on the first step, while in the second step existing highly optimized tools to perform GWAS association tests under a variety of models can be employed.
-
-.
+The advances presented here are on the first step, while in the second step existing, highly optimized tools to perform GWAS association tests under a variety of models can be employed.
+
 % We hypothesis this should increase GWAS power for several reasons.
@@
 $\epsilon$ is modeled as a Gaussian random variable that is time- and phylogeny-independent.
-The expectation of $\epsilon$ is 0, meaning environmental effects are equally likely to raise or lower the trait from the pathogen-determined level.
-The parameter $\sigma_\epsilon^2$ measures the between-host variance of the environmental effect.
+The expectation of $\epsilon$ is 0, meaning non-heritable effects are equally likely to raise or lower the trait from the pathogen-determined level.
+The parameter $\sigma_\epsilon^2$ measures the between-host variance of the non-heritable effect.
 
@@
 The expectation $\bm{\mu}_{OU}$ is defined by equation \ref{eq:OUmean}, the diagonal elements of the covariance matrix $\boldsymbol{\Sigma}_{OU}$ are defined by equation \ref{eq:OUVar}, and the off-diagonal elements of $\boldsymbol{\Sigma}_{OU}$ by equation \ref{eq:OUcov}.
-Similarly, let $\bm{\epsilon}$ be a vector of the environmental part of the trait for each individual.
+Similarly, let $\bm{\epsilon}$ be a vector of the non-heritable part of the trait for each individual.
 $\bm{\epsilon}$ is a realization of a Gaussian random vector $\bm{\mathcal{E}} \sim \mathcal{N}\big(\bm{0}, \boldsymbol{\Sigma}_\mathcal{E}\big)$, where $\boldsymbol{\Sigma}_\mathcal{E}$ is a diagonal matrix with diagonal elements equal to $\sigma^2_\epsilon$.
@@
 
-Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}_v$, the pathogen effect on the trait, taking into account all available information - measured trait values, the pathogen phylogeny, and inferred POUMM parameters.
+Importantly, equation \ref{eq:MuG} is the maximum likelihood estimate for $\bm{g}$, the pathogen effect on the trait, taking into account all available information - measured trait values, the pathogen phylogeny, and inferred POUMM parameters.
 This estimator is an inverse-variance weighted average of measured trait ($\bm{z}$) and information from the POUMM evolutionary model ($\bm{\mu}_{OU}$).
-In other words, $g$ will be closer to the measured trait value if the trait is not very heritable.
-If the trait is highly heritable, $g$ will be closer to the expected value under the POUMM, i.e.
+In other words, $\bm{g}$ will be closer to the measured trait value if the trait is not very heritable.
+If the trait is highly heritable, $\bm{g}$ will be closer to the expected value under the POUMM, i.e.
 take more information from the phylogenetic relationships between infecting strains.
@@
 We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in \cite{Hodcroft2014}.
-Then, we simulated pathogen trait values $g$ along this phylogeny using the POUMM package in R \citep{Mitov2017a-POUMM}.
+Then, we simulated pathogen trait values $\bm{g}$ along this phylogeny using the POUMM package in R \citep{Mitov2017a-POUMM}.
+This part of the simulation is illustrated in Figure \ref{fig:spVLModel}A.
 For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
@@
 10 random positions had an effect size of 0.2 on the trait and 10 had an effect size of -0.2.
-This parameterization produced roughly normally distributed host trait values centered at 0 with variance equal to 25\% of the total trait variance, which we constrained to 0.73 based on the variance in log spVL values measured by \cite{Mitov2018}.
+This part of the simulation is illustrated in Figure \ref{fig:spVLModel}B.
+Our parameterization produced roughly normally distributed host trait values centered at 0 with variance equal to 25\% of the total trait variance, which we constrained to 0.73 based on the variance in log spVL values measured by \cite{Mitov2018}.
 We used 25\% host heritability for spVL based on \cite{McLaren2015}.
 
-Finally, we sampled an additional random environmental effect for each tip from a normal distribution centered at 0.
+Finally, we sampled an additional random environmental effect for each tip from a normal distribution centered at 0, as illustrated in Figure \ref{fig:spVLModel}C.
 The variance of this distribution was scaled based on the pathogen heritability of the trait, from 0 (no affect) in the scenario with 75\% pathogen heritability and 25\% host heritability to 0.44 in the scenario with 15\% pathogen heritability and 25\% host heritability.
-Figure \ref{fig:sim-design} shows a high-level schematic of this simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
+Figure \ref{fig:sim-design} provides a more detailed schematic of this simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
 
@@
 To ensure a fair comparison, we scaled trait values to have the same mean, zero, as host genetic effects so as not to bias the root mean squared error (RMSE) by a constant factor.
-As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to viral genetic effects and environmental effects is fixed.
+As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to pathogen genetic effects and environmental effects is fixed.
 Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.
@@
 Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations.
-These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny.
-This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait.
-Figure \ref{fig:tip-trait-correlation} shows in more detail how trait values are correlated at closely clustered tips when an infectious disease trait is highly heritable and selection strength is low/ stochastic fluctuations are low.
+Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure \ref{fig:tip-trait-correlation}) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.
 
@@
 arboricola} QDR analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.
-RMSE = Root mean square error, TPR = True positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
+RMSE = root mean square error, TPR = true positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
 		\label{fig:simulationResults}
@@
 
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
-	\caption{A high-level schematic of the experimental setup for the two application datasets.
-For application (A) to HIV-1 spVL in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
-We fit the POUMM to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
-For application (B) to \emph{A.
-thaliana}-\emph{X.
-Arboricola} QDR from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
-We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
-	\label{fig:application-exp-setup}
-	\end{center}
-\end{figure}
-
 We applied our framework to empirical data from two different host-pathogen systems with different experimental setups (Figure \ref{fig:application-exp-setup}).
-First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,392 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018.
+First, we used data collected by the Swiss HIV Cohort Study (SHCS) from 1,493 individuals in Switzerland infected with HIV-1 subtype B between 1994 and 2018.
 The SHCS provided viral load measurements, $pol$ gene sequences, and human genotype data for these individuals.
@@
 
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
+	\caption{A high-level schematic of the experimental setup for the two application datasets.
+For (A) HIV-1 spVL in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
+We fit the POUMM to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
+For (B) \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
+We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
+	\label{fig:application-exp-setup}
+	\end{center}
+\end{figure}
+
 We compared our proposed GWAS framework with a more standard approach by performing two different GWAS on the same SHCS human genotypes.
+We retained 1,392 individuals of European ancestry for the GWAS.
 In the (i) ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable.
@@
 However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}).
-Finally Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date.
+Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date.
+We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}).
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
@@
 I.e., with reference to Figure \ref{fig:application-exp-setup}, we selected $z_{11}$ for the first sample, $z_{23}$ for the second sample, and so on.
-In the second GWAS, we used our (very uncertain) estimates for the non-pathogen effects on QDR for each pairing.
+In the second GWAS, we used our estimates for the non-pathogen effects on QDR for each pairing.
 Since our method did not utilize phylogenetic information in this case, the estimated non-pathogen part of the trait is simply the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
@@
 I.e., with reference to Figure \ref{fig:application-exp-setup}, we used a scaled version of $z_{11} - \bar{z}_1$ for the first sample, $z_{23} - \bar{z}_2$ for the first sample, and so on.
-Figure \ref{fig:gwas-results-qdr} shows that XXX.
+Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants.
+Q-Q plots show the distribution of p-values are also very similar (\ref{fig:qq-plots-qdr}).
+In the first, standard GWAS, one \emph{A.
+thaliana} loci just exceeds the threshold for significant association after correction for multiple testing.
+In the second, corrected GWAS, no \emph{A.
+thaliana} variants are significantly associated with QDR to \emph{X.
+arboricola}.
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.png}
+	\caption{Results from comparative GWAS on \emph{A.
+thaliana}-\emph{X.
+arboricola} QDR data.
+The two facets show association p-values for the same host \emph{A.
+thaliana} variants in GWAS with two different response variables.
+On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
+On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
+In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
+thaliana} types.
+The alternating shades correspond to different chromosomes.
+The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
+	\label{fig:gwas-results-qdr}
+	\end{center}
+\end{figure}
 
@@
 By using information from the pathogen phylogeny, we show that it is possible to improve GWAS power to detect host genetic variants associated with a disease trait.
-This improved power is envisioned to obtain a better understanding of which host factors are broadly protective against a disease versus which increase susceptibility or disease severity.
+This improved power is envisioned to contribute to a better understanding of which host factors are broadly protective against a disease versus which increase susceptibility or disease severity.
 
 % Contextualize the study & address limitations
-The main novelty of our approach is to use these POUMM parameter estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects.
-In simulations, we show that when trait heritability amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
+The main novelty of our approach is to estimate parameters governing the evolutionary dynamics of a trait in the pathogen population and use these estimates to correct infectious disease trait values prior to performing GWAS, thereby estimating and removing pathogen effects.
+In simulations, we show that when trait heritability due to shared pathogen ancestry amongst infection partners is greater than approximately 25\%, GWAS power to detect host genetic variants associated with the same trait is reduced.
 Our method can correct for this effect in certain evolutionary scenarios by using information from the full pathogen phylogeny.
@@
 It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
-However, such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect.
+However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.
 In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
@@
 However, the overall picture is consistent between the two GWAS (Figure \ref{fig:gwas-results}A).
-Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL is robust to our correction for pathogen effects.
 For \emph{A.
-thaliana} QDR, inferred phylogenetic correlations are too weak to justify our phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
-\cite{Wang2018Two-wayGenomes} found no significant \emph{A.
+thaliana} QDR to \emph{X.
+arboricola}, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types.
+Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consistent with previous GWAS results from \cite{Wang2018Two-wayGenomes}.
+That study found no significant \emph{A.
 thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects.
 As with HIV-1 spVL, our results do not challenge this previous finding.
+Therefore, we conclude that GWAS for host determinants of HIV-1 subtype B spVL and \emph{A.
+thaliana} determinants of QDR to \emph{X.
+arboricola} are robust to our correction for pathogen effects.
 
@@
 In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.
-In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
+In cases where a correction can be estimated and applied, the corrected trait values are envisioned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
 Further, additional model complexity can be added to the GWAS association tests.
@@
 % Summarize the paper
-In summary, we present a coherent infectious disease GWAS framework which takes the pathogen phylogeny into account when searching for host determinants of a disease trait.
-We further show that the pathogen phylogeny only has an impact on the GWAS outputs if heritability of the trait amongst infection partners $>$ 25\%.
-This means that for the systems studied here, spVL in HIV and QDR for X.
+In summary, we present a coherent infectious disease GWAS framework that takes the pathogen phylogeny into account when searching for host determinants of a disease trait.
+We further show that the pathogen phylogeny only has an impact on the GWAS outputs if heritability of the trait amongst infection partners is $>$ 25\%.
+This means that for the systems studied here, spVL in individuals living with HIV and QDR for X.
 arboricola infections in A.
-thaliana, the phylogeny correction does not change GWAS results and thus previously published results are not biased by the pathogen phylogeny.
+thaliana, the phylogenetic correction does not change GWAS results.
+Our findings indicate previously published GWAS results for these systems are not biased by shared evolutionary history amongst infecting pathogen strains.
 
@@
 
-Whenever possible, we tried to parameterize our simulation model for spVL using empirical data.
+Whenever possible, we tried to parameterize our simulation model using empirical data on the spVL trait.
 We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}.
 Other studies have estimated slightly lower values though (Table \ref{tab:spVLmeasurements}).
-After allotting 25\% of this variance to the host part of spVL $h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between the viral part $g$ and the environmental part $\epsilon$ in different ratios to assess estimator performance across a range of spVL heritabilities.
+After allotting 25\% of this variance to a host part of spVL $h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between a viral part $g$ and an environmental part $e$ in different ratios to assess estimator performance across a range of spVL heritabilities.
 $h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
@@
 The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table \ref{tab:sim-params}, Figure \ref{fig:sigma-vs-alpha}).
-Finally, the environmental part of spVL $\epsilon$ was generated from a normal distribution with mean 0.
+Finally, the environmental part of spVL $e$ was generated from a normal distribution with mean 0.
 For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.
@@
 Given the stricter correction for multiple testing in this second simulation setup, the TPR decreased significantly across all three GWAS response variables used.
-
-\begin{table}[H]
-	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
-	\begin{tabularx}{\linewidth}{p{1.5cm}ll} \toprule 
-		Variable & Expression & Definition \\ \midrule 
-		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & Total spVL variance \\ 
-		$H^2_h$ & 0.25 & Host heritability of spVL \\
-		$H^2_{\bar{t}}$ & varied & Viral heritability of spVL at $\bar{t}$ \\
-		$\sigma^2_{h}$ & $\sigma_{h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
-		$\sigma^2_{p}(\bar{t})$ & $\sigma^2_{g_{v}}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in viral part of spVL at $\bar{t}$ \\ 
-		$\sigma^2_{\epsilon}$ & $\sigma_{\epsilon}^2 = \sigma_z^2 - \sigma^2_{p} - \sigma^2_{h}$ & Variance in environmental part of spVL \\ 
-		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in viral phylogeny \\ 
-		$\bm{g}_v$ & $\bm{g}_v \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & Viral part of spVL for all individuals \\ 
-		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
-		$g_{0}$ & 4.5 log copies/mL & $g$ at the root of the phylogeny  \\ 
-		$\alpha$ & varied & Selection strength of OU process \\ 
-		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g_{v}}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
-		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Viral phylogeny \\ 
-		$g_{h_i}$ & \makecell{$g_{h_i} = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \\ \delta \sum_{j = M/2}^{j = M}{G_{ij}}$} & Host part of spVL for individual $i$ \\ 
-		$G_{N \times M}$ & \makecell{$G_{ij} \sim Binom(2, p)\ \\ \forall i \in {1...N}, \forall j \in {1...M}$} & Host genotype matrix \\
-		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
-		$\delta$ & 0.2 & Host variant effect size \\ 
-		$M$ & 20 & Number of causal host variants \\ 
-		$\epsilon_i$ & $ \epsilon_i \sim Norm(0, \sigma^2_{\epsilon})$ & Environmental part of spVL for individual $i$ \\ 
-		$N$ & 500 & Number of simulated samples \\ \bottomrule 
-	\end{tabularx}
-	\label{tab:sim-params}
-\end{table}
 
@@
 thaliana} and \emph{X.
-arboricola} genotyping and quantitative disease resistance (QDR) measurements are described in detail in \cite{Wang2018Two-wayGenomes}.
+arboricola} genotyping and quantitative disease resistance (QDR) measurements were generated by \cite{Wang2018Two-wayGenomes} and are described in detail in that publication.
 Briefly, \cite{Wang2018Two-wayGenomes} infected different \emph{A.
@@
 arboricola} pathogen genotype matrix.
-The inferred tree topology (Figure \ref{fig:qdr-on-tree}) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes} generated using the unweighted pair group method with arithmetic mean (UPGMA).
+The inferred tree topology (Figure \ref{fig:qdr-on-tree}) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes}, which was generated using the unweighted pair group method with arithmetic mean (UPGMA).
 Compared to UPGMA, the neighbor-joining method we used relaxes the assumptions of a strict molecular clock and sampling all at the same time-point.
 For the trait value to fit the POUMM, we calculated mean QDR across all leaves infected on all hosts for each \emph{X.
-arboricola} strain.
+arboricola} strain (see Figure \ref{fig:application-exp-setup}B) We used PLINK version 2.0 to select bi-allelic variants from the VCF file using option --max-alleles 2.
+We then used options --maf 0.1 and --max-maf 0.9 to remove variants with minor allele frequencies less than 0.1 as in \cite{Wang2018Two-wayGenomes}.
+After filtering, approximately 1.1 million genetic variants from \emph{A.
+thaliana} were retained for GWAS (Table \ref{tab:variant-filtering-qdr}).
 
@@
 arboricola} QDR, respectively.
-Table \ref{tab:POUMM-parameters-our-estimates} gives the posterior mean values used for subsequent calculations.
+Tables \ref{tab:POUMM-parameters-our-estimates} and \ref{tab:POUMM-parameters-our-estimates-qdr} give the posterior mean values used for subsequent calculations.
 
 \subsection*{Phylogenetic trait correction}
-We corrected HIV-1 spVL and \emph{X.
+We estimated the pathogen and non-pathogen effects on HIV-1 spVL and \emph{X.
 arboricola} mean QDR trait values using the method described in this paper.
@@
 arboricola} case, each sample corresponds to one \emph{X.
-arboricola} strain and the mean QDR score for that strain.
+arboricola} strain and the mean QDR score for that strain across all host types (see Figure \ref{fig:application-exp-setup}).
 To calculate the expected correlation in trait values between tips in the pathogen phylogeny, we used the function ``covVTipsGivenTreePOUMM'' in the same package.
@@
 \subsection*{Association testing}
-We performed two GWAS using the same human genotype data from the SHCS.
+We performed two comparative GWAS for each system, using the same host genotype data across the two GWAS.
 For the first ``GWAS with standard trait value'' we used the total (uncorrected) trait values ($\bm{z}$) as the response variable for association testing, replicating a standard GWAS set-up.
-For the second ``GWAS with estimated non-pathogen part of trait'' we replaced total trait values with the estimated non-pathogen component of the trait ($\hat{\bm{e}}$) as the response variable.
-Association testing was performed using a linear association model in PLINK version 2.3 \citep{Chang2015} with sex and the top 5 principle components of host genetic variation included as covariates.
-The sex and principle components covariates were included to reduce residual variance in spVL and control for confounding from host population structure, respectively.
-
-\subsection{Phylogenetic uncertainty}
+For the second ``GWAS with estimated non-pathogen part of trait'' we replaced total trait values with the estimated non-pathogen component of the trait ($\hat{\bm{\epsilon}}$) as the response variable.
+Association testing was performed using a linear association model in PLINK version 2.3 and 2.0, respectively \citep{Chang2015} with the top 5 principle components of host genetic variation included as covariates.
+For the HIV-1 spVL GWAS, we additionally included sex as a covariate.
+The sex and principle components covariates were included to reduce residual variance and control for confounding from host population structure, respectively.
+
+\subsection*{Phylogenetic uncertainty}
 Our method assumes the phylogeny accurately reflects the evolutionary relationships between pathogen strains.
@@
 For the HIV application, we additionally tested the sensitivity of the inference to phylogenetic uncertainty.
-We inferred the phylogeny again, this time using the IQ-TREE option ``wt'' to output all locally optimal trees.
-We fit the POUMM to two randomly selected trees from this set.
-Posterior mean parameter estimates were within the 95\% HPD intervals inferred from the maximum likelihood tree.
-GWAS results were also comparable.
+We inferred the phylogeny again, this time using the IQ-TREE option -wt to output all locally optimal trees.
+We fit the POUMM to two randomly selected trees from this set and repeated the trait correction and association testing steps using these trees and the corresponding POUMM parameter estimates.
 
@@
 
-Here we show the root mean square error (RMSE) of $\bm{z} - \bm{\bar{z}}$ as an estimate for $\bm{h}$ should be $\approx$ 0.74 in our simulation scheme. First we write the expression for the RMSE:
+Here we show the root mean square error (RMSE) of $\bm{z} - \bm{\bar{z}}$ as an estimate for the host part of the trait, denoted $\bm{h}$, should be $\approx$ 0.74 in our simulation scheme.
+First we write the expression for the RMSE:
 
@@
 
-Note that $z - \bar{z}$ differs from $h$ due to a viral effect and an environmental effect. So the term inside the square root equals the combined variance of these two effects:
+Note that under our simulation setup, $z - \bar{z}$ differs from $h$ due to the pathogen effect $g$ and the environmental effect $e$.
+So the term inside the square root equals the combined variance of these two effects:
 
 \begin{equation}
-    RMSE = \sqrt{\sigma^2_{p} + \sigma^2_\epsilon}
+    RMSE = \sqrt{\sigma^2_{g} + \sigma^2_e}
 \end{equation}
@@
 \begin{split}
-    \sigma^2_{h} + \sigma^2_{p} + \sigma^2_\epsilon = \sigma^2_z \\
-    0.25 * \sigma^2_{z} + \sigma^2_{p} + \sigma^2_\epsilon = \sigma^2_z \\
-    \sigma^2_{p} + \sigma^2_\epsilon = 0.75 * \sigma^2_z \\
-    \sigma^2_{p} + \sigma^2_\epsilon = 0.75 * 0.73 \\
-    \sigma^2_{p} + \sigma^2_\epsilon = 0.55
+    \sigma^2_{h} + \sigma^2_{g} + \sigma^2_e = \sigma^2_z \\
+    0.25 * \sigma^2_{z} + \sigma^2_{g} + \sigma^2_e = \sigma^2_z \\
+    \sigma^2_{g} + \sigma^2_e = 0.75 * \sigma^2_z \\
+    \sigma^2_{g} + \sigma^2_e = 0.75 * 0.73 \\
+    \sigma^2_{g} + \sigma^2_e = 0.55
 \end{split}
@@
 \newpage
+
+\begin{figure}[H]
+\begin{center}
+\includegraphics[width = 0.7\linewidth]{figures/sigma_vs_alpha.png}
+	\caption{Relationship between the selection strength parameter $\alpha$ and the intensity of stochastic evolutionary fluctuations parameter $\sigma$ at two different heritability ($H^2$) values in the  simulation scheme.
+$\sigma$ was determined as a function of $\alpha$ and $H^2$ under the POUMM (function given in Table \ref{tab:sim-params}).}
+	\label{fig:sigma-vs-alpha}
+	\end{center}
+\end{figure}
 
@@
 	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{h}$};
-	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
-	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{\epsilon}$};
+	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g}(\bar{t})$};
+	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{e}$};
 	\draw[taro] (H2h) -- (varh);
@@
 	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
-	\node[anchor = west] (blabel) at (-1, 4.5) {\textbf{B} Generating viral effects};
+	\node[anchor = west] (blabel) at (-1, 4.5) {\textbf{B} Generating pathogen effects};
 	\node[constnode] (theta) at (0, 1.5) {$\theta$};
@@
 	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
-	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g_{v}}(\bar{t})$};
+	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g}(\bar{t})$};
 	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
-	\node[snode] (v) at (3.6, 3.25) {$\vec{p}$};
+	\node[snode] (v) at (3.6, 3.25) {$\vec{g}$};
 	\draw[taro] (theta) -- (v);
@@
 	\node[constnode] (d) at (1.8, 1.5) {2};
-	\node[detnode] (p) at (3.6, 1.5) {$g$};
+	\node[detnode] (p) at (3.6, 1.5) {$p$};
 	\node[snode] (G) at (1.8, 3) {$G$};
@@
 	\node[anchor = west] (dlabel) at (-1, 3) {\textbf{D} Generating environmental effects};
-	\node[detnode] (vare) at (0, 0) {$\sigma^2_{\epsilon}$};
+	\node[detnode] (vare) at (0, 0) {$\sigma^2_{e}$};
 	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
-	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{\epsilon}$};
+	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{e}$};
 	\draw[taro] (zero) -- (e);
@@
 All parameters are defined in Table \ref{tab:sim-params}, as well as the values or expressions used for them.
-(A) shows how the variance in the simulated environmental effect $\sigma^2_{\epsilon}$ is smaller if the master pathogen heritability value $H^2_{\bar{t}}$ is higher and vice-versa.
+(A) shows how the variance in the simulated environmental effect $\sigma^2_{e}$ is smaller if the master pathogen heritability value $H^2_{\bar{t}}$ is higher and vice-versa.
 (B) shows the OU parameters and the pathogen phylogeny, which generate the Gaussian-distributed pathogen effects.
@@
 In other words, we use $\sigma$ to maintain the desired pathogen heritability while varying $\alpha$.
-We generate a new random phylogeny for each simulation.
 (C) shows how host genotypes are drawn to generate host effects.
@@
 We assume half the variants have a positive effect and half have a negative effect.
-The allele frequency $g$ for the causal variants set so that we achieve the desired variance in the host effects.
+The allele frequency $p$ for the causal variants set so that we achieve the desired variance in the host effects.
 (D) shows that the environmental effect is drawn from a Gaussian distribution with mean zero and variance as determined in part (A).}
 	\label{fig:sim-design}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.7\linewidth]{figures/sigma_vs_alpha.png}
-	\caption{Relationship between the selection strength parameter $\alpha$ and the intensity of stochastic evolutionary fluctuations parameter $\sigma$ at two different heritability ($H^2$) values in the  simulation scheme.
-$\sigma$ was determined as a function of $\alpha$ and $H^2$ under the POUMM (function given in Table \ref{tab:sim-params}).}
-	\label{fig:sigma-vs-alpha}
-	\end{center}
 \end{figure}
@@
 \begin{figure}[H]
+	\centering
+		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
+		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on HIV-1 spVL data from the SHCS. We ran two different MCMC chains to ensure the estimates converged.}
+\label{fig:poumm-parameters}
+\end{figure}
+
+\begin{figure}[H]
 \begin{center}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on SHCS data. We ran two different MCMC chains to ensure the estimates converged.}
-\label{fig:poumm-parameters}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
 	\includegraphics[width=\linewidth]{figures/qq_plots.png}
-	\caption{Quartile-quartile plots from association tests. The dashed green line shows the y = x line.}
+	\caption{Quartile-quartile plots from HIV-1 spVL association tests. The dashed green line shows the y = x line.}
 	\label{fig:qq-plots}
-\end{figure}
-
-\begin{figure}[H]
-	\includegraphics[width=\linewidth]{figures/host_genotype_pca.png}
-	\caption{SHCS individuals and HapMap3 individuals plotted along the top three principle components of genetic variation.
-Points with black borders are within the thresholds used to select individuals of likely European ancestry.}
-	\label{fig:PCA}
 \end{figure}
@@
 \begin{center}
-\includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas.png}
-	\caption{Results from comparative GWAS on \emph{A.
-thaliana}-\emph{X.
-arboricola} QDR data.
-The two facets show association p-values for the same host \emph{A.
-thaliana} variants in GWAS with two different response variables.
-On the left, we used unmodified (total) QDR values.
-On the right, we used our estimates for the non-pathogen effects on QDR.
-In this case, these values are total QDR minus a pathogen strain-specific scaled mean QDR.
-The alternating shades correspond to different chromosomes.
-The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
-	\label{fig:gwas-results-qdr}
+\includegraphics[width = \linewidth]{figures/qq_plots_arabidopsis_xanthomonas.png}
+	\caption{Quartile-quartile plots from \emph{A. thaliana}-\emph{X. arboricola} QDR association tests. The dashed blue line shows the y = x line.}
+	\label{fig:qq-plots-qdr}
 	\end{center}
@@
 
+\begin{figure}[H]
+	\includegraphics[width=\linewidth]{figures/host_genotype_pca.png}
+	\caption{SHCS individuals and HapMap3 individuals plotted along the top three principle components of genetic variation.
+Points with black borders are within the thresholds used to select individuals of likely European ancestry.}
+	\label{fig:PCA}
+\end{figure}
+
 \end{doublespace}  % don't doublespace tables
@@
 \newpage
+
+\begin{table}[H]
+	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
+	\begin{tabularx}{\linewidth}{p{1.5cm}p{6.5cm}l} \toprule 
+		Variable & Expression & Definition \\ \midrule 
+		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & Total spVL variance \\ 
+		$H^2_h$ & 0.25 & Host heritability of spVL \\
+		$H^2_{\bar{t}}$ & varied & Pathogen heritability of spVL at $\bar{t}$ \\
+		$\sigma^2_{h}$ & $\sigma_{h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
+		$\sigma^2_{g}(\bar{t})$ & $\sigma^2_{g}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in pathogen part of spVL at $\bar{t}$ \\ 
+		$\sigma^2_{e}$ & $\sigma_{e}^2 = \sigma_z^2 - \sigma^2_{g} - \sigma^2_{h}$ & Variance in environmental part of spVL \\ 
+		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in pathogen phylogeny \\ 
+		$\bm{g}$ & $\bm{g} \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & Pathogen part of spVL for all individuals \\ 
+		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
+		$g_{0}$ & 4.5 log copies/mL & $g$ at the root of the phylogeny  \\ 
+		$\alpha$ & varied & Selection strength of OU process \\ 
+		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
+		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Pathogen phylogeny \\ 
+		$h_i$ & $h_i = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & Host part of spVL for individual $i$ \\ 
+		$G_{N \times M}$ & \makecell{$G_{ij} \sim Binom(2, p)\ \\ \forall i \in {1...N}, \forall j \in {1...M}$} & Host genotype matrix \\
+		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
+		$\delta$ & 0.2 & Host variant effect size \\ 
+		$M$ & 20 & Number of causal host variants \\ 
+		$e_i$ & $ e_i \sim Norm(0, \sigma^2_{e})$ & Environmental part of spVL for individual $i$ \\ 
+		$N$ & 500 & Number of simulated samples \\ \bottomrule 
+	\end{tabularx}
+	\label{tab:sim-params}
+\end{table}
 
@@
 	$\alpha$ & 57.65 & (19.49, 95.2) \\
-	$\sigma_e$ & 0.54 & (0.43, 0.65) \\ 
+	$\sigma_\epsilon$ & 0.54 & (0.43, 0.65) \\ 
 	$H^2_{\bar{t}}$ & 0.45 & (0.24, 0.67) \\ \hline
@@
 	\label{tab:POUMM-parameters-our-estimates}
-\end{table}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for QDR based on \emph{A. thaliana}-\emph{X. arboricola} data data. HPD = Highest posterior density.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Parameter & Posterior mean & 95\% HPD \\ \hline 
-	${g_{0}}$ & 0.40 & (0.01, 0.78) \\
-	$\theta$ & 0.39 & (0.30, 0.49) \\
-	$\sigma$ & 0.03 & (0.0, 0.06) \\
-	$\alpha$ & 0.03 & (0.0, 0.05) \\
-	$\sigma_e$ & 0.18 & (0.08, 0.27) \\ 
-	$H^2_{\bar{t}}$ & 0.33 & (0.0, 0.77) \\ \hline
-	\end{tabularx}
-	\label{tab:POUMM-parameters-our-estimates-qdr}
 \end{table}
@@
 \label{tab:gwas-top-associations}
+\end{table}
+
+\begin{table}[H]
+\centering
+\caption{Results comparison using three different approximate maximum likelihood HIV-1 phylogenies (see Materials and Methods).}
+\begin{tabularx}{\linewidth}{p{1.8cm}p{2cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}p{1.8cm}} % {lrrrrrr} {llclcccc}
+\hline
+& \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\main tree}} & \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\tree 2}} & \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\tree 3}} \\
+  \hline
+ Variant & Effect size & p-value & Effect size & p-value & Effect size & p-value \\
+  \hline
+    rs59440261 & -0.22 & $2.6 \times 10^{-10}$ & -0.22 & $1.5 \times 10^{-10}$ & -0.24 & $1.4 \times 10^{-10}$ \\ 
+    rs1015164 & 0.078 & $8.5 \times 10^{-6}$ & 0.076 & $9.5 \times 10^{-6}$ & 0.083 & $9 \times 10^{-6}$ \\ 
+   \hline
+\end{tabularx}
+\label{tab:comp-gwas-mclaren-snps-alt-trees}
+\end{table}
+
+\begin{table}[H]
+\caption{POUMM parameter estimates for QDR based on \emph{A. thaliana}-\emph{X. arboricola} data. HPD = Highest posterior density.}
+	\begin{tabularx}{\linewidth}{XXX} \hline 
+	Parameter & Posterior mean & 95\% HPD \\ \hline 
+	${g_{0}}$ & 0.40 & (0.01, 0.78) \\
+	$\theta$ & 0.39 & (0.30, 0.49) \\
+	$\sigma$ & 0.03 & (0.0, 0.06) \\
+	$\alpha$ & 0.03 & (0.0, 0.05) \\
+	$\sigma_\epsilon$ & 0.18 & (0.08, 0.27) \\ 
+	$H^2_{\bar{t}}$ & 0.33 & (0.0, 0.77) \\ \hline
+	\end{tabularx}
+	\label{tab:POUMM-parameters-our-estimates-qdr}
 \end{table}
@@
 \begin{table}[H]
-\caption{Number of variants for \emph{A. thaliana} QDR GWAS after sequential filtering steps.}
+\caption{Number of variants for \emph{A.
+thaliana} QDR GWAS after sequential filtering steps.
+The last entry lists variants without GWAS p-values because PLINK assessed the correlation between predictor variables (the variant and the top 5 principle components of host genetic variation are predictors) to be too strong.
+This did not occur in the HIV-1 spVL GWAS.}
 	\begin{tabularx}{\linewidth}{XX}
@@
 	\hline 
-	Raw data & 11769920 \\
+	Raw data & 12883854 \\
+	Bi-allelic variants & 11769920 \\
 	Minor allele frequency $<$ 0.1 & 1743952 \\
-	NA p-value due to too-high correlation with a covariate & 1070541 \hline			
+	NA p-value (too-high covariate correlation) & 1070541 \\ \hline			
 	\end{tabularx}
@@
 \section*{Editors’ comments}
-AE: I have now received two reviews of your paper.
+I have now received two reviews of your paper.
 Both reviewers were excited about the methodological concept, but both where also concerned about 1) the perceived narrow application of the methodology and 2) the example based on data which are not accessible.
@@
 \vspace{5mm}
-\textcolor{blue}{We edited the introduction to make it clearer when this method is applicable: ``This framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.''}
-
-\textcolor{blue}{We also expanded the discussion to include the possibility of transcriptomic applications and highlight that even when this method is not anticipated to increase GWAS power, it is still interesting to fit the POUMM and estimate pathogen effects on different traits: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma \citep{An2018HostCarcinoma}, Hepatitis C treatment success \citep{Ansari2017Genome-to-genomeVirus}, and susceptibility to or severity of certain bacterial infections, e.g.
-\cite{Messina2016ImpactInfections, Donnenberg2015BacterialStudies}.
-Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model \citep{Rohlfs2014ModelingVariation}.
-Thus, one could also estimate pathogen effects on host gene expression.
-In particular, it will be of interest to understand the evolutionary dynamics of these traits in the respective pathogen populations.''}
+\textcolor{blue}{We edited the introduction to make it clearer when this method is applicable: ``The introduced framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.''}
+
+\textcolor{blue}{Then, as mentioned, we included a second application of our method to a phytopathogenic bacteria.}
+
+\textcolor{blue}{We also expanded the discussion section that gives more specific examples of future potential applications, including the possibility of transcriptomic applications, which we hadn't previously thought about but should be possible: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g.
+Donnenberg et al.
+(2015); Messina et al.
+(2016).
+Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014).
+Thus, one could also estimate pathogen effects on host gene expression.''}
 
@@
 arboricola}.
-The findings are highlighted in the discussion: ``For \emph{A.
-thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
-\cite{Wang2018Two-wayGenomes} found no significant \emph{A.
-thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects.
-As with HIV-1 spVL, our results do not challenge this previous finding.''}
+The findings are highlighted in the discussion: ``Given our estimates for trait heritability and selection strength on HIV-1 spVL and A.
+thaliana quantitative disease resistance to X.
+arboricola, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure 2)...
+For A.
+thaliana QDR to X.
+arboricola, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types.
+Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consitent with previous GWAS results from Wang et al.
+(2018).''}
     \item In the case of co-infection, as sometimes is the case of HIV and TB, how would the method perform? \\
-    \textcolor{blue}{We addressed this point in the discussion: ``Finally, additional model complexity can be added to the GWAS association tests.
+    \textcolor{blue}{We addressed this point in the discussion: ``Further, additional model complexity can be added to the GWAS association tests.
 For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power.
@@
 However, we note that McLaren et al.
-2015 found similar signal of association on chromosome 3 and 6 when they filtered to elite controllers in a standard GWAS setup.}
+(2015) found similar signal of association on chromosome 3 and 6 when they filtered to elite controllers in a standard GWAS setup.}
     \item In the case of determining whether viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, I am not sure I understand the premise of this hypothesis.
@@
 \\
-    \textcolor{blue}{We edited the introduction to make it clearer why one should consider pathogen effects in a host GWAS: ``Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait.
-Corrections for these types of population structure in human GWAS cohorts are well-developed and widely accepted  \citep{Astle2009, Price2006}.
-More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}...
-[our method] estimates and removes pathogen effects on the trait so that downstream GWAS association testing with host genetic variants can be performed as normal under the assumption of independent samples''}
+    \textcolor{blue}{We edited the introduction to remove the confounding and extraneous variables terminology and make it clearer why one should consider pathogen effects in a host GWAS: ``Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait.
+Corrections for these types of population structure in human GWAS cohorts are well- developed and widely accepted (Astle and Balding, 2009; Price et al., 2006).
+More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017).'' We also make it clearer how our method addresses this problem: ``In a first step, we fit an evolutionary model to trait data and the pathogen phylogeny.
+This first step provides an estimate of the correlation structure of the trait due to heritable pathogen effects.
+The estimate is used to remove pathogen effects on the trait.
+In the second step, the resulting corrected trait data is used in a GWAS with host genetic variants.
+The GWAS can be performed as normal under the assumption of independent samples.''}
     \item What type of clinical information/trait is included under the environmental effect? How the diversity in environmental effects across the population affects the method? \\
@@
 
-    \textcolor{blue}{In addition to the change described in response 1, we also added text to the new approaches section to make it clearer that our method is complementary to but does not replace other methods to account for diversity in environmental effects (we only account for heritable pathogen genetic effects): ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
-This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.''}
+    \textcolor{blue}{We also added text to the new approaches section to make it clearer that our method is complementary to but does not replace other methods to account for diversity in environmental effects (we only account for heritable pathogen genetic effects): ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
+This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
+We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes.''}
     \item Lines 194-196:  please expand on the parameters of the simulation in greater detail. \\
-    \textcolor{blue}{We edited this paragraph to provide a full description of the simulation scheme.}  
+    \textcolor{blue}{We added a full description of the simulation scheme: ``We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in Hodcroft et al.
+(2014)...
+Finally, we sampled an additional random environmental effect for each tip from a normal distribution centered at 0, as illustrated in Figure 1C.''}
     \item Lines 205-207: it would be best, if there is not a limitation in the number of words, to report here “why” instead of saying that it has been shown in SI. \\
-    \textcolor{blue}{We added the suggested explanation: ``As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to viral genetic effects and environmental effects is fixed.
+    \textcolor{blue}{We added the suggested explanation: ``As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to pathogen genetic effects and environmental effects is fixed.
 Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.''}
@@
     \textcolor{blue}{We edited this statement to be clear that we mean ``pathogen heritability''.
-We also tried to explain the concept of heritability due to the pathogen better in the introduction: ``Variation in infectious disease traits like viral load or infection severity can come from several sources.
+We also tried to explain the concept of heritability due to the pathogen better in the New Approaches section: ``Variation in infectious disease traits like viral load or infection severity can come from several sources.
 These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature.
@@
 This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
-Broad-sense pathogen heritability $H^2$ quantifies the fraction of total variance in a trait that is ``inherited'' from infection partner to infection partner, i.e., due to pathogen factors.''}
+We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes.
+Broad-sense pathogen heritability H2 quantifies the fraction of total variance in a trait that is “inherited” from infection partner to infection partner, i.e., due to pathogen factors.''}
 \end{enumerate}
@@
 arboricola} pathosystem to demonstrate the broader applicability of our method.
-Unfortunately neither dataset we analyzed had sufficiently highly correlated trait values amongst related pathogen strains to generate novel biological insights.
+Unfortunately, neither dataset we analyzed had sufficiently highly correlated trait values amongst related pathogen strains to generate novel biological insights.
 We edited the text to emphasize that an advantage of our approach is that one gets information on the evolutionary dynamics of the trait in  the pathogen population, even if GWAS results are not changed.}
@@
     \item How does the method compare to the ATOMM approach in https://www.pnas.org/content/115/24/E5440? Can this be applied to any of the simulated or real data here for comparison. \\
-    \textcolor{blue}{We cannot apply ATOMM directly to our simulated data because we do not simulate pathogen genotypes, rather we simulate only pathogen trait values.
-Further, ATOMM is only implemented for haploid alleles, making it not ideal for the HIV data analyzed.
+    \textcolor{blue}{We could not apply ATOMM directly to our simulated data because we do not simulate pathogen genotypes, rather we simulate only pathogen trait values.
+ATOMM is also only implemented for haploid alleles, making it not ideal for the HIV data analyzed.
 Instead, we were able to access the same dataset ATOMM was tested on in their original publication.
@@
 
-    \textcolor{blue}{We present the findings of this comparison in the discussion: ``For \emph{A.
-thaliana} QDR, inferred phylogenetic correlations are too weak to justify a phylogenetic correction (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
-\cite{Wang2018Two-wayGenomes} found no significant \emph{A.
-thaliana} variants associated with QDR using a linear mixed model jointly accounting for host genetic effects, pathogen genetic effects, and interaction effects.
-As with HIV-1 spVL, our results do not challenge this previous finding.''}
-
-    \textcolor{blue}{We also highlight the linear mixed model approach presented in ATOMM as an alternate means to take pathogen effects into accout: ``As in the \emph{A.
-thaliana}-\emph{X.
-arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using the corrected trait values in a GWAS.
-In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.''}
+    \textcolor{blue}{We present the findings of this comparison in the discussion: ``For comparison, we also fit the POUMM to quantitative disease resistance measurements from A.
+thaliana infected with the phytopathogenic bacteria X.
+arboricola.
+We estimated X.
+arboricola virulence heritability to be 33\% (95\% HPD 0 - 77\%).
+(Wang et al., 2018) originally estimated a QDR heritability of 44\% in this dataset, falling within the wide range of our estimate.
+We note that Wang et al.
+(2018) used a linear mixed model in which the experimental unit is QDR scored on individual leaves, whereas our estimate is based on much coarser binning of QDR scores into a mean score across all leaves on all host accessions and all replicates (N = 22)...
+For A.
+thaliana QDR to X.
+arboricola, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
+We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types.
+Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consistent with previous GWAS results from Wang et al.
+(2018).''}
+
+    \textcolor{blue}{We also highlight the linear mixed model approach presented in ATOMM as an alternate means to take pathogen effects into account: ``As in the A.
+thaliana-X.
+arboricola application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS.
+In this case, one may wish to use a linear mixed model as in (Wang et al., 2018), where the pathogen effect is co-estimated as a random effect.''}
     \item As I understand it the POUMM method assumes a 'correct' phylogeny.
@@
 \\
-    \textcolor{blue}{This is correct, our method assumes the phylogeny accurately reflects the evolutionary history of the pathogen strains.
-We tested the sensitivity of the method to phylogenetic uncertainty in the HIV application.
-This is described in the section ``'': ``''.}
+    \textcolor{blue}{This is correct.
+We added a section ``Phylogenetic uncertainty'' in the Materials and Methods to describe an extra sensitivity test for the HIV analysis: ``Our method assumes the phylogeny accurately reflects the evolutionary relationships between pathogen strains...
+For the HIV application, we additionally tested the sensitivity of the inference to phylogenetic uncertainty.
+We inferred the phylogeny again, this time using the IQ-TREE option -wt to output all locally optimal trees.
+We fit the POUMM to two randomly selected trees from this set and repeated the trait correction and association testing steps using these trees and the corresponding POUMM parameter estimates.'' The results are briefly described in the Results section: ``We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table S4).''}
 
@@
 \\
-    \textcolor{blue}{We implemented this suggestion and added two new main text figures.
-They are discussed in the section ``Estimator accuracy'': ``Figure \ref{fig:simulationTraitExample} gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations.
-These parameter combinations mean trait values are phylogentically correlated at clustered tips in the phylogeny.
-This information is the basis from which our method estimates the pathogen and corresponding non-pathogen effects on the trait.
-Figure \ref{fig:tip-trait-correlation} shows in more detail how trait values are correlated at closely clustered tips when an infectious disease trait is highly heritable and selection strength is low/ stochastic fluctuations are low.''}
+    \textcolor{blue}{We implemented this suggestion and added two new main text figures (Figures 3 and 4).
+They are discussed in the section ``Estimator accuracy'': ``Figure 3 gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations.
+Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
     \item Do the authors know how the computational requirements of their ML estimation method scales with sample size? Would it work for a large SARS-CoV-2 phylogeny for instance? \\
@@
 We additionally verified that the matrix inversion steps required for the last ML estimation step based on the fitted POUMM parameters ran successfully on a simulated tree of 10,000 tips using 4 cores with maximum memory use of 6500MB per core.
-We now highlight the method's scalability in the discussion: ``The method relies on the freely available R package POUMM \citep{Mitov2017a-POUMM}, which scales to trees of up to 10,000 tips \citep{Mitov2019}.''}
+We now highlight the method's scalability in the discussion: ``Our method relies on the freely available R package POUMM (Mitov and Stadler, 2017), which scales to trees of up to 10,000 tips (Mitov and Stadler, 2019).''}
     \item Fig S3 would be useful, but isn't really readable at present. Perhaps consider an interactive tree viewer to make this more useful. \\
@@
 thaliana}-\emph{X.
-arboricola} dataset, which is much more legible due to having fewer samples (Figure \ref{fig:qdr-on-tree}).
-We also added sub-sampled phylogenies from the simulation study annotated with trait values in Figure \ref{fig:simulationTraitExample} to demonstrate what phylogenetically-correlated trait values look like.
+arboricola} dataset, which is much more legible due to having fewer samples (Figure S8).
+We also added sub-sampled phylogenies from the simulation study annotated with trait values in Figure 3 to demonstrate what phylogenetically-correlated trait values look like in a successful simulation example.
 Finally, we now reference the highest correlation values expected between any two tips in the HIV-1 and \emph{X.
-arboricola} phylogenies to add a quantative element to the discussion of phylogenetic correlations: ``The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
-However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability...
-Given the posterior mean estimates for the POUMM parameters [using \emph{X.
-arboricola} data], expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application).
+arboricola} phylogenies to add a quantitative element to the discussion of phylogenetic correlations: ``The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
+However, Figure S5 shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.'' Then for the \emph{A.
+thaliana}-\emph{X.
+arboricola} application: ``Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application).
 Thus, the phylogeny is not very informative for a trait value correction.''}
     \item The authors I think only look at power/TPR. What about the effect on false positive rate? \\
-    \textcolor{blue}{We re-ran the simulations to assess the false positive rate and include the results in the Materials and methods section: ``For our main results (Figure \ref{fig:simulationResults}) we simulated 20 truly associated variants, as described above.
+    \textcolor{blue}{We re-ran the simulations to assess the false positive rate and include the results in the Materials and Methods section: ``For our main results (Figure 2) we simulated 20 truly associated variants, as described above.
 To also check the false positive rate (FPR), we re-ran the simulations with an additional 80 non-associated variants.
-Across all the association tests in this second simulation setup (7 $H^2$ levels $\times$ 10 $\alpha$ levels $\times$ 100 variants $\times$ 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005  using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value.
+Across all the association tests in this second simulation setup (7 $H^2$ levels × 10 $\alpha$ levels × 100 variants × 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005 using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value.
 These rates are comparable to the expected FPR of 0.0005 at significance level 0.05 corrected for 100 tests.
@@
     \item Is there any explanation why the simulated (actual) and estimated non-pathogen trait in fig 2B start to look quite different at stronger selection strengths. \\
-    \textcolor{blue}{We added a figure (Figure \ref{fig:tip-trait-correlation}) to illustrate how the correlation between clustered tips on the phylogenetic tree decreases with increasing selection strength/ increasing stochasticity.
-Without phylogenetic signal, our method has no information to estimate the pathogen effect on the trait.}
+    \textcolor{blue}{We added a figure (Figure 4) to illustrate how the correlation between clustered tips on the phylogenetic tree decreases with increasing selection strength/ increasing stochasticity.
+Without phylogenetic signal, our method has no information to estimate the pathogen effect on the trait.
+This is described in the section ``Estimator accuracy'': ``Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
     \item Using plink and removing samples with divergent ancestry is probably a little outdated for host GWAS.
@@
 arboricola}, so we feel that updating our example GWAS is beyond the scope of the current study.
-We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS...
-In cases where a correction can be estimated and applied, it is still possible to use any of the number of previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
+We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS....
+In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK (Chang et al., 2015), though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
 \end{enumerate}
@@
 Maybe remove 'two', as this will change anyway.\\
-    \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}.
-Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \citep{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \citep{collins_phylogenetic_nodate}.''}
+    \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017).
+Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in (Earle et al., 2016) and modified association tests that account for phylogenetic relationships amongst samples as in (Collins and Didelot, 2018).''}
     \item L110 'Of these, only pathogen effects are heritable from one transmission partner to another'.
@@
     \item Please add the link to the github code in the framework section at the end of the methods.\\
-    \textcolor{blue}{We added the link in the section ``Phylogenetic spVL correction'': ``All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.''}
+    \textcolor{blue}{We added the link in the section ``Phylogenetic trait correction'': ``All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.''}
     \item L283 'In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.' Could the authors give a biological example of this?
 Would antimicrobial resistance variation fit the bill?\\
+    \textcolor{blue}{We have not fit the POUMM to data from enough systems to have a clear example of this, so instead we suggest other systems that might fall into this case: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g.
+Donnenberg et al.
+(2015); Messina et al.
+(2016).
+Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014).
+Thus, one could also estimate pathogen effects on host gene expression.''}
+
     \textcolor{blue}{We anticipate antimicrobial resistance would be under strong selective pressure, but agree this could produce phylogenetically correlated trait values.
-We added a statement in the discussion: ``It is possible that traits under strong selection pressure would also be phylogenetically structured, for instance antimicrobial resistance, though such traits might violate the POUMM assumption that trait values vary as a random walk if they are caused by few mutations of strong affect.''}
+We added a statement in the discussion: ``It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
+However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.''}
 \end{enumerate}

### 2022-06-23 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 
-% \begin{flushright} % for MBE submission
-% Submission type: Article\\
-% Section: Methods
-% \end{flushright}
+\begin{flushright} % for MBE submission
+Submission type: Article\\
+Section: Methods
+\end{flushright}
 
@@
 More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}.
-Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \citep{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \citep{collins_phylogenetic_nodate}.
+Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \cite{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \cite{collins_phylogenetic_nodate}.
 % \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
@@
 GWAS using paired host-pathogen genotype data have previously been done to elucidate the marginal and interaction effects of host and pathogen genetic variants.
-Methods to account for microbial population structure when testing for marginal host associations or host-pathogen interaction effects include adding the microbial kinship matrix as a random effect in a linear mixed model as in \citep{Wang2018Two-wayGenomes} and using principle components derived from either this matrix or the pathogen phylogeny as covariates in a linear model as in \citep{Naret2018}.
+Methods to account for microbial population structure when testing for marginal host associations or host-pathogen interaction effects include adding the microbial kinship matrix as a random effect in a linear mixed model as in \cite{Wang2018Two-wayGenomes} and using principle components derived from either this matrix or the pathogen phylogeny as covariates in a linear model as in \cite{Naret2018}.
 These methods focus on capturing and accounting for correlations due to the pathogen phylogeny, without further investigating the nature of these correlations.
@@
 
-\begin{figure}[H]
-	\begin{center}
-	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
-	\caption{A high-level schematic of our POUMM-based simulation framework in the context of HIV-1 spVL.
-(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
-(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
-Each variant can be present in 0, 1, or 2 copies.
-Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
-(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
-These three effects sum to the trait value for each simulated individual.}
-		\label{fig:spVLModel}
-	\end{center}
-\end{figure}
+% Figure 1 location
 
@@
 	\item Construct a pathogen phylogeny using the pathogen genome sequences.
-	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with e.g. the R package POUMM \citep{Mitov2017a-POUMM}.
+	\item Estimate the parameters of the POUMM based on the trait values and the pathogen phylogeny. This can be done with the R package POUMM \citep{Mitov2017a-POUMM}.
 	\item Generate maximum-likelihood estimates for the pathogen and corresponding non-pathogen effects on the trait using equations \ref{eq:MuG} and \ref{eq:EHat}.
@@
 
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/simulation_results}
-		\caption{Results from the simulation study.
-We simulated host, pathogen, and environmental effects on a trait under the POUMM with different heritability (y-axis) and selection strength (x-axis) parameters.
-For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
-(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
-The RMSE is with reference to the true (simulated) host part of the trait values.
-Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
-(B) shows how GWAS power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
-Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
-The points highlight specific heritability and selection strength values from the \emph{A.
-thaliana}-\emph{X.
-arboricola} QDR analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.
-RMSE = root mean square error, TPR = true positive rate, QDR = quantitative disease resistance, spVL = set-point viral load.}
-		\label{fig:simulationResults}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
-		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves GWAS power (right side) and where it does not (left side).
-These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
-(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability 15 and 75\%, respectively.
-Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
-(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability 15 and 75\%.
-The solid line is the y=x line.
-Selection strength was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
-		\label{fig:simulationTraitExample}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
-	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
-These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
-Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
-Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability and low selection strength/ low stochastic fluctuations (upper left facet).}
-	\label{fig:tip-trait-correlation}
-	\end{center}
-\end{figure}
+% Figure 2-4 location
 
@@
 
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
-	\caption{A high-level schematic of the experimental setup for the two application datasets.
-For (A) HIV-1 spVL in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
-We fit the POUMM to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
-For (B) \emph{A.
-thaliana}-\emph{X.
-arboricola} QDR from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
-We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
-	\label{fig:application-exp-setup}
-	\end{center}
-\end{figure}
+% Figure 5 location
 
@@
 We retained 1,392 individuals of European ancestry for the GWAS.
-In the (i) ``GWAS with standard trait value'' we used the total trait value, our calculated spVL values, as the GWAS response variable.
+In the (i) ``GWAS with standard trait value'' we used the total trait value, calculated spVL values, as the GWAS response variable.
 In the (ii) ``GWAS with estimated non-pathogen part of trait'' we used our estimates for the non-pathogen effects on spVL.
@@
 Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date.
+Effect sizes are smaller with a phylogenetic correction and p-values are slightly increased.
 We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}).
@@
 
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/gwas_results}
-		\caption{Results from comparative GWAS on HIV-1 spVL data.
-(A) shows association p-values for the same host variants from the SHCS cohort in GWAS with two different response variables.
-On the left, we used unmodified (total) spVL values.
-On the right, we used our estimates for the non-pathogen effects on spVL.
-The alternating shades correspond to different chromosomes.
-(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
-Base positions are with reference to genome build GRCh37.
-The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
-Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
-The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
-		\label{fig:gwas-results}
-	\end{center}
-\end{figure}
-
-\begin{table}[H]
-\centering
-\caption{Top association results from \citet{McLaren2015} compared to results from this study.
-Results from this study are for host variants from the SHCS in GWAS with two different response variables.
-``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
-\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
-\hline
-& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
-  \hline
- Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
-  \hline
-    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
-    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:comp-gwas-mclaren-snps}
-\end{table}
+% Figure 6 location
+
+% Table 1 location
 
@@
 Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants.
-Q-Q plots show the distribution of p-values are also very similar (\ref{fig:qq-plots-qdr}).
+Q-Q plots show the distribution of p-values are also very similar (Figure \ref{fig:qq-plots-qdr}).
 In the first, standard GWAS, one \emph{A.
@@
 
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.png}
-	\caption{Results from comparative GWAS on \emph{A.
-thaliana}-\emph{X.
-arboricola} QDR data.
-The two facets show association p-values for the same host \emph{A.
-thaliana} variants in GWAS with two different response variables.
-On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
-On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
-In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
-thaliana} types.
-The alternating shades correspond to different chromosomes.
-The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
-	\label{fig:gwas-results-qdr}
-	\end{center}
-\end{figure}
+% Figure 7 location
 
@@
 In simulations, we showed this is the case when pathogen heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.
-It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
-However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.
 In summary, cohort-level, phylogenetically structured differences in the measured trait value are necessary for our approach to outperform state of the art methods.
@@
 Given our estimates for trait heritability and selection strength on HIV-1 spVL and \emph{A.
-thaliana} quantitative disease resistance to \emph{X.
+thaliana} QDR to \emph{X.
 arboricola}, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure \ref{fig:simulationResults}).
@@
 
-Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS.
+Our method has several limitations.
+When POUMM parameter estimates are highly uncertain, correcting trait values based on posterior mean or maximum likelihood parameter estimates neglects this uncertainty.
+Then, as in the \emph{A.
+thaliana}-\emph{X.
+arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS.
+In this case, one may wish to use a linear mixed model as in \cite{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.
+The expected correlation structure estimated under the POUMM could be used for the covariance of the random effect, taking the phylogeny into account differently but still utilizing information from the evolutionary model.
+Finally, as we show here, our method is not anticipated to be useful in certain evolutionary scenarios.
+For instance, traits like antimicrobial resistance may be under strong selection pressure and be highly heritable.
+In these instances, our simulations do not point to a large improvement when adding our pre-processing step.
+In any case, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect, meaning our approach would not apply.
+In this situation, one would rather account for antimicrobial resistance as a covariate in the GWAS association model.
+
+The primary advantage of our approach is that it is complementary to previously developed methods for infectious disease GWAS.
 First, it provides additional information on the evolutionary dynamics of the trait in the pathogen population.
 Then, it is a convenient pre-processing step for GWAS because it simply produces a corrected response variable for GWAS association tests.
-As in the \emph{A.
-thaliana}-\emph{X.
-arboricola} application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS.
-In this case, one may wish to use a linear mixed model as in \citep{Wang2018Two-wayGenomes}, where the pathogen effect is co-estimated as a random effect.
-In cases where a correction can be estimated and applied, the corrected trait values are envisioned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
+In cases where a correction can be estimated and applied using our method, the corrected trait values are envisioned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK \citep{Chang2015}, though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).
 Further, additional model complexity can be added to the GWAS association tests.
@@
 We further show that the pathogen phylogeny only has an impact on the GWAS outputs if heritability of the trait amongst infection partners is $>$ 25\%.
-This means that for the systems studied here, spVL in individuals living with HIV and QDR for X.
+For the systems studied here, spVL in individuals living with HIV and QDR for X.
 arboricola infections in A.
@@
 $g$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
-For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on fitting the same model to SCHS data (Table \ref{tab:POUMM-parameters-our-estimates}).
+For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on previous estimates of mean spVL (Table \ref{tab:spVLmeasurements}).
 This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}).
@@
 \subsection*{Phylogenetic trait correction}
-We estimated the pathogen and non-pathogen effects on HIV-1 spVL and \emph{X.
-arboricola} mean QDR trait values using the method described in this paper.
+We estimated the pathogen and non-pathogen effects on HIV-1 spVL in humans and \emph{X.
+arboricola} mean QDR in \emph{A.
+thaliana} using the method described in this paper.
 For each  individual, we estimated the pathogen part of the trait value using equation \ref{eq:MuG} and the corresponding non-pathogen part using equation \ref{eq:EHat}.
@@
 
+\newpage
+
+% Main text table
+
+\begin{table}[H]
+\centering
+\caption{Top association results from \citet{McLaren2015} compared to results from this study.
+Results from this study are for host variants from the SHCS in GWAS with two different response variables.
+``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
+\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
+\hline
+& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
+  \hline
+ Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
+  \hline
+    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
+    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
+   \hline
+\end{tabularx}
+\label{tab:comp-gwas-mclaren-snps}
+\end{table}
+
+% Main text figure legends
+
+\begin{figure}[H]
+	\begin{center}
+% 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
+(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
+(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
+Each variant can be present in 0, 1, or 2 copies.
+Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
+(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
+These three effects sum to the trait value for each simulated individual.}
+		\label{fig:spVLModel}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/simulation_results}
+		\caption{Results from the simulation study.
+We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters.
+For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
+(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
+The RMSE is with reference to the true (simulated) host part of the trait values.
+Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
+(B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
+Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
+The points highlight specific heritability and selection strength values from the \emph{A.
+thaliana}-\emph{X.
+arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
+		\label{fig:simulationResults}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
+		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
+These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
+(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively.
+Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
+(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%.
+The solid line is the y=x line.
+Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
+		\label{fig:simulationTraitExample}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    % \includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
+	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
+These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
+Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
+Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
+	\label{fig:tip-trait-correlation}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    % \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
+	\caption{A high-level schematic of the experimental setup for the two application datasets.
+For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
+We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
+For (B) \emph{A.
+thaliana}-\emph{X.
+arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
+We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
+	\label{fig:application-exp-setup}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/gwas_results}
+		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
+(A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables.
+On the left, we used unmodified (total) spVL values.
+On the right, we used our estimates for the non-pathogen effects on spVL.
+The alternating shades correspond to different chromosomes.
+(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
+Base positions are with reference to genome build GRCh37.
+The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
+Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
+The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
+		\label{fig:gwas-results}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    % \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.png}
+	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
+thaliana} quantitative disease resistance (QDR) to \emph{X.
+arboricola}.
+The two facets show association p-values for the same host \emph{A.
+thaliana} variants in GWAS with two different response variables.
+On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
+On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
+In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
+thaliana} types.
+The alternating shades correspond to different chromosomes.
+The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
+	\label{fig:gwas-results-qdr}
+	\end{center}
+\end{figure}
+
 %Additional ideas for supplemental materials:
@@
 
-Here we show the root mean square error (RMSE) of $\bm{z} - \bm{\bar{z}}$ as an estimate for the host part of the trait, denoted $\bm{h}$, should be $\approx$ 0.74 in our simulation scheme.
+Here we show the root mean square error (RMSE) of the scaled trait value for each individual, $z_i - \bar{z}$ as an estimate for the host part of the trait for each individual, ${h_i}$, should be $\approx$ 0.74 in our simulation scheme.
 First we write the expression for the RMSE:
@@
 \begin{equation}
-    RMSE = \sqrt{\frac{\sum_i^N(z - \bar{z} - h)^2}{N}}
+    RMSE = \sqrt{\frac{\sum_i^N(z_i - \bar{z} - h_i)^2}{N}}
 \end{equation}
 
-Note that under our simulation setup, $z - \bar{z}$ differs from $h$ due to the pathogen effect $g$ and the environmental effect $e$.
+Note that under our simulation setup, $z_i - \bar{z}$ differs from $h_i$ due to the individual pathogen effect $g_i$ and environmental effect $e_i$.
 So the term inside the square root equals the combined variance of these two effects:
@@
 
-Therefore, we can expect the RMSE for $\bm{z} - \bm{\bar{z}}$ as an estimate for $\bm{h}$ to be around $\sqrt{0.55} \approx 0.74$.
+Therefore, we can expect the RMSE for $z_i - \bar{z}$ as an estimate for ${h_i}$ to be around $\sqrt{0.55} \approx 0.74$.
 

### 2022-06-23 manuscript/response_to_reviewers.tex
Update on Overleaf.

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{xcolor}

\begin{document}

\newpage
\section*{Editors’ comments}
I have now received two reviews of your paper. Both reviewers were excited about the methodological concept, but both where also concerned about 1) the perceived narrow application of the methodology and 2) the example based on data which are not accessible. The reviewers have suggested ways to perhaps broaden the potential applications of the method and have even suggested additional (alternative) data sets that ARE accessible.  I strongly encourage you to utilize data that others can access to explore your methods. If you want to make insights on HIV evolution using restricted data, then please send such work to a medical journal. But for MBE, the expectation (especially for a methodological paper) is that the demonstrations of methods should be using data that are easily and openly accessible.  Hopefully, you will find these reviews helpful in revising your work.

\textcolor{blue}{We thank the editor for the chance to improve our manuscript and the two reviewers for their constructive feedback. We feel the quality of the manuscript improved thanks to their suggestions. Below, we highlight the main improvements in the revised manuscript. We also discuss two suggestions we did not implement. Finally, in the point-by-point responses, we highlight the specific changes to the manuscript based on each of the reviewers comments.}

\textcolor{blue}{The main improvement in the revised manuscript is that we added an analysis of an open dataset and re-phrased our abstract and introduction to focus on the broader applicability of our method, including what differentiates it from previous methods. We highlight that our two-step method provides additional insights on the evolutionary dynamics of a trait in a pathogen population and that it is compatible with a variety of downstream GWAS association testing methods. We removed the HIV-specific information from the introduction and moved the relevant comparisons to the discussion section.}

\textcolor{blue}{The only suggestions we did not implement were to re-run the HIV GWAS using a linear mixed model and to perform the HIV analysis again using elite controllers. In the first case, we rephrased the discussion to make it clearer that our method is envisioned as a pre-processing step that can be used prior to GWAS association testing under any model. Given the revised focus of the manuscript away from HIV-1 specific insights, we felt improving the HIV-1 GWAS association testing methodology was beyond the scope of the current manuscript. In the second case, our dataset does not contain enough elite controllers to perform the additional analysis.}

\section*{Reviewer 1}

The authors present a new GWAS method that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny. I do not have critiques about the framework, as the method the authors have developed seems novel and in principal, groundbreaking. However, I am finding difficult to grasp the usability of it outside of pathogens that integrates in host genome. What is the broad applicability of the method? Is this method only applicable to stVL in the context of HIV? What else could be investigated? Is it possible to pair transcriptomic data? Can author expand on this?

\vspace{5mm}
\textcolor{blue}{We edited the introduction to make it clearer when this method is applicable: ``The introduced framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.''}

\textcolor{blue}{Then, as mentioned, we included a second application of our method to a phytopathogenic bacteria.}

\textcolor{blue}{We also expanded the discussion section that gives more specific examples of future potential applications, including the possibility of transcriptomic applications, which we hadn't previously thought about but should be possible: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g. Donnenberg et al. (2015); Messina et al. (2016). Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014). Thus, one could also estimate pathogen effects on host gene expression.''}

\begin{enumerate}
    \item It would have been interesting to see how their methods performs with another type of pathogen, a virus that does not integrate, such SARS-CoV-2, or a bacterial one, such TB. \\
    \textcolor{blue}{We added an analysis of \emph{A. thaliana} quantitative disease resistance to the phytopathogenic bacteria \emph{X. arboricola}. The findings are highlighted in the discussion: ``Given our estimates for trait heritability and selection strength on HIV-1 spVL and A. thaliana quantitative disease resistance to X. arboricola, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure 2)... For A. thaliana QDR to X. arboricola, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$). We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types. Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consitent with previous GWAS results from Wang et al. (2018).''}
    \item In the case of co-infection, as sometimes is the case of HIV and TB, how would the method perform? \\
    \textcolor{blue}{We addressed this point in the discussion: ``Further, additional model complexity can be added to the GWAS association tests. For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power. In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.''}
    \item In the case of HIV, is it possible to compare the data with elite controllers? \\
    \textcolor{blue}{Unfortunately our dataset does not include a sufficient number of elite controllers to perform this comparison. The SHCS cohort includes 83 individuals classified as elite controllers, as defined by at least three measurements of plasma virus load (VL) $<$ 2000 RNA copies/ml over at least a 12-month period in the absence of antiviral therapy (Pereya et al, Science, 2010). spVL values range from 1.3 to 3.1 in this group (all very low). After filtering steps, only 8 of these individuals were retained for the GWAS, so the sample size is unfortunately not sufficient to do a comparative GWAS. However, we note that McLaren et al. (2015) found similar signal of association on chromosome 3 and 6 when they filtered to elite controllers in a standard GWAS setup.}
    \item In the case of determining whether viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, I am not sure I understand the premise of this hypothesis. Can authors elaborate on it more in detail? \\
    \textcolor{blue}{We edited the introduction to remove the confounding and extraneous variables terminology and make it clearer why one should consider pathogen effects in a host GWAS: ``Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait. Corrections for these types of population structure in human GWAS cohorts are well- developed and widely accepted (Astle and Balding, 2009; Price et al., 2006). More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017).'' We also make it clearer how our method addresses this problem: ``In a first step, we fit an evolutionary model to trait data and the pathogen phylogeny. This first step provides an estimate of the correlation structure of the trait due to heritable pathogen effects. The estimate is used to remove pathogen effects on the trait. In the second step, the resulting corrected trait data is used in a GWAS with host genetic variants. The GWAS can be performed as normal under the assumption of independent samples.''}
    \item What type of clinical information/trait is included under the environmental effect? How the diversity in environmental effects across the population affects the method? \\
    \textcolor{blue}{We edited the text to give examples of factors that would be grouped into the environmental effect under the POUMM model we use: ``Variation in infectious disease traits like viral load or infection severity can come from several sources. These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature... In the infectious disease GWAS case, we assume the heritable part comprises pathogen genetic factors and all other factors are non-heritable.''}

    \textcolor{blue}{We also added text to the new approaches section to make it clearer that our method is complementary to but does not replace other methods to account for diversity in environmental effects (we only account for heritable pathogen genetic effects): ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value. This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains. We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes.''}
    \item Lines 194-196:  please expand on the parameters of the simulation in greater detail. \\
    \textcolor{blue}{We added a full description of the simulation scheme: ``We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in Hodcroft et al. (2014)... Finally, we sampled an additional random environmental effect for each tip from a normal distribution centered at 0, as illustrated in Figure 1C.''}  
    \item Lines 205-207: it would be best, if there is not a limitation in the number of words, to report here “why” instead of saying that it has been shown in SI. \\
    \textcolor{blue}{We added the suggested explanation: ``As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to pathogen genetic effects and environmental effects is fixed. Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.''}
    \item Liens 212: “our method performs well when an infectious disease trait is highly heritable”, I found the concept of heritability in the infectious disease filed confusing. Are the authors talking about traits in the viral population? Is it possible to expand the sentence with more details for clarity? \\
    \textcolor{blue}{We edited this statement to be clear that we mean ``pathogen heritability''. We also tried to explain the concept of heritability due to the pathogen better in the New Approaches section: ``Variation in infectious disease traits like viral load or infection severity can come from several sources. These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature. GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value. This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains. We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes. Broad-sense pathogen heritability H2 quantifies the fraction of total variance in a trait that is “inherited” from infection partner to infection partner, i.e., due to pathogen factors.''}
\end{enumerate}

\section*{Reviewer 2}

In this paper, Nadeau and colleagues look to correct for pathogen heritability in studies of host association with infectious diseases. They do this by extending an existing method which models trait evolution as a walk along a phylogeny, and provide a new method which which decomposes a pathogen trait into pathogen heritability vs 'other effects' using maximum-likelihood. By using only the component corresponding to 'other effects' as the phenotype in a standard host GWAS, they are able to correct for dependence between pathogen samples due to shared ancestry. As the authors note, this method is rather convenient as it results in an adjusted phenotype which can be used with standard GWAS tools. They apply their method to simulated data, showing a range (high heritability and selection) where a power advantage should be gained. They also apply the method to HIV set point viral load (spVL).

The paper is based on a sound idea, and is a technically competent solution, with no major scientific issues I could identify. Code is provided; data available via managed access; derivations in methods were appropriately explained. The introduction and conclusion are both well-written and do not overstate the results.

The main comment I have is based perhaps more on the impact of the study. The study is very strongly focused on analysis of HIV-1 spVL, which is not really the ideal trait for this method (as seen from the simulations in fig 2; and fig S4 showing the trait changing not all that much). It has long been established that there are two clear host associations, which indeed are exactly what the authors find before and after their adjustment. [If the aim was to try and find more associations for this trait, it would likely be much better to collect lots more samples than change the association methods, as has been found in host GWAS studies of other traits]. To increase the relevance of this study, the authors may consider whether they can apply their approach to other pathogens/traits. At the same time, I also appreciate that host/pathogen data is hard to come by and may have access issues that are too difficult to overcome. However, I do think if any of these studies in other pathogens could also be analysed here, it would really add to the paper. The authors cite a study of HCV as well as some other viral studies as examples, and I would also point to two more in bacteria (https://www.pnas.org/content/115/24/E5440; https://www.nature.com/articles/s41467-019-09976-3), and one in malaria (https://www.biorxiv.org/content/10.1101/2021.03.30.437659v1). Perhaps similar data is now available for SARS-CoV-2 also?

\textcolor{blue}{We were able to access additional data from the \emph{A. thaliana}-\emph{X. arboricola} pathosystem to demonstrate the broader applicability of our method. Unfortunately, neither dataset we analyzed had sufficiently highly correlated trait values amongst related pathogen strains to generate novel biological insights. We edited the text to emphasize that an advantage of our approach is that one gets information on the evolutionary dynamics of the trait in  the pathogen population, even if GWAS results are not changed.}

\vspace{5mm}
I also have some more minor comments:

\begin{enumerate}
    \item How does the method compare to the ATOMM approach in https://www.pnas.org/content/115/24/E5440? Can this be applied to any of the simulated or real data here for comparison. \\
    \textcolor{blue}{We could not apply ATOMM directly to our simulated data because we do not simulate pathogen genotypes, rather we simulate only pathogen trait values. ATOMM is also only implemented for haploid alleles, making it not ideal for the HIV data analyzed. Instead, we were able to access the same dataset ATOMM was tested on in their original publication. This allows us to compare GWAS results from our method versus the ATOMM method, which accounts for the pathogen and host genome jointly in the same linear mixed model and allows tests for marginal effects of host and pathogen alleles as well as pairwise host allele $\times$ pathogen allele interaction effects.}

    \textcolor{blue}{We present the findings of this comparison in the discussion: ``For comparison, we also fit the POUMM to quantitative disease resistance measurements from A. thaliana infected with the phytopathogenic bacteria X. arboricola. We estimated X. arboricola virulence heritability to be 33\% (95\% HPD 0 - 77\%). (Wang et al., 2018) originally estimated a QDR heritability of 44\% in this dataset, falling within the wide range of our estimate. We note that Wang et al. (2018) used a linear mixed model in which the experimental unit is QDR scored on individual leaves, whereas our estimate is based on much coarser binning of QDR scores into a mean score across all leaves on all host accessions and all replicates (N = 22)... For A. thaliana QDR to X. arboricola, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$). We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types. Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consistent with previous GWAS results from Wang et al. (2018).''}

    \textcolor{blue}{We also highlight the linear mixed model approach presented in ATOMM as an alternate means to take pathogen effects into account: ``As in the A. thaliana-X. arboricola application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS. In this case, one may wish to use a linear mixed model as in (Wang et al., 2018), where the pathogen effect is co-estimated as a random effect.''}
    \item As I understand it the POUMM method assumes a 'correct' phylogeny. Do the authors have any thoughts about including uncertainty in the phylogeny (for example by using a posterior sample from e.g. mrbayes) and whether this may affect their approach? Relatedly, does a tree from just pol sequences (L342) give enough resolution? \\
    \textcolor{blue}{This is correct. We added a section ``Phylogenetic uncertainty'' in the Materials and Methods to describe an extra sensitivity test for the HIV analysis: ``Our method assumes the phylogeny accurately reflects the evolutionary relationships between pathogen strains... For the HIV application, we additionally tested the sensitivity of the inference to phylogenetic uncertainty. We inferred the phylogeny again, this time using the IQ-TREE option -wt to output all locally optimal trees. We fit the POUMM to two randomly selected trees from this set and repeated the trait correction and association testing steps using these trees and the corresponding POUMM parameter estimates.'' The results are briefly described in the Results section: ``We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table S4).''}

    \item It would be useful to include more histograms and scatterplots of traits before/after the adjustment, as in figure S4. In particular, doing this plot again for one of the simulations where it does improve power would be informative. In my opinion, these could be a main figure. \\
    \textcolor{blue}{We implemented this suggestion and added two new main text figures (Figures 3 and 4). They are discussed in the section ``Estimator accuracy'': ``Figure 3 gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations. Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
    \item Do the authors know how the computational requirements of their ML estimation method scales with sample size? Would it work for a large SARS-CoV-2 phylogeny for instance? \\
    \textcolor{blue}{The POUMM R package for estimating the POUMM model parameters runs in less than an hour on a tree of 10,000 tips and has been shown to scale to trees of up to 100,000 tips. We additionally verified that the matrix inversion steps required for the last ML estimation step based on the fitted POUMM parameters ran successfully on a simulated tree of 10,000 tips using 4 cores with maximum memory use of 6500MB per core. We now highlight the method's scalability in the discussion: ``Our method relies on the freely available R package POUMM (Mitov and Stadler, 2017), which scales to trees of up to 10,000 tips (Mitov and Stadler, 2019).''}
    \item Fig S3 would be useful, but isn't really readable at present. Perhaps consider an interactive tree viewer to make this more useful. \\
    \textcolor{blue}{We added the same figure for the \emph{A. thaliana}-\emph{X. arboricola} dataset, which is much more legible due to having fewer samples (Figure S8). We also added sub-sampled phylogenies from the simulation study annotated with trait values in Figure 3 to demonstrate what phylogenetically-correlated trait values look like in a successful simulation example. Finally, we now reference the highest correlation values expected between any two tips in the HIV-1 and \emph{X. arboricola} phylogenies to add a quantitative element to the discussion of phylogenetic correlations: ``The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45. However, Figure S5 shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.'' Then for the \emph{A. thaliana}-\emph{X. arboricola} application: ``Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application). Thus, the phylogeny is not very informative for a trait value correction.''}
    \item The authors I think only look at power/TPR. What about the effect on false positive rate? \\
    \textcolor{blue}{We re-ran the simulations to assess the false positive rate and include the results in the Materials and Methods section: ``For our main results (Figure 2) we simulated 20 truly associated variants, as described above. To also check the false positive rate (FPR), we re-ran the simulations with an additional 80 non-associated variants. Across all the association tests in this second simulation setup (7 $H^2$ levels × 10 $\alpha$ levels × 100 variants × 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005 using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value. These rates are comparable to the expected FPR of 0.0005 at significance level 0.05 corrected for 100 tests. Given the stricter correction for multiple testing in this second simulation setup, the TPR decreased significantly across all three GWAS response variables used.''}
    \item I don't think panel A of figure 2 is explained well by the caption or text, I could not understand what the RMSE is showing here. What does 'more accurate' look like in this plot? \\
    \textcolor{blue}{We edited the figure legend to make it clearer: ``(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario. The RMSE is with reference to the true (simulated) host part of the trait values. Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.''}
    \item Is there any explanation why the simulated (actual) and estimated non-pathogen trait in fig 2B start to look quite different at stronger selection strengths. \\
    \textcolor{blue}{We added a figure (Figure 4) to illustrate how the correlation between clustered tips on the phylogenetic tree decreases with increasing selection strength/ increasing stochasticity. Without phylogenetic signal, our method has no information to estimate the pathogen effect on the trait. This is described in the section ``Estimator accuracy'': ``Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
    \item Using plink and removing samples with divergent ancestry is probably a little outdated for host GWAS. A linear mixed model would likely provide more power still. In particular, I'd suggest warped-lmm (http://dx.doi.org/10.1038/ncomms5890), which also transforms the phenotype to increase power. Furthermore, I would be interested to see histograms/scatter plots of a few phenotypes: a) no adjustment; b) warped-lmm only; c) the author's method; d) both adjustments. Perhaps the methods are similar, or using both does increases power beyond either method alone. \\
    \textcolor{blue}{We highlight that our method is a pre-processing step before GWAS and can be combined with many different GWAS models implemented in softwares like PLINK, GEMMA, and warped-lmm. Our revised manuscript focuses more on the method rather than generating biological insights for HIV or \emph{X. arboricola}, so we feel that updating our example GWAS is beyond the scope of the current study. We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS.... In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK (Chang et al., 2015), though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
\end{enumerate}

\vspace{5mm}
Suggested text changes:

\begin{enumerate}
    \item The first sentence of the abstract is probably also true for many common diseases analysed in human GWAS (apart from perhaps the pathogen, but gut microbiome may be a factor too). \\
    \textcolor{blue}{We changed the wording to make it clearer what sets infectious disease GWAS apart: ``Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait.''}
    \item L84-86. There are more than two methods for this, see also https://doi.org/10.1371/journal.pgen.1007758; https://doi.org/10.1093/bioinformatics/bty539; https://doi.org/10.1099/mgen.0.000469. Maybe remove 'two', as this will change anyway.\\
    \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017). Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in (Earle et al., 2016) and modified association tests that account for phylogenetic relationships amongst samples as in (Collins and Didelot, 2018).''}
    \item L110 'Of these, only pathogen effects are heritable from one transmission partner to another'. This might be nit-picking but I could see this phrasing being misunderstood by some readers. Environment will be partially shared along a transmission chain, probably. Interaction effects could also be shared if e.g. hosts are likely to have the same ancestry.\\
    \textcolor{blue}{We improved the phrasing to be more accurate: ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value. This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.''}
    \item Equation 1 onwards. Using 'e' for the environmental effects is easy to get confused with more standard use of the number e = 2.718, especially when raised to a power. Consider changing this symbol.\\
    \textcolor{blue}{We changed the variable to $\epsilon$ and also changed the variable for the pathogen part of the trait to $g$ to match previous work on the POUMM (e.g. \citep{Mitov2017a-POUMM})}.
    \item Please add the link to the github code in the framework section at the end of the methods.\\
    \textcolor{blue}{We added the link in the section ``Phylogenetic trait correction'': ``All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.''}
    \item L283 'In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.' Could the authors give a biological example of this? Would antimicrobial resistance variation fit the bill?\\
    \textcolor{blue}{We have not fit the POUMM to data from enough systems to have a clear example of this, so instead we suggest other systems that might fall into this case: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g. Donnenberg et al. (2015); Messina et al. (2016). Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014). Thus, one could also estimate pathogen effects on host gene expression.''}

    \textcolor{blue}{We anticipate antimicrobial resistance would be under strong selective pressure, but agree this could produce phylogenetically correlated trait values. We added a statement in the discussion: ``It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance. However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.''}
\end{enumerate}
\end{document}

### 2022-06-23 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 	\begin{center}
-% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values}
+% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
@@
 \begin{center}
-    % \includegraphics[width = \linewidth]{figures/tip_trait_correlation.png}
+    % \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
 	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
@@
 \begin{center}
-    % \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.png}
+    % \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
 	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
@@
 \stopsection
-\newpage
-\section*{Editors’ comments}
-I have now received two reviews of your paper.
-Both reviewers were excited about the methodological concept, but both where also concerned about 1) the perceived narrow application of the methodology and 2) the example based on data which are not accessible.
-The reviewers have suggested ways to perhaps broaden the potential applications of the method and have even suggested additional (alternative) data sets that ARE accessible.
-I strongly encourage you to utilize data that others can access to explore your methods.
-If you want to make insights on HIV evolution using restricted data, then please send such work to a medical journal.
-But for MBE, the expectation (especially for a methodological paper) is that the demonstrations of methods should be using data that are easily and openly accessible.
-Hopefully, you will find these reviews helpful in revising your work.
-
-\textcolor{blue}{We thank the editor for the chance to improve our manuscript and the two reviewers for their constructive feedback.
-We feel the quality of the manuscript improved thanks to their suggestions.
-Below, we highlight the main improvements in the revised manuscript.
-We also discuss two suggestions we did not implement.
-Finally, in the point-by-point responses, we highlight the specific changes to the manuscript based on each of the reviewers comments.}
-
-\textcolor{blue}{The main improvement in the revised manuscript is that we added an analysis of an open dataset and re-phrased our abstract and introduction to focus on the broader applicability of our method, including what differentiates it from previous methods.
-We highlight that our two-step method provides additional insights on the evolutionary dynamics of a trait in a pathogen population and that it is compatible with a variety of downstream GWAS association testing methods.
-We removed the HIV-specific information from the introduction and moved the relevant comparisons to the discussion section.}
-
-\textcolor{blue}{The only suggestions we did not implement were to re-run the HIV GWAS using a linear mixed model and to perform the HIV analysis again using elite controllers.
-In the first case, we rephrased the discussion to make it clearer that our method is envisioned as a pre-processing step that can be used prior to GWAS association testing under any model.
-Given the revised focus of the manuscript away from HIV-1 specific insights, we felt improving the HIV-1 GWAS association testing methodology was beyond the scope of the current manuscript.
-In the second case, our dataset does not contain enough elite controllers to perform the additional analysis.}
-
-\section*{Reviewer 1}
-
-The authors present a new GWAS method that estimates and removes trait variability due to the pathogen using information from the full pathogen phylogeny.
-I do not have critiques about the framework, as the method the authors have developed seems novel and in principal, groundbreaking.
-However, I am finding difficult to grasp the usability of it outside of pathogens that integrates in host genome.
-What is the broad applicability of the method?
-Is this method only applicable to stVL in the context of HIV?
-What else could be investigated?
-Is it possible to pair transcriptomic data?
-Can author expand on this?
-
-\vspace{5mm}
-\textcolor{blue}{We edited the introduction to make it clearer when this method is applicable: ``The introduced framework relies on paired pathogen-host genotyping and is envisioned specifically for continuous-valued traits that are highly heritable from infection partner to infection partner.''}
-
-\textcolor{blue}{Then, as mentioned, we included a second application of our method to a phytopathogenic bacteria.}
-
-\textcolor{blue}{We also expanded the discussion section that gives more specific examples of future potential applications, including the possibility of transcriptomic applications, which we hadn't previously thought about but should be possible: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g.
-Donnenberg et al.
-(2015); Messina et al.
-(2016).
-Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014).
-Thus, one could also estimate pathogen effects on host gene expression.''}
-
-\begin{enumerate}
-    \item It would have been interesting to see how their methods performs with another type of pathogen, a virus that does not integrate, such SARS-CoV-2, or a bacterial one, such TB. \\
-    \textcolor{blue}{We added an analysis of \emph{A.
-thaliana} quantitative disease resistance to the phytopathogenic bacteria \emph{X.
-arboricola}.
-The findings are highlighted in the discussion: ``Given our estimates for trait heritability and selection strength on HIV-1 spVL and A.
-thaliana quantitative disease resistance to X.
-arboricola, our simulation results reveal that we cannot expect a significant improvement in GWAS power for these systems (Figure 2)...
-For A.
-thaliana QDR to X.
-arboricola, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
-We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types.
-Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consitent with previous GWAS results from Wang et al.
-(2018).''}
-    \item In the case of co-infection, as sometimes is the case of HIV and TB, how would the method perform? \\
-    \textcolor{blue}{We addressed this point in the discussion: ``Further, additional model complexity can be added to the GWAS association tests.
-For instance, our method does not account for co-infection, which might add additional variance to trait values and decrease GWAS power.
-In this case, one could add co-infection status as a covariate in the GWAS association test to account for this variable.''}
-    \item In the case of HIV, is it possible to compare the data with elite controllers? \\
-    \textcolor{blue}{Unfortunately our dataset does not include a sufficient number of elite controllers to perform this comparison.
-The SHCS cohort includes 83 individuals classified as elite controllers, as defined by at least three measurements of plasma virus load (VL) $<$ 2000 RNA copies/ml over at least a 12-month period in the absence of antiviral therapy (Pereya et al, Science, 2010).
-spVL values range from 1.3 to 3.1 in this group (all very low).
-After filtering steps, only 8 of these individuals were retained for the GWAS, so the sample size is unfortunately not sufficient to do a comparative GWAS.
-However, we note that McLaren et al.
-(2015) found similar signal of association on chromosome 3 and 6 when they filtered to elite controllers in a standard GWAS setup.}
-    \item In the case of determining whether viral effects may be confounding or extraneous variables that bias estimates of host genetic effects, I am not sure I understand the premise of this hypothesis.
-Can authors elaborate on it more in detail?
-\\
-    \textcolor{blue}{We edited the introduction to remove the confounding and extraneous variables terminology and make it clearer why one should consider pathogen effects in a host GWAS: ``Shared ancestry among individuals, especially between close relatives, can give rise to spurious genetic correlations with a trait.
-Corrections for these types of population structure in human GWAS cohorts are well- developed and widely accepted (Astle and Balding, 2009; Price et al., 2006).
-More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017).'' We also make it clearer how our method addresses this problem: ``In a first step, we fit an evolutionary model to trait data and the pathogen phylogeny.
-This first step provides an estimate of the correlation structure of the trait due to heritable pathogen effects.
-The estimate is used to remove pathogen effects on the trait.
-In the second step, the resulting corrected trait data is used in a GWAS with host genetic variants.
-The GWAS can be performed as normal under the assumption of independent samples.''}
-    \item What type of clinical information/trait is included under the environmental effect? How the diversity in environmental effects across the population affects the method? \\
-    \textcolor{blue}{We edited the text to give examples of factors that would be grouped into the environmental effect under the POUMM model we use: ``Variation in infectious disease traits like viral load or infection severity can come from several sources.
-These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature...
-In the infectious disease GWAS case, we assume the heritable part comprises pathogen genetic factors and all other factors are non-heritable.''}
-
-    \textcolor{blue}{We also added text to the new approaches section to make it clearer that our method is complementary to but does not replace other methods to account for diversity in environmental effects (we only account for heritable pathogen genetic effects): ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
-This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
-We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes.''}
-    \item Lines 194-196:  please expand on the parameters of the simulation in greater detail. \\
-    \textcolor{blue}{We added a full description of the simulation scheme: ``We first simulated a phylogeny of 500 tips with exponentially distributed branch lengths and mean root-to-tip time of 0.14 substitutions per site per year as in Hodcroft et al.
-(2014)...
-Finally, we sampled an additional random environmental effect for each tip from a normal distribution centered at 0, as illustrated in Figure 1C.''}
-    \item Lines 205-207: it would be best, if there is not a limitation in the number of words, to report here “why” instead of saying that it has been shown in SI. \\
-    \textcolor{blue}{We added the suggested explanation: ``As shown in the supplemental material, we can calculate the expected RMSE using the scaled trait value across scenarios in our simulation scheme because the variance in the trait due to pathogen genetic effects and environmental effects is fixed.
-Thus, we expect the RMSE using the scaled trait value to be 0.74 across all simulation scenarios.''}
-    \item Liens 212: “our method performs well when an infectious disease trait is highly heritable”, I found the concept of heritability in the infectious disease filed confusing.
-Are the authors talking about traits in the viral population?
-Is it possible to expand the sentence with more details for clarity?
-\\
-    \textcolor{blue}{We edited this statement to be clear that we mean ``pathogen heritability''.
-We also tried to explain the concept of heritability due to the pathogen better in the New Approaches section: ``Variation in infectious disease traits like viral load or infection severity can come from several sources.
-These include host genetic factors, pathogen genetic factors, interaction effects between the host and the pathogen, or non-genetic factors like healthcare quality or temperature.
-GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
-This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.
-We aim to remove this pathogen-induced correlation in the trait data prior to performing GWAS on the host genomes.
-Broad-sense pathogen heritability H2 quantifies the fraction of total variance in a trait that is “inherited” from infection partner to infection partner, i.e., due to pathogen factors.''}
-\end{enumerate}
-
-\section*{Reviewer 2}
-
-In this paper, Nadeau and colleagues look to correct for pathogen heritability in studies of host association with infectious diseases.
-They do this by extending an existing method which models trait evolution as a walk along a phylogeny, and provide a new method which which decomposes a pathogen trait into pathogen heritability vs 'other effects' using maximum-likelihood.
-By using only the component corresponding to 'other effects' as the phenotype in a standard host GWAS, they are able to correct for dependence between pathogen samples due to shared ancestry.
-As the authors note, this method is rather convenient as it results in an adjusted phenotype which can be used with standard GWAS tools.
-They apply their method to simulated data, showing a range (high heritability and selection) where a power advantage should be gained.
-They also apply the method to HIV set point viral load (spVL).
-
-The paper is based on a sound idea, and is a technically competent solution, with no major scientific issues I could identify.
-Code is provided; data available via managed access; derivations in methods were appropriately explained.
-The introduction and conclusion are both well-written and do not overstate the results.
-
-The main comment I have is based perhaps more on the impact of the study.
-The study is very strongly focused on analysis of HIV-1 spVL, which is not really the ideal trait for this method (as seen from the simulations in fig 2; and fig S4 showing the trait changing not all that much).
-It has long been established that there are two clear host associations, which indeed are exactly what the authors find before and after their adjustment.
-[If the aim was to try and find more associations for this trait, it would likely be much better to collect lots more samples than change the association methods, as has been found in host GWAS studies of other traits].
-To increase the relevance of this study, the authors may consider whether they can apply their approach to other pathogens/traits.
-At the same time, I also appreciate that host/pathogen data is hard to come by and may have access issues that are too difficult to overcome.
-However, I do think if any of these studies in other pathogens could also be analysed here, it would really add to the paper.
-The authors cite a study of HCV as well as some other viral studies as examples, and I would also point to two more in bacteria (https://www.pnas.org/content/115/24/E5440; https://www.nature.com/articles/s41467-019-09976-3), and one in malaria (https://www.biorxiv.org/content/10.1101/2021.03.30.437659v1).
-Perhaps similar data is now available for SARS-CoV-2 also?
-
-\textcolor{blue}{We were able to access additional data from the \emph{A.
-thaliana}-\emph{X.
-arboricola} pathosystem to demonstrate the broader applicability of our method.
-Unfortunately, neither dataset we analyzed had sufficiently highly correlated trait values amongst related pathogen strains to generate novel biological insights.
-We edited the text to emphasize that an advantage of our approach is that one gets information on the evolutionary dynamics of the trait in  the pathogen population, even if GWAS results are not changed.}
-
-\vspace{5mm}
-I also have some more minor comments:
-
-\begin{enumerate}
-    \item How does the method compare to the ATOMM approach in https://www.pnas.org/content/115/24/E5440? Can this be applied to any of the simulated or real data here for comparison. \\
-    \textcolor{blue}{We could not apply ATOMM directly to our simulated data because we do not simulate pathogen genotypes, rather we simulate only pathogen trait values.
-ATOMM is also only implemented for haploid alleles, making it not ideal for the HIV data analyzed.
-Instead, we were able to access the same dataset ATOMM was tested on in their original publication.
-This allows us to compare GWAS results from our method versus the ATOMM method, which accounts for the pathogen and host genome jointly in the same linear mixed model and allows tests for marginal effects of host and pathogen alleles as well as pairwise host allele $\times$ pathogen allele interaction effects.}
-
-    \textcolor{blue}{We present the findings of this comparison in the discussion: ``For comparison, we also fit the POUMM to quantitative disease resistance measurements from A.
-thaliana infected with the phytopathogenic bacteria X.
-arboricola.
-We estimated X.
-arboricola virulence heritability to be 33\% (95\% HPD 0 - 77\%).
-(Wang et al., 2018) originally estimated a QDR heritability of 44\% in this dataset, falling within the wide range of our estimate.
-We note that Wang et al.
-(2018) used a linear mixed model in which the experimental unit is QDR scored on individual leaves, whereas our estimate is based on much coarser binning of QDR scores into a mean score across all leaves on all host accessions and all replicates (N = 22)...
-For A.
-thaliana QDR to X.
-arboricola, the trait value correction does not utilize phylogenetic information because phylogenetic correlations between samples are too weak (maximum expected correlation between strains was $3.2 \times 10^{-12}$).
-We anyways corrected QDR trait values based on average QDR for each pathogen strain across the full range of host types.
-Results show slight decrease in p-values for the most-associated variants in this application as well, but the overall picture is consistent with previous GWAS results from Wang et al.
-(2018).''}
-
-    \textcolor{blue}{We also highlight the linear mixed model approach presented in ATOMM as an alternate means to take pathogen effects into account: ``As in the A.
-thaliana-X.
-arboricola application, fitting the POUMM may reveal that expected phylogenetic correlations between samples are not strong enough to justify using our method to correct trait values in a GWAS.
-In this case, one may wish to use a linear mixed model as in (Wang et al., 2018), where the pathogen effect is co-estimated as a random effect.''}
-    \item As I understand it the POUMM method assumes a 'correct' phylogeny.
-Do the authors have any thoughts about including uncertainty in the phylogeny (for example by using a posterior sample from e.g.
-mrbayes) and whether this may affect their approach?
-Relatedly, does a tree from just pol sequences (L342) give enough resolution?
-\\
-    \textcolor{blue}{This is correct.
-We added a section ``Phylogenetic uncertainty'' in the Materials and Methods to describe an extra sensitivity test for the HIV analysis: ``Our method assumes the phylogeny accurately reflects the evolutionary relationships between pathogen strains...
-For the HIV application, we additionally tested the sensitivity of the inference to phylogenetic uncertainty.
-We inferred the phylogeny again, this time using the IQ-TREE option -wt to output all locally optimal trees.
-We fit the POUMM to two randomly selected trees from this set and repeated the trait correction and association testing steps using these trees and the corresponding POUMM parameter estimates.'' The results are briefly described in the Results section: ``We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table S4).''}
-
-    \item It would be useful to include more histograms and scatterplots of traits before/after the adjustment, as in figure S4.
-In particular, doing this plot again for one of the simulations where it does improve power would be informative.
-In my opinion, these could be a main figure.
-\\
-    \textcolor{blue}{We implemented this suggestion and added two new main text figures (Figures 3 and 4).
-They are discussed in the section ``Estimator accuracy'': ``Figure 3 gives some intuition for how this correction works by contrasting simulated scenarios with high and low heritability and low selection strength/ low stochastic fluctuations.
-Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
-    \item Do the authors know how the computational requirements of their ML estimation method scales with sample size? Would it work for a large SARS-CoV-2 phylogeny for instance? \\
-    \textcolor{blue}{The POUMM R package for estimating the POUMM model parameters runs in less than an hour on a tree of 10,000 tips and has been shown to scale to trees of up to 100,000 tips.
-We additionally verified that the matrix inversion steps required for the last ML estimation step based on the fitted POUMM parameters ran successfully on a simulated tree of 10,000 tips using 4 cores with maximum memory use of 6500MB per core.
-We now highlight the method's scalability in the discussion: ``Our method relies on the freely available R package POUMM (Mitov and Stadler, 2017), which scales to trees of up to 10,000 tips (Mitov and Stadler, 2019).''}
-    \item Fig S3 would be useful, but isn't really readable at present. Perhaps consider an interactive tree viewer to make this more useful. \\
-    \textcolor{blue}{We added the same figure for the \emph{A.
-thaliana}-\emph{X.
-arboricola} dataset, which is much more legible due to having fewer samples (Figure S8).
-We also added sub-sampled phylogenies from the simulation study annotated with trait values in Figure 3 to demonstrate what phylogenetically-correlated trait values look like in a successful simulation example.
-Finally, we now reference the highest correlation values expected between any two tips in the HIV-1 and \emph{X.
-arboricola} phylogenies to add a quantitative element to the discussion of phylogenetic correlations: ``The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
-However, Figure S5 shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.'' Then for the \emph{A.
-thaliana}-\emph{X.
-arboricola} application: ``Given the posterior mean estimates for the POUMM parameters, expected correlation in trait values between tips were very low (maximum value $3.2 \times 10^{-12}$ compared to maximum value of 0.45 in the HIV-1 spVL application).
-Thus, the phylogeny is not very informative for a trait value correction.''}
-    \item The authors I think only look at power/TPR. What about the effect on false positive rate? \\
-    \textcolor{blue}{We re-ran the simulations to assess the false positive rate and include the results in the Materials and Methods section: ``For our main results (Figure 2) we simulated 20 truly associated variants, as described above.
-To also check the false positive rate (FPR), we re-ran the simulations with an additional 80 non-associated variants.
-Across all the association tests in this second simulation setup (7 $H^2$ levels × 10 $\alpha$ levels × 100 variants × 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005 using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value.
-These rates are comparable to the expected FPR of 0.0005 at significance level 0.05 corrected for 100 tests.
-Given the stricter correction for multiple testing in this second simulation setup, the TPR decreased significantly across all three GWAS response variables used.''}
-    \item I don't think panel A of figure 2 is explained well by the caption or text, I could not understand what the RMSE is showing here. What does 'more accurate' look like in this plot? \\
-    \textcolor{blue}{We edited the figure legend to make it clearer: ``(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
-The RMSE is with reference to the true (simulated) host part of the trait values.
-Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.''}
-    \item Is there any explanation why the simulated (actual) and estimated non-pathogen trait in fig 2B start to look quite different at stronger selection strengths. \\
-    \textcolor{blue}{We added a figure (Figure 4) to illustrate how the correlation between clustered tips on the phylogenetic tree decreases with increasing selection strength/ increasing stochasticity.
-Without phylogenetic signal, our method has no information to estimate the pathogen effect on the trait.
-This is described in the section ``Estimator accuracy'': ``Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
-    \item Using plink and removing samples with divergent ancestry is probably a little outdated for host GWAS.
-A linear mixed model would likely provide more power still.
-In particular, I'd suggest warped-lmm (http://dx.doi.org/10.1038/ncomms5890), which also transforms the phenotype to increase power.
-Furthermore, I would be interested to see histograms/scatter plots of a few phenotypes: a) no adjustment; b) warped-lmm only; c) the author's method; d) both adjustments.
-Perhaps the methods are similar, or using both does increases power beyond either method alone.
-\\
-    \textcolor{blue}{We highlight that our method is a pre-processing step before GWAS and can be combined with many different GWAS models implemented in softwares like PLINK, GEMMA, and warped-lmm.
-Our revised manuscript focuses more on the method rather than generating biological insights for HIV or \emph{X.
-arboricola}, so we feel that updating our example GWAS is beyond the scope of the current study.
-We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS....
-In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK (Chang et al., 2015), though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
-\end{enumerate}
-
-\vspace{5mm}
-Suggested text changes:
-
-\begin{enumerate}
-    \item The first sentence of the abstract is probably also true for many common diseases analysed in human GWAS (apart from perhaps the pathogen, but gut microbiome may be a factor too). \\
-    \textcolor{blue}{We changed the wording to make it clearer what sets infectious disease GWAS apart: ``Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait.''}
-    \item L84-86.
-There are more than two methods for this, see also https://doi.org/10.1371/journal.pgen.1007758; https://doi.org/10.1093/bioinformatics/bty539; https://doi.org/10.1099/mgen.0.000469.
-Maybe remove 'two', as this will change anyway.\\
-    \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017).
-Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in (Earle et al., 2016) and modified association tests that account for phylogenetic relationships amongst samples as in (Collins and Didelot, 2018).''}
-    \item L110 'Of these, only pathogen effects are heritable from one transmission partner to another'.
-This might be nit-picking but I could see this phrasing being misunderstood by some readers.
-Environment will be partially shared along a transmission chain, probably.
-Interaction effects could also be shared if e.g.
-hosts are likely to have the same ancestry.\\
-    \textcolor{blue}{We improved the phrasing to be more accurate: ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value.
-This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.''}
-    \item Equation 1 onwards.
-Using 'e' for the environmental effects is easy to get confused with more standard use of the number e = 2.718, especially when raised to a power.
-Consider changing this symbol.\\
-    \textcolor{blue}{We changed the variable to $\epsilon$ and also changed the variable for the pathogen part of the trait to $g$ to match previous work on the POUMM (e.g. \citep{Mitov2017a-POUMM})}.
-    \item Please add the link to the github code in the framework section at the end of the methods.\\
-    \textcolor{blue}{We added the link in the section ``Phylogenetic trait correction'': ``All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.''}
-    \item L283 'In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.' Could the authors give a biological example of this?
-Would antimicrobial resistance variation fit the bill?\\
-    \textcolor{blue}{We have not fit the POUMM to data from enough systems to have a clear example of this, so instead we suggest other systems that might fall into this case: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g.
-Donnenberg et al.
-(2015); Messina et al.
-(2016).
-Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014).
-Thus, one could also estimate pathogen effects on host gene expression.''}
-
-    \textcolor{blue}{We anticipate antimicrobial resistance would be under strong selective pressure, but agree this could produce phylogenetically correlated trait values.
-We added a statement in the discussion: ``It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
-However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.''}
-\end{enumerate}
 \end{document}

### 2022-06-23 manuscript/hiv-gwas.tex
Update on Overleaf.

@@
 
-\begin{flushright} % for MBE submission
-Submission type: Article\\
-Section: Methods
-\end{flushright}
+% \begin{flushright} % for MBE submission
+% Submission type: Article\\
+% Section: Methods
+% \end{flushright}
 
@@
 
-% Figure 1 location
+\begin{figure}[H]
+	\begin{center}
+	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
+(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
+(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
+Each variant can be present in 0, 1, or 2 copies.
+Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
+(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
+These three effects sum to the trait value for each simulated individual.}
+		\label{fig:spVLModel}
+	\end{center}
+\end{figure}
 
@@
 
-% Figure 2-4 location
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/simulation_results}
+		\caption{Results from the simulation study.
+We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters.
+For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
+(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
+The RMSE is with reference to the true (simulated) host part of the trait values.
+Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
+(B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
+Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
+The points highlight specific heritability and selection strength values from the \emph{A.
+thaliana}-\emph{X.
+arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
+		\label{fig:simulationResults}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
+		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
+These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
+(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively.
+Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
+(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%.
+The solid line is the y=x line.
+Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
+		\label{fig:simulationTraitExample}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
+	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
+These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
+Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
+Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
+	\label{fig:tip-trait-correlation}
+	\end{center}
+\end{figure}
 
@@
 
-% Figure 5 location
+\begin{figure}[H]
+\begin{center}
+    \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
+	\caption{A high-level schematic of the experimental setup for the two application datasets.
+For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
+We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
+For (B) \emph{A.
+thaliana}-\emph{X.
+arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
+We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
+	\label{fig:application-exp-setup}
+	\end{center}
+\end{figure}
 
@@
 
-% Figure 6 location
-
-% Table 1 location
+\begin{figure}[H]
+	\begin{center}
+		\includegraphics[width=\linewidth]{figures/gwas_results}
+		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
+(A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables.
+On the left, we used unmodified (total) spVL values.
+On the right, we used our estimates for the non-pathogen effects on spVL.
+The alternating shades correspond to different chromosomes.
+(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
+Base positions are with reference to genome build GRCh37.
+The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
+Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
+The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
+		\label{fig:gwas-results}
+	\end{center}
+\end{figure}
+
+\begin{table}[H]
+\centering
+\caption{Top association results from \citet{McLaren2015} compared to results from this study.
+Results from this study are for host variants from the SHCS in GWAS with two different response variables.
+``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
+\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
+\hline
+& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
+  \hline
+ Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
+  \hline
+    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
+    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
+   \hline
+\end{tabularx}
+\label{tab:comp-gwas-mclaren-snps}
+\end{table}
 
@@
 
-% Figure 7 location
+\begin{figure}[H]
+\begin{center}
+    \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
+	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
+thaliana} quantitative disease resistance (QDR) to \emph{X.
+arboricola}.
+The two facets show association p-values for the same host \emph{A.
+thaliana} variants in GWAS with two different response variables.
+On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
+On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
+In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
+thaliana} types.
+The alternating shades correspond to different chromosomes.
+The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
+	\label{fig:gwas-results-qdr}
+	\end{center}
+\end{figure}
 
@@
 \newpage
-
-% Main text table
-
-\begin{table}[H]
-\centering
-\caption{Top association results from \citet{McLaren2015} compared to results from this study.
-Results from this study are for host variants from the SHCS in GWAS with two different response variables.
-``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
-\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
-\hline
-& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
-  \hline
- Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
-  \hline
-    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
-    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:comp-gwas-mclaren-snps}
-\end{table}
-
-% Main text figure legends
-
-\begin{figure}[H]
-	\begin{center}
-% 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
-	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
-(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
-(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
-Each variant can be present in 0, 1, or 2 copies.
-Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
-(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
-These three effects sum to the trait value for each simulated individual.}
-		\label{fig:spVLModel}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-% 		\includegraphics[width=\linewidth]{figures/simulation_results}
-		\caption{Results from the simulation study.
-We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters.
-For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
-(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
-The RMSE is with reference to the true (simulated) host part of the trait values.
-Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
-(B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
-Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
-The points highlight specific heritability and selection strength values from the \emph{A.
-thaliana}-\emph{X.
-arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
-		\label{fig:simulationResults}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
-		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
-These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
-(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively.
-Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
-(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%.
-The solid line is the y=x line.
-Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
-		\label{fig:simulationTraitExample}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-    % \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
-	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
-These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
-Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
-Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
-	\label{fig:tip-trait-correlation}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-    % \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
-	\caption{A high-level schematic of the experimental setup for the two application datasets.
-For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
-We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
-For (B) \emph{A.
-thaliana}-\emph{X.
-arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
-We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
-	\label{fig:application-exp-setup}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-% 		\includegraphics[width=\linewidth]{figures/gwas_results}
-		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
-(A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables.
-On the left, we used unmodified (total) spVL values.
-On the right, we used our estimates for the non-pathogen effects on spVL.
-The alternating shades correspond to different chromosomes.
-(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
-Base positions are with reference to genome build GRCh37.
-The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
-Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
-The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
-		\label{fig:gwas-results}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-    % \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
-	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
-thaliana} quantitative disease resistance (QDR) to \emph{X.
-arboricola}.
-The two facets show association p-values for the same host \emph{A.
-thaliana} variants in GWAS with two different response variables.
-On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
-On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
-In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
-thaliana} types.
-The alternating shades correspond to different chromosomes.
-The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
-	\label{fig:gwas-results-qdr}
-	\end{center}
-\end{figure}
 

### 2022-06-23 scripts/R/make_manuscript_figures.rmd
add pdfs for all but gwas figures

@@
 top_row <- plot_grid(NULL, rmse_plot + guides(shape = "none"), rel_widths = c(1.03, 3), labels = c("", "A"))
-png(filename = "figures/simulation_results.png", width = linewidth, height = height, unit = "cm", res = 300)
+
 plot_grid(
@@
 )
-dev.off()
+ggsave(filename = "figures/simulation_results.pdf", width = linewidth, height = height, unit = "cm")
 ```
@@
         shared_theme
-ggsave(filename = "figures/tip_trait_correlation.png", width = linewidth, height = height * 2/3, unit = "cm")
+ggsave(filename = "figures/tip_trait_correlation.pdf", width = linewidth, height = height * 2/3, unit = "cm")
 
@@
           labels = c("", "", "C"))
-ggsave(filename = paste(FIGDIR, "example_simulation_trait_values.png", sep = "/"),
+ggsave(filename = paste(FIGDIR, "example_simulation_trait_values.pdf", sep = "/"),
       width = linewidth, height = height, unit = "cm")
@@
 
-png(
-        filename = paste(FIGDIR, "gwas_results.png", sep = "/"),
-        width = linewidth, height = height, unit = "cm", res = 300)
 plot_grid(aligned_plots[[1]], aligned_plots[[2]], nrow = 2, labels = c("A", "B"))
-dev.off()
+ggsave(
+        filename = paste(FIGDIR, "gwas_results.pdf", sep = "/"),
+        width = linewidth, height = height, unit = "cm"
+)
 ```
@@
 ```
+Make A. thaliana GWAS figures (here or use scripts/R/plot_manhattan_euler.R)
+```{R}
+# Load data
+print("Loading data")
+gwas_results_xanthomonas <- read.delim(
+    file = paste(OUTDIR_XANTHAMONAS, "gwas_results.maxmaf.pvals.txt", sep = "/"), 
+    sep = "", 
+    header = T
+)
+
+pt_size <- 1
+
+# Add genome position to data
+chr_lengths <- gwas_results_xanthomonas %>%
+        group_by(CHROM) %>%
+        summarise(chr_len = max(POS))
+chr_starts <- chr_lengths %>%
+        mutate(CHR_start = cumsum(as.numeric(chr_len)) - chr_len) %>%
+        select(-chr_len)
+gwas_results_annotated <- left_join(gwas_results_xanthomonas, chr_starts, by = "CHROM")
+gwas_results_annotated <- gwas_results_annotated %>%
+        arrange(CHROM, POS) %>%
+        mutate(BP_position = CHR_start + POS)
+
+# This is maybe not the best way, but makes sure tick marks and labels are at the center of each chromosome
+x_axis_ticks <- chr_starts
+max_BP_position <- max(gwas_results_annotated[gwas_results_annotated$CHROM == 5, "BP_position"])
+x_axis_ticks <- rbind(
+        x_axis_ticks,
+        data.frame(
+                CHROM = c(6),
+                CHR_start = c(max_BP_position)))
+x_axis_ticks$midpoint <- x_axis_ticks$CHR_start + (lead(x_axis_ticks$CHR_start) - x_axis_ticks$CHR_start)/2
+
+# Format data for plotting
+results_long <- tidyr::pivot_longer(
+        data = gwas_results_annotated,
+        cols = c("P_standard", "P_corrected"),
+        names_to = "gwas_type",
+        names_prefix = "P_",
+        values_to = "P")
+
+# Fix facet labels for plotting
+results_long$gwas_type <- factor(
+        results_long$gwas_type,
+        levels = c("standard", "corrected"),
+        labels = c(standard_gwas_name, corrected_gwas_name))
+
+# Make comparative GWAS overview plot
+print("Making plot")
+bonferroni_signifiance_threshold <- 0.05 / nrow(gwas_results_annotated)
+overview_comparison <- ggplot(
+        data = results_long,
+        aes(x = BP_position, y = -log10(P))) +
+        geom_point(aes(color = CHROM %% 2 == 0), size = pt_size) +
+        scale_color_manual(values = c("grey", "dark grey")) +
+        scale_x_continuous(
+                breaks = x_axis_ticks$midpoint,
+                labels = x_axis_ticks$CHROM,
+                expand = c(0.01, 0.01)) +
+        scale_y_continuous(limits = c(0, 16), expand = c(0, 0)) +
+        labs(x = "Chromosome", y = expression('-Log'[10]*'(p)')) +
+        geom_hline(
+                yintercept = -log10(bonferroni_signifiance_threshold),
+                linetype = "dashed") +
+        shared_theme +
+        theme(
+                panel.grid.major.x = element_blank(),
+                panel.grid.minor.x = element_blank(),
+                legend.position = "none"
+        ) +
+        facet_grid(. ~ gwas_type)
+
+print("Saving plot")
+ggsave(
+        plot = overview_comparison,
+        filename = paste(FIGDIR, "gwas_results_xanthomonas.pdf", sep = "/"),
+        width = linewidth, height = height / 2, unit = "cm"
+)
+```

### 2022-06-23 manuscript/hiv-gwas-manuscript-only.tex
Update on Overleaf.

--- 2022-06-23 manuscript/hiv-gwas.tex
@@
 Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \cite{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \cite{collins_phylogenetic_nodate}.
-% \citep{Earle2016} simultaneously test for pathogen lineage and locus-level associations with a trait using a linear mixed model framework.
-\citep{collins_phylogenetic_nodate} test for locus-level associations that are significantly stronger than expected under a null model considering shared pathogen ancestry.
 These approaches are designed to quantify genetic effects from one organism on a trait.
@@
 These methods focus on capturing and accounting for correlations due to the pathogen phylogeny, without further investigating the nature of these correlations.
-%However, the top principle components capture only some of the information from the full pathogen phylogeny. Furthermore, we would like to simultaneously address issues (i) and (ii).
 
@@
 The advances presented here are on the first step, while in the second step existing, highly optimized tools to perform GWAS association tests under a variety of models can be employed.
-
-% We hypothesis this should increase GWAS power for several reasons.
-For one, pathogen effects may be confounding or extraneous variables that produce distorted or spurious associations between host genetic variants and a disease trait in host GWAS.
-
-%This may occur if the virus affects the disease trait (the dependent variable in host GWAS) or if it affects both the disease trait and the host genotype (the independent variable in host GWAS).
-%For example, stratification in both the host and pathogen populations may yield spurious correlations in host GWAS that are really due to independent pathogen factors. 
-% Then, co-evolution between a pathogen and its host may yield distorted estimates for host genetic effects that are not broadly protective against all pathogen strains.
-Finally, trait values from a population are not truly independent samples, given that hosts closer in the transmission chain have more similar pathogen strains and therefore more similar trait values.
-
-% % Why is spVL and important trait
-% For HIV, GWAS have used a trait called set point viral load (spVL) to identify human variants associated with the severity of disease course.
-spVL is generally defined to be the average concentration of viral RNA copies in host plasma during the asymptomatic phase of infection in the absence of treatment (see e.g.
-\cite{Alizon2010}).
-In untreated individuals, spVL is predictive of the duration of asymptomatic infection \citep{Mellors1996} and infectiousness \citep{Quinn2000}.
-If viral load can be reduced to undetectable levels, an individual is effectively uninfectious and the risk of disease progression is massively reduced \citep{HHSARTGuidelines2019}.
-Notably, spVL varies by orders of magnitude between individuals \citep{Mellors1996}.
-Thus, spVL measurements point to a wide range in natural HIV control amongst individuals.
-
-% % How do we measure the extent to which genetic factors determine spVL
-% To understand HIV pathogenicity, it is important to understand to what extent spVL is determined by host genetic factors as opposed to pathogen genetic factors.
-Heritability is a key measure of how genetically-determined a trait is.
-On the host side, narrow-sense heritability, the fraction of total trait variance explained by additive genetic effects, is a typical output of GWAS.
-% % How heritable is spVL (human side)
-% Several GWAS have measured the narrow-sense heritability of spVL and identified associated host genetic variants \citep{Bartha2013, Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012}.
-The largest study to-date by \citet{McLaren2015} estimated that additive human genetic effects account for approximately 25\% of the variation in spVL amongst infected individuals.
-All but 5\% of this was attributed to two regions in the human genome, the major histocompatibility complex (MHC) and C-C motif chemokine receptor 5 (\emph{CCR5}).
-Both associations are biologically relevant: the MHC encodes proteins that present viral epitopes at the cell surface and \emph{CCR5} encodes a co-receptor for HIV-1 cell entry.
-In other words, MHC proteins match bits of the virus like puzzle pieces and display these to signal that a cell is infected.
-CCR5 proteins help the virus infect target cells.
-
-% How heritable is spVL (viral side)
-% On the pathogen side, broad-sense heritability $H^2$ is typically measured.
-This is the fraction of total trait variance due to any factors ``inherited'' from infection partner to infection partner (the most obvious being pathogen genetic factors).
-Broad-sense heritability is commonly measured by quantifying how correlated a trait is between infection partners or how phylogenetically structured a trait is based on a reconstructed pathogen phylogeny.
-For HIV spVL, estimates differ depending on the methods employed and the cohort studied (see \citet{Mitov2018} for a discussion of this uncertainty).
-Using phylogenetic methods on large UK and Swiss cohorts, \citet{Mitov2018} and \citet{Bertels2018} estimated that pathogen factors account for 21\% - 29\% of the variation in spVL amongst infected individuals.
-However, variation in the MHC is known to exert strong selective pressure on the virus \citep{Kloverpris2016, Nguyen2021SystematicLink}.
-If the virus can change its ``puzzle piece'' shape to escape MHC-presentation, infected cells can go undetected.
-This means that MHC variants affect spVL largely via selection on the virus \citep{Bartha2017}.
-In summary, human genetic factors play a role in determining spVL, but these effects may be due to interaction with specific viral genetic variants.
-
-% % The gap
-% Most of the GWAS for human genetic determinants of spVL conducted so far \citep{Dalmasso2008, Fellay2007, Pereyra2010, Fellay2009, Pelak2010, VanManen2011, McLaren2012, McLaren2015} do not explicitly consider any viral effect on spVL.
-In these GWAS, viral genetic effects are lumped in with residual variance due to other, non-genetic factors.
-This has several potential negative consequences.
-% n GWAS, hosts are typically selected to not share recent ancestry.
-Further, shared environmental variables are included as covariates to avoid spurious associations.
-We argue similar considerations must be taken
-% for pathogen effects that heritable from one transmission partner to another \citep{Leventhal2016}.
 
@@
 The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}).
-As shown in Figure \ref{fig:sigma-vs-alpha}, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization.
+As shown in Figure S1, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization.
 For each $H^2$ and $\alpha$ value considered in the simulation, we recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
@@
 In summary, we show that it is theoretically possible to improve GWAS power for heritable infectious disease traits by estimating and removing pathogen effects using information from the pathogen phylogeny.
-
-% More accurate estimation of the non-viral part of spVL does not guarantee improved GWAS power.
-For example, if the viral part of spVL were constant across all viruses, association results (significance and effect size) would not be effected by a viral contribution to the trait.
-% Under our simulation set-up a standard GWAS would have $\approx$ 30\% power to detect a host variant explaining 1.25\% of the variability in spVL.
-This matches simulation results (figure \ref{fig:simulationResults}B, scaled spVL).
 
@@
 
-\newpage
-
-%Additional ideas for supplemental materials:
-%\begin{itemize}
-%	\item Table of genotype samples by project (genotyping chip, number of samples, spVL distribution)
-%	\item Link to github with all the code used to run simulations and generate figures
-%	\item Example spVL plots
-%	\item Proof of robustness of method to inferred tree -- NA because I just re-ran the _whole_ thing and it's not that different
-%\end{itemize}
-
-\newpage
-\startsetups
-    \setcounter{figure}{0}
-    \renewcommand{\thefigure}{S\arabic{figure}}
-    \setcounter{table}{0}
-    \renewcommand{\thetable}{S\arabic{table}}
-\stopsetups
-\section*{Supplemental Material}\label{supp}
-\beginsupplement
-\setcounter{page}{1}
-
-\begin{doublespace}
-
-\subsection*{Expected results from simulations}
-
-Here we show the root mean square error (RMSE) of the scaled trait value for each individual, $z_i - \bar{z}$ as an estimate for the host part of the trait for each individual, ${h_i}$, should be $\approx$ 0.74 in our simulation scheme.
-First we write the expression for the RMSE:
-
-\begin{equation}
-    RMSE = \sqrt{\frac{\sum_i^N(z_i - \bar{z} - h_i)^2}{N}}
-\end{equation}
-
-Note that under our simulation setup, $z_i - \bar{z}$ differs from $h_i$ due to the individual pathogen effect $g_i$ and environmental effect $e_i$.
-So the term inside the square root equals the combined variance of these two effects:
-
-\begin{equation}
-    RMSE = \sqrt{\sigma^2_{g} + \sigma^2_e}
-\end{equation}
-
-We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{h}$, are fixed parameters in our simulation scheme.
-
-\begin{equation}
-\begin{split}
-    \sigma^2_{h} + \sigma^2_{g} + \sigma^2_e = \sigma^2_z \\
-    0.25 * \sigma^2_{z} + \sigma^2_{g} + \sigma^2_e = \sigma^2_z \\
-    \sigma^2_{g} + \sigma^2_e = 0.75 * \sigma^2_z \\
-    \sigma^2_{g} + \sigma^2_e = 0.75 * 0.73 \\
-    \sigma^2_{g} + \sigma^2_e = 0.55
-\end{split}
-\end{equation}
-
-Therefore, we can expect the RMSE for $z_i - \bar{z}$ as an estimate for ${h_i}$ to be around $\sqrt{0.55} \approx 0.74$.
-
-\newpage
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.7\linewidth]{figures/sigma_vs_alpha.png}
-	\caption{Relationship between the selection strength parameter $\alpha$ and the intensity of stochastic evolutionary fluctuations parameter $\sigma$ at two different heritability ($H^2$) values in the  simulation scheme.
-$\sigma$ was determined as a function of $\alpha$ and $H^2$ under the POUMM (function given in Table \ref{tab:sim-params}).}
-	\label{fig:sigma-vs-alpha}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{tikzpicture}
-
-	% Variance 
-	\begin{scope}[xshift = 0cm, yshift = 0cm]
-	\node[anchor = west] (alabel) at (-1, 4) {\textbf{A} Variance partitioning};
-	\node[constnode] (H2h) at (0, 0) {0.25};
-	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
-	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {\textcolor{red}{$H^2_{\bar{t}}$}};
-	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{h}$};
-	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g}(\bar{t})$};
-	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{e}$};
-	\draw[taro] (H2h) -- (varh);
-	\draw[taro] (varz) -- (varh);
-	\draw[taro] (varz) -- (varv);
-	\draw[taro] (H2) -- (varv);
-	\draw[taro] (varh) -- (vare);
-	\draw[taro] (varv) -- (vare);
-	\draw[taro] (varz) -- (vare);
-	\end{scope}
-
-	% Virus
-	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
-	\node[anchor = west] (blabel) at (-1, 4.5) {\textbf{B} Generating pathogen effects};
-	\node[constnode] (theta) at (0, 1.5) {$\theta$};
-	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
-	\node[constnode] (alpha) at (3.6, 1.5) {\textcolor{red}{$\alpha$}};
-	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
-	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
-	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g}(\bar{t})$};
-	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
-	\node[snode] (v) at (3.6, 3.25) {$\vec{g}$};
-	\draw[taro] (theta) -- (v);
-	\draw[taro] (g0) -- (v);
-	\draw[taro] (alpha) -- (v);
-	\draw[taro] (sigma) -- (v);
-	\draw[taro] (tree) -- (v);
-	\draw[taro] (varv) -- (sigma);
-	\draw[taro] (tbar) -- (sigma);
-	\draw[taro] (tbar) -- (tree);
-	\draw[taro] (alpha) -- (sigma);
-	\node[white] (varpic) at (3.6+0.9, -1.75) {};
-	\draw[taro] (varpic) -- (varv);
-	\node (vdist) at ($(v) + (1.8, 0.3)$) {Gaussian};
-	\draw (v) -- (vdist);
-	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
-	\end{scope}
-
-	% Host 
-	\begin{scope}[xshift = 0cm, yshift = -7.5cm]
-	\node[anchor = west] (clabel) at (-1, 5.55) {\textbf{C} Generating host effects};
-	\node[constnode] (d) at (1.8, 1.5) {2};
-	\node[detnode] (p) at (3.6, 1.5) {$p$};
-	\node[snode] (G) at (1.8, 3) {$G$};
-	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
-	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
-	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{h}$};
-	\node[constnode] (delta) at (3.6, 3) {$\delta$};
-	\node[detnode] (h) at (3.6, 4.5) {$\vec{h}$};
-	\node[constnode] (M) at (5.4, 3) {$M$};
-	\draw[taro] (d) -- (G);
-	\draw[taro] (p) -- (G);
-	\draw[taro] (varh) -- (p);
-	\draw[taro] (delta) -- (p);
-	\draw[taro] (M) -- (p);
-	\draw[taro] (M) -- (h);
-	\draw[taro] (delta) -- (h);
-	\draw[taro] (G) -- (h);
-	\node[white] (varpic) at (3.6, -1.5) {};
-	\draw[taro] (varpic) -- (varh);
-	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Gaussian};
-	\draw (h) -- (hdist);
-	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
-	\draw (G) -- (gdist);
-	\end{scope}
-
-	% Environment
-	\begin{scope}[xshift = 9cm, yshift = -6.5cm]
-	\node[anchor = west] (dlabel) at (-1, 3) {\textbf{D} Generating environmental effects};
-	\node[detnode] (vare) at (0, 0) {$\sigma^2_{e}$};
-	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
-	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{e}$};
-	\draw[taro] (zero) -- (e);
-	\draw[taro] (vare) -- (e);
-	\node[white] (varpic) at (0, -1.5) {};
-	\draw[taro] (varpic) -- (vare);
-	\node (edist) at ($(e) + (1.8, 0.3)$) {Gaussian};
-	\draw (e) -- (edist);
-	\end{scope}
-
-	\end{tikzpicture}
-	\caption{A graphical model representation of our simulation scheme, following the recommendations in \cite{Hohna2014}.
-Variables in solid squares are constants, with the two master control variables that we vary from simulation to simulation highlighted in red.
-Variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
-Arrows represent dependencies among variables and the dashed square represents repetition.
-All parameters are defined in Table \ref{tab:sim-params}, as well as the values or expressions used for them.
-(A) shows how the variance in the simulated environmental effect $\sigma^2_{e}$ is smaller if the master pathogen heritability value $H^2_{\bar{t}}$ is higher and vice-versa.
-(B) shows the OU parameters and the pathogen phylogeny, which generate the Gaussian-distributed pathogen effects.
-The OU parameters $\theta$ and $g_0$ are fixed, whereas $\sigma$ is a deterministic function of the variance in the pathogen effect and the value of $\alpha$.
-In other words, we use $\sigma$ to maintain the desired pathogen heritability while varying $\alpha$.
-(C) shows how host genotypes are drawn to generate host effects.
-The host genotype matrix $G$ contains the number of copies (0, 1, or 2) for each of $M$ causal variants with effect size $\delta$.
-We assume half the variants have a positive effect and half have a negative effect.
-The allele frequency $p$ for the causal variants set so that we achieve the desired variance in the host effects.
-(D) shows that the environmental effect is drawn from a Gaussian distribution with mean zero and variance as determined in part (A).}
-	\label{fig:sim-design}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.7\linewidth]{figures/spvl_calculation_comparison.png}
-	\caption{A comparison of different ways to calculate spVL based on viral load measurements provided by the SHCS.
-The stricter filtering excludes all measurements possibly $<$ 6 months after infection and after treatment or AIDS, whereas the more lenient filtering excludes only measurements after treatment.
-We used the lenient filter, mean measurement values because these correlate well with the values from the stricter filter but allow us to retain many more individuals from the cohort for our study.}
-	\label{fig:spvl-calc-comparison}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on HIV-1 spVL data from the SHCS. We ran two different MCMC chains to ensure the estimates converged.}
-\label{fig:poumm-parameters}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/spvl_on_tree.png}
-	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}
-	\label{fig:spvl-on-tree}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.462\linewidth]{figures/trait_values.png}
-	\caption{A comparison of measured (calculated) spVL values versus our estimated non-pathogen effect on spVL for each SHCS cohort member used in the study.
-The histograms show the marginal distribution of each value across the individuals.}
-	\label{fig:trait-comparison}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{figures/qq_plots.png}
-	\caption{Quartile-quartile plots from HIV-1 spVL association tests. The dashed green line shows the y = x line.}
-	\label{fig:qq-plots}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/qdr_on_tree_arboricola.png}
-	\caption{Inferred \emph{X. arboricola} phylogeny with tips colored by (A) calculated QDR, (B) estimated non-pathogen effects on QDR and (C) estimated pathogen effects on QDR.}
-	\label{fig:qdr-on-tree}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates_arboricola.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on \emph{A.
-thaliana}-\emph{X.
-arboricola} data.
-We ran two different MCMC chains to ensure the estimates converged.}
-\label{fig:poumm-parameters-qdr}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.462\linewidth]{figures/trait_values_arboricola.png}
-	\caption{A comparison of measured (calculated) mean QDR values across all \emph{A.
-thaliana} accession pairings and replicates  versus our estimated non-pathogen effect on mean QDR for each \emph{X.
-arboricola} pathogen strain.
-The histograms show the marginal distribution of each value across the strains.
-The Pearson correlation coefficient is 1.}
-	\label{fig:trait-comparison-qdr}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{figures/qq_plots_arabidopsis_xanthomonas.png}
-	\caption{Quartile-quartile plots from \emph{A. thaliana}-\emph{X. arboricola} QDR association tests. The dashed blue line shows the y = x line.}
-	\label{fig:qq-plots-qdr}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\includegraphics[width=\linewidth]{figures/host_genotype_pca.png}
-	\caption{SHCS individuals and HapMap3 individuals plotted along the top three principle components of genetic variation.
-Points with black borders are within the thresholds used to select individuals of likely European ancestry.}
-	\label{fig:PCA}
-\end{figure}
-
-\end{doublespace}  % don't doublespace tables
-
-\newpage
-
-\begin{table}[H]
-	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
-	\begin{tabularx}{\linewidth}{p{1.5cm}p{6.5cm}l} \toprule 
-		Variable & Expression & Definition \\ \midrule 
-		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & Total spVL variance \\ 
-		$H^2_h$ & 0.25 & Host heritability of spVL \\
-		$H^2_{\bar{t}}$ & varied & Pathogen heritability of spVL at $\bar{t}$ \\
-		$\sigma^2_{h}$ & $\sigma_{h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
-		$\sigma^2_{g}(\bar{t})$ & $\sigma^2_{g}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in pathogen part of spVL at $\bar{t}$ \\ 
-		$\sigma^2_{e}$ & $\sigma_{e}^2 = \sigma_z^2 - \sigma^2_{g} - \sigma^2_{h}$ & Variance in environmental part of spVL \\ 
-		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in pathogen phylogeny \\ 
-		$\bm{g}$ & $\bm{g} \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & Pathogen part of spVL for all individuals \\ 
-		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
-		$g_{0}$ & 4.5 log copies/mL & $g$ at the root of the phylogeny  \\ 
-		$\alpha$ & varied & Selection strength of OU process \\ 
-		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
-		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Pathogen phylogeny \\ 
-		$h_i$ & $h_i = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & Host part of spVL for individual $i$ \\ 
-		$G_{N \times M}$ & \makecell{$G_{ij} \sim Binom(2, p)\ \\ \forall i \in {1...N}, \forall j \in {1...M}$} & Host genotype matrix \\
-		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
-		$\delta$ & 0.2 & Host variant effect size \\ 
-		$M$ & 20 & Number of causal host variants \\ 
-		$e_i$ & $ e_i \sim Norm(0, \sigma^2_{e})$ & Environmental part of spVL for individual $i$ \\ 
-		$N$ & 500 & Number of simulated samples \\ \bottomrule 
-	\end{tabularx}
-	\label{tab:sim-params}
-\end{table}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for spVL based on SHCS data. HPD = Highest posterior density.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Parameter & Posterior mean & 95\% HPD \\ \hline 
-	${g_{0}}$ & 4.23 & (1.72, 6.71) \\
-	$\theta$ & 4.47 & (4.37, 4.58) \\
-	$\sigma$ & 5.25 & (2.37, 7.9) \\
-	$\alpha$ & 57.65 & (19.49, 95.2) \\
-	$\sigma_\epsilon$ & 0.54 & (0.43, 0.65) \\ 
-	$H^2_{\bar{t}}$ & 0.45 & (0.24, 0.67) \\ \hline
-	\end{tabularx}
-	\label{tab:POUMM-parameters-our-estimates}
-\end{table}
-
-% latex table generated in R 4.1.1 by xtable 1.8-4 package
-% Sat Oct 30 10:31:16 2021
-\begin{table}[ht]
-\caption{Effect size and p-values from the top most strongly associated variants in the $\emph{CCR5}$ and MCR regions from each of the two GWAS performed in our study.
-``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with the estimated non-pathogen part of the trait.
-Entries above the dividing line are the top-associated variants from the ``Standard'' GWAS and entries below the dividing line are the top-associated variants from the ``Corrected'' GWAS.
-Many entries overlap between the two.}
-\begin{tabularx}{\linewidth}{XXXXXXX}
-  \hline
-    Region & Position & Variant & Standard effect size & Standard p-value & Corrected effect size & Corrected p-value \\ 
-  \hline
-    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46540932 & rs10865942 & -0.16 & $8.4 \times 10^{-9}$ & -0.081 & $4 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
-    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
-    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
-    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
-    MHC & 31391401 & rs4418214 & -0.34 & $2.4 \times 10^{-14}$ & -0.18 & $2.5 \times 10^{-12}$ \\ 
-    MHC & 31400137 & rs138130755 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31400705 & rs138117378 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31402358 & rs148792134 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31409677 & rs140991764 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    \hline
-    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46556835 & rs6808142 & 0.15 & $8.3 \times 10^{-8}$ & 0.082 & $3.3 \times 10^{-7}$ \\ 
-    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
-    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
-    MHC & 31367874 & rs111281598 & -0.37 & $1.5 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
-    MHC & 31376266 & rs73400361 & -0.37 & $1.4 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
-    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:gwas-top-associations}
-\end{table}
-
-\begin{table}[H]
-\centering
-\caption{Results comparison using three different approximate maximum likelihood HIV-1 phylogenies (see Materials and Methods).}
-\begin{tabularx}{\linewidth}{p{1.8cm}p{2cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}p{1.8cm}} % {lrrrrrr} {llclcccc}
-\hline
-& \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\main tree}} & \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\tree 2}} & \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\tree 3}} \\
-  \hline
- Variant & Effect size & p-value & Effect size & p-value & Effect size & p-value \\
-  \hline
-    rs59440261 & -0.22 & $2.6 \times 10^{-10}$ & -0.22 & $1.5 \times 10^{-10}$ & -0.24 & $1.4 \times 10^{-10}$ \\ 
-    rs1015164 & 0.078 & $8.5 \times 10^{-6}$ & 0.076 & $9.5 \times 10^{-6}$ & 0.083 & $9 \times 10^{-6}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:comp-gwas-mclaren-snps-alt-trees}
-\end{table}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for QDR based on \emph{A. thaliana}-\emph{X. arboricola} data. HPD = Highest posterior density.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Parameter & Posterior mean & 95\% HPD \\ \hline 
-	${g_{0}}$ & 0.40 & (0.01, 0.78) \\
-	$\theta$ & 0.39 & (0.30, 0.49) \\
-	$\sigma$ & 0.03 & (0.0, 0.06) \\
-	$\alpha$ & 0.03 & (0.0, 0.05) \\
-	$\sigma_\epsilon$ & 0.18 & (0.08, 0.27) \\ 
-	$H^2_{\bar{t}}$ & 0.33 & (0.0, 0.77) \\ \hline
-	\end{tabularx}
-	\label{tab:POUMM-parameters-our-estimates-qdr}
-\end{table}
-
-\begin{table}[H]
-\caption{Summary statistics for log spVL in previously sampled populations.
-$\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$).
-Values from \protect\citep{Blanquart2017, Mitov2018} are empirical; values from \protect\citep{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Measurement & Value & Reference \\ \hline 
-	$\bar{z}$ & $\approx$ 4.5 & \citet{Mitov2018} \\
-	$\bar{z}$ &  4.4 & \citet{Blanquart2017} \\
-	$\bar{z}$ & $\approx$ 4.5 & \citet{Bonhoeffer2015} \\
-	$\sigma^2_z$ &  0.73 &  \citet{Mitov2018} \\
-	$\sigma^2_z$ &  0.50 & \citet{Blanquart2017} \\
-	$\sigma^2_z$ & $\approx$ 0.5 & \citet{Bonhoeffer2015} \\ \hline 
-	\end{tabularx}
-	\label{tab:spVLmeasurements}
-\end{table}
-%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for spVL from previous studies.}
-	\begin{tabularx}{\linewidth}{llXXX}
-	\hline 
-	Parameter & Value (Uncertainty) & Reference & Notes \\
-	\hline 
-	$g_0$ & 5.54 (4.04 - 7.25) & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
-	$\theta$ & 4.45 (4.41 - 4.49) & \citet{Mitov2018} & \\
-	$\theta$ &  4.0 (1.6 - 4.)) & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
-	$\theta$ & 4.1 (3.5 - 4.9) & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
-	$\alpha$ & 28.78 (16.64 - 46.93) & \citet{Mitov2018} & \\
-	$\alpha$ & 32.7 (0.03 - 57.6) & \citet{Bertels2018} & \\
-	$\alpha$ & 7.6 (1.2 - 10) & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
-	$\sigma$ & 2.97 (1.95 - 4.37) & \citet{Mitov2018} & \\
-	$\sigma$ & 1.3 (0.66 -1.87) &  \citet{Blanquart2017} & \\
-	$\sigma_e$ & 0.77 (0.73, 0.8) & \citet{Mitov2018} & \\
-	$\sigma_e$ & 0.61 (0.54, 0.65) &  \citet{Blanquart2017} & \\
-	\hline 
-	\end{tabularx}
-	\label{tab:POUMMparams-prev-estimated}
-\end{table}
-
-\begin{table}[H]
-\caption{Number of samples for HIV-1 spVL GWAS after sequential filtering steps.}
-    \begin{tabularx}{\linewidth}{XX}
-    \hline
-    Sample filter & Number of samples remaining \\
-    \hline
-    Subtype B pol sequences & 1516 \\
-    With paired spVL measurement & 1516 \\
-    $>$ 750 characters in sequence & 1493 \\
-    Individual is of European ancestry & 1396 \\
-    Kinship coefficient $>$ 0.09375 & 1392 \\ 
-    \hline
-    \end{tabularx}
-    \label{tab:sample-filtering}
-\end{table}
-
-\begin{table}[H]
-\caption{Number of variants for HIV-1 spVL GWAS after sequential filtering steps.}
-	\begin{tabularx}{\linewidth}{XX}
-	\hline
-	Variant filter & Number of variants remaining \\
-	\hline 
-	Raw data & 76979521 \\
-	Missing genotype rate $>$ 0.05 & 11590002 \\
-	Hardy-Weinburg exact test p-value $< 5 \times 10^{-5}$ & 11589246 \\
-	Minor allele frequency $<$ 0.01 & 6228626 \\ \hline			
-	\end{tabularx}
-	\label{tab:variant-filtering}
-\end{table}
-
-\begin{table}[H]
-\caption{Number of variants for \emph{A.
-thaliana} QDR GWAS after sequential filtering steps.
-The last entry lists variants without GWAS p-values because PLINK assessed the correlation between predictor variables (the variant and the top 5 principle components of host genetic variation are predictors) to be too strong.
-This did not occur in the HIV-1 spVL GWAS.}
-	\begin{tabularx}{\linewidth}{XX}
-	\hline
-	Variant filter & Number of variants remaining \\
-	\hline 
-	Raw data & 12883854 \\
-	Bi-allelic variants & 11769920 \\
-	Minor allele frequency $<$ 0.1 & 1743952 \\
-	NA p-value (too-high covariate correlation) & 1070541 \\ \hline			
-	\end{tabularx}
-	\label{tab:variant-filtering-qdr}
-\end{table}
-
-\stopsection
 \end{document}

### 2022-06-23 manuscript/response_to_reviewers.tex
Update on Overleaf.

@@
 arboricola}, so we feel that updating our example GWAS is beyond the scope of the current study.
-We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS....
-In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK (Chang et al., 2015), though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
+We added this explanation to the discussion: ``In cases where a correction can be estimated and applied using our method, the corrected trait values are envisioned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK (Chang et al., 2015), though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
 \end{enumerate}
@@
     \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017).
-Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in (Earle et al., 2016) and modified association tests that account for phylogenetic relationships amongst samples as in (Collins and Didelot, 2018).''}
+Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in Earle et al., (2016) and modified association tests that account for phylogenetic relationships amongst samples as in Collins and Didelot, (2018).''}
     \item L110 'Of these, only pathogen effects are heritable from one transmission partner to another'.
@@
     \textcolor{blue}{We anticipate antimicrobial resistance would be under strong selective pressure, but agree this could produce phylogenetically correlated trait values.
-We added a statement in the discussion: ``It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance.
-However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.''}
+We added a statement in the discussion: ``Finally, as we show here, our method is not anticipated to be useful in certain evolutionary scenarios.
+For instance, traits like antimicrobial resistance may be under strong selection pressure and be highly heritable.
+In these instances, our simulations do not point to a large improvement when adding our pre-processing step.
+In any case, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect, meaning our approach would not apply.
+In this situation, one would rather account for antimicrobial resistance as a covariate in the GWAS association model.''}
 \end{enumerate}

### 2022-06-23 manuscript/hiv-gwas-manuscript-only.tex
Update on Overleaf.

@@
 
-\begin{figure}[H]
-	\begin{center}
-	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
-	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
-(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
-(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
-Each variant can be present in 0, 1, or 2 copies.
-Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
-(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
-These three effects sum to the trait value for each simulated individual.}
-		\label{fig:spVLModel}
-	\end{center}
-\end{figure}
-
 PMMs have previously been applied to the study of infectious disease traits using two different types of random processes to model trait evolution.
@@
 For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
-The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}).
+The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table S1).
 As shown in Figure S1, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization.
@@
 The variance of this distribution was scaled based on the pathogen heritability of the trait, from 0 (no affect) in the scenario with 75\% pathogen heritability and 25\% host heritability to 0.44 in the scenario with 15\% pathogen heritability and 25\% host heritability.
-Figure \ref{fig:sim-design} provides a more detailed schematic of this simulation framework and Table \ref{tab:sim-params} gives the value or expression for each parameter.
+Figure S2 provides a more detailed schematic of this simulation framework and Table S1 gives the value or expression for each parameter.
 
@@
 
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/simulation_results}
-		\caption{Results from the simulation study.
-We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters.
-For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
-(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
-The RMSE is with reference to the true (simulated) host part of the trait values.
-Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
-(B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
-Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
-The points highlight specific heritability and selection strength values from the \emph{A.
-thaliana}-\emph{X.
-arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
-		\label{fig:simulationResults}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
-		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
-These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
-(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively.
-Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
-(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%.
-The solid line is the y=x line.
-Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
-		\label{fig:simulationTraitExample}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-    \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
-	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
-These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
-Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
-Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
-	\label{fig:tip-trait-correlation}
-	\end{center}
-\end{figure}
-
 \subsection*{Application to HIV-1 set-point viral load}
@@
 We followed the method outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
-Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
-We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}).
+Figure S3 shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
+We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure S4, Table S2).
 To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}.
 The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
-However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.
-Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
-
-\begin{figure}[H]
-\begin{center}
-    \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
-	\caption{A high-level schematic of the experimental setup for the two application datasets.
-For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
-We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
-For (B) \emph{A.
-thaliana}-\emph{X.
-arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
-We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
-	\label{fig:application-exp-setup}
-	\end{center}
-\end{figure}
+However, Figure S5 shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.
+Finally, figure S6 shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
 
@@
 Figure \ref{fig:gwas-results}A shows that results are qualitatively similar between the two GWAS.
-Q-Q plots show the distribution of p-values are very similar as well (Figure \ref{fig:qq-plots}).
+Q-Q plots show the distribution of p-values are very similar as well (Figure S7).
 Figure \ref{fig:gwas-results}B shows how the strength of association changed for some variants in the MHC and \emph{CCR5} regions.
@@
 Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
-However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}).
+However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table S3).
 Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date.
 Effect sizes are smaller with a phylogenetic correction and p-values are slightly increased.
-We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}).
+We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table S4).
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
-
-\begin{figure}[H]
-	\begin{center}
-		\includegraphics[width=\linewidth]{figures/gwas_results}
-		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
-(A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables.
-On the left, we used unmodified (total) spVL values.
-On the right, we used our estimates for the non-pathogen effects on spVL.
-The alternating shades correspond to different chromosomes.
-(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
-Base positions are with reference to genome build GRCh37.
-The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
-Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
-The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
-		\label{fig:gwas-results}
-	\end{center}
-\end{figure}
-
-\begin{table}[H]
-\centering
-\caption{Top association results from \citet{McLaren2015} compared to results from this study.
-Results from this study are for host variants from the SHCS in GWAS with two different response variables.
-``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
-\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
-\hline
-& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
-  \hline
- Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
-  \hline
-    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
-    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:comp-gwas-mclaren-snps}
-\end{table}
 
@@
 thaliana} types (Figure \ref{fig:application-exp-setup}B).
-Figure \ref{fig:qdr-on-tree}A shows the inferred \emph{X.
+Figure S8A shows the inferred \emph{X.
 arboricola} pathogen phylogeny annotated with the  mean QDR trait value used for each strain.
 Mean QDR  was generally low, varying between 0.11 for strain NL\_P126 and 0.78 for strain FOR\_F21.
-Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table \ref{tab:POUMM-parameters-our-estimates-qdr}).
-These values deviated significantly from the respective priors (Figure \ref{fig:poumm-parameters-qdr}).
-Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table \ref{tab:POUMM-parameters-our-estimates-qdr}).
+Fitting the POUMM yielded very low selection strength $\alpha$ and intensity of stochastic fluctuations  $\sigma$ parameter estimates (posterior mean 0.03 with 95\% HPD 0.0 - 0.05 and 0.03 with 95\% HPD 0.0 - 0.06, respectively; Table S5).
+These values deviated significantly from the respective priors (Figure S9).
+Heritability, on the other hand, was quite uncertain (posterior mean 0.33 with 95\% HPD 0.0 - 0.77; Table S5).
 The posterior mean selection strength and heritability values are also shown in the context of the simulation study as points on Figure \ref{fig:simulationResults}.
@@
 Thus, the phylogeny is not very informative for a trait value correction.
-Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure \ref{fig:trait-comparison-qdr}).
+Indeed, the estimated pathogen part of the QDR trait calculated by our method is simply a scaling of the total QDR trait value (Figure S10).
 We anyways selected 22 random host-pathogen strain pairings to perform a comparative GWAS analogous to that for HIV-1 spVL, where each host is infected with a single pathogen strain.
@@
 Figure \ref{fig:gwas-results-qdr} shows that results are qualitatively similar between the two GWAS, with a slight decrease in association strength for the top-associated variants.
-Q-Q plots show the distribution of p-values are also very similar (Figure \ref{fig:qq-plots-qdr}).
+Q-Q plots show the distribution of p-values are also very similar (Figure S11).
 In the first, standard GWAS, one \emph{A.
@@
 arboricola}.
-
-\begin{figure}[H]
-\begin{center}
-    \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
-	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
-thaliana} quantitative disease resistance (QDR) to \emph{X.
-arboricola}.
-The two facets show association p-values for the same host \emph{A.
-thaliana} variants in GWAS with two different response variables.
-On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
-On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
-In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
-thaliana} types.
-The alternating shades correspond to different chromosomes.
-The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
-	\label{fig:gwas-results-qdr}
-	\end{center}
-\end{figure}
 
@@
 We set the total variance in spVL to 0.73 log copies$^2$ mL$^{-2}$ based on UK cohort data \citep{Mitov2018}.
-Other studies have estimated slightly lower values though (Table \ref{tab:spVLmeasurements}).
+Other studies have estimated slightly lower values though (Table S6).
 After allotting 25\% of this variance to a host part of spVL $h$ based on results by \citet{McLaren2015}, we partitioned the remaining variance between a viral part $g$ and an environmental part $e$ in different ratios to assess estimator performance across a range of spVL heritabilities.
 $h$ was simulated as the sum of contributions from 20 causal host genetic variants, 10 of which had an effect size of 0.2 log copies mL$^{-1}$ and 10 of which had an effect size of -0.2 log copies mL$^{-1}$.
-Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $h$ had the appropriate variance (see Table \ref{tab:sim-params}).
+Host genetic variants were generated from a binomial distribution with probability $p$ calculated such that $h$ had the appropriate variance (see Table S1).
 We generated a random viral phylogeny with branch lengths on the same time scale as a previously inferred UK cohort HIV tree \citep{Hodcroft2014} using the R package ape \citep{Paradis2018}.
 $g$ was simulated by running an OU process along the phylogeny using the R package POUMM \citep{Mitov2017a-POUMM} and sampling values at the tips.
-For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on previous estimates of mean spVL (Table \ref{tab:spVLmeasurements}).
-This is similar to values previously inferred for HIV (Table \ref{tab:POUMMparams-prev-estimated}).
+For the OU parameters $\theta$ and $g_{0}$ we used 4.5 log copies mL$^{-1}$ based on previous estimates of mean spVL (Table S6).
+This is similar to values previously inferred for HIV (Table S7).
 To assess our estimator's performance under a range of evolutionary scenarios, we co-varied the heritability $H^2$ and selection strength $\alpha$ parameters.
-The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table \ref{tab:sim-params}, Figure \ref{fig:sigma-vs-alpha}).
+The intensity of random fluctuations $\sigma$ was determined based on these parameters (Table S1, Figure S1).
 Finally, the environmental part of spVL $e$ was generated from a normal distribution with mean 0.
-For a full graphical model representation of the simulation scheme, see Figure \ref{fig:sim-design}.
+For a full graphical model representation of the simulation scheme, see Figure S2.
 
@@
 We calculated spVL as the arithmetic mean of viral RNA measurements made prior to the start of antiretroviral treatment.
-For a comparison of several different filtering methods, see Figure \ref{fig:spvl-calc-comparison}.
+For a comparison of several different filtering methods, see Figure S3.
 
-For GWAS, we retained data from 1,392 of the 1,493 SHCS individuals with European ancestry who were not closely related to other individuals in the cohort (Table \ref{tab:sample-filtering}).
+For GWAS, we retained data from 1,392 of the 1,493 SHCS individuals with European ancestry who were not closely related to other individuals in the cohort (Table S8).
 These were 227 females and 1165 males.
-Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure \ref{fig:PCA}).
+Ancestry was determined by plotting individuals along the three primary axes of genotypic variation from a combined dataset of SHCS samples and HapMap populations (Figure S12).
 Kinship was evaluated using PLINK version 2.3 \citep{Chang2015}; we used the --king-cutoff option to exclude one from each pair of individuals with a kinship coefficient $>$ 0.09375.
@@
 We used the options --maf 0.01, --geno 0.01, and --hwe 0.00005 to remove variants with minor allele frequency less than 0.01, missing call rate greater than 0.05, or Hardy-Weinberg equilibrium exact test p-value less than 5x10$^-5$.
-After quality filtering, approximately 6.2 million genetic variants from the 1,392 individuals were retained for GWAS (Table \ref{tab:variant-filtering}).
+After quality filtering, approximately 6.2 million genetic variants from the 1,392 individuals were retained for GWAS (Table S9).
 
@@
 arboricola} pathogen genotype matrix.
-The inferred tree topology (Figure \ref{fig:qdr-on-tree}) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes}, which was generated using the unweighted pair group method with arithmetic mean (UPGMA).
+The inferred tree topology (Figure S8) closely matches the hierarchical clustering presented in \citep{Wang2018Two-wayGenomes}, which was generated using the unweighted pair group method with arithmetic mean (UPGMA).
 Compared to UPGMA, the neighbor-joining method we used relaxes the assumptions of a strict molecular clock and sampling all at the same time-point.
@@
 After filtering, approximately 1.1 million genetic variants from \emph{A.
-thaliana} were retained for GWAS (Table \ref{tab:variant-filtering-qdr}).
+thaliana} were retained for GWAS (Table S10).
 
@@
 The first 2x10$^5$ samples of each chain were used for automatic adjustment of the MCMC proposal distribution.
-Figures \ref{fig:poumm-parameters} and \ref{fig:poumm-parameters-qdr} show the posterior distributions for inferred parameters for HIV-1 spVL and \emph{X.
+Figures S4 and S9 show the posterior distributions for inferred parameters for HIV-1 spVL and \emph{X.
 arboricola} QDR, respectively.
-Tables \ref{tab:POUMM-parameters-our-estimates} and \ref{tab:POUMM-parameters-our-estimates-qdr} give the posterior mean values used for subsequent calculations.
+Tables S2 and S5 give the posterior mean values used for subsequent calculations.
 
@@
 
+\newpage
+
+\section*{Tables}
+
+\begin{table}[H]
+\centering
+\caption{Top association results from \citet{McLaren2015} compared to results from this study.
+Results from this study are for host variants from the SHCS in GWAS with two different response variables.
+``Standard trait value'' means we used the unmodified (total) spVL value and ``Estimated non-pathogen part of trait'' means we used our estimates for the non-pathogen effects on spVL.}
+\begin{tabularx}{\linewidth}{p{1cm}p{1.8cm}p{2.4cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}} % {lrrrrrr} {llclcccc}
+\hline
+& & \makecell{McLaren \\ et al.}  & \multicolumn{2}{l}{\makecell{Standard \\ trait value}} & \multicolumn{2}{l}{\makecell{Estimated non-pathogen \\ part of trait}} \\ 
+  \hline
+ Region & Variant & p-value & Effect size & p-value & Effect size & p-value \\
+  \hline
+    MHC & rs59440261 & $2.0 \times 10^{-83}$ & -0.4 & $3.3 \times 10^{-11}$ & -0.22 & $2.6 \times 10^{-10}$ \\ 
+    \emph{CCR5} & rs1015164 & $1.5 \times 10^{-19}$ & 0.15 & $7.5 \times 10^{-7}$ & 0.078 & $8.5 \times 10^{-6}$ \\ 
+   \hline
+\end{tabularx}
+\label{tab:comp-gwas-mclaren-snps}
+\end{table}
+
+\newpage
+\section*{Figure legends}
+
+\begin{figure}[H]
+	\begin{center}
+% 	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
+(A) shows how the viral effects on spVL evolve along the viral phylogeny according to an Ornstein-Uhlenbeck process.
+(B) shows how human host genetic effects are the sum of independent effects from several causal variants.
+Each variant can be present in 0, 1, or 2 copies.
+Half the variants have a positive effect of size $\delta$ and half have a negative effect of size $\delta$.
+(C) shows how other environmental effects are independently drawn from a Gaussian distribution centered at 0.
+These three effects sum to the trait value for each simulated individual.}
+		\label{fig:spVLModel}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/simulation_results}
+		\caption{Results from the simulation study.
+We simulated host, pathogen, and environmental effects on a trait under the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) with different heritability ($H^2$; y-axis) and selection strength ($\alpha$; x-axis) parameters.
+For each simulated dataset, we applied our method to estimate the non-pathogen effects and performed GWAS with these values.
+(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario.
+The RMSE is with reference to the true (simulated) host part of the trait values.
+Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.
+(B) shows how genome-wide association study (GWAS) power can improve given the true, simulated non-pathogen effect on spVL (left) and using our estimate for this value (middle) compared to using the scaled trait value (right).
+Each tile's color corresponds to the average value across 20 simulated datasets of 500 samples.
+The points highlight specific heritability and selection strength values from the \emph{A.
+thaliana}-\emph{X.
+arboricola} quantitative disease resistance (QDR) analysis, HIV-1 spVL analysis, and four simulated scenarios that are presented in more detail in Figure \ref{fig:tip-trait-correlation}.}
+		\label{fig:simulationResults}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
+		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
+These examples correspond to two of the unfilled points in Figure \ref{fig:simulationResults}.
+(A) and (B) show total trait values for 12 randomly selected tips from the simulated phylogeny with pathogen heritability $H^2$ of 15 and 75\%, respectively.
+Depending on the pathogen heritability, trait values are more or less correlated at clustered tips.
+(C) compares our method's estimate for the non-pathogen part of trait values (y-axis) with true simulated host trait values (x-axis) with pathogen heritability of 15 and 75\%.
+The solid line is the y=x line.
+Selection strength $\alpha$ was fixed to 0.1 time $^{-1}$ for both scenarios and all other parameters were fixed as in the full simulation study.}
+		\label{fig:simulationTraitExample}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    % \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
+	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
+These examples correspond to the four unfilled points in Figure \ref{fig:simulationResults}.
+Correlations are calculated for pairs of tips binned by phylogenetic distance (into deciles) across the 20 replicate simulations for each of the four evolutionary scenarios.
+Trait values are only noticeably correlated for closely clustered tips under the scenario with high pathogen heritability $H^2$ and low selection strength $\alpha$/ low stochastic fluctuations $\sigma$ (upper left facet).}
+	\label{fig:tip-trait-correlation}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    % \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
+	\caption{A high-level schematic of the experimental setup for the two application datasets.
+For (A) HIV-1 set-point viral load (spVL) in the Swiss HIV Cohort Study, data are paired viral and human genotypes and associated spVL measurements.
+We fit the phylogenetic Ornstein-Uhlenbeck mixed model (POUMM) to the viral phylogeny and spVL values associated with each infected individual ($z_1$, $z_2$, ..., $z_{1493}$).
+For (B) \emph{A.
+thaliana}-\emph{X.
+arboricola} quantitative disease resistance (QDR) from \cite{Wang2018Two-wayGenomes}, data are bacterial and plant genotypes with QDR measurements for all possible combinations of pathogen and host plant strains.
+We fit the POUMM to the bacterial phylogeny and mean QDR calculated for each pathogen strain across all the hosts plant types ($\bar{z}_1$, $\bar{z}_2$, ..., $\bar{z}_{22}$).}
+	\label{fig:application-exp-setup}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+% 		\includegraphics[width=\linewidth]{figures/gwas_results}
+		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
+(A) shows association p-values for the same host variants from the Swiss HIV cohort in GWAS with two different response variables.
+On the left, we used unmodified (total) spVL values.
+On the right, we used our estimates for the non-pathogen effects on spVL.
+The alternating shades correspond to different chromosomes.
+(B) compares the strength of association for variants in the \emph{CCR5} and MHC regions between the two GWAS (positions 45.4 - 47Mb on chromosome 3 and 29.5 - 33.5Mb on chromosome 6 for the \emph{CCR5} and MHC, respectively).
+Base positions are with reference to genome build GRCh37.
+The color of each point represents the difference in -log$_{10}$ p-value between the two GWAS.
+Red means taking into account phylogenetic information decreased the strength of association and blue means it increased it.
+The dashed lines show genome-wide significance at p = $5 \times 10^{-8}$.}
+		\label{fig:gwas-results}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+\begin{center}
+    % \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
+	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
+thaliana} quantitative disease resistance (QDR) to \emph{X.
+arboricola}.
+The two facets show association p-values for the same host \emph{A.
+thaliana} variants in GWAS with two different response variables.
+On the left, we used unmodified (total) QDR values for each of the 22 selected host-pathogen pairings on which these results are based.
+On the right, we used our estimates for the non-pathogen effects on QDR for these samples.
+In this case, estimated non-pathogen effects are the specific QDR for each selected host-pathogen pairing, minus mean QDR for the respective pathogen strain, calculated across all the host \emph{A.
+thaliana} types.
+The alternating shades correspond to different chromosomes.
+The dashed lines show significance at significance level 0.05 with a Bonferroni correction for multiple testing.}
+	\label{fig:gwas-results-qdr}
+	\end{center}
+\end{figure}
+
 \end{document}

### 2022-07-18 scripts/R/make_manuscript_figures.rmd
fix fig 2: was using sensitivity analysis results for FPR with 100 tests, should be for main analysis with 20 tests

@@
 accuracy_simulation_results <- read.delim("output/2021-08-31_estimator_accuracy_MLE.txt")
-tpr_simulation_results_old <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
-tpr_simulation_results <- read.delim("output_revisions/2022-06-13_GWAS_TPR_MLE.txt")
+tpr_simulation_results <- read.delim("output/2021-08-31_GWAS_TPR_MLE.txt")
+N_TESTS <- 20
+# tpr_simulation_results <- read.delim("output_revisions/2022-06-13_GWAS_TPR_MLE.txt")  # includes non-causal variants to assess FPR
+# N_TESTS <- 100
 example_simulation_results <- read.delim("output_revisions/2022-06-11_simulateFullDataset.txt")
@@
         significance.correction = "Bonferroni",
-        N.tests = 100)
+        N.tests = N_TESTS)
 data$TP <- ifelse(data$p.value < sig.level & data$is.causal.variant, T, F)
@@
 
-plot_grid(aligned_plots[[1]], aligned_plots[[2]], nrow = 2, labels = c("A", "B"))
+p <- plot_grid(aligned_plots[[1]], aligned_plots[[2]], nrow = 2, labels = c("A", "B"))
 ggsave(
-        filename = paste(FIGDIR, "gwas_results.pdf", sep = "/"),
+        plot = p,
+        filename = paste(FIGDIR, "gwas_results.tiff", sep = "/"),
+        width = linewidth, height = height, unit = "cm", compression = "lzw"
+)
+ggsave(
+        plot = p,
+        filename = paste(FIGDIR, "gwas_results.png", sep = "/"),
         width = linewidth, height = height, unit = "cm"
@@
         plot = overview_comparison,
-        filename = paste(FIGDIR, "gwas_results_xanthomonas.pdf", sep = "/"),
+        filename = paste(FIGDIR, "gwas_results_xanthomonas.tiff", sep = "/"),
         width = linewidth, height = height / 2, unit = "cm"
 )
-```
+ggsave(
+        plot = overview_comparison,
+        filename = paste(FIGDIR, "gwas_results_xanthomonas.png", sep = "/"),
+        width = linewidth, height = height / 2, unit = "cm"
+)
+```

### 2022-07-18 scripts/R/fit_poumm_xanthamonas.R
update code comments

@@ -18,7 +18,6 @@ require(POUMM)
 require(bayesmeta)
 require(dplyr)
 require(ggplot2)
-# require(la)
 
 set.seed(SEED)  # Seed for POUMM parameter inference
 
@@ -54,13 +53,13 @@ POUMM_spec <- POUMM::specifyPOUMM_ATH2tMeanSeG0(
   parPriorMCMC = function(par) {
       # alpha
       dexp(par[1], rate = 0.02, log = T) +
-      # g0 or theta
+      # theta
       dnorm(par[2], mean = 0.4, sd = 0.2, log = T) + 
       # H2
       dunif(par[3],  min = 0, max = 1, log = T) + 
       # SigmaE
       bayesmeta::drayleigh(par[4], scale = 1/sqrt(2*0.02), log = T) + 
-      # g0 or theta
+      # g0
       dnorm(par[5], mean = 0.4, sd = 0.2, log = T)
   },
   nAdaptMCMC = N_ADAPT_MCMC,

### 2022-07-19 manuscript_proof/manuscript.tex
Update on Overleaf.

--- 2022-06-23 manuscript/hiv-gwas.tex
@@
 	\begin{center}
-	\includegraphics[width = \linewidth]{figures/simulation_setup.png}
+	\includegraphics[width = \linewidth]{simulation_setup.png}
 	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{figures/simulation_results}
+		\includegraphics[width=\linewidth]{simulation_results}
 		\caption{Results from the simulation study.
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{figures/example_simulation_trait_values.pdf}
+		\includegraphics[width=\linewidth]{example_simulation_trait_values.pdf}
 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{figures/tip_trait_correlation.pdf}
+    \includegraphics[width = \linewidth]{tip_trait_correlation.pdf}
 	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{figures/application_experimental_setups.pdf}
+    \includegraphics[width = \linewidth]{application_experimental_setups.pdf}
 	\caption{A high-level schematic of the experimental setup for the two application datasets.
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{figures/gwas_results}
+		\includegraphics[width=\linewidth]{gwas_results}
 		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{figures/gwas_results_xanthomonas_maxmaf.pdf}
+    \includegraphics[width = \linewidth]{gwas_results_xanthomonas}
 	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
@@
 
-\bibliographystyle{manuscript/my_natbib}
-\bibliography{manuscript/references_synced}
+\bibliographystyle{my_natbib}
+\bibliography{references_synced}
 
@@
 \begin{center}
-\includegraphics[width = 0.7\linewidth]{figures/sigma_vs_alpha.png}
+\includegraphics[width = 0.7\linewidth]{sigma_vs_alpha.png}
 	\caption{Relationship between the selection strength parameter $\alpha$ and the intensity of stochastic evolutionary fluctuations parameter $\sigma$ at two different heritability ($H^2$) values in the  simulation scheme.
@@
 \begin{center}
-\includegraphics[width = 0.7\linewidth]{figures/spvl_calculation_comparison.png}
+\includegraphics[width = 0.7\linewidth]{spvl_calculation_comparison.png}
 	\caption{A comparison of different ways to calculate spVL based on viral load measurements provided by the SHCS.
@@
 	\centering
-		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates.png}
+		\includegraphics[width=0.7\linewidth]{poumm_parameter_estimates.png}
 		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on HIV-1 spVL data from the SHCS. We ran two different MCMC chains to ensure the estimates converged.}
@@
 \begin{center}
-\includegraphics[width = \linewidth]{figures/spvl_on_tree.png}
+\includegraphics[width = \linewidth]{spvl_on_tree.png}
 	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}
@@
 \begin{center}
-\includegraphics[width = 0.462\linewidth]{figures/trait_values.png}
+\includegraphics[width = 0.462\linewidth]{trait_values.png}
 	\caption{A comparison of measured (calculated) spVL values versus our estimated non-pathogen effect on spVL for each SHCS cohort member used in the study.
@@
 	\centering
-	\includegraphics[width=\linewidth]{figures/qq_plots.png}
+	\includegraphics[width=\linewidth]{qq_plots.png}
 	\caption{Quartile-quartile plots from HIV-1 spVL association tests. The dashed green line shows the y = x line.}
@@
 \begin{center}
-\includegraphics[width = \linewidth]{figures/qdr_on_tree_arboricola.png}
+\includegraphics[width = \linewidth]{qdr_on_tree_arboricola.png}
 	\caption{Inferred \emph{X. arboricola} phylogeny with tips colored by (A) calculated QDR, (B) estimated non-pathogen effects on QDR and (C) estimated pathogen effects on QDR.}
@@
 	\centering
-		\includegraphics[width=0.7\linewidth]{figures/poumm_parameter_estimates_arboricola.png}
+		\includegraphics[width=0.7\linewidth]{poumm_parameter_estimates_arboricola.png}
 		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on \emph{A.
@@
 \begin{center}
-\includegraphics[width = 0.462\linewidth]{figures/trait_values_arboricola.png}
+\includegraphics[width = 0.462\linewidth]{trait_values_arboricola.png}
 	\caption{A comparison of measured (calculated) mean QDR values across all \emph{A.
@@
 \begin{center}
-\includegraphics[width = \linewidth]{figures/qq_plots_arabidopsis_xanthomonas.png}
+\includegraphics[width = \linewidth]{qq_plots_arabidopsis_xanthomonas.png}
 	\caption{Quartile-quartile plots from \emph{A. thaliana}-\emph{X. arboricola} QDR association tests. The dashed blue line shows the y = x line.}
@@
 \begin{figure}[H]
-	\includegraphics[width=\linewidth]{figures/host_genotype_pca.png}
+	\includegraphics[width=\linewidth]{host_genotype_pca.png}
 	\caption{SHCS individuals and HapMap3 individuals plotted along the top three principle components of genetic variation.

### 2022-07-19 manuscript_proof/manuscript.tex
Update on Overleaf.

@@
 	\begin{center}
-	\includegraphics[width = \linewidth]{simulation_setup.png}
+	\includegraphics[width = \linewidth]{fig1_Stadler_MBE-21-1179.pdf}
 	\caption{A high-level schematic of our phylogenetic Ornstein-Uhlenbeck mixed model (POUMM)-based simulation framework in the context of HIV-1 set-point viral load (spVL).
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{simulation_results}
+		\includegraphics[width=\linewidth]{fig2_Stadler_MBE-21-1179.pdf}
 		\caption{Results from the simulation study.
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{example_simulation_trait_values.pdf}
+		\includegraphics[width=\linewidth]{fig3_Stadler_MBE-21-1179.pdf}
 		\caption{Simulated data from two evolutionary scenarios where a phylogenetic correction to trait values improves genome-wide association study (GWAS) power (right side) and where it does not (left side).
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{tip_trait_correlation.pdf}
+    \includegraphics[width = \linewidth]{fig4_Stadler_MBE-21-1179.pdf}
 	\caption{Correlations between trait values in pairs of tips in four simulated scenarios.
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{application_experimental_setups.pdf}
+    \includegraphics[width = \linewidth]{fig5_Stadler_MBE-21-1179.pdf}
 	\caption{A high-level schematic of the experimental setup for the two application datasets.
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{gwas_results}
+		\includegraphics[width=\linewidth]{fig6_Stadler_MBE-21-1179}
 		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{gwas_results_xanthomonas}
+    \includegraphics[width = \linewidth]{fig7_Stadler_MBE-21-1179}
 	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.
@@
 
-\newpage
-
-%Additional ideas for supplemental materials:
-%\begin{itemize}
-%	\item Table of genotype samples by project (genotyping chip, number of samples, spVL distribution)
-%	\item Link to github with all the code used to run simulations and generate figures
-%	\item Example spVL plots
-%	\item Proof of robustness of method to inferred tree -- NA because I just re-ran the _whole_ thing and it's not that different
-%\end{itemize}
-
-\newpage
-\startsetups
-    \setcounter{figure}{0}
-    \renewcommand{\thefigure}{S\arabic{figure}}
-    \setcounter{table}{0}
-    \renewcommand{\thetable}{S\arabic{table}}
-\stopsetups
-\section*{Supplemental Material}\label{supp}
-\beginsupplement
-\setcounter{page}{1}
-
-\begin{doublespace}
-
-\subsection*{Expected results from simulations}
-
-Here we show the root mean square error (RMSE) of the scaled trait value for each individual, $z_i - \bar{z}$ as an estimate for the host part of the trait for each individual, ${h_i}$, should be $\approx$ 0.74 in our simulation scheme.
-First we write the expression for the RMSE:
-
-\begin{equation}
-    RMSE = \sqrt{\frac{\sum_i^N(z_i - \bar{z} - h_i)^2}{N}}
-\end{equation}
-
-Note that under our simulation setup, $z_i - \bar{z}$ differs from $h_i$ due to the individual pathogen effect $g_i$ and environmental effect $e_i$.
-So the term inside the square root equals the combined variance of these two effects:
-
-\begin{equation}
-    RMSE = \sqrt{\sigma^2_{g} + \sigma^2_e}
-\end{equation}
-
-We can calculate the variance due to these two effects because the total variance in spVL $\sigma^2_z$, and the fraction of the total variance due to host genetic effects, $\sigma^2_{h}$, are fixed parameters in our simulation scheme.
-
-\begin{equation}
-\begin{split}
-    \sigma^2_{h} + \sigma^2_{g} + \sigma^2_e = \sigma^2_z \\
-    0.25 * \sigma^2_{z} + \sigma^2_{g} + \sigma^2_e = \sigma^2_z \\
-    \sigma^2_{g} + \sigma^2_e = 0.75 * \sigma^2_z \\
-    \sigma^2_{g} + \sigma^2_e = 0.75 * 0.73 \\
-    \sigma^2_{g} + \sigma^2_e = 0.55
-\end{split}
-\end{equation}
-
-Therefore, we can expect the RMSE for $z_i - \bar{z}$ as an estimate for ${h_i}$ to be around $\sqrt{0.55} \approx 0.74$.
-
-\newpage
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.7\linewidth]{sigma_vs_alpha.png}
-	\caption{Relationship between the selection strength parameter $\alpha$ and the intensity of stochastic evolutionary fluctuations parameter $\sigma$ at two different heritability ($H^2$) values in the  simulation scheme.
-$\sigma$ was determined as a function of $\alpha$ and $H^2$ under the POUMM (function given in Table \ref{tab:sim-params}).}
-	\label{fig:sigma-vs-alpha}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\begin{tikzpicture}
-
-	% Variance 
-	\begin{scope}[xshift = 0cm, yshift = 0cm]
-	\node[anchor = west] (alabel) at (-1, 4) {\textbf{A} Variance partitioning};
-	\node[constnode] (H2h) at (0, 0) {0.25};
-	\node[constnode] (varz) at ($(H2h) + (1.8, 0)$) {$\sigma^2_z$};
-	\node[constnode] (H2) at ($(H2h) + (3.6, 0)$) {\textcolor{red}{$H^2_{\bar{t}}$}};
-	\node[detnode] (varh) at ($(H2h) + (0.9, 1.5)$) {$\sigma^2_{h}$};
-	\node[detnode] (varv) at ($(varz) + (0.9, 1.5)$) {\tiny $\sigma^2_{g}(\bar{t})$};
-	\node[detnode] (vare) at ($(varh) + (0.9, 1.5)$) {$\sigma^2_{e}$};
-	\draw[taro] (H2h) -- (varh);
-	\draw[taro] (varz) -- (varh);
-	\draw[taro] (varz) -- (varv);
-	\draw[taro] (H2) -- (varv);
-	\draw[taro] (varh) -- (vare);
-	\draw[taro] (varv) -- (vare);
-	\draw[taro] (varz) -- (vare);
-	\end{scope}
-
-	% Virus
-	\begin{scope}[xshift = 6cm, yshift = -0.5cm]
-	\node[anchor = west] (blabel) at (-1, 4.5) {\textbf{B} Generating pathogen effects};
-	\node[constnode] (theta) at (0, 1.5) {$\theta$};
-	\node[constnode] (g0) at (1.8, 1.5) {$g_{0}$};
-	\node[constnode] (alpha) at (3.6, 1.5) {\textcolor{red}{$\alpha$}};
-	\node[detnode] (sigma) at (5.4, 1.5) {$\sigma$};
-	\node[constnode] (tree) at (7.2, 1.5) {$\Psi$};
-	\node[detnode] (varv) at (3.6+0.9, 0) {\tiny $\sigma^2_{g}(\bar{t})$};
-	\node[constnode] (tbar) at (3.6+0.9+1.8, 0) {$\bar{t}$};
-	\node[snode] (v) at (3.6, 3.25) {$\vec{g}$};
-	\draw[taro] (theta) -- (v);
-	\draw[taro] (g0) -- (v);
-	\draw[taro] (alpha) -- (v);
-	\draw[taro] (sigma) -- (v);
-	\draw[taro] (tree) -- (v);
-	\draw[taro] (varv) -- (sigma);
-	\draw[taro] (tbar) -- (sigma);
-	\draw[taro] (tbar) -- (tree);
-	\draw[taro] (alpha) -- (sigma);
-	\node[white] (varpic) at (3.6+0.9, -1.75) {};
-	\draw[taro] (varpic) -- (varv);
-	\node (vdist) at ($(v) + (1.8, 0.3)$) {Gaussian};
-	\draw (v) -- (vdist);
-	\draw[decoration={brace, mirror, raise=18pt}, sloped, decorate] (7.2, 1.5 - .5) -- node[below = 18pt] {OU parameters} (7.2, 1.5 + .5);
-	\end{scope}
-
-	% Host 
-	\begin{scope}[xshift = 0cm, yshift = -7.5cm]
-	\node[anchor = west] (clabel) at (-1, 5.55) {\textbf{C} Generating host effects};
-	\node[constnode] (d) at (1.8, 1.5) {2};
-	\node[detnode] (p) at (3.6, 1.5) {$p$};
-	\node[snode] (G) at (1.8, 3) {$G$};
-	\node[rectangle, dashed, very thick, inner sep=4mm, draw=black!100, fit = (G) ] (Gplate) {};
-	\node[anchor=north west, inner sep=3pt] at (Gplate.north west) {\tiny $N\times M$};
-	\node[detnode] (varh) at (3.6, 0) {$\sigma^2_{h}$};
-	\node[constnode] (delta) at (3.6, 3) {$\delta$};
-	\node[detnode] (h) at (3.6, 4.5) {$\vec{h}$};
-	\node[constnode] (M) at (5.4, 3) {$M$};
-	\draw[taro] (d) -- (G);
-	\draw[taro] (p) -- (G);
-	\draw[taro] (varh) -- (p);
-	\draw[taro] (delta) -- (p);
-	\draw[taro] (M) -- (p);
-	\draw[taro] (M) -- (h);
-	\draw[taro] (delta) -- (h);
-	\draw[taro] (G) -- (h);
-	\node[white] (varpic) at (3.6, -1.5) {};
-	\draw[taro] (varpic) -- (varh);
-	\node (hdist) at ($(h) + (2.7, 0.3)$) {$\approx$ Gaussian};
-	\draw (h) -- (hdist);
-	\node (gdist) at ($(G) + (-2, 0.6)$) {Binomial};
-	\draw (G) -- (gdist);
-	\end{scope}
-
-	% Environment
-	\begin{scope}[xshift = 9cm, yshift = -6.5cm]
-	\node[anchor = west] (dlabel) at (-1, 3) {\textbf{D} Generating environmental effects};
-	\node[detnode] (vare) at (0, 0) {$\sigma^2_{e}$};
-	\node[constnode] (zero) at ($(vare) + (1.8, 0)$) {0};
-	\node[snode] (e) at ($(vare) + (1.8/2, 1.5)$) {$\vec{e}$};
-	\draw[taro] (zero) -- (e);
-	\draw[taro] (vare) -- (e);
-	\node[white] (varpic) at (0, -1.5) {};
-	\draw[taro] (varpic) -- (vare);
-	\node (edist) at ($(e) + (1.8, 0.3)$) {Gaussian};
-	\draw (e) -- (edist);
-	\end{scope}
-
-	\end{tikzpicture}
-	\caption{A graphical model representation of our simulation scheme, following the recommendations in \cite{Hohna2014}.
-Variables in solid squares are constants, with the two master control variables that we vary from simulation to simulation highlighted in red.
-Variables in solid circles are realizations of random variables and variables in dashed circles are determined as a function of other variables.
-Arrows represent dependencies among variables and the dashed square represents repetition.
-All parameters are defined in Table \ref{tab:sim-params}, as well as the values or expressions used for them.
-(A) shows how the variance in the simulated environmental effect $\sigma^2_{e}$ is smaller if the master pathogen heritability value $H^2_{\bar{t}}$ is higher and vice-versa.
-(B) shows the OU parameters and the pathogen phylogeny, which generate the Gaussian-distributed pathogen effects.
-The OU parameters $\theta$ and $g_0$ are fixed, whereas $\sigma$ is a deterministic function of the variance in the pathogen effect and the value of $\alpha$.
-In other words, we use $\sigma$ to maintain the desired pathogen heritability while varying $\alpha$.
-(C) shows how host genotypes are drawn to generate host effects.
-The host genotype matrix $G$ contains the number of copies (0, 1, or 2) for each of $M$ causal variants with effect size $\delta$.
-We assume half the variants have a positive effect and half have a negative effect.
-The allele frequency $p$ for the causal variants set so that we achieve the desired variance in the host effects.
-(D) shows that the environmental effect is drawn from a Gaussian distribution with mean zero and variance as determined in part (A).}
-	\label{fig:sim-design}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.7\linewidth]{spvl_calculation_comparison.png}
-	\caption{A comparison of different ways to calculate spVL based on viral load measurements provided by the SHCS.
-The stricter filtering excludes all measurements possibly $<$ 6 months after infection and after treatment or AIDS, whereas the more lenient filtering excludes only measurements after treatment.
-We used the lenient filter, mean measurement values because these correlate well with the values from the stricter filter but allow us to retain many more individuals from the cohort for our study.}
-	\label{fig:spvl-calc-comparison}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-		\includegraphics[width=0.7\linewidth]{poumm_parameter_estimates.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on HIV-1 spVL data from the SHCS. We ran two different MCMC chains to ensure the estimates converged.}
-\label{fig:poumm-parameters}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{spvl_on_tree.png}
-	\caption{Inferred HIV-1 $pol$ gene phylogeny with tips colored by (A) calculated spVL, (B) estimated non-pathogen effects on spVL and (C) estimated pathogen effects on spVL.}
-	\label{fig:spvl-on-tree}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.462\linewidth]{trait_values.png}
-	\caption{A comparison of measured (calculated) spVL values versus our estimated non-pathogen effect on spVL for each SHCS cohort member used in the study.
-The histograms show the marginal distribution of each value across the individuals.}
-	\label{fig:trait-comparison}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{qq_plots.png}
-	\caption{Quartile-quartile plots from HIV-1 spVL association tests. The dashed green line shows the y = x line.}
-	\label{fig:qq-plots}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{qdr_on_tree_arboricola.png}
-	\caption{Inferred \emph{X. arboricola} phylogeny with tips colored by (A) calculated QDR, (B) estimated non-pathogen effects on QDR and (C) estimated pathogen effects on QDR.}
-	\label{fig:qdr-on-tree}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\centering
-		\includegraphics[width=0.7\linewidth]{poumm_parameter_estimates_arboricola.png}
-		\caption{Posterior distributions compared to the prior for POUMM parameter estimates based on \emph{A.
-thaliana}-\emph{X.
-arboricola} data.
-We ran two different MCMC chains to ensure the estimates converged.}
-\label{fig:poumm-parameters-qdr}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = 0.462\linewidth]{trait_values_arboricola.png}
-	\caption{A comparison of measured (calculated) mean QDR values across all \emph{A.
-thaliana} accession pairings and replicates  versus our estimated non-pathogen effect on mean QDR for each \emph{X.
-arboricola} pathogen strain.
-The histograms show the marginal distribution of each value across the strains.
-The Pearson correlation coefficient is 1.}
-	\label{fig:trait-comparison-qdr}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-\begin{center}
-\includegraphics[width = \linewidth]{qq_plots_arabidopsis_xanthomonas.png}
-	\caption{Quartile-quartile plots from \emph{A. thaliana}-\emph{X. arboricola} QDR association tests. The dashed blue line shows the y = x line.}
-	\label{fig:qq-plots-qdr}
-	\end{center}
-\end{figure}
-
-\begin{figure}[H]
-	\includegraphics[width=\linewidth]{host_genotype_pca.png}
-	\caption{SHCS individuals and HapMap3 individuals plotted along the top three principle components of genetic variation.
-Points with black borders are within the thresholds used to select individuals of likely European ancestry.}
-	\label{fig:PCA}
-\end{figure}
-
-\end{doublespace}  % don't doublespace tables
-
-\newpage
-
-\begin{table}[H]
-	\caption{Simulation model parameters. For a full graphical model representation of the simulation scheme, including how these parameters are related, see Figure \ref{fig:sim-design}.}
-	\begin{tabularx}{\linewidth}{p{1.5cm}p{6.5cm}l} \toprule 
-		Variable & Expression & Definition \\ \midrule 
-		$\sigma^2_z$ &  0.73 log copies$^{2}$/mL$^{2}$ & Total spVL variance \\ 
-		$H^2_h$ & 0.25 & Host heritability of spVL \\
-		$H^2_{\bar{t}}$ & varied & Pathogen heritability of spVL at $\bar{t}$ \\
-		$\sigma^2_{h}$ & $\sigma_{h}^2 = 0.25*\sigma_z^2$ & Variance in host part of spVL \\
-		$\sigma^2_{g}(\bar{t})$ & $\sigma^2_{g}(\bar{t}) = H^2_{\bar{t}}*\sigma_z^2$ & Variance in pathogen part of spVL at $\bar{t}$ \\ 
-		$\sigma^2_{e}$ & $\sigma_{e}^2 = \sigma_z^2 - \sigma^2_{g} - \sigma^2_{h}$ & Variance in environmental part of spVL \\ 
-		$\bar{t}$ & 0.14 substitutions site$^{-1}$ yr$^{-1}$ & Mean root-tip time in pathogen phylogeny \\ 
-		$\bm{g}$ & $\bm{g} \sim Norm(\bm{\mu}_{OU}, \boldsymbol{\Sigma}_{OU})$ & Pathogen part of spVL for all individuals \\ 
-		$\theta$ & 4.5 log copies/mL & Optimal spVL value \\
-		$g_{0}$ & 4.5 log copies/mL & $g$ at the root of the phylogeny  \\ 
-		$\alpha$ & varied & Selection strength of OU process \\ 
-		$\sigma$ & $\sigma = \sqrt{\frac{2\alpha\sigma^2_{g}(\bar{t})}{1 - exp(-2\alpha\bar{t})}}$ & Time-unit standard deviation of OU process \\ 
-		$\Psi$ & branch lengths $\sim Exp(\bar{t})$ & Pathogen phylogeny \\ 
-		$h_i$ & $h_i = \delta \sum_{j = 1}^{j = M/2}{G_{ij}} - \delta \sum_{j = M/2}^{j = M}{G_{ij}}$ & Host part of spVL for individual $i$ \\ 
-		$G_{N \times M}$ & \makecell{$G_{ij} \sim Binom(2, p)\ \\ \forall i \in {1...N}, \forall j \in {1...M}$} & Host genotype matrix \\
-		$p$ & $p = \frac{1}{2} - \sqrt{\frac{1}{4} - \frac{H_{h}^2\sigma^2_z}{2\delta^2M}}$ &  Host variant allele frequency \\ 
-		$\delta$ & 0.2 & Host variant effect size \\ 
-		$M$ & 20 & Number of causal host variants \\ 
-		$e_i$ & $ e_i \sim Norm(0, \sigma^2_{e})$ & Environmental part of spVL for individual $i$ \\ 
-		$N$ & 500 & Number of simulated samples \\ \bottomrule 
-	\end{tabularx}
-	\label{tab:sim-params}
-\end{table}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for spVL based on SHCS data. HPD = Highest posterior density.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Parameter & Posterior mean & 95\% HPD \\ \hline 
-	${g_{0}}$ & 4.23 & (1.72, 6.71) \\
-	$\theta$ & 4.47 & (4.37, 4.58) \\
-	$\sigma$ & 5.25 & (2.37, 7.9) \\
-	$\alpha$ & 57.65 & (19.49, 95.2) \\
-	$\sigma_\epsilon$ & 0.54 & (0.43, 0.65) \\ 
-	$H^2_{\bar{t}}$ & 0.45 & (0.24, 0.67) \\ \hline
-	\end{tabularx}
-	\label{tab:POUMM-parameters-our-estimates}
-\end{table}
-
-% latex table generated in R 4.1.1 by xtable 1.8-4 package
-% Sat Oct 30 10:31:16 2021
-\begin{table}[ht]
-\caption{Effect size and p-values from the top most strongly associated variants in the $\emph{CCR5}$ and MCR regions from each of the two GWAS performed in our study.
-``Standard'' means the GWAS with standard spVL trait values and ``Corrected'' means the GWAS with the estimated non-pathogen part of the trait.
-Entries above the dividing line are the top-associated variants from the ``Standard'' GWAS and entries below the dividing line are the top-associated variants from the ``Corrected'' GWAS.
-Many entries overlap between the two.}
-\begin{tabularx}{\linewidth}{XXXXXXX}
-  \hline
-    Region & Position & Variant & Standard effect size & Standard p-value & Corrected effect size & Corrected p-value \\ 
-  \hline
-    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46540932 & rs10865942 & -0.16 & $8.4 \times 10^{-9}$ & -0.081 & $4 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
-    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
-    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
-    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
-    MHC & 31391401 & rs4418214 & -0.34 & $2.4 \times 10^{-14}$ & -0.18 & $2.5 \times 10^{-12}$ \\ 
-    MHC & 31400137 & rs138130755 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31400705 & rs138117378 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31402358 & rs148792134 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    MHC & 31409677 & rs140991764 & -0.46 & $1 \times 10^{-12}$ & -0.26 & $2.6 \times 10^{-12}$ \\ 
-    \hline
-    $\emph{CCR5}$ & 46531144 & rs9845968 & -0.16 & $5.6 \times 10^{-9}$ & -0.083 & $1.2 \times 10^{-7}$ \\ 
-    $\emph{CCR5}$ & 46537849 & rs867620 & -0.16 & $3.2 \times 10^{-9}$ & -0.085 & $6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46539864 & rs11130092 & -0.16 & $1.1 \times 10^{-9}$ & -0.087 & $2.6 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46541147 & rs7430431 & -0.17 & $9.2 \times 10^{-10}$ & -0.088 & $2.3 \times 10^{-8}$ \\ 
-    $\emph{CCR5}$ & 46556835 & rs6808142 & 0.15 & $8.3 \times 10^{-8}$ & 0.082 & $3.3 \times 10^{-7}$ \\ 
-    MHC & 31274380 & rs9264942 & -0.21 & $4.5 \times 10^{-13}$ & -0.12 & $3.7 \times 10^{-13}$ \\ 
-    MHC & 31321919 & rs1055821 & -0.33 & $9.4 \times 10^{-13}$ & -0.19 & $1.4 \times 10^{-12}$ \\ 
-    MHC & 31367874 & rs111281598 & -0.37 & $1.5 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
-    MHC & 31376266 & rs73400361 & -0.37 & $1.4 \times 10^{-12}$ & -0.22 & $2.1 \times 10^{-12}$ \\ 
-    MHC & 31380034 & rs112243036 & -0.32 & $9.9 \times 10^{-16}$ & -0.17 & $3.7 \times 10^{-14}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:gwas-top-associations}
-\end{table}
-
-\begin{table}[H]
-\centering
-\caption{Results comparison using three different approximate maximum likelihood HIV-1 phylogenies (see Materials and Methods).}
-\begin{tabularx}{\linewidth}{p{1.8cm}p{2cm}p{1.8cm}p{2cm}p{1.8cm}p{2cm}p{1.8cm}} % {lrrrrrr} {llclcccc}
-\hline
-& \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\main tree}} & \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\tree 2}} & \multicolumn{2}{l}{\makecell{Estimated\\non-pathogen\\part of trait\\tree 3}} \\
-  \hline
- Variant & Effect size & p-value & Effect size & p-value & Effect size & p-value \\
-  \hline
-    rs59440261 & -0.22 & $2.6 \times 10^{-10}$ & -0.22 & $1.5 \times 10^{-10}$ & -0.24 & $1.4 \times 10^{-10}$ \\ 
-    rs1015164 & 0.078 & $8.5 \times 10^{-6}$ & 0.076 & $9.5 \times 10^{-6}$ & 0.083 & $9 \times 10^{-6}$ \\ 
-   \hline
-\end{tabularx}
-\label{tab:comp-gwas-mclaren-snps-alt-trees}
-\end{table}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for QDR based on \emph{A. thaliana}-\emph{X. arboricola} data. HPD = Highest posterior density.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Parameter & Posterior mean & 95\% HPD \\ \hline 
-	${g_{0}}$ & 0.40 & (0.01, 0.78) \\
-	$\theta$ & 0.39 & (0.30, 0.49) \\
-	$\sigma$ & 0.03 & (0.0, 0.06) \\
-	$\alpha$ & 0.03 & (0.0, 0.05) \\
-	$\sigma_\epsilon$ & 0.18 & (0.08, 0.27) \\ 
-	$H^2_{\bar{t}}$ & 0.33 & (0.0, 0.77) \\ \hline
-	\end{tabularx}
-	\label{tab:POUMM-parameters-our-estimates-qdr}
-\end{table}
-
-\begin{table}[H]
-\caption{Summary statistics for log spVL in previously sampled populations.
-$\bar{z}$ is average spVL (log copies/mL) and $\sigma^2_z$ is variance in measured spVL ( log $\textrm{copies}^2/\textrm{mL}^2$).
-Values from \protect\citep{Blanquart2017, Mitov2018} are empirical; values from \protect\citep{Bonhoeffer2015} were estimated by fitting a normal distribution to the data.}
-	\begin{tabularx}{\linewidth}{XXX} \hline 
-	Measurement & Value & Reference \\ \hline 
-	$\bar{z}$ & $\approx$ 4.5 & \citet{Mitov2018} \\
-	$\bar{z}$ &  4.4 & \citet{Blanquart2017} \\
-	$\bar{z}$ & $\approx$ 4.5 & \citet{Bonhoeffer2015} \\
-	$\sigma^2_z$ &  0.73 &  \citet{Mitov2018} \\
-	$\sigma^2_z$ &  0.50 & \citet{Blanquart2017} \\
-	$\sigma^2_z$ & $\approx$ 0.5 & \citet{Bonhoeffer2015} \\ \hline 
-	\end{tabularx}
-	\label{tab:spVLmeasurements}
-\end{table}
-%\todo{I guessed at bar(z) from figure in Mitov2018, search supplement for exact}
-
-\begin{table}[H]
-\caption{POUMM parameter estimates for spVL from previous studies.}
-	\begin{tabularx}{\linewidth}{llXXX}
-	\hline 
-	Parameter & Value (Uncertainty) & Reference & Notes \\
-	\hline 
-	$g_0$ & 5.54 (4.04 - 7.25) & \citet{Mitov2018} & 8,483 UK HIV cohort individuals, $pol$ tree \\
-	$\theta$ & 4.45 (4.41 - 4.49) & \citet{Mitov2018} & \\
-	$\theta$ &  4.0 (1.6 - 4.)) & \citet{Bertels2018} & 3,036 SHCS individuals, $pol$ tree \\
-	$\theta$ & 4.1 (3.5 - 4.9) & \citet{Blanquart2017} & 1,581 subtype B individuals from Europe, whole genome tree \\
-	$\alpha$ & 28.78 (16.64 - 46.93) & \citet{Mitov2018} & \\
-	$\alpha$ & 32.7 (0.03 - 57.6) & \citet{Bertels2018} & \\
-	$\alpha$ & 7.6 (1.2 - 10) & \citet{Blanquart2017} & **limited $\alpha$ to $\le$ 10 \\
-	$\sigma$ & 2.97 (1.95 - 4.37) & \citet{Mitov2018} & \\
-	$\sigma$ & 1.3 (0.66 -1.87) &  \citet{Blanquart2017} & \\
-	$\sigma_e$ & 0.77 (0.73, 0.8) & \citet{Mitov2018} & \\
-	$\sigma_e$ & 0.61 (0.54, 0.65) &  \citet{Blanquart2017} & \\
-	\hline 
-	\end{tabularx}
-	\label{tab:POUMMparams-prev-estimated}
-\end{table}
-
-\begin{table}[H]
-\caption{Number of samples for HIV-1 spVL GWAS after sequential filtering steps.}
-    \begin{tabularx}{\linewidth}{XX}
-    \hline
-    Sample filter & Number of samples remaining \\
-    \hline
-    Subtype B pol sequences & 1516 \\
-    With paired spVL measurement & 1516 \\
-    $>$ 750 characters in sequence & 1493 \\
-    Individual is of European ancestry & 1396 \\
-    Kinship coefficient $>$ 0.09375 & 1392 \\ 
-    \hline
-    \end{tabularx}
-    \label{tab:sample-filtering}
-\end{table}
-
-\begin{table}[H]
-\caption{Number of variants for HIV-1 spVL GWAS after sequential filtering steps.}
-	\begin{tabularx}{\linewidth}{XX}
-	\hline
-	Variant filter & Number of variants remaining \\
-	\hline 
-	Raw data & 76979521 \\
-	Missing genotype rate $>$ 0.05 & 11590002 \\
-	Hardy-Weinburg exact test p-value $< 5 \times 10^{-5}$ & 11589246 \\
-	Minor allele frequency $<$ 0.01 & 6228626 \\ \hline			
-	\end{tabularx}
-	\label{tab:variant-filtering}
-\end{table}
-
-\begin{table}[H]
-\caption{Number of variants for \emph{A.
-thaliana} QDR GWAS after sequential filtering steps.
-The last entry lists variants without GWAS p-values because PLINK assessed the correlation between predictor variables (the variant and the top 5 principle components of host genetic variation are predictors) to be too strong.
-This did not occur in the HIV-1 spVL GWAS.}
-	\begin{tabularx}{\linewidth}{XX}
-	\hline
-	Variant filter & Number of variants remaining \\
-	\hline 
-	Raw data & 12883854 \\
-	Bi-allelic variants & 11769920 \\
-	Minor allele frequency $<$ 0.1 & 1743952 \\
-	NA p-value (too-high covariate correlation) & 1070541 \\ \hline			
-	\end{tabularx}
-	\label{tab:variant-filtering-qdr}
-\end{table}
-
-\stopsection
 \end{document}

### 2022-07-19 manuscript_proof/manuscript.tex
Update on Overleaf.

@@
 
-% Set up supplemental figure numbering
-\newcommand{\beginsupplement}{\setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{S\arabic{table}}}
-
 \begin{document}
-
-% \begin{flushright} % for MBE submission
-% Submission type: Article\\
-% Section: Methods
-% \end{flushright}
 
@@
 For the simulation, we considered a range of pathogen heritability parameter values $H^2$, from 15 to 75\%, and a range of selection strength parameters values $\alpha$, from 0.1 to 60 time$^{-1}$.
-The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table \ref{tab:sim-params}).
-As shown in Figure \ref{fig:sigma-vs-alpha}, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization.
+The intensity of stochastic fluctuations parameter $\sigma$ was determined based on $H^2$ and $\alpha$ (a re-arrangement of equation \ref{eq:OUVar}, equation given in Table S1).
+As shown in Figure S1, higher $\alpha$ values correspond to higher $\sigma$ values to maintain constant $H^2$ under this parameterization.
 For each $H^2$ and $\alpha$ value considered in the simulation, we recorded the simulated pathogen part of the trait value for each tip in the phylogeny.
@@
 We followed the method outlined above to estimate the pathogen and non-pathogen effects on spVL for the cohort from these data.
-Figure \ref{fig:spvl-calc-comparison} shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
-We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure \ref{fig:poumm-parameters}, Table \ref{tab:POUMM-parameters-our-estimates}).
+Figure S3 shows the calculated (total) spVL values, which vary between approximately 1 and 6 log copies/mL in the cohort.
+We estimated spVL heritability in this cohort to be 45\% (95\% highest posterior density, HPD, 24 - 67\%) and selection strength to be 58 time$^{-1}$ (95\% HPD 19 - 95) (Figure S4, Table S2).
 To put these values into the context of our simulation study, they are shown as points on Figure \ref{fig:simulationResults}.
 The highest expected correlation in trait values between any two tips in the HIV-1 phylogeny under the POUMM was 0.45.
-However, Figure \ref{fig:spvl-on-tree} shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.
-Finally, figure \ref{fig:trait-comparison} shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
+However, Figure S5 shows that this trait is not obviously phylogenetically structured in the cohort in general, despite high heritability.
+Finally, Figure S6 shows that the estimated non-pathogen effects on spVL correlate quite strongly with total spVL.
 
@@
 Association strength increased for some variants in the MHC, for example, SNP rs9265880 had the greatest increase in significance in the MHC region, from a p-value of $3.5 \times 10^{-07}$ to $7.7 \times 10^{-09}$.
-However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table \ref{tab:gwas-top-associations}).
+However, the top-associated variants in the MHC and \emph{CCR5} regions were consistent regardless of the GWAS response variable used (Table S3).
 Finally, Table \ref{tab:comp-gwas-mclaren-snps} shows how our GWAS results compare for the two top-associated SNPs identified by \citet{McLaren2015}, who performed the largest standard GWAS for HIV spVL to date.
 Effect sizes are smaller with a phylogenetic correction and p-values are slightly increased.
-We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table \ref{tab:comp-gwas-mclaren-snps-alt-trees}).
+We repeated the analysis using three different approximate maximum-likelihood phylogenies and these results were consistent (see Materials and Methods; Table S4).
 In summary, there are no clear patterns that point to new regions of association in the human genome with spVL when we take into account the pathogen phylogeny.
@@
 	\begin{center}
-		\includegraphics[width=\linewidth]{fig6_Stadler_MBE-21-1179}
+		\includegraphics[width=\linewidth]{fig6_Stadler_MBE-21-1179.jpg}
 		\caption{Results from comparative genome-wide association studies (GWAS) on HIV-1 set-point viral load (spVL) data.
@@
 \begin{center}
-    \includegraphics[width = \linewidth]{fig7_Stadler_MBE-21-1179}
+    \includegraphics[width = \linewidth]{fig7_Stadler_MBE-21-1179.jpg}
 	\caption{Results from comparative genome-wide association studies (GWAS) on \emph{A.

### 2022-07-19 manuscript_proof/manuscript.tex
Update on Overleaf.

@@
 The members of the SHCS are:
-Abela I, Aebi-Popp K, Anagnostopoulos A, Battegay M, Bernasconi E, Braun DL, Bucher HC, Calmy A, Cavassini M, Ciuffi A, Dollenmaier G, Egger M, Elzi L, Fehr J, Fellay J, Furrer H, Fux CA, Günthard HF (President of the SHCS), Hachfeld A, Haerry D (deputy of "Positive Council"), Hasse B, Hirsch HH, Hoffmann M, Hösli I, Huber M, Kahlert CR (Chairman of the Mother & Child Substudy), Kaiser L, Keiser O, Klimkait T, Kouyos RD, Kovari H, Kusejko K (Head of Data Centre), Martinetti G, Martinez de Tejada B, Marzolini C, Metzner KJ, Müller N, Nemeth J, Nicca D, Paioni P, Pantaleo G, Perreau M, Rauch A (Chairman of the Scientific Board), Schmid P, Speck R, Stöckle M (Chairman of the Clinical and Laboratory Committee), Tarr P, Trkola A, Wandeler G, Yerly S.
+Abela I, Aebi-Popp K, Anagnostopoulos A, Battegay M, Bernasconi E, Braun DL, Bucher HC, Calmy A, Cavassini M, Ciuffi A, Dollenmaier G, Egger M, Elzi L, Fehr J, Fellay J, Furrer H, Fux CA, Günthard HF (President of the SHCS), Hachfeld A, Haerry D (deputy of "Positive Council"), Hasse B, Hirsch HH, Hoffmann M, Hösli I, Huber M, Kahlert CR (Chairman of the Mother \& Child Substudy), Kaiser L, Keiser O, Klimkait T, Kouyos RD, Kovari H, Kusejko K (Head of Data Centre), Martinetti G, Martinez de Tejada B, Marzolini C, Metzner KJ, Müller N, Nemeth J, Nicca D, Paioni P, Pantaleo G, Perreau M, Rauch A (Chairman of the Scientific Board), Schmid P, Speck R, Stöckle M (Chairman of the Clinical and Laboratory Committee), Tarr P, Trkola A, Wandeler G, Yerly S.
 
-The Swiss HIV Cohort Study is supported by the Swiss National Science Foundation (grant #201369), by SHCS project #858 and by the SHCS research foundation.
+The Swiss HIV Cohort Study is supported by the Swiss National Science Foundation (grant \#201369), by SHCS project \#858 and by the SHCS research foundation.
 Furthermore, the SHCS drug resistance database is supported by the Yvonne Jacob Foundation (to HFG).

### 2022-07-21 manuscript_proof/manuscript.tex
Update on Overleaf.

@@
 More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure \citep{Power2017}.
-Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in \cite{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \cite{collins_phylogenetic_nodate}.
+Microbial GWAS-specific phylogenetic methods to account for population structure include explicitly testing for lineage-specific effects as in \cite{Earle2016} and modified association tests that account for phylogenetic relationships amongst samples as in \cite{collins_phylogenetic_nodate}.
 These approaches are designed to quantify genetic effects from one organism on a trait.
